### 2016-04-12 hoist.tex
Added points

\begin{section}

1. Code hoisting in adjacent blocks in a diamond/triangle like structure.
2. O(n) Algorithm.
3. Hoisting equivalent computations in non-sibling BBs.
4. Hoisting across switch blocks.
5. Hoisting to improve PRE.
6. Hoisting to reduce critical path in OoO execution.
7. Ranking/Wavefront algorithm to hoist across switch blocks.
\end{section}

### 2016-04-14 hoist.tex
adding concepts for GCM for reducing the liveness

@@
 7. Ranking/Wavefront algorithm to hoist across switch blocks.
+
+Global code motion.
+ bb1: b = ...
+ bb2: c = ...
+ bb3: a = b + c (to be moved to b5)
+ bb4: ... = a
+ old-live-range = distance(bb1,bb3) + distance(bb2,bb3) + distance (bb3,bb4)
+ new-live-rance = distance(bb1,bb5) + distance(bb2,bb5) + distance (bb5,bb4)
+ distance(bbx, bby) = total instruction count in the path from bbx to bby
+ If the new live-range is less than the old one it will be a good candidate
+ for gcm. When both the ranges will be same:
+  - It is a simple copy of kind a = b.
+  - One of the operands is not a Instruction/Register.
+  - In these cases we need to have some heuristic or we can ignore them.
+
+ The decision for bb5 can be made by whether hoist/sink is beneficial.
+
+ Loads can be moved early as long as there is load available in each branch
+ or there is already a load/store from/to the same underlying object.
+ Stores can be moved up if there is a store/load to/from the same object.
+ The idea is to establish that the object has memory allocated to it.
+ Moving load/store together may help with locality of reference.
+
+ PRE via Global Value Numbering
+ Remove redundant instructions which are:
+  - Already available in the dominator.
+  - Are in one of the sibling branches, i.e., the instruction is used at
+    a point which shares a common dominator where all the use-operands
+    are available.
+ In some cases it is possible to generate redundancy by restructuring the code
+ (Ref. Ras Bodik), but that I'll leave for the next iteration of this patch.
+
 \end{section}

### 2016-04-14 hoist.tex
adding hoist vs sink

@@
 
+ Hoisting also reduces critical path length of execution in out of order machines (but not in sequential machines), by exposing ILP before the conditional where the instruction was hoisted.
+This feature has already been identified in the previous patch which improves ray-trace.
+
+ Concerns:
+ Safety of load instructions to be checked. We cannot hoist loads until all the paths in the parent BBs have the same load. This is to comply with C semantics, although I'm not sure about this yet.
+ Finding a cost model in case optimistic hoisting does not give desired results.
+
+It seems, code hoisting is beneficial in cases even if the computations are not redundant.
+Say c = f(a, b) is an instruction.
+Hoisting up will reduce the liveness of registers a and b, but will only increase the liveness of c.
+So we gain 2:1 even when the computation is not redundant.
+For loads this is not the case because, load takes only one operand so the liveness remains the same, additionally hoisting too much loads can have adversely affect the cache behavior.
+On the other hand sinking loads may improve the cache behavior, because we load as late as possible. But sinking computations may increase the live range.
+We might want to to improve the code-hoisting to a global code motion algorithm.
+
 \end{section}

### 2016-04-22 hoist.tex
Added advantages of code-hoisting

@@
 
+\section{Hoisting bottom up with optimistic approach}
+Optimistic means collect the set of all potential redundancies and then discard the bad ones.
+
+Bottom up:
+Sort the basic blocks of potentially redundant computations based on DFS-in numbers.
+Then start evaluating the safety checks for first two, if good then move to second one. That way we can even
+partially hoist computations should we find that we cannot remove all the redundant computations.
+
+We can also split the set of to partially hoist multiple times, if hoisting globally is not possible.
+
+In the case when more than one computation from the same basic block shows up as redundant.
+e.g.
+BB1
+load a;
+....
+store .;
+....
+load a;
+
+BB2
+load a
+
+BB0 -> BB1
+BB0 -> BB2
+
+In this case both the loads from BB1 will show up as redundant (according to the current algorithm), but the complete
+hoising will fail as the second load in BB1 is not safe to hoist. By following the bottom-up approach, the sorted list
+of BBs will appear like this:
+
+{ BB1, BB2(first load), BB2(second load) }
+
+Now we will start evaluating the safety of BB1-l1 and BB2-l2 which will pass, then BB12-l1 and BB2-l2 will fail.
+So we can revert back to BB12-l1 and hoist them to BB0. Doing the top down approach and finding the partial set
+to hoist instructions partially would result in a combinatorial explosion.
+
+GCC: 23286 has interesting test cases.
+
+\section{Enabling LICM by code-hoisting}
+loop:
+if (a)
+  i1;
+else
+  i1;
+
+Here i1 is redundant and can be hoisted out of loop. But the LICM will not do because it does not reason about instructions in the conditional.
+By code-hoisting i1 will be hoisted out of conditional which will benefit the LICM.
+
+\section{Improving the Inliner heuristics}
+Reducing the number of instructions also reduces the inline cost and hence improves the inliner heuristics.
+
+\section{Improving vectorization}
+It will also benefit vectorizer by reducing the number
+of basic blocks, should all the instructions in conditional get hoisted.
+
 \end{section}

### 2016-05-04 hoist.tex
hoisting scalars past a call

@@
 
+\section{Hoisting scalars}
+Scalars are the easiest to hoist because we do not have to analyze for mem-refs. As long as all the
+operands are available (which in SSA form we do, or may be there are chained dependencies), the scalar can
+be hoisted. Care must be taken in case of hoisting scalars too far, as that may increase register pressure
+and result in spills. For example hoisting a scalar past a call. In that case the call may result in
+save and restore of the register the scalar may be defined to. In our current implementation
+we do not hoist scalars past a call. Another way to mitigate this problem would be to
+reinstantiate (rematerialize) the computation after a call (may be as a different pass).
+
+\section{Hoisting calls}
+Calls can be hoisted by value-numbering all the arguments. For safety we have to check that no other side-effects are there
+between call and the hoisting point, just like loads and stores.
+
+\section{Hoisting stores}
+We need to value number the address and the value being stored.
+
 \section{Hoisting bottom up with optimistic approach}

### 2016-06-14 hoist.tex
Adding sections
partition

@@
+\section{Introduction}
+
+\section{Global Value Numbering}
+\section{Code Hoisting}
+Code hoisting does not remove redundancies by itself but it enables removal of redundancies by other passes.
+e.g., Enable inlining.
+\section{Related Work}
+
+\section{Cost models}
+\subsection{In the presence of calls}
+\subsection{Hoisting too far away}
+
+\section{Comparative analysis of different approaches to hoisting}
+\subsection{Optimistic approach}
+Collect GVN of all the instructions in the beginning.
+\subsection{Pessimistic approach}
+Start from a branch.
+
 \begin{section}
@@
 save and restore of the register the scalar may be defined to. In our current implementation
-we do not hoist scalars past a call. Another way to mitigate this problem would be to
+we do not hoist scalars past a call. Another way to mitigate this problem is be to
 reinstantiate (rematerialize) the computation after a call (may be as a different pass).
@@
 Sort the basic blocks of potentially redundant computations based on DFS-in numbers.
+TODO: Explain why sorting by DFS-in numbers.
 Then start evaluating the safety checks for first two, if good then move to second one. That way we can even
@@
 
+\section{Partition the list of hoisting candidates to maximize hoisting}
+The set of candidates are sorted in the order of their DFS-in numbers. So that candidates close to
+each other in the list are also closer (??) in the CFG.
+
+In the case where all the candidates cannot be hoisted at a common hoist point because of:
+\begin{enumerate}
+\item Safety conditions are not met.
+\item Live range would increase too much.
+\end{enumerate}
+
+We partition the list at the point where the above mentioned checks fail.
+TODO: Put an example to illustrate the point.
+
 \section{Enabling LICM by code-hoisting}

### 2016-07-02 hoist.tex
added motivating example for hoist

@@
 
-Global code motion.
- bb1: b = ...
- bb2: c = ...
- bb3: a = b + c (to be moved to b5)
- bb4: ... = a
- old-live-range = distance(bb1,bb3) + distance(bb2,bb3) + distance (bb3,bb4)
- new-live-rance = distance(bb1,bb5) + distance(bb2,bb5) + distance (bb5,bb4)
- distance(bbx, bby) = total instruction count in the path from bbx to bby
- If the new live-range is less than the old one it will be a good candidate
- for gcm. When both the ranges will be same:
-  - It is a simple copy of kind a = b.
-  - One of the operands is not a Instruction/Register.
-  - In these cases we need to have some heuristic or we can ignore them.
+\section{Hoisting to improve performance}
+Hoisting also reduces critical path length of execution in out of order machines (but not in sequential machines), by exposing ILP before the conditional where the instruction was hoisted.
 
- The decision for bb5 can be made by whether hoist/sink is beneficial.
+\begin{program}
+float foo(float d, float min, float max, float a)
+{
+  float tmin;
+  float tmax;
 
- Loads can be moved early as long as there is load available in each branch
- or there is already a load/store from/to the same underlying object.
- Stores can be moved up if there is a store/load to/from the same object.
- The idea is to establish that the object has memory allocated to it.
- Moving load/store together may help with locality of reference.
+  float inv = 1.0f / d;
+  if (inv >= 0) {
+    tmin = (min - a) * inv;
+    tmax = (max - a) * inv;
+  } else {
+    tmin = (max - a) * inv;
+    tmax = (min - a) * inv;
+  }
+  return tmax + tmin;
+}
 
- PRE via Global Value Numbering
- Remove redundant instructions which are:
-  - Already available in the dominator.
-  - Are in one of the sibling branches, i.e., the instruction is used at
-    a point which shares a common dominator where all the use-operands
-    are available.
- In some cases it is possible to generate redundancy by restructuring the code
- (Ref. Ras Bodik), but that I'll leave for the next iteration of this patch.
+\end{program}
 
- Hoisting also reduces critical path length of execution in out of order machines (but not in sequential machines), by exposing ILP before the conditional where the instruction was hoisted.
-This feature has already been identified in the previous patch which improves ray-trace.
+In this program the computations of tmax and tmin and identical in both the branches but they cannot be evaluated before inv. Since inv depends on a division operation
+which is generally more expensive than the addition, subtraction and multiplication operations as required to evaluate tmax and tmin, the total latency of computation across
+each branch is:
+
+O(div) + 2(O(sub) + O(mul))
+Or for out of order processors with two add units and two multiply units:
+O(div) + O(sub) + O(mul)
+
+But if we hoist the computation of tmax and tmin outside the conditionals, the C code version would look like this:
+\begin{program}
+float foo(float d, float min, float max, float a)
+{
+  float tmin;
+  float tmax;
+
+  tmin' = (min - a);
+  tmax' = (max - a);
+
+  float inv = 1.0f / d;
+  tmin' = tmin' * inv;
+  tmax' = tmax' * inv;
+
+  if (inv >= 0) {
+    tmin = tmin';
+    tmax = tmax';
+  } else {
+    tmin = tmax';
+    tmax = tmin';
+  }
+
+  return tmax + tmin;
+}
+
+\end{program}
+
+In this code the two subtractions and the div operations can execute in parallel. So the total number of cycles will be
+O(div) + O(mul)
+since O(div) is usually greater than O(sub)
+
+Of course, a sinking optimization could just sink the computation of both tmax and tmin, and since the final operation is an addition (aasociative under fast math), the entire if-else
+code would go away.
+
+\begin{program}
+float foo(float d, float min, float max, float a)
+{
+  float tmin;
+  float tmax;
+
+  tmin' = (min - a);
+  tmax' = (max - a);
+
+  float inv = 1.0f / d;
+  tmin' = tmin' * inv;
+  tmax' = tmax' * inv;
+
+  return tmax' + tmin';
+}
+\end{program}
 

### 2016-07-02 hoist.tex
remove

@@
 
- Concerns:
- Safety of load instructions to be checked. We cannot hoist loads until all the paths in the parent BBs have the same load. This is to comply with C semantics, although I'm not sure about this yet.
- Finding a cost model in case optimistic hoisting does not give desired results.
-
-It seems, code hoisting is beneficial in cases even if the computations are not redundant.
-Say c = f(a, b) is an instruction.
-Hoisting up will reduce the liveness of registers a and b, but will only increase the liveness of c.
-So we gain 2:1 even when the computation is not redundant.
-For loads this is not the case because, load takes only one operand so the liveness remains the same, additionally hoisting too much loads can have adversely affect the cache behavior.
-On the other hand sinking loads may improve the cache behavior, because we load as late as possible. But sinking computations may increase the live range.
-We might want to to improve the code-hoisting to a global code motion algorithm.
-
 \section{Hoisting scalars}

### 2016-07-29 hoist.tex
adding topics

@@
 \subsection{In the presence of calls}
+Hoisting scalars across calls is tricky because it can increase the number of spills.
+
 \subsection{Hoisting too far away}
+\begin{enumerate}
+\item When the hoist point is too far.
+\item When the number of instructions we may cross is very high.
+\end{enumerate}
 
@@
 \subsection{Pessimistic approach}
-Start from a branch.
+Start from a branch. Hoist one by one iteratively.
+\subsection{Triangle approach}
+
+\section{Using MemorySSA}
+\subsection{Iterative update to memory ssa during the hoisting process}
+
+\section{Without MemorySSA using the dominance information}
 

### 2016-08-01 hoist.tex
formatting. NFC
restructuring

@@
 \section{Code Hoisting}
-Code hoisting does not remove redundancies by itself but it enables removal of redundancies by other passes.
-e.g., Enable inlining.
+Code hoisting does not remove redundancies by itself but it enables
+removal of redundancies by other passes.  e.g., Enable inlining.
 \section{Related Work}
-
-\section{Cost models}
-\subsection{In the presence of calls}
-Hoisting scalars across calls is tricky because it can increase the number of spills.
-
-\subsection{Hoisting too far away}
-\begin{enumerate}
-\item When the hoist point is too far.
-\item When the number of instructions we may cross is very high.
-\end{enumerate}
-
-\section{Comparative analysis of different approaches to hoisting}
-\subsection{Optimistic approach}
-Collect GVN of all the instructions in the beginning.
-\subsection{Pessimistic approach}
-Start from a branch. Hoist one by one iteratively.
-\subsection{Triangle approach}
-
-\section{Using MemorySSA}
-\subsection{Iterative update to memory ssa during the hoisting process}
-
-\section{Without MemorySSA using the dominance information}
 
@@
 \section{Hoisting to improve performance}
-Hoisting also reduces critical path length of execution in out of order machines (but not in sequential machines), by exposing ILP before the conditional where the instruction was hoisted.
+Hoisting also reduces critical path length of execution in out of
+order machines (but not in sequential machines), by exposing ILP
+before the conditional where the instruction was hoisted.
 
@@
 
-In this program the computations of tmax and tmin and identical in both the branches but they cannot be evaluated before inv. Since inv depends on a division operation
-which is generally more expensive than the addition, subtraction and multiplication operations as required to evaluate tmax and tmin, the total latency of computation across
-each branch is:
+In this program the computations of tmax and tmin and identical in
+both the branches but they cannot be evaluated before inv. Since inv
+depends on a division operation which is generally more expensive than
+the addition, subtraction and multiplication operations as required to
+evaluate tmax and tmin, the total latency of computation across each
+branch is:
 
@@
 
-But if we hoist the computation of tmax and tmin outside the conditionals, the C code version would look like this:
+But if we hoist the computation of tmax and tmin outside the
+conditionals, the C code version would look like this:
 \begin{program}
@@
 
-In this code the two subtractions and the div operations can execute in parallel. So the total number of cycles will be
-O(div) + O(mul)
+In this code the two subtractions and the div operations can execute
+in parallel. So the total number of cycles will be O(div) + O(mul)
 since O(div) is usually greater than O(sub)
 
-Of course, a sinking optimization could just sink the computation of both tmax and tmin, and since the final operation is an addition (aasociative under fast math), the entire if-else
-code would go away.
+Of course, a sinking optimization could just sink the computation of
+both tmax and tmin, and since the final operation is an addition
+(asociative under fast math), the entire if-else code would go away.
 
@@
 \section{Hoisting scalars}
-Scalars are the easiest to hoist because we do not have to analyze for mem-refs. As long as all the
-operands are available (which in SSA form we do, or may be there are chained dependencies), the scalar can
-be hoisted. Care must be taken in case of hoisting scalars too far, as that may increase register pressure
-and result in spills. For example hoisting a scalar past a call. In that case the call may result in
-save and restore of the register the scalar may be defined to. In our current implementation
-we do not hoist scalars past a call. Another way to mitigate this problem is be to
-reinstantiate (rematerialize) the computation after a call (may be as a different pass).
+Scalars are the easiest to hoist because we do not have to analyze
+them for aliasing memory references. As long as all the operands are
+available (which in SSA form we do, or may be there are chained
+dependencies), the scalar can be hoisted. Care must be taken in case
+of hoisting scalars too far, as that may increase register pressure
+and result in spills. For example hoisting a scalar past a call. In
+that case the call may result in save and restore of the register the
+scalar may be defined to. In our current implementation we do not
+hoist scalars past a call. Another way to mitigate this problem is be
+to reinstantiate (rematerialize) the computation after a call (may be
+as a different pass).
 
-\section{Hoisting calls}
-Calls can be hoisted by value-numbering all the arguments. For safety we have to check that no other side-effects are there
-between call and the hoisting point, just like loads and stores.
+\section{Using MemorySSA}
+\subsection{Iterative update to memory ssa during the hoisting process}
+
+\section{Without MemorySSA using the dominance information}
 
@@
 
+\section{Hoisting calls}
+Calls can be hoisted by value-numbering all the arguments. For safety
+we have to check that there are no other side-effects between a call and
+its hoisting point, just like loads and stores.
+
+\section{Comparative analysis of different approaches to hoisting}
+\subsection{Optimistic approach}
+Collect GVN of all the instructions in the beginning.
+\subsection{Pessimistic approach}
+Start from a branch. Hoist one by one iteratively.
+\subsection{Triangle approach}
+
 \section{Hoisting bottom up with optimistic approach}
-Optimistic means collect the set of all potential redundancies and then discard the bad ones.
+Optimistic means collect the set of all potential redundancies and
+then discard the bad ones.
 
-Bottom up:
-Sort the basic blocks of potentially redundant computations based on DFS-in numbers.
-TODO: Explain why sorting by DFS-in numbers.
-Then start evaluating the safety checks for first two, if good then move to second one. That way we can even
-partially hoist computations should we find that we cannot remove all the redundant computations.
+Bottom up: Sort the basic blocks of potentially redundant computations
+based on DFS-in numbers.  TODO: Explain why sorting by DFS-in numbers.
+Then start evaluating the safety checks for first two, if good then
+move to second one. That way we can even partially hoist computations
+should we find that we cannot remove all the redundant computations.
 
-We can also split the set of to partially hoist multiple times, if hoisting globally is not possible.
+We can also split the set of to partially hoist multiple times, if
+hoisting globally is not possible.
 
-In the case when more than one computation from the same basic block shows up as redundant.
+In the case when more than one computation from the same basic block
+shows up as redundant.
+
 e.g.
@@
 
-In this case both the loads from BB1 will show up as redundant (according to the current algorithm), but the complete
-hoising will fail as the second load in BB1 is not safe to hoist. By following the bottom-up approach, the sorted list
-of BBs will appear like this:
+In this case both the loads from BB1 will show up as redundant
+(according to the current algorithm), but the complete hoising will
+fail as the second load in BB1 is not safe to hoist. By following the
+bottom-up approach, the sorted list of BBs will appear like this:
 
@@
 
-Now we will start evaluating the safety of BB1-l1 and BB2-l2 which will pass, then BB12-l1 and BB2-l2 will fail.
-So we can revert back to BB12-l1 and hoist them to BB0. Doing the top down approach and finding the partial set
-to hoist instructions partially would result in a combinatorial explosion.
+Now we will start evaluating the safety of BB1-l1 and BB2-l2 which
+will pass, then BB12-l1 and BB2-l2 will fail.  So we can revert back
+to BB12-l1 and hoist them to BB0. Doing the top down approach and
+finding the partial set to hoist instructions partially would result
+in a combinatorial explosion.
 
@@
 \section{Partition the list of hoisting candidates to maximize hoisting}
-The set of candidates are sorted in the order of their DFS-in numbers. So that candidates close to
-each other in the list are also closer (??) in the CFG.
+The set of candidates are sorted in the order of their DFS-in
+numbers. So that candidates close to each other in the list are also
+closer (??) in the CFG.
 
-In the case where all the candidates cannot be hoisted at a common hoist point because of:
+In the case where all the candidates cannot be hoisted at a common
+hoist point because of:
 \begin{enumerate}
@@
 
+\section{Cost models}
+\subsection{In the presence of calls}
+Hoisting scalars across calls is tricky because it can increase the
+number of spills.
+
+\subsection{Hoisting too far away}
+\begin{enumerate}
+\item When the hoist point is too far.
+\item When the number of instructions we may cross is very high.
+\end{enumerate}
+
 \section{Enabling LICM by code-hoisting}
@@
 
-Here i1 is redundant and can be hoisted out of loop. But the LICM will not do because it does not reason about instructions in the conditional.
-By code-hoisting i1 will be hoisted out of conditional which will benefit the LICM.
+Here i1 is redundant and can be hoisted out of loop. But the LICM will
+not do because it does not reason about instructions in the
+conditional.  By code-hoisting i1 will be hoisted out of conditional
+which will benefit the LICM.
 
 \section{Improving the Inliner heuristics}
-Reducing the number of instructions also reduces the inline cost and hence improves the inliner heuristics.
+Reducing the number of instructions also reduces the inline cost and
+hence improves the inliner heuristics.
 

### 2016-08-03 hoist.tex
formalism for partitioning candidates

@@
 
+In order to hoist a subset of identical instructions, we partition the
+list of all candidates in a way to maximise the total number of
+hoistings.  By sorting the list of all the candidates in the
+increasing order of their DFS-in numbers, we make sure that candidates
+closer in the list have their common dominator nearby.
+
+Essentially,
+abs(DFS-in(I1) - DFS-in(I2)) < abs(DFS-in(I1) - DFS-in(I3))
+=> dist(I1, nearest-common-dom(I1, I2)) <= dist(I1, nearest-common-dom(I1, I3))
+
+This means for candidates closer to each other w.r.t. DFS numbers, we
+would make fewer checks for preconditions.
+
 We partition the list at the point where the above mentioned checks fail.
+
 TODO: Put an example to illustrate the point.

### 2016-08-03 hoist.tex
pessimistic approach

@@
 Collect GVN of all the instructions in the beginning.
+
+\subsection{Triangle approach}
+This is the most basic approach to code-hoisting when the instructions are
+hoisted to immediate parent in a diamond like structure. When there are
+identical computations in both the branches of a conditional, and all the
+legality checks are met, the instruction can be hoisted to the predecessor basic
+block. This approach is simple to implement and has minimal impact on the
+register pressure.
+
 \subsection{Pessimistic approach}
-Start from a branch. Hoist one by one iteratively.
-\subsection{Triangle approach}
+To extend the triangle approach to the entire function, the algorithm traverses
+the basic blocks in the invers depth-first order. All the leaf nodes are visited
+before the non-leaf nodes because instructions are hoisted upwards. We start
+from a branch and collect the GVN of all the instructions. Then in the sibling
+basic block we find if there are any identical computations. Once we find an
+identical computation we try to hoist it to the nearest common dominator which
+is the immediate predecessor of both the basic blocks. We traverse the basic
+block from top to down so that the once an instruction is hoisted it creates
+opportunities for other dependent instructions. After all the sibling basic
+blocks have all the identical computations hoisted, we continue the hoisting
+process with their parent and its sibling of their parent until we reach the
+beginning of the function.
 
@@
 Essentially,
-abs(DFS-in(I1) - DFS-in(I2)) < abs(DFS-in(I1) - DFS-in(I3))
-=> dist(I1, nearest-common-dom(I1, I2)) <= dist(I1, nearest-common-dom(I1, I3))
+abs(DFSIn(I1) - DFSIn(I2)) < abs(DFSIn(I1) - DFSIn(I3))
+=> bbDist(I1, nearest-common-dom(I1, I2)) <= bbDist(I1, nearest-common-dom(I1, I3))
 
-This means for candidates closer to each other w.r.t. DFS numbers, we
-would make fewer checks for preconditions.
+This means for candidates closer to each other w.r.t. DFS numbers:
+\begin{enumerate}
+\item would make fewer checks for preconditions.
+\item hoisting point will be closer so, the intersection of live-range of the instruction
+with other instructions will be minimal.
+\end{enumerate}
 
-We partition the list at the point where the above mentioned checks fail.
+In our current implementation we partition the list at the point where the
+legality/profitability checks fail.
 
@@
 \section{Improving the Inliner heuristics}
-Reducing the number of instructions also reduces the inline cost and
+Reducing the number of instructions also reduces the inlining cost and
 hence improves the inliner heuristics.

### 2016-08-03 hoist.tex
optimistic approach

@@
 \section{Comparative analysis of different approaches to hoisting}
-\subsection{Optimistic approach}
-Collect GVN of all the instructions in the beginning.
-
 \subsection{Triangle approach}
@@
 beginning of the function.
+
+\subsection{Optimistic approach}
+In this approach, the goal is to maximise the total number of hoistings in the
+entire function.  This algorithm is very useful when optimizing for code-size.
+We see a lot of computations hoisted \ref{experimental-results}. In order to do
+that, we collect the GVN of all the instructions in the function. After that we
+iterate on the list of instructions having identical GVN. After that we find the
+common dominator dominating all the identical computations and try to hoist the
+instruction in that basic block, if feasible. Many times it is not possible to
+hoist all the instructions to one common dominator, due to legality/profitability
+constraints. In those cases, the algorithmn would partion the list of computations
+a partially hoist a set of computations to their common dominator. For more details
+see \ref{sec:partition}
+
+\subsubsection{pros,cons}
+Naturally, the amount of hoisting depends on the generality of the GVN
+algorithm. Current llvm GVN implementation finds out equivalence between
+dependent instructions to some extent.
 
@@
 \section{Partition the list of hoisting candidates to maximize hoisting}
+\cite{sec:partition}
 The set of candidates are sorted in the order of their DFS-in

### 2016-08-04 hoist.tex
make it compile

@@
+\documentclass{sig-alternate}
+\usepackage{comment}
+\usepackage{amsmath}
+\usepackage{hyperref}
+\usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{graphviz}
+\usepackage{auto-pst-pdf}
+\usepackage{etoolbox}
+\usepackage{flushend}
+\usepackage{needspace}
+
+\makeatletter
+\preto{\@verbatim}{\topsep=1pt \partopsep=0pt}
+\makeatother
+
+\pagenumbering{arabic}
+
+\begin{document}
+\def \SCoP {SCoP}
+\def \GCC {GCC}
+\def \LLVM {LLVM}
+\def \SESE {SESE}
+\def \CFG {CFG}
+\def \SSA {SSA}
+\def \scev {scev}
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\title{GVN-hoist: hoisting expressions from control flow}
+
+\toappear{
+   \hrule \vspace{5pt}
+   LLVM-dev 2016
+}
+\numberofauthors{2}
+
+\author{
+\alignauthor
+Aditya Kumar\\
+       \affaddr{Samsung Austin R\&D Center}\\
+       \email{aditya.k7@samsung.com}
+\alignauthor
+Sebastian Pop\\
+       \affaddr{Samsung Austin R\&D Center}\\
+       \email{s.pop@samsung.com}
+}
+
+\maketitle
+\begin{abstract}
+Code hoisting does not remove redundancies by itself but it enables
+removal of redundancies by other passes.  e.g., Enable inlining.
+\end{abstract}
+
 \section{Introduction}
+
+graph algorithms \cite{tarjan},
+SSA \cite{cytron}
+GVN \cite{rosen1988global,click1995global}
+CFG, DOM \cite{dragonbook}
+SESE, SEME \cite{sese}
 
@@
 \section{Related Work}
-
-\begin{section}
 
@@
 
-\begin{program}
+\begin{verbatim}
 float foo(float d, float min, float max, float a)
@@
 }
-
-\end{program}
+\end{verbatim}
 
@@
 conditionals, the C code version would look like this:
-\begin{program}
+\begin{verbatim}
 float foo(float d, float min, float max, float a)
@@
 
-\end{program}
+\end{verbatim}
 
@@
 }
-\end{program}
+\end{verbatim}
 
@@
 
-\end{section}
+\bibliographystyle{abbrv}
+{\small
+\bibliography{Bibliography}
+}
+\end{document}

### 2016-08-04 hoist.tex
clean a bit the related work

@@
 
+Why hoisting expressions:
+\begin{itemize}
+\item OOO processors: good for speculative execution of branches
+\item more ILP for in-order processors
+\item code size
+\item optimize for register pressure
+\item improve inlining heuristics
+\end{itemize}
+
+\begin{enumerate}
+\item Code hoisting in adjacent blocks in a diamond/triangle like structure.
+\item O(n) Algorithm.
+\item Hoisting equivalent computations in non-sibling BBs.
+\item Hoisting across switch blocks.
+\item Hoisting to improve PRE.
+\item Hoisting to reduce critical path in OoO execution.
+\item Ranking/Wavefront algorithm to hoist across switch blocks.
+\end{enumerate}
+
+\section{Related Work}
+
+A lot of bug reports in GCC and LLVM bugzillas, no papers on this topic yet.
+
+Code hoisting does not remove redundancies by itself but it enables
+removal of redundancies by other passes.  e.g., Enable inlining.
+
 graph algorithms \cite{tarjan},
@@
 
-\section{Global Value Numbering}
-\section{Code Hoisting}
-Code hoisting does not remove redundancies by itself but it enables
-removal of redundancies by other passes.  e.g., Enable inlining.
-\section{Related Work}
-
-1. Code hoisting in adjacent blocks in a diamond/triangle like structure.
-2. O(n) Algorithm.
-3. Hoisting equivalent computations in non-sibling BBs.
-4. Hoisting across switch blocks.
-5. Hoisting to improve PRE.
-6. Hoisting to reduce critical path in OoO execution.
-7. Ranking/Wavefront algorithm to hoist across switch blocks.
+\subsection{Global Value Numbering}
 

### 2016-08-04 hoist.tex
reorganize the sections

@@
 
-\section{Hoisting to improve performance}
+\section{Code hoisting}
 Hoisting also reduces critical path length of execution in out of
@@
 
-\section{Hoisting scalars}
+\subsection{Hoisting scalars}
 Scalars are the easiest to hoist because we do not have to analyze
@@
 
-\section{Using MemorySSA}
+\subsection{Using MemorySSA}
 \subsection{Iterative update to memory ssa during the hoisting process}
 
-\section{Without MemorySSA using the dominance information}
+\subsection{Without MemorySSA using the dominance information}
 
-\section{Hoisting stores}
+\subsection{Hoisting stores}
 We need to value number the address and the value being stored.
 
-\section{Hoisting calls}
+\subsection{Hoisting calls}
 Calls can be hoisted by value-numbering all the arguments. For safety
@@
 
-\section{Comparative analysis of different approaches to hoisting}
+\subsection{Cost models}
+
+\subsection{Reduce register pressure}
+Hoisting scalars across calls is tricky because it can increase the
+number of spills.
+
+\subsection{In the presence of calls}
+Caller saved registers.
+
+\subsection{Hoisting too far away}
+\begin{enumerate}
+\item When the hoist point is too far.
+\item When the number of instructions we may cross is very high.
+\end{enumerate}
+
+\subsection{Enabling LICM by code-hoisting}
+loop:
+if (a)
+  i1;
+else
+  i1;
+
+Here i1 is redundant and can be hoisted out of loop. But the LICM will
+not do because it does not reason about instructions in the
+conditional.  By code-hoisting i1 will be hoisted out of conditional
+which will benefit the LICM.
+
+\subsection{Improving the Inliner heuristics}
+Reducing the number of instructions also reduces the inlining cost and
+hence improves the inliner heuristics.
+
+\subsection{Improving vectorization}
+It will also benefit vectorizer by reducing the number
+of basic blocks, should all the instructions in conditional get hoisted.
+
+\section{Experimental Evaluation}
+
+\subsection{Comparative analysis of different approaches to hoisting}
 \subsection{Triangle approach}
@@
 
-\section{Hoisting bottom up with optimistic approach}
+\subsection{Hoisting bottom up with optimistic approach}
 Optimistic means collect the set of all potential redundancies and
@@
 
-\section{Partition the list of hoisting candidates to maximize hoisting}
+\subsection{Partition the list of hoisting candidates to maximize hoisting}
 \cite{sec:partition}
@@
 
-\section{Cost models}
-\subsection{In the presence of calls}
-Hoisting scalars across calls is tricky because it can increase the
-number of spills.
-
-\subsection{Hoisting too far away}
-\begin{enumerate}
-\item When the hoist point is too far.
-\item When the number of instructions we may cross is very high.
-\end{enumerate}
-
-\section{Enabling LICM by code-hoisting}
-loop:
-if (a)
-  i1;
-else
-  i1;
-
-Here i1 is redundant and can be hoisted out of loop. But the LICM will
-not do because it does not reason about instructions in the
-conditional.  By code-hoisting i1 will be hoisted out of conditional
-which will benefit the LICM.
-
-\section{Improving the Inliner heuristics}
-Reducing the number of instructions also reduces the inlining cost and
-hence improves the inliner heuristics.
-
-\section{Improving vectorization}
-It will also benefit vectorizer by reducing the number
-of basic blocks, should all the instructions in conditional get hoisted.
+\section{Conclusion and Future Work}
 

### 2016-08-04 hoist.tex
rework introduction and add MSSA ref

@@
 \begin{abstract}
-Code hoisting does not remove redundancies by itself but it enables
-removal of redundancies by other passes.  e.g., Enable inlining.
+  Code hoisting identifies identical computations and replaces them with a
+  single expression in the common dominator.  Code hoisting is often confused
+  with CSE (Common Subexpression Elimination) although is not bound to find
+  redundancies within a single basic block.  It could also be confused with PRE,
+  although the main goal is not to remove redundancies: it enables removal of
+  redundancies by other passes like LICM.  The main goal of code hoisting is to
+  reduce code size with added benefit of exposing instruction level parallelism
+  and reduced register pressure.
 \end{abstract}
@@
 SESE, SEME \cite{sese}
+MemorySSA \cite{novillo2007memory}
 

### 2016-08-04 hoist.tex
register pressure

@@
 \subsection{Reduce register pressure}
+Following example explains how code hoisting can actually reduce the register pressure.
+Consider the following example where the labels prefixed with 'p' represent the position of
+instruction in a basic block (names prefixed with bb).
+
+bb0: p0: b = ...
+bb1: p1: c = ...
+bb2: p2: if (c) jump bb3 else jump bb4
+bb3: p3: a0 = b + c
+bb4: p4: a1 = b + c
+bb5: p5: d = phi {a0(bb3), a1(bb4)}
+
+If we measure distance(px, py) as total instruction count in the path from the position of px to py
+
+old-live-range = max(distance(p0, p3) + distance(p1, p3) , distance(p0, p4) + distance(p1, p4)) + distance (p3, p5)
+new-live-rance = distance(p0, p2) + distance(p1, p2) + distance (p2, p5)
+
+If the new live-range is less than the old one it will be a good candidate
+for hoist. There are cases when both the ranges will be same:
+\begin{enumerate}
+  - There is only one operand on the right hand side e.g., an assignment operation.
+  - One of the operands is not a Instruction/Register.
+\end{enumerate}
+
+In a special case where the instruction to be hoisted has the last use of its
+operands then the code hoisting will always reduce the register pressure if it
+has two register operands because the gain in live-range will be in the ratio of
+2:1. Based on the above formulae we can also deduce that, as long as there is
+one register operand in the right hand side with its last use, code hoisting
+will either decrease or maintain the register pressure.
+
 Hoisting scalars across calls is tricky because it can increase the

### 2016-08-04 hoist.tex
calls

@@
 \subsection{Reduce register pressure}
+\label{hoist:reg-pressure}
 Following example explains how code hoisting can actually reduce the register pressure.
@@
 
+\begin{verbatim}
 bb0: p0: b = ...
@@
 
-If we measure distance(px, py) as total instruction count in the path from the position of px to py
+If we measure distance(px, py) as total instruction
+count in the path from the position of px to py
 
-old-live-range = max(distance(p0, p3) + distance(p1, p3) , distance(p0, p4) + distance(p1, p4)) + distance (p3, p5)
-new-live-rance = distance(p0, p2) + distance(p1, p2) + distance (p2, p5)
+old-live-range = max(distance(p0, p3) + distance(p1, p3) ,
+                     distance(p0, p4) + distance(p1, p4))
+                 + distance (p3, p5)
+new-live-rance = distance(p0, p2) + distance(p1, p2)
+                 + distance (p2, p5)
+\end{verbatim}
 
@@
 \begin{enumerate}
-  - There is only one operand on the right hand side e.g., an assignment operation.
-  - One of the operands is not a Instruction/Register.
+  \item There is only one operand on the right hand side e.g., an assignment operation.
+  \item One of the operands is not a Instruction/Register.
 \end{enumerate}
@@
 
-Hoisting scalars across calls is tricky because it can increase the
-number of spills.
+\subsection{In the presence of calls}
+Hoisting scalars across calls is tricky because it can increase the number of
+spills. During the frame lowering of calls, the argument registers, in general,
+the caller saved registers are saved because they might be modified by the
+callee and after the call they are restored. So before the call the register
+pressure is high because the number of available registers are reduced by the
+number of caller saved registers. In that situation if a computation is hoisted
+across the call, that would increase the total number of registers required by
+one, thus contributing to the register pressure.
 
-\subsection{In the presence of calls}
-Caller saved registers.
+However, in the two special cases discussed in Section-\ref{hoist:reg-pressure},
+it will be okay to hoist.
 
@@
 \section{Experimental Evaluation}
+\label{sec:experimental-results}
 
@@
 entire function.  This algorithm is very useful when optimizing for code-size.
-We see a lot of computations hoisted \ref{experimental-results}. In order to do
-that, we collect the GVN of all the instructions in the function. After that we
-iterate on the list of instructions having identical GVN. After that we find the
-common dominator dominating all the identical computations and try to hoist the
-instruction in that basic block, if feasible. Many times it is not possible to
-hoist all the instructions to one common dominator, due to legality/profitability
-constraints. In those cases, the algorithmn would partion the list of computations
-a partially hoist a set of computations to their common dominator. For more details
-see \ref{sec:partition}
+We see a lot of computations hoisted (see
+Section~\ref{sec:experimental-results}. In order to do that, we collect the GVN
+of all the instructions in the function. After that we iterate on the list of
+instructions having identical GVN. After that we find the common dominator
+dominating all the identical computations and try to hoist the instruction in
+that basic block, if feasible. Many times it is not possible to hoist all the
+instructions to one common dominator, due to legality/profitability
+constraints. In those cases, the algorithmn would partion the list of
+computations a partially hoist a set of computations to their common
+dominator. For more details see Section~\ref{subsec:partition}.
 
@@
 \subsection{Partition the list of hoisting candidates to maximize hoisting}
-\cite{sec:partition}
+\label{subsec:partition}
 The set of candidates are sorted in the order of their DFS-in

### 2016-08-04 hoist.tex
fix typo
hoisting too far away

@@
 \subsection{Hoisting too far away}
+When we load a value from memory, we require an extra register to keep the value
+around. Hoisting a load means we need to keep the value alive for a longer
+period. That could result in spills later in the register allocation.
+
+Also, if there are several instructions in between the hoisting point and the
+instruction to be hoisted, the instruction to be hoisted crosses several
+instructions while hoisting, it means we are adding one register to all the
+live-ranges spanning the instructions. That could result in spills.
+
 \begin{enumerate}

### 2016-08-04 hoist.tex
loads stores

@@
 
+\subsection{Hoisting loads}
+In order to value number stores, we value number (hash) the address from where
+the value is to be loaded. Then we need to compute the hoisting point just like
+we did for the scalars. At the hoisting point we compute if the operands to the
+load (an address) is available. If it is not available we can try to compute the
+addresss if that is possible.  Along the path from current position of the load
+instruction through the hoiting point, we find if there are aliasing stores, we
+find if any of the basic blocks are landing pads or targets of indirect
+branches. In those cases we discard the candidate. Since the load increases the
+register pressure by one (unless the operand to load becomes dead at the load),
+we evaluate if it is beneficial to hoist the load.
+
 \subsection{Hoisting stores}
-We need to value number the address and the value being stored.
+In order to value number stores, we value number (hash) the address as well as
+the value to be stored to that address. For the legality, we need to check the
+dependency requirements just like hoisting loads. We need to check that the
+operands of store are available at the hoisting point, that there are no
+aliasing loads or store along the path from the current position through the
+hoisting point and, none of the basic blocks along the path are landing pads.
+Once all the legality checks are satisfied, it is a good candidate to be
+hoisted.
 

### 2016-08-04 hoist.tex
hoisting too far away

@@
 However, in the two special cases discussed in Section-\ref{hoist:reg-pressure},
-it will be okay to hoist.
+it will be okay to hoist because the regiter pressure would not change.
 
@@
 instructions while hoisting, it means we are adding one register to all the
-live-ranges spanning the instructions. That could result in spills.
+live-ranges spanning the instructions. That could result in spills. In the
+current implementation we choose to hoist if the number of instructions crossed
+is below a threshold. Ideally, it should be okay to hoist all the instructions
+and a later live-range-splitting pass should make the right decision of
+rematerializing the instruction should it be beneficial to do so. But the
+current live-range splitting pass of llvm is not making the optimial decision
+and we have found spills if the threshold is exceeded. The threshold was
+computed as a result of tuning the llvm testsuite \cite{llvm-nightly} and spec
+benchmark \cite{Henning2000}.
 
@@
 \begin{enumerate}
-\item would make fewer checks for preconditions.
-\item hoisting point will be closer so, the intersection of live-range of the instruction
-with other instructions will be minimal.
+\item would make fewer checks for legality and profitability.
+\item hoisting point will be closer so, the intersection of live-range of the
+  instruction with other instructions will be minimal.
 \end{enumerate}

### 2016-08-04 hoist.tex
rewrite abstract

@@
 
-\title{GVN-hoist: hoisting expressions from control flow}
+\title{GVN-Hoist: Hoisting Computations from Branches}
 
@@
 \begin{abstract}
-  Code hoisting identifies identical computations and replaces them with a
-  single expression in the common dominator.  Code hoisting is often confused
-  with CSE (Common Subexpression Elimination) although is not bound to find
-  redundancies within a single basic block.  It could also be confused with PRE,
-  although the main goal is not to remove redundancies: it enables removal of
-  redundancies by other passes like LICM.  The main goal of code hoisting is to
-  reduce code size with added benefit of exposing instruction level parallelism
-  and reduced register pressure.
+  Code hoisting identifies identical computations across the program with the
+  help of a Global Value Numbering (GVN) analysis, identifies a hoisting
+  position that dominates all the identical computations and in which all
+  operands are avaialable, duplicates one of the expressions in the hoisting
+  position, and removes all the computations now redundant in their original
+  position.
+
+  Code hoisting is often confused with CSE (Common Subexpression Elimination)
+  although hoisting is not bound to find redundancies within a single basic
+  block.  Code hoisting could also be confused with Partial Redundancy
+  Elimination (PRE) or even with Global Common Subexpression Elimination (GCSE)
+  although the main goal of code hoisting is not to remove redundancies: code
+  hoisting effectively exposes redundancies and enables other passes like LICM
+  to remove more redundancies.  The main goal of code hoisting is to reduce code
+  size with the added benefit of exposing instruction level parallelism and
+  reduced register pressure.
 \end{abstract}
@@
 \subsection{Enabling LICM by code-hoisting}
+
+\begin{verbatim}
 loop:
@@
   i1;
+\end{verbatim}
 
@@
 e.g.
+\begin{verbatim}
 BB1
@@
 BB0 -> BB2
+\end{verbatim}
 

### 2016-08-05 hoist.tex
calls

@@
 the caller saved registers are saved because they might be modified by the
-callee and after the call they are restored. So before the call the register
-pressure is high because the number of available registers are reduced by the
-number of caller saved registers. In that situation if a computation is hoisted
-across the call, that would increase the total number of registers required by
-one, thus contributing to the register pressure.
+callee and after the call they are restored \cite{frame-lowering}. So before the
+call, the register pressure is high because the number of available registers
+are reduced by the number of caller saved registers. In that situation if a
+computation is hoisted across the call, that would increase the total number of
+registers required by one, thus contributing to the register pressure.
 
@@
 it will be okay to hoist because the regiter pressure would not change.
+
+Hoisting memory references also require precise analysis of all the memory
+addresses accessed by the call. In our implementation, since we only analyze the
+callee declaration the analysis is very conservative. In the presence of pure
+calls, loads can be hoisted but stores can't. Also, if the call throws exceptions,
+or it it may not return, memory references cannot be hoisted.
 

### 2016-08-05 hoist.tex
some introduction

@@
   to remove more redundancies.  The main goal of code hoisting is to reduce code
-  size with the added benefit of exposing instruction level parallelism and
+  size with the added benefit of exposing more instruction level parallelism and
   reduced register pressure.
@@
 \section{Introduction}
+
+Classical scalar optimizations like CSE \cite{dragonbook} work very well on
+single basic blocks.  With time, we have seen the transition to more complex
+passes like GCSE that work across basic blocks, or other passes like the
+macroblock formation that create a region within
 

### 2016-08-05 hoist.tex
restructure sections

@@
 
+\subsection{Enabling LICM by code-hoisting}
+
+\begin{verbatim}
+loop:
+if (a)
+  i1;
+else
+  i1;
+\end{verbatim}
+
+Here i1 is redundant and can be hoisted out of loop. But the LICM will
+not do because it does not reason about instructions in the
+conditional.  By code-hoisting i1 will be hoisted out of conditional
+which will benefit the LICM.
+
+\subsection{Improving the Inliner heuristics}
+Reducing the number of instructions also reduces the inlining cost and
+hence improves the inliner heuristics.
+
+\subsection{Improving vectorization}
+It will also benefit vectorizer by reducing the number
+of basic blocks, should all the instructions in conditional get hoisted.
+
 \section{Related Work}
@@
 
+\newpage
+
 \section{Code hoisting}
+
+\subsection{Overview}
+
 Hoisting also reduces critical path length of execution in out of
@@
 
-\subsection{Hoisting scalars}
+\subsubsection{Hoisting scalars}
 Scalars are the easiest to hoist because we do not have to analyze
@@
 
-\subsection{Using MemorySSA}
-\subsection{Iterative update to memory ssa during the hoisting process}
-
-\subsection{Without MemorySSA using the dominance information}
-
-\subsection{Hoisting loads}
+\subsubsection{Hoisting loads}
 In order to value number stores, we value number (hash) the address from where
@@
 
-\subsection{Hoisting stores}
+\subsubsection{Hoisting stores}
 In order to value number stores, we value number (hash) the address as well as
@@
 
-\subsection{Hoisting calls}
+\subsubsection{Hoisting calls}
 Calls can be hoisted by value-numbering all the arguments. For safety
@@
 
-\subsection{Cost models}
+\subsection{Algorithm}
+A good full page.
 
-\subsection{Reduce register pressure}
-\label{hoist:reg-pressure}
-Following example explains how code hoisting can actually reduce the register pressure.
-Consider the following example where the labels prefixed with 'p' represent the position of
-instruction in a basic block (names prefixed with bb).
-
-\begin{verbatim}
-bb0: p0: b = ...
-bb1: p1: c = ...
-bb2: p2: if (c) jump bb3 else jump bb4
-bb3: p3: a0 = b + c
-bb4: p4: a1 = b + c
-bb5: p5: d = phi {a0(bb3), a1(bb4)}
-
-If we measure distance(px, py) as total instruction
-count in the path from the position of px to py
-
-old-live-range = max(distance(p0, p3) + distance(p1, p3) ,
-                     distance(p0, p4) + distance(p1, p4))
-                 + distance (p3, p5)
-new-live-rance = distance(p0, p2) + distance(p1, p2)
-                 + distance (p2, p5)
-\end{verbatim}
-
-If the new live-range is less than the old one it will be a good candidate
-for hoist. There are cases when both the ranges will be same:
-\begin{enumerate}
-  \item There is only one operand on the right hand side e.g., an assignment operation.
-  \item One of the operands is not a Instruction/Register.
-\end{enumerate}
-
-In a special case where the instruction to be hoisted has the last use of its
-operands then the code hoisting will always reduce the register pressure if it
-has two register operands because the gain in live-range will be in the ratio of
-2:1. Based on the above formulae we can also deduce that, as long as there is
-one register operand in the right hand side with its last use, code hoisting
-will either decrease or maintain the register pressure.
-
-\subsection{In the presence of calls}
-Hoisting scalars across calls is tricky because it can increase the number of
-spills. During the frame lowering of calls, the argument registers, in general,
-the caller saved registers are saved because they might be modified by the
-callee and after the call they are restored \cite{frame-lowering}. So before the
-call, the register pressure is high because the number of available registers
-are reduced by the number of caller saved registers. In that situation if a
-computation is hoisted across the call, that would increase the total number of
-registers required by one, thus contributing to the register pressure.
-
-However, in the two special cases discussed in Section-\ref{hoist:reg-pressure},
-it will be okay to hoist because the regiter pressure would not change.
-
-Hoisting memory references also require precise analysis of all the memory
-addresses accessed by the call. In our implementation, since we only analyze the
-callee declaration the analysis is very conservative. In the presence of pure
-calls, loads can be hoisted but stores can't. Also, if the call throws exceptions,
-or it it may not return, memory references cannot be hoisted.
-
-\subsection{Hoisting too far away}
-When we load a value from memory, we require an extra register to keep the value
-around. Hoisting a load means we need to keep the value alive for a longer
-period. That could result in spills later in the register allocation.
-
-Also, if there are several instructions in between the hoisting point and the
-instruction to be hoisted, the instruction to be hoisted crosses several
-instructions while hoisting, it means we are adding one register to all the
-live-ranges spanning the instructions. That could result in spills. In the
-current implementation we choose to hoist if the number of instructions crossed
-is below a threshold. Ideally, it should be okay to hoist all the instructions
-and a later live-range-splitting pass should make the right decision of
-rematerializing the instruction should it be beneficial to do so. But the
-current live-range splitting pass of llvm is not making the optimial decision
-and we have found spills if the threshold is exceeded. The threshold was
-computed as a result of tuning the llvm testsuite \cite{llvm-nightly} and spec
-benchmark \cite{Henning2000}.
-
-\begin{enumerate}
-\item When the hoist point is too far.
-\item When the number of instructions we may cross is very high.
-\end{enumerate}
-
-\subsection{Enabling LICM by code-hoisting}
-
-\begin{verbatim}
-loop:
-if (a)
-  i1;
-else
-  i1;
-\end{verbatim}
-
-Here i1 is redundant and can be hoisted out of loop. But the LICM will
-not do because it does not reason about instructions in the
-conditional.  By code-hoisting i1 will be hoisted out of conditional
-which will benefit the LICM.
-
-\subsection{Improving the Inliner heuristics}
-Reducing the number of instructions also reduces the inlining cost and
-hence improves the inliner heuristics.
-
-\subsection{Improving vectorization}
-It will also benefit vectorizer by reducing the number
-of basic blocks, should all the instructions in conditional get hoisted.
-
-\section{Experimental Evaluation}
-\label{sec:experimental-results}
-
-\subsection{Comparative analysis of different approaches to hoisting}
-\subsection{Triangle approach}
-This is the most basic approach to code-hoisting when the instructions are
-hoisted to immediate parent in a diamond like structure. When there are
-identical computations in both the branches of a conditional, and all the
-legality checks are met, the instruction can be hoisted to the predecessor basic
-block. This approach is simple to implement and has minimal impact on the
-register pressure.
-
-\subsection{Pessimistic approach}
-To extend the triangle approach to the entire function, the algorithm traverses
-the basic blocks in the invers depth-first order. All the leaf nodes are visited
-before the non-leaf nodes because instructions are hoisted upwards. We start
-from a branch and collect the GVN of all the instructions. Then in the sibling
-basic block we find if there are any identical computations. Once we find an
-identical computation we try to hoist it to the nearest common dominator which
-is the immediate predecessor of both the basic blocks. We traverse the basic
-block from top to down so that the once an instruction is hoisted it creates
-opportunities for other dependent instructions. After all the sibling basic
-blocks have all the identical computations hoisted, we continue the hoisting
-process with their parent and its sibling of their parent until we reach the
-beginning of the function.
-
-\subsection{Optimistic approach}
-In this approach, the goal is to maximise the total number of hoistings in the
-entire function.  This algorithm is very useful when optimizing for code-size.
-We see a lot of computations hoisted (see
-Section~\ref{sec:experimental-results}. In order to do that, we collect the GVN
-of all the instructions in the function. After that we iterate on the list of
-instructions having identical GVN. After that we find the common dominator
-dominating all the identical computations and try to hoist the instruction in
-that basic block, if feasible. Many times it is not possible to hoist all the
-instructions to one common dominator, due to legality/profitability
-constraints. In those cases, the algorithmn would partion the list of
-computations a partially hoist a set of computations to their common
-dominator. For more details see Section~\ref{subsec:partition}.
-
-\subsubsection{pros,cons}
-Naturally, the amount of hoisting depends on the generality of the GVN
-algorithm. Current llvm GVN implementation finds out equivalence between
-dependent instructions to some extent.
-
-\subsection{Hoisting bottom up with optimistic approach}
+\subsubsection{Hoisting bottom up with optimistic approach}
 Optimistic means collect the set of all potential redundancies and
@@
 
-\subsection{Partition the list of hoisting candidates to maximize hoisting}
+\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
 \label{subsec:partition}
@@
 
+\subsection{Cost models}
+
+\subsubsection{Reduce register pressure}
+\label{hoist:reg-pressure}
+Following example explains how code hoisting can actually reduce the register pressure.
+Consider the following example where the labels prefixed with 'p' represent the position of
+instruction in a basic block (names prefixed with bb).
+
+\begin{verbatim}
+bb0: p0: b = ...
+bb1: p1: c = ...
+bb2: p2: if (c) jump bb3 else jump bb4
+bb3: p3: a0 = b + c
+bb4: p4: a1 = b + c
+bb5: p5: d = phi {a0(bb3), a1(bb4)}
+
+If we measure distance(px, py) as total instruction
+count in the path from the position of px to py
+
+old-live-range = max(distance(p0, p3) + distance(p1, p3) ,
+                     distance(p0, p4) + distance(p1, p4))
+                 + distance (p3, p5)
+new-live-rance = distance(p0, p2) + distance(p1, p2)
+                 + distance (p2, p5)
+\end{verbatim}
+
+If the new live-range is less than the old one it will be a good candidate
+for hoist. There are cases when both the ranges will be same:
+\begin{enumerate}
+  \item There is only one operand on the right hand side e.g., an assignment operation.
+  \item One of the operands is not a Instruction/Register.
+\end{enumerate}
+
+In a special case where the instruction to be hoisted has the last use of its
+operands then the code hoisting will always reduce the register pressure if it
+has two register operands because the gain in live-range will be in the ratio of
+2:1. Based on the above formulae we can also deduce that, as long as there is
+one register operand in the right hand side with its last use, code hoisting
+will either decrease or maintain the register pressure.
+
+\subsubsection{In the presence of calls}
+Hoisting scalars across calls is tricky because it can increase the number of
+spills. During the frame lowering of calls, the argument registers, in general,
+the caller saved registers are saved because they might be modified by the
+callee and after the call they are restored \cite{frame-lowering}. So before the
+call, the register pressure is high because the number of available registers
+are reduced by the number of caller saved registers. In that situation if a
+computation is hoisted across the call, that would increase the total number of
+registers required by one, thus contributing to the register pressure.
+
+However, in the two special cases discussed in Section-\ref{hoist:reg-pressure},
+it will be okay to hoist because the regiter pressure would not change.
+
+Hoisting memory references also require precise analysis of all the memory
+addresses accessed by the call. In our implementation, since we only analyze the
+callee declaration the analysis is very conservative. In the presence of pure
+calls, loads can be hoisted but stores can't. Also, if the call throws exceptions,
+or it it may not return, memory references cannot be hoisted.
+
+\subsubsection{Hoisting too far away}
+When we load a value from memory, we require an extra register to keep the value
+around. Hoisting a load means we need to keep the value alive for a longer
+period. That could result in spills later in the register allocation.
+
+Also, if there are several instructions in between the hoisting point and the
+instruction to be hoisted, the instruction to be hoisted crosses several
+instructions while hoisting, it means we are adding one register to all the
+live-ranges spanning the instructions. That could result in spills. In the
+current implementation we choose to hoist if the number of instructions crossed
+is below a threshold. Ideally, it should be okay to hoist all the instructions
+and a later live-range-splitting pass should make the right decision of
+rematerializing the instruction should it be beneficial to do so. But the
+current live-range splitting pass of llvm is not making the optimial decision
+and we have found spills if the threshold is exceeded. The threshold was
+computed as a result of tuning the llvm testsuite \cite{llvm-nightly} and spec
+benchmark \cite{Henning2000}.
+
+\begin{enumerate}
+\item When the hoist point is too far.
+\item When the number of instructions we may cross is very high.
+\end{enumerate}
+
+\subsection{Other Approaches}
+
+\subsubsection{Triangle approach}
+This is the most basic approach to code-hoisting when the instructions are
+hoisted to immediate parent in a diamond like structure. When there are
+identical computations in both the branches of a conditional, and all the
+legality checks are met, the instruction can be hoisted to the predecessor basic
+block. This approach is simple to implement and has minimal impact on the
+register pressure.
+
+\subsubsection{Pessimistic approach}
+To extend the triangle approach to the entire function, the algorithm traverses
+the basic blocks in the invers depth-first order. All the leaf nodes are visited
+before the non-leaf nodes because instructions are hoisted upwards. We start
+from a branch and collect the GVN of all the instructions. Then in the sibling
+basic block we find if there are any identical computations. Once we find an
+identical computation we try to hoist it to the nearest common dominator which
+is the immediate predecessor of both the basic blocks. We traverse the basic
+block from top to down so that the once an instruction is hoisted it creates
+opportunities for other dependent instructions. After all the sibling basic
+blocks have all the identical computations hoisted, we continue the hoisting
+process with their parent and its sibling of their parent until we reach the
+beginning of the function.
+
+\subsubsection{Optimistic approach}
+In this approach, the goal is to maximise the total number of hoistings in the
+entire function.  This algorithm is very useful when optimizing for code-size.
+We see a lot of computations hoisted (see
+Section~\ref{sec:experimental-results}. In order to do that, we collect the GVN
+of all the instructions in the function. After that we iterate on the list of
+instructions having identical GVN. After that we find the common dominator
+dominating all the identical computations and try to hoist the instruction in
+that basic block, if feasible. Many times it is not possible to hoist all the
+instructions to one common dominator, due to legality/profitability
+constraints. In those cases, the algorithmn would partion the list of
+computations a partially hoist a set of computations to their common
+dominator. For more details see Section~\ref{subsec:partition}.
+
+\subsubsection{pros,cons}
+Naturally, the amount of hoisting depends on the generality of the GVN
+algorithm. Current llvm GVN implementation finds out equivalence between
+dependent instructions to some extent.
+
+\subsubsection{PRE based hoisting}
+
+\newpage
+
+\section{Experimental Evaluation}
+\label{sec:experimental-results}
+
+\subsection{Comparative analysis of different approaches to hoisting}
+
 \section{Conclusion and Future Work}

### 2016-08-07 hoist.tex
overview

@@
 
-Hoisting also reduces critical path length of execution in out of
-order machines (but not in sequential machines), by exposing ILP
-before the conditional where the instruction was hoisted.
+Code hoisting can also reduce the critical path length of execution in out of
+order machines. As more instructios are available at the hoisting point, the
+hardware has more instructions to reorder. Following example illustrates how
+hoisting can improve performance by exposing more ILP.
 
@@
 the addition, subtraction and multiplication operations as required to
-evaluate tmax and tmin, the total latency of computation across each
+evaluate tmax and tmin. The total latency of computation across each
 branch is:
@@
 In this code the two subtractions and the div operations can execute
-in parallel. So the total number of cycles will be O(div) + O(mul)
-since O(div) is usually greater than O(sub)
+in parallel. So the total number of cycles will be max(O(div), O(sub)) + O(mul)
+= O(div) + O(mul); since O(div) is usually greater than O(sub)
 
-Of course, a sinking optimization could just sink the computation of
-both tmax and tmin, and since the final operation is an addition
-(asociative under fast math), the entire if-else code would go away.
+Of course, a partial redundancy elimination pass could just remove the entire
+if-block because final operation is an addition (asociative under fast math).
 

### 2016-08-07 hoist.tex
overview

@@
 {
-  float tmin;
-  float tmax;
+  float tmin, tmax, inv;
 
-  float inv = 1.0f / d;
+  inv = 1.0f / d;
   if (inv >= 0) {
@@
 
-In this program the computations of tmax and tmin and identical in
-both the branches but they cannot be evaluated before inv. Since inv
-depends on a division operation which is generally more expensive than
-the addition, subtraction and multiplication operations as required to
-evaluate tmax and tmin. The total latency of computation across each
-branch is:
+In this program the computations of tmax and tmin are identical to the
+computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
+depends on inv which depends on a division operation which is generally more
+expensive than the addition, subtraction and multiplication operations. The
+total latency of computation across each branch is:
 
@@
 
-But if we hoist the computation of tmax and tmin outside the
+Now if the computation of tmax and tmin are hoisted outside the
 conditionals, the C code version would look like this:
@@
 {
-  float tmin;
-  float tmax;
+  float tmin, tmax, tmin1, tmax1, inv;
 
-  tmin' = (min - a);
-  tmax' = (max - a);
+  tmin1 = (min - a);
+  tmax1 = (max - a);
 
-  float inv = 1.0f / d;
-  tmin' = tmin' * inv;
-  tmax' = tmax' * inv;
+  inv = 1.0f / d;
+  tmin1 = tmin1 * inv;
+  tmax1 = tmax1 * inv;
 
   if (inv >= 0) {
-    tmin = tmin';
-    tmax = tmax';
+    tmin = tmin1;
+    tmax = tmax1;
   } else {
-    tmin = tmax';
-    tmax = tmin';
+    tmin = tmax1;
+    tmax = tmin1;
   }
@@
 
-In this code the two subtractions and the div operations can execute
-in parallel. So the total number of cycles will be max(O(div), O(sub)) + O(mul)
-= O(div) + O(mul); since O(div) is usually greater than O(sub)
+In this code the two subtractions and the division operations can be executed in
+parallel because there are no dependencies among them. So the total number of
+cycles will be max(O(div), O(sub)) + O(mul) = O(div) + O(mul); since O(div) is
+usually much greater than O(sub) \cite{x86,aarch64}
 
@@
 {
-  float tmin;
-  float tmax;
+  float tmin1, tmax1, inv;
 
-  tmin' = (min - a);
-  tmax' = (max - a);
+  tmin1 = (min - a);
+  tmax1 = (max - a);
 
-  float inv = 1.0f / d;
-  tmin' = tmin' * inv;
-  tmax' = tmax' * inv;
+  inv = 1.0f / d;
+  tmin1 = tmin1 * inv;
+  tmax1 = tmax1 * inv;
 
-  return tmax' + tmin';
+  return tmax1 + tmin1;
 }

### 2016-08-08 hoist.tex
scalar load store call

@@
 \subsubsection{Hoisting scalars}
-Scalars are the easiest to hoist because we do not have to analyze
-them for aliasing memory references. As long as all the operands are
-available (which in SSA form we do, or may be there are chained
-dependencies), the scalar can be hoisted. Care must be taken in case
-of hoisting scalars too far, as that may increase register pressure
-and result in spills. For example hoisting a scalar past a call. In
-that case the call may result in save and restore of the register the
-scalar may be defined to. In our current implementation we do not
-hoist scalars past a call. Another way to mitigate this problem is be
-to reinstantiate (rematerialize) the computation after a call (may be
-as a different pass).
+Scalars are the easiest to hoist because we do not have to analyze them for
+aliasing memory references. As long as all the operands are available, the
+scalar can be hoisted. Care must be taken in case of hoisting scalars too far,
+as that may increase register pressure and result in spills. For example
+hoisting a scalar past a call (Section-\ref{cost:across-calls}). In our current
+implementation we hoist scalars past a call only when optimizing for code-side
+(-Os). Another way to mitigate this problem is be to reinstantiate
+(rematerialize) the computation after a call (may be as a different pass).
 
@@
 In order to value number stores, we value number (hash) the address from where
-the value is to be loaded. Then we need to compute the hoisting point just like
-we did for the scalars. At the hoisting point we compute if the operands to the
-load (an address) is available. If it is not available we can try to compute the
-addresss if that is possible.  Along the path from current position of the load
-instruction through the hoiting point, we find if there are aliasing stores, we
+the value is to be loaded. Then we need to compute the hoisting point similar to
+the hoisting of scalars. At the hoisting point the availability of operands to the
+load (an address) is checked. If it is not available we try to compute the
+addresss if that is possible.  Along the path, from current position of the load
+instruction through the hoisting point, we find if there are aliasing stores, we
 find if any of the basic blocks are landing pads or targets of indirect
@@
 register pressure by one (unless the operand to load becomes dead at the load),
-we evaluate if it is beneficial to hoist the load.
+we also evaluate if it is beneficial to hoist the load.
 
@@
 In order to value number stores, we value number (hash) the address as well as
-the value to be stored to that address. For the legality, we need to check the
-dependency requirements just like hoisting loads. We need to check that the
-operands of store are available at the hoisting point, that there are no
-aliasing loads or store along the path from the current position through the
-hoisting point and, none of the basic blocks along the path are landing pads.
-Once all the legality checks are satisfied, it is a good candidate to be
-hoisted.
+the value to be stored at that address. For the legality, we need to check the
+dependency requirements similar to the hoisting of loads. We need to check that
+the operands of the store instruction are available at the hoisting point, that
+there are no aliasing loads or store along the path from the current position
+through the hoisting point and, none of the basic blocks along the path have
+their address taken. Once all the legality checks are satisfied, it is a good
+candidate to be hoisted.
 
@@
 we have to check that there are no other side-effects between a call and
-its hoisting point, just like loads and stores.
+its hoisting point, just like we did for loads and stores.
 
@@
 \subsubsection{In the presence of calls}
+\label{cost:across-calls}
 Hoisting scalars across calls is tricky because it can increase the number of

### 2016-08-08 hoist.tex
calls

@@
 
+For legality we have to check that the side-effects of call does not intersect
+any side-effect between the instruciton to be hoisted and its hoisting point.
+
 \subsubsection{Hoisting scalars}
@@
 \subsubsection{Hoisting calls}
-Calls can be hoisted by value-numbering all the arguments. For safety
-we have to check that there are no other side-effects between a call and
-its hoisting point, just like we did for loads and stores.
+llvm has built in facility to value-number calls. If there are calls in sibling
+branches with identical value numbers, they can be hoisted. For legality we have
+to check that the loads/stores in the call does not create hazard with any
+loads/stores (in the caller) in the path from callsite through the hoisting
+point, just like we did for loads and stores.
 

### 2016-08-08 hoist.tex
overview of the algorithm

@@
 
-\subsection{Overview}
-
 Code hoisting can also reduce the critical path length of execution in out of
@@
 
-For legality we have to check that the side-effects of call does not intersect
-any side-effect between the instruciton to be hoisted and its hoisting point.
+\subsection{Overview}
+Similar to most compiler optimization passes the code-hoisting can also be
+divided into following five steps:
+
+\begin{enumerate}
+\item Checking for candidacy. Finding if a set of instructions perform identical
+  computation.
+\item Checking for legality. Finding a hoisting point i.e. a common dominator of
+  the set of instructions. If such a dominator is found then checking if all the
+  use-operands of the set of instructions are available or not.  Checking that
+  the side-effects of the computations does not intersect any side-effect
+  between the instructions to be hoisted and their hoisting point. In some cases
+  if the operands are not available, it is possible to reinstantiate
+  (remateralize) the use-operands, thus making the passing the legality check.
+\item Checking for profitability. Establishing a set of cost models tune for
+  performance. Like decreasing code size, minimizing spills etc.
+\item Making the transformation. Putting a copy of the computation at the
+  hoisting point and removing all the computations made redundant by the new
+  copy.
+\item Verification. Performing a set of checks (e.g., consistency of use-defs
+  and SSA semantics) to establish that program invariants are maintained.
+\end{enumerate}
 

### 2016-08-08 hoist.tex
restructure the overview of algo

--- 2016-08-08 hoist.tex
@@
 
-\newpage
-
-\section{Code hoisting}
-
-\subsection{Overview}
-
 Code hoisting can also reduce the critical path length of execution in out of
@@
 
-For legality we have to check that the side-effects of call does not intersect
-any side-effect between the instruciton to be hoisted and its hoisting point.
+\newpage
+
+\section{Code hoisting}
+Similar to most compiler optimization passes the code-hoisting pass can also be
+divided into the following five steps:
+
+\subsection{Finding the candidates}
+Finding if a set of instructions perform identical computations. Since this code
+hoisting is based on GVN, all we need to do is to compare the value number of
+computations.
+
+\subsection{Checking for legality}
+Finding a hoisting point i.e. a common dominator of the set of instructions. If
+such a dominator is found then checking if all the use-operands of the set of
+instructions are available or not. In some cases if the operands are not
+available, it is possible to reinstantiate (remateralize) the use-operands, thus
+making the passing the legality check. Checking that the side-effects of the
+computations does not intersect with any side-effect between the instructions to
+be hoisted and their hoisting point. For memory operations like loads, stores,
+calls etc., it is also required to check that all the paths from the hoisting
+point to the end of the function should have the exact memory operation
+available, in order to guarantee correctness.
 
@@
 
-\subsection{Algorithm}
-A good full page.
+\subsection{Checking for profitability}
+Establishing a set of cost models tune for performance. Like decreasing code
+size, minimizing spills etc.
 
-\subsubsection{Hoisting bottom up with optimistic approach}
+\subsection{Making the transformation}
+Putting a copy of the computation at the hoisting point and removing all the
+computations made redundant by the new copy.
+
+\subsection{Verification}
+Performing a set of checks (e.g., consistency of use-defs and SSA semantics) to
+establish that program invariants are maintained.
+
+\subsection{Hoisting bottom up with optimistic approach}
 Optimistic means collect the set of all potential redundancies and

### 2016-08-08 hoist.tex
finding the candidates

@@
 Finding if a set of instructions perform identical computations. Since this code
-hoisting is based on GVN, all we need to do is to compare the value number of
-computations.
+hoisting is based on GVN, we need the facility to compute GVN of instructions.
+In llvm \cite{llvm}, GVN infrastructure is already set up and works well for
+most kind of instructions except loads and store so we compute the GVN of loads
+and stores separately. In order to value number stores, we value number (hash)
+the address from where the value is to be loaded. For stores, we value number
+(hash) the address as well as the value to be stored at that address.
+
+The process of computing GVN can be on-demand (as we come across an instruction)
+or, precomputed (computing GVN of all the instructions beforehand). Which
+process to choose is determined by the scope of code-hoisting we want to
+perform. A pessimistic approach where we want to hoist limited set of
+instructions from the sibing branches as we iterate the DFS tree bottom-up, it
+is sufficient to compute values on-demand. Whereas, in the optimistic approach,
+where we want to hoist as many instructions as possible, would require values to
+be precomputed.
+
+Whatever approach we take, to find a candidate all we need to do is to compare
+the value number of computations.
 

### 2016-08-08 hoist.tex
profitability

@@
 \subsection{Checking for legality}
-Finding a hoisting point i.e. a common dominator of the set of instructions. If
+Finding a hoisting point i.e., a common dominator of the set of instructions. If
 such a dominator is found then checking if all the use-operands of the set of
@@
 aliasing memory references. As long as all the operands are available, the
-scalar can be hoisted. Care must be taken in case of hoisting scalars too far,
-as that may increase register pressure and result in spills. For example
-hoisting a scalar past a call (Section-\ref{cost:across-calls}). In our current
-implementation we hoist scalars past a call only when optimizing for code-side
-(-Os). Another way to mitigate this problem is be to reinstantiate
-(rematerialize) the computation after a call (may be as a different pass).
+scalar can be hoisted.
 
 \subsubsection{Hoisting loads}
-In order to value number stores, we value number (hash) the address from where
-the value is to be loaded. Then we need to compute the hoisting point similar to
-the hoisting of scalars. At the hoisting point the availability of operands to the
-load (an address) is checked. If it is not available we try to compute the
-addresss if that is possible.  Along the path, from current position of the load
-instruction through the hoisting point, we find if there are aliasing stores, we
-find if any of the basic blocks are landing pads or targets of indirect
-branches. In those cases we discard the candidate. Since the load increases the
-register pressure by one (unless the operand to load becomes dead at the load),
-we also evaluate if it is beneficial to hoist the load.
+At the hoisting point the availability of operands to the load (an address) is
+checked. If it is not available we try to compute the addresss if that is
+possible.  Along the path, from current position of the load instruction through
+the hoisting point, we find if there are aliasing stores, we find if any of the
+basic blocks are landing pads or targets of indirect branches. In those cases we
+discard the candidate.
 
 \subsubsection{Hoisting stores}
-In order to value number stores, we value number (hash) the address as well as
-the value to be stored at that address. For the legality, we need to check the
-dependency requirements similar to the hoisting of loads. We need to check that
-the operands of the store instruction are available at the hoisting point, that
-there are no aliasing loads or store along the path from the current position
-through the hoisting point and, none of the basic blocks along the path have
-their address taken. Once all the legality checks are satisfied, it is a good
-candidate to be hoisted.
+We need to check the dependency requirements similar to the hoisting of
+loads. We need to check that the operands of the store instruction are available
+at the hoisting point, that there are no aliasing loads or store along the path
+from the current position through the hoisting point and, none of the basic
+blocks along the path have their address taken.
 
 \subsubsection{Hoisting calls}
-llvm has built in facility to value-number calls. If there are calls in sibling
-branches with identical value numbers, they can be hoisted. For legality we have
-to check that the loads/stores in the call does not create hazard with any
-loads/stores (in the caller) in the path from callsite through the hoisting
+We have to check that the loads/stores in the call does not create hazard with
+any loads/stores (in the caller) in the path from callsite through the hoisting
 point, just like we did for loads and stores.
@@
 size, minimizing spills etc.
+
+\subsubsection{Hoisting scalars}
+Care must be taken in case of hoisting scalars too far, as that may increase
+register pressure and result in spills. For example hoisting a scalar past a
+call (Section-\ref{cost:across-calls}). In our current implementation we hoist
+scalars past a call only when optimizing for code-side (-Os). Another way to
+mitigate this problem is be to reinstantiate (rematerialize) the computation
+after a call (may be as a different pass).
+
+\subsubsection{Hoisting loads}
+A load instruction introduces a register where the value loaded will be kept,
+the register pressure increases by one (unless the operand to load becomes dead
+at the load). On the other hand, loading a value early will reduce the stall
+during execution should the value is not in the cache. We generally prefer to
+hoist load except the hoisting point is too far (this distance is computed by
+looking at the experimental results of representative benchmarks see
+Section-\ref{sec:experimental-results}).
+
+\subsubsection{Hoisting stores}
+Since stores do not increase the live-range of any registers, and in some cases
+it ends the liveness of registers, we hoist all the stores.
+
+\subsubsection{Hoisting calls}
+Currently we hoist all the calls.
 

### 2016-08-08 hoist.tex
legality

@@
 process to choose is determined by the scope of code-hoisting we want to
-perform. A pessimistic approach where we want to hoist limited set of
-instructions from the sibing branches as we iterate the DFS tree bottom-up, it
-is sufficient to compute values on-demand. Whereas, in the optimistic approach,
-where we want to hoist as many instructions as possible, would require values to
-be precomputed.
+perform. In a pessimistic approach (Section-\ref{subsec:pessimistic}) where we
+want to hoist limited set of instructions from the sibing branches as we iterate
+the DFS tree bottom-up, it is sufficient to compute values on-demand. Whereas,
+in the optimistic approach (Section-\ref{subsec:optimistic}), where we want to
+hoist as many instructions as possible, would require values to be precomputed.
 
@@
 available, it is possible to reinstantiate (remateralize) the use-operands, thus
-making the passing the legality check. Checking that the side-effects of the
-computations does not intersect with any side-effect between the instructions to
-be hoisted and their hoisting point. For memory operations like loads, stores,
-calls etc., it is also required to check that all the paths from the hoisting
-point to the end of the function should have the exact memory operation
-available, in order to guarantee correctness.
+passing the legality check. Checking that the side-effects of the computations
+does not intersect with any side-effects between the instructions to be hoisted
+and their hoisting point. For memory operations like loads, stores, calls etc.,
+it is also required to check that all the paths from the hoisting point to the
+end of the function should execute the exact instruction, in order to guarantee
+correctness. Moreover, hoisting of memory operations should not be done on paths
+which have indirect branch targets e.g., landing pad, case statements, goto
+labels etc., because it becomes tricky to prove that all the paths from hoisting
+point to the end of the function would execute the instruction.
 
@@
 possible.  Along the path, from current position of the load instruction through
-the hoisting point, we find if there are aliasing stores, we find if any of the
-basic blocks are landing pads or targets of indirect branches. In those cases we
+the hoisting point, we find if there are aliasing stores. In those cases we
 discard the candidate.
@@
 at the hoisting point, that there are no aliasing loads or store along the path
-from the current position through the hoisting point and, none of the basic
-blocks along the path have their address taken.
+from the current position through the hoisting point.
 
 \subsubsection{Hoisting calls}
-We have to check that the loads/stores in the call does not create hazard with
-any loads/stores (in the caller) in the path from callsite through the hoisting
-point, just like we did for loads and stores.
+For hoisting the call instructions, we check that the loads/stores in the call
+does not create hazard with any loads/stores (in the caller) in the path from
+callsite through the hoisting point, just like we did for loads and stores.
 
@@
 mitigate this problem is be to reinstantiate (rematerialize) the computation
-after a call (may be as a different pass).
+after a call (may be as a different optimization pass).
 
@@
 \subsection{Hoisting bottom up with optimistic approach}
+\label{subsec:optimistic}
 Optimistic means collect the set of all potential redundancies and
@@
 \subsubsection{Pessimistic approach}
+\label{subsec:pessimistic}
 To extend the triangle approach to the entire function, the algorithm traverses

### 2016-08-08 hoist.tex
contrib
partition

@@
 
+\subsection{Contributions of this paper}
+\begin{itemize}
+\item A systematic approach to writing a compiler optimization pass.
+\item gvn hoisting
+\item partition algorithm
+\item cost model to reduce live-range and reduce spills
+\end{itemize}
+
 \section{Related Work}
@@
 \section{Code hoisting}
-Similar to most compiler optimization passes the code-hoisting pass can also be
-divided into the following five steps:
+For the code hoisting we follow a very systematic approach to writing a compiler
+optimization pass. As such this pass can be divided into the following five
+steps:
 
@@
 end of the function should execute the exact instruction, in order to guarantee
-correctness. Moreover, hoisting of memory operations should not be done on paths
-which have indirect branch targets e.g., landing pad, case statements, goto
-labels etc., because it becomes tricky to prove that all the paths from hoisting
-point to the end of the function would execute the instruction.
+correctness. Moreover, hoisting of memory operations is tricky on paths which
+have indirect branch targets e.g., landing pad, case statements, goto labels
+etc., because it becomes difficult to prove that all the paths from hoisting
+point to the end of the function would execute the instruction. In our current
+implementation we discard hoisting on such paths.
+
+In the optimistic approach (Section-\ref{subsec:optimistic}), it is possible
+that a common hoisting point of all the instructions is either too far away, or
+not possible. In these cases, it is still possible to `partially' hoist a subset
+of instructions by splitting the set of candidates and finding a closer hoisting
+point for each subset. For more details see Section-\ref{subsec:partition}.
 
@@
 \subsubsection{Hoisting loads}
-At the hoisting point the availability of operands to the load (an address) is
-checked. If it is not available we try to compute the addresss if that is
+The availability of operands to the load (an address) is checked at the hoisting
+point. If it is not available we try to rematerialize the addresss if that is
 possible.  Along the path, from current position of the load instruction through
@@
 computations a partially hoist a set of computations to their common
-dominator. For more details see Section~\ref{subsec:partition}.
+dominator. For more details see Section-\ref{subsec:partition}.
 

### 2016-08-08 hoist.tex
add more text on introduction

@@
 
-Classical scalar optimizations like CSE \cite{dragonbook} work very well on
-single basic blocks.  With time, we have seen the transition to more complex
-passes like GCSE that work across basic blocks, or other passes like the
-macroblock formation that create a region within
+Compiler techniques to remove redundant computations are composed of an analysis
+phase that detects identical computations in the program and a transformation
+phase that reduces the number of run-time computations.  Classical scalar
+optimizations like CSE \cite{dragonbook} work very well on single basic blocks.
+When it comes to detect redundancies across basic blocks these techniques fall
+short: more complex passes like GCSE and PRE have been designed to handle these
+cases based on dataflow analysis \cite{morel1979global}.  At first these
+techniques were described in the classical data-flow analysis framework, and
+later the use of the SSA representation lowered the cost in terms of compilation
+time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
+techniques in the main stream: nowadays SSA based PRE is available in every
+industrial compiler.
 
-Why hoisting expressions:
+This paper describes code hoisting, a technique that uses the information
+computed for PRE to detect identical computations and has a transformation phase
+whose goal differs from PRE: it removes identical computations from different
+branches of execution.  These identical computations would not be redundant at
+run-time and thus the number of run-time computations is not reduced.  Instead,
+the goals of code hoisting are:
 \begin{itemize}
-\item OOO processors: good for speculative execution of branches
-\item more ILP for in-order processors
-\item code size
-\item optimize for register pressure
-\item improve inlining heuristics
+\item to reduce the code size of the program,
+\item to expose more instruction level parallelism by hoisting identical
+  computations to be executed earlier, and thus helping to schedule heavy
+  computations earlier in order to avoid pipeline bubbles,
+\item to help out-of-order processors with speculative execution of branches: by
+  hoisting expressions out of branches, code hoisting can effectively reduce
+  the amount of code to be speculatively executed,
+\item to decrease the number of live variables and thus to reduce register
+  pressure,
+\item to improve function inlining heuristics by reducing the size of a function body.
 \end{itemize}

### 2016-08-08 hoist.tex
calls

@@
 \subsubsection{Hoisting calls}
-Currently we hoist all the calls.
+Currently we hoist all the calls that are suitable candidates for hoisting.
 
 \subsection{Making the transformation}
-Putting a copy of the computation at the hoisting point and removing all the
-computations made redundant by the new copy.
+Once all the legality checks and profitability checks are satisfied, a copy of
+the computation is inserted at the hoisting point and all the computations made
+redundant by the new copy are removed. After that SSA semantics is restored by
+updating the intermediate representation (IR) to reflect changes.
 

### 2016-08-08 hoist.tex
more on intro

@@
 \begin{itemize}
-\item to reduce the code size of the program,
-\item to expose more instruction level parallelism by hoisting identical
-  computations to be executed earlier, and thus helping to schedule heavy
-  computations earlier in order to avoid pipeline bubbles,
+\item to reduce the code size of the program;
+\item to improve function inlining heuristics: functions become cheaper to
+  inline by reducing their code size;
+\item to expose more instruction level parallelism: by hoisting identical
+  computations to be executed earlier, schedulers can move heavy computations
+  earlier in order to avoid pipeline bubbles;
 \item to help out-of-order processors with speculative execution of branches: by
-  hoisting expressions out of branches, code hoisting can effectively reduce
-  the amount of code to be speculatively executed,
-\item to decrease the number of live variables and thus to reduce register
-  pressure,
-\item to improve function inlining heuristics by reducing the size of a function body.
+  hoisting expressions out of branches, code hoisting can effectively reduce the
+  amount of code to be speculatively executed and can reduce the critical path;
+\item to reduce register pressure: by moving computations closer to the
+  definitions of their operands;
+\item to improve passes that do not work well with branches:
+  \begin{itemize}
+  \item to improve loop vectorization by reducing a loop with control flow to a
+    loop with a single BB, should all the instructions in a conditional get
+    hoisted and sinked;
+  \item to enable more loop invariant code motion (LICM): as LICM does not
+    reason about instructions in the context of loops with conditional branches,
+    code-hoisting is needed to move instructions out of conditional expressions
+    and expose them to LICM.
+  \end{itemize}
 \end{itemize}
 
-\begin{enumerate}
-\item Code hoisting in adjacent blocks in a diamond/triangle like structure.
-\item O(n) Algorithm.
-\item Hoisting equivalent computations in non-sibling BBs.
-\item Hoisting across switch blocks.
-\item Hoisting to improve PRE.
-\item Hoisting to reduce critical path in OoO execution.
-\item Ranking/Wavefront algorithm to hoist across switch blocks.
-\end{enumerate}
-
-\subsection{Enabling LICM by code-hoisting}
-
-\begin{verbatim}
-loop:
-if (a)
-  i1;
-else
-  i1;
-\end{verbatim}
-
-Here i1 is redundant and can be hoisted out of loop. But the LICM will
-not do because it does not reason about instructions in the
-conditional.  By code-hoisting i1 will be hoisted out of conditional
-which will benefit the LICM.
-
-\subsection{Improving the Inliner heuristics}
-Reducing the number of instructions also reduces the inlining cost and
-hence improves the inliner heuristics.
-
-\subsection{Improving vectorization}
-It will also benefit vectorizer by reducing the number
-of basic blocks, should all the instructions in conditional get hoisted.
+The complexity of code hoisting is linear in number of instructions that could
+be hoisted in the program, matching the complexity of PRE on SSA form.  The
+analysis phase is based on the Global Value Numbering (GVN), the same analysis
+used for PRE, followed by the computation of a partition of identical
+expressions to be hoisted in a same location to guarantee safety properties and
+program performance, and followed by a simple code generation that adds the
+identified instruction in the hoisting point and removes all the now redundant
+expressions.
 
@@
 
+\begin{itemize}
+\item Code hoisting in adjacent blocks in a diamond/triangle like structure.
+\item Hoisting equivalent computations in non-sibling BBs.
+\item Hoisting across switch blocks.
+\item Ranking/Wavefront algorithm to hoist across switch blocks.
+\end{itemize}
+
 \section{Related Work}
@@
 A lot of bug reports in GCC and LLVM bugzillas, no papers on this topic yet.
-
-Code hoisting does not remove redundancies by itself but it enables
-removal of redundancies by other passes.  e.g., Enable inlining.
 
@@
 In llvm \cite{llvm}, GVN infrastructure is already set up and works well for
-most kind of instructions except loads and store so we compute the GVN of loads
-and stores separately. In order to value number stores, we value number (hash)
+most kind of instructions except loads and stores so we compute the GVN of loads
+and stores separately. In order to value number loads, we value number (hash)
 the address from where the value is to be loaded. For stores, we value number
@@
 such a dominator is found then checking if all the use-operands of the set of
-instructions are available or not. In some cases if the operands are not
+instructions are available or not. In some cases when the operands are not
 available, it is possible to reinstantiate (remateralize) the use-operands, thus
@@
 point to the end of the function would execute the instruction. In our current
-implementation we discard hoisting on such paths.
+implementation we discard hoisting through such paths.
 
@@
 that a common hoisting point of all the instructions is either too far away, or
-not possible. In these cases, it is still possible to `partially' hoist a subset
-of instructions by splitting the set of candidates and finding a closer hoisting
-point for each subset. For more details see Section-\ref{subsec:partition}.
+not legally possible. In these cases, it is still possible to `partially' hoist
+a subset of instructions by splitting the set of candidates and finding a closer
+hoisting point for each subset. For more details see
+Section-\ref{subsec:partition}.
 
@@
 \subsubsection{Hoisting loads}
-The availability of operands to the load (an address) is checked at the hoisting
-point. If it is not available we try to rematerialize the addresss if that is
+The availability of operand to the load (an address) is checked at the hoisting
+point. If that is not available we try to rematerialize the addresss if
 possible.  Along the path, from current position of the load instruction through
-the hoisting point, we find if there are aliasing stores. In those cases we
-discard the candidate.
+the hoisting point, we find if there are aliasing writes to memory. In those
+cases we discard the candidate.
 
@@
 For hoisting the call instructions, we check that the loads/stores in the call
-does not create hazard with any loads/stores (in the caller) in the path from
+would not hazard with any loads/stores (in the caller) along the path from the
 callsite through the hoisting point, just like we did for loads and stores.
@@
 \subsection{Making the transformation}
-Once all the legality checks and profitability checks are satisfied, a copy of
-the computation is inserted at the hoisting point and all the computations made
-redundant by the new copy are removed. After that SSA semantics is restored by
-updating the intermediate representation (IR) to reflect changes.
+Once all the legality checks and profitability checks are satisfied for a set of
+identical instructions, they are suitable candidates for hoisting. A copy of the
+computation is inserted at the hoisting point along with any instructions which
+needed to be rematerialized. Thereafter, all the computations made redundant by
+the new copy are removed. After that SSA semantics is restored by updating the
+intermediate representation (IR) to reflect changes.
 

### 2016-08-09 hoist.tex
fully redundant
reorganize policies and cost model

--- 2016-08-08 hoist.tex
@@
 \item to reduce the code size of the program;
+\item to remove fully redundant instructions: although this is also done by
+  GVN-PRE.
 \item to improve function inlining heuristics: functions become cheaper to
@@
 
-\subsection{Verification}
+\subsection{Verifying the post-conditions}
 Performing a set of checks (e.g., consistency of use-defs and SSA semantics) to
@@
 
+\section{Code hoisting policies}
+The amount of hoisting depends on the whether we collect information about all
+the instructions before finding candidates (optimistic), or on demand
+(pessimistic). It also depends on the generality of the GVN algorithm, however,
+that analysis is beyond the scope of this paper.
+
 \subsection{Hoisting bottom up with optimistic approach}
 \label{subsec:optimistic}
-Optimistic means collect the set of all potential redundancies and
-then discard the bad ones.
+In this approach, the goal is to maximise the total number of hoistings in the
+entire function.  This algorithm is very useful when optimizing for code-size.
+In order to do that, we collect the GVN of all the instructions in the
+function. After that we iterate on the list of instructions having identical
+GVNs. After that we find the common dominator dominating all the identical
+computations and try to hoist the instruction in that basic block, if
+feasible. Many times it is not possible to hoist all the instructions to one
+common dominator, due to legality/profitability constraints. In those cases, the
+algorithmn would partion the list of computations and partially hoist a set of
+computations to their common dominator. For more details see
+Section-\ref{subsec:partition}.
 
@@
 
-\subsection{Cost models}
+\subsubsection{Triangle approach}
+This is the most basic approach to code-hoisting when the instructions are
+hoisted to immediate parent in a diamond like structure. When there are
+identical computations in both the branches of a conditional, and all the
+legality checks are met, the instruction can be hoisted to the predecessor basic
+block. This approach is simple to implement and has minimal impact on the
+register pressure.
+
+\subsubsection{Pessimistic approach}
+\label{subsec:pessimistic}
+To extend the triangle approach to the entire function, the algorithm traverses
+the basic blocks in the invers depth-first order. All the leaf nodes are visited
+before the non-leaf nodes because instructions are hoisted upwards. We start
+from a branch and collect the GVN of all the instructions. Then in the sibling
+basic block we find if there are any identical computations. Once we find an
+identical computation we try to hoist it to the nearest common dominator which
+is the immediate predecessor of both the basic blocks. We traverse the basic
+block from top to down so that the once an instruction is hoisted it creates
+opportunities for other dependent instructions. After all the sibling basic
+blocks have all the identical computations hoisted, we continue the hoisting
+process with their parent and its sibling of their parent until we reach the
+beginning of the function.
+
+\subsubsection{PRE based hoisting}
+
+\section{Cost models}
 
@@
 
-\subsection{Other Approaches}
-
-\subsubsection{Triangle approach}
-This is the most basic approach to code-hoisting when the instructions are
-hoisted to immediate parent in a diamond like structure. When there are
-identical computations in both the branches of a conditional, and all the
-legality checks are met, the instruction can be hoisted to the predecessor basic
-block. This approach is simple to implement and has minimal impact on the
-register pressure.
-
-\subsubsection{Pessimistic approach}
-\label{subsec:pessimistic}
-To extend the triangle approach to the entire function, the algorithm traverses
-the basic blocks in the invers depth-first order. All the leaf nodes are visited
-before the non-leaf nodes because instructions are hoisted upwards. We start
-from a branch and collect the GVN of all the instructions. Then in the sibling
-basic block we find if there are any identical computations. Once we find an
-identical computation we try to hoist it to the nearest common dominator which
-is the immediate predecessor of both the basic blocks. We traverse the basic
-block from top to down so that the once an instruction is hoisted it creates
-opportunities for other dependent instructions. After all the sibling basic
-blocks have all the identical computations hoisted, we continue the hoisting
-process with their parent and its sibling of their parent until we reach the
-beginning of the function.
-
-\subsubsection{Optimistic approach}
-In this approach, the goal is to maximise the total number of hoistings in the
-entire function.  This algorithm is very useful when optimizing for code-size.
-We see a lot of computations hoisted (see
-Section~\ref{sec:experimental-results}. In order to do that, we collect the GVN
-of all the instructions in the function. After that we iterate on the list of
-instructions having identical GVN. After that we find the common dominator
-dominating all the identical computations and try to hoist the instruction in
-that basic block, if feasible. Many times it is not possible to hoist all the
-instructions to one common dominator, due to legality/profitability
-constraints. In those cases, the algorithmn would partion the list of
-computations a partially hoist a set of computations to their common
-dominator. For more details see Section-\ref{subsec:partition}.
-
-\subsubsection{pros,cons}
-Naturally, the amount of hoisting depends on the generality of the GVN
-algorithm. Current llvm GVN implementation finds out equivalence between
-dependent instructions to some extent.
-
-\subsubsection{PRE based hoisting}
-
 \newpage

### 2016-08-09 hoist.tex
optimistic

@@
 \subsection{Checking for legality}
+\label{subsec:legality}
 Finding a hoisting point i.e., a common dominator of the set of instructions. If
@@
 \subsection{Checking for profitability}
+\label{subsec:profitability}
 Establishing a set of cost models tune for performance. Like decreasing code
@@
 entire function.  This algorithm is very useful when optimizing for code-size.
-In order to do that, we collect the GVN of all the instructions in the
-function. After that we iterate on the list of instructions having identical
-GVNs. After that we find the common dominator dominating all the identical
-computations and try to hoist the instruction in that basic block, if
-feasible. Many times it is not possible to hoist all the instructions to one
-common dominator, due to legality/profitability constraints. In those cases, the
-algorithmn would partion the list of computations and partially hoist a set of
-computations to their common dominator. For more details see
-Section-\ref{subsec:partition}.
+In order to do that, we collect the GVN of all the instructions in the function
+and iterate on the list of instructions having identical GVNs. After that we
+find the common dominator dominating all the identical computations and perform
+legality checks (Section-\ref{subsec:legality}). Many times it is not possible
+to hoist all the instructions to one common dominator, due to legality
+constraints e.g., intersecting side-effects or, profitability constraints e.g.,
+hoisting point too far. In those cases, the algorithmn would partition the list
+of computations and partially hoist a set of computations to their common
+dominator. For more details see Section-\ref{subsec:partition}.
+
+\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
+\label{subsec:partition}
+The set of candidates are sorted in the order of their DFS-in
+numbers. So that candidates close to each other in the list are also
+closer (??) in the CFG.
+
+In order to hoist a subset of identical instructions, we partition the list of
+all candidates in a way to maximise the total number of hoistings.  By sorting
+the list of all the candidates in the increasing order of their DFS-in numbers,
+we make sure that candidates closer in the list have their common dominator
+nearby in cases when there are no fully redundant instructions.
+
+Essentially,
+abs(DFSIn(I1) - DFSIn(I2)) < abs(DFSIn(I1) - DFSIn(I3))
+=> bbDist(I1, nearest-common-dom(I1, I2)) <= bbDist(I1, nearest-common-dom(I1, I3))
+
+In case of fully redundant computations the nearby elements may not be 'closer'
+than the separated ones. TODO: Explain.
+
+This means for candidates closer to each other w.r.t. DFS numbers:
+\begin{enumerate}
+\item would make fewer checks for legality and profitability.
+\item hoisting point will be closer so, the intersection of live-range of the
+  instruction with other instructions will be minimal.
+\end{enumerate}
+
+In the case where all the candidates cannot be hoisted at a common hoist point
+because of:
+\begin{enumerate}
+\item Safety conditions are not met.
+\item Live range would increase too much.
+\end{enumerate}
+
+In our current implementation we partition the list at the point where the
+legality checks fail.
+
+TODO: Put an example to illustrate the point.
 
@@
 GCC: 23286 has interesting test cases.
-
-\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
-\label{subsec:partition}
-The set of candidates are sorted in the order of their DFS-in
-numbers. So that candidates close to each other in the list are also
-closer (??) in the CFG.
-
-In the case where all the candidates cannot be hoisted at a common
-hoist point because of:
-\begin{enumerate}
-\item Safety conditions are not met.
-\item Live range would increase too much.
-\end{enumerate}
-
-In order to hoist a subset of identical instructions, we partition the
-list of all candidates in a way to maximise the total number of
-hoistings.  By sorting the list of all the candidates in the
-increasing order of their DFS-in numbers, we make sure that candidates
-closer in the list have their common dominator nearby.
-
-Essentially,
-abs(DFSIn(I1) - DFSIn(I2)) < abs(DFSIn(I1) - DFSIn(I3))
-=> bbDist(I1, nearest-common-dom(I1, I2)) <= bbDist(I1, nearest-common-dom(I1, I3))
-
-This means for candidates closer to each other w.r.t. DFS numbers:
-\begin{enumerate}
-\item would make fewer checks for legality and profitability.
-\item hoisting point will be closer so, the intersection of live-range of the
-  instruction with other instructions will be minimal.
-\end{enumerate}
-
-In our current implementation we partition the list at the point where the
-legality/profitability checks fail.
-
-TODO: Put an example to illustrate the point.
 

### 2016-08-09 hoist.tex
time complexity moved down, triangle and pessimistic removed

@@
 
-The complexity of code hoisting is linear in number of instructions that could
-be hoisted in the program, matching the complexity of PRE on SSA form.  The
-analysis phase is based on the Global Value Numbering (GVN), the same analysis
-used for PRE, followed by the computation of a partition of identical
-expressions to be hoisted in a same location to guarantee safety properties and
-program performance, and followed by a simple code generation that adds the
-identified instruction in the hoisting point and removes all the now redundant
-expressions.
-
 \subsection{Contributions of this paper}
 \begin{itemize}
-\item A systematic approach to writing a compiler optimization pass.
-\item gvn hoisting
+\item hoisting of identical computations
 \item partition algorithm
+\item code hoisting can remove fully redundant computations
+\item code hoisting can actually improve performance in some cases
 \item cost model to reduce live-range and reduce spills
-\end{itemize}
-
-\begin{itemize}
-\item Code hoisting in adjacent blocks in a diamond/triangle like structure.
-\item Hoisting equivalent computations in non-sibling BBs.
-\item Hoisting across switch blocks.
-\item Ranking/Wavefront algorithm to hoist across switch blocks.
 \end{itemize}
@@
 The amount of hoisting depends on the whether we collect information about all
-the instructions before finding candidates (optimistic), or on demand
-(pessimistic). It also depends on the generality of the GVN algorithm, however,
-that analysis is beyond the scope of this paper.
+the instructions before finding candidates (optimistic).
+
+In the pessimistic approach, the basic blocks are traversed in the inverse
+depth-first order, computing the GVN of instructions as they come by. The GVN of
+sibling branches are compared for equality. Once such a candidate is found, it
+is hoisted in the common dominator. All the leaf nodes are visited before the
+non-leaf nodes (bottom-up) because instructions are hoisted upwards. The
+pessimistic algorithm is fast, results in fewer spills but hoists very less
+instructions. We have implemented optimistic approach because it is more general
+and can be tuned down to closely mimic pessimistic approach by changing a flag.
+
+It also depends on the generality of the GVN algorithm, however, that analysis
+is beyond the scope of this paper.
 
@@
 entire function.  This algorithm is very useful when optimizing for code-size.
-In order to do that, we collect the GVN of all the instructions in the function
-and iterate on the list of instructions having identical GVNs. After that we
-find the common dominator dominating all the identical computations and perform
-legality checks (Section-\ref{subsec:legality}). Many times it is not possible
-to hoist all the instructions to one common dominator, due to legality
-constraints e.g., intersecting side-effects or, profitability constraints e.g.,
-hoisting point too far. In those cases, the algorithmn would partition the list
-of computations and partially hoist a set of computations to their common
-dominator. For more details see Section-\ref{subsec:partition}.
+We collect the GVN of all the instructions in the function and iterate on the
+list of instructions having identical GVNs. After that we find the common
+dominator dominating all such identical computations and perform legality checks
+(Section-\ref{subsec:legality}). Many times it is not possible to hoist all the
+instructions to one common dominator, due to legality constraints e.g.,
+intersecting side-effects or, profitability constraints e.g., hoisting point too
+far. In those cases, this algorithmn would partition the list of identical
+computations and partially hoist a set of computations to their common
+dominator. The partition algorithm is described as follows:
 
@@
 \label{subsec:partition}
-The set of candidates are sorted in the order of their DFS-in
-numbers. So that candidates close to each other in the list are also
-closer (??) in the CFG.
+The set of candidates are sorted in the order of their DFS-in numbers. So that
+candidates close to each other in the list are also closer (??) in the CFG.
 
@@
 
-\subsubsection{Triangle approach}
-This is the most basic approach to code-hoisting when the instructions are
-hoisted to immediate parent in a diamond like structure. When there are
-identical computations in both the branches of a conditional, and all the
-legality checks are met, the instruction can be hoisted to the predecessor basic
-block. This approach is simple to implement and has minimal impact on the
-register pressure.
-
-\subsubsection{Pessimistic approach}
-\label{subsec:pessimistic}
-To extend the triangle approach to the entire function, the algorithm traverses
-the basic blocks in the invers depth-first order. All the leaf nodes are visited
-before the non-leaf nodes because instructions are hoisted upwards. We start
-from a branch and collect the GVN of all the instructions. Then in the sibling
-basic block we find if there are any identical computations. Once we find an
-identical computation we try to hoist it to the nearest common dominator which
-is the immediate predecessor of both the basic blocks. We traverse the basic
-block from top to down so that the once an instruction is hoisted it creates
-opportunities for other dependent instructions. After all the sibling basic
-blocks have all the identical computations hoisted, we continue the hoisting
-process with their parent and its sibling of their parent until we reach the
-beginning of the function.
-
-\subsubsection{PRE based hoisting}
+\subsubsection{Time complexity of algorithm}
+The complexity of code hoisting is linear in number of instructions that could
+be hoisted in the program, matching the complexity of PRE on SSA form.  The
+analysis phase is based on the Global Value Numbering (GVN), the same analysis
+used for PRE, followed by the computation of a partition of identical
+expressions to be hoisted in a same location to guarantee safety properties and
+program performance, and followed by a simple code generation that adds the
+identified instruction in the hoisting point and removes all the now redundant
+expressions.
 

### 2016-08-09 hoist.tex
remove PRE related comment
reword contributions

@@
 \item to reduce the code size of the program;
-\item to remove fully redundant instructions: although this is also done by
-  GVN-PRE.
 \item to improve function inlining heuristics: functions become cheaper to
@@
 
-\subsection{Contributions of this paper}
+The main contributions of this paper are:
 \begin{itemize}
-\item hoisting of identical computations
-\item partition algorithm
-\item code hoisting can remove fully redundant computations
-\item code hoisting can actually improve performance in some cases
+\item a new algorithm to hoist computations from branches
 \item cost model to reduce live-range and reduce spills
+\item performance evaluation of the implementation in LLVM
+\item comparison against other algorithms for code hoisting
 \end{itemize}

### 2016-08-09 hoist.tex
policies and partition algo

@@
 \section{Code hoisting policies}
-The amount of hoisting depends on the whether we collect information about all
-the instructions before finding candidates (optimistic).
+The amount of hoisting depends on whether we collect GVN of instructions
+on-demand (pessimistic) or, before finding candidates (optimistic).It also
+depends on the generality of the GVN algorithm, however, that analysis is beyond
+the scope of this paper.
 
+\subsection{Hoisting bottom up with pessimistic approach}
+\label{subsec:pessimistic}
 In the pessimistic approach, the basic blocks are traversed in the inverse
@@
 is hoisted in the common dominator. All the leaf nodes are visited before the
-non-leaf nodes (bottom-up) because instructions are hoisted upwards. The
-pessimistic algorithm is fast, results in fewer spills but hoists very less
-instructions. We have implemented optimistic approach because it is more general
-and can be tuned down to closely mimic pessimistic approach by changing a flag.
-
-It also depends on the generality of the GVN algorithm, however, that analysis
-is beyond the scope of this paper.
+non-leaf nodes (bottom-up) because instructions are hoisted upwards.
 
@@
 far. In those cases, this algorithmn would partition the list of identical
-computations and partially hoist a set of computations to their common
-dominator. The partition algorithm is described as follows:
+instructions into subsets which can be partially hoisted to their respective
+common dominators. The partition algorithm is described as follows:
 
@@
 \label{subsec:partition}
-The set of candidates are sorted in the order of their DFS-in numbers. So that
-candidates close to each other in the list are also closer (??) in the CFG.
-
 In order to hoist a subset of identical instructions, we partition the list of
@@
 GCC: 23286 has interesting test cases.
+
+The pessimistic algorithm is fast, results in fewer spills but hoists very less
+instructions. We have implemented optimistic approach because it is more general
+and can be tuned down to closely mimic pessimistic approach by changing a flag.
 

### 2016-08-09 hoist.tex
fix

@@
   block.  Code hoisting could also be confused with Partial Redundancy
-  Elimination (PRE) or even with Global Common Subexpression Elimination (GCSE)
+  Elimination (PRE) or with Global Common Subexpression Elimination (GCSE)
   although the main goal of code hoisting is not to remove redundancies: code
@@
 branches of execution.  These identical computations would not be redundant at
-run-time and thus the number of run-time computations is not reduced.  Instead,
-the goals of code hoisting are:
+run-time and as the number of run-time computations is not reduced, code
+hoisting is not a redundancy elimination pass.  The goals of code hoisting are:
 \begin{itemize}

### 2016-08-10 hoist.tex
partition

@@
 
-In our current implementation we partition the list at the point where the
-legality checks fail.
+In our current implementation we keep as many candidates in one set as possible
+unless (greedy approach). We split the list at a point where the legality checks
+fail to hoist subset of candidates which are legal to hoist and, start finding
+new hoisting point for the remaining ones.
 
 TODO: Put an example to illustrate the point.
-
-Bottom up: Sort the basic blocks of potentially redundant computations
-based on DFS-in numbers.  TODO: Explain why sorting by DFS-in numbers.
-Then start evaluating the safety checks for first two, if good then
-move to second one. That way we can even partially hoist computations
-should we find that we cannot remove all the redundant computations.
-
-We can also split the set of to partially hoist multiple times, if
-hoisting globally is not possible.
 

### 2016-08-10 hoist.tex
dfs depth

@@
 
-Essentially,
-abs(DFSIn(I1) - DFSIn(I2)) < abs(DFSIn(I1) - DFSIn(I3))
-=> bbDist(I1, nearest-common-dom(I1, I2)) <= bbDist(I1, nearest-common-dom(I1, I3))
+\begin{verbatim}
+Essentially if,
+DFS-I1I2 = abs(DFSIn(I1) - DFSIn(I2))
+DFS-I1I3 = abs(DFSIn(I1) - DFSIn(I3))
+BBDistI1I2 = bbDist(I1, nearest-common-dom(I1, I2))
+BBDistI1I3 = bbDist(I1, nearest-common-dom(I1, I3))
+
+Then,
+DFS-I1I2 < DFS-I1I3 => BBDistI1I2 <= BBDistI1I3
+\end{verbatim}
 

### 2016-08-11 hoist.tex
formalisatio of partition algorithm

--- 2016-08-10 hoist.tex
@@
 dominator dominating all such identical computations and perform legality checks
-(Section-\ref{subsec:legality}). Many times it is not possible to hoist all the
+(Section-\ref{subsec:legality}). Often times it is not possible to hoist all the
 instructions to one common dominator, due to legality constraints e.g.,
@@
 all candidates in a way to maximise the total number of hoistings.  By sorting
-the list of all the candidates in the increasing order of their DFS-in numbers,
-we make sure that candidates closer in the list have their common dominator
-nearby in cases when there are no fully redundant instructions.
+the list of all the candidates in the increasing order of their depth first
+search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
+candidates closer in the list have their common dominator nearby in cases when
+there are no fully redundant instructions. Essentially if,
 
-Essentially,
-abs(DFSIn(I1) - DFSIn(I2)) < abs(DFSIn(I1) - DFSIn(I3))
-=> bbDist(I1, nearest-common-dom(I1, I2)) <= bbDist(I1, nearest-common-dom(I1, I3))
+\begin{verbatim}
+BasicBlock B1, B2, B3 // B1 != B2 != B3
+DFSIn(B1) = depth first discovery time stamp
+DFSIn(B1,B2) = abs(DFSIn(B1) - DFSIn(B2))
+DFSIn(B1,B3) = abs(DFSIn(B1) - DFSIn(B3))
+BBDist(B1,B2) = BBDist(B1, nearest-common-dom(B1, B2))
+BBDist(B1,B3) = BBDist(B1, nearest-common-dom(B1, B3))
 
-In case of fully redundant computations the nearby elements may not be 'closer'
-than the separated ones. TODO: Explain.
+Then,
+DFSIn(B1,B2) < DFSIn(B1,B3) implies
+ BBDist(B1,B2) <= BBDist(B1,B3)
+\end{verbatim}
 
-This means for candidates closer to each other w.r.t. DFS numbers:
+Same property would hold for instructions I1, I2, I3 provided they are not fully
+redundant i.e., none of the basic blocks containing I1, I2 and I3 dominate the
+other.
+
+\begin{verbatim}
+DFSIn(I1) = DFSIn of containing basic block
+DFSIn(I1,I2) = abs(DFSIn(I1) - DFSIn(I2))
+DFSIn(I1,I3) = abs(DFSIn(I1) - DFSIn(I3))
+BBDist(I1,I2) = BBDist(I1, nearest-common-dom(I1, I2))
+BBDist(I1,I3) = BBDist(I1, nearest-common-dom(I1, I3))
+
+Then,
+ DFSIn(I1,I2) < DFSIn(I1,I3) implies
+  BBDist(I1,I2) <= BBDist(I1,I3)
+\end{verbatim}
+
+It says that if I1 is closer to I2 than I3 in the list of candidates sorted by
+DFSIn numbers, then the nearest common dominator of I1 and I2 will be closer (in
+terms of basic block distance) to I1, than nearest common dominator of I1 and
+I3.
+
+In the presence of fully redundant computations in the list of hoistable
+candidates this equation may not hold. For example, if there is another
+instruction I4 which is dominated by I1, but I4 is present in a basic block
+which is farther than DFSIn(I1, I2) i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
+however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$
+
+So by sorting candidates w.r.t. their DFSIn numbers:
 \begin{enumerate}

### 2016-08-11 hoist.tex
formalisation of partition algorithm

@@
 DFSIn(B1) = depth first discovery time stamp
-DFSIn(B1,B2) = abs(DFSIn(B1) - DFSIn(B2))
-DFSIn(B1,B3) = abs(DFSIn(B1) - DFSIn(B3))
-BBDist(B1,B2) = BBDist(B1, nearest-common-dom(B1, B2))
-BBDist(B1,B3) = BBDist(B1, nearest-common-dom(B1, B3))
+DFSIn(B1,B2) = |DFSIn(B1) - DFSIn(B2)|
+DFSIn(B1,B3) = |DFSIn(B1) - DFSIn(B3)|
+
+Depth(B1) = depth of tree from the root node.
+// When B1 dominates B2
+BBDist(B1, B2) = |Depth(B1) - Depth(B2)|
+
+NCD(B1, B2) = nearest common dominator of B1 and B2
+BBDist(B1,B2) = BBDist(B1, NCD(B1, B2))
+BBDist(B1,B3) = BBDist(B1, NCD(B1, B3))
 
@@
 Same property would hold for instructions I1, I2, I3 provided they are not fully
-redundant i.e., none of the basic blocks containing I1, I2 and I3 dominate the
-other.
+redundant i.e., none of the basic blocks containing I1, I2 and I3 respectively
+dominate the other:
 
 \begin{verbatim}
-DFSIn(I1) = DFSIn of containing basic block
-DFSIn(I1,I2) = abs(DFSIn(I1) - DFSIn(I2))
-DFSIn(I1,I3) = abs(DFSIn(I1) - DFSIn(I3))
-BBDist(I1,I2) = BBDist(I1, nearest-common-dom(I1, I2))
-BBDist(I1,I3) = BBDist(I1, nearest-common-dom(I1, I3))
+Instruction I1, I2, I3
+// I1, I2, I3 are not fully redundant
+// B1, B2, B3 contains I1, I2, I3 respectively
+DFSIn(I1) = DFSIn(B1)
+BBDist(I1, I2) = BBDist(B1, B2)
+DFSIn(I1) = DFSIn(B1)
+DFSIn(I1,I3) = DFSIn(B1, B3)
+NCD(I1, I2) = NCD(B1, B2)
 

### 2016-08-11 hoist.tex
restructure

@@
 
+\subsection{Illustrative Example}
+In the case when more than one computation from the same basic block
+shows up as redundant.
+
+e.g.
+\begin{verbatim}
+BB1
+load a;
+....
+store .;
+....
+load a;
+
+BB2
+load a
+
+BB0 -> BB1
+BB0 -> BB2
+\end{verbatim}
+
+In this case both the loads from BB1 will show up as redundant
+(according to the current algorithm), but the complete hoising will
+fail as the second load in BB1 is not safe to hoist. By following the
+bottom-up approach, the sorted list of BBs will appear like this:
+
+{ BB1, BB2(first load), BB2(second load) }
+
+Now we will start evaluating the safety of BB1-l1 and BB2-l2 which
+will pass, then BB12-l1 and BB2-l2 will fail.  So we can revert back
+to BB12-l1 and hoist them to BB0. Doing the top down approach and
+finding the partial set to hoist instructions partially would result
+in a combinatorial explosion.
+
+GCC: 23286 has interesting test cases.
+
 \section{Code hoisting policies}
 The amount of hoisting depends on whether we collect GVN of instructions
-on-demand (pessimistic) or, before finding candidates (optimistic).It also
+before finding candidates (optimistic) or, on-demand (pessimistic). It also
 depends on the generality of the GVN algorithm, however, that analysis is beyond
 the scope of this paper.
-
-\subsection{Hoisting bottom up with pessimistic approach}
-\label{subsec:pessimistic}
-In the pessimistic approach, the basic blocks are traversed in the inverse
-depth-first order, computing the GVN of instructions as they come by. The GVN of
-sibling branches are compared for equality. Once such a candidate is found, it
-is hoisted in the common dominator. All the leaf nodes are visited before the
-non-leaf nodes (bottom-up) because instructions are hoisted upwards.
 
@@
 \begin{verbatim}
-BasicBlock B1, B2, B3 // B1 != B2 != B3
+// B1 != B2 != B3
+BasicBlock B1, B2, B3
 DFSIn(B1) = depth first discovery time stamp
@@
 which is farther than DFSIn(I1, I2) i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
-however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$
+however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$.
 
@@
 \item would make fewer checks for legality and profitability.
-\item hoisting point will be closer so, the intersection of live-range of the
+\item hoisting point will be closer, so the intersection of live-range of the
   instruction with other instructions will be minimal.
@@
 
-In the case where all the candidates cannot be hoisted at a common hoist point
-because of:
-\begin{enumerate}
-\item Safety conditions are not met.
-\item Live range would increase too much.
-\end{enumerate}
+In our current implementation we keep as many candidates in one set as possible
+(greedy approach). We split the list at a point where the legality checks fail
+to hoist subset of candidates which are legal to hoist and then start finding
+new hoisting point for the remaining ones. So there are two limitations of the
+current implementation: first, sorting by DFSIn numbers does not give the right
+order when there are fully redundant instructions, second, it lacks cost model
+for partitioning in order to preserve/improve performance.
 
-In our current implementation we keep as many candidates in one set as possible
-unless (greedy approach). We split the list at a point where the legality checks
-fail to hoist subset of candidates which are legal to hoist and, start finding
-new hoisting point for the remaining ones.
-
-TODO: Put an example to illustrate the point.
-
-In the case when more than one computation from the same basic block
-shows up as redundant.
-
-e.g.
-\begin{verbatim}
-BB1
-load a;
-....
-store .;
-....
-load a;
-
-BB2
-load a
-
-BB0 -> BB1
-BB0 -> BB2
-\end{verbatim}
-
-In this case both the loads from BB1 will show up as redundant
-(according to the current algorithm), but the complete hoising will
-fail as the second load in BB1 is not safe to hoist. By following the
-bottom-up approach, the sorted list of BBs will appear like this:
-
-{ BB1, BB2(first load), BB2(second load) }
-
-Now we will start evaluating the safety of BB1-l1 and BB2-l2 which
-will pass, then BB12-l1 and BB2-l2 will fail.  So we can revert back
-to BB12-l1 and hoist them to BB0. Doing the top down approach and
-finding the partial set to hoist instructions partially would result
-in a combinatorial explosion.
-
-GCC: 23286 has interesting test cases.
+\subsection{Hoisting bottom up with pessimistic approach}
+\label{subsec:pessimistic}
+In the pessimistic approach, the basic blocks are traversed in the inverse
+depth-first order, computing the GVN of instructions as they come by. The GVN of
+sibling branches are compared for equality. Once such a candidate is found, it
+is hoisted in the common dominator. All the leaf nodes are visited before the
+non-leaf nodes (bottom-up) because instructions are hoisted upwards.
 
@@
 
-\subsubsection{Time complexity of algorithm}
+\subsection{Time complexity of algorithm}
 The complexity of code hoisting is linear in number of instructions that could

### 2016-08-11 hoist.tex
cost models

@@
 new hoisting point for the remaining ones. So there are two limitations of the
-current implementation: first, sorting by DFSIn numbers does not give the right
-order when there are fully redundant instructions, second, it lacks cost model
-for partitioning in order to preserve/improve performance.
+current implementation of the partition algorithm: First, sorting by DFSIn
+numbers does not give the desired order when there are fully redundant
+instructions; Second, it lacks cost model for partitioning in order to
+preserve/improve performance \ref{sec:future-work}.
 
@@
 \label{hoist:reg-pressure}
-Following example explains how code hoisting can actually reduce the register pressure.
-Consider the following example where the labels prefixed with 'p' represent the position of
-instruction in a basic block (names prefixed with bb).
+Following example explains how code hoisting can actually reduce the register
+pressure.  Consider the following example where the labels prefixed with 'P'
+represent the position of instruction in a basic block (names prefixed with 'B').
 
 \begin{verbatim}
-bb0: p0: b = ...
-bb1: p1: c = ...
-bb2: p2: if (c) jump bb3 else jump bb4
-bb3: p3: a0 = b + c
-bb4: p4: a1 = b + c
-bb5: p5: d = phi {a0(bb3), a1(bb4)}
+B0: P0: b = 1
+    goto B1
 
-If we measure distance(px, py) as total instruction
-count in the path from the position of px to py
+B1: P1: c = 2
+    goto B2
 
-old-live-range = max(distance(p0, p3) + distance(p1, p3) ,
-                     distance(p0, p4) + distance(p1, p4))
-                 + distance (p3, p5)
-new-live-rance = distance(p0, p2) + distance(p1, p2)
-                 + distance (p2, p5)
+B2: P2: if c is true then goto B3 else goto B4
+
+B3: P3: a0 = b + c
+    goto B5
+
+B4: P4: a1 = b + c
+    goto B5
+
+B5: P5: d = phi {a0(B3), a1(B4)}
+
+If we measure D(Px,Py) as total instruction
+count in the path from the position of Px to Py
+
+live-range(a0) = D(P0,P3) + D(P1,P3) + D(P3,P5)
+               = 6 + 4 + 2 = 12
+live-range(a1) = D(P0,P4) + D(P1,P4) + D(P4,P5)
+               = 6 + 4 + 2 = 12
+
+old-live-range = max(live-range(a0),live-range(a1))
+               = 12
+
+After hoisting a0 and a1 are removed and a copy
+of a0 as a01 is placed in B2 just before P2.
+
+live-range(a01) = D(P0,P2-1) + D(P1,P2-1) + D(P2 -1,P5)
+                = 4 + 3 + 3 = 10
+new-live-range = live-range(a01)
 \end{verbatim}
 
-If the new live-range is less than the old one it will be a good candidate
-for hoist. There are cases when both the ranges will be same:
-\begin{enumerate}
-  \item There is only one operand on the right hand side e.g., an assignment operation.
-  \item One of the operands is not a Instruction/Register.
-\end{enumerate}
+If the new live-range is less than the old one it will be a good candidate for
+hoisting. The live ranges can remain same as well if there is only one operand
+on the right hand side e.g., an assignment operation or, one of the operands is
+not a register. That means hoisting upwards will decrease the live-range of its
+use but increase the live-range of its definition.
 
@@
 one register operand in the right hand side with its last use, code hoisting
-will either decrease or maintain the register pressure.
+will either decrease or preserve the register pressure.
 
@@
 
-However, in the two special cases discussed in Section-\ref{hoist:reg-pressure},
-it will be okay to hoist because the regiter pressure would not change.
+However, in the special case discussed in Section-\ref{hoist:reg-pressure},
+it will be okay to hoist because the register pressure would not decrease.
 
@@
 \section{Conclusion and Future Work}
+\label{sec:future-work}
 

### 2016-08-11 hoist.tex
memref
cost models

@@
 \section{Cost models}
+Similar to any compiler optimization pass, there are several cost functions that
+are deployed to tune for optimal combination of performance and code-size.
+Since this is mostly a code-size optimization pass, the goal is to not regress
+in performance across popular benchmarks at the same time reduce code size as
+much as possible. Following are the cost models which are implemented:
 
-\subsubsection{Reduce register pressure}
+\subsection{Reduce register pressure}
 \label{hoist:reg-pressure}
@@
 
-\subsubsection{In the presence of calls}
+\subsection{In the presence of calls}
 \label{cost:across-calls}
@@
 computation is hoisted across the call, that would increase the total number of
-registers required by one, thus contributing to the register pressure.
+registers required by one, thus contributing to the register pressure. However,
+in the special case discussed in Section-\ref{hoist:reg-pressure}, it will be
+okay to hoist because the register pressure would not decrease.
 
-However, in the special case discussed in Section-\ref{hoist:reg-pressure},
-it will be okay to hoist because the register pressure would not decrease.
+Hoisting loads/stores across calls also require precise analysis of all the
+memory addresses accessed by the call. Our implementation being an
+intraprocedural pass, the analysis is very conservative. In the presence of pure
+calls, loads can be hoisted but stores can't. Also, if the call throws
+exceptions, or it it may not return, memory references cannot be hoisted.
 
-Hoisting memory references also require precise analysis of all the memory
-addresses accessed by the call. In our implementation, since we only analyze the
-callee declaration the analysis is very conservative. In the presence of pure
-calls, loads can be hoisted but stores can't. Also, if the call throws exceptions,
-or it it may not return, memory references cannot be hoisted.
-
-\subsubsection{Hoisting too far away}
-When we load a value from memory, we require an extra register to keep the value
-around. Hoisting a load means we need to keep the value alive for a longer
-period. That could result in spills later in the register allocation.
-
-Also, if there are several instructions in between the hoisting point and the
+\subsection{Hoisting too far away}
+If there are several instructions in between the hoisting point and the
 instruction to be hoisted, the instruction to be hoisted crosses several
-instructions while hoisting, it means we are adding one register to all the
+instructions while hoisting, that means we are adding one register to all the
 live-ranges spanning the instructions. That could result in spills. In the
@@
 is below a threshold. Ideally, it should be okay to hoist all the instructions
-and a later live-range-splitting pass should make the right decision of
-rematerializing the instruction should it be beneficial to do so. But the
-current live-range splitting pass of llvm is not making the optimial decision
-and we have found spills if the threshold is exceeded. The threshold was
-computed as a result of tuning the llvm testsuite \cite{llvm-nightly} and spec
-benchmark \cite{Henning2000}.
+and a later a live-range-splitting \cite{cooper1998live} pass should make the
+right decision of rematerializing the instruction should it be beneficial to do
+so. But the current live-range splitting pass of llvm is not making the optimial
+decision and we have found spills if the threshold is exceeded. The threshold
+was computed as a result of tuning the llvm testsuite \cite{llvm-nightly} and
+spec benchmark \cite{Henning2000}.
 
-\begin{enumerate}
-\item When the hoist point is too far.
-\item When the number of instructions we may cross is very high.
-\end{enumerate}
+Also, hoisting a load increases the register pressure by one across all the
+instructions which the load would cross. That could result in spills later in
+the register allocation.
+
+However, in the special case discussed in Section-\ref{hoist:reg-pressure}, it
+will be okay to hoist because the register pressure would not decrease.
 

### 2016-08-11 hoist.tex
add related work, move ILP to cost model

@@
 \begin{itemize}
-\item a new algorithm to hoist computations from branches
-\item cost model to reduce live-range and reduce spills
-\item performance evaluation of the implementation in LLVM
-\item comparison against other algorithms for code hoisting
+\item a new algorithm to hoist computations from branches,
+\item cost model to reduce live-range and reduce spills,
+\item performance evaluation of the implementation in LLVM,
+\item comparison against other algorithms for code hoisting.
 \end{itemize}
+
+\newpage
 
@@
 
-graph algorithms \cite{tarjan},
-SSA \cite{cytron}
-GVN \cite{rosen1988global,click1995global}
-CFG, DOM \cite{dragonbook}
-SESE, SEME \cite{sese}
-MemorySSA \cite{novillo2007memory}
-
-\subsection{Global Value Numbering}
-
-Code hoisting can also reduce the critical path length of execution in out of
-order machines. As more instructios are available at the hoisting point, the
-hardware has more instructions to reorder. Following example illustrates how
-hoisting can improve performance by exposing more ILP.
-
-\begin{verbatim}
-float foo(float d, float min, float max, float a)
-{
-  float tmin, tmax, inv;
-
-  inv = 1.0f / d;
-  if (inv >= 0) {
-    tmin = (min - a) * inv;
-    tmax = (max - a) * inv;
-  } else {
-    tmin = (max - a) * inv;
-    tmax = (min - a) * inv;
-  }
-  return tmax + tmin;
-}
-\end{verbatim}
-
-In this program the computations of tmax and tmin are identical to the
-computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
-depends on inv which depends on a division operation which is generally more
-expensive than the addition, subtraction and multiplication operations. The
-total latency of computation across each branch is:
-
-O(div) + 2(O(sub) + O(mul))
-Or for out of order processors with two add units and two multiply units:
-O(div) + O(sub) + O(mul)
-
-Now if the computation of tmax and tmin are hoisted outside the
-conditionals, the C code version would look like this:
-\begin{verbatim}
-float foo(float d, float min, float max, float a)
-{
-  float tmin, tmax, tmin1, tmax1, inv;
-
-  tmin1 = (min - a);
-  tmax1 = (max - a);
-
-  inv = 1.0f / d;
-  tmin1 = tmin1 * inv;
-  tmax1 = tmax1 * inv;
-
-  if (inv >= 0) {
-    tmin = tmin1;
-    tmax = tmax1;
-  } else {
-    tmin = tmax1;
-    tmax = tmin1;
-  }
-
-  return tmax + tmin;
-}
-
-\end{verbatim}
-
-In this code the two subtractions and the division operations can be executed in
-parallel because there are no dependencies among them. So the total number of
-cycles will be max(O(div), O(sub)) + O(mul) = O(div) + O(mul); since O(div) is
-usually much greater than O(sub) \cite{x86,aarch64}
-
-Of course, a partial redundancy elimination pass could just remove the entire
-if-block because final operation is an addition (asociative under fast math).
-
-\begin{verbatim}
-float foo(float d, float min, float max, float a)
-{
-  float tmin1, tmax1, inv;
-
-  tmin1 = (min - a);
-  tmax1 = (max - a);
-
-  inv = 1.0f / d;
-  tmin1 = tmin1 * inv;
-  tmax1 = tmax1 * inv;
-
-  return tmax1 + tmin1;
-}
-\end{verbatim}
-
 \newpage
@@
 \section{Code hoisting}
+
+\subsection{}
+The algorithm for code hoisting uses several common representations of the
+program that we shortly describe below:
+\begin{itemize}
+\item Control Flow Graph (CFG) and the Dominance (DOM) and Post-Dominance (PDOM)
+  relations \cite{dragonbook};
+\item Single Entry Single Exit (SESE) \cite{sese} and Single Entry Multiple Exit
+  (SEME) regions;
+\item Static Single Assignment (SSA) \cite{cytron};
+\item Global Value Numbering (GVN) \cite{rosen1988global,click1995global}: to
+  identify similar computations compilers use GVN.  Each expression is given a
+  unique number and the expressions that the compiler can prove identical are
+  given the same number;
+\item Memory SSA \cite{novillo2007memory}: memory operations that the compiler
+  is able to prove in dependence are linked through use-def chains;
+\end{itemize}
+
 For the code hoisting we follow a very systematic approach to writing a compiler
@@
 
+\subsubsection{Exposing more ILP}
+Code hoisting can also reduce the critical path length of execution in out of
+order machines. As more instructios are available at the hoisting point, the
+hardware has more instructions to reorder. Following example illustrates how
+hoisting can improve performance by exposing more ILP.
+
+\begin{verbatim}
+float foo(float d, float min, float max, float a)
+{
+  float tmin, tmax, inv;
+
+  inv = 1.0f / d;
+  if (inv >= 0) {
+    tmin = (min - a) * inv;
+    tmax = (max - a) * inv;
+  } else {
+    tmin = (max - a) * inv;
+    tmax = (min - a) * inv;
+  }
+  return tmax + tmin;
+}
+\end{verbatim}
+
+In this program the computations of tmax and tmin are identical to the
+computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
+depends on inv which depends on a division operation which is generally more
+expensive than the addition, subtraction and multiplication operations. The
+total latency of computation across each branch is:
+
+O(div) + 2(O(sub) + O(mul))
+Or for out of order processors with two add units and two multiply units:
+O(div) + O(sub) + O(mul)
+
+Now if the computation of tmax and tmin are hoisted outside the
+conditionals, the C code version would look like this:
+\begin{verbatim}
+float foo(float d, float min, float max, float a)
+{
+  float tmin, tmax, tmin1, tmax1, inv;
+
+  tmin1 = (min - a);
+  tmax1 = (max - a);
+
+  inv = 1.0f / d;
+  tmin1 = tmin1 * inv;
+  tmax1 = tmax1 * inv;
+
+  if (inv >= 0) {
+    tmin = tmin1;
+    tmax = tmax1;
+  } else {
+    tmin = tmax1;
+    tmax = tmin1;
+  }
+
+  return tmax + tmin;
+}
+
+\end{verbatim}
+
+In this code the two subtractions and the division operations can be executed in
+parallel because there are no dependencies among them. So the total number of
+cycles will be max(O(div), O(sub)) + O(mul) = O(div) + O(mul); since O(div) is
+usually much greater than O(sub) \cite{x86,aarch64}
+
+Of course, a partial redundancy elimination pass could just remove the entire
+if-block because final operation is an addition (asociative under fast math).
+
+\begin{verbatim}
+float foo(float d, float min, float max, float a)
+{
+  float tmin1, tmax1, inv;
+
+  tmin1 = (min - a);
+  tmax1 = (max - a);
+
+  inv = 1.0f / d;
+  tmin1 = tmin1 * inv;
+  tmax1 = tmax1 * inv;
+
+  return tmax1 + tmin1;
+}
+\end{verbatim}
+
 \newpage

### 2016-08-12 hoist.tex
future work

@@
 \section{Cost models}
+\label{sec:cost-models}
 Similar to any compiler optimization pass, there are several cost functions that
@@
 
-\subsubsection{Exposing more ILP}
+\subsection{Exposing more ILP}
 Code hoisting can also reduce the critical path length of execution in out of
@@
 \label{sec:future-work}
+We have presented the GVN based code hoisting algorithm. The primary goal is to
+reduce the code size but it benefits performance in some cases as well. To
+preserve performance and not hoist too much we have implemented several cost
+models \ref{sec:cost-models}. Since those cost models depend on a set of thresholds,
+it requires tuning, as such, we used representative benchmarks to tune them.
 

### 2016-08-14 hoist.tex
math

@@
 total latency of computation across each branch is:
-
-O(div) + 2(O(sub) + O(mul))
-Or for out of order processors with two add units and two multiply units:
-O(div) + O(sub) + O(mul)
+$O(div) + 2(O(sub) + O(mul))$
+Or, for out of order processors with two add units and two multiply units:
+$O(div) + O(sub) + O(mul)$
 
@@
 parallel because there are no dependencies among them. So the total number of
-cycles will be max(O(div), O(sub)) + O(mul) = O(div) + O(mul); since O(div) is
-usually much greater than O(sub) \cite{x86,aarch64}
+cycles will be $max(O(div), O(sub)) + O(mul) = O(div) + O(mul)$; since $O(div)$ is
+usually much greater than $O(sub)$ \cite{x86,aarch64}
 

### 2016-08-14 hoist.tex
related work

@@
 \section{Related Work}
+
+Rosen et al. explain moving computations successors to remove redundancies
+\cite{rosen1988global}. Their algorithm iterates on computations of same rank
+and move the code with identical computations from the sibling branch. This is a
+pessimistic approach (TODO: is it?). Also there is no notion of partially
+hoisting the computations so their approach may result is missing many hoisting
+opportunities.
 

### 2016-08-14 hoist.tex
move example

@@
 \subsection{Illustrative Example}
-In the case when more than one computation from the same basic block
-shows up as redundant.
+Code hoisting can also reduce the critical path length of execution in out of
+order machines. As more instructios are available at the hoisting point, the
+hardware has more instructions to reorder. Following example illustrates how
+hoisting can improve performance by exposing more ILP.
 
-e.g.
 \begin{verbatim}
-BB1
-load a;
-....
-store .;
-....
-load a;
+float foo(float d, float min, float max, float a)
+{
+  float tmin, tmax, inv;
 
-BB2
-load a
-
-BB0 -> BB1
-BB0 -> BB2
+  inv = 1.0f / d;
+  if (inv >= 0) {
+    tmin = (min - a) * inv;
+    tmax = (max - a) * inv;
+  } else {
+    tmin = (max - a) * inv;
+    tmax = (min - a) * inv;
+  }
+  return tmax + tmin;
+}
 \end{verbatim}
 
-In this case both the loads from BB1 will show up as redundant
-(according to the current algorithm), but the complete hoising will
-fail as the second load in BB1 is not safe to hoist. By following the
-bottom-up approach, the sorted list of BBs will appear like this:
+In this program the computations of tmax and tmin are identical to the
+computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
+depends on inv which depends on a division operation which is generally more
+expensive than the addition, subtraction and multiplication operations. The
+total latency of computation across each branch is:
+$O(div) + 2(O(sub) + O(mul))$
+Or, for out of order processors with two add units and two multiply units:
+$O(div) + O(sub) + O(mul)$
 
-{ BB1, BB2(first load), BB2(second load) }
+Now if the computation of tmax and tmin are hoisted outside the
+conditionals, the C code version would look like this:
+\begin{verbatim}
+float foo(float d, float min, float max, float a)
+{
+  float tmin, tmax, tmin1, tmax1, inv;
 
-Now we will start evaluating the safety of BB1-l1 and BB2-l2 which
-will pass, then BB12-l1 and BB2-l2 will fail.  So we can revert back
-to BB12-l1 and hoist them to BB0. Doing the top down approach and
-finding the partial set to hoist instructions partially would result
-in a combinatorial explosion.
+  tmin1 = (min - a);
+  tmax1 = (max - a);
+
+  inv = 1.0f / d;
+  tmin1 = tmin1 * inv;
+  tmax1 = tmax1 * inv;
+
+  if (inv >= 0) {
+    tmin = tmin1;
+    tmax = tmax1;
+  } else {
+    tmin = tmax1;
+    tmax = tmin1;
+  }
+
+  return tmax + tmin;
+}
+
+\end{verbatim}
+
+In this code the two subtractions and the division operations can be executed in
+parallel because there are no dependencies among them. So the total number of
+cycles will be $max(O(div), O(sub)) + O(mul) = O(div) + O(mul)$; since $O(div)$ is
+usually much greater than $O(sub)$ \cite{x86,aarch64}
+
+Of course, a partial redundancy elimination pass could just remove the entire
+if-block because final operation is an addition (asociative under fast math).
+
+\begin{verbatim}
+float foo(float d, float min, float max, float a)
+{
+  float tmin1, tmax1, inv;
+
+  tmin1 = (min - a);
+  tmax1 = (max - a);
+
+  inv = 1.0f / d;
+  tmin1 = tmin1 * inv;
+  tmax1 = tmax1 * inv;
+
+  return tmax1 + tmin1;
+}
+\end{verbatim}
 
@@
 
-\subsection{Exposing more ILP}
-Code hoisting can also reduce the critical path length of execution in out of
-order machines. As more instructios are available at the hoisting point, the
-hardware has more instructions to reorder. Following example illustrates how
-hoisting can improve performance by exposing more ILP.
-
-\begin{verbatim}
-float foo(float d, float min, float max, float a)
-{
-  float tmin, tmax, inv;
-
-  inv = 1.0f / d;
-  if (inv >= 0) {
-    tmin = (min - a) * inv;
-    tmax = (max - a) * inv;
-  } else {
-    tmin = (max - a) * inv;
-    tmax = (min - a) * inv;
-  }
-  return tmax + tmin;
-}
-\end{verbatim}
-
-In this program the computations of tmax and tmin are identical to the
-computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
-depends on inv which depends on a division operation which is generally more
-expensive than the addition, subtraction and multiplication operations. The
-total latency of computation across each branch is:
-$O(div) + 2(O(sub) + O(mul))$
-Or, for out of order processors with two add units and two multiply units:
-$O(div) + O(sub) + O(mul)$
-
-Now if the computation of tmax and tmin are hoisted outside the
-conditionals, the C code version would look like this:
-\begin{verbatim}
-float foo(float d, float min, float max, float a)
-{
-  float tmin, tmax, tmin1, tmax1, inv;
-
-  tmin1 = (min - a);
-  tmax1 = (max - a);
-
-  inv = 1.0f / d;
-  tmin1 = tmin1 * inv;
-  tmax1 = tmax1 * inv;
-
-  if (inv >= 0) {
-    tmin = tmin1;
-    tmax = tmax1;
-  } else {
-    tmin = tmax1;
-    tmax = tmin1;
-  }
-
-  return tmax + tmin;
-}
-
-\end{verbatim}
-
-In this code the two subtractions and the division operations can be executed in
-parallel because there are no dependencies among them. So the total number of
-cycles will be $max(O(div), O(sub)) + O(mul) = O(div) + O(mul)$; since $O(div)$ is
-usually much greater than $O(sub)$ \cite{x86,aarch64}
-
-Of course, a partial redundancy elimination pass could just remove the entire
-if-block because final operation is an addition (asociative under fast math).
-
-\begin{verbatim}
-float foo(float d, float min, float max, float a)
-{
-  float tmin1, tmax1, inv;
-
-  tmin1 = (min - a);
-  tmax1 = (max - a);
-
-  inv = 1.0f / d;
-  tmin1 = tmin1 * inv;
-  tmax1 = tmax1 * inv;
-
-  return tmax1 + tmin1;
-}
-\end{verbatim}
-
 \newpage

### 2016-08-14 hoist.tex
ranking

@@
 needed to be rematerialized. Thereafter, all the computations made redundant by
-the new copy are removed. After that SSA semantics is restored by updating the
-intermediate representation (IR) to reflect changes.
+the new copy are removed. After that SSA form is restored by updating the
+intermediate representation (IR) to reflect changes. After one iteration of
+algorithm runs through the entire function, it creates more oppportunities for
+\italics{higher ranked} computations \cite{rosen1988global}. Currently, we rerun
+the algorithm until there are no more instructions left to be
+hoisted. Obviously, this is not the most optimal approach and can be improved by
+ranking the computations \cite{rosen1988global}.
 
@@
 
+Currently, we rerun the algorithm until there are no more instructions left to
+be hoisted. This is not the most optimal approach and results in expensive
+analyses to be recomputed. This can be improved by ranking the computations
+\cite{rosen1988global}.
+
 \bibliographystyle{abbrv}

### 2016-08-17 hoist.tex
grammar rewriting

@@
 \subsection{Finding the candidates}
-Finding if a set of instructions perform identical computations. Since this code
-hoisting is based on GVN, we need the facility to compute GVN of instructions.
-In llvm \cite{llvm}, GVN infrastructure is already set up and works well for
-most kind of instructions except loads and stores so we compute the GVN of loads
-and stores separately. In order to value number loads, we value number (hash)
-the address from where the value is to be loaded. For stores, we value number
-(hash) the address as well as the value to be stored at that address.
+The first step is to find if a set of instructions perform identical
+computations. Since our code hoisting algorithm is based on GVN, we need the
+facility to compute GVN of instructions.  In llvm \cite{llvm}, GVN
+infrastructure is already set up and works well for most kind of instructions
+except loads and stores so we compute the GVN of loads and stores separately. In
+order to value number loads, we value number (hash) the address from where the
+value is to be loaded. For stores, we value number (hash) the address as well as
+the value to be stored at that address.
 
@@
 \label{subsec:legality}
-Finding a hoisting point i.e., a common dominator of the set of instructions. If
-such a dominator is found then checking if all the use-operands of the set of
+Since the equality of candidates is purely based on the value they compute, we
+need to establish if hoisting them to a common dominator would be feasible. One
+a common dominator is found then we check if all the use-operands of the set of
 instructions are available or not. In some cases when the operands are not
 available, it is possible to reinstantiate (remateralize) the use-operands, thus
-passing the legality check. Checking that the side-effects of the computations
-does not intersect with any side-effects between the instructions to be hoisted
-and their hoisting point. For memory operations like loads, stores, calls etc.,
-it is also required to check that all the paths from the hoisting point to the
-end of the function should execute the exact instruction, in order to guarantee
-correctness. Moreover, hoisting of memory operations is tricky on paths which
-have indirect branch targets e.g., landing pad, case statements, goto labels
-etc., because it becomes difficult to prove that all the paths from hoisting
-point to the end of the function would execute the instruction. In our current
-implementation we discard hoisting through such paths.
+passing the legality check. Subsequently, it is checked that the side-effects of
+the computations does not intersect with any side-effects between the
+instructions to be hoisted and their hoisting point. For memory operations like
+loads, stores, calls etc., it is also required to check that all the paths from
+the hoisting point to the end of the function should execute the exact
+instruction, in order to guarantee correctness. Moreover, hoisting of memory
+operations is tricky on paths which have indirect branch targets e.g., landing
+pad, case statements, goto labels etc., because it becomes difficult to prove
+that all the paths from hoisting point to the end of the function would execute
+the instruction. In our current implementation we discard hoisting through such
+paths.
 
@@
 Scalars are the easiest to hoist because we do not have to analyze them for
-aliasing memory references. As long as all the operands are available, the
-scalar can be hoisted.
+aliasing memory references. As long as all the operands are available (or can be
+made available by rematerialization), the scalar can be hoisted.
 

### 2016-08-17 hoist.tex
profitability

@@
 \label{subsec:profitability}
-Establishing a set of cost models tune for performance. Like decreasing code
-size, minimizing spills etc.
+After the legality checks have passed, we check for profitability of hoisting.
+That takes into account the impact code-hoisting would have on various
+parameters that affects runtime performance e.g., impact on live-range, gain in
+the code size.  We have established a set of cost models \ref{sec:cost-models}
+for each parameter and tuned them for performance against representative
+benchmarks.
 
 \subsubsection{Hoisting scalars}
-Care must be taken in case of hoisting scalars too far, as that may increase
-register pressure and result in spills. For example hoisting a scalar past a
-call (Section-\ref{cost:across-calls}). In our current implementation we hoist
-scalars past a call only when optimizing for code-side (-Os). Another way to
-mitigate this problem is be to reinstantiate (rematerialize) the computation
+Since scalars are the majority of instructions which are hoisted, we pay special
+attention in case of hoisting scalars too far, as that may increase register
+pressure and result in spills. For example hoisting a scalar past a call
+(Section-\ref{cost:across-calls}).  In our current implementation we hoist
+scalars past a call only when optimizing for code-side (-Os). Ideally, a later
+stage of live-range splitting pass should split the live-ranges for optimal
+performance, however, that is not the case with llvm as we have found
+regressions when scalars are hoisted too far \ref{sec:cost-models}. Another way
+to mitigate this problem is be to reinstantiate (rematerialize) the computation
 after a call (may be as a different optimization pass).
@@
 algorithm runs through the entire function, it creates more oppportunities for
-\italics{higher ranked} computations \cite{rosen1988global}. Currently, we rerun
+\emph{higher ranked} computations \cite{rosen1988global}. Currently, we rerun
 the algorithm until there are no more instructions left to be

### 2016-08-18 hoist.tex
results on llvm-testsuite

@@
 \label{sec:experimental-results}
+We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146 and the results are:
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|}
+      \hline
+      Metric                   	     & Number\\\hline
+      Instructions hoisted & 21616 \\\hline
+      Instructions removed & 30438 \\\hline
+      Loads hoisted        & 14802 \\\hline
+      Loads removed        & 20719 \\\hline
+      Stores hoisted       & 15    \\\hline
+      Stores removed       & 15    \\\hline
+      Calls hoisted        & 8     \\\hline
+      Calls removed        & 8     \\\hline
+\end{tabular}
+  \end{center}
+  \caption{Code hoisting metrics on llvm-testsuite}
+  \label{tab:results}
+\end{table}
 

### 2016-08-19 hoist.tex
exp results

--- 2016-08-17 hoist.tex
@@
 run-time and as the number of run-time computations is not reduced, code
-hoisting is not a redundancy elimination pass.  The goals of code hoisting are:
+hoisting is not a redundancy elimination pass. This is also different from
+global code scheduling \cite{dragonbook} in the sense that code-hoisting will
+only hoist computations when there are two or more identical computation sharing
+a common dominator. The goals of code hoisting are:
+
 \begin{itemize}
@@
 \item to expose more instruction level parallelism: by hoisting identical
-  computations to be executed earlier, schedulers can move heavy computations
-  earlier in order to avoid pipeline bubbles;
+  computations to be executed earlier, instruction schedulers can move heavy
+  computations earlier in order to avoid pipeline bubbles;
 \item to help out-of-order processors with speculative execution of branches: by
@@
 \label{sec:experimental-results}
+We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146 and the
+results are listed in Table-\ref{tab:results}. The table lists the number of
+scalars, loads, stores and calls hoisted as well as removed. For each category,
+the number of instructions removed is greater or equal to the number of
+instructions hoisted because each hoisting is performed only when at least one
+identical computation is found.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|}
+      \hline
+      Metric                   	     & Number\\\hline
+      Scalars hoisted      & 6791  \\\hline
+      Scalars removed      & 9696  \\\hline
+      Loads hoisted        & 14802 \\\hline
+      Loads removed        & 20719 \\\hline
+      Stores hoisted       & 15    \\\hline
+      Stores removed       & 15    \\\hline
+      Calls hoisted        & 8     \\\hline
+      Calls removed        & 8     \\\hline
+      Total Instructions hoisted & 21616 \\\hline
+      Total Instructions removed & 30438 \\\hline
+\end{tabular}
+  \end{center}
+  \caption{Code hoisting metrics on llvm-testsuite}
+  \label{tab:results}
+\end{table}
 

### 2016-08-19 hoist.tex
analysis

@@
 
+Loads are hoisted the most followed by scalars, stores and calls in decreasing
+order.  This was the common trend in all our experiments. One reason why loads
+are hoisted more than scalars is the early execution of this pass in the llvm
+pass pipeline where there are many loads (?? Verify). This order may change should this pass
+be scheduled later.
+
 \begin{table}[h!]
@@
 
+TODO: Table for decrease in the code-size of final binaries.
+
 \subsection{Comparative analysis of different approaches to hoisting}

### 2016-08-19 hoist.tex
load most

--- 2016-08-19 hoist.tex
@@
 
+Loads are hoisted the most followed by scalars, stores and calls in decreasing
+order.  This was the common trend in all our experiments. One reason why loads
+are hoisted the most is the early execution of this pass (before mem2reg) in the
+llvm pass pipeline. Passes like mem2reg, instcombine might actually remove
+those loads so this order may change should this pass be scheduled later.
+
 \begin{table}[h!]
@@
 
-\subsection{Comparative analysis of different approaches to hoisting}
+TODO: Table for decrease in the code-size of final binaries.
 

### 2016-08-19 hoist.tex
static metrics

@@
 We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146 and the
-results are listed in Table-\ref{tab:results}. The table lists the number of
-scalars, loads, stores and calls hoisted as well as removed. For each category,
-the number of instructions removed is greater or equal to the number of
-instructions hoisted because each hoisting is performed only when at least one
-identical computation is found.
+results are listed in Table-\ref{tab:hoist-results}. The table lists the number
+of scalars, loads, stores and calls hoisted as well as removed. For each
+category, the number of instructions removed is greater or equal to the number
+of instructions hoisted because each hoisting is performed only when at least
+one identical computation is found.
 
@@
   \caption{Code hoisting metrics on llvm-testsuite}
-  \label{tab:results}
+  \label{tab:hoist-results}
 \end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|}
+      \hline
+      Metric               & Before & After              \\\hline
+      Call sites deleted, not inlined             & 1988    & 1988   \\\hline
+      Functions deleted (all callers found)       & 38250   & 38255  \\\hline
+      Functions inlined                           & 154986  & 154985 \\\hline
+      Allocas merged together                     & 212     & 212    \\\hline
+      Caller-callers analyzed                     & 193042  & 193092 \\\hline
+      Call sites analyzed                         & 414336  & 414381 \\\hline
+      Rematerialized defs for spilling            & 18321   & 18326  \\\hline
+      Rematerialized defs for splitting           & 5719    & 5842   \\\hline
+      Spill slots allocated                       & 42912   & 42970  \\\hline
+      Spilled live ranges                         & 61330   & 61362  \\\hline
+      Spills inserted                             & 50724   & 50784  \\\hline
+\end{tabular}
+  \end{center}
+  \caption{Static metrics before and after code-hoisting on llvm-testsuite}
+  \label{tab:static-results}
+\end{table}
+
+Other static metrics are listed in Table-\ref{tab:static-results}. Here we can
+see that except for rematerializing defs for splitting, which has an overhead of ~2\%, all
+other parameters have less than 1\% overhead. This is to explain why the performance does not
+go down with our implementation (and cost-model) of code hoisting pass.
 

### 2016-08-22 hoist.tex
related work
future work

@@
 A lot of bug reports in GCC and LLVM bugzillas, no papers on this topic yet.
+
+Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
+code hoisting in a data flow framework.
 
@@
 
-\newpage
-
 \section{Experimental Evaluation}
 \label{sec:experimental-results}
-We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146 and the
+We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146) and the
 results are listed in Table-\ref{tab:hoist-results}. The table lists the number
@@
 
+With the implementation of code-hoisting in llvm, the passes which rely on the
+code-size/instruction-count to make optimization decisions needs to be
+revisited. The first candidate would be the inliner. We have seen different
+inlining decisions (See Table-\ref{tab:code-size}) before and after
+code-hoisting was enabled.  Since inliner has several magic numbers tuned for
+the previous pass layout, it would need some improvement.
+
 \bibliographystyle{abbrv}

### 2016-08-22 hoist.tex
code-size

@@
 
-TODO: Table for decrease in the code-size of final binaries.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|}
+      \hline
+      Code-size metric  (.text)                   & Number   \\\hline
+      Total  benchmarks                           & 497      \\\hline
+      Total  gained in size                       & 39       \\\hline
+      Total  decrease in size                     & 58       \\\hline
+      Median decrease in size                     & 2.9\%    \\\hline
+      Median increase in size                     & 2.4\%    \\\hline
+    \end{tabular}
+  \end{center}
+  \caption{Code size metrics on llvm-testsuite}
+  \label{tab:code-size}
+\end{table}
+
+While benchmarking llvm-testsuite we see both increase as well as decrease in
+the codesizes of the final binaries. Since the pass runs early, it has affect on
+many optimizations which rely on number of instructions, lenth of the use-def
+chain etc metrics which are affected by code-hoisting. Specially the inliner
+where the total inline cost (in llvm) is heavily dependent on the total count of
+instructions in the caller and callee. Various code-size metrics are shown in
+Table-\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
+5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
+by 35.38\%.
 

### 2016-08-22 hoist.tex
outlier

@@
 5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
-by 35.38\%.
+by 35.38\%. In this benchmark 3 more (15 as compared to 12) functions got
+inlined and because of that 10 more (81 vs. 71) vector instructions got
+generated, 3 calls got hoisted/sunk as (compared to 0), one loop got unswitched
+(compared to 0), 6 high latency machine instructions got hoisted out of loop, 59
+(compared ot 30) machine instructions got hoisted out of loop, 70 (compared to
+39) machine instructions were sunk.
 

### 2016-08-23 hoist.tex
adding author
fix author alignment
some fixes to introduction

@@
        \email{s.pop@samsung.com}
+\and
+\alignauthor
+Daniel Berlin\\
+       \affaddr{Google Inc.}\\
+       \email{dberlin@dberlin.org}
 }
@@
 \begin{abstract}
-  Code hoisting identifies identical computations across the program with the
+  Code-hoisting identifies identical computations across the program with the
   help of a Global Value Numbering (GVN) analysis, identifies a hoisting
@@
 
-  Code hoisting is often confused with CSE (Common Subexpression Elimination)
-  although hoisting is not bound to find redundancies within a single basic
-  block.  Code hoisting could also be confused with Partial Redundancy
+  Code-hoisting is often confused with CSE (Common Subexpression Elimination)
+  although code-hoisting is not bound to find redundancies within a single basic
+  block.  Code-hoisting could also be confused with Partial Redundancy
   Elimination (PRE) or with Global Common Subexpression Elimination (GCSE)
-  although the main goal of code hoisting is not to remove redundancies: code
-  hoisting effectively exposes redundancies and enables other passes like LICM
-  to remove more redundancies.  The main goal of code hoisting is to reduce code
-  size with the added benefit of exposing more instruction level parallelism and
-  reduced register pressure.
+  although the main goal of code-hoisting is not to remove redundancies:
+  code-hoisting effectively exposes redundancies and enables other passes like
+  LICM to remove more redundancies.  The main goal of code-hoisting is to reduce
+  code size with the added benefit of exposing more instruction level
+  parallelism and reduced register pressure.
 \end{abstract}
@@
 
-This paper describes code hoisting, a technique that uses the information
+This paper describes code-hoisting, a technique that uses the information
 computed for PRE to detect identical computations and has a transformation phase
 whose goal differs from PRE: it removes identical computations from different
-branches of execution.  These identical computations would not be redundant at
-run-time and as the number of run-time computations is not reduced, code
-hoisting is not a redundancy elimination pass. This is also different from
-global code scheduling \cite{dragonbook} in the sense that code-hoisting will
-only hoist computations when there are two or more identical computation sharing
-a common dominator. The goals of code hoisting are:
+branches of execution.  These identical computations in different branches of
+execution are not redundant computations at run-time and the number of run-time
+computations is not reduced. Code-hoisting is not a redundancy elimination pass,
+and thus it has different cost function and heuristics than PRE or CSE.
+Code-hoisting is also different from global code scheduling
+\cite{dragonbook,click1995global} in the sense that code-hoisting will only
+hoist computations when there are identical computations sharing a common
+dominator. The goals of code hoisting are:
 
@@
 \item a new algorithm to hoist computations from branches,
-\item cost model to reduce live-range and reduce spills,
+\item cost models to reduce live-range and reduce spills,
 \item performance evaluation of the implementation in LLVM,
@@
 \end{itemize}
-
-\newpage
 

### 2016-08-23 hoist.tex
add hoisting bug refs
cleanup finding candidates

@@
 
-A lot of bug reports in GCC and LLVM bugzillas, no papers on this topic yet.
+A lot of bug reports in GCC and LLVM bugzillas
+\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, no papers on this topic yet.
 
@@
 
-\subsection{}
 The algorithm for code hoisting uses several common representations of the
@@
 \item Memory SSA \cite{novillo2007memory}: memory operations that the compiler
-  is able to prove in dependence are linked through use-def chains;
+  is able to prove in dependence are linked through use-def chains.
 \end{itemize}
 
-For the code hoisting we follow a very systematic approach to writing a compiler
-optimization pass. As such this pass can be divided into the following five
-steps:
+The code-hoisting pass can be divided into the following steps that we will
+describe in the rest of this section:
+\begin{itemize}
+\item find candidates to hoist and compute hoisting position,
+\item check whether moving the code is legal and profitable,
+\item transform the code.
+\end{itemize}
 
 \subsection{Finding the candidates}
-The first step is to find if a set of instructions perform identical
-computations. Since our code hoisting algorithm is based on GVN, we need the
-facility to compute GVN of instructions.  In llvm \cite{llvm}, GVN
-infrastructure is already set up and works well for most kind of instructions
-except loads and stores so we compute the GVN of loads and stores separately. In
-order to value number loads, we value number (hash) the address from where the
-value is to be loaded. For stores, we value number (hash) the address as well as
-the value to be stored at that address.
+The first step is to find a set of instructions that perform identical
+computations: this is performed by a linear scan of all instructions of the
+program and classifying all instructions by their value given by GVN.  The
+current implementation of GVN in LLVM has some limitations when it comes to
+loads and stores so we compute the GVN of loads and stores separately.  Our
+short-time solution to value number loads is to hash the address from where the
+value is to be loaded. For stores, we value number the address as well as the
+value to be stored at that address.
+
+Another limitation of the current GVN implementation in LLVM is that the
+instructions dependent on the loads will not get numbered correctly, and so
+after hoisting all candidates we need to rerun the GVN analysis in order to
+discover new candidates now available after having hoisted load instructions.
+This limitation should be addressed in a new implementation of the GVN based on
+MemorySSA, that would better account for equivalent loads and their dependent
+instructions.
 
@@
 perform. In a pessimistic approach (Section-\ref{subsec:pessimistic}) where we
-want to hoist limited set of instructions from the sibing branches as we iterate
+want to hoist limited set of instructions from the sibling branches as we iterate
 the DFS tree bottom-up, it is sufficient to compute values on-demand. Whereas,
@@
 
-Whatever approach we take, to find a candidate all we need to do is to compare
-the value number of computations.
+Once the instructions have been classified in equivalence classes, we compute
+for each group of equivalent instructions a point in the program that is both
+legal and profitable for the code-hoisting.
 

### 2016-08-23 hoist.tex
reword some more

@@
 \begin{itemize}
-\item find candidates to hoist and compute hoisting position,
-\item check whether moving the code is legal and profitable,
+\item find candidates: instructions computing same values,
+\item compute a point in the program where it is both legal and profitable to
+  move the code,
 \item transform the code.
@@
 computations: this is performed by a linear scan of all instructions of the
-program and classifying all instructions by their value given by GVN.  The
-current implementation of GVN in LLVM has some limitations when it comes to
+program and classifying all instructions by their value given by GVN.
+
+The current implementation of GVN in LLVM has some limitations when it comes to
 loads and stores so we compute the GVN of loads and stores separately.  Our
@@
 process to choose is determined by the scope of code-hoisting we want to
-perform. In a pessimistic approach (Section-\ref{subsec:pessimistic}) where we
-want to hoist limited set of instructions from the sibling branches as we iterate
-the DFS tree bottom-up, it is sufficient to compute values on-demand. Whereas,
-in the optimistic approach (Section-\ref{subsec:optimistic}), where we want to
-hoist as many instructions as possible, would require values to be precomputed.
+perform. In a pessimistic approach, as described in
+Section-\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
+from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
+to compute GVN values on-demand. Whereas, in the optimistic approach, as
+described in Section-\ref{subsec:optimistic}, we want to hoist as many
+instructions as possible, and it would require GVN values to be precomputed.
 
@@
 for each group of equivalent instructions a point in the program that is both
-legal and profitable for the code-hoisting.
+legal and profitable for the instructions to be moved to.
 

### 2016-08-23 hoist.tex
reword some more

@@
 Since the equality of candidates is purely based on the value they compute, we
-need to establish if hoisting them to a common dominator would be feasible. One
-a common dominator is found then we check if all the use-operands of the set of
-instructions are available or not. In some cases when the operands are not
+need to establish if hoisting them to a common dominator would be feasible. Once
+a common dominator is found, we check whether all the use-operands of the set of
+instructions are available at that position. In some cases when the operands are not
 available, it is possible to reinstantiate (remateralize) the use-operands, thus
-passing the legality check. Subsequently, it is checked that the side-effects of
-the computations does not intersect with any side-effects between the
-instructions to be hoisted and their hoisting point. For memory operations like
-loads, stores, calls etc., it is also required to check that all the paths from
-the hoisting point to the end of the function should execute the exact
-instruction, in order to guarantee correctness. Moreover, hoisting of memory
-operations is tricky on paths which have indirect branch targets e.g., landing
-pad, case statements, goto labels etc., because it becomes difficult to prove
-that all the paths from hoisting point to the end of the function would execute
-the instruction. In our current implementation we discard hoisting through such
-paths.
+passing the legality check.
 
-In the optimistic approach (Section-\ref{subsec:optimistic}), it is possible
-that a common hoisting point of all the instructions is either too far away, or
-not legally possible. In these cases, it is still possible to `partially' hoist
-a subset of instructions by splitting the set of candidates and finding a closer
-hoisting point for each subset. For more details see
+Subsequently, it is checked that the side-effects of the computations does not
+intersect with any side-effects between the instructions to be hoisted and their
+hoisting point. For memory operations like loads, stores, calls etc., it is also
+required to check that all the paths from the hoisting point to the end of the
+function should execute the exact instruction, in order to guarantee
+correctness.
+
+Moreover, hoisting of memory operations is tricky on paths which have indirect
+branch targets e.g., landing pad, case statements, goto labels etc., because it
+becomes difficult to prove that all the paths from hoisting point to the end of
+the function would execute the instruction. In our current implementation we
+discard hoisting through such paths.
+
+In the optimistic approach, described in Section-\ref{subsec:optimistic}, it is
+possible that a common hoisting point of all the instructions is either too far
+away, or not legally possible. In these cases, it is still possible to
+`partially' hoist a subset of instructions by splitting the set of candidates
+and finding a closer hoisting point for each subset. For more details see
 Section-\ref{subsec:partition}.
@@
 aliasing memory references. As long as all the operands are available (or can be
-made available by rematerialization), the scalar can be hoisted.
+made available by rematerialization), the scalar computation can be hoisted.
 
@@
 point. If that is not available we try to rematerialize the addresss if
-possible.  Along the path, from current position of the load instruction through
-the hoisting point, we find if there are aliasing writes to memory. In those
-cases we discard the candidate.
+possible.  Along the path, from current position of the load instruction
+backwards on the control flow to the hoisting point, we check whether there are
+writes to memory that may alias with the load, in which case we discard the
+candidate.
 
 \subsubsection{Hoisting stores}
-We need to check the dependency requirements similar to the hoisting of
-loads. We need to check that the operands of the store instruction are available
-at the hoisting point, that there are no aliasing loads or store along the path
-from the current position through the hoisting point.
+For stores, we check the dependency requirements similar to the hoisting of
+loads. We check that the operands of the store instruction are available at the
+hoisting point, that there are no aliasing loads or store along the path from
+the current position to the hoisting point.
 
 \subsubsection{Hoisting calls}
-For hoisting the call instructions, we check that the loads/stores in the call
-would not hazard with any loads/stores (in the caller) along the path from the
-callsite through the hoisting point, just like we did for loads and stores.
+Call instructions can be divided into three categories: those calls equivalent
+to purely scalar computations, calls reading from memory, and most of the time,
+without further information, calls have to be classified as writing to memory,
+that is the most restrictive form.  Each category of call instructions will be
+handled as described for scalar, load, and store instructions.
 

### 2016-08-23 hoist.tex
more cleanups

@@
 
-\subsection{Finding the candidates}
+\subsection{Finding candidates to hoist}
 The first step is to find a set of instructions that perform identical
@@
 perform. In a pessimistic approach, as described in
-Section-\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
+Section~\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
 from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
 to compute GVN values on-demand. Whereas, in the optimistic approach, as
-described in Section-\ref{subsec:optimistic}, we want to hoist as many
+described in Section~\ref{subsec:optimistic}, we want to hoist as many
 instructions as possible, and it would require GVN values to be precomputed.
@@
 
-\subsection{Checking for legality}
+\subsection{Legality check}
 \label{subsec:legality}
@@
 
-In the optimistic approach, described in Section-\ref{subsec:optimistic}, it is
+In the optimistic approach, described in Section~\ref{subsec:optimistic}, it is
 possible that a common hoisting point of all the instructions is either too far
@@
 and finding a closer hoisting point for each subset. For more details see
-Section-\ref{subsec:partition}.
+Section~\ref{subsec:partition}.
 
-\subsubsection{Hoisting scalars}
+\subsubsection{Legality of hoisting scalars}
 Scalars are the easiest to hoist because we do not have to analyze them for
@@
 
-\subsubsection{Hoisting loads}
+\subsubsection{Legality of hoisting loads}
 The availability of operand to the load (an address) is checked at the hoisting
@@
 
-\subsubsection{Hoisting stores}
+\subsubsection{Legality of hoisting stores}
 For stores, we check the dependency requirements similar to the hoisting of
@@
 
-\subsubsection{Hoisting calls}
+\subsubsection{Legality of hoisting calls}
 Call instructions can be divided into three categories: those calls equivalent
@@
 
-\subsection{Checking for profitability}
+\subsection{Profitability check}
 \label{subsec:profitability}
@@
 That takes into account the impact code-hoisting would have on various
-parameters that affects runtime performance e.g., impact on live-range, gain in
-the code size.  We have established a set of cost models \ref{sec:cost-models}
+parameters that affect runtime performance e.g., impact on live-range, gain in
+the code size.  We have established a set of cost models described in Section~\ref{sec:cost-models}
 for each parameter and tuned them for performance against representative
@@
 
-\subsubsection{Hoisting scalars}
+\subsubsection{Profitability of hoisting scalars}
 Since scalars are the majority of instructions which are hoisted, we pay special
 attention in case of hoisting scalars too far, as that may increase register
-pressure and result in spills. For example hoisting a scalar past a call
-(Section-\ref{cost:across-calls}).  In our current implementation we hoist
-scalars past a call only when optimizing for code-side (-Os). Ideally, a later
-stage of live-range splitting pass should split the live-ranges for optimal
-performance, however, that is not the case with llvm as we have found
-regressions when scalars are hoisted too far \ref{sec:cost-models}. Another way
-to mitigate this problem is be to reinstantiate (rematerialize) the computation
-after a call (may be as a different optimization pass).
+pressure and result in spills. For example hoisting a scalar past a call, as
+described in Section~\ref{cost:across-calls}.  In our current implementation we
+hoist scalars past a call only when optimizing for code-side (-Os). Ideally, a
+later stage of live-range splitting pass should split the live-ranges for
+optimal performance, however, that is not the case with llvm as we have found
+regressions when scalars are hoisted too far, as in
+Section~\ref{sec:cost-models}. Another way to mitigate this problem is be to
+reinstantiate (rematerialize) the computation after a call (may be as a
+different optimization pass).
 
-\subsubsection{Hoisting loads}
+\subsubsection{Profitability of hoisting loads}
 A load instruction introduces a register where the value loaded will be kept,
@@
 hoist load except the hoisting point is too far (this distance is computed by
-looking at the experimental results of representative benchmarks see
-Section-\ref{sec:experimental-results}).
+looking at the experimental results of representative benchmarks, see
+Section~\ref{sec:experimental-results}).
 
-\subsubsection{Hoisting stores}
+\subsubsection{Profitability of hoisting stores}
 Since stores do not increase the live-range of any registers, and in some cases
@@
 
-\subsubsection{Hoisting calls}
+\subsubsection{Profitability of hoisting calls}
 Currently we hoist all the calls that are suitable candidates for hoisting.
 
-\subsection{Making the transformation}
+\subsection{Code generation}
 Once all the legality checks and profitability checks are satisfied for a set of
@@
 list of instructions having identical GVNs. After that we find the common
-dominator dominating all such identical computations and perform legality checks
-(Section-\ref{subsec:legality}). Often times it is not possible to hoist all the
-instructions to one common dominator, due to legality constraints e.g.,
-intersecting side-effects or, profitability constraints e.g., hoisting point too
-far. In those cases, this algorithmn would partition the list of identical
-instructions into subsets which can be partially hoisted to their respective
-common dominators. The partition algorithm is described as follows:
+dominator dominating all such identical computations and perform legality
+checks, as described in Section~\ref{subsec:legality}. Often times it is not
+possible to hoist all the instructions to one common dominator, due to legality
+constraints e.g., intersecting side-effects or, profitability constraints e.g.,
+hoisting point too far. In those cases, this algorithmn would partition the list
+of identical instructions into subsets which can be partially hoisted to their
+respective common dominators. The partition algorithm is described as follows:
 
@@
 instructions; Second, it lacks cost model for partitioning in order to
-preserve/improve performance \ref{sec:future-work}.
+preserve/improve performance, see Section~\ref{sec:future-work}.
 
@@
 registers required by one, thus contributing to the register pressure. However,
-in the special case discussed in Section-\ref{hoist:reg-pressure}, it will be
+in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
 okay to hoist because the register pressure would not decrease.
@@
 
-However, in the special case discussed in Section-\ref{hoist:reg-pressure}, it
+However, in the special case discussed in Section~\ref{hoist:reg-pressure}, it
 will be okay to hoist because the register pressure would not decrease.
@@
 We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146) and the
-results are listed in Table-\ref{tab:hoist-results}. The table lists the number
+results are listed in Table~\ref{tab:hoist-results}. The table lists the number
 of scalars, loads, stores and calls hoisted as well as removed. For each
@@
 
-Other static metrics are listed in Table-\ref{tab:static-results}. Here we can
+Other static metrics are listed in Table~\ref{tab:static-results}. Here we can
 see that except for rematerializing defs for splitting, which has an overhead of ~2\%, all
@@
 instructions in the caller and callee. Various code-size metrics are shown in
-Table-\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
+Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
 5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
@@
 preserve performance and not hoist too much we have implemented several cost
-models \ref{sec:cost-models}. Since those cost models depend on a set of thresholds,
-it requires tuning, as such, we used representative benchmarks to tune them.
+models described in Section~\ref{sec:cost-models}. Since those cost models
+depend on a set of thresholds, it requires tuning, as such, we used
+representative benchmarks to tune them.
 
@@
 revisited. The first candidate would be the inliner. We have seen different
-inlining decisions (See Table-\ref{tab:code-size}) before and after
+inlining decisions in Table~\ref{tab:code-size}, before and after
 code-hoisting was enabled.  Since inliner has several magic numbers tuned for

### 2016-08-23 hoist.tex
more cleanups

@@
 \subsection{Code generation}
-Once all the legality checks and profitability checks are satisfied for a set of
+Once all the legality and profitability checks are satisfied for a set of
 identical instructions, they are suitable candidates for hoisting. A copy of the
@@
 needed to be rematerialized. Thereafter, all the computations made redundant by
-the new copy are removed. After that SSA form is restored by updating the
-intermediate representation (IR) to reflect changes. After one iteration of
-algorithm runs through the entire function, it creates more oppportunities for
-\emph{higher ranked} computations \cite{rosen1988global}. Currently, we rerun
-the algorithm until there are no more instructions left to be
-hoisted. Obviously, this is not the most optimal approach and can be improved by
-ranking the computations \cite{rosen1988global}.
+the new copy are removed, and the SSA form is restored by updating the
+intermediate representation (IR) to reflect the changes.
 
-\subsection{Verifying the post-conditions}
-Performing a set of checks (e.g., consistency of use-defs and SSA semantics) to
-establish that program invariants are maintained.
+After one iteration of algorithm runs through the entire function, it creates
+more oppportunities for \emph{higher ranked} computations
+\cite{rosen1988global}. Currently, this is a limitation of the GVN analysis
+pass, and so we rerun the code-hoisting algorithm until there are no more
+instructions left to be hoisted.  Obviously, this is not the most optimal
+approach and can be improved by ranking the computations \cite{rosen1988global},
+or by improving the GVN analysis to correctly number loads and dependent
+instructions.
+
+Finally after the transform is done, we verify a set of post-conditions to
+establish that program invariants are maintained: e.g., consistency of
+use-defs, and SSA semantics.
+
+\newpage
 

### 2016-08-24 hoist.tex
fix
related work

@@
        \email{s.pop@samsung.com}
-\and
-\alignauthor
-Daniel Berlin\\
-       \affaddr{Google Inc.}\\
-       \email{dberlin@dberlin.org}
+%\and
+%\alignauthor
+%Daniel Berlin\\
+%       \affaddr{Google Inc.}\\
+%       \email{dberlin@dberlin.org}
 }
@@
 pessimistic approach (TODO: is it?). Also there is no notion of partially
-hoisting the computations so their approach may result is missing many hoisting
+hoisting the computations so their approach may result in missing many hoisting
 opportunities.
@@
 code hoisting in a data flow framework.
+
+GCC implementation: Hoists top down in a predecessor where the value is in
+ANTIC\_IN. This approach may have the problem of hoisting the expression too far
+and to mititage that it only hoists expressions when at least one successor
+computes it. Hence the algorithm will hoist a limited set of computations.  One
+advantage of GCC implementation is that it works in sync with the GVN-PRE such
+that when new hoisting opportunities are created by GVN-PRE, code-hoisting will
+hoist them.
 
@@
 analyses to be recomputed. This can be improved by ranking the computations
-\cite{rosen1988global}.
+\cite{rosen1988global}. Also gvn-hoist runs very early in the pass pipeline, it
+will be good to evaluate the codesize/performance impact when it is run in sync
+with gvn-pre just like gcc does.
 

### 2016-08-24 hoist.tex
remove Dan Berlin from authors
abstract

@@
 \begin{abstract}
-  Code-hoisting identifies identical computations across the program with the
-  help of a Global Value Numbering (GVN) analysis, identifies a hoisting
-  position that dominates all the identical computations and in which all
-  operands are avaialable, duplicates one of the expressions in the hoisting
-  position, and removes all the computations now redundant in their original
-  position.
+  Code-hoisting identifies identical computations across the program and hoists
+  them to a common dominator so as to save code size.  Although the main goal of
+  code-hoisting is not to remove redundancies: it effectively exposes
+  redundancies and enables other passes like LICM to remove more redundancies.
+  The main goal of code-hoisting is to reduce code size with the added benefit
+  of exposing more instruction level parallelism and reduced register pressure.
 
-  Code-hoisting is often confused with CSE (Common Subexpression Elimination)
-  although code-hoisting is not bound to find redundancies within a single basic
-  block.  Code-hoisting could also be confused with Partial Redundancy
-  Elimination (PRE) or with Global Common Subexpression Elimination (GCSE)
-  although the main goal of code-hoisting is not to remove redundancies:
-  code-hoisting effectively exposes redundancies and enables other passes like
-  LICM to remove more redundancies.  The main goal of code-hoisting is to reduce
-  code size with the added benefit of exposing more instruction level
-  parallelism and reduced register pressure.
+  We present a code hoisting pass that we implemented in llvm. It is based on
+  Global Value Numbering infrastructure available in llvm. The experimental
+  results show an average of 2.5\% savings in code size, although the code size
+  increases in many cases because it enables more inlining. This is an
+  optimistic algorithm in the sense that we consider all identical computations
+  in a function as potential candidates to be hoisted. We make an extra effort
+  to hoist candidates by partitioning the potential candidates in a way to
+  enable partial hoisting in case common hoisting points for all the candidates
+  cannot be found. We also formalize cases when register pressure will reduce as
+  a result of hoisting.
 \end{abstract}

### 2016-08-24 hoist.tex
vbe listed

@@
 Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
-code hoisting in a data flow framework.
+code hoisting in a data flow framework. A list of Very Busy Expressions are
+computed which are hoisted in a basic block where the expression is anticipable
+(all the operands are available). This algorithm would hoist as far as possible
+without regarding the impact on register pressure and as such a cost model will
+be required.
 
-GCC implementation: Hoists top down in a predecessor where the value is in
-ANTIC\_IN. This approach may have the problem of hoisting the expression too far
-and to mititage that it only hoists expressions when at least one successor
-computes it. Hence the algorithm will hoist a limited set of computations.  One
-advantage of GCC implementation is that it works in sync with the GVN-PRE such
-that when new hoisting opportunities are created by GVN-PRE, code-hoisting will
-hoist them.
+GCC recently got code-hoisting which is implemented as part of GVN-PRE. The
+algorithm hoists top down in a predecessor where the value is in ANTIC\_IN. This
+approach may have the problem of hoisting the expression too far and to mititage
+that it only hoists expressions when at least one successor computes it. Hence
+the algorithm will hoist a limited set of computations.  One advantage of GCC
+implementation is that it works in sync with the GVN-PRE such that when new
+hoisting opportunities are created by GVN-PRE, code-hoisting will hoist them.
 

### 2016-08-25 hoist.tex
add description of current implementation of code hoisting in llvm

@@
 
-  We present a code hoisting pass that we implemented in llvm. It is based on
-  Global Value Numbering infrastructure available in llvm. The experimental
+  We present a code hoisting pass that we implemented in LLVM. It is based on
+  Global Value Numbering infrastructure available in LLVM. The experimental
   results show an average of 2.5\% savings in code size, although the code size
@@
 
+There are a lot of bug reports in GCC and LLVM bugzillas
+\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
+a more powerful code hoist transform.  The current LLVM implementation of code
+hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
+hoisting from identical basic blocks: the instructions of two sibling basic
+blocks are read in the same time, and all the instructions of the blocks are
+hoisted to the common parent block as long as the compiler is able to prove that
+the instructions are equivalent.  This implementation does not allow for an easy
+extension: first in terms of compilation time overhead the implementation would
+become quadratic in number of instructions to bisimulate, and second the
+equivalence of instructions currently uses $$I1->isIdenticalToWhenDefined(I2)$$
+would need to be rewritten to be more general, leading to using a mechanism
+similar to GVN.
+
 Rosen et al. explain moving computations successors to remove redundancies
 \cite{rosen1988global}. Their algorithm iterates on computations of same rank
-and move the code with identical computations from the sibling branch. This is a
-pessimistic approach (TODO: is it?). Also there is no notion of partially
-hoisting the computations so their approach may result in missing many hoisting
-opportunities.
-
-A lot of bug reports in GCC and LLVM bugzillas
-\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, no papers on this topic yet.
+and move the code with identical computations from the sibling branch. Also
+there is no notion of partially hoisting the computations so their approach may
+result in missing many hoisting opportunities.
 

### 2016-08-25 hoist.tex
add more detailed description of GCC hoisting from Danny, and added an Ack section

@@
 would need to be rewritten to be more general, leading to using a mechanism
-similar to GVN.
+similar to the idea described in this paper based on GVN.
 
@@
 Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
-code hoisting in a data flow framework. A list of Very Busy Expressions are
-computed which are hoisted in a basic block where the expression is anticipable
-(all the operands are available). This algorithm would hoist as far as possible
-without regarding the impact on register pressure and as such a cost model will
-be required.
+code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
+are computed which are hoisted in a basic block where the expression is
+anticipable (all the operands are available). This algorithm would hoist as far
+as possible without regarding the impact on register pressure and as such a cost
+model will be required.  Also the description of VBE is based on the classic
+dataflow model and an adaptation to a sparse SSA representation is required.
 
-GCC recently got code-hoisting which is implemented as part of GVN-PRE. The
-algorithm hoists top down in a predecessor where the value is in ANTIC\_IN. This
-approach may have the problem of hoisting the expression too far and to mititage
-that it only hoists expressions when at least one successor computes it. Hence
-the algorithm will hoist a limited set of computations.  One advantage of GCC
-implementation is that it works in sync with the GVN-PRE such that when new
-hoisting opportunities are created by GVN-PRE, code-hoisting will hoist them.
+GCC recently got code-hoisting which is implemented as part of GVN-PRE: it uses
+the set of ANTIC and AVAIL value expressions computed for PRE.  The algorithm
+hoists top down in a predecessor where the value is in ANTIC\_IN.  It uses
+ANTIC[B] to know what expressions will be computed on every path from B to exit,
+and can be computed in B: that is the safety condition.  It uses AVAIL[B] to
+subtract out those values already being computed (because they are already
+available): this is a redundancy condition.  The cost function is: for each
+hoist candidate, if all successors of B are dominated by B, then we know
+insertion into B will eliminate all the remaining computations.  It then checks
+to see if at least one successor of B has the value available.  This avoids
+hoisting it way up the chain to ANTIC.  It also checks to ensure that B has
+multiple successors, since hoisting in a straight line is pointless.  The
+algorithm continues on down the dominator tree, iterating with PRE until no more
+changes.  One advantage of GCC implementation is that it works in sync with the
+GVN-PRE such that when new hoisting opportunities are created by GVN-PRE,
+code-hoisting will hoist them.
 
@@
 
+\section{Acknowledgments}
+We would like to thank Daniel Berlin for his code reviews and for his feedback
+on earlier versions of this paper.
+
 \bibliographystyle{abbrv}

### 2016-08-25 hoist.tex
added more missing refs, and added perf improvement for illustration example

@@
 
-GCC recently got code-hoisting which is implemented as part of GVN-PRE: it uses
-the set of ANTIC and AVAIL value expressions computed for PRE.  The algorithm
-hoists top down in a predecessor where the value is in ANTIC\_IN.  It uses
-ANTIC[B] to know what expressions will be computed on every path from B to exit,
-and can be computed in B: that is the safety condition.  It uses AVAIL[B] to
-subtract out those values already being computed (because they are already
-available): this is a redundancy condition.  The cost function is: for each
-hoist candidate, if all successors of B are dominated by B, then we know
-insertion into B will eliminate all the remaining computations.  It then checks
-to see if at least one successor of B has the value available.  This avoids
-hoisting it way up the chain to ANTIC.  It also checks to ensure that B has
-multiple successors, since hoisting in a straight line is pointless.  The
+GCC recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
+part of GVN-PRE: it uses the set of ANTIC and AVAIL value expressions computed
+for PRE.  The algorithm hoists top down in a predecessor where the value is in
+ANTIC.  It uses ANTIC[B] to know what expressions will be computed on every
+path from B to exit, and can be computed in B: that is the safety condition.  It
+uses AVAIL[B] to subtract out those values already being computed (because they
+are already available): this is a redundancy condition.  The cost function is:
+for each hoist candidate, if all successors of B are dominated by B, then we
+know insertion into B will eliminate all the remaining computations.  It then
+checks to see if at least one successor of B has the value available.  This
+avoids hoisting it way up the chain to ANTIC.  It also checks to ensure that B
+has multiple successors, since hoisting in a straight line is pointless.  The
 algorithm continues on down the dominator tree, iterating with PRE until no more
@@
 
-\subsection{Illustrative Example}
-Code hoisting can also reduce the critical path length of execution in out of
+\subsection{Illustrative Example} \label{subsec:example}
+Code hoisting can reduce the critical path length of execution in out of
 order machines. As more instructios are available at the hoisting point, the
@@
 \begin{verbatim}
-float foo(float d, float min, float max, float a)
+float fun(float d, float min, float max, float a)
 {
@@
 total latency of computation across each branch is:
-$O(div) + 2(O(sub) + O(mul))$
+$C(div) + 2(C(sub) + C(mul))$
 Or, for out of order processors with two add units and two multiply units:
-$O(div) + O(sub) + O(mul)$
+$C(div) + C(sub) + C(mul)$
 
@@
 parallel because there are no dependencies among them. So the total number of
-cycles will be $max(O(div), O(sub)) + O(mul) = O(div) + O(mul)$; since $O(div)$ is
-usually much greater than $O(sub)$ \cite{x86,aarch64}
+cycles will be $max(C(div), C(sub)) + C(mul) = C(div) + C(mul)$; since $C(div)$ is
+usually much greater than $C(sub)$ \cite{x86,aarch64}.
 
-Of course, a partial redundancy elimination pass could just remove the entire
-if-block because final operation is an addition (asociative under fast math).
+A partial redundancy elimination pass could now remove the entire if-block
+because final operation is an addition (asociative under fast math), and the end
+result is:
 
 \begin{verbatim}
-float foo(float d, float min, float max, float a)
+float fun(float d, float min, float max, float a)
 {
@@
 
-GCC: 23286 has interesting test cases.
+%GCC: 23286 has interesting test cases.
 
@@
 \label{subsec:optimistic}
-In this approach, the goal is to maximise the total number of hoistings in the
+In this approach, the goal is to maximize the total number of hoistings in the
 entire function.  This algorithm is very useful when optimizing for code-size.
@@
 In order to hoist a subset of identical instructions, we partition the list of
-all candidates in a way to maximise the total number of hoistings.  By sorting
+all candidates in a way to maximize the total number of hoistings.  By sorting
 the list of all the candidates in the increasing order of their depth first
@@
 
+The code shown in Section~\ref{subsec:example} is a reduced test that appears in
+a hot loop of a proprietary benchmark.  When the expressions are hoisted from
+the conditional clauses, the overall performance of that benchmark improves by
+$15\%$ on an out-of-order processor due to increased instruction level
+parallelism, and better scheduling of the instructions, accommodating for the
+long latency of the division operation.
+
 \section{Conclusion and Future Work}

### 2016-08-25 hoist.tex
more cleanups

@@
 \begin{document}
-\def \SCoP {SCoP}
 \def \GCC {GCC}
@@
 \def \SSA {SSA}
-\def \scev {scev}
+\def \MemorySSA {MemorySSA}
+\def \PRE {PRE}
+\def \GVN {GVN}
+\def \LLVMTestSuite {\LLVM{} test-suite}
 
@@
 
-\title{GVN-Hoist: Hoisting Computations from Branches}
+\title{\GVN{}-Hoist: Hoisting Computations from Branches}
 
@@
    \hrule \vspace{5pt}
-   LLVM-dev 2016
+   \LLVM{}-dev 2016
 }
@@
 
-  We present a code hoisting pass that we implemented in LLVM. It is based on
-  Global Value Numbering infrastructure available in LLVM. The experimental
+  We present a code hoisting pass that we implemented in \LLVM{}. It is based on
+  Global Value Numbering infrastructure available in \LLVM{}. The experimental
   results show an average of 2.5\% savings in code size, although the code size
@@
 When it comes to detect redundancies across basic blocks these techniques fall
-short: more complex passes like GCSE and PRE have been designed to handle these
+short: more complex passes like GCSE and \PRE{} have been designed to handle these
 cases based on dataflow analysis \cite{morel1979global}.  At first these
 techniques were described in the classical data-flow analysis framework, and
-later the use of the SSA representation lowered the cost in terms of compilation
+later the use of the \SSA{} representation lowered the cost in terms of compilation
 time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
-techniques in the main stream: nowadays SSA based PRE is available in every
+techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
 industrial compiler.
@@
 This paper describes code-hoisting, a technique that uses the information
-computed for PRE to detect identical computations and has a transformation phase
-whose goal differs from PRE: it removes identical computations from different
+computed for \PRE{} to detect identical computations and has a transformation phase
+whose goal differs from \PRE{}: it removes identical computations from different
 branches of execution.  These identical computations in different branches of
@@
 computations is not reduced. Code-hoisting is not a redundancy elimination pass,
-and thus it has different cost function and heuristics than PRE or CSE.
+and thus it has different cost function and heuristics than \PRE{} or CSE.
 Code-hoisting is also different from global code scheduling
@@
 \item cost models to reduce live-range and reduce spills,
-\item performance evaluation of the implementation in LLVM,
+\item performance evaluation of the implementation in \LLVM{},
 \item comparison against other algorithms for code hoisting.
@@
 
-There are a lot of bug reports in GCC and LLVM bugzillas
+There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
 \cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
-a more powerful code hoist transform.  The current LLVM implementation of code
+a more powerful code hoist transform.  The current \LLVM{} implementation of code
 hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
@@
 would need to be rewritten to be more general, leading to using a mechanism
-similar to the idea described in this paper based on GVN.
+similar to the idea described in this paper based on \GVN{}.
 
@@
 model will be required.  Also the description of VBE is based on the classic
-dataflow model and an adaptation to a sparse SSA representation is required.
+dataflow model and an adaptation to a sparse \SSA{} representation is required.
 
-GCC recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
-part of GVN-PRE: it uses the set of ANTIC and AVAIL value expressions computed
-for PRE.  The algorithm hoists top down in a predecessor where the value is in
+\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
+part of \GVN{}-\PRE{}: it uses the set of ANTIC and AVAIL value expressions computed
+for \PRE{}.  The algorithm hoists top down in a predecessor where the value is in
 ANTIC.  It uses ANTIC[B] to know what expressions will be computed on every
@@
 has multiple successors, since hoisting in a straight line is pointless.  The
-algorithm continues on down the dominator tree, iterating with PRE until no more
-changes.  One advantage of GCC implementation is that it works in sync with the
-GVN-PRE such that when new hoisting opportunities are created by GVN-PRE,
+algorithm continues on down the dominator tree, iterating with \PRE{} until no more
+changes.  One advantage of \GCC{} implementation is that it works in sync with the
+\GVN{}-\PRE{} such that when new hoisting opportunities are created by \GVN{}-\PRE{},
 code-hoisting will hoist them.
@@
 \begin{itemize}
-\item Control Flow Graph (CFG) and the Dominance (DOM) and Post-Dominance (PDOM)
+\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
   relations \cite{dragonbook};
-\item Single Entry Single Exit (SESE) \cite{sese} and Single Entry Multiple Exit
+\item Single Entry Single Exit (\SESE{}) \cite{sese} and Single Entry Multiple Exit
   (SEME) regions;
-\item Static Single Assignment (SSA) \cite{cytron};
-\item Global Value Numbering (GVN) \cite{rosen1988global,click1995global}: to
-  identify similar computations compilers use GVN.  Each expression is given a
+\item Static Single Assignment (\SSA{}) \cite{cytron};
+\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
+  identify similar computations compilers use \GVN{}.  Each expression is given a
   unique number and the expressions that the compiler can prove identical are
   given the same number;
-\item Memory SSA \cite{novillo2007memory}: memory operations that the compiler
+\item Memory \SSA{} \cite{novillo2007memory}: memory operations that the compiler
   is able to prove in dependence are linked through use-def chains.
@@
 computations: this is performed by a linear scan of all instructions of the
-program and classifying all instructions by their value given by GVN.
+program and classifying all instructions by their value given by \GVN{}.
 
-The current implementation of GVN in LLVM has some limitations when it comes to
-loads and stores so we compute the GVN of loads and stores separately.  Our
+The current implementation of \GVN{} in \LLVM{} has some limitations when it comes to
+loads and stores so we compute the \GVN{} of loads and stores separately.  Our
 short-time solution to value number loads is to hash the address from where the
@@
 
-Another limitation of the current GVN implementation in LLVM is that the
+Another limitation of the current \GVN{} implementation in \LLVM{} is that the
 instructions dependent on the loads will not get numbered correctly, and so
-after hoisting all candidates we need to rerun the GVN analysis in order to
+after hoisting all candidates we need to rerun the \GVN{} analysis in order to
 discover new candidates now available after having hoisted load instructions.
-This limitation should be addressed in a new implementation of the GVN based on
-MemorySSA, that would better account for equivalent loads and their dependent
+This limitation should be addressed in a new implementation of the \GVN{} based on
+\MemorySSA{}, that would better account for equivalent loads and their dependent
 instructions.
 
-The process of computing GVN can be on-demand (as we come across an instruction)
-or, precomputed (computing GVN of all the instructions beforehand). Which
+The process of computing \GVN{} can be on-demand (as we come across an instruction)
+or, precomputed (computing \GVN{} of all the instructions beforehand). Which
 process to choose is determined by the scope of code-hoisting we want to
@@
 from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
-to compute GVN values on-demand. Whereas, in the optimistic approach, as
+to compute \GVN{} values on-demand. Whereas, in the optimistic approach, as
 described in Section~\ref{subsec:optimistic}, we want to hoist as many
-instructions as possible, and it would require GVN values to be precomputed.
+instructions as possible, and it would require \GVN{} values to be precomputed.
 
@@
 later stage of live-range splitting pass should split the live-ranges for
-optimal performance, however, that is not the case with llvm as we have found
+optimal performance, however, that is not the case with \LLVM{} as we have found
 regressions when scalars are hoisted too far, as in
@@
 needed to be rematerialized. Thereafter, all the computations made redundant by
-the new copy are removed, and the SSA form is restored by updating the
+the new copy are removed, and the \SSA{} form is restored by updating the
 intermediate representation (IR) to reflect the changes.
@@
 more oppportunities for \emph{higher ranked} computations
-\cite{rosen1988global}. Currently, this is a limitation of the GVN analysis
+\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
 pass, and so we rerun the code-hoisting algorithm until there are no more
@@
 approach and can be improved by ranking the computations \cite{rosen1988global},
-or by improving the GVN analysis to correctly number loads and dependent
+or by improving the \GVN{} analysis to correctly number loads and dependent
 instructions.
@@
 establish that program invariants are maintained: e.g., consistency of
-use-defs, and SSA semantics.
+use-defs, and \SSA{} semantics.
 
@@
 total latency of computation across each branch is:
-$C(div) + 2(C(sub) + C(mul))$
+$C_{div} + 2(C_{sub} + C_{mul})$
 Or, for out of order processors with two add units and two multiply units:
-$C(div) + C(sub) + C(mul)$
+$C_{div} + C_{sub} + C_{mul}$
 
@@
 parallel because there are no dependencies among them. So the total number of
-cycles will be $max(C(div), C(sub)) + C(mul) = C(div) + C(mul)$; since $C(div)$ is
-usually much greater than $C(sub)$ \cite{x86,aarch64}.
-
-A partial redundancy elimination pass could now remove the entire if-block
-because final operation is an addition (asociative under fast math), and the end
-result is:
-
-\begin{verbatim}
-float fun(float d, float min, float max, float a)
-{
-  float tmin1, tmax1, inv;
-
-  tmin1 = (min - a);
-  tmax1 = (max - a);
-
-  inv = 1.0f / d;
-  tmin1 = tmin1 * inv;
-  tmax1 = tmax1 * inv;
-
-  return tmax1 + tmin1;
-}
-\end{verbatim}
+cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since $C_{div}$ is
+usually much greater than $C_{sub}$ \cite{x86,aarch64}.
 
@@
 
+\newpage
+
 \section{Code hoisting policies}
-The amount of hoisting depends on whether we collect GVN of instructions
+The amount of hoisting depends on whether we collect \GVN{} of instructions
 before finding candidates (optimistic) or, on-demand (pessimistic). It also
-depends on the generality of the GVN algorithm, however, that analysis is beyond
+depends on the generality of the \GVN{} algorithm, however, that analysis is beyond
 the scope of this paper.
@@
 entire function.  This algorithm is very useful when optimizing for code-size.
-We collect the GVN of all the instructions in the function and iterate on the
-list of instructions having identical GVNs. After that we find the common
+We collect the \GVN{} of all the instructions in the function and iterate on the
+list of instructions having identical \GVN{}s. After that we find the common
 dominator dominating all such identical computations and perform legality
@@
 possible to hoist all the instructions to one common dominator, due to legality
-constraints e.g., intersecting side-effects or, profitability constraints e.g.,
+constraints, e.g., intersecting side-effects, or profitability constraints, e.g.,
 hoisting point too far. In those cases, this algorithmn would partition the list
@@
 Same property would hold for instructions I1, I2, I3 provided they are not fully
-redundant i.e., none of the basic blocks containing I1, I2 and I3 respectively
+redundant, i.e., none of the basic blocks containing I1, I2 and I3 respectively
 dominate the other:
@@
 instruction I4 which is dominated by I1, but I4 is present in a basic block
-which is farther than DFSIn(I1, I2) i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
+which is farther than $DFSIn(I1, I2)$, i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
 however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$.
@@
 new hoisting point for the remaining ones. So there are two limitations of the
-current implementation of the partition algorithm: First, sorting by DFSIn
+current implementation of the partition algorithm: first, sorting by DFSIn
 numbers does not give the desired order when there are fully redundant
-instructions; Second, it lacks cost model for partitioning in order to
+instructions; second, it lacks cost model for partitioning in order to
 preserve/improve performance, see Section~\ref{sec:future-work}.
@@
 In the pessimistic approach, the basic blocks are traversed in the inverse
-depth-first order, computing the GVN of instructions as they come by. The GVN of
+depth-first order, computing the \GVN{} of instructions as they come by. The \GVN{} of
 sibling branches are compared for equality. Once such a candidate is found, it
@@
 
-The pessimistic algorithm is fast, results in fewer spills but hoists very less
+The pessimistic algorithm is fast, results in fewer spills and also hoists fewer
 instructions. We have implemented optimistic approach because it is more general
@@
 The complexity of code hoisting is linear in number of instructions that could
-be hoisted in the program, matching the complexity of PRE on SSA form.  The
-analysis phase is based on the Global Value Numbering (GVN), the same analysis
-used for PRE, followed by the computation of a partition of identical
+be hoisted in the program, matching the complexity of \PRE{} on \SSA{} form.  The
+analysis phase is based on the Global Value Numbering (\GVN{}), the same analysis
+used for \PRE{}, followed by the computation of a partition of identical
 expressions to be hoisted in a same location to guarantee safety properties and
@@
 hoisting. The live ranges can remain same as well if there is only one operand
-on the right hand side e.g., an assignment operation or, one of the operands is
+on the right hand side, e.g., an assignment operation or, one of the operands is
 not a register. That means hoisting upwards will decrease the live-range of its
@@
 the caller saved registers are saved because they might be modified by the
-callee and after the call they are restored \cite{frame-lowering}. So before the
+callee and after the call they are restored. So before the
 call, the register pressure is high because the number of available registers
@@
 calls, loads can be hoisted but stores can't. Also, if the call throws
-exceptions, or it it may not return, memory references cannot be hoisted.
+exceptions, or if it may not return, memory references cannot be hoisted.
 
@@
 right decision of rematerializing the instruction should it be beneficial to do
-so. But the current live-range splitting pass of llvm is not making the optimial
-decision and we have found spills if the threshold is exceeded. The threshold
-was computed as a result of tuning the llvm testsuite \cite{llvm-nightly} and
-spec benchmark \cite{Henning2000}.
+so. But the current live-range splitting pass of \LLVM{} is not making the
+optimial decision and we have found spills if the threshold is exceeded. The
+threshold was computed as a result of tuning the \LLVMTestSuite{}, SPEC Cpu2000,
+and Cpu2006 benchmarks \cite{Henning2000}.
 
@@
 \label{sec:experimental-results}
-We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146) and the
+We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146) and the
 results are listed in Table~\ref{tab:hoist-results}. The table lists the number
@@
 are hoisted the most is the early execution of this pass (before mem2reg) in the
-llvm pass pipeline. Passes like mem2reg, instcombine might actually remove
+\LLVM{} pass pipeline. Passes like mem2reg, instcombine might actually remove
 those loads so this order may change should this pass be scheduled later.
@@
   \end{center}
-  \caption{Code hoisting metrics on llvm-testsuite}
+  \caption{Code hoisting metrics on \LLVMTestSuite{}}
   \label{tab:hoist-results}
@@
   \end{center}
-  \caption{Static metrics before and after code-hoisting on llvm-testsuite}
+  \caption{Static metrics before and after code-hoisting on \LLVMTestSuite{}}
   \label{tab:static-results}
@@
   \end{center}
-  \caption{Code size metrics on llvm-testsuite}
+  \caption{Code size metrics on \LLVMTestSuite{}}
   \label{tab:code-size}
@@
 
-While benchmarking llvm-testsuite we see both increase as well as decrease in
-the codesizes of the final binaries. Since the pass runs early, it has affect on
-many optimizations which rely on number of instructions, lenth of the use-def
-chain etc metrics which are affected by code-hoisting. Specially the inliner
-where the total inline cost (in llvm) is heavily dependent on the total count of
-instructions in the caller and callee. Various code-size metrics are shown in
+While benchmarking \LLVMTestSuite{} we see both increase as well as decrease in
+the codesizes of the final binaries. Since the pass runs early, it affects many
+optimizations which rely on the number of instructions, lenth of the use-def
+chain, and other metrics. For instance, the inliner is impacted by a decrease in
+the number of instructions in the caller and callee, as its heuristics estimate
+the size of functions to be inlined. Various code-size metrics are shown in
 Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
 5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
-by 35.38\%. In this benchmark 3 more (15 as compared to 12) functions got
-inlined and because of that 10 more (81 vs. 71) vector instructions got
-generated, 3 calls got hoisted/sunk as (compared to 0), one loop got unswitched
-(compared to 0), 6 high latency machine instructions got hoisted out of loop, 59
-(compared ot 30) machine instructions got hoisted out of loop, 70 (compared to
-39) machine instructions were sunk.
+by 35.38\%. In this benchmark 3 more functions got inlined (15 as compared to
+12) and because of that 10 more vector instructions got generated (81 vs. 71), 3
+calls got hoisted/sunk as (compared to 0), one loop got unswitched (compared to
+0), 6 high latency machine instructions got hoisted out of loop, 59 (compared to
+30) machine instructions got hoisted out of loop, 70 (compared to 39) machine
+instructions were sunk.
 
-The code shown in Section~\ref{subsec:example} is a reduced test that appears in
-a hot loop of a proprietary benchmark.  When the expressions are hoisted from
+The code shown in Section~\ref{subsec:example} is a reduced example that appears
+in a hot loop of a proprietary benchmark.  When the expressions are hoisted from
 the conditional clauses, the overall performance of that benchmark improves by
@@
 \label{sec:future-work}
-We have presented the GVN based code hoisting algorithm. The primary goal is to
+We have presented the \GVN{} based code hoisting algorithm. The primary goal is to
 reduce the code size but it benefits performance in some cases as well. To
@@
 analyses to be recomputed. This can be improved by ranking the computations
-\cite{rosen1988global}. Also gvn-hoist runs very early in the pass pipeline, it
+\cite{rosen1988global}. Also \GVN{}-hoist runs very early in the pass pipeline, it
 will be good to evaluate the codesize/performance impact when it is run in sync
-with gvn-pre just like gcc does.
+with \GVN{}-\PRE{} just like \GCC{} does.
 
-With the implementation of code-hoisting in llvm, the passes which rely on the
+With the implementation of code-hoisting in \LLVM{}, the passes which rely on the
 code-size/instruction-count to make optimization decisions needs to be

### 2016-08-25 hoist.tex
add discussion about GEP hoisting at -Oz

@@
 
+In order to reduce register pressure, we have restricted hoisting of address
+computations away from their corresponding loads and stores when the loads and
+stores cannot be moved.  This restriction is lifted when aggressively optimizing
+for size (in -Oz).  We have seen several uplifts and degradations in performance
+of the generated code on several benchmarks, and to avoid the degradations we
+decided to only enable hoisting of getElementPtrs without their associated
+memory operations when optimizing for size.  This restriction may be lifted in
+the future, either when the register allocation rematerialization pass has been
+improved to catch the regressions, or by implementing more precise heuristics
+for code hoisting.
+
 \section{Experimental Evaluation}

### 2016-08-26 hoist.tex
abstract

@@
   Code-hoisting identifies identical computations across the program and hoists
-  them to a common dominator so as to save code size.  Although the main goal of
-  code-hoisting is not to remove redundancies: it effectively exposes
-  redundancies and enables other passes like LICM to remove more redundancies.
-  The main goal of code-hoisting is to reduce code size with the added benefit
-  of exposing more instruction level parallelism and reduced register pressure.
+  them to a common dominator so as to save code size.  Not only code-hoisting
+  saves code size, a) It exposes redundancies in some cases, b) It exposes more
+  instruction level parallelism in the basic block a computation is hoisted, and
+  c) It enables other passes like LICM to remove more redundancies.
 
@@
   results show an average of 2.5\% savings in code size, although the code size
-  increases in many cases because it enables more inlining. This is an
-  optimistic algorithm in the sense that we consider all identical computations
+  also increases in many cases because it enables more inlining. This is an
+  optimistic algorithm in the sense that it considers all identical computations
   in a function as potential candidates to be hoisted. We make an extra effort
@@
   cannot be found. We also formalize cases when register pressure will reduce as
-  a result of hoisting.
+  a result of hoisting and why sorting the list of potential candidates
+  w.r.t. their DFS numbers helps hoist more candidates with less compile time
+  overhead.
 \end{abstract}
@@
 However, in the special case discussed in Section~\ref{hoist:reg-pressure}, it
-will be okay to hoist because the register pressure would not decrease.
+will be okay to hoist because the register pressure would not increase.
 

### 2016-08-26 hoist.tex
contrib

@@
 phase that reduces the number of run-time computations.  Classical scalar
-optimizations like CSE \cite{dragonbook} work very well on single basic blocks.
-When it comes to detect redundancies across basic blocks these techniques fall
+optimizations like CSE \cite{dragonbook} work very well on single basic blocks but
+when it comes to detect redundancies across basic blocks these techniques fall
 short: more complex passes like GCSE and \PRE{} have been designed to handle these
@@
 This paper describes code-hoisting, a technique that uses the information
-computed for \PRE{} to detect identical computations and has a transformation phase
-whose goal differs from \PRE{}: it removes identical computations from different
-branches of execution.  These identical computations in different branches of
-execution are not redundant computations at run-time and the number of run-time
-computations is not reduced. Code-hoisting is not a redundancy elimination pass,
-and thus it has different cost function and heuristics than \PRE{} or CSE.
-Code-hoisting is also different from global code scheduling
-\cite{dragonbook,click1995global} in the sense that code-hoisting will only
-hoist computations when there are identical computations sharing a common
+computed for \PRE{} to detect identical computations and has a transformation
+phase whose goal differs from \PRE{}: it removes identical computations from
+different branches of execution.  These identical computations in different
+branches of execution are not redundant computations at run-time and the number
+of run-time computations is not reduced. Code-hoisting is not a redundancy
+elimination pass, and thus it has different cost function and heuristics than
+\PRE{} or CSE.  Code-hoisting is also different from global code scheduling
+\cite{dragonbook,click1995global} in the sense that it will only hoist
+computations when there are identical computations sharing a common
 dominator. The goals of code hoisting are:
@@
 \item to improve function inlining heuristics: functions become cheaper to
-  inline by reducing their code size;
+  inline when thier code size decreases because inliner heuristics in \LLVM{}
+  depends on instruction count.
 \item to expose more instruction level parallelism: by hoisting identical
@@
 \begin{itemize}
-\item a new algorithm to hoist computations from branches,
+\item a new optimistic algorithm to hoist computations from branches,
 \item cost models to reduce live-range and reduce spills,
+\item a technique to maximize hoisting in an optimistic approach by partitioning
+  the list of potential candidates sorted by their DFS visit number,
 \item performance evaluation of the implementation in \LLVM{},

### 2016-08-26 hoist.tex
machine compiler flags

@@
 \label{sec:experimental-results}
-We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146) and the
-results are listed in Table~\ref{tab:hoist-results}. The table lists the number
-of scalars, loads, stores and calls hoisted as well as removed. For each
-category, the number of instructions removed is greater or equal to the number
-of instructions hoisted because each hoisting is performed only when at least
-one identical computation is found.
+We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146). All
+the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
+optimization level.  The results for code-hoisting are listed in
+Table~\ref{tab:hoist-results}. The table lists the number of scalars, loads,
+stores and calls hoisted as well as removed. For each category, the number of
+instructions removed is greater or equal to the number of instructions hoisted
+because each hoisting is performed only when at least one identical computation
+is found.
 

### 2016-10-24 slides.tex
Added the slides for llvm-dev

\documentclass{beamer}
\usepackage{hyperref}
\begin{document}

\def \GVN {GVN}

\title{\GVN{}-Hoist: Hoisting Computations from Branches}
\author{Sebastian Pop and Aditya Kumar}
\institute{SARC: Samsung Austin R\&D Center}
\date{\today}

\frame{\titlepage}

\frame{\frametitle{\GVN{}-Hoist: Hoisting Computations from Branches}
  \begin{itemize}
    \item Identifies identical computations in a function
    \item Hoist identical computations to a common dominator
    \item Uses GVN infrastructure of LLVM
    \item Reduces code size
  \end{itemize}
}

\frame{\frametitle{Open source tools}
  \begin{itemize}
    \item Compiler: gcc, llvm
    \item Gprof
    \item Linux Perf
    \item Create\_gcov
    \item Gooda
  \end{itemize}
}

\frame{\frametitle{PGO: Compiler instrumentation}
  \begin{itemize}
    \item Compiler inserts probes in the code.
    \item Instrumented program collects profile as it executes.
    \item Gives very precise profile information.
    \item Good for benchmarking, analyzing small programs.
    \item gprof/gcov used to process sampled output and display.
  \end{itemize}
}

\frame{\frametitle{PGO: Compiler instrumentation (Useful Commands)}
  gprof:
  \begin{itemize}
    \item gcc -fprofile-generate test.cpp
    \item gprof [exe] [gmon.out] $>$ [outfile]
  \end{itemize}
  gcov:
  \begin{itemize}
    \item gcc -fprofile-arcs -ftest-coverage test.cpp
    \item gcov [options] test.cpp
  \end{itemize}
}

\frame{\frametitle{AutoFDO: Non Intrusive}
  \begin{itemize}
    \item Linux perf collects profiles (Sample based)
    \item No compiler instrumentation, debug-info required while running.
    \item Negligible overhead (1-2\%) [Google]
    \item Run on different machines and at different time intervals. 
    \item Good for system wide profiling.
  \end{itemize}
}

\frame{\frametitle{AutoFDO: Non Intrusive (Useful Commands)}
  \begin{itemize}
    \item perf record/report/stat
    \item collecting profile: perf record -b [exe] [-o perf.data]
    \item report: perf report [perf.data]
  \end{itemize}
}

\frame{\frametitle{Results [Baptiste Wicht et al.]}
    \begin{tabular}{| l | p {2.5cm} | p {3cm} | }
        \hline
        PGO Kind & Avg. Run Time Overhead (\%) & Performance Improvement (\%) \\ \hline
        Instrumented & 16 (highest 53) & 7 to 14  \\ \hline
        AutoFDO & 1 to 2 & 5 to 10 \\ \hline
        AutoFDO+LTO & 1 to 2 & 10 to 30 \\ \hline
    \end{tabular}
}

\frame{\frametitle{Challenges for non-intrusive AutoFDO}
  \begin{itemize}
    \item Value based profiling not available.
    \item Only works on recent Intel machines.
    \item Tradeoff between accuracy of profile and overhead.
    \item Goal is to be as close to the instrumented profile.
  \end{itemize}
}

\frame{\frametitle{Challenges for non-intrusive AutoFDO...}
  \begin{itemize}
    \item Calculating missing edge/basic-block frequencies based on Equivalence class.
    \item Calculating accurate profile of execution paths sharing single line of code based on DWARF discriminator [Baptiste Wicht et al.].
  \end{itemize}
}

\frame{\frametitle{create\_gcov}
    \begin{itemize}
        \item Creates coverage file from perf.data to be used by the compiler.
        \item optimize with profile information: gcc -fauto-profile=file.gcov -O2 test.cpp
        \item Source: https://github.com/google/autofdo
     \end{itemize}
}

\frame{\frametitle{gooda}
    \begin{itemize}
        \item gooda-analyzer
        \item gooda-visualizer
        \item Source: https://github.com/David-Levinthal/gooda
        \item Requires patching linux perf to link against libpfm4 [See: gooda-analyzer/README] 
        \item Requires LBR (Available in Intel Sandy Bridge, Ivy Bridge, Westmere)
    \end{itemize}
}

\frame{\frametitle{Interesting Areas}
  Tools related:
  \begin{itemize}
    \item Extend PGO infrastructure in gcc/llvm:
      currently PGO benefits inliner, code layout, register allocation.
    \item Compiler cost model based on profile to enable demand driven optimization.
    \item Online optimizations driven by hardware performance monitoring [Schneider et al.].
  \end{itemize}
  Architecture specific:
  \begin{itemize}
    \item No AutoFDO tool for generating value profiles.
    \item No LBR for ARM.
  \end{itemize}

}

\frame{\frametitle{References}
    \begin{itemize}
    \item \url{http://static.googleusercontent.com/media/research.google.com/en//pubs/archive/36575.pdf}
    \item Baptiste Wicht et al. ``Hardware Counted Profile-Guided Optimization.''
    \item \url{http://www.burningcutlery.com/derek/docs/instant-profiling-CGO13.pdf}
    \item \url{https://github.com/David-Levinthal/gooda}
    \item \url{https://github.com/google/autofdo}
    \end{itemize}
}

\end{document}

### 2016-10-28 libcxx-slides.tex
libcxx-slides

\documentclass{beamer}
\usepackage{hyperref}
\begin{document}

\def \GVN {GVN}

\title{libcxx}
\author{Sebastian Pop and Aditya Kumar}
\institute{SARC: Samsung Austin R\&D Center}
\date{\today}

\frame{\titlepage}

\frame{\frametitle{libcxx-performance improvements}
  \begin{itemize}
  \item PR21192 - Reading from stdin is 1-2 orders of magnitude slower than using libstdc++
    - implementation of sync\_with\_stdio doesn't disable synchronization \url{http://stackoverflow.com/q/33837349}

  \item PR19708 - std::find is significantly slower than libstdc++
    - using Duff's device
    - unrolling the loop
    - enabling vectorization \url{http://stackoverflow.com/q/2741859}

  \item PR20837 - libc++'s std::sort is O(N^2) in the worst case (instead of O(N*ln(N))).
    - switching to heapsort for large size

  \item PR26886 - libc++'s std::stable\_sort also has a worst-case complexity issue

  \item PR15456 - A faster implementation of std::function is possible

  \item PR16747 and PR21275 - Our unordered\_multimap insert is much slower than libstdc++'s

  \item Removing Heap usage

  \item Missing features (and performance issues) in <regex>

  \item Non-conforming allocators in the test suite

  \item Identifying missing attributes in functions

  \item Debug mode for the library
  \end{itemize}
}

\end{document}

### 2016-11-01 hoist.tex
move files around

@@
   them to a common dominator so as to save code size.  Not only code-hoisting
-  saves code size, a) It exposes redundancies in some cases, b) It exposes more
+  saves code size, it exposes redundancies in some cases, it exposes more
   instruction level parallelism in the basic block a computation is hoisted, and
-  c) It enables other passes like LICM to remove more redundancies.
+  it enables other passes like LICM to remove more redundancies.
 
@@
 phase that reduces the number of run-time computations.  Classical scalar
-optimizations like CSE \cite{dragonbook} work very well on single basic blocks but
+optimizations like CSE \cite{dragonbook} work very well on single basic blocks and
 when it comes to detect redundancies across basic blocks these techniques fall
@@
 \item to improve function inlining heuristics: functions become cheaper to
-  inline when thier code size decreases because inliner heuristics in \LLVM{}
+  inline when their code size decreases because inliner heuristics in \LLVM{}
   depends on instruction count.

### 2016-11-01 other/libcxx-slides.tex
move files around

--- 2016-10-28 libcxx-slides.tex
@@
 
+  \item clang+libc++ vs. gcc+libstdc++ performance update. Constructor of vector in libstdc++ contains memcpy
+    in the final assembly.
+
+Benchmark	                        Time (ns)	CPU (ns)	Iterations	Time (ns)	CPU (ns)	Iterations	%(Time)
+BM_ConstructIterIter/vector_char/1024	2125	2127	283762	67	67	10205902	3071.641791
+BM_ConstructIterIter/vector_size_t/1024	382	383	1830386	223	223	3138652	71.30044843
+BM_ConstructIterIter/vector_string/1024	557109	557641	1235	354504	354834	1980	57.15168235
+
   \item Removing Heap usage

### 2016-11-01 slides/gvn-hoist.tex
move files around

--- 2016-10-24 slides.tex
@@
     \item Hoist identical computations to a common dominator
-    \item Uses GVN infrastructure of LLVM
+    \item Uses GVN and MemorySSA
     \item Reduces code size
@@
 
-\frame{\frametitle{Open source tools}
-  \begin{itemize}
-    \item Compiler: gcc, llvm
-    \item Gprof
-    \item Linux Perf
-    \item Create\_gcov
-    \item Gooda
-  \end{itemize}
-}
-
-\frame{\frametitle{PGO: Compiler instrumentation}
-  \begin{itemize}
-    \item Compiler inserts probes in the code.
-    \item Instrumented program collects profile as it executes.
-    \item Gives very precise profile information.
-    \item Good for benchmarking, analyzing small programs.
-    \item gprof/gcov used to process sampled output and display.
-  \end{itemize}
-}
-
-\frame{\frametitle{PGO: Compiler instrumentation (Useful Commands)}
-  gprof:
-  \begin{itemize}
-    \item gcc -fprofile-generate test.cpp
-    \item gprof [exe] [gmon.out] $>$ [outfile]
-  \end{itemize}
-  gcov:
-  \begin{itemize}
-    \item gcc -fprofile-arcs -ftest-coverage test.cpp
-    \item gcov [options] test.cpp
-  \end{itemize}
-}
-
-\frame{\frametitle{AutoFDO: Non Intrusive}
-  \begin{itemize}
-    \item Linux perf collects profiles (Sample based)
-    \item No compiler instrumentation, debug-info required while running.
-    \item Negligible overhead (1-2\%) [Google]
-    \item Run on different machines and at different time intervals. 
-    \item Good for system wide profiling.
-  \end{itemize}
-}
-
-\frame{\frametitle{AutoFDO: Non Intrusive (Useful Commands)}
-  \begin{itemize}
-    \item perf record/report/stat
-    \item collecting profile: perf record -b [exe] [-o perf.data]
-    \item report: perf report [perf.data]
-  \end{itemize}
-}
-
-\frame{\frametitle{Results [Baptiste Wicht et al.]}
-    \begin{tabular}{| l | p {2.5cm} | p {3cm} | }
-        \hline
-        PGO Kind & Avg. Run Time Overhead (\%) & Performance Improvement (\%) \\ \hline
-        Instrumented & 16 (highest 53) & 7 to 14  \\ \hline
-        AutoFDO & 1 to 2 & 5 to 10 \\ \hline
-        AutoFDO+LTO & 1 to 2 & 10 to 30 \\ \hline
-    \end{tabular}
-}
-
-\frame{\frametitle{Challenges for non-intrusive AutoFDO}
-  \begin{itemize}
-    \item Value based profiling not available.
-    \item Only works on recent Intel machines.
-    \item Tradeoff between accuracy of profile and overhead.
-    \item Goal is to be as close to the instrumented profile.
-  \end{itemize}
-}
-
-\frame{\frametitle{Challenges for non-intrusive AutoFDO...}
-  \begin{itemize}
-    \item Calculating missing edge/basic-block frequencies based on Equivalence class.
-    \item Calculating accurate profile of execution paths sharing single line of code based on DWARF discriminator [Baptiste Wicht et al.].
-  \end{itemize}
-}
-
-\frame{\frametitle{create\_gcov}
-    \begin{itemize}
-        \item Creates coverage file from perf.data to be used by the compiler.
-        \item optimize with profile information: gcc -fauto-profile=file.gcov -O2 test.cpp
-        \item Source: https://github.com/google/autofdo
-     \end{itemize}
-}
-
-\frame{\frametitle{gooda}
-    \begin{itemize}
-        \item gooda-analyzer
-        \item gooda-visualizer
-        \item Source: https://github.com/David-Levinthal/gooda
-        \item Requires patching linux perf to link against libpfm4 [See: gooda-analyzer/README] 
-        \item Requires LBR (Available in Intel Sandy Bridge, Ivy Bridge, Westmere)
-    \end{itemize}
-}
-
-\frame{\frametitle{Interesting Areas}
-  Tools related:
-  \begin{itemize}
-    \item Extend PGO infrastructure in gcc/llvm:
-      currently PGO benefits inliner, code layout, register allocation.
-    \item Compiler cost model based on profile to enable demand driven optimization.
-    \item Online optimizations driven by hardware performance monitoring [Schneider et al.].
-  \end{itemize}
-  Architecture specific:
-  \begin{itemize}
-    \item No AutoFDO tool for generating value profiles.
-    \item No LBR for ARM.
-  \end{itemize}
-
-}
-
 \frame{\frametitle{References}
     \begin{itemize}
-    \item \url{http://static.googleusercontent.com/media/research.google.com/en//pubs/archive/36575.pdf}
-    \item Baptiste Wicht et al. ``Hardware Counted Profile-Guided Optimization.''
-    \item \url{http://www.burningcutlery.com/derek/docs/instant-profiling-CGO13.pdf}
-    \item \url{https://github.com/David-Levinthal/gooda}
-    \item \url{https://github.com/google/autofdo}
     \end{itemize}

### 2016-11-01 slides/gvn-hoist.tex
add page numbers to slides
slides

@@
 \usepackage{hyperref}
+\usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{graphviz}
+\usepackage{tikz}
+\usepackage{dot2texi}
+\usepackage{listings}
+
+\pgfdeclarelayer{background}
+\pgfdeclarelayer{foreground}
+\pgfsetlayers{background,main,foreground}
+
 \begin{document}
-
 \def \GVN {GVN}
@@
 \institute{SARC: Samsung Austin R\&D Center}
-\date{\today}
+\date{November 3, 2016}
+
+\definecolor{myblue}{rgb}{0.0, 0.0, 0.5}
+\definecolor{myred}{rgb}{0.5, 0.0, 0.0}
+\definecolor{mygreen}{rgb}{0.0, 0.5, 0.0}
+\lstset{language=C++,
+  basicstyle=\ttfamily,
+  keywordstyle=\color{myblue}\ttfamily,
+  stringstyle=\color{myred}\ttfamily,
+  commentstyle=\color{mygreen}\ttfamily,
+  morecomment=[l][\color{magenta}]{\#}
+}
+\addtobeamertemplate{navigation symbols}{}{%
+    \usebeamerfont{footline}%
+    \usebeamercolor[fg]{footline}%
+    \hspace{1em}%
+    \insertframenumber/\inserttotalframenumber
+}
 
@@
 
-\frame{\frametitle{References}
+\frame{\frametitle{CFGSimplify's code hoisting}
   \begin{itemize}
+  \item identifies computations to be hoisted at the beginning of BB
+  \item stops at first difference
+  \item very fast
+  \item catches a lot of low hanging fruit
+  \end{itemize}
+}
+
+\frame{\frametitle{Code size reduction}
+  \begin{itemize}
+  \item an average of 2.5\% savings in code size
+  \item improve inline heuristics
+  \item more functions inlined
   \end{itemize}

### 2016-11-02 slides/gvn-hoist.tex
algorithm overview
vn and insertion points

@@
 
+\frame{\frametitle{\GVN{}-Hoist: Algorithm}
+  \begin{itemize}
+  \item dfs number all instructions
+  \item compute value number of scalars, loads, stores, calls
+  \item compute insertion points of each type of instructions separately
+  \item hoist expressions
+  \end{itemize}
+}
+
+\frame{\frametitle{\GVN{}-Hoist: Algorithm-collecting value numbers}
+  \begin{itemize}
+  \item For loads value number the pointer operand
+  \item For stores value number the pointer operand and the value being stored
+  \item For calls, scalars use the GVN infrastructure
+  \end{itemize}
+}
+
+\frame{\frametitle{\GVN{}-Hoist: Algorithm-compute insertion points}
+  \begin{itemize}
+  \item For a set of instructions having the same value number, compute a common insertion point.
+  \item When no common insertion points can be found, partition the candidates into a smaller set of
+    hoistable candidates.
+  \item Partition such that instructions with closeby DFS numbers are together.
+  \end{itemize}
+}
+
 \frame{\frametitle{CFGSimplify's code hoisting}

### 2016-11-02 slides/gvn-hoist.tex
hoist
partition and cost models

--- 2016-11-01 slides/gvn-hoist.tex
@@
 
+\frame{\frametitle{Optimistic \GVN{}-hoist Algorithm}
+  \begin{itemize}
+  \item DFS number all instructions
+  \item Compute value number of scalars, loads, stores, calls
+  \item Compute insertion points of each type of instructions
+  \item Hoist expressions
+  \end{itemize}
+}
+
+\frame{\frametitle{\GVN{}-Hoist: Algorithm-collecting value numbers}
+  \begin{itemize}
+  \item Loads: value number the pointer operand
+  \item Stores: value number the pointer operand and the value being stored
+  \item Calls, Scalars: use the existing GVN infrastructure
+  \end{itemize}
+}
+
+\frame{\frametitle{\GVN{}-Hoist: Algorithm-compute insertion points}
+ Insertion Point: A location where all the operands are either available
+    or, can be made available.
+
+  \begin{itemize}
+  \item Compute a common insertion point for a set of instructions having the same GVN.
+    (Similar to VBEs but not as strict)
+  \item Partition the candidates into a smaller set of hoistable candidates when
+    no common insertion points can be found (The instructions are sorted by their DFS numbers)
+  \end{itemize}
+}
+
+\frame{\frametitle{\GVN{}-Hoist: Algorithm-hoist expressions}
+  \begin{itemize}
+  \item Scalars: Just move one of the instrutions to the hoisting point and remove others. Update SSA.
+  \item Loads and Stores: Try to make geps available (if not), and then hoist. Update SSA and memory SSA.
+  \item Calls: Hoist just like scalars. Update SSA.
+  \end{itemize}
+}
+
 \frame{\frametitle{CFGSimplify's code hoisting}
@@
   \item catches a lot of low hanging fruit
+  \end{itemize}
+}
+
+\frame{\frametitle{Cost models}
+  \begin{itemize}
+  \item Do not hoist too far
+  \item Do not hoist GEPs
   \end{itemize}

### 2016-11-02 slides/gvn-hoist.tex
cost model

@@
     \item Hoist identical computations to a common dominator
-    \item Uses GVN and MemorySSA
     \item Reduces code size
+    \item Reduces critical path lenth in some cases by exposing more ILP.
   \end{itemize}
 }
+
+; TODO: Examples of hoisting
 
@@
   \item Hoist expressions
+  \item Propagate changes by upating SSA
   \end{itemize}
@@
   \item Calls: Hoist just like scalars. Update SSA.
+  \item Set minimum alignment among all the candidates for loads and stores, maximum for allocas.
   \end{itemize}
@@
   \begin{itemize}
-  \item Do not hoist too far
+  \item Limit the number of Basic Blocks in the path between initial position
+    and the hoisting point
+  \item Limit the number of instructions between the initial position and the
+    beginning of its Basic Block
   \item Do not hoist GEPs
+  \item Limit the number of dependent instructions to be hoisted
   \end{itemize}

### 2016-11-02 slides/gvn-hoist.tex
add eval section

@@
   \begin{itemize}
-    \item Identifies identical computations in a function
-    \item Hoist identical computations to a common dominator
-    \item Reduces code size
-    \item Reduces critical path lenth in some cases by exposing more ILP.
+    \item identifies identical computations in a function
+    \item hoist identical computations to a common dominator
+    \item reduces code size
+    \item reduces critical path lenth by exposing more ILP
   \end{itemize}
@@
 
-; TODO: Examples of hoisting
+\begin{frame}[fragile]{Example of hoisting}
+  \begin{lstlisting}
+  if (inv >= 0) {
+    tmin = (min - a) * inv;
+    tmax = (max - a) * inv;
+  } else {
+    tmin = (max - a) * inv;
+    tmax = (min - a) * inv;
+  }
+  \end{lstlisting}
+
+  $\hspace{2cm}\Big\downarrow$
+
+  \begin{lstlisting}
+  x = (min - a) * inv;
+  y = (max - a) * inv;
+  if (inv >= 0) {
+    tmin = x;
+    tmax = y;
+  } else {
+    tmin = y;
+    tmax = x;
+  }
+  \end{lstlisting}
+\end{frame}
 
@@
   \begin{itemize}
-  \item DFS number all instructions
-  \item Compute value number of scalars, loads, stores, calls
-  \item Compute insertion points of each type of instructions
-  \item Hoist expressions
-  \item Propagate changes by upating SSA
+  \item compute value number of scalars, loads, stores, calls
+  \item compute insertion points of each type of instructions
+  \item hoist expressions
+  \item propagate changes by upating SSA
   \end{itemize}
@@
   \begin{itemize}
-  \item Loads: value number the pointer operand
-  \item Stores: value number the pointer operand and the value being stored
-  \item Calls, Scalars: use the existing GVN infrastructure
+  \item loads: value number the pointer operand
+  \item stores: value number the pointer operand and the value being stored
+  \item calls, Scalars: use the existing GVN infrastructure
   \end{itemize}
@@
   \begin{itemize}
-  \item Scalars: Just move one of the instrutions to the hoisting point and remove others. Update SSA.
+  \item Scalars: Just move one of the instructions to the hoisting point and remove others. Update SSA.
   \item Loads and Stores: Try to make geps available (if not), and then hoist. Update SSA and memory SSA.
@@
   \begin{itemize}
-  \item identifies computations to be hoisted at the beginning of BB
+  \item hoists computations at the beginning of BB
   \item stops at first difference
-  \item very fast
-  \item catches a lot of low hanging fruit
+  \item very fast: disabling slows the compiler: $1688 \to 1692$ Bn insns
   \end{itemize}
+}
+
+\frame{\frametitle{GVN hoisting}
+  \begin{itemize}
+  \item $1\%$ compile time overhead: $1678 \to 1692$ Bn insns
+  \item more hoists than CFG-simplify: $15048 \to 25318$
+  \end{itemize}
+
+  \begin{center}
+    \begin{tabular}{|l|c|}
+      \hline
+      Scalars hoisted      & 8960  \\\hline
+      Scalars removed      & 11940 \\\hline
+      Loads hoisted        & 16301 \\\hline
+      Loads removed        & 22690 \\\hline
+      Stores hoisted       & 50    \\\hline
+      Stores removed       & 50    \\\hline
+      Calls hoisted        & 7     \\\hline
+      Calls removed        & 7     \\\hline
+      Total Instructions hoisted & 25318 \\\hline
+      Total Instructions removed & 34687 \\\hline
+    \end{tabular}
+  \end{center}
 }
@@
 \frame{\frametitle{Code size reduction}
-  \begin{itemize}
-  \item an average of 2.5\% savings in code size
-  \item improve inline heuristics
-  \item more functions inlined
-  \end{itemize}
+  \begin{center}
+    \begin{tabular}{|l|c|}
+      \hline
+      Code-size metric  (.text)                   & Number   \\\hline
+      Total  benchmarks                           & 497      \\\hline
+      Total  gained in size                       & 39       \\\hline
+      Total  decrease in size                     & 58       \\\hline
+      Median decrease in size                     & 2.9\%    \\\hline
+      Median increase in size                     & 2.4\%    \\\hline
+    \end{tabular}
+  \end{center}
 }

### 2016-11-02 slides/gvn-hoist.tex
clean up

@@
   \item hoist expressions
-  \item propagate changes by upating SSA
+  \item propagate changes by updating SSA
   \end{itemize}
@@
 \frame{\frametitle{\GVN{}-Hoist: Algorithm-collecting value numbers}
+  current GVN not accurate for loads and stores:
   \begin{itemize}
-  \item loads: value number the pointer operand
-  \item stores: value number the pointer operand and the value being stored
-  \item calls, Scalars: use the existing GVN infrastructure
+  \item scalars: use the existing GVN infrastructure
+  \item loads: VN the pointer operand
+  \item stores: VN the pointer operand and the value being stored
+  \item calls: as stores, loads, or scalars (following side effects)
   \end{itemize}
@@
   \begin{itemize}
-  \item Compute a common insertion point for a set of instructions having the same GVN.
+  \item Compute a common insertion point for a set of instructions having the same GVN
     (Similar to VBEs but not as strict)
   \item Partition the candidates into a smaller set of hoistable candidates when
-    no common insertion points can be found (The instructions are sorted by their DFS numbers)
+    no common insertion points can be found
   \end{itemize}
@@
   \begin{itemize}
-  \item Scalars: Just move one of the instructions to the hoisting point and remove others. Update SSA.
-  \item Loads and Stores: Try to make geps available (if not), and then hoist. Update SSA and memory SSA.
-  \item Calls: Hoist just like scalars. Update SSA.
-  \item Set minimum alignment among all the candidates for loads and stores, maximum for allocas.
+  \item scalars: just move one of the instructions to the hoisting point and remove others; update SSA
+  \item loads and stores: try to make geps available, then hoist; update SSA and memory SSA
   \end{itemize}
@@
   \begin{itemize}
-  \item Limit the number of Basic Blocks in the path between initial position
+  \item limit the number of basic blocks in the path between initial position
     and the hoisting point
-  \item Limit the number of instructions between the initial position and the
-    beginning of its Basic Block
-  \item Do not hoist GEPs
-  \item Limit the number of dependent instructions to be hoisted
+  \item limit the number of instructions between the initial position and the
+    beginning of its basic block
+  \item do not hoist GEPs
+  \item limit the number of dependent instructions to be hoisted
   \end{itemize}

### 2016-11-03 slides/gvn-hoist.tex
clean up
move cfgsimplify in introduction
more good stuff

@@
 \usepackage{listings}
+\usepackage{adjustbox}
 
@@
 \def \GVN {GVN}
+\def \SSA {SSA}
+\def \MemorySSA {Memory-\SSA{}}
 
@@
 
-\frame{\frametitle{\GVN{}-Hoist: Hoisting Computations from Branches}
+\frame{\frametitle{CFGSimplify's code hoisting}
   \begin{itemize}
-    \item identifies identical computations in a function
-    \item hoist identical computations to a common dominator
-    \item reduces code size
-    \item reduces critical path lenth by exposing more ILP
+  \item hoists computations at the beginning of BB
+  \item uses operands equality to detect same computations
+  \item stops at first difference
+  \item very fast: disabling it slows the compiler: {\small $1688 \to 1692$ Bn insns
+    (callgrind compiling the test-suite on x86\_64-linux)}
   \end{itemize}
@@
 
-\begin{frame}[fragile]{Example of hoisting}
+\begin{frame}[fragile]{CFGSimplify limits}
+  \begin{columns}[T,onlytextwidth] % align columns
+    \column{.4\textwidth}
+    \begin{block}{\small Original program}
       \begin{lstlisting}
+inv = 1/d;
 if (inv >= 0) {
-    tmin = (min - a) * inv;
-    tmax = (max - a) * inv;
+  tmin = min * inv;
+  tmax = max * inv;
 } else {
-    tmin = (max - a) * inv;
-    tmax = (min - a) * inv;
+  tmin = max * inv;
+  tmax = min * inv;
 }
       \end{lstlisting}
-
-  $\hspace{2cm}\Big\downarrow$
-
+    \end{block}
+    \pause
+    \column{.1\textwidth}
+    \begin{center}
+      \vspace{2cm}
+      $\longrightarrow$
+    \end{center}
+    \column{.4\textwidth}
+    \begin{block}{\small Expressions hoisted}
       \begin{lstlisting}
-  x = (min - a) * inv;
-  y = (max - a) * inv;
+inv = 1/d;
+x = min * inv;
+y = max * inv;
 if (inv >= 0) {
@@
       \end{lstlisting}
+    \end{block}
+  \end{columns}
 \end{frame}
 
-\frame{\frametitle{Optimistic \GVN{}-hoist Algorithm}
+\frame{\frametitle{\GVN{}-Hoist: Hoisting Computations from Branches}
   \begin{itemize}
-  \item compute value number of scalars, loads, stores, calls
-  \item compute insertion points of each type of instructions
-  \item hoist expressions
-  \item propagate changes by updating SSA
+  \item removes all limitations of CFGSimplify implementation
+  \item works across several BBs: hoists to a common dominator
+  \item hoist past ld/st side effects: uses \MemorySSA{} for fast dependence analysis
+  \item reduces code size
+  \item reduces critical path length by exposing more ILP
   \end{itemize}
@@
 
-\frame{\frametitle{\GVN{}-Hoist: Algorithm-collecting value numbers}
-  current GVN not accurate for loads and stores:
+\frame{\frametitle{Optimistic \GVN{}-hoist Algorithm}
+  \begin{enumerate}
+  \item compute value number of scalars, loads, stores, calls
+  \item compute insertion points of each type of instructions
+  \item hoist expressions and propagate changes by updating \SSA{}
+  \end{enumerate}
+}
+
+\begin{frame}[fragile]{\GVN{}: Value Numbering Example and Limitations}
+  \begin{columns}[T,onlytextwidth] % align columns
+    \column{.3\textwidth}
+    \begin{block}{\small Simple program}
+      \begin{lstlisting}
+a = x + y
+b = x + 1
+c = y + 1
+d = b + c
+e = a + 2
+f = load d
+g = load e
+      \end{lstlisting}
+    \end{block}
+    \pause
+    \column{.1\textwidth}
+    \begin{center}
+      \vspace{1cm}
+      $\longrightarrow$
+    \end{center}
+    \column{.4\textwidth}
+    \begin{block}{\small Value Numbering}
+      \begin{lstlisting}
+(a, 1)
+(b, 2)
+(c, 3)
+(d, 4)
+(e, 4)
+      \end{lstlisting}
+    \end{block}
+    \pause
+    \begin{block}{\small Limitations to current \GVN{} implementation}
+      \begin{lstlisting}
+(f, 5)
+(g, 6)
+// should be (g, 5)
+      \end{lstlisting}
+    \end{block}
+  \end{columns}
+
+\end{frame}
+
+\frame{\frametitle{\GVN{}-Hoist Step 1: Collect Value Numbers}
   \begin{itemize}
-  \item scalars: use the existing GVN infrastructure
-  \item loads: VN the pointer operand
-  \item stores: VN the pointer operand and the value being stored
-  \item calls: as stores, loads, or scalars (following side effects)
+  \item scalars: use the existing \GVN{} infrastructure
+  \end{itemize}
+  current \GVN{} not accurate for loads and stores: use ad-hoc change
+  \begin{itemize}
+  \item loads: VN the gep
+  \item stores: VN the gep and stored value
+  \item calls: as stores, loads, or scalars (following calls' side-effects)
   \end{itemize}
@@
 
-\frame{\frametitle{\GVN{}-Hoist: Algorithm-compute insertion points}
-  Insertion Point: A location where all the operands are either available
-    or, can be made available.
+\frame{\frametitle{\GVN{}-Hoist Step 2: Compute Insertion Points}
+  insertion point: location where all the operands are available
 
   \begin{itemize}
-  \item Compute a common insertion point for a set of instructions having the same GVN
-    (Similar to VBEs but not as strict)
-  \item Partition the candidates into a smaller set of hoistable candidates when
+  \item compute a common insertion point for a set of instructions having the same \GVN{}
+    (similar to VBEs but not as strict)
+  \item partition the candidates into a smaller set of hoistable candidates when
     no common insertion points can be found
@@
 
-\frame{\frametitle{\GVN{}-Hoist: Algorithm-hoist expressions}
+\frame{\frametitle{\GVN{}-Hoist Step 3: Move the Code}
   \begin{itemize}
-  \item scalars: just move one of the instructions to the hoisting point and remove others; update SSA
-  \item loads and stores: try to make geps available, then hoist; update SSA and memory SSA
+  \item scalars: just move one of the instructions to the hoisting point and remove others; update \SSA{}
+  \item loads and stores: make geps available, then hoist; update \SSA{} and \MemorySSA{}
   \end{itemize}
@@
 
-\frame{\frametitle{CFGSimplify's code hoisting}
+\newlength\someheight
+\setlength\someheight{4cm}
+
+\begin{frame}[fragile]{Example}
+\begin{adjustbox}{width=\textwidth,height=\someheight,keepaspectratio}
+\begin{lstlisting}
+define float @f(float %d, float %min, float %max, float %a) {
+entry:
+  %div = fdiv float 1.000000e+00, %d
+  %cmp = fcmp oge float %div, 0.000000e+00
+  br i1 %cmp, label %if.then, label %if.else
+
+if.then:                                          ; preds = %entry
+  %sub = fsub float %min, %a
+  %mul = fmul float %sub, %div
+  %sub1 = fsub float %max, %a
+  %mul2 = fmul float %sub1, %div
+  br label %if.end
+
+if.else:                                          ; preds = %entry
+  %sub3 = fsub float %max, %a
+  %mul4 = fmul float %sub3, %div
+  %sub5 = fsub float %min, %a
+  %mul6 = fmul float %sub5, %div
+  br label %if.end
+
+if.end:                                           ; preds = %if.else, %if.then
+  %tmax.0 = phi float [ %mul2, %if.then ], [ %mul6, %if.else ]
+  %tmin.0 = phi float [ %mul, %if.then ], [ %mul4, %if.else ]
+  %add = fadd float %tmax.0, %tmin.0
+  ret float %add
+}
+\end{lstlisting}
+\end{adjustbox}
+\end{frame}
+
+\begin{frame}[fragile]{Example}
+\begin{adjustbox}{width=\textwidth,height=\someheight,keepaspectratio}
+\begin{lstlisting}
+define float @f(float %d, float %min, float %max, float %a) {
+entry:
+  %div = fdiv float 1.000000e+00, %d
+  %cmp = fcmp oge float %div, 0.000000e+00
+  %sub1 = fsub float %max, %a
+  %sub = fsub float %min, %a
+  %mul2 = fmul float %sub1, %div
+  %mul = fmul float %sub, %div
+  br i1 %cmp, label %if.then, label %if.else
+
+if.then:                                          ; preds = %entry
+  br label %if.end
+
+if.else:                                          ; preds = %entry
+  br label %if.end
+
+if.end:                                           ; preds = %if.else, %if.then
+  %tmax.0 = phi float [ %mul2, %if.then ], [ %mul, %if.else ]
+  %tmin.0 = phi float [ %mul, %if.then ], [ %mul2, %if.else ]
+  %add = fadd float %tmax.0, %tmin.0
+  ret float %add
+}
+\end{lstlisting}
+\end{adjustbox}
+\end{frame}
+
+\frame{\frametitle{Cost models}
+  tuned on x86\_64 and AArch64 Linux: test-suite, SPEC 2k, 2k6, \ldots{}
   \begin{itemize}
-  \item hoists computations at the beginning of BB
-  \item stops at first difference
-  \item very fast: disabling slows the compiler: $1688 \to 1692$ Bn insns
+  \item limit the number of basic blocks in the path between initial position
+    and the hoisting point
+  \item limit the number of instructions between the initial position and the
+    beginning of its basic block
+  \item do not hoist GEPs (except at -Os)
+  \item limit the number of dependent instructions to be hoisted
   \end{itemize}
@@
 
-\frame{\frametitle{GVN hoisting}
+\frame{\frametitle{Knobs}
   \begin{itemize}
-  \item $1\%$ compile time overhead: $1678 \to 1692$ Bn insns
-  \item more hoists than CFG-simplify: $15048 \to 25318$
+  \item {\bf -enable-gvn-hoist:} enable the \GVN{}-hoist pass {\small(default = on)}
+  \item {\bf -Os, -Oz:} allow GEPs to be hoisted independently of ld/st
+  \item {\bf -gvn-hoist-max-bbs:} max number of basic blocks on the path between
+    hoisting locations {\small (default = 4, unlimited = -1)}
+  \item {\bf -gvn-hoist-max-depth:} hoist instructions from the beginning of the BB up
+    to the maximum specified depth {\small (default = 100, unlimited = -1)}
+  \item {\bf -gvn-hoist-max-chain-length:} maximum length of dependent chains to hoist
+    {\small (default = 10, unlimited = -1)}
+  \item {\bf -gvn-max-hoisted:} max number of instructions to hoist {\small (default unlimited = -1)}
+  \end{itemize}
+}
+\frame{\frametitle{\GVN{}-Hoist: Evaluation}
+  \begin{itemize}
+  \item $<1\%$ compile time overhead: $1678 \to 1692$ Bn insns \\
+    {\small (callgrind compiling the test-suite at -O3 on x86\_64-linux)}
+  \item more hoists than CFG-simplify: $15048 \to 25318$ \\
+    {\small (compiling the test-suite for x86\_64 at -O3)}
   \end{itemize}
@@
 
-\frame{\frametitle{Cost models}
-  \begin{itemize}
-  \item limit the number of basic blocks in the path between initial position
-    and the hoisting point
-  \item limit the number of instructions between the initial position and the
-    beginning of its basic block
-  \item do not hoist GEPs
-  \item limit the number of dependent instructions to be hoisted
-  \end{itemize}
-}
-
 \frame{\frametitle{Code size reduction}
@@
   \end{center}
+  \begin{itemize}
+  \item test-suite compiled at -O3 for x86\_64-linux
+  \item increase in size due to more inlining
+  \item many effects due to early scheduling of the pass
+  \end{itemize}
+}
+
+\frame{\frametitle{Discussion}
+  \begin{itemize}
+  \item schedule \GVN{}-hoist pass several times?
+  \item remove CFGSimplify's hoisting?
+  \item hoist + sink interactions (discuss with James Molloy)
+  \item early scheduling in opt needs tuning with target info?
+  \item make \GVN{}-hoist more aggressive for -Os and -Oz?
+  \item need a better \GVN{} implementation?
+  \item \MemorySSA{} is easy to use and fast: {\bf \color{myred}so please use it!} \\
+    {\small(thanks Danny, Georges, and others)}
+  \end{itemize}
 }

### 2016-11-03 slides/gvn-hoist.tex
fix

--- 2016-11-02 slides/gvn-hoist.tex
@@
 
-\frame{\frametitle{\GVN{}-Hoist: Hoisting Computations from Branches}
+\frame{\frametitle{CFGSimplify's code hoisting}
   \begin{itemize}
-    \item identifies identical computations in a function
-    \item hoist identical computations to a common dominator
-    \item reduces code size
-    \item reduces critical path lenth by exposing more ILP
+  \item hoists computations at the beginning of BB
+  \item stops at first difference
+  \item very fast: disabling slows the compiler: $1688 \to 1692$ Bn insns
   \end{itemize}
@@
 
-\begin{frame}[fragile]{Example of hoisting}
+\begin{frame}[fragile]{CFGSimplify limits}
   \begin{lstlisting}
@@
 
+\frame{\frametitle{\GVN{}-Hoist: Hoisting Computations from Branches}
+  \begin{itemize}
+    \item identifies identical computations in a function
+    \item hoist identical computations to a common dominator
+    \item reduces code size
+    \item reduces critical path length by exposing more ILP
+  \end{itemize}
+}
+
 \frame{\frametitle{Optimistic \GVN{}-hoist Algorithm}
@@
   \item compute insertion points of each type of instructions
-  \item hoist expressions
-  \item propagate changes by updating SSA
+  \item hoist expressions and propagate changes by updating SSA
   \end{itemize}
@@
 
+\begin{frame}[fragile]{Value Numbering}
+
+  \begin{columns}[T,onlytextwidth] % align columns
+    \column{.3\textwidth}
+    \begin{block}{\small Simple program}
+      \begin{lstlisting}
+a = x + y
+b = x + 1
+c = y + 1
+d = b + c
+e = a + 2
+      \end{lstlisting}
+    \end{block}
+    \pause
+    \column{.2\textwidth}
+    \begin{center}
+      \vspace{1cm}
+      $\rightarrow$
+    \end{center}
+    \column{.3\textwidth}
+    \begin{block}{\small Value Numbering}
+      \begin{lstlisting}
+a -> 1
+b -> 2
+c -> 3
+d -> 4
+e -> 4
+      \end{lstlisting}
+    \end{block}
+  \end{columns}
+
+\end{frame}
+
 \frame{\frametitle{\GVN{}-Hoist: Algorithm-collecting value numbers}
-  current GVN not accurate for loads and stores:
   \begin{itemize}
@@
   \end{itemize}
+  current GVN not accurate for loads and stores: use ad-hoc change
 }
@@
 
-\frame{\frametitle{CFGSimplify's code hoisting}
+\frame{\frametitle{Cost models}
   \begin{itemize}
-  \item hoists computations at the beginning of BB
-  \item stops at first difference
-  \item very fast: disabling slows the compiler: $1688 \to 1692$ Bn insns
+  \item limit the number of basic blocks in the path between initial position
+    and the hoisting point
+  \item limit the number of instructions between the initial position and the
+    beginning of its basic block
+  \item do not hoist GEPs
+  \item limit the number of dependent instructions to be hoisted
   \end{itemize}

### 2016-11-03 slides/gvn-hoist.tex
adding example
simplify

--- 2016-11-03 slides/gvn-hoist.tex
@@
 if (inv >= 0) {
-  tmin = min * inv;
-  tmax = max * inv;
+  ta = a * inv;
+  tb = b * inv;
 } else {
-  tmin = max * inv;
-  tmax = min * inv;
+  ta = b * inv;
+  tb = a * inv;
 }
@@
 inv = 1/d;
-x = min * inv;
-y = max * inv;
+x = a * inv;
+y = b * inv;
 if (inv >= 0) {
-  tmin = x;
-  tmax = y;
+  ta = x;
+  tb = y;
 } else {
-  tmin = y;
-  tmax = x;
+  ta = y;
+  tb = x;
 }

### 2016-11-03 slides/gvn-hoist.tex
simplify
remove llvm ir example: it is not really needed

@@
       \begin{lstlisting}
-inv = 1/d;
-if (inv >= 0) {
-  ta = a * inv;
-  tb = b * inv;
+i = 1/d;
+if (i >= 0) {
+  u = a * i;
+  v = b * i;
 } else {
-  ta = b * inv;
-  tb = a * inv;
+  u = b * i;
+  v = a * i;
 }
@@
       \begin{lstlisting}
-inv = 1/d;
-x = a * inv;
-y = b * inv;
-if (inv >= 0) {
-  ta = x;
-  tb = y;
+i = 1/d;
+x = a * i;
+y = b * i;
+if (i >= 0) {
+  u = x;
+  v = y;
 } else {
-  ta = y;
-  tb = x;
+  u = y;
+  v = x;
 }
@@
 
-\newlength\someheight
-\setlength\someheight{4cm}
-
-\begin{frame}[fragile]{Example}
-\begin{adjustbox}{width=\textwidth,height=\someheight,keepaspectratio}
-\begin{lstlisting}
-define float @f(float %d, float %min, float %max, float %a) {
-entry:
-  %div = fdiv float 1.000000e+00, %d
-  %cmp = fcmp oge float %div, 0.000000e+00
-  br i1 %cmp, label %if.then, label %if.else
-
-if.then:                                          ; preds = %entry
-  %sub = fsub float %min, %a
-  %mul = fmul float %sub, %div
-  %sub1 = fsub float %max, %a
-  %mul2 = fmul float %sub1, %div
-  br label %if.end
-
-if.else:                                          ; preds = %entry
-  %sub3 = fsub float %max, %a
-  %mul4 = fmul float %sub3, %div
-  %sub5 = fsub float %min, %a
-  %mul6 = fmul float %sub5, %div
-  br label %if.end
-
-if.end:                                           ; preds = %if.else, %if.then
-  %tmax.0 = phi float [ %mul2, %if.then ], [ %mul6, %if.else ]
-  %tmin.0 = phi float [ %mul, %if.then ], [ %mul4, %if.else ]
-  %add = fadd float %tmax.0, %tmin.0
-  ret float %add
-}
-\end{lstlisting}
-\end{adjustbox}
-\end{frame}
-
-\begin{frame}[fragile]{Example}
-\begin{adjustbox}{width=\textwidth,height=\someheight,keepaspectratio}
-\begin{lstlisting}
-define float @f(float %d, float %min, float %max, float %a) {
-entry:
-  %div = fdiv float 1.000000e+00, %d
-  %cmp = fcmp oge float %div, 0.000000e+00
-  %sub1 = fsub float %max, %a
-  %sub = fsub float %min, %a
-  %mul2 = fmul float %sub1, %div
-  %mul = fmul float %sub, %div
-  br i1 %cmp, label %if.then, label %if.else
-
-if.then:                                          ; preds = %entry
-  br label %if.end
-
-if.else:                                          ; preds = %entry
-  br label %if.end
-
-if.end:                                           ; preds = %if.else, %if.then
-  %tmax.0 = phi float [ %mul2, %if.then ], [ %mul, %if.else ]
-  %tmin.0 = phi float [ %mul, %if.then ], [ %mul2, %if.else ]
-  %add = fadd float %tmax.0, %tmin.0
-  ret float %add
-}
-\end{lstlisting}
-\end{adjustbox}
-\end{frame}
-
 \frame{\frametitle{Cost models}

### 2016-11-14 hoist.tex
hoisting at -Oz

@@
 
+When aggressively for code-size (at optimization level -Oz), we allow hoisting
+of scalars even when the expression is not very busy. This creates unnecessary
+computation on paths where the expression is not needed so the program may run slow
+but reduces the code size. TODO: Experimental results.
+
 \subsubsection{Profitability of hoisting loads}
@@
 during execution should the value is not in the cache. We generally prefer to
-hoist load except the hoisting point is too far (this distance is computed by
+hoist load except when the hoisting point is too far (this distance is computed by
 looking at the experimental results of representative benchmarks, see

### 2016-11-30 other/libcxx-slides.tex
cgo

@@
 \begin{document}
-
-\def \GVN {GVN}
 
@@
 
-\frame{\frametitle{libcxx-performance improvements}
+\frame{\frametitle{libcxx-performance analysis}
+  \begin{itemize}
+  \item Improving standard library algorithms like string::find, basic_streambuf::xsgetn
+  \item inlining the constructor and destructor of std::string
+  \item systematic analysis of attributes in the standard library
+  \item comparative analysis of clang-libc++ vs. gcc-libstdc++ on representative benchmarks.
+  \item clang+libc++ vs. gcc+libstdc++ performance update. Constructor of vector in libstdc++ contains memcpy
+    in the final assembly.
+  \end{itemize}
+
   \begin{itemize}
@@
 
-  \item clang+libc++ vs. gcc+libstdc++ performance update. Constructor of vector in libstdc++ contains memcpy
-    in the final assembly.
-
 Benchmark	                                Time (ns)	CPU (ns)	Iterations	Time (ns)	CPU (ns)	Iterations	%(Time)
@@
 
-  \item Removing Heap usage
-
-  \item Missing features (and performance issues) in <regex>
-
-  \item Non-conforming allocators in the test suite
-
   \item Identifying missing attributes in functions
 
-  \item Debug mode for the library
   \end{itemize}

### 2016-11-30 other/libcxx-slides.tex
abstract

@@
 \frame{\frametitle{libcxx-performance analysis}
+
+Abstract: We will discuss the improvements and future work on libcxx. This
+includes the improvements on standard library algorithms like string::find and
+basic_streambuf::xsgetn. These algorithms were suboptimal and we got huge
+improvements after optimizing them. Similarly, we enabled the inlining of
+constructor and destructor of std::string. We will present a systematic analysis
+of function attributes in libc++ and the places where we added missing
+attributes. Finally, we will present a comparative analysis of clang-libc++
+vs. gcc-libstdc++ on representative benchmarks
+
+Keywords: libc++, libstdc++, C++, performance analysis, function attributes
+
   \begin{itemize}

### 2016-12-05 hoist.tex
interference

@@
   amount of code to be speculatively executed and can reduce the critical path;
-\item to reduce register pressure: by moving computations closer to the
-  definitions of their operands;
+\item to reduce interference/register pressure: by moving computations closer to
+  the definitions of their operands in cases where hoisting would reduce the
+  interference. For example if all the use operands of an expression of the form
+  a = b op c are last usages (a kill) then hoisting will reduce the interference
+  by two but increase by one.
 \item to improve passes that do not work well with branches:

### 2017-04-14 hoist.tex
simd benefits
Updated abstract

@@
 
-\title{\GVN{}-Hoist: Hoisting Computations from Branches}
+\title{SSA based global code motion of identical computations}
 
@@
 \begin{abstract}
-  Code-hoisting identifies identical computations across the program and hoists
-  them to a common dominator so as to save code size.  Not only code-hoisting
-  saves code size, it exposes redundancies in some cases, it exposes more
-  instruction level parallelism in the basic block a computation is hoisted, and
-  it enables other passes like LICM to remove more redundancies.
+We present a global code motion compiler optimization which schedules
+identical computations across the program so as to save code size.  Not only
+this code motion saves code size, it exposes redundancies in some cases, it
+exposes more instruction level parallelism in the basic block when the
+computations are hoisted, and it enables other passes like loop invariant
+motion to remove more redundancies. The cost model to drive the code motion is
+based on live range analysis on SSA representation such that the (virtual)
+register pressure does not increase.
 
-  We present a code hoisting pass that we implemented in \LLVM{}. It is based on
-  Global Value Numbering infrastructure available in \LLVM{}. The experimental
-  results show an average of 2.5\% savings in code size, although the code size
-  also increases in many cases because it enables more inlining. This is an
-  optimistic algorithm in the sense that it considers all identical computations
-  in a function as potential candidates to be hoisted. We make an extra effort
-  to hoist candidates by partitioning the potential candidates in a way to
-  enable partial hoisting in case common hoisting points for all the candidates
-  cannot be found. We also formalize cases when register pressure will reduce as
-  a result of hoisting and why sorting the list of potential candidates
-  w.r.t. their DFS numbers helps hoist more candidates with less compile time
-  overhead.
+We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
+infrastructure available in \LLVM{}. The experimental results show an average
+of 2.5\% savings in code size in llvm testsuite, although the code size also
+increases in many cases because it enables more inlining. We have also seen
+imrpovements of approximately 5\% in a couple of SPEC 2006 benchmarks viz. gcc
+and mcf, moreover, register spills reduced for almost all the SPEC 2006
+bencmarks when compiled for X86\_64. This is an optimistic algorithm in the
+sense that it considers all identical computations in a function as potential
+candidates. We make an extra effort to hoist candidates by partitioning the
+potential candidates in a way to enable partial hoisting in case common
+hoisting points for all the candidates cannot be found. We also formalize how
+register pressure will reduce as a result of code-motion and why sorting the
+list of potential candidates w.r.t. their depth first numbers helps hoist more
+candidates with less compile time overhead.
 \end{abstract}
@@
   by two but increase by one.
+\item for SIMD architectures which execute all branches, code hoisting will
+  reduce the total number of instructions to be executed.
 \item to improve passes that do not work well with branches:

### 2017-04-14 hoist.tex
abstract

@@
 phase that reduces the number of run-time computations.  Classical scalar
-optimizations like CSE \cite{dragonbook} work very well on single basic blocks and
+optimizations like CSE \cite{dragonbook} work very well on single basic blocks but
 when it comes to detect redundancies across basic blocks these techniques fall
@@
 
-This paper describes code-hoisting, a technique that uses the information
-computed for \PRE{} to detect identical computations and has a transformation
-phase whose goal differs from \PRE{}: it removes identical computations from
-different branches of execution.  These identical computations in different
-branches of execution are not redundant computations at run-time and the number
-of run-time computations is not reduced. Code-hoisting is not a redundancy
+This paper describes code-motion of identical computations, a technique that
+uses the information computed for \PRE{} to detect identical computations but
+has a transformation phase whose goal differs from \PRE{}: it removes identical
+computations from different branches of execution.  These identical computations
+in different branches of execution are not redundant computations at run-time
+and the number of run-time computations is not reduced. It is not a redundancy
 elimination pass, and thus it has different cost function and heuristics than
-\PRE{} or CSE.  Code-hoisting is also different from global code scheduling
-\cite{dragonbook,click1995global} in the sense that it will only hoist
-computations when there are identical computations sharing a common
-dominator. The goals of code hoisting are:
+\PRE{} or CSE. It is more similar to global code scheduling
+\cite{dragonbook,click1995global} in the sense that it will only move
+computations. The goals of code hoisting are:
 
@@
     loop with a single BB, should all the instructions in a conditional get
-    hoisted and sinked;
+    hoisted or sinked;
   \item to enable more loop invariant code motion (LICM): as LICM does not

### 2017-04-14 hoist.tex
goals

@@
 \cite{dragonbook,click1995global} in the sense that it will only move
-computations. The goals of code hoisting are:
+computations. The goals of this code motion are:
 
@@
 \item to help out-of-order processors with speculative execution of branches: by
-  hoisting expressions out of branches, code hoisting can effectively reduce the
-  amount of code to be speculatively executed and can reduce the critical path;
-\item to reduce interference/register pressure: by moving computations closer to
-  the definitions of their operands in cases where hoisting would reduce the
-  interference. For example if all the use operands of an expression of the form
-  a = b op c are last usages (a kill) then hoisting will reduce the interference
-  by two but increase by one.
-\item for SIMD architectures which execute all branches, code hoisting will
-  reduce the total number of instructions to be executed.
+  hoisting/sinking expressions out of branches, code hoisting can effectively
+  reduce the amount of code to be speculatively executed and can reduce the
+  critical path;
+\item to reduce interference/register pressure: For example if all the use
+  operands of an instruction (I) of the form a = b op c are last usages (a kill)
+  then hoisting will reduce the interference by two but increase by one across
+  the path I gets hoisted.
+\item for SIMD architectures, which execute all branches, it will reduce the
+  total number of instructions to be executed.
 \item to improve passes that do not work well with branches:
@@
 \begin{itemize}
-\item a new optimistic algorithm to hoist computations from branches,
+\item a new optimistic algorithm to move computations out of branches,
 \item cost models to reduce live-range and reduce spills,
@@
   the list of potential candidates sorted by their DFS visit number,
-\item performance evaluation of the implementation in \LLVM{},
-\item comparison against other algorithms for code hoisting.
+\item performance evaluation of the implementation in \LLVM{}.
 \end{itemize}

### 2017-04-14 hoist.tex
remove dfsin formalisms

@@
   relations \cite{dragonbook};
-\item Single Entry Single Exit (\SESE{}) \cite{sese} and Single Entry Multiple Exit
-  (SEME) regions;
 \item Static Single Assignment (\SSA{}) \cite{cytron};
@@
 
-\subsection{Finding candidates to hoist}
+\subsection{Finding candidates to move}
 The first step is to find a set of instructions that perform identical
@@
 
-The current implementation of \GVN{} in \LLVM{} has some limitations when it comes to
-loads and stores so we compute the \GVN{} of loads and stores separately.  Our
-short-time solution to value number loads is to hash the address from where the
-value is to be loaded. For stores, we value number the address as well as the
-value to be stored at that address.
+The current implementation of \GVN{} in \LLVM{} has some limitations when it
+comes to loads and stores so we compute the \GVN{} of loads and stores
+separately.  Our solution to value number loads is to hash the address from
+where the value is to be loaded. For stores, we value number the address as well
+as the value to be stored at that address.
 
@@
 candidates closer in the list have their common dominator nearby in cases when
-there are no fully redundant instructions. Essentially if,
+there are no fully redundant instructions. In Figure-\ref{fig:dfsin}, if B3 B4
+B5 and B6 have identical computations and for some reason they cannot be hoisted
+together at B0, then we partition the set of hoistable candidates in their DFSin
+order.
 
-\begin{verbatim}
-// B1 != B2 != B3
-BasicBlock B1, B2, B3
-DFSIn(B1) = depth first discovery time stamp
-DFSIn(B1,B2) = |DFSIn(B1) - DFSIn(B2)|
-DFSIn(B1,B3) = |DFSIn(B1) - DFSIn(B3)|
-
-Depth(B1) = depth of tree from the root node.
-// When B1 dominates B2
-BBDist(B1, B2) = |Depth(B1) - Depth(B2)|
-
-NCD(B1, B2) = nearest common dominator of B1 and B2
-BBDist(B1,B2) = BBDist(B1, NCD(B1, B2))
-BBDist(B1,B3) = BBDist(B1, NCD(B1, B3))
-
-Then,
-DFSIn(B1,B2) < DFSIn(B1,B3) implies
- BBDist(B1,B2) <= BBDist(B1,B3)
-\end{verbatim}
-
-Same property would hold for instructions I1, I2, I3 provided they are not fully
-redundant, i.e., none of the basic blocks containing I1, I2 and I3 respectively
-dominate the other:
-
-\begin{verbatim}
-Instruction I1, I2, I3
-// I1, I2, I3 are not fully redundant
-// B1, B2, B3 contains I1, I2, I3 respectively
-DFSIn(I1) = DFSIn(B1)
-BBDist(I1, I2) = BBDist(B1, B2)
-DFSIn(I1) = DFSIn(B1)
-DFSIn(I1,I3) = DFSIn(B1, B3)
-NCD(I1, I2) = NCD(B1, B2)
-
-Then,
- DFSIn(I1,I2) < DFSIn(I1,I3) implies
-  BBDist(I1,I2) <= BBDist(I1,I3)
-\end{verbatim}
-
-It says that if I1 is closer to I2 than I3 in the list of candidates sorted by
-DFSIn numbers, then the nearest common dominator of I1 and I2 will be closer (in
-terms of basic block distance) to I1, than nearest common dominator of I1 and
-I3.
-
-In the presence of fully redundant computations in the list of hoistable
-candidates this equation may not hold. For example, if there is another
-instruction I4 which is dominated by I1, but I4 is present in a basic block
-which is farther than $DFSIn(I1, I2)$, i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
-however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$.
+\begin{figure}
+\centering
+%rankdir=LR;
+\digraph[scale=0.35]{abc}{
+  B0 -> B1
+  B0 -> B2
+  B1 -> B3
+  B1 -> B4
+  B2 -> B5
+  B2 -> B6
+}
+\caption{CFG to illustrate partitioning}
+\label{fig:dfsin}
+\end{figure}
 

### 2017-04-14 hoist.tex
remove verbose dfsin arguments

@@
 search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
-candidates closer in the list have their common dominator nearby in cases when
-there are no fully redundant instructions. In Figure-\ref{fig:dfsin}, if B3 B4
-B5 and B6 have identical computations and for some reason they cannot be hoisted
-together at B0, then we partition the set of hoistable candidates in their DFSin
-order.
+candidates closer in the list have their common dominator nearby. In
+Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and for
+some reason they cannot be hoisted together at B0, then we partition the set of
+hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
+be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
+B1 and those in B5, B6 to be hoisted at B2.
 
@@
 
-So by sorting candidates w.r.t. their DFSIn numbers:
-\begin{enumerate}
-\item would make fewer checks for legality and profitability.
-\item hoisting point will be closer, so the intersection of live-range of the
-  instruction with other instructions will be minimal.
-\end{enumerate}
-
 In our current implementation we keep as many candidates in one set as possible
@@
 to hoist subset of candidates which are legal to hoist and then start finding
-new hoisting point for the remaining ones. So there are two limitations of the
-current implementation of the partition algorithm: first, sorting by DFSIn
-numbers does not give the desired order when there are fully redundant
-instructions; second, it lacks cost model for partitioning in order to
-preserve/improve performance, see Section~\ref{sec:future-work}.
+new hoisting point for the remaining ones.
 

### 2017-04-14 hoist.tex
remove partial hoist

@@
   hoisting/sinking expressions out of branches, code hoisting can effectively
-  reduce the amount of code to be speculatively executed and can reduce the
+  reduce the amount of code to be speculatively executed and hence reduce the
   critical path;
@@
 \GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
-part of \GVN{}-\PRE{}: it uses the set of ANTIC and AVAIL value expressions computed
-for \PRE{}.  The algorithm hoists top down in a predecessor where the value is in
-ANTIC.  It uses ANTIC[B] to know what expressions will be computed on every
-path from B to exit, and can be computed in B: that is the safety condition.  It
-uses AVAIL[B] to subtract out those values already being computed (because they
-are already available): this is a redundancy condition.  The cost function is:
-for each hoist candidate, if all successors of B are dominated by B, then we
-know insertion into B will eliminate all the remaining computations.  It then
-checks to see if at least one successor of B has the value available.  This
-avoids hoisting it way up the chain to ANTIC.  It also checks to ensure that B
-has multiple successors, since hoisting in a straight line is pointless.  The
-algorithm continues on down the dominator tree, iterating with \PRE{} until no more
-changes.  One advantage of \GCC{} implementation is that it works in sync with the
-\GVN{}-\PRE{} such that when new hoisting opportunities are created by \GVN{}-\PRE{},
-code-hoisting will hoist them.
+part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
+expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
+basic block B i.e., values computed on all paths from B to exit and
+AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
+top down to a predecessor.  It uses to know what expressions will be computed on
+every path from B to exit, and can be computed in B: that is the safety
+condition.  It uses AVAIL\_OUT[B] to subtract out those values already being
+computed.  The cost function is: for each hoist candidate, if all successors of
+B are dominated by B, then we know insertion into B will eliminate all the
+remaining computations.  It then checks to see if at least one successor of B
+has the value available.  This avoids hoisting it way up the chain to ANTIC.  It
+also checks to ensure that B has multiple successors, since hoisting in a
+straight line is pointless.  The algorithm continues top down the dominator
+tree, iterating with \PRE{} until no more changes.  One advantage of \GCC{}
+implementation is that it works in sync with the \GVN{}-\PRE{} such that when
+new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
+hoist them.
 
@@
 
-When aggressively for code-size (at optimization level -Oz), we allow hoisting
-of scalars even when the expression is not very busy. This creates unnecessary
-computation on paths where the expression is not needed so the program may run slow
-but reduces the code size. TODO: Experimental results.
-
 \subsubsection{Profitability of hoisting loads}
@@
 Currently we hoist all the calls that are suitable candidates for hoisting.
+
+Since stores and calls are hoisted the least \ref{table:experimental-results}
+the performance does not change much whether they are hoisted or not.
 

### 2017-04-14 hoist.tex
cleanup from the write

@@
 
-The code-hoisting pass can be divided into the following steps that we will
+The code-motion pass can be broadly divided into the following steps that we will
 describe in the rest of this section:
 \begin{itemize}
-\item find candidates: instructions computing same values,
+\item find candidates suitable for code-motion
 \item compute a point in the program where it is both legal and profitable to
   move the code,
-\item transform the code.
+\item transform the code and update data structures to continue iterative code
+  motion.
 \end{itemize}
@@
 the new copy are removed, and the \SSA{} form is restored by updating the
-intermediate representation (IR) to reflect the changes.
+intermediate representation (IR) to reflect the changes. At the same time
+MemorySSA is also updated to get up-to-date information about memory references.
 
@@
 
-Finally after the transform is done, we verify a set of post-conditions to
-establish that program invariants are maintained: e.g., consistency of
-use-defs, and \SSA{} semantics.
+Finally after the transformation is done, we verify a set of post-conditions to
+establish that program invariants are maintained: e.g., consistency of use-defs,
+and \SSA{} semantics.
 

### 2017-04-14 hoist.tex
adding Barany and Shobaki

@@
 
+Barany \cite{barany2013} presented a global scheduler with ILP formulation with
+a goal to minimize register pressure. The results they got were not very
+promising. It may be because they only used the scheduler for smaller functions
+(<1000 instructions) also they compiled the benchmarks for ARM-Cortex which is
+more resilient to register pressure because it has more registers compared to
+say X86.
+
+Shobaki \cite{shobaki2013} also recently presented a combinatorial global
+scheduler with reasonable performance improvements. It is possible that both
+Barany and Shobaki's implmentation will have similar results when compiled for
+same target. Also, both suffer the same problem, although Shobaki not so much,
+of large compile times which is not feasible for industrial compilers like gcc
+and LLVM.
+
+We got reduction in register spills on all the SPEC2006 benchmarks and
+reasonable performance improvements with very low compile time overhead.
+
+% TODO: References
+% Sink instructions
+% Ranking of instructions from Zadeck
+% Liveness analysis from Ramakrishna
+
 \newpage
 
-\section{Code hoisting}
+\section{Code motion}
 
-The algorithm for code hoisting uses several common representations of the
+The algorithm for code motion uses several common representations of the
 program that we shortly describe below:

### 2017-04-14 hoist.tex
thank

@@
 % Liveness analysis from Ramakrishna
+% Experimental results: Register spills, compile time, run time.
 
@@
 We would like to thank Daniel Berlin for his code reviews and for his feedback
-on earlier versions of this paper.
+on earlier versions of this paper and Brian Grayson for motivating examples that
+started this work.
 

### 2017-04-14 hoist.tex
register pressure model

@@
 parameters that affect runtime performance e.g., impact on live-range, gain in
-the code size.  We have established a set of cost models described in Section~\ref{sec:cost-models}
-for each parameter and tuned them for performance against representative
-benchmarks.
+the code size.  We have established a set of cost models described in
+Section~\ref{sec:cost-models} for each parameter and tuned them for performance
+against representative benchmarks.
 
@@
 \label{hoist:reg-pressure}
-Following example explains how code hoisting can actually reduce the register
-pressure.  Consider the following example where the labels prefixed with 'P'
-represent the position of instruction in a basic block (names prefixed with 'B').
+Hoisting upwards will decrease the live-range of its use, if it is a last use (a
+kill), but increase the live-range of its definition. Conversely, sinking will
+decrease the live-range of the defined register but increase the live-range for
+killed operands. If the live-range after code-motion is less than before it will
+be moved.
+
+Essentially, as long as there is one killed operand , code hoisting will either
+decrease or preserve the register pressure. Similarly, code-sinking will either
+decrease or preserve the register pressure as long as there is one operand
+killed at most.
+
+Following example explains how code motion of identical computations can reduce
+the register pressure.  Consider the following example where the labels prefixed
+with 'P' represent the position of instruction in a basic block (names prefixed
+with 'B').
 
 \begin{verbatim}
-B0: P0: b = 1
+B0: b = m
     goto B1
 
-B1: P1: c = 2
+B1: c = n
     goto B2
 
-B2: P2: if c is true then goto B3 else goto B4
+B2: if c is true then goto B3 else goto B4
 
-B3: P3: a0 = b + c
-    goto B5
+B3: a0 = b<kill> + c<kill>
 
-B4: P4: a1 = b + c
-    goto B5
-
-B5: P5: d = phi {a0(B3), a1(B4)}
-
-If we measure D(Px,Py) as total instruction
-count in the path from the position of Px to Py
-
-live-range(a0) = D(P0,P3) + D(P1,P3) + D(P3,P5)
-               = 6 + 4 + 2 = 12
-live-range(a1) = D(P0,P4) + D(P1,P4) + D(P4,P5)
-               = 6 + 4 + 2 = 12
-
-old-live-range = max(live-range(a0),live-range(a1))
-               = 12
+B4: a1 = b<kill> + c<kill>
 
 After hoisting a0 and a1 are removed and a copy
-of a0 as a01 is placed in B2 just before P2.
-
-live-range(a01) = D(P0,P2-1) + D(P1,P2-1) + D(P2 -1,P5)
-                = 4 + 3 + 3 = 10
-new-live-range = live-range(a01)
+of a0 as a01 is placed in B2 just before the branch.
 \end{verbatim}
 
-If the new live-range is less than the old one it will be a good candidate for
-hoisting. The live ranges can remain same as well if there is only one operand
-on the right hand side, e.g., an assignment operation or, one of the operands is
-not a register. That means hoisting upwards will decrease the live-range of its
-use but increase the live-range of its definition.
-
-In a special case where the instruction to be hoisted has the last use of its
-operands then the code hoisting will always reduce the register pressure if it
-has two register operands because the gain in live-range will be in the ratio of
-2:1. Based on the above formulae we can also deduce that, as long as there is
-one register operand in the right hand side with its last use, code hoisting
-will either decrease or preserve the register pressure.
+In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
+reduce the register pressure in B3 and B4 because two operands will be freed.
 

### 2017-04-14 hoist.tex
remove pessimistic

@@
 AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
-top down to a predecessor.  It uses to know what expressions will be computed on
-every path from B to exit, and can be computed in B: that is the safety
-condition.  It uses AVAIL\_OUT[B] to subtract out those values already being
-computed.  The cost function is: for each hoist candidate, if all successors of
-B are dominated by B, then we know insertion into B will eliminate all the
-remaining computations.  It then checks to see if at least one successor of B
-has the value available.  This avoids hoisting it way up the chain to ANTIC.  It
-also checks to ensure that B has multiple successors, since hoisting in a
-straight line is pointless.  The algorithm continues top down the dominator
-tree, iterating with \PRE{} until no more changes.  One advantage of \GCC{}
-implementation is that it works in sync with the \GVN{}-\PRE{} such that when
-new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
-hoist them.
+top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
+be computed on every path from B to exit, and can be computed in B.  It uses
+AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
+function is: for each hoist candidate, if all successors of B are dominated by
+B, then we know insertion into B will eliminate all the remaining computations.
+It then checks to see if at least one successor of B has the value available.
+This avoids hoisting it way up the chain to ANTIC.  It also checks to ensure
+that B has multiple successors, since hoisting in a straight line is pointless.
+The algorithm continues top down the dominator tree, iterating with \PRE{} until
+no more changes.  One advantage of \GCC{} implementation is that it works in
+sync with the \GVN{}-\PRE{} such that when new hoisting opportunities are
+created by \GVN{}-\PRE{}, code-hoisting will hoist them.
 
@@
 
-The process of computing \GVN{} can be on-demand (as we come across an instruction)
-or, precomputed (computing \GVN{} of all the instructions beforehand). Which
-process to choose is determined by the scope of code-hoisting we want to
-perform. In a pessimistic approach, as described in
-Section~\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
-from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
-to compute \GVN{} values on-demand. Whereas, in the optimistic approach, as
-described in Section~\ref{subsec:optimistic}, we want to hoist as many
-instructions as possible, and it would require \GVN{} values to be precomputed.
+The process of computing \GVN{} can be on-demand (as we come across an
+instruction) or, precomputed (computing \GVN{} of all the instructions
+beforehand). Which process to choose is determined by the scope of code-hoisting
+we want to perform. In a pessimistic approach, we want to hoist a limited set of
+instructions from the sibling branches as we iterate the DFS tree bottom-up, it
+is sufficient to compute \GVN{} values on-demand. Whereas, in the optimistic
+approach, as described in Section~\ref{subsec:optimistic}, we want to move as
+many instructions as possible, and it would require \GVN{} values to be
+precomputed.
 
@@
 Since the equality of candidates is purely based on the value they compute, we
-need to establish if hoisting them to a common dominator would be feasible. Once
-a common dominator is found, we check whether all the use-operands of the set of
-instructions are available at that position. In some cases when the operands are not
-available, it is possible to reinstantiate (remateralize) the use-operands, thus
-passing the legality check.
+need to establish if hoisting them to a common dominator or sinking them to a
+common post-dominator would be feasible. Once a common dominator
+(post-dominator) is found, we check whether all the use-operands of the set of
+instructions are available at that position. In some cases when the operands are
+not available, it is possible to reinstantiate (remateralize) the use-operands,
+thus passing the legality check.
 
 Subsequently, it is checked that the side-effects of the computations does not
-intersect with any side-effects between the instructions to be hoisted and their
-hoisting point. For memory operations like loads, stores, calls etc., it is also
-required to check that all the paths from the hoisting point to the end of the
-function should execute the exact instruction, in order to guarantee
-correctness.
+intersect with any side-effects between the instructions to be hoisted/sunk and
+their hoisting/sinking point.
 
-Moreover, hoisting of memory operations is tricky on paths which have indirect
+Moreover, moving memory operations is tricky on paths which have indirect
 branch targets e.g., landing pad, case statements, goto labels etc., because it
-becomes difficult to prove that all the paths from hoisting point to the end of
-the function would execute the instruction. In our current implementation we
-discard hoisting through such paths.
+becomes difficult to prove safety checks. In our current implementation we
+discard candidates on those paths.
 
-In the optimistic approach, described in Section~\ref{subsec:optimistic}, it is
-possible that a common hoisting point of all the instructions is either too far
-away, or not legally possible. In these cases, it is still possible to
-`partially' hoist a subset of instructions by splitting the set of candidates
-and finding a closer hoisting point for each subset. For more details see
+In the approach, described in Section~\ref{subsec:optimistic}, it is possible
+that a common hoisting point of all the instructions is either too far away, or
+not legally possible. In these cases, it is still possible to `partially' hoist
+a subset of instructions by splitting the set of candidates and finding a closer
+hoisting point for each subset. For more details see
 Section~\ref{subsec:partition}.

### 2017-04-14 hoist.tex
merge cost-model and profitability

@@
 
-\subsection{Profitability check}
+\subsection{Profitability check (Cost models)}
 \label{subsec:profitability}
@@
 against representative benchmarks.
+
+\section{Cost models}
+\label{sec:cost-models}
+Similar to any compiler optimization pass, there are several cost functions that
+are deployed to tune for optimal combination of performance and code-size.
+Since this is mostly a code-size optimization pass, the goal is to not regress
+in performance across popular benchmarks at the same time reduce code size as
+much as possible. Following are the cost models which are implemented:
+
+\subsection{Reduce register pressure}
+\label{hoist:reg-pressure}
+Hoisting upwards will decrease the live-range of its use, if it is a last use (a
+kill), but increase the live-range of its definition. Conversely, sinking will
+decrease the live-range of the defined register but increase the live-range for
+killed operands. If the live-range after code-motion is less than before it will
+be moved.
+
+Essentially, as long as there is one killed operand , code hoisting will either
+decrease or preserve the register pressure. Similarly, code-sinking will either
+decrease or preserve the register pressure as long as there is one operand
+killed at most.
+
+Following example explains how code motion of identical computations can reduce
+the register pressure.  Consider the following example where the labels prefixed
+with 'P' represent the position of instruction in a basic block (names prefixed
+with 'B').
+
+\begin{verbatim}
+    b = m
+    c = n
+B0: if c is true then goto B3 else goto B4
+
+B1: a0 = b<kill> + c<kill>
+
+B2: a1 = b<kill> + c<kill>
+
+After hoisting a0 and a1 are removed and a copy
+of a0 as a01 is placed in B0 just before the branch.
+\end{verbatim}
+
+In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
+reduce the register pressure in B3 and B4 because two operands will be freed.
+
+Ideally, it should be okay to hoist all the instructions and a later a
+live-range-splitting \cite{cooper1998live} pass should make the right decision
+of rematerializing the instruction should it be beneficial to do so. But the
+current live-range splitting pass of \LLVM{} is not making the optimial decision
+and we have found regressions while hoisting aggressively.
+
+Moreover, LLVM has a `getelementptr' instruction which computes the address
+where a load or a store would happen. It is a scalar computation and gets
+hoisted frequently even if the loads/stores would not get hoisted.  In order to
+reduce register pressure while hoisting loads, we have restricted hoisting of
+address computations away from their corresponding loads and stores when the
+loads and stores cannot be moved.  This restriction is only to mitigate the
+limitations of LLVM's register allocator and may be lifted in the future, when
+the register allocation rematerialization pass has been improved to catch the
+regressions.
+
+\subsection{In the presence of calls}
+\label{cost:across-calls}
+Hoisting scalars across calls is tricky because it can increase the number of
+spills. During the frame lowering of calls, the argument registers, in general,
+the caller saved registers are saved because they might be modified by the
+callee and after the call they are restored. So before the
+call, the register pressure is high because the number of available registers
+are reduced by the number of caller saved registers. In that situation if a
+computation is hoisted across the call, that would increase the total number of
+registers required by one, thus contributing to the register pressure. However,
+in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
+okay to hoist because the register pressure would not decrease.
+
+Hoisting loads/stores across calls also require precise analysis of all the
+memory addresses accessed by the call. Our implementation being an
+intraprocedural pass, the analysis is very conservative. In the presence of pure
+calls, loads can be hoisted but stores can't. Also, if the call throws
+exceptions, or if it may not return, memory references cannot be hoisted.
 
@@
 
-\subsection{Hoisting bottom up with pessimistic approach}
-\label{subsec:pessimistic}
-In the pessimistic approach, the basic blocks are traversed in the inverse
-depth-first order, computing the \GVN{} of instructions as they come by. The \GVN{} of
-sibling branches are compared for equality. Once such a candidate is found, it
-is hoisted in the common dominator. All the leaf nodes are visited before the
-non-leaf nodes (bottom-up) because instructions are hoisted upwards.
-
-The pessimistic algorithm is fast, results in fewer spills and also hoists fewer
-instructions. We have implemented optimistic approach because it is more general
-and can be tuned down to closely mimic pessimistic approach by changing a flag.
+\section{Implementation Details}
+\label{sec:implementation-details}
 
@@
 expressions.
-
-\section{Cost models}
-\label{sec:cost-models}
-Similar to any compiler optimization pass, there are several cost functions that
-are deployed to tune for optimal combination of performance and code-size.
-Since this is mostly a code-size optimization pass, the goal is to not regress
-in performance across popular benchmarks at the same time reduce code size as
-much as possible. Following are the cost models which are implemented:
-
-\subsection{Reduce register pressure}
-\label{hoist:reg-pressure}
-Hoisting upwards will decrease the live-range of its use, if it is a last use (a
-kill), but increase the live-range of its definition. Conversely, sinking will
-decrease the live-range of the defined register but increase the live-range for
-killed operands. If the live-range after code-motion is less than before it will
-be moved.
-
-Essentially, as long as there is one killed operand , code hoisting will either
-decrease or preserve the register pressure. Similarly, code-sinking will either
-decrease or preserve the register pressure as long as there is one operand
-killed at most.
-
-Following example explains how code motion of identical computations can reduce
-the register pressure.  Consider the following example where the labels prefixed
-with 'P' represent the position of instruction in a basic block (names prefixed
-with 'B').
-
-\begin{verbatim}
-B0: b = m
-    goto B1
-
-B1: c = n
-    goto B2
-
-B2: if c is true then goto B3 else goto B4
-
-B3: a0 = b<kill> + c<kill>
-
-B4: a1 = b<kill> + c<kill>
-
-After hoisting a0 and a1 are removed and a copy
-of a0 as a01 is placed in B2 just before the branch.
-\end{verbatim}
-
-In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
-reduce the register pressure in B3 and B4 because two operands will be freed.
-
-\subsection{In the presence of calls}
-\label{cost:across-calls}
-Hoisting scalars across calls is tricky because it can increase the number of
-spills. During the frame lowering of calls, the argument registers, in general,
-the caller saved registers are saved because they might be modified by the
-callee and after the call they are restored. So before the
-call, the register pressure is high because the number of available registers
-are reduced by the number of caller saved registers. In that situation if a
-computation is hoisted across the call, that would increase the total number of
-registers required by one, thus contributing to the register pressure. However,
-in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
-okay to hoist because the register pressure would not decrease.
-
-Hoisting loads/stores across calls also require precise analysis of all the
-memory addresses accessed by the call. Our implementation being an
-intraprocedural pass, the analysis is very conservative. In the presence of pure
-calls, loads can be hoisted but stores can't. Also, if the call throws
-exceptions, or if it may not return, memory references cannot be hoisted.
-
-\subsection{Hoisting too far away}
-If there are several instructions in between the hoisting point and the
-instruction to be hoisted, the instruction to be hoisted crosses several
-instructions while hoisting, that means we are adding one register to all the
-live-ranges spanning the instructions. That could result in spills. In the
-current implementation we choose to hoist if the number of instructions crossed
-is below a threshold. Ideally, it should be okay to hoist all the instructions
-and a later a live-range-splitting \cite{cooper1998live} pass should make the
-right decision of rematerializing the instruction should it be beneficial to do
-so. But the current live-range splitting pass of \LLVM{} is not making the
-optimial decision and we have found spills if the threshold is exceeded. The
-threshold was computed as a result of tuning the \LLVMTestSuite{}, SPEC Cpu2000,
-and Cpu2006 benchmarks \cite{Henning2000}.
-
-Also, hoisting a load increases the register pressure by one across all the
-instructions which the load would cross. That could result in spills later in
-the register allocation.
-
-However, in the special case discussed in Section~\ref{hoist:reg-pressure}, it
-will be okay to hoist because the register pressure would not increase.
-
-In order to reduce register pressure, we have restricted hoisting of address
-computations away from their corresponding loads and stores when the loads and
-stores cannot be moved.  This restriction is lifted when aggressively optimizing
-for size (in -Oz).  We have seen several uplifts and degradations in performance
-of the generated code on several benchmarks, and to avoid the degradations we
-decided to only enable hoisting of getElementPtrs without their associated
-memory operations when optimizing for size.  This restriction may be lifted in
-the future, either when the register allocation rematerialization pass has been
-improved to catch the regressions, or by implementing more precise heuristics
-for code hoisting.
 

### 2017-04-14 hoist.tex
ranking and sinkability

@@
 
-\section{Code hoisting policies}
+\section{SSA based global code motion of identical computations}
+\label{sec:implementation-details}
+% Ranking of instructions from Zadeck
+% Liveness analysis from Ramakrishna
+
 The amount of hoisting depends on whether we collect \GVN{} of instructions
 before finding candidates (optimistic) or, on-demand (pessimistic). It also
-depends on the generality of the \GVN{} algorithm, however, that analysis is beyond
-the scope of this paper.
+depends on the generality of the \GVN{} algorithm. We have implemented a
+optimistic global code motion of identical computation which uses the liveness
+analysis as illustrated in Das \cite{das2012} and ranking expressions explained
+in Rosen \cite{rosen1988global}.
 
-\subsection{Hoisting bottom up with optimistic approach}
+\subsection{Optimistic code motion algorithm}
 \label{subsec:optimistic}
-In this approach, the goal is to maximize the total number of hoistings in the
-entire function.  This algorithm is very useful when optimizing for code-size.
 We collect the \GVN{} of all the instructions in the function and iterate on the
-list of instructions having identical \GVN{}s. After that we find the common
-dominator dominating all such identical computations and perform legality
-checks, as described in Section~\ref{subsec:legality}. Often times it is not
-possible to hoist all the instructions to one common dominator, due to legality
-constraints, e.g., intersecting side-effects, or profitability constraints, e.g.,
-hoisting point too far. In those cases, this algorithmn would partition the list
-of identical instructions into subsets which can be partially hoisted to their
-respective common dominators. The partition algorithm is described as follows:
+list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
+sinking. So first we find the common dominator dominating all such identical
+computations and perform legality checks, as described in
+Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
+instructions to one common dominator, due to legality constraints, e.g.,
+intersecting side-effects, or profitability constraints, e.g., increasing
+register pressure. In those cases, this algorithmn would partition the list of
+identical instructions into subsets which can be partially hoisted to their
+respective common dominators. However, it should be noted that we only hoist
+very busy expressions to avoid checking for undefined behaviors resulting because
+of introducing extra computation in an execution path.
+
+Code hoisting opens new opportunities for other hoistable candidates which were
+of higher rank (depended on candidates which got hoisted). Ideally we could
+iterate on lower ranking expressions first and then proceed to higher ranking
+expressions in the same iteration but LLVM's GVN infrastucture does not compute
+equivalence classes in a effective way. We found it simpler to just recompute
+the value numbers and start finding hoistable candidates again.
+
+For sinking however, higher ranked expressions would be sunk first. And it would
+be illegal to sink higher ranked identical expressions if they are not
+anticipable in the common post-dominator. For example:
+
+\begin{verbatim}
+B0: i0 = load B
+B1: i1 = load A
+    c1 = i1 + 10
+    d1 = i0 + 20
+    goto B3
+B2: i2 = load A
+    c2 = i2 + 10
+    d2 = i0 + 20
+    goto B3
+
+B3: phi(c1, c2)
+    phi(d1, d2)
+\end{verbatim}
+
+In this example (c1, c2) or (d1, d2) could be sunk first. Since (c1, c2) depend
+on i1 and i2 respectively which are also in their original basic blocks, c1 and
+c2 are not anticipable in B3. So without knowing the sinkability of `i1' and
+`i2' it would be illegal to sink (c1, c2) to B3. On the other hand (d1, d2) can
+safely be sunk because their operands are readily available at the sink point
+i.e., B3. It should also be noted that, just because the expressions are
+identical and operands are available it still requires a unique post-dominator
+PHI to use the exact same values to be legally sinkable.
+
+The partition algorithm is described as follows:
+% TODO: Explain why all the identical computation cannot be hoisted.
 
@@
 
-\section{Implementation Details}
-\label{sec:implementation-details}
-
 \subsection{Time complexity of algorithm}
@@
 \label{sec:experimental-results}
+% Experimental results: Register spills, compile time, run time.
 We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146). All

### 2017-04-15 hoist.tex
add parse-stat script to parse llvm stats

@@
 
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|}
+      \hline
+Number of spills                &base   &code-motion &\%loss \\\hline
+400.perlbench	&2542	&2481	&0.97\\\hline
+401.bzip2	        &718	&707	&0.98\\\hline
+403.gcc	        &5778	&5710	&0.98\\\hline
+429.mcf	        &14	&17	&1.21\\\hline
+433.milc	        &616	&624	&1.01\\\hline
+444.namd	        &3185	&3223	&1.01\\\hline
+445.gobmk	        &2166	&2170	&1.00\\\hline
+447.dealII	        &10038	&9151	&0.91\\\hline
+450.soplex	        &1104	&1114	&1.00\\\hline
+453.povray	        &5199	&5176	&0.99\\\hline
+456.hmmer	        &1195	&1195	&1   \\\hline
+458.sjeng	        &168	&169	&1.00\\\hline
+462.libquantum	&183	&183	&1   \\\hline
+464.h264ref	&3382	&3423	&1.01\\\hline
+470.lbm	        &30	&30	&1   \\\hline
+471.omnetpp	&521	&522	&1.00\\\hline
+473.astar	        &176	&196	&1.11\\\hline
+482.sphinx3	&603	&594	&0.98\\\hline
+483.xalancbmk	&5101	&4941	&0.96\\\hline
+Grand Total	                &42719	&41626	&0.97\\\hline
+    \end{tabular}
+  \end{center}
+  \caption{Number of spills on SPEC2006}
+  \label{tab:code-size}
+\end{table}
+
 While benchmarking \LLVMTestSuite{} we see both increase as well as decrease in

### 2017-04-15 other/parse-stat.py

@@ -1,0 +1,40 @@
+import glob
+
+d = {}
+def parse(files):
+  for f in files:
+    with open(f, 'r') as fs:
+        for l in fs:
+          #print l
+          p = l.split(":")
+          n = p[0]
+          val = p[1].strip()
+          valname = val.split("-")
+          #print n, val
+          #print valname[0].strip().split()[0]
+          key = n + '_' + '_'.join(valname[1].strip().split())
+          #print key
+          d[key] = 0 # to initialize each key to zero
+          #return
+  for f in files:
+    with open(f, 'r') as fs:
+        for l in fs:
+          p = l.split(":")
+          n = p[0]
+          val = p[1].strip()
+          valname = val.split("-")
+          #print n, val
+          ins_val = valname[0].strip().split()[0]
+          key = n + '_' + '_'.join(valname[1].strip().split())
+          #print n, val[0]
+          d[key] += int(ins_val)
+
+          
+
+# pass the list of insts.csv files to be parsed.
+files = glob.glob("gvn_hoist*")
+parse(files)
+print d
+files = glob.glob("base_reganal*")
+parse(files)
+print d

### 2017-04-15 other/go-reganal-stat.sh
get compile stats with script

@@ -1,0 +1,18 @@
+#!/bin/bash -xe
+files='gvn_hoist_reganal
+base_reganal'
+
+for f in $files; do
+  grep 'regalloc.*spills inserted' `find . -name $f` -a > $f.spills;
+  grep 'inline.*Number of functions inlined' `find . -name $f` -a > $f.inlinecost;
+  grep 'inline.*functions deleted because all callers found' `find . -name $f` -a > $f.f_deleted.inlinecost;
+  grep 'licm.*Number of instructions hoisted out of loop' `find . -name $f` -a > $f.hoist.licm;
+  grep 'licm.*Number of instructions sunk out of loop' `find . -name $f` -a > $f.sunk.licm;
+done
+
+grep 'gvn.hoist.*instructions hoisted' `find . -name gvn_hoist_reganal` -a > gvn_hoist_reganal.hoist.insn
+grep 'gvn.hoist.*instructions removed' `find . -name gvn_hoist_reganal` -a > gvn_hoist_reganal.remov.insn
+grep 'gvn.hoist.*loads hoisted' `find . -name gvn_hoist_reganal` -a > gvn_hoist_reganal.hoist.loads
+grep 'gvn.hoist.*loads removed' `find . -name gvn_hoist_reganal` -a > gvn_hoist_reganal.remov.loads
+grep 'gvn.hoist.*calls hoisted' `find . -name gvn_hoist_reganal` -a > gvn_hoist_reganal.hoist.calls
+grep 'gvn.hoist.*instructions sunk' `find . -name gvn_hoist_reganal` -a > gvn_hoist_reganal.sink.insn

### 2017-04-15 acmart-pacmpl-template.tex
acm sigplan format

--- 2017-04-15 hoist.tex
@@
-\documentclass{sig-alternate}
+%% For double-blind review submission
+\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
+%\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
+%% For single-blind review submission
+%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
+%% For final camera-ready submission
+%\documentclass[acmlarge]{acmart}\settopmatter{}
+
+%% Note: Authors migrating a paper from PACMPL format to traditional
+%% SIGPLAN proceedings format should change 'acmlarge' to
+%% 'sigplan,10pt'.
+
+%% Some recommended packages.
+\usepackage{booktabs}   %% For formal tables:
+                        %% http://ctan.org/pkg/booktabs
+\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+                        %% http://ctan.org/pkg/subcaption
 \usepackage{comment}
@@
 
-\makeatletter
-\preto{\@verbatim}{\topsep=1pt \partopsep=0pt}
-\makeatother
+\makeatletter\if@ACM@journal\makeatother
 
-\pagenumbering{arabic}
+\makeatletter\preto{\@verbatim}{\topsep=1pt \partopsep=0pt}\makeatother
+
+%% Journal information (used by PACMPL format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{1}
+\acmArticle{1}
+\acmYear{2017}
+\acmMonth{1}
+\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\else\makeatother
+%% Conference information (used by SIGPLAN proceedings format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+\acmYear{2017}
+\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\fi
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission
+\setcopyright{none}             %% For review submission
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2017}           %% If different from \acmYear
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
 
 \begin{document}
+
+%% Title information
+\title[SSA GCM]{SSA based global code motion of identical computations}         %% [Short Title] is optional;
+                                        %% when present, will be used in
+                                        %% header instead of Full Title.
+%\titlenote{with title note}             %% \titlenote is optional;
+                                        %% can be repeated if necessary;
+                                        %% contents suppressed with 'anonymous'
+%\subtitle{Subtitle}                     %% \subtitle is optional
+%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
+                                        %% can be repeated if necessary;
+                                        %% contents suppressed with 'anonymous'
+
 \def \GCC {GCC}
@@
 
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
+%% Author information
+%% Contents and number of authors suppressed with 'anonymous'.
+%% Each author should be introduced by \author, followed by
+%% \authornote (optional), \orcid (optional), \affiliation, and
+%% \email.
+%% An author may have multiple affiliations and/or emails; repeat the
+%% appropriate command.
+%% Many elements are not rendered, but should be provided for metadata
+%% extraction tools.
 
-\title{SSA based global code motion of identical computations}
+%% Author with single affiliation.
 
-\toappear{
-   \hrule \vspace{5pt}
-   \LLVM{}-dev 2016
+\author{Aditya Kumar}
+%\authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Senior Compiler Engineer}
+  %\department{Department1}              %% \department is recommended
+  \institution{Samsung Austin R\&D Center}            %% \institution is required
+  %\streetaddress{Street1 Address1}
+  \city{Austin}
+  \state{Texas}
+  %\postcode{Post-Code1}
+  \country{USA}
 }
-\numberofauthors{2}
+\email{aditya.k7@samsung.com}
 
-\author{
-\alignauthor
-Aditya Kumar\\
-       \affaddr{Samsung Austin R\&D Center}\\
-       \email{aditya.k7@samsung.com}
-\alignauthor
-Sebastian Pop\\
-       \affaddr{Samsung Austin R\&D Center}\\
+\author{Sebastian Pop}
+%\authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Senior Staff Compiler Engineer}
+  %\department{Department1}              %% \department is recommended
+  \institution{Samsung Austin R\&D Center}            %% \institution is required
+  %\streetaddress{Street1 Address1}
+  \city{Austin}
+  \state{Texas}
+  %\postcode{Post-Code1}
+  \country{USA}
+}
 \email{s.pop@samsung.com}
-%\and
-%\alignauthor
-%Daniel Berlin\\
-%       \affaddr{Google Inc.}\\
-%       \email{dberlin@dberlin.org}
-}
 
-\maketitle
+%% Paper note
+%% The \thanks command may be used to create a "paper note" ---
+%% similar to a title note or an author note, but not explicitly
+%% associated with a particular element.  It will appear immediately
+%% above the permission/copyright statement.
+\thanks{with paper note}                %% \thanks is optional
+                                        %% can be repeated if necesary
+                                        %% contents suppressed with 'anonymous'
+
+%% Abstract
+%% Note: \begin{abstract}...\end{abstract} environment must come
+%% before \maketitle command
 \begin{abstract}
@@
 
+%% 2012 ACM Computing Classification System (CSS) concepts
+%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011008</concept_id>
+<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003456.10003457.10003521.10003525</concept_id>
+<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Compilers}
+\ccsdesc[500]{Software and its engineering~Source code generation}
+\ccsdesc[500]{General and reference~Performance}
+
+%\ccsdesc[300]{Social and professional topics~History of programming languages}
+%% End of generated code
+
+%% Keywords
+%% comma separated list
+\keywords{Optimizng Compilers, GCC, LLVM, Code Generation, Global Scheduling}  %% \keywords is optional
+
+%% \maketitle
+%% Note: \maketitle command must come after title commands, author
+%% commands, abstract environment, Computing Classification System
+%% environment and commands, and keywords command.
+\maketitle
+
 \section{Introduction}
-
 Compiler techniques to remove redundant computations are composed of an analysis
@@
 \label{sec:experimental-results}
-% Experimental results: Register spills, compile time, run time.
+% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
 We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146). All
@@
 
-\bibliographystyle{abbrv}
-{\small
+%% Acknowledgments
+\begin{acks}                            %% acks environment is optional
+                                        %% contents suppressed with 'anonymous'
+  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
+  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
+  %% acknowledge financial support and will be used by metadata
+  %% extraction tools.
+  This material is based upon work supported by the
+  \grantsponsor{GS100000001}{National Science
+    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  conclusions or recommendations expressed in this material are those
+  of the author and do not necessarily reflect the views of the
+  National Science Foundation.
+\end{acks}
+
+%% Bibliography
 \bibliography{Bibliography}
-}
+
+%% Appendix
+\appendix
+\section{Appendix}
+
+Text of appendix \ldots
+
 \end{document}

### 2017-04-15 acmart-pacmpl-template.tex
dj graph

@@
 
-Rosen et al. explain moving computations successors to remove redundancies
-\cite{rosen1988global}. Their algorithm iterates on computations of same rank
-and move the code with identical computations from the sibling branch. Also
-there is no notion of partially hoisting the computations so their approach may
-result in missing many hoisting opportunities.
+Rosen \cite{rosen1988global} explain moving computations from successors. Their
+algorithm iterates on computations of same rank and move the code with identical
+computations from the sibling branch to a common dominator.
 
@@
 Barany and Shobaki's implmentation will have similar results when compiled for
-same target. Also, both suffer the same problem, although Shobaki not so much,
-of large compile times which is not feasible for industrial compilers like gcc
-and LLVM.
+same target. Also, both suffer from the same problem, although Shobaki not so
+much, of large compile times which is not feasible for industrial compilers like
+gcc and LLVM.
 
-We got reduction in register spills on all the SPEC2006 benchmarks and
-reasonable performance improvements with very low compile time overhead.
+We got reduction in register spills on some SPEC2006 benchmarks, reduction in
+code-size and some performance improvements with very low compile time overhead.
 
@@
   relations \cite{dragonbook};
+\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
+  dominator tree with join-edges to keep track of data-flow in a program. We use
+  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012}.
 \item Static Single Assignment (\SSA{}) \cite{cytron};
@@
 \end{itemize}
+
+\subsection{Liveness using DJ Graph}
+The DJ Graph is very useful for computing liveness because the values will only
+flow either through dominator edges or the join edges (where we insert a PHI).
 

### 2017-04-15 acmart-pacmpl-template.tex
contrib

@@
 \item a new optimistic algorithm to move computations out of branches,
-\item cost models to reduce live-range and reduce spills,
+\item a cost model to reduce interference and reduce spills,
 \item a technique to maximize hoisting in an optimistic approach by partitioning
   the list of potential candidates sorted by their DFS visit number,
-\item performance evaluation of the implementation in \LLVM{}.
+\item experimental evaluation of our implementation in \LLVM{} which combines
+  SSA based liveness analysis, and ranking of expressions to move very busy
+  expressions in order to reduce code-size (and improve performance in some
+  cases).
 \end{itemize}
@@
 
-Rosen \cite{rosen1988global} explain moving computations from successors. Their
-algorithm iterates on computations of same rank and move the code with identical
-computations from the sibling branch to a common dominator.
-
 Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
@@
 dataflow model and an adaptation to a sparse \SSA{} representation is required.
+
+Rosen \cite{rosen1988global} also introduced moving computations from successors
+using value numbering. Their algorithm iterates on computations of same rank and
+move the code with identical computations from the sibling branch to a common
+dominator if they are very busy \cite{steven1997advanced}.
 

### 2017-04-15 acmart-pacmpl-template.tex
cost model

@@
   critical path;
-\item to reduce interference/register pressure: For example if all the use
-  operands of an instruction (I) of the form a = b op c are last usages (a kill)
-  then hoisting will reduce the interference by two but increase by one across
-  the path I gets hoisted.
+\item to reduce interference/register pressure;
 \item for SIMD architectures, which execute all branches, it will reduce the
@@
 promising. It may be because they only used the scheduler for smaller functions
-(<1000 instructions) also they compiled the benchmarks for ARM-Cortex which is
+($<1000$ instructions) also they compiled the benchmarks for ARM-Cortex which is
 more resilient to register pressure because it has more registers compared to
@@
 
-\subsection{Profitability check (Cost models)}
-\label{subsec:profitability}
+\section{Profitability check (Cost models)}
+\label{sec:cost-models}
 After the legality checks have passed, we check for profitability of hoisting.
@@
 parameters that affect runtime performance e.g., impact on live-range, gain in
-the code size.  We have established a set of cost models described in
-Section~\ref{sec:cost-models} for each parameter and tuned them for performance
-against representative benchmarks.
-
-\section{Cost models}
-\label{sec:cost-models}
-Similar to any compiler optimization pass, there are several cost functions that
-are deployed to tune for optimal combination of performance and code-size.
-Since this is mostly a code-size optimization pass, the goal is to not regress
-in performance across popular benchmarks at the same time reduce code size as
-much as possible. Following are the cost models which are implemented:
+the code size. Since this is mostly a code-size optimization pass, the goal is
+to not regress in performance across popular benchmarks at the same time reduce
+code size as much as possible. Following are the cost models which are
+implemented:
 

### 2017-04-15 acmart-pacmpl-template.tex
move example in introduction

@@
 \cite{dragonbook,click1995global} in the sense that it will only move
-computations. The goals of this code motion are:
+computations. We will now consider an example that explains how code hoisting can
+also improve performance.
 
+\subsection{Illustrative Example}
+\label{subsec:example}
+Code hoisting can reduce the critical path length of execution in out of
+order machines. As more instructios are available at the hoisting point, the
+hardware has more instructions to reorder. Following example illustrates how
+hoisting can improve performance by exposing more ILP.
+
+\begin{verbatim}
+float fun(float d, float min, float max, float a) {
+  float tmin, tmax, inv;
+
+  inv = 1.0f / d;
+  if (inv >= 0) {
+    tmin = (min - a) * inv;
+    tmax = (max - a) * inv;
+  } else {
+    tmin = (max - a) * inv;
+    tmax = (min - a) * inv;
+  }
+  return tmax + tmin;
+}
+\end{verbatim}
+
+In this program the computations of tmax and tmin are identical to the
+computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
+depends on inv which depends on a division operation which is generally more
+expensive than the addition, subtraction and multiplication operations. The
+total latency of computation across each branch is:
+$C_{div} + 2(C_{sub} + C_{mul})$
+Or, for out of order processors with two add units and two multiply units:
+$C_{div} + C_{sub} + C_{mul}$
+
+Now if the computation of tmax and tmin are hoisted outside the
+conditionals, the C code version would look like this:
+\begin{verbatim}
+float fun(float d, float min, float max, float a) {
+  float tmin, tmax, tmin1, tmax1, inv;
+
+  tmin1 = (min - a);
+  tmax1 = (max - a);
+
+  inv = 1.0f / d;
+  tmin1 = tmin1 * inv;
+  tmax1 = tmax1 * inv;
+
+  if (inv >= 0) {
+    tmin = tmin1;
+    tmax = tmax1;
+  } else {
+    tmin = tmax1;
+    tmax = tmin1;
+  }
+
+  return tmax + tmin;
+}
+
+\end{verbatim}
+
+In this code the two subtractions and the division operations can be executed in
+parallel because there are no dependencies among them. So the total number of
+cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since $C_{div}$ is
+usually much greater than $C_{sub}$ \cite{x86,aarch64}.
+
+%GCC: 23286 has interesting test cases.
+
+In fact there are several advantages of code-motion of this kind:
 \begin{itemize}
-\item to reduce the code size of the program;
-\item to improve function inlining heuristics: functions become cheaper to
+\item It reduces the code size of the program;
+\item It improves function inlining heuristics: functions become cheaper to
   inline when their code size decreases because inliner heuristics in \LLVM{}
   depends on instruction count.
-\item to expose more instruction level parallelism: by hoisting identical
-  computations to be executed earlier, instruction schedulers can move heavy
-  computations earlier in order to avoid pipeline bubbles;
-\item to help out-of-order processors with speculative execution of branches: by
-  hoisting/sinking expressions out of branches, code hoisting can effectively
-  reduce the amount of code to be speculatively executed and hence reduce the
-  critical path;
-\item to reduce interference/register pressure;
-\item for SIMD architectures, which execute all branches, it will reduce the
+\item It expose more instruction level parallelism to the later compiler
+  passes. By hoisting identical computations to be executed earlier, instruction
+  schedulers can move heavy computations earlier in order to avoid pipeline
+  bubbles;
+\item It reduces branch misprediction penalty on out-of-order processors with
+  speculative execution of branches: by hoisting/sinking expressions out of
+  branches, it can effectively reduce the amount of code to be speculatively
+  executed and hence reduce the critical path;
+\item It reduces interference/register pressure when appropriate cost-model is
+  applied. We have used a cost model in our implementation which hoists/sinks only
+  when register pressure would potentially reduce.
+\item For SIMD architectures, which execute all branches, it will reduce the
   total number of instructions to be executed.
-\item to improve passes that do not work well with branches:
+\item It improves passes that do not work well with branches:
   \begin{itemize}
@@
 There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
-\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
-a more powerful code hoist transform.  The current \LLVM{} implementation of code
-hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
-hoisting from identical basic blocks: the instructions of two sibling basic
-blocks are read in the same time, and all the instructions of the blocks are
-hoisted to the common parent block as long as the compiler is able to prove that
-the instructions are equivalent.  This implementation does not allow for an easy
-extension: first in terms of compilation time overhead the implementation would
-become quadratic in number of instructions to bisimulate, and second the
-equivalence of instructions currently uses $$I1->isIdenticalToWhenDefined(I2)$$
-would need to be rewritten to be more general, leading to using a mechanism
-similar to the idea described in this paper based on \GVN{}.
+\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
+a more powerful code hoist transform.  The current \LLVM{} implementation of
+code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
+basic blocks: the instructions of two sibling basic blocks are read in the same
+time, and all the instructions of the blocks are hoisted to the common parent
+block as long as the compiler is able to prove that the instructions are
+equivalent.  This implementation does not allow for an easy extension: first in
+terms of compilation time overhead the implementation is quadratic in number of
+instructions to bisimulate and second, the equivalence of instructions is
+computed by comparing the operands which is neither general nor scalabale.
 
@@
 code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
-are computed which are hoisted in a basic block where the expression is
+computed which are hoisted in a basic block where the expression is
 anticipable (all the operands are available). This algorithm would hoist as far
@@
 Rosen \cite{rosen1988global} also introduced moving computations from successors
-using value numbering. Their algorithm iterates on computations of same rank and
-move the code with identical computations from the sibling branch to a common
-dominator if they are very busy \cite{steven1997advanced}.
+but using value numbering. Their algorithm iterates on computations of same rank
+and move the code with identical computations from the sibling branch to a
+common dominator if they are very busy \cite{steven1997advanced}. We also
+incorporate ranking of expressions in our implementation.
 
@@
 \item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
-  identify similar computations compilers use \GVN{}.  Each expression is given a
-  unique number and the expressions that the compiler can prove identical are
-  given the same number;
-\item Memory \SSA{} \cite{novillo2007memory}: memory operations that the compiler
-  is able to prove in dependence are linked through use-def chains.
+  identify similar computations compilers use \GVN{}.  Each expression is given
+  a unique number and the expressions that the compiler can prove to be
+  identical are given the same number;
+\item Memory \SSA{} \cite{novillo2007memory}: it is a factored use-def chain of
+  memory operations that the compiler is able to prove are dependent.
 \end{itemize}
@@
 \newpage
-
-\subsection{Illustrative Example} \label{subsec:example}
-Code hoisting can reduce the critical path length of execution in out of
-order machines. As more instructios are available at the hoisting point, the
-hardware has more instructions to reorder. Following example illustrates how
-hoisting can improve performance by exposing more ILP.
-
-\begin{verbatim}
-float fun(float d, float min, float max, float a)
-{
-  float tmin, tmax, inv;
-
-  inv = 1.0f / d;
-  if (inv >= 0) {
-    tmin = (min - a) * inv;
-    tmax = (max - a) * inv;
-  } else {
-    tmin = (max - a) * inv;
-    tmax = (min - a) * inv;
-  }
-  return tmax + tmin;
-}
-\end{verbatim}
-
-In this program the computations of tmax and tmin are identical to the
-computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
-depends on inv which depends on a division operation which is generally more
-expensive than the addition, subtraction and multiplication operations. The
-total latency of computation across each branch is:
-$C_{div} + 2(C_{sub} + C_{mul})$
-Or, for out of order processors with two add units and two multiply units:
-$C_{div} + C_{sub} + C_{mul}$
-
-Now if the computation of tmax and tmin are hoisted outside the
-conditionals, the C code version would look like this:
-\begin{verbatim}
-float fun(float d, float min, float max, float a)
-{
-  float tmin, tmax, tmin1, tmax1, inv;
-
-  tmin1 = (min - a);
-  tmax1 = (max - a);
-
-  inv = 1.0f / d;
-  tmin1 = tmin1 * inv;
-  tmax1 = tmax1 * inv;
-
-  if (inv >= 0) {
-    tmin = tmin1;
-    tmax = tmax1;
-  } else {
-    tmin = tmax1;
-    tmax = tmin1;
-  }
-
-  return tmax + tmin;
-}
-
-\end{verbatim}
-
-In this code the two subtractions and the division operations can be executed in
-parallel because there are no dependencies among them. So the total number of
-cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since $C_{div}$ is
-usually much greater than $C_{sub}$ \cite{x86,aarch64}.
-
-%GCC: 23286 has interesting test cases.
 

### 2017-04-15 acmart-pacmpl-template.tex
contrib and click

@@
 
-The main contributions of this paper are:
+There have been a lot of work both in industry and academia to hoist and sink
+code out of branches, and in general global scheduling
+\cite{click1995global}. Some relate code-hoisting to code-size optimization
+\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
+scheduling to improve performance. Most of the recent work on global scheduling
+are done using ILP which results in prohibitively high compile time. To the best
+of our knowledge we have not found any reference which explored code-motion of
+identical computation in as much detail as we have done. We have also open
+sourced the implementation for easy verification. The main contributions of this
+paper are:
 \begin{itemize}
@@
 
+Click \cite{click1995global} describe aggressive global code motion to first
+schedule all the instructions as early as possible. This results in very long
+live ranges which is mitigated by again scheduling all the instructions as late
+as possible. They report sppeedup of as high as $23\%$.
+
 Barany \cite{barany2013} presented a global scheduler with ILP formulation with
@@
 \subsection{Liveness using DJ Graph}
-The DJ Graph is very useful for computing liveness because the values will only
-flow either through dominator edges or the join edges (where we insert a PHI).
+DJ-Graph \cite{Sreedhar1996} is a data structure that augments the dominator
+tree with join-edges to keep track of data-flow in a program. We use DJ-Graph to
+compute liveness of variables as illustrated in \cite{das2012}.  It is very
+efficient for computing liveness because the values will only flow either
+through dominator edges or the join edges (where we insert a PHI).
 

### 2017-04-15 acmart-pacmpl-template.tex
liveness

@@
 DJ-Graph \cite{Sreedhar1996} is a data structure that augments the dominator
-tree with join-edges to keep track of data-flow in a program. We use DJ-Graph to
-compute liveness of variables as illustrated in \cite{das2012}.  It is very
-efficient for computing liveness because the values will only flow either
-through dominator edges or the join edges (where we insert a PHI).
+tree with join-edges to keep track of data-flow in a program. We use DJ-Graph
+and Merge Sets to compute liveness of variables as illustrated in
+\cite{das2012}.  It is very efficient for computing liveness and does not
+require any bitvectors to be maintained for each basic block. The underlying
+simplicity of liveness computation is due to SSA form, where the the values only
+flow (from def to use) either through dominator edges or the join edges (where
+we insert a PHI). The DJ-graph contains both these edges which allows for
+computation of merge sets for each basic block i.e., a set of all basic blocks
+where the values can flow from a particular basic block.
+
+We compute merge sets of the control flow graph which remains
+same through the code-motion transformation. For code-motion we only want to
+know if a use operand is a kill (to compute changes in register pressure). For
+that we only need to know whether the use is also required later in the
+execution path. A variable is live out of a basic block B if it is used in the
+merge set of B. We compute the live-out relation on-demand when profitability of
+hoistable/sinkable candidates is to be evaluated.
 

### 2017-04-15 acmart-pacmpl-template.tex
isliveout

@@
 \usepackage{needspace}
+\usepackage{algpseudocode}
 
@@
 
-We compute merge sets of the control flow graph which remains
-same through the code-motion transformation. For code-motion we only want to
-know if a use operand is a kill (to compute changes in register pressure). For
-that we only need to know whether the use is also required later in the
-execution path. A variable is live out of a basic block B if it is used in the
-merge set of B. We compute the live-out relation on-demand when profitability of
-hoistable/sinkable candidates is to be evaluated.
+We compute merge sets of the control flow graph which remains same through the
+code-motion transformation. For code-motion we only want to know if a use
+operand is a kill (to compute changes in register pressure). For that we only
+need to know whether the use is also required later in the execution path. A
+variable is live out of a basic block B if it is used in the merge set of B. We
+compute the live-out relation on-demand when profitability of hoistable/sinkable
+candidates is to be evaluated. A simplified version of isLiveOutUsingMergeSet is
+presented here:
+
+\begin{verbatim}
+  Compute if variable a is liveout from basic block n
+  Input: Node N, Variable a
+
+  if a is defined in N then
+    if a is used outside any basic block other than N then
+      return true;
+    else return false;
+  endif
+
+  Ms(n) = null;
+  // Mergeset of N is the union of merge sets of its successors
+  for w in successors(n) do
+    Ms(n) = Ms(n) U Mr(w);
+  endfor
+
+  // Iterate over all the uses of a and see if any intersect with the
+  // merge set of N.
+  for t in users(a) do
+    b = basic_block(t)
+    while (b != null) and (b != def_bb(a)) do
+      if b ∩ Ms(n) then
+        return true;
+      endif
+      b = dom-parent(b);
+    endwhile
+  endfor
+  return false;
+\end{verbatim}
+
+The original algorithm presented in \cite{das2012} has mismatched types in terms
+of uses and nodes because each node can have many uses. Also, while iterating on
+the dominator of each user of a variable we may reach to the beginning of a
+function, in that case the inner while loop needs to terminate. These two cases
+were missing from the algorithm and we came across them during implementation.
 

### 2017-04-15 acmart-pacmpl-template.tex
algos

@@
 computation of merge sets for each basic block i.e., a set of all basic blocks
-where the values can flow from a particular basic block.
+where the values can flow from a particular basic block. We have implemented
+merge-set computation from DJ-graph as illustrated in \cite{das2005}. A simplified
+version of merge set that we implemented is presented here:
+
+\begin{verbatim}
+// Return true if the merge set of source node of
+// a visited J-edge (incoming edge of lnode) is not
+// the subset of the merge set of lnode.
+bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet M)
+   for (all incoming edges to lnode) do
+     Let e = Incoming edge
+     if (e is in JE && e in V) then
+       Let snode = Source Node of e
+       if (M(snode)!(Subset) M(lnode)) then
+         return true
+       end if
+     end if
+   end for
+return false
+\end{verbatim}
+
+\begin{verbatim}
+// Compute merge set top-down in breadth first order.
+bool constructMergeSet_1(BFSList B, JEdges JE, DomLevel DL)
+repeat = False
+// List of visited edges
+Visited V
+for n in B do
+  for e in (all incoming edges to n) do
+    if (e is in JE && e not in V) then
+      V(e) = true
+      Let snode = Source Node of e
+      Let tnode = Target Node of e
+      Let tmp = snode
+      Let lnode = NULL
+      while (DL(tmp) ≥ DL(tnode)) do
+        Merge(tmp) = Merge(tmp) U Merge(tnode) U {tnode}
+        lnode = tmp
+        tmp = dom-parent(tmp)
+      end while
+      repeat = still_inconsistent(lnode, JE, Merge)
+    end if
+  end for
+end while
+return repeat
+
+// Construct Merge set of each node in control-flow-graph G of a function.
+void  constructMergeSet(CFG G) {
+  B = Breadth First Order of G
+  JE = JEdges of G
+  DL = List of Path length (from root) of each node in G.
+
+  do // Call until a fixed point is reached.
+    Repeat = constructMergeSet_1(B, JE, DL);
+  while (Repeat);
+}
+\end{verbatim}
 
@@
 \begin{verbatim}
-  Compute if variable a is liveout from basic block n
+bool isLiveOutUsingMergeSet()
+  // Compute if variable a is liveout from basic block n
   Input: Node N, Variable a
@@
 The original algorithm presented in \cite{das2012} has mismatched types in terms
-of uses and nodes because each node can have many uses. Also, while iterating on
-the dominator of each user of a variable we may reach to the beginning of a
-function, in that case the inner while loop needs to terminate. These two cases
-were missing from the algorithm and we came across them during implementation.
+of uses and nodes because each node can have many instructions and hence many
+uses. Also, while iterating on the dominator of each user of a variable we may
+reach to the beginning of a function, in that case the inner while loop needs to
+terminate. These two cases were missing from the algorithm and we came across
+them during implementation.
 

### 2017-04-16 acmart-pacmpl-template.tex
legality

@@
 of our knowledge we have not found any reference which explored code-motion of
-identical computation in as much detail as we have done. We have also open
-sourced the implementation for easy verification. The main contributions of this
-paper are:
+identical computation in as much detail as we have done. A part of our
+implementation (aggressive code hoisting) is already merged in LLVM trunk,
+however, a general implementation of code-motion of identical expressions is
+still missing from GCC and LLVM trunk. The main contributions of this paper are:
 \begin{itemize}
@@
 computations: this is performed by a linear scan of all instructions of the
-program and classifying all instructions by their value given by \GVN{}.
+program and classifying all instructions by their value given by \GVN{}. We
+could computed available and anticipable sets as computed by GCC's code-hoisting
+but that would be a lot of data structures to maintain at each basic block
+level. For GCC it makes sense because their code-hoisting is integrated with
+GVN-PRE which already has those data structures available.
 
@@
 where the value is to be loaded. For stores, we value number the address as well
-as the value to be stored at that address.
-
-Another limitation of the current \GVN{} implementation in \LLVM{} is that the
-instructions dependent on the loads will not get numbered correctly, and so
-after hoisting all candidates we need to rerun the \GVN{} analysis in order to
-discover new candidates now available after having hoisted load instructions.
-This limitation should be addressed in a new implementation of the \GVN{} based on
-\MemorySSA{}, that would better account for equivalent loads and their dependent
-instructions.
+as the value to be stored at that address. Another limitation of the current
+\GVN{} implementation in \LLVM{} is that the instructions dependent on the loads
+will not get numbered correctly, and so after hoisting all candidates we need to
+rerun the \GVN{} analysis in order to discover new candidates now available
+after having hoisted load instructions.  This limitation should be addressed in
+a new implementation of the \GVN{} based on \MemorySSA{}, that would better
+account for equivalent loads and their dependent instructions.
 
@@
 
-Once the instructions have been classified in equivalence classes, we compute
-for each group of equivalent instructions a point in the program that is both
+Once the instructions have been classified into equivalence classes, we compute
+for each group of equivalent instructions, a point in the program that is both
 legal and profitable for the instructions to be moved to.
@@
 \label{subsec:legality}
-Since the equality of candidates is purely based on the value they compute, we
+Since the equality of candidates is purely based on the value numbers, we also
 need to establish if hoisting them to a common dominator or sinking them to a
-common post-dominator would be feasible. Once a common dominator
-(post-dominator) is found, we check whether all the use-operands of the set of
-instructions are available at that position. In some cases when the operands are
-not available, it is possible to reinstantiate (remateralize) the use-operands,
-thus passing the legality check.
-
-Subsequently, it is checked that the side-effects of the computations does not
-intersect with any side-effects between the instructions to be hoisted/sunk and
-their hoisting/sinking point.
-
-Moreover, moving memory operations is tricky on paths which have indirect
-branch targets e.g., landing pad, case statements, goto labels etc., because it
-becomes difficult to prove safety checks. In our current implementation we
-discard candidates on those paths.
-
-In the approach, described in Section~\ref{subsec:optimistic}, it is possible
-that a common hoisting point of all the instructions is either too far away, or
-not legally possible. In these cases, it is still possible to `partially' hoist
-a subset of instructions by splitting the set of candidates and finding a closer
-hoisting point for each subset. For more details see
-Section~\ref{subsec:partition}.
+common post-dominator would be legal. Once a common dominator (post-dominator)
+is found, we check whether all the use-operands of the set of instructions are
+available at that position. It is possible reinstantiate (remateralize) the
+use-operands in some cases when the operands are not available and make it legal
+to move the instruction.
+
+Subsequently, it is checked that the side-effects of the computations (if any)
+do not intersect with any side-effects between the instructions to be
+hoisted/sunk and their hoisting/sinking point. It is also necessary to check if
+there are indirect branch targets e.g., landing pad, case statements, goto
+labels etc., along the path because it becomes difficult to prove safety checks
+in those cases. In our current implementation we discard candidates on those
+paths.
+
+Code-motion basically consists of two parts i.e., hoisting and sinking. The
+unique part about this algorithm is that it only moves identical
+computations. An immediate guarantee of this approach is gain in code size (the
+final executable may be of larger size because of more inlining). The algorithm
+prefers hoisting to sinking. If the dependency of a hoistable candidate is in
+the same basic block as the candidate, then the dependency must also be
+hoistable otherwise hoisting will be illegal or would require a complicated code
+generation to make it legal. The current algorithm discards cases if the
+dependency is neither hoistable nor rematerializable.
 
@@
 aliasing memory references. As long as all the operands are available (or can be
-made available by rematerialization), the scalar computation can be hoisted.
+made available by rematerialization), the scalar computations can be hoisted.
 
@@
 writes to memory that may alias with the load, in which case we discard the
-candidate.
+candidate. We use the MemorySSA infrastructure of \LLVM{} to iterate on the
+use-def chains available for memory references.
 
@@
 For stores, we check the dependency requirements similar to the hoisting of
-loads. We check that the operands of the store instruction are available at the
-hoisting point, that there are no aliasing loads or store along the path from
-the current position to the hoisting point.
+loads using the MemorySSA of \LLVM{}. We check that the operands of the store
+instruction are available at the hoisting point, that there are no aliasing
+loads or store along the path from the current position to the hoisting point.
 
@@
 without further information, calls have to be classified as writing to memory,
-that is the most restrictive form.  Each category of call instructions will be
+that is the most restrictive form.  Each category of call instructions is
 handled as described for scalar, load, and store instructions.
 
-\section{Profitability check (Cost models)}
-\label{sec:cost-models}
-After the legality checks have passed, we check for profitability of hoisting.
-That takes into account the impact code-hoisting would have on various
-parameters that affect runtime performance e.g., impact on live-range, gain in
-the code size. Since this is mostly a code-size optimization pass, the goal is
-to not regress in performance across popular benchmarks at the same time reduce
-code size as much as possible. Following are the cost models which are
-implemented:
-
-\subsection{Reduce register pressure}
-\label{hoist:reg-pressure}
-Hoisting upwards will decrease the live-range of its use, if it is a last use (a
-kill), but increase the live-range of its definition. Conversely, sinking will
-decrease the live-range of the defined register but increase the live-range for
-killed operands. If the live-range after code-motion is less than before it will
-be moved.
-
-Essentially, as long as there is one killed operand , code hoisting will either
-decrease or preserve the register pressure. Similarly, code-sinking will either
-decrease or preserve the register pressure as long as there is one operand
-killed at most.
-
-Following example explains how code motion of identical computations can reduce
-the register pressure.  Consider the following example where the labels prefixed
-with 'P' represent the position of instruction in a basic block (names prefixed
-with 'B').
-
-\begin{verbatim}
-    b = m
-    c = n
-B0: if c is true then goto B3 else goto B4
-
-B1: a0 = b<kill> + c<kill>
-
-B2: a1 = b<kill> + c<kill>
-
-After hoisting a0 and a1 are removed and a copy
-of a0 as a01 is placed in B0 just before the branch.
-\end{verbatim}
-
-In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
-reduce the register pressure in B3 and B4 because two operands will be freed.
-
-Ideally, it should be okay to hoist all the instructions and a later a
-live-range-splitting \cite{cooper1998live} pass should make the right decision
-of rematerializing the instruction should it be beneficial to do so. But the
-current live-range splitting pass of \LLVM{} is not making the optimial decision
-and we have found regressions while hoisting aggressively.
-
-Moreover, LLVM has a `getelementptr' instruction which computes the address
-where a load or a store would happen. It is a scalar computation and gets
-hoisted frequently even if the loads/stores would not get hoisted.  In order to
-reduce register pressure while hoisting loads, we have restricted hoisting of
-address computations away from their corresponding loads and stores when the
-loads and stores cannot be moved.  This restriction is only to mitigate the
-limitations of LLVM's register allocator and may be lifted in the future, when
-the register allocation rematerialization pass has been improved to catch the
-regressions.
-
-\subsection{In the presence of calls}
-\label{cost:across-calls}
-Hoisting scalars across calls is tricky because it can increase the number of
-spills. During the frame lowering of calls, the argument registers, in general,
-the caller saved registers are saved because they might be modified by the
-callee and after the call they are restored. So before the
-call, the register pressure is high because the number of available registers
-are reduced by the number of caller saved registers. In that situation if a
-computation is hoisted across the call, that would increase the total number of
-registers required by one, thus contributing to the register pressure. However,
-in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
-okay to hoist because the register pressure would not decrease.
-
-Hoisting loads/stores across calls also require precise analysis of all the
-memory addresses accessed by the call. Our implementation being an
-intraprocedural pass, the analysis is very conservative. In the presence of pure
-calls, loads can be hoisted but stores can't. Also, if the call throws
-exceptions, or if it may not return, memory references cannot be hoisted.
-
-\subsubsection{Profitability of hoisting scalars}
-Since scalars are the majority of instructions which are hoisted, we pay special
-attention in case of hoisting scalars too far, as that may increase register
-pressure and result in spills. For example hoisting a scalar past a call, as
-described in Section~\ref{cost:across-calls}.  In our current implementation we
-hoist scalars past a call only when optimizing for code-side (-Os). Ideally, a
-later stage of live-range splitting pass should split the live-ranges for
-optimal performance, however, that is not the case with \LLVM{} as we have found
-regressions when scalars are hoisted too far, as in
-Section~\ref{sec:cost-models}. Another way to mitigate this problem is be to
-reinstantiate (rematerialize) the computation after a call (may be as a
-different optimization pass).
-
-\subsubsection{Profitability of hoisting loads}
-A load instruction introduces a register where the value loaded will be kept,
-the register pressure increases by one (unless the operand to load becomes dead
-at the load). On the other hand, loading a value early will reduce the stall
-during execution should the value is not in the cache. We generally prefer to
-hoist load except when the hoisting point is too far (this distance is computed by
-looking at the experimental results of representative benchmarks, see
-Section~\ref{sec:experimental-results}).
-
-\subsubsection{Profitability of hoisting stores}
-Since stores do not increase the live-range of any registers, and in some cases
-it ends the liveness of registers, we hoist all the stores.
-
-\subsubsection{Profitability of hoisting calls}
-Currently we hoist all the calls that are suitable candidates for hoisting.
-
-Since stores and calls are hoisted the least \ref{table:experimental-results}
-the performance does not change much whether they are hoisted or not.
-
-\subsection{Code generation}
-Once all the legality and profitability checks are satisfied for a set of
-identical instructions, they are suitable candidates for hoisting. A copy of the
-computation is inserted at the hoisting point along with any instructions which
-needed to be rematerialized. Thereafter, all the computations made redundant by
-the new copy are removed, and the \SSA{} form is restored by updating the
-intermediate representation (IR) to reflect the changes. At the same time
-MemorySSA is also updated to get up-to-date information about memory references.
-
-After one iteration of algorithm runs through the entire function, it creates
-more oppportunities for \emph{higher ranked} computations
-\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
-pass, and so we rerun the code-hoisting algorithm until there are no more
-instructions left to be hoisted.  Obviously, this is not the most optimal
-approach and can be improved by ranking the computations \cite{rosen1988global},
-or by improving the \GVN{} analysis to correctly number loads and dependent
-instructions.
-
-Finally after the transformation is done, we verify a set of post-conditions to
-establish that program invariants are maintained: e.g., consistency of use-defs,
-and \SSA{} semantics.
-
-\newpage
-
-\newpage
-
-\section{SSA based global code motion of identical computations}
-\label{sec:implementation-details}
-% Ranking of instructions from Zadeck
-% Liveness analysis from Ramakrishna
-
-The amount of hoisting depends on whether we collect \GVN{} of instructions
-before finding candidates (optimistic) or, on-demand (pessimistic). It also
-depends on the generality of the \GVN{} algorithm. We have implemented a
-optimistic global code motion of identical computation which uses the liveness
-analysis as illustrated in Das \cite{das2012} and ranking expressions explained
-in Rosen \cite{rosen1988global}.
-
-\subsection{Optimistic code motion algorithm}
-\label{subsec:optimistic}
-We collect the \GVN{} of all the instructions in the function and iterate on the
-list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
-sinking. So first we find the common dominator dominating all such identical
-computations and perform legality checks, as described in
-Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
-instructions to one common dominator, due to legality constraints, e.g.,
-intersecting side-effects, or profitability constraints, e.g., increasing
-register pressure. In those cases, this algorithmn would partition the list of
-identical instructions into subsets which can be partially hoisted to their
-respective common dominators. However, it should be noted that we only hoist
-very busy expressions to avoid checking for undefined behaviors resulting because
-of introducing extra computation in an execution path.
-
-Code hoisting opens new opportunities for other hoistable candidates which were
-of higher rank (depended on candidates which got hoisted). Ideally we could
-iterate on lower ranking expressions first and then proceed to higher ranking
-expressions in the same iteration but LLVM's GVN infrastucture does not compute
-equivalence classes in a effective way. We found it simpler to just recompute
-the value numbers and start finding hoistable candidates again.
-
-For sinking however, higher ranked expressions would be sunk first. And it would
-be illegal to sink higher ranked identical expressions if they are not
-anticipable in the common post-dominator. For example:
+\subsubsection{Legality of sinking expressions}
+One might wonder that if a set of expressions compute the same value then why
+not just hoist them. The problem is, sometimes, hoisting is not upward-safe
+\ref{click1995global} e.g., if the expressions are in a landing pad
+etc. Sometimes sinking may reduce the register pressure e.g., when the use
+operands are not kills.
+
+%TODO: Explain why all the identical computation cannot be hoisted.
+For sinking, higher ranked expressions would be sunk first. And it would be
+illegal to sink higher ranked identical expressions if they are not anticipable
+in the common post-dominator. For example:
 
@@
 i.e., B3. It should also be noted that, just because the expressions are
-identical and operands are available it still requires a unique post-dominator
+identical and operands are available, it still requires a unique post-dominating
 PHI to use the exact same values to be legally sinkable.
 
-The partition algorithm is described as follows:
-% TODO: Explain why all the identical computation cannot be hoisted.
-
-\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
+A general global scheduling algorithm also requires checks for undefinedness
+when introducing a new computation along a path. Since only very busy
+expressions are moved in the current implementation, there is no need to check
+for undefinedness resulting due to movement of instruction. This simplifies the
+implementation.
+
+\subsubsection{Partitioning the list of hoistable candidates to maximize hoisting}
 \label{subsec:partition}
+%The partition algorithm is described as follows:
+In the approach, described in Section~\ref{subsec:optimistic}, it is possible
+that a common hoisting point of all the instructions is either too far away, or
+not legally possible. In these cases, it is still possible to `partially' hoist
+a subset of instructions by splitting the set of candidates and finding a closer
+hoisting point for each subset.
+
 In order to hoist a subset of identical instructions, we partition the list of
@@
 new hoisting point for the remaining ones.
+
+\section{Profitability check (Cost models)}
+\label{sec:cost-models}
+After the legality checks have passed, we check if a code-motion is profitable.
+That takes into account the impact code-motion would have on various parameters
+that might affect runtime performance e.g., impact on live-range, gain in the
+code size. Since this is mostly a code-size optimization pass, the goal is to
+not regress in performance across popular benchmarks at the same time reduce
+code size as much as possible. Following are the cost models which are
+implemented:
+
+\subsection{Reduce register pressure}
+\label{hoist:reg-pressure}
+Hoisting upwards will decrease the live-range of its use, if it is a last use (a
+kill), but increase the live-range of its definition. Conversely, sinking will
+decrease the live-range of the defined register but increase the live-range for
+killed operands. If the live-range after code-motion is less than before it will
+be moved.
+
+Essentially, as long as there is one killed operand , code hoisting will either
+decrease or preserve the register pressure. Similarly, code-sinking will either
+decrease or preserve the register pressure as long as there is one operand
+killed at most.
+
+Following example explains how code motion of identical computations can reduce
+the register pressure.  Consider the following example where the labels prefixed
+with 'P' represent the position of instruction in a basic block (names prefixed
+with 'B').
+
+\begin{verbatim}
+    b = m
+    c = n
+B0: if c is true then goto B3 else goto B4
+
+B1: a0 = b<kill> + c<kill>
+
+B2: a1 = b<kill> + c<kill>
+
+After hoisting a0 and a1 are removed and a copy
+of a0 as a01 is placed in B0 just before the branch.
+\end{verbatim}
+
+In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
+reduce the register pressure in B3 and B4 because two operands will be freed.
+
+Ideally, it should be okay to hoist all the instructions and a later a
+live-range-splitting \cite{cooper1998live} pass should make the right decision
+of rematerializing the instruction should it be beneficial to do so. But the
+current live-range splitting pass of \LLVM{} is not making the optimial decision
+and we have found regressions while hoisting aggressively.
+
+Moreover, LLVM has a `getelementptr' instruction which computes the address
+where a load or a store would happen. It is a scalar computation and gets
+hoisted frequently even if the loads/stores would not get hoisted.  In order to
+reduce register pressure while hoisting loads, we have restricted hoisting of
+address computations away from their corresponding loads and stores when the
+loads and stores cannot be moved.  This restriction is only to mitigate the
+limitations of LLVM's register allocator and may be lifted in the future, when
+the register allocation rematerialization pass has been improved to catch the
+regressions.
+
+\subsection{In the presence of calls}
+\label{cost:across-calls}
+Hoisting scalars across calls is tricky because it can increase the number of
+spills. During the frame lowering of calls, the argument registers, in general,
+the caller saved registers are saved because they might be modified by the
+callee and after the call they are restored. So before the
+call, the register pressure is high because the number of available registers
+are reduced by the number of caller saved registers. In that situation if a
+computation is hoisted across the call, that would increase the total number of
+registers required by one, thus contributing to the register pressure. However,
+in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
+okay to hoist because the register pressure would not decrease.
+
+Hoisting loads/stores across calls also require precise analysis of all the
+memory addresses accessed by the call. Our implementation being an
+intraprocedural pass, the analysis is very conservative. In the presence of pure
+calls, loads can be hoisted but stores can't. Also, if the call throws
+exceptions, or if it may not return, memory references cannot be hoisted.
+
+\subsubsection{Profitability of hoisting scalars}
+Since scalars are the majority of instructions which are hoisted, we pay special
+attention in case of hoisting scalars too far, as that may increase register
+pressure and result in spills. For example hoisting a scalar past a call, as
+described in Section~\ref{cost:across-calls}.  In our current implementation we
+hoist scalars past a call only when optimizing for code-side (-Os). Ideally, a
+later stage of live-range splitting pass should split the live-ranges for
+optimal performance, however, that is not the case with \LLVM{} as we have found
+regressions when scalars are hoisted too far, as in
+Section~\ref{sec:cost-models}. Another way to mitigate this problem is be to
+reinstantiate (rematerialize) the computation after a call (may be as a
+different optimization pass).
+
+\subsubsection{Profitability of hoisting loads}
+A load instruction introduces a register where the value loaded will be kept,
+the register pressure increases by one (unless the operand to load becomes dead
+at the load). On the other hand, loading a value early will reduce the stall
+during execution should the value is not in the cache. We generally prefer to
+hoist load except when the hoisting point is too far (this distance is computed by
+looking at the experimental results of representative benchmarks, see
+Section~\ref{sec:experimental-results}).
+
+\subsubsection{Profitability of hoisting stores}
+Since stores do not increase the live-range of any registers, and in some cases
+it ends the liveness of registers, we hoist all the stores.
+
+\subsubsection{Profitability of hoisting calls}
+Currently we hoist all the calls that are suitable candidates for hoisting.
+
+Since stores and calls are hoisted the least \ref{table:experimental-results}
+the performance does not change much whether they are hoisted or not.
+
+\subsection{Code generation}
+Once all the legality and profitability checks are satisfied for a set of
+identical instructions, they are suitable candidates for hoisting. A copy of the
+computation is inserted at the hoisting point along with any instructions which
+needed to be rematerialized. Thereafter, all the computations made redundant by
+the new copy are removed, and the \SSA{} form is restored by updating the
+intermediate representation (IR) to reflect the changes. At the same time
+MemorySSA is also updated to get up-to-date information about memory references.
+
+After one iteration of algorithm runs through the entire function, it creates
+more oppportunities for \emph{higher ranked} computations
+\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
+pass, and so we rerun the code-hoisting algorithm until there are no more
+instructions left to be hoisted.  Obviously, this is not the most optimal
+approach and can be improved by ranking the computations \cite{rosen1988global},
+or by improving the \GVN{} analysis to correctly number loads and dependent
+instructions.
+
+Finally after the transformation is done, we verify a set of post-conditions to
+establish that program invariants are maintained: e.g., consistency of use-defs,
+and \SSA{} semantics.
+
+\newpage
+
+\newpage
+
+\section{SSA based global code motion of identical computations}
+\label{sec:implementation-details}
+% Ranking of instructions from Zadeck
+% Liveness analysis from Ramakrishna
+
+The amount of hoisting depends on whether we collect \GVN{} of instructions
+before finding candidates (optimistic) or, on-demand (pessimistic). It also
+depends on the generality of the \GVN{} algorithm. We have implemented a
+optimistic global code motion of identical computation which uses the liveness
+analysis as illustrated in Das \cite{das2012} and ranking expressions explained
+in Rosen \cite{rosen1988global}.
+
+\subsection{Optimistic code motion algorithm}
+\label{subsec:optimistic}
+We collect the \GVN{} of all the instructions in the function and iterate on the
+list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
+sinking. So first we find the common dominator dominating all such identical
+computations and perform legality checks, as described in
+Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
+instructions to one common dominator, due to legality constraints, e.g.,
+intersecting side-effects, or profitability constraints, e.g., increasing
+register pressure. In those cases, this algorithmn would partition
+(Section~\ref{subsec:partition}) the list of identical instructions into subsets
+which can be partially hoisted to their respective common dominators. However,
+it should be noted that we only hoist very busy expressions to avoid checking
+for undefined behaviors resulting because of introducing extra computation in an
+execution path.
+
+Code hoisting opens new opportunities for other hoistable candidates which were
+of higher rank (depended on candidates which got hoisted). Ideally we could
+iterate on lower ranking expressions first and then proceed to higher ranking
+expressions in the same iteration but LLVM's GVN infrastucture does not compute
+equivalence classes in a effective way. We found it simpler to just recompute
+the value numbers and start finding hoistable candidates again.
 
@@
 
-Currently, we rerun the algorithm until there are no more instructions left to
-be hoisted. This is not the most optimal approach and results in expensive
-analyses to be recomputed. This can be improved by ranking the computations
-\cite{rosen1988global}. Also \GVN{}-hoist runs very early in the pass pipeline, it
-will be good to evaluate the codesize/performance impact when it is run in sync
-with \GVN{}-\PRE{} just like \GCC{} does.
+In general it is a good idea to start with lower ranked expression first such
+that maximum hoisting can happen in one iteration, however, current
+implementation does not rank the expressions and iteratively finds a fixed point
+when no more candidates are available. Even this implementation converges
+quickly and no significant compile time regression have been observed because of
+code hoisting pass. This is not the most optimal approach and results in
+multiple data structures to be recomputed. This can be improved by ranking the
+computations \cite{rosen1988global}. Also \GVN{}-hoist runs very early in the
+pass pipeline, it will be good to evaluate the codesize/performance impact when
+it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.
 

### 2017-04-16 acmart-pacmpl-template.tex
ackn

@@
 
-\newpage
-
 \section{Code motion}
@@
 
-\section{Profitability check (Cost models)}
+\subsection{Profitability check (Cost models)}
 \label{sec:cost-models}
@@
 
-\subsection{Reduce register pressure}
+\subsubsection{Reduce register pressure}
 \label{hoist:reg-pressure}
@@
 In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
-reduce the register pressure in B3 and B4 because two operands will be freed.
+reduce the register pressure in B3 and B4 because two registers will be freed.
 
@@
 
-\subsection{In the presence of calls}
+\subsubsection{Hoisting expression across call}
 \label{cost:across-calls}
-Hoisting scalars across calls is tricky because it can increase the number of
-spills. During the frame lowering of calls, the argument registers, in general,
-the caller saved registers are saved because they might be modified by the
-callee and after the call they are restored. So before the
-call, the register pressure is high because the number of available registers
-are reduced by the number of caller saved registers. In that situation if a
-computation is hoisted across the call, that would increase the total number of
-registers required by one, thus contributing to the register pressure. However,
-in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
-okay to hoist because the register pressure would not decrease.
+Even hoisting scalars across calls is tricky because it can increase the number
+of spills. During the frame lowering of calls, the argument registers, in
+general, the caller saved registers are saved because they might be modified by
+the callee and after the call they are restored. So before the call, the
+register pressure is high because the number of available registers are reduced
+by the number of caller saved registers. In that situation if a computation
+that increases register pressure is not profitable to hoist.
 
 Hoisting loads/stores across calls also require precise analysis of all the
-memory addresses accessed by the call. Our implementation being an
-intraprocedural pass, the analysis is very conservative. In the presence of pure
-calls, loads can be hoisted but stores can't. Also, if the call throws
-exceptions, or if it may not return, memory references cannot be hoisted.
+memory addresses accessed by the call. The current implementation being an
+intraprocedural pass, cannot hoist aggressively across calls. In the presence of
+pure calls, loads can be hoisted but stores can't. Also, if the call throws
+exceptions, or if it may not return, nothing can be hoisted.
 
@@
 and \SSA{} semantics.
-
-\newpage
-
-\newpage
 
@@
 
-\section{Acknowledgments}
-We would like to thank Daniel Berlin for his code reviews and for his feedback
-on earlier versions of this paper and Brian Grayson for motivating examples that
-started this work.
-
 %% Acknowledgments
@@
   %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
+We would like to thank Daniel Berlin for his code reviews and for his feedback
+on earlier versions of this paper and Brian Grayson for motivating examples that
+started this work.
 \end{acks}
@@
 %% Appendix
-\appendix
-\section{Appendix}
-
-Text of appendix \ldots
+%\appendix
+%\section{Appendix}
+%Text of appendix \ldots
 

### 2017-04-16 acmart-pacmpl-template.tex
overall algorithm

@@
 
+% Explain the overall algorithm, explain how anticipability is computed on demand
+% explain the pinned-instructions/barriers.
+\begin{verbatim}
+Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
+
+Compute GVN of each expression in the function
+Compute DJ Graph of function and compute mergeset based on that
+
+For each value number VN with 2 or more instructions:
+  sort the instructions according to DFSIn numbers
+  if first two I1, I2 are hoistable:
+    if they are anticipable at hoist-point:
+      if they are profitable to be hoisted at hoist-point:
+        proceed to check hoistability of subsequent instruction with same VN
+      else
+        proceed to check if I2 and I3 (if available) is hoistable
+
+For each hoistable VN:
+  move the first one I1 to the hoist-point
+  update the use of all other candidates to refer to I1
+  remove all others.
+  if I1 has memory references (a load, store etc.)
+    update MemorySSA of I1 and others to point to the MemorySSA reference of I1
+    remove MemorySSA reference of all others which were deleted
+  update statistics
+
+If any of candidates were hoisted then repeat hoisting
+Once no more candidates are hoistable proceed to sinking
+\end{verbatim}
+
+\begin{verbatim}
+Analyses available: Dominator Tree, DFS Numbering, Memory SSA, Mergeset
+Compute GVN of each expression in the function
+
+For each value number VN with 2 or more instructions:
+  if there are two instructions with same immediate successor as post-dominator
+    if both are only used in the same PHI-node of the successor
+      if both have dependencies that are available or can be made available at the sink point
+         mark VN as sinkable
+
+For each pair I1, I2 of sinkable instructions:
+  Move the first one I1 to the sink point which is just after all the PHI-nodes
+  in the post-dominator
+  update the use of PHI with I1
+  Remove I2 and PHI
+  if I1 is a memory reference:
+    update MemorySSA of I1 and others to point to the MemorySSA reference of I1
+    remove MemorySSA reference of all others which were deleted
+  update statistics
+
+\end{verbatim}
+
 Code hoisting opens new opportunities for other hoistable candidates which were

### 2017-04-16 acmart-pacmpl-template.tex
downward safety

@@
 program and classifying all instructions by their value given by \GVN{}. We
-could computed available and anticipable sets as computed by GCC's code-hoisting
+could compute available and anticipable sets as computed by GCC's code-hoisting
 but that would be a lot of data structures to maintain at each basic block
@@
 
+computing barriers:
+
+\begin{verbatim}
+computing downward-safety of hoistable instructions at hoist-point
+Worklist = list of all basic blocks of hoistable instructions
+for each basic block B in the dominator tree starting at hoist-point:
+  if Worklist is empty
+    return false // Path exists! not downward safe
+  if B is in Worklist:
+    remove B from worklist
+    remove subtree with root at B // Available at B => downward safe from B
+  if B is a leaf basic block:
+    return false  // Path exists! not downward safe
+  if B dominates hoist-point:
+    return false  // Back edge
+
+\end{verbatim}
+
 % Explain the overall algorithm, explain how anticipability is computed on demand
@@
   if first two I1, I2 are hoistable:
-    if they are anticipable at hoist-point:
+    if they are downward-safe at hoist-point:
       if they are profitable to be hoisted at hoist-point:

### 2017-04-16 acmart-pacmpl-template.tex
remove separate profitability args

@@
 
-\subsubsection{Profitability of hoisting scalars}
-Since scalars are the majority of instructions which are hoisted, we pay special
-attention in case of hoisting scalars too far, as that may increase register
-pressure and result in spills. For example hoisting a scalar past a call, as
-described in Section~\ref{cost:across-calls}.  In our current implementation we
-hoist scalars past a call only when optimizing for code-side (-Os). Ideally, a
-later stage of live-range splitting pass should split the live-ranges for
-optimal performance, however, that is not the case with \LLVM{} as we have found
-regressions when scalars are hoisted too far, as in
-Section~\ref{sec:cost-models}. Another way to mitigate this problem is be to
-reinstantiate (rematerialize) the computation after a call (may be as a
-different optimization pass).
-
-\subsubsection{Profitability of hoisting loads}
-A load instruction introduces a register where the value loaded will be kept,
-the register pressure increases by one (unless the operand to load becomes dead
-at the load). On the other hand, loading a value early will reduce the stall
-during execution should the value is not in the cache. We generally prefer to
-hoist load except when the hoisting point is too far (this distance is computed by
-looking at the experimental results of representative benchmarks, see
-Section~\ref{sec:experimental-results}).
-
-\subsubsection{Profitability of hoisting stores}
-Since stores do not increase the live-range of any registers, and in some cases
-it ends the liveness of registers, we hoist all the stores.
-
-\subsubsection{Profitability of hoisting calls}
-Currently we hoist all the calls that are suitable candidates for hoisting.
-
-Since stores and calls are hoisted the least \ref{table:experimental-results}
-the performance does not change much whether they are hoisted or not.
+There is a special case for hoisting load instructions when the hoist-point is
+the predecessor basic block for all the loads. Even if the register pressure
+would increase, we prefer to hoist loads in this case. The reason being, that
+would make the loaded value available by the time it is used. Also, because
+stores and calls are hoisted the least \ref{table:experimental-results}, the
+performance does not change much whether they are hoisted or not.
 

### 2017-04-16 acmart-pacmpl-template.tex
profit

@@
 where a load or a store would happen. It is a scalar computation and gets
-hoisted frequently even if the loads/stores would not get hoisted.  In order to
-reduce register pressure while hoisting loads, we have restricted hoisting of
-address computations away from their corresponding loads and stores when the
-loads and stores cannot be moved.  This restriction is only to mitigate the
-limitations of LLVM's register allocator and may be lifted in the future, when
-the register allocation rematerialization pass has been improved to catch the
-regressions.
+hoisted frequently even if the corresponding loads/stores would not get hoisted.
+In order to reduce register pressure while hoisting loads, we have restricted
+hoisting of address computations away from their corresponding loads and stores
+when the loads and stores cannot be moved.  This restriction is only to mitigate
+the limitations of LLVM's register allocator and may be lifted in the future,
+when the register allocation rematerialization pass has been improved to catch
+these regressions.
 
@@
 \label{sec:implementation-details}
+% Move this before legality and profitability??
 % Ranking of instructions from Zadeck

### 2017-04-16 acmart-pacmpl-template.tex
taylor

@@
 
-Once the instructions have been classified into equivalence classes, we compute
-for each group of equivalent instructions, a point in the program that is both
-legal and profitable for the instructions to be moved to.
+Once the instructions have been classified into congruence classes
+\cite{briggs1997} i.e., a set of instructions which compute the same value, we
+compute for each group of congruent instructions, a point in the program that is
+both legal and profitable for the instructions to be moved to.
 
@@
 approach and can be improved by ranking the computations \cite{rosen1988global},
-or by improving the \GVN{} analysis to correctly number loads and dependent
-instructions.
+or by improving the \GVN{} analysis to correctly populate congruence classes.
 

### 2017-04-16 acmart-pacmpl-template.tex
barriers

@@
 \subsection{Finding candidates to move}
+\label{subsec:finding-candidates}
 The first step is to find a set of instructions that perform identical
@@
 paths.
-
-Code-motion basically consists of two parts i.e., hoisting and sinking. The
-unique part about this algorithm is that it only moves identical
-computations. An immediate guarantee of this approach is gain in code size (the
-final executable may be of larger size because of more inlining). The algorithm
-prefers hoisting to sinking. If the dependency of a hoistable candidate is in
-the same basic block as the candidate, then the dependency must also be
-hoistable otherwise hoisting will be illegal or would require a complicated code
-generation to make it legal. The current algorithm discards cases if the
-dependency is neither hoistable nor rematerializable.
 
@@
 % Liveness analysis from Ramakrishna
-
-The amount of hoisting depends on whether we collect \GVN{} of instructions
-before finding candidates (optimistic) or, on-demand (pessimistic). It also
-depends on the generality of the \GVN{} algorithm. We have implemented a
-optimistic global code motion of identical computation which uses the liveness
-analysis as illustrated in Das \cite{das2012} and ranking expressions explained
-in Rosen \cite{rosen1988global}.
+In this section we will present an overview of the algorithm we implemented in
+\LLVM{}.  The amount of hoisting depends on whether we collect \GVN{} of
+instructions before finding candidates (optimistic) or, on-demand
+(pessimistic). It also depends on the generality of the \GVN{} algorithm as
+mentioned earlier in Section~\ref{subsec:finding-candidates}. We have
+implemented a optimistic global code motion of congruent instructions which uses
+the liveness analysis as illustrated in Das \cite{das2012} and ranking
+expressions explained in Rosen \cite{rosen1988global}.
+
+Code-motion basically consists of two parts i.e., hoisting and sinking. This
+implementation only moves congruent instructions. An immediate guarantee of this
+approach is gain in code size (the final executable may be of larger size
+because of more inlining). The algorithm prefers hoisting to sinking. If the
+dependency of a hoistable candidate is in the same basic block as the candidate,
+then the dependency must also be hoistable otherwise hoisting will be illegal or
+would require a complicated code generation to make it legal. The current
+algorithm discards cases if the dependency is neither hoistable nor
+rematerializable.
 
@@
 intersecting side-effects, or profitability constraints, e.g., increasing
-register pressure. In those cases, this algorithmn would partition
+register pressure; in those cases, this algorithmn would partition
 (Section~\ref{subsec:partition}) the list of identical instructions into subsets
@@
 
+\begin{verbatim}
 computing barriers:
+for each basic block B in a function:
+  barrier_found = false
+  for each instruction I in B:
+    if I does not guarantee progress:
+      mark I as a barrier instruction
+      barrier_found = true
+      break;
+
+  // Find the last barrier below which instructions can be sunk.
+  // If there was no barrier in B, any instruction satisfying
+  // other legality checks can be sunk.
+  if barrier_found:
+    for each instruction I in B in reverse order:
+      if I does not guarantee progress:
+        mark I as a sink barrier
+        break
+\end{verbatim}
 

### 2017-04-16 acmart-pacmpl-template.tex
barriers

@@
 and move the code with identical computations from the sibling branch to a
-common dominator if they are very busy \cite{steven1997advanced}. We also
-incorporate ranking of expressions in our implementation.
+common dominator if they are very busy \cite{steven1997advanced}.
 
@@
 execution path.
+
+Barriers are based on the concept of pinned instructions \cite{click1995global}
+but extended to adapt to LLVM IR. Since a basic block in LLVM IR is actually an
+extended basic block because there might be calls in the middle of a basic block
+which might not return. So barriers can be present in the middle of a basic
+block. Essentially, any instruction that cannot not guarantee progress is marked
+as a barrier. In the absense of context (as in our current implementation), some
+instructions which might be safe are still classified a barrier e.g., volatile
+loads/stores, calls with missing attributes.
 

### 2017-04-16 acmart-pacmpl-template.tex
time

@@
 candidates closer in the list have their common dominator nearby. In
-Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and for
-some reason they cannot be hoisted together at B0, then we partition the set of
-hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
-be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
-B1 and those in B5, B6 to be hoisted at B2.
+Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and
+for some reason they cannot be hoisted together at B0, then we partition the set
+of hoistable candidates in their DFSIn order. In this case the DFSIn ordering
+would be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be
+hoisted at B1 and those in B5, B6 to be hoisted at B2. Even if (B3, B4) could
+not be hoisted, for some reason, to B1, (B5, B6) can still be hoisted -- if
+legal and profitable -- to B2.
 
@@
 \subsection{Time complexity of algorithm}
-The complexity of code hoisting is linear in number of instructions that could
-be hoisted in the program, matching the complexity of \PRE{} on \SSA{} form.  The
-analysis phase is based on the Global Value Numbering (\GVN{}), the same analysis
-used for \PRE{}, followed by the computation of a partition of identical
-expressions to be hoisted in a same location to guarantee safety properties and
-program performance, and followed by a simple code generation that adds the
-identified instruction in the hoisting point and removes all the now redundant
-expressions.
+The complexity of code hoisting is linear in number of instructions that are
+candidates for code-motion, matching the complexity of \PRE{} on \SSA{} form.
+The analyses computed for this pass are Global Value Numbering, Computation of
+DJ Graph and MergeSets, Marking Barriers, all are linear in number of
+instructions in a function. Liveness analysis expensive but only performed
+on-demand for hoistable candidates so it does not affect the overall compile
+time by much. Other analyses like Alias Analysis, Memory SSA and Dominator Tree
+are already available in the LLVM pass pipeline. Although we recompute GVN, and
+Barriers for each iteration of the code-motion, we have not seen signinficant
+increase in the compilation times. We have also provided appropriate compiler
+flags to expedite code-motion by bailing out with fewer iterations, or skip the
+liveness based profitability analysis to aggressively move the code as long as
+they are legal (Section~\ref{sec:experimental-results}).
+
+The analysis is followed by a simple code generation that adds the identified
+instruction in the destination point and removes all instructions rendered
+redundant as a result of code-motion.
 

### 2017-04-16 acmart-pacmpl-template.tex
refactoring

@@
 \section{Introduction}
+\label{sec:intro}
 Compiler techniques to remove redundant computations are composed of an analysis
@@
 
-This paper describes code-motion of identical computations, a technique that
-uses the information computed for \PRE{} to detect identical computations but
-has a transformation phase whose goal differs from \PRE{}: it removes identical
-computations from different branches of execution.  These identical computations
-in different branches of execution are not redundant computations at run-time
-and the number of run-time computations is not reduced. It is not a redundancy
-elimination pass, and thus it has different cost function and heuristics than
-\PRE{} or CSE. It is more similar to global code scheduling
-\cite{dragonbook,click1995global} in the sense that it will only move
-computations. We will now consider an example that explains how code hoisting can
-also improve performance.
-
-\subsection{Illustrative Example}
-\label{subsec:example}
-Code hoisting can reduce the critical path length of execution in out of
-order machines. As more instructios are available at the hoisting point, the
-hardware has more instructions to reorder. Following example illustrates how
+This paper describes code-motion of congruent instructions \cite{briggs1997}, a
+technique that uses the information computed for \PRE{} to detect identical
+computations but has a transformation phase whose goal differs from \PRE{}: it
+removes identical computations from different branches of execution.  These
+identical computations in different branches of execution are not redundant
+computations at run-time and the number of run-time computations is not
+reduced. It is not a redundancy elimination pass, and thus it has different cost
+function and heuristics than \PRE{} or CSE. It is more similar to global code
+scheduling \cite{dragonbook,click1995global} in the sense that it will only move
+computations. Code hoisting can reduce the critical path length of execution in
+out of order machines. As more instructios are available at the hoisting point,
+the hardware has more instructions to reorder. Following example illustrates how
 hoisting can improve performance by exposing more ILP.
@@
 
-In this code the two subtractions and the division operations can be executed in
-parallel because there are no dependencies among them. So the total number of
-cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since $C_{div}$ is
-usually much greater than $C_{sub}$ \cite{x86,aarch64}.
+In this code the division operation can be executed in parallel with the two
+subtractions because there are no dependencies among them. So the total number
+of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
+$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}.
 
@@
 \begin{itemize}
-\item It reduces the code size of the program;
-\item It improves function inlining heuristics: functions become cheaper to
-  inline when their code size decreases because inliner heuristics in \LLVM{}
-  depends on instruction count.
+\item It helps reduce the code size of the program by replacing multiple
+  instructions with one although the final code size may be higher because of
+  improved inlining heuristics; functions become cheaper to inline when their
+  code size decreases because inliner heuristics in \LLVM{} depends on
+  instruction count.
 \item It expose more instruction level parallelism to the later compiler
@@
   total number of instructions to be executed.
-\item It improves passes that do not work well with branches:
-  \begin{itemize}
-  \item to improve loop vectorization by reducing a loop with control flow to a
-    loop with a single BB, should all the instructions in a conditional get
+\item It may improve loop vectorization by reducing a loop with control flow to
+  a loop with a single BB, should all the instructions in a conditional get
   hoisted or sinked;
-  \item to enable more loop invariant code motion (LICM): as LICM does not
-    reason about instructions in the context of loops with conditional branches,
-    code-hoisting is needed to move instructions out of conditional expressions
-    and expose them to LICM.
-  \end{itemize}
+\item It enables more loop invariant code motion (LICM): as LICM passes, in
+  general, cannot effectively reason about instructions within conditional
+  branches the context of loops, code-hoisting is needed to move instructions
+  out of conditional expressions and expose them to LICM.
 \end{itemize}
@@
 implementation (aggressive code hoisting) is already merged in LLVM trunk,
-however, a general implementation of code-motion of identical expressions is
-still missing from GCC and LLVM trunk. The main contributions of this paper are:
+however, a general implementation of code-motion of congruent (but not
+redundant) expressions is still missing from GCC and LLVM trunk. The main
+contributions of this paper are:
 \begin{itemize}
-\item a new optimistic algorithm to move computations out of branches,
-\item a cost model to reduce interference and reduce spills,
+\item a new optimistic algorithm to move congruent instructiosn out of branches,
+\item a cost model to reduce interference and hence, reduce spills,
 \item a technique to maximize hoisting in an optimistic approach by partitioning
@@
 code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
-basic blocks: the instructions of two sibling basic blocks are read in the same
+basic blocks: the instructions of two sibling basic blocks are read at the same
 time, and all the instructions of the blocks are hoisted to the common parent
@@
 
-Rosen \cite{rosen1988global} also introduced moving computations from successors
-but using value numbering. Their algorithm iterates on computations of same rank
-and move the code with identical computations from the sibling branch to a
-common dominator if they are very busy \cite{steven1997advanced}.
+Rosen \cite{rosen1988global} also briefly discuss hoisting computations with
+identical value numbers from immediate successors. Their algorithm iterates on
+computations of same rank and move the code with identical computations from the
+sibling branch to a common dominator if they are very busy
+\cite{steven1997advanced}. The cost-model to mitigate register pressure is
+missing, also there is no mention of sinking congruent instructions.
 
@@
 
-Barany \cite{barany2013} presented a global scheduler with ILP formulation with
-a goal to minimize register pressure. The results they got were not very
-promising. It may be because they only used the scheduler for smaller functions
-($<1000$ instructions) also they compiled the benchmarks for ARM-Cortex which is
-more resilient to register pressure because it has more registers compared to
-say X86.
+Barany \cite{barany2013} presented a global scheduler with integer linear
+programming (ILP) formulation with a goal to minimize register pressure. The
+results they got were not very promising. It may be because they only used the
+scheduler for smaller functions ($<1000$ instructions) also they compiled the
+benchmarks for ARM-Cortex which is more resilient to register pressure because
+it has more registers compared to say X86.
 
@@
 Barany and Shobaki's implmentation will have similar results when compiled for
-same target. Also, both suffer from the same problem, although Shobaki not so
-much, of large compile times which is not feasible for industrial compilers like
-gcc and LLVM.
+same target architecture. Also, both suffer from the same problem, although
+Shobaki not so much, of large compile times which is not feasible for industrial
+compilers like gcc and LLVM.
 
@@
 \begin{itemize}
-\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
-  relations \cite{dragonbook};
+\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
+  Control Flow Graph (\CFG{}).
 \item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
@@
 \begin{itemize}
-\item find candidates suitable for code-motion
+\item find candidates (congruent instructions) suitable for code-motion,
 \item compute a point in the program where it is both legal and profitable to
   move the code,
-\item transform the code and update data structures to continue iterative code
-  motion.
+\item move the code to hoist point or sink point, and
+\item update data structures to continue iterative code motion.
 \end{itemize}
@@
 version of merge set that we implemented is presented here:
-
-\begin{verbatim}
-// Return true if the merge set of source node of
-// a visited J-edge (incoming edge of lnode) is not
-// the subset of the merge set of lnode.
-bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet M)
-   for (all incoming edges to lnode) do
-     Let e = Incoming edge
-     if (e is in JE && e in V) then
-       Let snode = Source Node of e
-       if (M(snode)!(Subset) M(lnode)) then
-         return true
-       end if
-     end if
-   end for
-return false
-\end{verbatim}
 
@@
 }
+\end{verbatim}
+
+\begin{verbatim}
+// Return true if the merge set of source node of
+// a visited J-edge (incoming edge of lnode) is not
+// the subset of the merge set of lnode.
+bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet M)
+   for (all incoming edges to lnode) do
+     Let e = Incoming edge
+     if (e is in JE && e in V) then
+       Let snode = Source Node of e
+       if (M(snode)!(Subset) M(lnode)) then
+         return true
+       end if
+     end if
+   end for
+return false
 \end{verbatim}
@@
 
-The code shown in Section~\ref{subsec:example} is a reduced example that appears
-in a hot loop of a proprietary benchmark.  When the expressions are hoisted from
-the conditional clauses, the overall performance of that benchmark improves by
+The code shown in Section~\ref{sec:intro} is a reduced example that appears in a
+hot loop of a proprietary benchmark.  When the expressions are hoisted from the
+conditional clauses, the overall performance of that benchmark improves by
 $15\%$ on an out-of-order processor due to increased instruction level

### 2017-04-17 acmart-pacmpl-template.tex
iskill

@@
 
-\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
+\GCC{}ecently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
 part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
@@
 It then checks to see if at least one successor of B has the value available.
-This avoids hoisting it way up the chain to ANTIC.  It also checks to ensure
-that B has multiple successors, since hoisting in a straight line is pointless.
-The algorithm continues top down the dominator tree, iterating with \PRE{} until
-no more changes.  One advantage of \GCC{} implementation is that it works in
-sync with the \GVN{}-\PRE{} such that when new hoisting opportunities are
-created by \GVN{}-\PRE{}, code-hoisting will hoist them.
+This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
+ensure that B has multiple successors, since hoisting in a straight line is
+pointless.  The algorithm continues top down the dominator tree, working in
+tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
+implementation is that it works in sync with the \GVN{}-\PRE{} such that when
+new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
+hoist them.
 
@@
 results they got were not very promising. It may be because they only used the
-scheduler for smaller functions ($<1000$ instructions) also they compiled the
+scheduler for smaller functions ($<1000$ instructions); also, they compiled the
 benchmarks for ARM-Cortex which is more resilient to register pressure because
-it has more registers compared to say X86.
+it has more registers compared to X86, for example.
 
@@
 
+\begin{verbatim}
+// return true if op is the last use of the
+// operand in I.
+bool isKill(Operand op, Instruction I) {
+  if (isLiveOutUsingMergeSet(op))
+    return false;
+  B = basic_block(I);
+  for each I1 after I in B:
+    if (I1 uses op)
+      return false;
+
+return true;
+}
+\end{verbatim}
+
 The original algorithm presented in \cite{das2012} has mismatched types in terms
@@
 terminate. These two cases were missing from the algorithm and we came across
-them during implementation.
+them during implementation. If a variable is not live-out of a basic block, it
+still may be used later in the basic block. To establish whether a use of a
+variable is a kill we iterate on all the subsequent instructions in a basic
+block checking for uses should isLiveOutUsingMergeSet return false.
 

### 2017-04-17 acmart-pacmpl-template.tex
mover partition to profitability

@@
 \label{subsec:finding-candidates}
-The first step is to find a set of instructions that perform identical
-computations: this is performed by a linear scan of all instructions of the
-program and classifying all instructions by their value given by \GVN{}. We
-could compute available and anticipable sets as computed by GCC's code-hoisting
-but that would be a lot of data structures to maintain at each basic block
-level. For GCC it makes sense because their code-hoisting is integrated with
-GVN-PRE which already has those data structures available.
+The first step is to find a set of congruent instructions
+\cite{briggs1997}. This is performed by a linear scan of all instructions of the
+program and classifying them by their value numbers. We could compute available
+and anticipable sets as computed by GCC's code-hoisting but that would be a lot
+of data structures to maintain at each basic block level. For GCC it makes sense
+because their code-hoisting is integrated with GVN-PRE which already has those
+data structures available.
 
@@
 
-Once the instructions have been classified into congruence classes
-\cite{briggs1997} i.e., a set of instructions which compute the same value, we
-compute for each group of congruent instructions, a point in the program that is
-both legal and profitable for the instructions to be moved to.
+Once the instructions have been classified into congruence classes, we compute
+for each group of congruent instructions, a point in the program that is both
+legal and profitable for the instructions to be moved to.
 
@@
 backwards on the control flow to the hoisting point, we check whether there are
-writes to memory that may alias with the load, in which case we discard the
-candidate. We use the MemorySSA infrastructure of \LLVM{} to iterate on the
-use-def chains available for memory references.
+writes to memory that may alias with the load, in which case the candidate is
+discarded. To iterate on the use-def chains for memory references the MemorySSA
+infrastructure of \LLVM{} is used.
 
@@
 loads using the MemorySSA of \LLVM{}. We check that the operands of the store
-instruction are available at the hoisting point, that there are no aliasing
+instruction are available at the hoisting point, and that there are no aliasing
 loads or store along the path from the current position to the hoisting point.
@@
 
+Hoisting loads/stores across calls also require precise analysis of all the
+memory addresses accessed by the call. The current implementation being an
+intraprocedural pass, cannot hoist aggressively across calls. In the presence of
+pure calls, loads can be hoisted but stores can't. Also, if a call throws
+exceptions, or if it may not return, nothing can be hoisted across that call.
+
 \subsubsection{Legality of sinking expressions}
-One might wonder that if a set of expressions compute the same value then why
-not just hoist them. The problem is, sometimes, hoisting is not upward-safe
-\ref{click1995global} e.g., if the expressions are in a landing pad
-etc. Sometimes sinking may reduce the register pressure e.g., when the use
-operands are not kills.
-
-%TODO: Explain why all the identical computation cannot be hoisted.
-For sinking, higher ranked expressions would be sunk first. And it would be
-illegal to sink higher ranked identical expressions if they are not anticipable
-in the common post-dominator. For example:
+Sometimes, hoisting is not upward-safe \ref{click1995global} e.g., if the
+expressions are in a landing pad etc., in that case sinking of those expressions
+may reduce code size. Sinking may also reduce the register pressure in some
+cases e.g., when the use operands are not kills.For sinking, higher ranked
+expressions would be sunk first. And it would be illegal to sink higher ranked
+identical expressions if they are not anticipable in the common
+post-dominator. For example:
 
@@
 
-In this example (c1, c2) or (d1, d2) could be sunk first. Since (c1, c2) depend
-on i1 and i2 respectively which are also in their original basic blocks, c1 and
-c2 are not anticipable in B3. So without knowing the sinkability of `i1' and
-`i2' it would be illegal to sink (c1, c2) to B3. On the other hand (d1, d2) can
-safely be sunk because their operands are readily available at the sink point
-i.e., B3. It should also be noted that, just because the expressions are
-identical and operands are available, it still requires a unique post-dominating
-PHI to use the exact same values to be legally sinkable.
+In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
+(c1, c2) depend on i1 and i2 respectively which are also in their original basic
+blocks, c1 and c2 are not anticipable in B3. So without knowing the sinkability
+of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
+(d1, d2) can safely be sunk because their operands are readily available at the
+sink point i.e., B3. It should also be noted that, just because the expressions
+are identical and operands are available, it still requires a unique
+post-dominating PHI to use the exact same values to be legally sinkable.
 
@@
 expressions are moved in the current implementation, there is no need to check
-for undefinedness resulting due to movement of instruction. This simplifies the
+for undefinedness resulting due to movement of instructions. This simplifies the
 implementation.
+
+\subsection{Profitability check (Cost models)}
+\label{sec:cost-models}
+After the legality checks have passed, we check if a code-motion is profitable.
+That takes into account the impact code-motion would have on various parameters
+that might affect runtime performance e.g., impact on live-range, gain in the
+code size. Since this is mostly a code-size optimization pass, the goal is to
+not regress in performance across popular benchmarks at the same time reduce
+code size as much as possible. Following are the cost models which are
+implemented:
+
+\subsubsection{Reduce register pressure}
+\label{hoist:reg-pressure}
+Hoisting upwards will decrease the live-range of its use, if it is a last use (a
+kill), but increase the live-range of its definition. Conversely, sinking will
+decrease the live-range of the defined register but increase the live-range for
+killed operands. If the live-range after code-motion is less than before it will
+be moved.
+
+Essentially, as long as there is one killed operand , code hoisting will either
+decrease or preserve the register pressure. Similarly, code-sinking will either
+decrease or preserve the register pressure as long as there is one operand
+killed at most.
+
+Following example explains how code motion of identical computations can reduce
+the register pressure.  Consider the following example where the labels prefixed
+with 'P' represent the position of instruction in a basic block (names prefixed
+with 'B').
+
+\begin{verbatim}
+    b = m
+    c = n
+B0: if c is true then goto B3 else goto B4
+
+B1: a0 = b<kill> + c<kill>
+
+B2: a1 = b<kill> + c<kill>
+
+After hoisting a0 and a1 are removed and a copy
+of a0 as a01 is placed in B0 just before the branch.
+\end{verbatim}
+
+In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
+reduce the register pressure in B3 and B4 because two registers will be freed.
+
+Ideally, it should be okay to hoist all the instructions and a later a
+live-range-splitting \cite{cooper1998live} pass should make the right decision
+of rematerializing the instruction should it be beneficial to do so. But the
+current live-range splitting pass of \LLVM{} is not making the optimial decision
+and we have found regressions while hoisting aggressively.
+
+Moreover, LLVM has a `getelementptr' instruction which computes the address
+where a load or a store would happen. It is a scalar computation and gets
+hoisted frequently even if the corresponding loads/stores would not get hoisted.
+In order to reduce register pressure while hoisting loads, we have restricted
+hoisting of address computations away from their corresponding loads and stores
+when the loads and stores cannot be moved.  This restriction is only to mitigate
+the limitations of LLVM's register allocator and may be lifted in the future,
+when the register allocation rematerialization pass has been improved to catch
+these regressions.
+
+\subsubsection{Hoisting expression across call}
+\label{cost:across-calls}
+Even hoisting scalars across calls is tricky because it can increase the number
+of spills. During the frame lowering of calls, the argument registers, in
+general, the caller saved registers are saved because they might be modified by
+the callee and after the call they are restored. So before the call, the
+register pressure is high because the number of available registers are reduced
+by the number of caller saved registers. In that situation if a computation
+that increases register pressure is not profitable to hoist.
 
@@
 In the approach, described in Section~\ref{subsec:optimistic}, it is possible
-that a common hoisting point of all the instructions is either too far away, or
-not legally possible. In these cases, it is still possible to `partially' hoist
-a subset of instructions by splitting the set of candidates and finding a closer
-hoisting point for each subset.
+that a common hoisting point for all the instructions does not satisfy legality
+or profitability checks. In these cases, it is still possible to `partially'
+hoist a subset of instructions by splitting the set of candidates and finding a
+closer hoisting point for each subset.
 
@@
 candidates closer in the list have their common dominator nearby. In
-Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and
-for some reason they cannot be hoisted together at B0, then we partition the set
-of hoistable candidates in their DFSIn order. In this case the DFSIn ordering
-would be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be
-hoisted at B1 and those in B5, B6 to be hoisted at B2. Even if (B3, B4) could
-not be hoisted, for some reason, to B1, (B5, B6) can still be hoisted -- if
-legal and profitable -- to B2.
+Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
+for some reason, they cannot be hoisted at B0, then we partition the set of
+hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
+be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
+B1 and those in B5, B6 to be hoisted at B2. Even if (B3, B4) could not be
+hoisted, for some reason, to B1, (B5, B6) can still be hoisted -- if legal and
+profitable -- to B2.
 
@@
 new hoisting point for the remaining ones.
-
-\subsection{Profitability check (Cost models)}
-\label{sec:cost-models}
-After the legality checks have passed, we check if a code-motion is profitable.
-That takes into account the impact code-motion would have on various parameters
-that might affect runtime performance e.g., impact on live-range, gain in the
-code size. Since this is mostly a code-size optimization pass, the goal is to
-not regress in performance across popular benchmarks at the same time reduce
-code size as much as possible. Following are the cost models which are
-implemented:
-
-\subsubsection{Reduce register pressure}
-\label{hoist:reg-pressure}
-Hoisting upwards will decrease the live-range of its use, if it is a last use (a
-kill), but increase the live-range of its definition. Conversely, sinking will
-decrease the live-range of the defined register but increase the live-range for
-killed operands. If the live-range after code-motion is less than before it will
-be moved.
-
-Essentially, as long as there is one killed operand , code hoisting will either
-decrease or preserve the register pressure. Similarly, code-sinking will either
-decrease or preserve the register pressure as long as there is one operand
-killed at most.
-
-Following example explains how code motion of identical computations can reduce
-the register pressure.  Consider the following example where the labels prefixed
-with 'P' represent the position of instruction in a basic block (names prefixed
-with 'B').
-
-\begin{verbatim}
-    b = m
-    c = n
-B0: if c is true then goto B3 else goto B4
-
-B1: a0 = b<kill> + c<kill>
-
-B2: a1 = b<kill> + c<kill>
-
-After hoisting a0 and a1 are removed and a copy
-of a0 as a01 is placed in B0 just before the branch.
-\end{verbatim}
-
-In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
-reduce the register pressure in B3 and B4 because two registers will be freed.
-
-Ideally, it should be okay to hoist all the instructions and a later a
-live-range-splitting \cite{cooper1998live} pass should make the right decision
-of rematerializing the instruction should it be beneficial to do so. But the
-current live-range splitting pass of \LLVM{} is not making the optimial decision
-and we have found regressions while hoisting aggressively.
-
-Moreover, LLVM has a `getelementptr' instruction which computes the address
-where a load or a store would happen. It is a scalar computation and gets
-hoisted frequently even if the corresponding loads/stores would not get hoisted.
-In order to reduce register pressure while hoisting loads, we have restricted
-hoisting of address computations away from their corresponding loads and stores
-when the loads and stores cannot be moved.  This restriction is only to mitigate
-the limitations of LLVM's register allocator and may be lifted in the future,
-when the register allocation rematerialization pass has been improved to catch
-these regressions.
-
-\subsubsection{Hoisting expression across call}
-\label{cost:across-calls}
-Even hoisting scalars across calls is tricky because it can increase the number
-of spills. During the frame lowering of calls, the argument registers, in
-general, the caller saved registers are saved because they might be modified by
-the callee and after the call they are restored. So before the call, the
-register pressure is high because the number of available registers are reduced
-by the number of caller saved registers. In that situation if a computation
-that increases register pressure is not profitable to hoist.
-
-Hoisting loads/stores across calls also require precise analysis of all the
-memory addresses accessed by the call. The current implementation being an
-intraprocedural pass, cannot hoist aggressively across calls. In the presence of
-pure calls, loads can be hoisted but stores can't. Also, if the call throws
-exceptions, or if it may not return, nothing can be hoisted.
 

### 2017-04-17 acmart-pacmpl-template.tex
refactor

@@
 killed operands. If the live-range after code-motion is less than before it will
-be moved.
-
-Essentially, as long as there is one killed operand , code hoisting will either
-decrease or preserve the register pressure. Similarly, code-sinking will either
-decrease or preserve the register pressure as long as there is one operand
-killed at most.
-
-Following example explains how code motion of identical computations can reduce
-the register pressure.  Consider the following example where the labels prefixed
-with 'P' represent the position of instruction in a basic block (names prefixed
-with 'B').
+be moved. Essentially, as long as there is one killed operand , code hoisting
+will either decrease or preserve the register pressure.  Similarly, code-sinking
+will either decrease or preserve the register pressure as long as there is one
+operand killed at most.  Following example explains how code motion of identical
+computations can reduce the register pressure.  Consider the following example
+where the labels prefixed with 'P' represent the position of instruction in a
+basic block (names prefixed with 'B').
 
@@
 
-\subsubsection{Hoisting expression across call}
+\subsubsection{Hoisting an expression across a call}
 \label{cost:across-calls}
 Even hoisting scalars across calls is tricky because it can increase the number
-of spills. During the frame lowering of calls, the argument registers, in
-general, the caller saved registers are saved because they might be modified by
-the callee and after the call they are restored. So before the call, the
-register pressure is high because the number of available registers are reduced
-by the number of caller saved registers. In that situation if a computation
-that increases register pressure is not profitable to hoist.
+of spills. During the frame lowering of calls, the argument registers also known
+as the caller saved registers are saved because they might be modified by the
+callee and after the call they are restored. So before the call, the register
+pressure is high because the number of available registers are reduced by the
+number of caller saved registers. In that situation a computation that increases
+register pressure is not profitable to hoist.
 
@@
 
+% TODO: Rewrite the following.
 In our current implementation we keep as many candidates in one set as possible
@@
 Once all the legality and profitability checks are satisfied for a set of
-identical instructions, they are suitable candidates for hoisting. A copy of the
+congruent instructions, they are suitable candidates for hoisting. A copy of the
 computation is inserted at the hoisting point along with any instructions which

### 2017-04-17 acmart-pacmpl-template.tex
adding codesize and spills

@@
       Metric               & Before & After              \\\hline
-      Call sites deleted, not inlined             & 1988    & 1988   \\\hline
       Functions deleted (all callers found)       & 38250   & 38255  \\\hline
       Functions inlined                           & 154986  & 154985 \\\hline
-      Allocas merged together                     & 212     & 212    \\\hline
-      Caller-callers analyzed                     & 193042  & 193092 \\\hline
-      Call sites analyzed                         & 414336  & 414381 \\\hline
-      Rematerialized defs for spilling            & 18321   & 18326  \\\hline
-      Rematerialized defs for splitting           & 5719    & 5842   \\\hline
-      Spill slots allocated                       & 42912   & 42970  \\\hline
-      Spilled live ranges                         & 61330   & 61362  \\\hline
       Spills inserted                             & 50724   & 50784  \\\hline
+      LICM
 \end{tabular}
@@
       \hline
-Number of spills                &base   &code-motion &\%loss \\\hline
-400.perlbench	&2542	&2481	&0.97\\\hline
-401.bzip2	        &718	&707	&0.98\\\hline
-403.gcc	        &5778	&5710	&0.98\\\hline
-429.mcf	        &14	&17	&1.21\\\hline
-433.milc	        &616	&624	&1.01\\\hline
-444.namd	        &3185	&3223	&1.01\\\hline
-445.gobmk	        &2166	&2170	&1.00\\\hline
-447.dealII	        &10038	&9151	&0.91\\\hline
-450.soplex	        &1104	&1114	&1.00\\\hline
-453.povray	        &5199	&5176	&0.99\\\hline
-456.hmmer	        &1195	&1195	&1   \\\hline
-458.sjeng	        &168	&169	&1.00\\\hline
-462.libquantum	&183	&183	&1   \\\hline
-464.h264ref	&3382	&3423	&1.01\\\hline
-470.lbm	        &30	&30	&1   \\\hline
-471.omnetpp	&521	&522	&1.00\\\hline
-473.astar	        &176	&196	&1.11\\\hline
-482.sphinx3	&603	&594	&0.98\\\hline
-483.xalancbmk	&5101	&4941	&0.96\\\hline
-Grand Total	                &42719	&41626	&0.97\\\hline
+Number of spills  &base(-Ofast) &code-motion(-Ofast) &\%loss \\\hline
+400.perlbench	  & 2539	& 2493	& 0.98
+401.bzip2	  & 718	        & 707	& 0.98
+403.gcc	          & 5782	& 5722	& 0.99
+429.mcf	          & 14	        & 17	& 1.21
+433.milc	  & 635	        & 642	& 1.01
+444.namd	  & 3223	& 3274	& 1.01
+445.gobmk	  & 2166	& 2173	& 1.00
+447.dealII	  & 11074	& 10234	& 0.92
+450.soplex	  & 1122	& 1125	& 1.00
+453.povray	  & 4701	& 4692	& 1.09
+456.hmmer	  & 1197	& 1274	& 1.06
+458.sjeng	  & 168	        & 176	& 1.05
+462.libquantum	  & 118	        & 118	& 1
+464.h264ref	  & 3379	& 3454	& 1.02
+470.lbm	          & 33	        & 33	& 1
+471.omnetpp	  & 524	        & 527	& 1.00
+473.astar	  & 180	        & 201	& 1.12
+482.sphinx3	  & 674	        & 670	& 0.99
+483.xalancbmk	  & 5102	& 4965	& 0.97
+Grand Total	  & 43349	& 42497	& 0.98
     \end{tabular}
   \end{center}
-  \caption{Number of spills on SPEC2006}
+  \caption{Number of spills on SPEC2006 at -Ofast}
+  \label{tab:code-size}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|}
+      \hline
+
+Code size         & base(ofast)	& ofast+gvnhoist&	increase in codesize \\\hline
+400.perlbench	  &  2469594	& 2478714	& 1.00\\\hline
+401.bzip2	  &  865902	& 867502	& 1.00 \\\hline
+403.gcc	          &  4379291	& 4437467	& 1.01 \\\hline
+429.mcf	          &  794909	& 794749	& 1.00 \\\hline
+433.milc	  &  1009553	& 1010209	& 1.00 \\\hline
+444.namd	  &  282794	& 286658	& 1.01 \\\hline
+445.gobmk	  &  2474193	& 2494673	& 1.01 \\\hline
+447.dealII	  &  3577836	& 3605672	& 1.01 \\\hline
+450.soplex	  &  411254	& 409034	& 0.99 \\\hline
+453.povray	  &  1114405	& 1114241	& 1.00 \\\hline
+456.hmmer	  &  1421823	& 1425287	& 1.00 \\\hline
+458.sjeng	  &  941066	& 940626	& 1.00 \\\hline
+462.libquantum	  &  911028	& 911124	& 1.00 \\\hline
+464.h264ref	  &  1677417	& 1680841	& 1.00 \\\hline
+470.lbm	          &  795444	& 795444	& 1    \\\hline
+471.omnetpp	  &  661099	& 658411	& 0.99 \\\hline
+473.astar	  &  44296	& 45224		& 1.02 \\\hline
+482.sphinx3	  &  1337445	& 1334437	& 1.00 \\\hline
+483.xalancbmk	  &  4293600	& 4247108	& 0.99 \\\hline
+    \end{tabular}
+  \end{center}
+  \caption{Number of spills on SPEC2006 at -Ofast}
   \label{tab:code-size}

### 2017-04-17 acmart-pacmpl-template.tex
stats

@@
 
+      TODO: Check if putting code-size or spills number is okay or only % is required.
 Code size         & base(ofast)	& ofast+gvnhoist&	increase in codesize \\\hline
-400.perlbench	  &  2469594	& 2478714	& 1.00\\\hline
+400.perlbench	  &  2469594	& 2478714	& 1.00 \\\hline
 401.bzip2	  &  865902	& 867502	& 1.00 \\\hline
@@
 
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|}
+      \hline
+ hoist licm	sunk licm		functions inlined		f deleted	
+ base	& gvn	& base	& gvn	& base	& gvn	& base	& gvn
+ 400.perlbench	  & 2428	& 2480	& 36	& 35	& 1562	& 1565	& 232	& 234
+ 401.bzip2	  & 721	& 825	& 3	& 3	& 172	& 180	& 32	& 33
+ 403.gcc	          & 8867	& 10079	& 143	& 182	& 5034	& 5080	& 1303	& 1304
+ 429.mcf	          & 55	& 51	& 1	& 1	& 4	& 4	& 	& 
+ 433.milc	  & 2537	& 2795	& 3	& 3	& 75	& 81	& 	& 
+ 444.namd	  & 2204	& 2206	& 240	& 240	& 387	& 390	& 57	& 55
+ 445.gobmk	  & 3090	& 3107	& 18	& 19	& 735	& 775	& 230	& 234
+ 447.dealII	  & 59114	& 95884	& 60	& 8	& 100891	& 100692	& 30219	& 29966
+ 450.soplex	  & 4668	& 4848	& 18	& 17	& 9801	& 9784	& 3125	& 3112
+ 453.povray	  & 9533	& 9079	& 23	& 32	& 5522	& 5532	& 1202	& 1201
+ 456.hmmer	  & 3200	& 3232	& 3	& 8	& 562	& 562	& 103	& 101
+ 458.sjeng	  & 400	& 446	& 	& 	& 127	& 135	& 18	& 18
+ 462.libquantum	  & 362	& 386	& 	& 	& 87	& 87	& 21	& 21
+ 464.h264ref	  & 17075	& 17153	& 4	& 4	& 338	& 382	& 72	& 72
+ 470.lbm	          & 39	& 39	& 	& 	& 13	& 13	& 4	& 4
+ 471.omnetpp	  & 565	& 661	& 32	& 30	& 4790	& 4843	& 1321	& 1316
+ 473.astar	  & 691	& 754	& 	& 	& 359	& 363	& 113	& 113
+ 482.sphinx3	  & 1599	& 1693	& 11	& 9	& 198	& 194	& 53	& 53
+ 483.xalancbmk	  & 9997	& 11453	& 56	& 54	& 69534	& 68839	& 29875	& 29519
+
+    \end{tabular}
+  \end{center}
+  \caption{Number of spills on SPEC2006 at -Ofast}
+  \label{tab:code-size}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|}
+      \hline
+ code-motion metric &  I hoisted	I removed	loads hoisted	loads removed	calls hoisted	I sunk	stores hoisted	stores removed
+ 400.perlbench	  &  2387	& 2456	& 1493	& 1560	& 1	& 4	& 38	& 38
+ 401.bzip2	  &  176	& 300	& 149	& 273	& 	& 	& 	& 
+ 403.gcc	          &  5670	& 5761	& 3109	& 3191	& 	& 1	& 3	& 3
+ 429.mcf	          &  42	& 43	& 37	& 38	& 	& 	& 	& 
+ 433.milc	  &  128	& 130	& 117	& 119	& 1	& 	& 	& 
+ 444.namd	  &  262	& 264	& 148	& 150	& 1	& 	& 	& 
+ 445.gobmk	  &  1021	& 1052	& 705	& 734	& 1	& 1	& 3	& 3
+ 447.dealII	  &  10455	& 12325	& 6394	& 8210	& 32	& 	& 20	& 20
+ 450.soplex	  &  2389	& 2731	& 1437	& 1702	& 11	& 	& 14	& 14
+ 453.povray	  &  1967	& 2269	& 1389	& 1661	& 14	& 	& 4	& 4
+ 456.hmmer	  &  408	& 427	& 319	& 338	& 6	& 	& 1	& 1
+ 458.sjeng	  &  372	& 399	& 234	& 251	& 9	& 1	& 1	& 1
+ 462.libquantum	  &  49	& 49	& 32	& 32	& 	& 	& 	& 
+ 464.h264ref	  &  2953	& 3438	& 2238	& 2656	& 1	& 1	& 2	& 2
+ 470.lbm	          &  23	& 23	& 19	& 19	& 	& 	& 	& 
+ 471.omnetpp	  &  391	& 408	& 297	& 314	& 2	& 	& 	& 
+ 473.astar	  &  202	& 222	& 140	& 158	& 	& 	& 	& 
+ 482.sphinx3	  &  168	& 172	& 136	& 140	& 	& 	& 	& 
+ 483.xalancbmk	  &  5894	& 6162	& 3301	& 3495	& 3	& 2	& 4	& 4
+        &         &         &         &         &         &         & 
+    \end{tabular}
+  \end{center}
+  \caption{Number of spills on SPEC2006 at -Ofast}
+  \label{tab:code-size}
+\end{table}
+
 While benchmarking \LLVMTestSuite{} we see both increase as well as decrease in

### 2017-04-17 acmart-pacmpl-template.tex
format tables

--- 2017-04-17 acmart-pacmpl-template.tex
@@
 Number of spills  &base(-Ofast) &code-motion(-Ofast) &\%loss \\\hline
-400.perlbench	  & 2539	& 2493	& 0.98
-401.bzip2	  & 718	        & 707	& 0.98
-403.gcc	          & 5782	& 5722	& 0.99
-429.mcf	          & 14	        & 17	& 1.21
-433.milc	  & 635	        & 642	& 1.01
-444.namd	  & 3223	& 3274	& 1.01
-445.gobmk	  & 2166	& 2173	& 1.00
-447.dealII	  & 11074	& 10234	& 0.92
-450.soplex	  & 1122	& 1125	& 1.00
-453.povray	  & 4701	& 4692	& 1.09
-456.hmmer	  & 1197	& 1274	& 1.06
-458.sjeng	  & 168	        & 176	& 1.05
-462.libquantum	  & 118	        & 118	& 1
-464.h264ref	  & 3379	& 3454	& 1.02
-470.lbm	          & 33	        & 33	& 1
-471.omnetpp	  & 524	        & 527	& 1.00
-473.astar	  & 180	        & 201	& 1.12
-482.sphinx3	  & 674	        & 670	& 0.99
-483.xalancbmk	  & 5102	& 4965	& 0.97
-Grand Total	  & 43349	& 42497	& 0.98
+400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
+401.bzip2	  & 718	        & 707	& 0.98 \\\hline
+403.gcc	          & 5782	& 5722	& 0.99 \\\hline
+429.mcf	          & 14	        & 17	& 1.21 \\\hline
+433.milc	  & 635	        & 642	& 1.01 \\\hline
+444.namd	  & 3223	& 3274	& 1.01 \\\hline
+445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
+447.dealII	  & 11074	& 10234	& 0.92 \\\hline
+450.soplex	  & 1122	& 1125	& 1.00 \\\hline
+453.povray	  & 4701	& 4692	& 1.09 \\\hline
+456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
+458.sjeng	  & 168	        & 176	& 1.05 \\\hline
+462.libquantum	  & 118	        & 118	& 1    \\\hline
+464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
+470.lbm	          & 33	        & 33	& 1    \\\hline
+471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
+473.astar	  & 180	        & 201	& 1.12 \\\hline
+482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
+483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
+Grand Total	  & 43349	& 42497	& 0.98 \\\hline
     \end{tabular}
@@
       \hline
-
+% TODO: Check if putting code-size or spills number is okay or only %age is required.
 Code size         & base(ofast)	& ofast+gvnhoist&	increase in codesize \\\hline
-400.perlbench	  &  2469594	& 2478714	& 1.00\\\hline
+400.perlbench	  &  2469594	& 2478714	& 1.00 \\\hline
 401.bzip2	  &  865902	& 867502	& 1.00 \\\hline
@@
 
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+      \hline
+ benchmark &      hoist licm	&hoist  &sunk licm &sunk&finline& i     &f deleted & f \\\hline
+ benchmark        & base        & gvn	& base	& gvn	& base	& gvn	& base	& gvn  \\\hline
+ 400.perlbench	  & 2428	& 2480	& 36	& 35	& 1562	& 1565	& 232	& 234  \\\hline
+ 401.bzip2	  & 721	        & 825	& 3	& 3	& 172	& 180	& 32	& 33   \\\hline
+ 403.gcc	  & 8867	& 10079	& 143	& 182	& 5034	& 5080	& 1303	& 130  \\\hline
+ 429.mcf	  & 55	        & 51	& 1	& 1	& 4	& 4	& 	&      \\\hline
+ 433.milc	  & 2537	& 2795	& 3	& 3	& 75	& 81	& 	&      \\\hline
+ 444.namd	  & 2204	& 2206	& 240	& 240	& 387	& 390	& 57	& 55   \\\hline
+ 445.gobmk	  & 3090	& 3107	& 18	& 19	& 735	& 775	& 230	& 234  \\\hline
+ 447.dealII	  & 59114	& 95884	& 60	& 8	& 100891& 100692& 30219	& 29966\\\hline
+ 450.soplex	  & 4668	& 4848	& 18	& 17	& 9801	& 9784	& 3125	& 3112 \\\hline
+ 453.povray	  & 9533	& 9079	& 23	& 32	& 5522	& 5532	& 1202	& 1201 \\\hline
+ 456.hmmer	  & 3200	& 3232	& 3	& 8	& 562	& 562	& 103	& 101  \\\hline
+ 458.sjeng	  & 400	        & 446	& 	& 	& 127	& 135	& 18	& 18   \\\hline
+ 462.libquantum	  & 362	        & 386	& 	& 	& 87	& 87	& 21	& 21   \\\hline
+ 464.h264ref	  & 17075	& 17153	& 4	& 4	& 338	& 382	& 72	& 72   \\\hline
+ 470.lbm	  & 39	        & 39	& 	& 	& 13	& 13	& 4	& 4    \\\hline
+ 471.omnetpp	  & 565	        & 661	& 32	& 30	& 4790	& 4843	& 1321	& 1316 \\\hline
+ 473.astar	  & 691	        & 754	& 	& 	& 359	& 363	& 113	& 113  \\\hline
+ 482.sphinx3	  & 1599	& 1693	& 11	& 9	& 198	& 194	& 53	& 53   \\\hline
+ 483.xalancbmk	  & 9997	& 11453	& 56	& 54	& 69534	& 68839	& 29875	& 29519\\\hline
+
+    \end{tabular}
+  \end{center}
+  \caption{Number of spills on SPEC2006 at -Ofast}
+  \label{tab:code-size}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+      \hline
+ code-motion metric &  I hoisted&I removed	&loads hoisted	&loads removed	&calls hoisted	&I sunk	&stores hoisted	&stores removed \\\hline
+ 400.perlbench	  &  2387	& 2456	& 1493	& 1560	& 1	& 4	& 38	& 38 \\\hline
+ 401.bzip2	  &  176	& 300	& 149	& 273	& 	& 	& 	&    \\\hline
+ 403.gcc	  &  5670	& 5761	& 3109	& 3191	& 	& 1	& 3	& 3  \\\hline
+ 429.mcf	  &  42	        & 43	& 37	& 38	& 	& 	& 	&    \\\hline
+ 433.milc	  &  128	& 130	& 117	& 119	& 1	& 	& 	&    \\\hline
+ 444.namd	  &  262	& 264	& 148	& 150	& 1	& 	& 	&    \\\hline
+ 445.gobmk	  &  1021	& 1052	& 705	& 734	& 1	& 1	& 3	& 3  \\\hline
+ 447.dealII	  &  10455	& 12325	& 6394	& 8210	& 32	& 	& 20	& 20 \\\hline
+ 450.soplex	  &  2389	& 2731	& 1437	& 1702	& 11	& 	& 14	& 14 \\\hline
+ 453.povray	  &  1967	& 2269	& 1389	& 1661	& 14	& 	& 4	& 4  \\\hline
+ 456.hmmer	  &  408	& 427	& 319	& 338	& 6	& 	& 1	& 1  \\\hline
+ 458.sjeng	  &  372	& 399	& 234	& 251	& 9	& 1	& 1	& 1  \\\hline
+ 462.libquantum	  &  49	        & 49	& 32	& 32	& 	& 	& 	&    \\\hline
+ 464.h264ref	  &  2953	& 3438	& 2238	& 2656	& 1	& 1	& 2	& 2  \\\hline
+ 470.lbm	  &  23	        & 23	& 19	& 19	& 	& 	& 	&    \\\hline
+ 471.omnetpp	  &  391	& 408	& 297	& 314	& 2	& 	& 	&    \\\hline
+ 473.astar	  &  202	& 222	& 140	& 158	& 	& 	& 	&    \\\hline
+ 482.sphinx3	  &  168	& 172	& 136	& 140	& 	& 	& 	&    \\\hline
+ 483.xalancbmk	  &  5894	& 6162	& 3301	& 3495	& 3	& 2	& 4	& 4  \\\hline
+    \end{tabular}
+  \end{center}
+  \caption{Number of spills on SPEC2006 at -Ofast}
+  \label{tab:code-size}
+\end{table}
+
 While benchmarking \LLVMTestSuite{} we see both increase as well as decrease in

### 2017-04-17 acmart-pacmpl-template.tex
tables

--- 2017-04-17 acmart-pacmpl-template.tex
@@
 
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{|l|c|}
-      \hline
-      Metric               & Number\\\hline
-      Scalars hoisted      & 6791  \\\hline
-      Scalars removed      & 9696  \\\hline
-      Loads hoisted        & 14802 \\\hline
-      Loads removed        & 20719 \\\hline
-      Stores hoisted       & 15    \\\hline
-      Stores removed       & 15    \\\hline
-      Calls hoisted        & 8     \\\hline
-      Calls removed        & 8     \\\hline
-      Total Instructions hoisted & 21616 \\\hline
-      Total Instructions removed & 30438 \\\hline
-\end{tabular}
-  \end{center}
-  \caption{Code hoisting metrics on \LLVMTestSuite{}}
-  \label{tab:hoist-results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{|l|c|c|}
-      \hline
-      Metric               & Before & After              \\\hline
-      Functions deleted (all callers found)       & 38250   & 38255  \\\hline
-      Functions inlined                           & 154986  & 154985 \\\hline
-      Spills inserted                             & 50724   & 50784  \\\hline
-      LICM
-\end{tabular}
-  \end{center}
-  \caption{Static metrics before and after code-hoisting on \LLVMTestSuite{}}
-  \label{tab:static-results}
-\end{table}
-
 Other static metrics are listed in Table~\ref{tab:static-results}. Here we can
@@
   \begin{center}
-    \begin{tabular}{|l|c|}
+    \begin{tabular}{|l|r|r|r|}
       \hline
-      Code-size metric  (.text)                   & Number   \\\hline
-      Total  benchmarks                           & 497      \\\hline
-      Total  gained in size                       & 39       \\\hline
-      Total  decrease in size                     & 58       \\\hline
-      Median decrease in size                     & 2.9\%    \\\hline
-      Median increase in size                     & 2.4\%    \\\hline
+Number of spills  &base(-Ofast) &code-motion(-Ofast) &\%change \\\hline
+400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
+401.bzip2	  & 718	        & 707	& 0.98 \\\hline
+403.gcc	          & 5782	& 5722	& 0.99 \\\hline
+429.mcf	          & 14	        & 17	& 1.21 \\\hline
+433.milc	  & 635	        & 642	& 1.01 \\\hline
+444.namd	  & 3223	& 3274	& 1.01 \\\hline
+445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
+447.dealII	  & 11074	& 10234	& 0.92 \\\hline
+450.soplex	  & 1122	& 1125	& 1.00 \\\hline
+453.povray	  & 4701	& 4692	& 1.09 \\\hline
+456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
+458.sjeng	  & 168	        & 176	& 1.05 \\\hline
+462.libquantum	  & 118	        & 118	& 1.00 \\\hline
+464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
+470.lbm	          & 33	        & 33	& 1.00 \\\hline
+471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
+473.astar	  & 180	        & 201	& 1.12 \\\hline
+482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
+483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
+Grand Total	  & 43349	& 42497	& 0.98 \\\hline
     \end{tabular}
   \end{center}
-  \caption{Code size metrics on \LLVMTestSuite{}}
-  \label{tab:code-size}
+  \caption{Number of spills on before and after code-motion on SPEC2006 at -Ofast}
+  \label{tab:spills}
 \end{table}
@@
       \hline
-Number of spills  &base(-Ofast) &code-motion(-Ofast) &\%loss \\\hline
-400.perlbench	  & 2539	& 2493	& 0.98
-401.bzip2	  & 718	        & 707	& 0.98
-403.gcc	          & 5782	& 5722	& 0.99
-429.mcf	          & 14	        & 17	& 1.21
-433.milc	  & 635	        & 642	& 1.01
-444.namd	  & 3223	& 3274	& 1.01
-445.gobmk	  & 2166	& 2173	& 1.00
-447.dealII	  & 11074	& 10234	& 0.92
-450.soplex	  & 1122	& 1125	& 1.00
-453.povray	  & 4701	& 4692	& 1.09
-456.hmmer	  & 1197	& 1274	& 1.06
-458.sjeng	  & 168	        & 176	& 1.05
-462.libquantum	  & 118	        & 118	& 1
-464.h264ref	  & 3379	& 3454	& 1.02
-470.lbm	          & 33	        & 33	& 1
-471.omnetpp	  & 524	        & 527	& 1.00
-473.astar	  & 180	        & 201	& 1.12
-482.sphinx3	  & 674	        & 670	& 0.99
-483.xalancbmk	  & 5102	& 4965	& 0.97
-Grand Total	  & 43349	& 42497	& 0.98
+% TODO: Check if putting code-size or spills number is okay or only %age is required.
+% llvm trunk vs. llvm trunk with code-motion
+Code size         &\% age	increase in codesize \\\hline
+400.perlbench	  & 1.00 \\\hline
+401.bzip2	  & 1.00 \\\hline
+403.gcc	          & 1.01 \\\hline
+429.mcf	          & 1.00 \\\hline
+433.milc	  & 1.00 \\\hline
+444.namd	  & 1.01 \\\hline
+445.gobmk	  & 1.01 \\\hline
+447.dealII	  & 1.01 \\\hline
+450.soplex	  & 0.99 \\\hline
+453.povray	  & 1.00 \\\hline
+456.hmmer	  & 1.00 \\\hline
+458.sjeng	  & 1.00 \\\hline
+462.libquantum	  & 1.00 \\\hline
+464.h264ref	  & 1.00 \\\hline
+470.lbm	          & 1.00 \\\hline
+471.omnetpp	  & 0.99 \\\hline
+473.astar	  & 1.02 \\\hline
+482.sphinx3	  & 1.00 \\\hline
+483.xalancbmk	  & 0.99 \\\hline
     \end{tabular}
   \end{center}
-  \caption{Number of spills on SPEC2006 at -Ofast}
+  \caption{Change in code size with and without code-motion on SPEC2006 at -Ofast}
   \label{tab:code-size}
@@
   \begin{center}
-    \begin{tabular}{|l|c|c|c|}
+    \begin{tabular}{|l|r|r|r|r|r|r|r|r|}
       \hline
-
-Code size         & base(ofast)	& ofast+gvnhoist&	increase in codesize \\\hline
-400.perlbench	  &  2469594	& 2478714	& 1.00\\\hline
-401.bzip2	  &  865902	& 867502	& 1.00 \\\hline
-403.gcc	          &  4379291	& 4437467	& 1.01 \\\hline
-429.mcf	          &  794909	& 794749	& 1.00 \\\hline
-433.milc	  &  1009553	& 1010209	& 1.00 \\\hline
-444.namd	  &  282794	& 286658	& 1.01 \\\hline
-445.gobmk	  &  2474193	& 2494673	& 1.01 \\\hline
-447.dealII	  &  3577836	& 3605672	& 1.01 \\\hline
-450.soplex	  &  411254	& 409034	& 0.99 \\\hline
-453.povray	  &  1114405	& 1114241	& 1.00 \\\hline
-456.hmmer	  &  1421823	& 1425287	& 1.00 \\\hline
-458.sjeng	  &  941066	& 940626	& 1.00 \\\hline
-462.libquantum	  &  911028	& 911124	& 1.00 \\\hline
-464.h264ref	  &  1677417	& 1680841	& 1.00 \\\hline
-470.lbm	          &  795444	& 795444	& 1    \\\hline
-471.omnetpp	  &  661099	& 658411	& 0.99 \\\hline
-473.astar	  &  44296	& 45224		& 1.02 \\\hline
-482.sphinx3	  &  1337445	& 1334437	& 1.00 \\\hline
-483.xalancbmk	  &  4293600	& 4247108	& 0.99 \\\hline
+  &  \multicolumn{2}{|c|}{Hoisted by LICM} &\multicolumn{2}{|c|}{Sunk by LICM} &\multicolumn{2}{|c|}{Functions Inlined} &  \multicolumn{2}{|c|}{Functions Deleted} \\\hline
+ Benchmark        & trunk       & GCM	& trunk	& GCM	& trunk	& GCM	& trunk	& GCM  \\\hline
+ 400.perlbench	  & 2428	& 2480	& 36	& 35	& 1562	& 1565	& 232	& 234  \\\hline
+ 401.bzip2	  & 721	        & 825	& 3	& 3	& 172	& 180	& 32	& 33   \\\hline
+ 403.gcc	  & 8867	& 10079	& 143	& 182	& 5034	& 5080	& 1303	& 130  \\\hline
+ 429.mcf	  & 55	        & 51	& 1	& 1	& 4	& 4	& 0 	& 0    \\\hline
+ 433.milc	  & 2537	& 2795	& 3	& 3	& 75	& 81	& 0	& 0    \\\hline
+ 444.namd	  & 2204	& 2206	& 240	& 240	& 387	& 390	& 57	& 55   \\\hline
+ 445.gobmk	  & 3090	& 3107	& 18	& 19	& 735	& 775	& 230	& 234  \\\hline
+ 447.dealII	  & 59114	& 95884	& 60	& 8	& 100891& 100692& 30219	& 29966\\\hline
+ 450.soplex	  & 4668	& 4848	& 18	& 17	& 9801	& 9784	& 3125	& 3112 \\\hline
+ 453.povray	  & 9533	& 9079	& 23	& 32	& 5522	& 5532	& 1202	& 1201 \\\hline
+ 456.hmmer	  & 3200	& 3232	& 3	& 8	& 562	& 562	& 103	& 101  \\\hline
+ 458.sjeng	  & 400	        & 446	& 0	& 0	& 127	& 135	& 18	& 18   \\\hline
+ 462.libquantum	  & 362	        & 386	& 0 	& 0	& 87	& 87	& 21	& 21   \\\hline
+ 464.h264ref	  & 17075	& 17153	& 4	& 4	& 338	& 382	& 72	& 72   \\\hline
+ 470.lbm	  & 39	        & 39	& 0	& 0	& 13	& 13	& 4	& 4    \\\hline
+ 471.omnetpp	  & 565	        & 661	& 32	& 30	& 4790	& 4843	& 1321	& 1316 \\\hline
+ 473.astar	  & 691	        & 754	& 0	& 0	& 359	& 363	& 113	& 113  \\\hline
+ 482.sphinx3	  & 1599	& 1693	& 11	& 9	& 198	& 194	& 53	& 53   \\\hline
+ 483.xalancbmk	  & 9997	& 11453	& 56	& 54	& 69534	& 68839	& 29875	& 29519\\\hline
     \end{tabular}
   \end{center}
-  \caption{Number of spills on SPEC2006 at -Ofast}
-  \label{tab:code-size}
+  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast}
+  \label{tab:stats}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|r|r|r|r|r|r|r|r|}
+      \hline
+ code-motion      & Instructions&Instructions   & Loads	        &Loads	        &Stores &Stores &Calls &Instructions \\\hline
+ metric           &  hoisted    &removed	&hoisted        &removed        &hoisted&removed&hoisted & sunk     \\\hline
+ 400.perlbench	  &  2387	& 2456		& 1493		& 1560		& 38	& 38	& 1	& 4	       \\\hline
+ 401.bzip2	  &  176	& 300		& 149		& 273		& 0	& 0	& 0 	& 0		\\\hline
+ 403.gcc	  &  5670	& 5761		& 3109		& 3191		& 3	& 3	& 0	& 1	       \\\hline
+ 429.mcf	  &  42	        & 43		& 37		& 38		& 0	& 0	& 0 	& 0		\\\hline
+ 433.milc	  &  128	& 130		& 117		& 119		& 0	& 0	& 1	& 0		\\\hline
+ 444.namd	  &  262	& 264		& 148		& 150		& 0	& 0	& 1	& 0		\\\hline
+ 445.gobmk	  &  1021	& 1052		& 705		& 734		& 3	& 3	& 1	& 1	       \\\hline
+ 447.dealII	  &  10455	& 12325		& 6394		& 8210		& 20	& 20	& 32	& 0	       \\\hline
+ 450.soplex	  &  2389	& 2731		& 1437		& 1702		& 14	& 14	& 11	& 0	       \\\hline
+ 453.povray	  &  1967	& 2269		& 1389		& 1661		& 4	& 4	& 14	& 0	       \\\hline
+ 456.hmmer	  &  408	& 427		& 319		& 338		& 1	& 1	& 6	& 0	       \\\hline
+ 458.sjeng	  &  372	& 399		& 234		& 251		& 1	& 1	& 9	& 1	       \\\hline
+ 462.libquantum	  &  49	        & 49		& 32		& 32		& 0	& 0	& 0	& 0		\\\hline
+ 464.h264ref	  &  2953	& 3438		& 2238		& 2656		& 2	& 2	& 1	& 1	       \\\hline
+ 470.lbm	  &  23	        & 23		& 19		& 19		& 0	& 0	& 0	& 0		\\\hline
+ 471.omnetpp	  &  391	& 408		& 297		& 314		& 0	& 0	& 2	& 0		\\\hline
+ 473.astar	  &  202	& 222		& 140		& 158		& 0	& 0	& 0	& 0		\\\hline
+ 482.sphinx3	  &  168	& 172		& 136		& 140		& 0	& 0	& 0	& 0		\\\hline
+ 483.xalancbmk	  &  5894	& 6162		& 3301		& 3495		& 4	& 4	& 3	& 2	       \\\hline
+    \end{tabular}
+  \end{center}
+  \caption{Code motion stats on SPEC2006 at -Ofast}
+  \label{tab:code-motion-metric}
 \end{table}

### 2017-04-17 acmart-pacmpl-template.tex
fixes

@@
 \def \LLVMTestSuite {\LLVM{} test-suite}
+\def \SPEC {SPEC Cpu2006}
 
@@
 \begin{verbatim}
+// Return true if the merge set of source node of
+// a visited J-edge (incoming edge of lnode) is not
+// the subset of the merge set of lnode.
+bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet M)
+   for (all incoming edges to lnode) do
+     Let e = Incoming edge
+     if (e is in JE && e in V) then
+       Let snode = Source Node of e
+       if (M(snode) is not a subset of M(lnode)) then
+         return true
+       end if
+     end if
+   end for
+return false
+
 // Compute merge set top-down in breadth first order.
@@
 
-\begin{verbatim}
-// Return true if the merge set of source node of
-// a visited J-edge (incoming edge of lnode) is not
-// the subset of the merge set of lnode.
-bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet M)
-   for (all incoming edges to lnode) do
-     Let e = Incoming edge
-     if (e is in JE && e in V) then
-       Let snode = Source Node of e
-       if (M(snode)!(Subset) M(lnode)) then
-         return true
-       end if
-     end if
-   end for
-return false
-\end{verbatim}
-
 We compute merge sets of the control flow graph which remains same through the
@@
 % Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
-We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146). All
+We ran \SPEC{} with the patch. All
 the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
 optimization level.  The results for code-hoisting are listed in
-Table~\ref{tab:hoist-results}. The table lists the number of scalars, loads,
+Table~\ref{tab:code-motion-metric}. The table lists the number of scalars, loads,
 stores and calls hoisted as well as removed. For each category, the number of
@@
 
-Other static metrics are listed in Table~\ref{tab:static-results}. Here we can
+Other static metrics are listed in Table~\ref{tab:stats} and in Table~\ref{tab:code-motion-metric}.
+Here we can
 see that except for rematerializing defs for splitting, which has an overhead of ~2\%, all

### 2017-04-17 acmart-pacmpl-template.tex
macro fixes

@@
 \def \LLVM {LLVM}
-\def \SESE {SESE}
 \def \CFG {CFG}
@@
 \def \GVN {GVN}
-\def \LLVMTestSuite {\LLVM{} test-suite}
 \def \SPEC {SPEC Cpu2006}
@@
 motion to remove more redundancies. The cost model to drive the code motion is
-based on live range analysis on SSA representation such that the (virtual)
+based on live range analysis on \SSA{} representation such that the (virtual)
 register pressure does not increase.
@@
 infrastructure available in \LLVM{}. The experimental results show an average
-of 2.5\% savings in code size in llvm testsuite, although the code size also
+of 2.5\% savings in code size on \SPEC{}, although the code size also
 increases in many cases because it enables more inlining. We have also seen
-imrpovements of approximately 5\% in a couple of SPEC 2006 benchmarks viz. gcc
-and mcf, moreover, register spills reduced for almost all the SPEC 2006
+execution time improvements of approximately 5\% in a couple of \SPEC{} benchmarks viz. \GCC{}
+and mcf, moreover, register spills reduced for almost all the \SPEC{}
 bencmarks when compiled for X86\_64. This is an optimistic algorithm in the
@@
 %% comma separated list
-\keywords{Optimizng Compilers, GCC, LLVM, Code Generation, Global Scheduling}  %% \keywords is optional
+\keywords{Optimizng Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional
 
@@
 identical computation in as much detail as we have done. A part of our
-implementation (aggressive code hoisting) is already merged in LLVM trunk,
+implementation (aggressive code hoisting) is already merged in \LLVM{} trunk,
 however, a general implementation of code-motion of congruent (but not
-redundant) expressions is still missing from GCC and LLVM trunk. The main
+redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
 contributions of this paper are:
@@
 \item experimental evaluation of our implementation in \LLVM{} which combines
-  SSA based liveness analysis, and ranking of expressions to move very busy
+  \SSA{} based liveness analysis, and ranking of expressions to move very busy
   expressions in order to reduce code-size (and improve performance in some
@@
 Shobaki not so much, of large compile times which is not feasible for industrial
-compilers like gcc and LLVM.
+compilers like \GCC{} and \LLVM{}.
 
@@
 require any bitvectors to be maintained for each basic block. The underlying
-simplicity of liveness computation is due to SSA form, where the the values only
+simplicity of liveness computation is due to \SSA{} form, where the the values only
 flow (from def to use) either through dominator edges or the join edges (where
@@
 program and classifying them by their value numbers. We could compute available
-and anticipable sets as computed by GCC's code-hoisting but that would be a lot
-of data structures to maintain at each basic block level. For GCC it makes sense
-because their code-hoisting is integrated with GVN-PRE which already has those
+and anticipable sets as computed by \GCC{'}s code-hoisting but that would be a lot
+of data structures to maintain at each basic block level. For \GCC{} it makes sense
+because their code-hoisting is integrated with \GVN{}-\PRE{} which already has those
 data structures available.
@@
 For stores, we check the dependency requirements similar to the hoisting of
-loads using the MemorySSA of \LLVM{}. We check that the operands of the store
+loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
 instruction are available at the hoisting point, and that there are no aliasing
@@
 
-Moreover, LLVM has a `getelementptr' instruction which computes the address
+Moreover, \LLVM{} has a `getelementptr' instruction which computes the address
 where a load or a store would happen. It is a scalar computation and gets
@@
 when the loads and stores cannot be moved.  This restriction is only to mitigate
-the limitations of LLVM's register allocator and may be lifted in the future,
+the limitations of \LLVM{'}s register allocator and may be lifted in the future,
 when the register allocation rematerialization pass has been improved to catch
@@
 }
-\caption{CFG to illustrate partitioning}
+\caption{\CFG{} to illustrate partitioning}
 \label{fig:dfsin}
@@
 intermediate representation (IR) to reflect the changes. At the same time
-MemorySSA is also updated to get up-to-date information about memory references.
+\MemorySSA{} is also updated to get up-to-date information about memory references.
 
@@
 
-\section{SSA based global code motion of identical computations}
+\section{\SSA{} based global code motion of identical computations}
 \label{sec:implementation-details}
@@
 Barriers are based on the concept of pinned instructions \cite{click1995global}
-but extended to adapt to LLVM IR. Since a basic block in LLVM IR is actually an
+but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
 extended basic block because there might be calls in the middle of a basic block
@@
 iterate on lower ranking expressions first and then proceed to higher ranking
-expressions in the same iteration but LLVM's GVN infrastucture does not compute
+expressions in the same iteration but \LLVM{'}s \GVN{} infrastucture does not compute
 equivalence classes in a effective way. We found it simpler to just recompute
@@
 on-demand for hoistable candidates so it does not affect the overall compile
-time by much. Other analyses like Alias Analysis, Memory SSA and Dominator Tree
-are already available in the LLVM pass pipeline. Although we recompute GVN, and
+time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
+are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
 Barriers for each iteration of the code-motion, we have not seen signinficant
@@
 
-While benchmarking \LLVMTestSuite{} we see both increase as well as decrease in
+While benchmarking \SPEC{} we see both increase as well as decrease in
 the codesizes of the final binaries. Since the pass runs early, it affects many

### 2017-04-17 paper.tex
less sinks

--- 2017-04-17 acmart-pacmpl-template.tex
@@
 \LLVM{} pass pipeline. Passes like mem2reg, instcombine might actually remove
-those loads so this order may change should this pass be scheduled later.
+those loads so this order may change should this pass be scheduled later. The number
+of instructions sunk is very less because \LLVM{} has a code sinking pass inside SimplifyCFG
+which is run before the global code-motion and sinks some instructions. Another reason is
+that instructions for which dependencies are not directly available in the successor is
+not sunk for now. Sinking those instructions would require a look ahead on the sinkability
+of the dependencies. We plan to implement this in near future.
 
@@
   \begin{center}
-    \begin{tabular}{|l|r|r|r|r|r|r|r|r|}
+    \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|}
       \hline
- code-motion      & Instructions&Instructions   & Loads	        &Loads	        &Stores &Stores &Calls &Instructions \\\hline
- metric           &  hoisted    &removed	&hoisted        &removed        &hoisted&removed&hoisted & sunk     \\\hline
- 400.perlbench	  &  2387	& 2456		& 1493		& 1560		& 38	& 38	& 1	& 4	       \\\hline
- 401.bzip2	  &  176	& 300		& 149		& 273		& 0	& 0	& 0 	& 0		\\\hline
- 403.gcc	  &  5670	& 5761		& 3109		& 3191		& 3	& 3	& 0	& 1	       \\\hline
- 429.mcf	  &  42	        & 43		& 37		& 38		& 0	& 0	& 0 	& 0		\\\hline
- 433.milc	  &  128	& 130		& 117		& 119		& 0	& 0	& 1	& 0		\\\hline
- 444.namd	  &  262	& 264		& 148		& 150		& 0	& 0	& 1	& 0		\\\hline
- 445.gobmk	  &  1021	& 1052		& 705		& 734		& 3	& 3	& 1	& 1	       \\\hline
- 447.dealII	  &  10455	& 12325		& 6394		& 8210		& 20	& 20	& 32	& 0	       \\\hline
- 450.soplex	  &  2389	& 2731		& 1437		& 1702		& 14	& 14	& 11	& 0	       \\\hline
- 453.povray	  &  1967	& 2269		& 1389		& 1661		& 4	& 4	& 14	& 0	       \\\hline
- 456.hmmer	  &  408	& 427		& 319		& 338		& 1	& 1	& 6	& 0	       \\\hline
- 458.sjeng	  &  372	& 399		& 234		& 251		& 1	& 1	& 9	& 1	       \\\hline
- 462.libquantum	  &  49	        & 49		& 32		& 32		& 0	& 0	& 0	& 0		\\\hline
- 464.h264ref	  &  2953	& 3438		& 2238		& 2656		& 2	& 2	& 1	& 1	       \\\hline
- 470.lbm	  &  23	        & 23		& 19		& 19		& 0	& 0	& 0	& 0		\\\hline
- 471.omnetpp	  &  391	& 408		& 297		& 314		& 0	& 0	& 2	& 0		\\\hline
- 473.astar	  &  202	& 222		& 140		& 158		& 0	& 0	& 0	& 0		\\\hline
- 482.sphinx3	  &  168	& 172		& 136		& 140		& 0	& 0	& 0	& 0		\\\hline
- 483.xalancbmk	  &  5894	& 6162		& 3301		& 3495		& 4	& 4	& 3	& 2	       \\\hline
+ code-motion  &\multicolumn{2}{|c|}{Total Instructions} &\multicolumn{2}{|c|}{Scalars}  & \multicolumn{2}{|c|}{Loads} & \multicolumn{1}{|c|}{Stores} &Calls   &Instructions    \\\hline
+ metric           &  hoisted    &removed  &hoisted &removed&hoisted     &removed        &hoisted=removed&hoisted & sunk \\\hline
+ 400.perlbench	  &  2387	& 2456	  &  855  & 857	   & 1493	& 1560		& 38	& 1	& 4	        \\\hline
+ 401.bzip2	  &  176	& 300	  &  27	  & 27	   & 149	& 273		& 0	& 0 	& 0		\\\hline
+ 403.gcc	  &  5670	& 5761	  &  2558 & 2567   & 3109	& 3191		& 3	& 0	& 1	        \\\hline
+ 429.mcf	  &  42	        & 43	  &  5	  & 5	   & 37 	& 38		& 0	& 0 	& 0		\\\hline
+ 433.milc	  &  128	& 130	  &  10	  & 10	   & 117	& 119		& 0	& 1	& 0		\\\hline
+ 444.namd	  &  262	& 264	  &  113  & 113	   & 148	& 150		& 0	& 1	& 0		\\\hline
+ 445.gobmk	  &  1021	& 1052	  &  312  & 314	   & 705	& 734		& 3	& 1	& 1	        \\\hline
+ 447.dealII	  &  10455	& 12325	  &  4009 & 4063   & 6394	& 8210		& 20	& 32	& 0	        \\\hline
+ 450.soplex	  &  2389	& 2731	  &  927  & 1004   & 1437	& 1702		& 14	& 11	& 0	        \\\hline
+ 453.povray	  &  1967	& 2269	  &  560  & 590	   & 1389	& 1661		& 4	& 14	& 0	        \\\hline
+ 456.hmmer	  &  408	& 427	  &  82	  & 82	   & 319	& 338		& 1	& 6	& 0	        \\\hline
+ 458.sjeng	  &  372	& 399	  &  128  & 138	   & 234	& 251		& 1	& 9	& 1	        \\\hline
+ 462.libquantum	  &  49	        & 49	  &  17	  & 17	   & 32	      	& 32		& 0	& 0	& 0		\\\hline
+ 464.h264ref	  &  2953	& 3438	  &  712  & 779	   & 2238	& 2656		& 2	& 1	& 1	        \\\hline
+ 470.lbm	  &  23	        & 23	  &  4	  & 4	   & 19		& 19		& 0	& 0	& 0		\\\hline
+ 471.omnetpp	  &  391	& 408	  &  92	  & 92	   & 297	& 314		& 0	& 2	& 0		\\\hline
+ 473.astar	  &  202	& 222	  &  62	  & 64	   & 140	& 158		& 0	& 0	& 0		\\\hline
+ 482.sphinx3	  &  168	& 172	  &  32	  & 32	   & 136	& 140		& 0	& 0	& 0		\\\hline
+ 483.xalancbmk	  &  5894	& 6162	  &  2586 & 2660   & 3301	& 3495		& 4	& 3	& 2	        \\\hline
+ Grand total      &34957	&38631	  &13091  &13418   &21694	&25041	        &90	&82     &10             \\\hline
     \end{tabular}

### 2017-04-17 paper.tex
spills
rebase

@@
 \LLVM{} pass pipeline. Passes like mem2reg, instcombine might actually remove
-those loads so this order may change should this pass be scheduled later. The number
-of instructions sunk is very less because \LLVM{} has a code sinking pass inside SimplifyCFG
-which is run before the global code-motion and sinks some instructions. Another reason is
-that instructions for which dependencies are not directly available in the successor is
-not sunk for now. Sinking those instructions would require a look ahead on the sinkability
-of the dependencies. We plan to implement this in near future.
+those loads so this order may change should this pass be scheduled later.
 

### 2017-04-17 paper.tex
comments

--- 2017-04-17 paper.tex
@@
 
+Spills are listed in Table~\cite{tab:spills}. Spills were reduced by an average of 2\% on the
+\SPEC{} benchmark suite. This is a result of cost-model which limits the code-motion to those
+candidates which do not increase the register pressure (except for loads Section~\cite{sec:cost-models},
+overall spills still decreases.
+
 \begin{table}[h!]
@@
 483.xalancbmk	  & 0.99 \\\hline
+Grand total       & 1.00 \\\hline
     \end{tabular}
@@
 \end{table}
+
+Code size remained almost the same as a result of code-motion as shown in Table~\cite{tab:code-size}.
+Some bencmarks like 473.astar increased in code size by almost 2\%. This is because once the code-size
+of a function decreases it becomes cheaper to inline. As we can see in Table-\cite{tab:stats}, four
+more functions got inlined in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
+all other benchmarks got more functions inlined.
 
@@
 \end{table}
+
+Some useful statc metric are presented in Table~\cite{tab:stats}, which shows how code-motion
+impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
+code out of the loop. It may hoist the code to the loop's preheader or sink the code to loop's
+post-dominator \cite{steven1997advanced}. As we can see number of instructions hoisted increased
+significantly in several benchmarks, although number of instructions sunk did not change by much.
+Number of inlined functiosn also improved for most benchmarks. In general when more functions
+were inlined, more functions were deleted as in 400.perlbench, 445.gobmk.
 

### 2017-04-17 paper.tex
fixes to section 1

@@
 computations. Code hoisting can reduce the critical path length of execution in
-out of order machines. As more instructios are available at the hoisting point,
-the hardware has more instructions to reorder. Following example illustrates how
-hoisting can improve performance by exposing more ILP.
+out-of-order machines. As more instructions are available at the hoisting point,
+the hardware has more independent instructions to reorder. The following example
+illustrates how hoisting can improve performance by exposing more ILP.
 
@@
   inv = 1.0f / d;
-  if (inv >= 0) {
-    tmin = (min - a) * inv;
-    tmax = (max - a) * inv;
-  } else {
-    tmin = (max - a) * inv;
-    tmax = (min - a) * inv;
-  }
+  if (inv >= 0) { tmin = (min - a) * inv; tmax = (max - a) * inv; }
+  else          { tmin = (max - a) * inv; tmax = (min - a) * inv; }
   return tmax + tmin;
@@
 computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
-depends on inv which depends on a division operation which is generally more
-expensive than the addition, subtraction and multiplication operations. The
-total latency of computation across each branch is:
-$C_{div} + 2(C_{sub} + C_{mul})$
-Or, for out of order processors with two add units and two multiply units:
-$C_{div} + C_{sub} + C_{mul}$
+depend on inv which depends on a division operation which is generally more
+expensive than the addition, subtraction, and multiplication operations. The
+total latency of computation across each branch is: $C_{div} + 2(C_{sub} +
+C_{mul})$.  For an out-of-order processor with two add units and two multiply
+units, the total latency of computation is $C_{div} + C_{sub} + C_{mul}$.
 
@@
 float fun(float d, float min, float max, float a) {
-  float tmin, tmax, tmin1, tmax1, inv;
-
-  tmin1 = (min - a);
-  tmax1 = (max - a);
-
-  inv = 1.0f / d;
-  tmin1 = tmin1 * inv;
-  tmax1 = tmax1 * inv;
-
-  if (inv >= 0) {
-    tmin = tmin1;
-    tmax = tmax1;
-  } else {
-    tmin = tmax1;
-    tmax = tmin1;
-  }
-
+  float tmin, tmax;
+  float inv = 1.0f / d;
+  float x = (min - a) * inv;
+  float y = (max - a) * inv;
+  if (inv >= 0) { tmin = x; tmax = y; }
+  else          { tmin = y; tmax = x; }
   return tmax + tmin;

### 2017-04-17 paper.tex
code size

@@
 Code size remained almost the same as a result of code-motion as shown in Table~\cite{tab:code-size}.
+The code size listed here is the text size (from linux size command) of the final executables for each benchmark.
 Some bencmarks like 473.astar increased in code size by almost 2\%. This is because once the code-size
@@
 all other benchmarks got more functions inlined.
+
+\begin{table}[h]
+  \begin{center}
+    \begin{tabular}{|l|c|c|c|}
+      \hline
+  Benchmark       &   Baseline	        &  GCM & \%Incrase     \\\hline
+  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
+  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
+  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
+  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
+  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
+  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
+  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
+  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
+  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
+  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
+  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
+  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
+  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
+  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
+  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
+  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
+  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
+  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
+  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
+    \end{tabular}
+  \end{center}
+  \caption{Change in compilation time with and without code-motion on SPEC2006 at -Ofast}
+  \label{tab:compile-time}
+\end{table}
+
+The compilation time actually reduced for most of the benchmarks as a result of code motion.
+This is because as code-motion removes (congruent) instructions, the number of instructions
+to be processed by later compilation passes also reduces. The compilation time listed here
+is the total instruction count as outputted from valgrind (--tool=cachegrind).
 

### 2017-04-17 paper.tex
fixes to section 1

@@
 float fun(float d, float min, float max, float a) {
-  float tmin, tmax;
   float inv = 1.0f / d;
@@
   float y = (max - a) * inv;
-  if (inv >= 0) { tmin = x; tmax = y; }
-  else          { tmin = y; tmax = x; }
+  float tmin = x, tmax = y;
+  if (inv < 0) { tmin = y; tmax = x; }
   return tmax + tmin;
@@
 \begin{itemize}
-\item It helps reduce the code size of the program by replacing multiple
+\item it helps reduce the code size of the program by replacing multiple
   instructions with one although the final code size may be higher because of
   improved inlining heuristics; functions become cheaper to inline when their
-  code size decreases because inliner heuristics in \LLVM{} depends on
-  instruction count.
-\item It expose more instruction level parallelism to the later compiler
+  code size decreases because the inliner heuristics depend on instruction
+  count;
+\item it exposes more instruction level parallelism to the later compiler
   passes. By hoisting identical computations to be executed earlier, instruction
@@
   bubbles;
-\item It reduces branch misprediction penalty on out-of-order processors with
-  speculative execution of branches: by hoisting/sinking expressions out of
+\item it reduces branch misprediction penalty on out-of-order processors with
+  speculative execution of branches: by hoisting or sinking expressions out of
   branches, it can effectively reduce the amount of code to be speculatively
   executed and hence reduce the critical path;
-\item It reduces interference/register pressure when appropriate cost-model is
-  applied. We have used a cost model in our implementation which hoists/sinks only
-  when register pressure would potentially reduce.
-\item For SIMD architectures, which execute all branches, it will reduce the
-  total number of instructions to be executed.
-\item It may improve loop vectorization by reducing a loop with control flow to
+\item it reduces interference or register pressure with an appropriate
+  cost-model: we have used a cost model in our implementation which only hoists
+  or sinks when the register pressure is reduced;
+\item it reduces the total number of instructions to be executed for SIMD
+  architectures which execute all code in branches based on masking or
+  predication;
+\item it may improve loop vectorization by reducing a loop with control flow to
   a loop with a single BB, should all the instructions in a conditional get
   hoisted or sinked;
-\item It enables more loop invariant code motion (LICM): as LICM passes, in
+\item it enables more loop invariant code motion (LICM): as LICM passes, in
   general, cannot effectively reason about instructions within conditional
-  branches the context of loops, code-hoisting is needed to move instructions
+  branches in the context of loops, code-hoisting is needed to move instructions
   out of conditional expressions and expose them to LICM.
@@
 implementation (aggressive code hoisting) is already merged in \LLVM{} trunk,
-however, a general implementation of code-motion of congruent (but not
-redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
+however, a general implementation of code-motion of congruent (non redundant)
+expressions is still missing from \GCC{} and \LLVM{} trunk. The main
 contributions of this paper are:
 \begin{itemize}
-\item a new optimistic algorithm to move congruent instructiosn out of branches,
-\item a cost model to reduce interference and hence, reduce spills,
+\item a new optimistic algorithm to move congruent instructiosn out of branches;
+\item a cost model to reduce interference and hence, reduce spills;
 \item a technique to maximize hoisting in an optimistic approach by partitioning
-  the list of potential candidates sorted by their DFS visit number,
+  the list of potential candidates sorted by their DFS visit number;
 \item experimental evaluation of our implementation in \LLVM{} which combines
   \SSA{} based liveness analysis, and ranking of expressions to move very busy
-  expressions in order to reduce code-size (and improve performance in some
-  cases).
+  expressions in order to reduce code-size and improve performance in some
+  cases.
 \end{itemize}
@@
 
-\GCC{}ecently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
+\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
 part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value

### 2017-04-17 paper.tex
fixes section 2

@@
 \def \GVN {GVN}
-\def \SPEC {SPEC Cpu2006}
+\def \SPEC {SPEC Cpu 2006}
 
@@
 
-We got reduction in register spills on some SPEC2006 benchmarks, reduction in
-code-size and some performance improvements with very low compile time overhead.
+With the algorithm described in the next section, we got a
+reduction in register spills, a reduction in code-size and some performance
+improvements with very low compilation time overhead, on \SPEC{} benchmarks.
 

### 2017-04-17 paper.tex
fix section 3

@@
 \item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
-  Control Flow Graph (\CFG{}).
+  Control Flow Graph (\CFG{});
 \item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
   dominator tree with join-edges to keep track of data-flow in a program. We use
-  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012}.
+  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
 \item Static Single Assignment (\SSA{}) \cite{cytron};
@@
   identical are given the same number;
-\item Memory \SSA{} \cite{novillo2007memory}: it is a factored use-def chain of
-  memory operations that the compiler is able to prove are dependent.
+\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
+  memory operations that the compiler is unable to prove
+  independent. \MemorySSA{} accelerates the access to the alias analysis
+  information.
 \end{itemize}
@@
 
-\subsection{Liveness using DJ Graph}
-DJ-Graph \cite{Sreedhar1996} is a data structure that augments the dominator
-tree with join-edges to keep track of data-flow in a program. We use DJ-Graph
-and Merge Sets to compute liveness of variables as illustrated in
+\subsection{Liveness using DJ-graph}
+DJ-graph \cite{Sreedhar1996} is a data structure that augments the dominator
+tree with join-edges to keep track of data-flow in a program. We use DJ-graph
+and merge-sets to compute liveness of variables as illustrated in
 \cite{das2012}.  It is very efficient for computing liveness and does not
 require any bitvectors to be maintained for each basic block. The underlying
-simplicity of liveness computation is due to \SSA{} form, where the the values only
-flow (from def to use) either through dominator edges or the join edges (where
-we insert a PHI). The DJ-graph contains both these edges which allows for
-computation of merge sets for each basic block i.e., a set of all basic blocks
-where the values can flow from a particular basic block. We have implemented
-merge-set computation from DJ-graph as illustrated in \cite{das2005}. A simplified
-version of merge set that we implemented is presented here:
+simplicity of liveness computation is due to \SSA{} form, where the the values
+only flow (from def to use) either through dominator edges or the join edges
+(where we insert a PHI). The DJ-graph contains both these edges which allows for
+computation of merge-sets for each basic block, i.e., a set of all basic blocks
+where the values can flow from a particular basic block. We have implemented the
+merge-set computation based on a DJ-graph as illustrated in \cite{das2005}. A
+simplified version of the merge-set that we implemented is presented here:
 
 \begin{verbatim}
-// Return true if the merge set of source node of
-// a visited J-edge (incoming edge of lnode) is not
-// the subset of the merge set of lnode.
-bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet M)
-   for (all incoming edges to lnode) do
-     Let e = Incoming edge
-     if (e is in JE && e in V) then
-       Let snode = Source Node of e
-       if (M(snode) is not a subset of M(lnode)) then
+
+// Return true if the merge-set of source node of a visited J-edge (incoming edge of
+// lnode) is not the subset of the merge-set of lnode.
+bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet Merge)
+  for e in (all incoming edges to lnode)
+    if (e is in JE && e in V)
+      if (Merge(Source Node of e) is not a subset of Merge(lnode))
         return true
-       end if
-     end if
-   end for
   return false
 
-// Compute merge set top-down in breadth first order.
+// Compute a merge-set top-down in breadth first order.
 bool constructMergeSet_1(BFSList B, JEdges JE, DomLevel DL)
-repeat = False
-// List of visited edges
-Visited V
-for n in B do
-  for e in (all incoming edges to n) do
-    if (e is in JE && e not in V) then
+  repeat = false
+  V = empty list of visited edges
+  for n in B
+    for e in (all incoming edges to n)
+      if (e is in JE && e not in V)
         V(e) = true
-      Let snode = Source Node of e
-      Let tnode = Target Node of e
-      Let tmp = snode
-      Let lnode = NULL
-      while (DL(tmp) ≥ DL(tnode)) do
+        let tmp = Source Node of e
+        let tnode = Target Node of e
+        let lnode = NULL
+        while (DL(tmp) >= DL(tnode))
           Merge(tmp) = Merge(tmp) U Merge(tnode) U {tnode}
@@
           tmp = dom-parent(tmp)
-      end while
         repeat = still_inconsistent(lnode, JE, Merge)
-    end if
-  end for
-end while
   return repeat
 
-// Construct Merge set of each node in control-flow-graph G of a function.
+// Construct merge-set of each node in Control-Flow-Graph G of a function.
 void  constructMergeSet(CFG G) {
@@
   do // Call until a fixed point is reached.
-    Repeat = constructMergeSet_1(B, JE, DL);
-  while (Repeat);
+    Repeat = constructMergeSet_1(B, JE, DL)
+  while (Repeat)
 }
@@
 
-We compute merge sets of the control flow graph which remains same through the
-code-motion transformation. For code-motion we only want to know if a use
-operand is a kill (to compute changes in register pressure). For that we only
+We compute merge sets of the control flow graph which remains the same through
+the code-motion transformation. For code-motion we only want to know if a use
+operand is a kill (to compute changes in register pressure.) For that we only
 need to know whether the use is also required later in the execution path. A
@@
 \begin{verbatim}
-bool isLiveOutUsingMergeSet()
-  // Compute if variable a is liveout from basic block n
-  Input: Node N, Variable a
-
-  if a is defined in N then
-    if a is used outside any basic block other than N then
-      return true;
-    else return false;
-  endif
-
-  Ms(n) = null;
+// Return true if variable A is liveout from basic block N
+bool isLiveOutUsingMergeSet(Node N, Variable A)
+  if (A is defined in N)
+    if (A is used outside any basic block other than N)
+      return true
+    return false
+
+  Ms(N) = null
   // Mergeset of N is the union of merge sets of its successors
-  for w in successors(n) do
-    Ms(n) = Ms(n) U Mr(w);
-  endfor
-
-  // Iterate over all the uses of a and see if any intersect with the
-  // merge set of N.
-  for t in users(a) do
-    b = basic_block(t)
-    while (b != null) and (b != def_bb(a)) do
-      if b ∩ Ms(n) then
-        return true;
-      endif
-      b = dom-parent(b);
-    endwhile
-  endfor
-  return false;
-\end{verbatim}
-
-\begin{verbatim}
-// return true if op is the last use of the
-// operand in I.
-bool isKill(Operand op, Instruction I) {
+  for W in successors(N)
+    Ms(N) = Ms(N) U Mr(W)
+
+  // Iterate over all the uses of A and see if any intersect with the merge-set of N.
+  for T in users(A)
+    B = basic_block(T)
+    while (B != null and B != def_bb(A))
+      if B in Ms(N) return true
+      B = dom-parent(B)
+  return false
+
+// Return true if op is the last use of the operand in I.
+bool isKill(Operand op, Instruction I)
   if (isLiveOutUsingMergeSet(op))
-    return false;
-  B = basic_block(I);
-  for each I1 after I in B:
+    return false
+
+  for each I1 after I in basic_block(I)
     if (I1 uses op)
-      return false;
-
-return true;
-}
+      return false
+  return true
 \end{verbatim}
@@
 Compute GVN of each expression in the function
-Compute DJ Graph of function and compute mergeset based on that
+Compute DJ-graph of function and compute mergeset based on that
 
@@
 The analyses computed for this pass are Global Value Numbering, Computation of
-DJ Graph and MergeSets, Marking Barriers, all are linear in number of
+DJ-graph and merge-sets, Marking Barriers, all are linear in number of
 instructions in a function. Liveness analysis expensive but only performed

### 2017-04-17 paper.tex
anonymous and change title
fix section 3

@@
 %% For double-blind review submission
-\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
 %\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
@@
 %% Title information
-\title[SSA GCM]{SSA based global code motion of identical computations}         %% [Short Title] is optional;
+\title[SSA GCM]{Global code motion of congruent instructions}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 them during implementation. If a variable is not live-out of a basic block, it
-still may be used later in the basic block. To establish whether a use of a
+still may be used later in the basic block: to establish whether a use of a
 variable is a kill we iterate on all the subsequent instructions in a basic
-block checking for uses should isLiveOutUsingMergeSet return false.
+block checking for uses.
 
@@
 comes to loads and stores so we compute the \GVN{} of loads and stores
-separately.  Our solution to value number loads is to hash the address from
-where the value is to be loaded. For stores, we value number the address as well
-as the value to be stored at that address. Another limitation of the current
-\GVN{} implementation in \LLVM{} is that the instructions dependent on the loads
-will not get numbered correctly, and so after hoisting all candidates we need to
-rerun the \GVN{} analysis in order to discover new candidates now available
-after having hoisted load instructions.  This limitation should be addressed in
-a new implementation of the \GVN{} based on \MemorySSA{}, that would better
-account for equivalent loads and their dependent instructions.
+separately.  Our solution is to value number the address from where the value of
+a load is to be read from memory. For stores, we value number the address as
+well as the value to be stored at that address. Another limitation of the
+current \GVN{} implementation in \LLVM{} is that the instructions dependent on
+the loads will not get numbered correctly, and so after hoisting all candidates
+we need to rerun the \GVN{} analysis in order to discover new candidates now
+available after having hoisted load instructions.  This limitation should be
+addressed in a new implementation of the \GVN{} based on \MemorySSA{}, that
+would better account for equivalent loads and their dependent instructions.
 
@@
 is found, we check whether all the use-operands of the set of instructions are
-available at that position. It is possible reinstantiate (remateralize) the
+available at that position. It is possible to reinstantiate or remateralize the
 use-operands in some cases when the operands are not available and make it legal
@@
 hoisted/sunk and their hoisting/sinking point. It is also necessary to check if
-there are indirect branch targets e.g., landing pad, case statements, goto
+there are indirect branch targets, e.g., landing pad, case statements, goto
 labels etc., along the path because it becomes difficult to prove safety checks
@@
 \subsubsection{Legality of sinking expressions}
-Sometimes, hoisting is not upward-safe \ref{click1995global} e.g., if the
-expressions are in a landing pad etc., in that case sinking of those expressions
+Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
+expressions are in a landing pad, sinking of those expressions
 may reduce code size. Sinking may also reduce the register pressure in some
-cases e.g., when the use operands are not kills.For sinking, higher ranked
+cases, e.g., when the use operands are not kills. For sinking, higher ranked
 expressions would be sunk first. And it would be illegal to sink higher ranked

### 2017-04-17 paper.tex
fix section 3.4

@@
 expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
-basic block B i.e., values computed on all paths from B to exit and
+basic block B, i.e., values computed on all paths from B to exit and
 AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
@@
 (d1, d2) can safely be sunk because their operands are readily available at the
-sink point i.e., B3. It should also be noted that, just because the expressions
+sink point, i.e., B3. It should also be noted that, just because the expressions
 are identical and operands are available, it still requires a unique
@@
 That takes into account the impact code-motion would have on various parameters
-that might affect runtime performance e.g., impact on live-range, gain in the
+that might affect runtime performance, e.g., impact on live-range, gain in the
 code size. Since this is mostly a code-size optimization pass, the goal is to
-not regress in performance across popular benchmarks at the same time reduce
+not regress in performance across popular benchmarks and at the same time to reduce
 code size as much as possible. Following are the cost models which are
@@
 killed operands. If the live-range after code-motion is less than before it will
-be moved. Essentially, as long as there is one killed operand , code hoisting
+be moved. Essentially, as long as there is one killed operand, code hoisting
 will either decrease or preserve the register pressure.  Similarly, code-sinking
@@
 B2: a1 = b<kill> + c<kill>
+\end{verbatim}
 
@@
 of a0 as a01 is placed in B0 just before the branch.
-\end{verbatim}
 
@@
 
-Code-motion basically consists of two parts i.e., hoisting and sinking. This
+Code-motion basically consists of two parts, i.e., hoisting and sinking. This
 implementation only moves congruent instructions. An immediate guarantee of this
@@
 as a barrier. In the absense of context (as in our current implementation), some
-instructions which might be safe are still classified a barrier e.g., volatile
+instructions which might be safe are still classified a barrier, e.g., volatile
 loads/stores, calls with missing attributes.

### 2017-04-17 paper.tex
fix section 3.4.1

@@
 will either decrease or preserve the register pressure as long as there is one
-operand killed at most.  Following example explains how code motion of identical
-computations can reduce the register pressure.  Consider the following example
-where the labels prefixed with 'P' represent the position of instruction in a
-basic block (names prefixed with 'B').
+operand killed at most.  The following example explains how code motion of
+identical computations can reduce the register pressure.
 
@@
 
-After hoisting a0 and a1 are removed and a copy
-of a0 as a01 is placed in B0 just before the branch.
-
-In this case, since `b' and `c' are killed in `a0' and `a1', hoisting them will
-reduce the register pressure in B3 and B4 because two registers will be freed.
-
-Ideally, it should be okay to hoist all the instructions and a later a
+After hoisting a0 and a1 are removed and a copy of a0 as a01 is placed in B0
+just before the branch. Since `b' and `c' are killed in `a0' and `a1', hoisting
+the expressions will reduce the register pressure because two registers will be
+freed.
+
+Ideally, it should be okay to hoist all the instructions and a later
 live-range-splitting \cite{cooper1998live} pass should make the right decision
@@
 current live-range splitting pass of \LLVM{} is not making the optimial decision
-and we have found regressions while hoisting aggressively.
+and we have found performance regressions while hoisting aggressively.
 

### 2017-04-17 paper.tex
perf data

@@
 \begin{abstract}
-  We present a global code motion compiler optimization which schedules
-identical computations across the program so as to save code size.  Not only
-this code motion saves code size, it exposes redundancies in some cases, it
-exposes more instruction level parallelism in the basic block when the
-computations are hoisted, and it enables other passes like loop invariant
-motion to remove more redundancies. The cost model to drive the code motion is
-based on live range analysis on \SSA{} representation such that the (virtual)
-register pressure does not increase.
+We present a global code motion compiler optimization which schedules congruent
+instructions across the program.  Not only this code motion saves code size, it
+exposes redundancies in some cases, it exposes more instruction level
+parallelism in the basic block to which instructions are moved, and it enables
+other passes like loop invariant motion to remove more redundancies. The cost
+model to drive the code motion is based on live range analysis on \SSA{}
+representation such that the (virtual) register pressure does not increase.
 
 We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
-infrastructure available in \LLVM{}. The experimental results show an average
-of 2.5\% savings in code size on \SPEC{}, although the code size also
-increases in many cases because it enables more inlining. We have also seen
-execution time improvements of approximately 5\% in a couple of \SPEC{} benchmarks viz. \GCC{}
-and mcf, moreover, register spills reduced for almost all the \SPEC{}
-bencmarks when compiled for X86\_64. This is an optimistic algorithm in the
-sense that it considers all identical computations in a function as potential
-candidates. We make an extra effort to hoist candidates by partitioning the
-potential candidates in a way to enable partial hoisting in case common
-hoisting points for all the candidates cannot be found. We also formalize how
-register pressure will reduce as a result of code-motion and why sorting the
-list of potential candidates w.r.t. their depth first numbers helps hoist more
-candidates with less compile time overhead.
+infrastructure available in \LLVM{}. The experimental results show an average of
+2\% reduction in the number of spills, and saving of 1\% on overall compile time
+on \SPEC{}. This being a code-size optimization, enables more inlining in
+majority of the benchmarks. We have also seen execution
+time improvements of approximately 5\% in a couple of \SPEC{} benchmarks
+viz. \GCC{} and mcf, moreover, register spills reduced for almost all the
+\SPEC{} bencmarks when compiled for X86\_64. This is an optimistic algorithm in
+the sense that it considers all identical computations in a function as
+potential candidates. We make an extra effort to hoist candidates by
+partitioning the potential candidates in a way to enable partial hoisting in
+case common hoisting points for all the candidates cannot be found. We also
+formalize how register pressure will reduce as a result of code-motion and why
+sorting the list of potential candidates w.r.t. their depth first numbers helps
+hoist more candidates with less compile time overhead.
 \end{abstract}
@@
 where a load or a store would happen. It is a scalar computation and gets
-hoisted frequently even if the corresponding loads/stores would not get hoisted.
-In order to reduce register pressure while hoisting loads, we have restricted
-hoisting of address computations away from their corresponding loads and stores
-when the loads and stores cannot be moved.  This restriction is only to mitigate
-the limitations of \LLVM{'}s register allocator and may be lifted in the future,
-when the register allocation rematerialization pass has been improved to catch
-these regressions.
+hoisted frequently even if the corresponding loads/stores would not be legal to
+hoist.  In order to reduce the register pressure while hoisting loads, we have
+restricted hoisting of address computations away from their corresponding loads
+and stores when the loads and stores cannot be moved. This restriction is only
+to mitigate the limitations of \LLVM{'}s register allocator and may be lifted in
+the future, when the register allocation rematerialization pass has been
+improved to catch these regressions.
 
@@
 In our current implementation we keep as many candidates in one set as possible
-(greedy approach). We split the list at a point where the legality checks fail
+(greedy approach.) We split the list at a point where the legality checks fail
 to hoist subset of candidates which are legal to hoist and then start finding
@@
 would make the loaded value available by the time it is used. Also, because
-stores and calls are hoisted the least \ref{table:experimental-results}, the
-performance does not change much whether they are hoisted or not.
+stores and calls are hoisted the least, see Table~\ref{tab:code-motion-metric},
+the performance does not change much whether they are hoisted or not.
 
@@
 approach and can be improved by ranking the computations \cite{rosen1988global},
-or by improving the \GVN{} analysis to correctly populate congruence classes.
+or by improving the \GVN{} analysis to correctly populate congruence classes as
+the program is modified by the code generation.
 
@@
 the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
-optimization level.  The results for code-hoisting are listed in
+optimization level.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|c|c|}
+      \hline
+% TODO: Check if putting code-size or spills number is okay or only %age is required.
+% llvm trunk vs. llvm trunk with code-motion
+Benchmark       &\% change in performance \% change in codesize & \\\hline
+400.perlbench	& 1.00  & 1.00 \\\hline
+401.bzip2	& 1.00  & 1.00 \\\hline
+403.gcc	        & 1.00  & 1.01 \\\hline
+429.mcf	        & 1.03  & 1.00 \\\hline
+433.milc	& 1.00  & 1.00 \\\hline
+444.namd	& 1.00  & 1.01 \\\hline
+445.gobmk	& 0.99  & 1.01 \\\hline
+447.dealII	& 0.95  & 1.01 \\\hline
+450.soplex	& 0.99  & 0.99 \\\hline
+453.povray	& 0.97  & 1.00 \\\hline
+456.hmmer	& 0.99  & 1.00 \\\hline
+458.sjeng	& 1.02  & 1.00 \\\hline
+462.libquantum	& 1.00  & 1.00 \\\hline
+464.h264ref	& 1.00  & 1.00 \\\hline
+470.lbm	        & 1.00  & 1.00 \\\hline
+471.omnetpp	& 1.00  & 0.99 \\\hline
+473.astar	& 1.00  & 1.02 \\\hline
+482.sphinx3	& 0.99  & 1.00 \\\hline
+483.xalancbmk	& 1.01  & 0.99 \\\hline
+Geomean         & 0.997 & 1.00 \\\hline
+    \end{tabular}
+  \end{center}
+  \caption{Execution time and code size with and without code-motion on SPEC2006 at -Ofast}
+  \label{tab:code-size}
+\end{table}
+
+Overall there was only a minor change in performance regressed with a few ups
+(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
+etc.). The code-motion pass was run twice in the pass pipeline (at -Ofast),
+first after EarlyCSE pass and then after the inliner. Because EarlyCSE removes
+local redundancies, GCM would have to analyze less redundant instructions, also
+as inliner creates more redundancies, GCM would have more opportunities to move
+congruent instructions out of branches.
+
+Code size also remained almost the same as a result of code-motion as shown in
+Table~\cite{tab:code-size}.  The code size listed here is the text size (from
+linux size command) of the final executables for each benchmark.  Some bencmarks
+like 473.astar increased in code size by almost 2\%. This is because once the
+code-size of a function decreases it becomes cheaper to inline. As we can see in
+Table-\cite{tab:stats}, four more functions got inlined in 473.astar. Except for
+447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
+more functions inlined.
+
+  The results for code-hoisting are listed in
 Table~\ref{tab:code-motion-metric}. The table lists the number of scalars, loads,
@@
 
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{|l|c|c|c|}
-      \hline
-% TODO: Check if putting code-size or spills number is okay or only %age is required.
-% llvm trunk vs. llvm trunk with code-motion
-Code size         &\% age	increase in codesize \\\hline
-400.perlbench	  & 1.00 \\\hline
-401.bzip2	  & 1.00 \\\hline
-403.gcc	          & 1.01 \\\hline
-429.mcf	          & 1.00 \\\hline
-433.milc	  & 1.00 \\\hline
-444.namd	  & 1.01 \\\hline
-445.gobmk	  & 1.01 \\\hline
-447.dealII	  & 1.01 \\\hline
-450.soplex	  & 0.99 \\\hline
-453.povray	  & 1.00 \\\hline
-456.hmmer	  & 1.00 \\\hline
-458.sjeng	  & 1.00 \\\hline
-462.libquantum	  & 1.00 \\\hline
-464.h264ref	  & 1.00 \\\hline
-470.lbm	          & 1.00 \\\hline
-471.omnetpp	  & 0.99 \\\hline
-473.astar	  & 1.02 \\\hline
-482.sphinx3	  & 1.00 \\\hline
-483.xalancbmk	  & 0.99 \\\hline
-Grand total       & 1.00 \\\hline
-    \end{tabular}
-  \end{center}
-  \caption{Change in code size with and without code-motion on SPEC2006 at -Ofast}
-  \label{tab:code-size}
-\end{table}
-
-Code size remained almost the same as a result of code-motion as shown in Table~\cite{tab:code-size}.
-The code size listed here is the text size (from linux size command) of the final executables for each benchmark.
-Some bencmarks like 473.astar increased in code size by almost 2\%. This is because once the code-size
-of a function decreases it becomes cheaper to inline. As we can see in Table-\cite{tab:stats}, four
-more functions got inlined in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
-all other benchmarks got more functions inlined.
-
 \begin{table}[h]
@@
       \hline
-  Benchmark       &   Baseline	        &  GCM & \%Incrase     \\\hline
+  Benchmark       &   Baseline	        &  GCM & \%Increase     \\\hline
   400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline

### 2017-04-17 paper.tex
experiment

--- 2017-04-17 paper.tex
@@
 % Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
-We ran \SPEC{} with the patch. All
-the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
-optimization level.
+We ran \SPEC{} with the patch. All the experiments were conducted on x86\_64
+Ubuntu-Linux machine and at -Ofast optimization level. Each sub benchmark was
+run three times and the best result was taken. We collected execution time,
+code-size which is listed in Table~\ref{tab:code-size}, other compiler stats are
+listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
+Table~\ref{tab:code-motion-metric}.  We also measured compile time
+Table~\ref{tab:compile-time} to see the impact of code-motion.
 
@@
 The results for code-hoisting are listed in
-Table~\ref{tab:code-motion-metric}. The table lists the number of scalars, loads,
-stores and calls hoisted as well as removed. For each category, the number of
-instructions removed is greater or equal to the number of instructions hoisted
-because each hoisting is performed only when at least one identical computation
-is found.
+Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
+loads, stores and calls hoisted as well as removed. For each category, the
+number of instructions removed is greater or equal to the number of instructions
+hoisted because each hoisting is performed only when at least one identical
+computation is found.
 
@@
 
-Other static metrics are listed in Table~\ref{tab:stats} and in Table~\ref{tab:code-motion-metric}.
-Here we can
-see that except for rematerializing defs for splitting, which has an overhead of ~2\%, all
-other parameters have less than 1\% overhead. This is to explain why the performance does not
-go down with our implementation (and cost-model) of code hoisting pass.
+Other static metrics are listed in Table~\ref{tab:stats} and in
+Table~\ref{tab:code-motion-metric}.  Here we can see that except for
+rematerializing defs for splitting, which has an overhead of ~2\%, all other
+parameters have less than 1\% overhead. This is to explain why the performance
+does not go down with our implementation (and cost-model) of code hoisting pass.
 
@@
  code-motion  &\multicolumn{2}{|c|}{Total Instructions} &\multicolumn{2}{|c|}{Scalars}  & \multicolumn{2}{|c|}{Loads} & \multicolumn{1}{|c|}{Stores} &Calls   &Instructions    \\\hline
- metric           &  hoisted    &removed  &hoisted &removed&hoisted     &removed        &hoisted=removed&hoisted & sunk \\\hline
+ metric           &  H          &R        &H      &R       &H           &R              &H=R    &H & sunk \\\hline
  400.perlbench	  &  2387	& 2456	  &  855  & 857	   & 1493	& 1560		& 38	& 1	& 4	        \\\hline
@@
   \end{center}
-  \caption{Code motion stats on SPEC2006 at -Ofast}
+  \caption{Code motion stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are hoisted the most followed by scalars, stores and calls}
   \label{tab:code-motion-metric}

### 2017-04-17 paper.tex
fuse section 4 into 3.5 codegen

@@
 
-\subsection{Code generation}
+\subsection{Code generation for the optimistic code motion algorithm}
+\label{subsec:optimistic}
 Once all the legality and profitability checks are satisfied for a set of
-congruent instructions, they are suitable candidates for hoisting. A copy of the
-computation is inserted at the hoisting point along with any instructions which
-needed to be rematerialized. Thereafter, all the computations made redundant by
-the new copy are removed, and the \SSA{} form is restored by updating the
-intermediate representation (IR) to reflect the changes. At the same time
-\MemorySSA{} is also updated to get up-to-date information about memory references.
+congruent instructions, they are suitable candidates for hoisting or sinking. A
+copy of the computation is inserted at the hoisting/sinking point along with any
+instructions which needed to be rematerialized. Thereafter, all the computations
+made redundant by the new copy are removed, and the \SSA{} form is restored by
+updating the intermediate representation (IR) to reflect the changes. At the
+same time \MemorySSA{} is also updated to get up-to-date information about
+memory references.
 
@@
 
-\section{\SSA{} based global code motion of identical computations}
-\label{sec:implementation-details}
-% Move this before legality and profitability??
-% Ranking of instructions from Zadeck
-% Liveness analysis from Ramakrishna
-In this section we will present an overview of the algorithm we implemented in
-\LLVM{}.  The amount of hoisting depends on whether we collect \GVN{} of
-instructions before finding candidates (optimistic) or, on-demand
-(pessimistic). It also depends on the generality of the \GVN{} algorithm as
-mentioned earlier in Section~\ref{subsec:finding-candidates}. We have
-implemented a optimistic global code motion of congruent instructions which uses
-the liveness analysis as illustrated in Das \cite{das2012} and ranking
-expressions explained in Rosen \cite{rosen1988global}.
+The amount of hoisting depends on whether we collect \GVN{} of instructions
+before finding candidates (optimistic) or, on-demand (pessimistic). It also
+depends on the generality of the \GVN{} algorithm as mentioned earlier in
+Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
+global code motion of congruent instructions which uses the liveness analysis as
+illustrated in Das \cite{das2012} and ranking expressions explained in Rosen
+\cite{rosen1988global}.
 
@@
 approach is gain in code size (the final executable may be of larger size
-because of more inlining). The algorithm prefers hoisting to sinking. If the
+because of more inlining.) The algorithm prefers hoisting to sinking. If the
 dependency of a hoistable candidate is in the same basic block as the candidate,
-then the dependency must also be hoistable otherwise hoisting will be illegal or
+then the dependency must also be hoistable, otherwise hoisting will be illegal or
 would require a complicated code generation to make it legal. The current
@@
 
-\subsection{Optimistic code motion algorithm}
-\label{subsec:optimistic}
 We collect the \GVN{} of all the instructions in the function and iterate on the
@@
 which might not return. So barriers can be present in the middle of a basic
-block. Essentially, any instruction that cannot not guarantee progress is marked
-as a barrier. In the absense of context (as in our current implementation), some
-instructions which might be safe are still classified a barrier, e.g., volatile
+block. Essentially, any instruction that cannot guarantee progress is marked
+as a barrier. In the absence of context (as in our current implementation), some
+instructions which might be safe are still classified as barriers, e.g., volatile
 loads/stores, calls with missing attributes.

### 2017-04-17 paper.tex
abstract

@@
 \begin{abstract}
-We present a global code motion compiler optimization which schedules congruent
-instructions across the program.  Not only this code motion saves code size, it
+We present a global code motion (GCM) compiler optimization which schedules
+congruent instructions across the program.  Not only this saves code size, it
 exposes redundancies in some cases, it exposes more instruction level
@@
 model to drive the code motion is based on live range analysis on \SSA{}
-representation such that the (virtual) register pressure does not increase.
+representation such that the (virtual) register pressure does not increase, which
+translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.
 
 We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
-infrastructure available in \LLVM{}. The experimental results show an average of
-2\% reduction in the number of spills, and saving of 1\% on overall compile time
-on \SPEC{}. This being a code-size optimization, enables more inlining in
-majority of the benchmarks. We have also seen execution
-time improvements of approximately 5\% in a couple of \SPEC{} benchmarks
-viz. \GCC{} and mcf, moreover, register spills reduced for almost all the
-\SPEC{} bencmarks when compiled for X86\_64. This is an optimistic algorithm in
-the sense that it considers all identical computations in a function as
-potential candidates. We make an extra effort to hoist candidates by
+infrastructure available in \LLVM{}. The experimental results show an average
+saving of 1\% on overall compile time on \SPEC{}. GCM enables more inlining and
+exposes more loop invariant code motion opportunities in majority of the
+benchmarks. We have also seen execution time improvements in a few of \SPEC{}
+benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
+\SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
+algorithm in the sense that it considers all identical computations in a
+function as potential candidates. We make an extra effort to hoist candidates by
 partitioning the potential candidates in a way to enable partial hoisting in
 case common hoisting points for all the candidates cannot be found. We also
-formalize how register pressure will reduce as a result of code-motion and why
-sorting the list of potential candidates w.r.t. their depth first numbers helps
-hoist more candidates with less compile time overhead.
+formalize how register pressure will reduce as a result of code-motion, and how
+code-motion increases instruction level parallelism thereby enabling more out of
+order execution on modern architectures.
 \end{abstract}
@@
 
+\subsubsection{Barriers}
+Barriers are based on the concept of pinned instructions \cite{click1995global}
+but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is
+actually an extended basic block because there might be calls in the middle of a
+basic block which might not return. Essentially, any instruction that cannot
+guarantee progress is marked as a barrier. In the absence of context, as in our
+current implementation, some instructions which might be safe are still
+classified as barriers, e.g., volatile loads/stores, calls with missing
+attributes.
+
+\begin{verbatim}
+computing barriers:
+for each basic block B in a function:
+  barrier_found = false
+  for each instruction I in B:
+    if I does not guarantee progress:
+      mark I as a barrier instruction
+      barrier_found = true
+      break;
+
+  // Find the last barrier below which instructions can be sunk.
+  // If there was no barrier in B, any instruction satisfying
+  // other legality checks can be sunk.
+  if barrier_found:
+    for each instruction I in B in reverse order:
+      if I does not guarantee progress:
+        mark I as a sink barrier
+        break
+\end{verbatim}
+
+\subsubsection{Downward-safety of movable instructions}
+
+In order to be able to move an instruction that has side effects, we must
+guarantee that the side-effects of that instructions before and after its motion
+will produce the same side-effects.  For hoisting in particular, we need to
+prove that on all execution paths from the hoisting point to the end of the
+program the side effects appear in the same order.
+
+\begin{verbatim}
+computing downward-safety of hoistable instructions at hoist-point
+  Worklist = list of all basic blocks of hoistable instructions
+  for each basic block B in the dominator tree starting at hoist-point:
+    if Worklist is empty
+      return false // Path exists! not downward safe
+    if B is in Worklist:
+      remove B from worklist
+      remove subtree with root at B // Available at B => downward safe from B
+    if B is a leaf basic block:
+      return false  // Path exists! not downward safe
+    if B dominates hoist-point:
+      return false  // Back edge
+  return true
+\end{verbatim}
+
 \subsection{Profitability check (Cost models)}
@@
 \begin{verbatim}
-    b = m
+B0: b = m
     c = n
-B0: if c is true then goto B3 else goto B4
+    if c is true then goto B3 else goto B4
 
@@
 
-Barriers are based on the concept of pinned instructions \cite{click1995global}
-but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
-extended basic block because there might be calls in the middle of a basic block
-which might not return. So barriers can be present in the middle of a basic
-block. Essentially, any instruction that cannot guarantee progress is marked
-as a barrier. In the absence of context (as in our current implementation), some
-instructions which might be safe are still classified as barriers, e.g., volatile
-loads/stores, calls with missing attributes.
-
-\begin{verbatim}
-computing barriers:
-for each basic block B in a function:
-  barrier_found = false
-  for each instruction I in B:
-    if I does not guarantee progress:
-      mark I as a barrier instruction
-      barrier_found = true
-      break;
-
-  // Find the last barrier below which instructions can be sunk.
-  // If there was no barrier in B, any instruction satisfying
-  // other legality checks can be sunk.
-  if barrier_found:
-    for each instruction I in B in reverse order:
-      if I does not guarantee progress:
-        mark I as a sink barrier
-        break
-\end{verbatim}
-
-\begin{verbatim}
-computing downward-safety of hoistable instructions at hoist-point
-Worklist = list of all basic blocks of hoistable instructions
-for each basic block B in the dominator tree starting at hoist-point:
-  if Worklist is empty
-    return false // Path exists! not downward safe
-  if B is in Worklist:
-    remove B from worklist
-    remove subtree with root at B // Available at B => downward safe from B
-  if B is a leaf basic block:
-    return false  // Path exists! not downward safe
-  if B dominates hoist-point:
-    return false  // Back edge
-
-\end{verbatim}
-
 % Explain the overall algorithm, explain how anticipability is computed on demand
-% explain the pinned-instructions/barriers.
 \begin{verbatim}
@@
 iterate on lower ranking expressions first and then proceed to higher ranking
-expressions in the same iteration but \LLVM{'}s \GVN{} infrastucture does not compute
-equivalence classes in a effective way. We found it simpler to just recompute
-the value numbers and start finding hoistable candidates again.
+expressions in the same iteration but \LLVM{'}s \GVN{} infrastucture does not
+compute equivalence classes in a effective way. We found it simpler to just
+recompute the value numbers and start finding hoistable candidates again.
 
@@
   \end{center}
-  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast}
+  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast. GCM improves LICM and inlining}
   \label{tab:stats}

### 2017-04-17 paper.tex
table

--- 2017-04-17 paper.tex
@@
   \begin{center}
-    \begin{tabular}{|l|r|r|r|r|r|r|r|r|}
+    \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
       \hline
-  &  \multicolumn{2}{|c|}{Hoisted by LICM} &\multicolumn{2}{|c|}{Sunk by LICM} &\multicolumn{2}{|c|}{Functions Inlined} &  \multicolumn{2}{|c|}{Functions Deleted} \\\hline
- Benchmark        & trunk       & GCM	& trunk	& GCM	& trunk	& GCM	& trunk	& GCM  \\\hline
- 400.perlbench	  & 2428	& 2480	& 36	& 35	& 1562	& 1565	& 232	& 234  \\\hline
- 401.bzip2	  & 721	        & 825	& 3	& 3	& 172	& 180	& 32	& 33   \\\hline
- 403.gcc	  & 8867	& 10079	& 143	& 182	& 5034	& 5080	& 1303	& 130  \\\hline
- 429.mcf	  & 55	        & 51	& 1	& 1	& 4	& 4	& 0 	& 0    \\\hline
- 433.milc	  & 2537	& 2795	& 3	& 3	& 75	& 81	& 0	& 0    \\\hline
- 444.namd	  & 2204	& 2206	& 240	& 240	& 387	& 390	& 57	& 55   \\\hline
- 445.gobmk	  & 3090	& 3107	& 18	& 19	& 735	& 775	& 230	& 234  \\\hline
- 447.dealII	  & 59114	& 95884	& 60	& 8	& 100891& 100692& 30219	& 29966\\\hline
- 450.soplex	  & 4668	& 4848	& 18	& 17	& 9801	& 9784	& 3125	& 3112 \\\hline
- 453.povray	  & 9533	& 9079	& 23	& 32	& 5522	& 5532	& 1202	& 1201 \\\hline
- 456.hmmer	  & 3200	& 3232	& 3	& 8	& 562	& 562	& 103	& 101  \\\hline
- 458.sjeng	  & 400	        & 446	& 0	& 0	& 127	& 135	& 18	& 18   \\\hline
- 462.libquantum	  & 362	        & 386	& 0 	& 0	& 87	& 87	& 21	& 21   \\\hline
- 464.h264ref	  & 17075	& 17153	& 4	& 4	& 338	& 382	& 72	& 72   \\\hline
- 470.lbm	  & 39	        & 39	& 0	& 0	& 13	& 13	& 4	& 4    \\\hline
- 471.omnetpp	  & 565	        & 661	& 32	& 30	& 4790	& 4843	& 1321	& 1316 \\\hline
- 473.astar	  & 691	        & 754	& 0	& 0	& 359	& 363	& 113	& 113  \\\hline
- 482.sphinx3	  & 1599	& 1693	& 11	& 9	& 198	& 194	& 53	& 53   \\\hline
- 483.xalancbmk	  & 9997	& 11453	& 56	& 54	& 69534	& 68839	& 29875	& 29519\\\hline
+  &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
+ Benchmark        & Base        & GCM	& \%  &Base     & GCM   &\%    & Base	& GCM	  &\%   & Base	& GCM	& \% \\\hline
+ 400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
+ 401.bzip2	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
+ 403.gcc	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
+ 429.mcf	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
+ 433.milc	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
+ 444.namd	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
+ 445.gobmk	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
+ 447.dealII	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
+ 450.soplex	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
+ 453.povray	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
+ 456.hmmer	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
+ 458.sjeng	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
+ 462.libquantum	  & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
+ 464.h264ref	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
+ 470.lbm	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
+ 471.omnetpp	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
+ 473.astar	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
+ 482.sphinx3	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
+ 483.xalancbmk	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
     \end{tabular}
   \end{center}
-  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast. GCM improves LICM and inlining}
+  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
   \label{tab:stats}
@@
   \end{center}
-  \caption{Code motion stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are hoisted the most followed by scalars, stores and calls}
+  \caption{GCM stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are hoisted the most followed by scalars, stores and calls}
   \label{tab:code-motion-metric}

### 2017-04-17 paper.tex
change algos

@@
 scheduling \cite{dragonbook,click1995global} in the sense that it will only move
-computations. Code hoisting can reduce the critical path length of execution in
-out-of-order machines. As more instructions are available at the hoisting point,
-the hardware has more independent instructions to reorder. The following example
-illustrates how hoisting can improve performance by exposing more ILP.
+computations. Code hoisting, for example, can reduce the critical path length of
+execution in out-of-order machines. As more instructions are available at the
+hoisting point, the hardware has more independent instructions to reorder. The
+following example illustrates how hoisting can improve performance by exposing
+more ILP.
 
@@
 
-In fact there are several advantages of code-motion of this kind:
+In fact, there are several advantages of GCM.
 \begin{itemize}
@@
 scheduling to improve performance. Most of the recent work on global scheduling
-are done using ILP which results in prohibitively high compile time. To the best
-of our knowledge we have not found any reference which explored code-motion of
-identical computation in as much detail as we have done. A part of our
-implementation (aggressive code hoisting) is already merged in \LLVM{} trunk,
-however, a general implementation of code-motion of congruent (non redundant)
-expressions is still missing from \GCC{} and \LLVM{} trunk. The main
-contributions of this paper are:
+are done using integer linear programming (ILP), which results in prohibitively
+high compile time. To the best of our knowledge we have not found any reference
+which explored code-motion of congruent instructions in as much detail as we
+have done. A part of our implementation (aggressive code hoisting) is already
+merged in \LLVM{} trunk, however, a general implementation of code-motion of
+congruent (non redundant) expressions is still missing from \GCC{} and \LLVM{}
+trunk. The main contributions of this paper are:
 \begin{itemize}
@@
   \SSA{} based liveness analysis, and ranking of expressions to move very busy
-  expressions in order to reduce code-size and improve performance in some
+  expressions in order to reduce register spills and improve performance in some
   cases.
@@
 
-Barany \cite{barany2013} presented a global scheduler with integer linear
-programming (ILP) formulation with a goal to minimize register pressure. The
-results they got were not very promising. It may be because they only used the
-scheduler for smaller functions ($<1000$ instructions); also, they compiled the
-benchmarks for ARM-Cortex which is more resilient to register pressure because
-it has more registers compared to X86, for example.
+Barany \cite{barany2013} presented a global scheduler with ILP formulation with
+a goal to minimize register pressure. The results they got were not very
+promising. It may be because they only used the scheduler for smaller functions
+($<1000$ instructions); also, they compiled the benchmarks for ARM-Cortex which
+is more resilient to register pressure because it has more registers compared to
+X86, for example.
 
@@
 
-With the algorithm described in the next section, we got a
-reduction in register spills, a reduction in code-size and some performance
-improvements with very low compilation time overhead, on \SPEC{} benchmarks.
+With the algorithm described in the next section, we got a reduction in register
+spills, improved inliner heuristics, improved compile time and some performance
+improvements on \SPEC{} benchmarks.
 
@@
 
-The code-motion pass can be broadly divided into the following steps that we will
+The GCM pass can be broadly divided into the following steps that we will
 describe in the rest of this section:
@@
 compute the live-out relation on-demand when profitability of hoistable/sinkable
-candidates is to be evaluated. A simplified version of isLiveOutUsingMergeSet is
-presented here:
+candidates is to be evaluated. A simplified version of
+\emph{isLiveOutUsingMergeSet} is presented here:
 
@@
 variable is a kill we iterate on all the subsequent instructions in a basic
-block checking for uses.
+block checking for uses, the \emph{isKill} function uses
+\emph{isLiveOutUsingMergeSet} to find out if an operand is killed in an
+instruction.
 
@@
 hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
-be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
-B1 and those in B5, B6 to be hoisted at B2. Even if (B3, B4) could not be
-hoisted, for some reason, to B1, (B5, B6) can still be hoisted -- if legal and
-profitable -- to B2.
+be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
+B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
+could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
+still be hoisted -- if legal and profitable -- to B2.
 
@@
 approach is gain in code size (the final executable may be of larger size
-because of more inlining.) The algorithm prefers hoisting to sinking. If the
+because of more inlining). The algorithm prefers hoisting to sinking. If the
 dependency of a hoistable candidate is in the same basic block as the candidate,
@@
 \begin{verbatim}
+int doCodeHoisting()
   Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
@@
 
+  return number_of_hoisted_candidates
+
 If any of candidates were hoisted then repeat hoisting
+\end{verbatim}
+
 Once no more candidates are hoistable proceed to sinking
-\end{verbatim}
 
@@
 % llvm trunk vs. llvm trunk with code-motion
-Benchmark       &\% change in performance \% change in codesize & \\\hline
+Benchmark       &\% change in performance & \% change in codesize \\\hline
 400.perlbench	& 1.00  & 1.00 \\\hline

### 2017-04-17 paper.tex
split gcm

@@
 
+\begin{verbatim}
+void doGCM()
+  Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
+  Compute DJ-graph of function and compute mergeset based on that
+  do
+    Compute GVN of each expression in the function
+    int N = doCodeHoisting()
+  while (N > 0) // If any of candidates were hoisted then repeat hoisting
+  doCodeSinking()
+\end{verbatim}
+
 We collect the \GVN{} of all the instructions in the function and iterate on the
@@
 int doCodeHoisting()
-  Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
-
-  Compute GVN of each expression in the function
-  Compute DJ-graph of function and compute mergeset based on that
-
   For each value number VN with 2 or more instructions:
@@
   return number_of_hoisted_candidates
-
-If any of candidates were hoisted then repeat hoisting
 \end{verbatim}
 
-Once no more candidates are hoistable proceed to sinking
+After no more candidates are hoistable, sinking is performed. Sinking is only
+performed once becase currently there are very few sinkable candidates
+(Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
+GCM) has a (suboptimal) code-sinking which sinks several instructions already.
 
 \begin{verbatim}
-Analyses available: Dominator Tree, DFS Numbering, Memory SSA, Mergeset
-Compute GVN of each expression in the function
-
+void doCodeSinking()
   For each value number VN with 2 or more instructions:
@@
 code-size of a function decreases it becomes cheaper to inline. As we can see in
-Table-\cite{tab:stats}, four more functions got inlined in 473.astar. Except for
+Table-\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got

### 2017-04-17 paper.tex
conclusion

@@
 \SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
-algorithm in the sense that it considers all identical computations in a
+algorithm in the sense that it considers all congruent instructions in a
 function as potential candidates. We make an extra effort to hoist candidates by
@@
     if there are two instructions with same immediate successor as post-dominator
-      if both are only used in the same PHI-node of the successor
+      if both are only used in the same PHI-node (PN) of the successor
         if both have dependencies that are available or can be made available at the sink point
@@
     in the post-dominator
-    update the use of PHI with I1
-    Remove I2 and PHI
+    update all the uses of PHI (PN) with I1
+    Remove I2 and PN
     if I1 is a memory reference:
@@
 the number of instructions in the caller and callee, as its heuristics estimate
-the size of functions to be inlined. Various code-size metrics are shown in
-Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
-5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
-by 35.38\%. In this benchmark 3 more functions got inlined (15 as compared to
-12) and because of that 10 more vector instructions got generated (81 vs. 71), 3
-calls got hoisted/sunk as (compared to 0), one loop got unswitched (compared to
-0), 6 high latency machine instructions got hoisted out of loop, 59 (compared to
-30) machine instructions got hoisted out of loop, 70 (compared to 39) machine
-instructions were sunk.
+the size of functions to be inlined.
 
@@
 \label{sec:future-work}
-We have presented the \GVN{} based code hoisting algorithm. The primary goal is to
-reduce the code size but it benefits performance in some cases as well. To
-preserve performance and not hoist too much we have implemented several cost
-models described in Section~\ref{sec:cost-models}. Since those cost models
-depend on a set of thresholds, it requires tuning, as such, we used
-representative benchmarks to tune them.
-
-In general it is a good idea to start with lower ranked expression first such
-that maximum hoisting can happen in one iteration, however, current
-implementation does not rank the expressions and iteratively finds a fixed point
-when no more candidates are available. Even this implementation converges
-quickly and no significant compile time regression have been observed because of
-code hoisting pass. This is not the most optimal approach and results in
-multiple data structures to be recomputed. This can be improved by ranking the
-computations \cite{rosen1988global}. Also \GVN{}-hoist runs very early in the
-pass pipeline, it will be good to evaluate the codesize/performance impact when
-it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.
-
-With the implementation of code-hoisting in \LLVM{}, the passes which rely on the
+We have presented the global code motion of congruent computations in SSA
+form. We saw that it improves inlining and LICM opportunities, reduces register
+spills, it also improves performance in some cases. To preserve performance and
+not hoist too much we have register pressure aware cost models described in
+Section~\ref{sec:cost-models}. A part of this implementation is already merged
+in LLVM trunk as GVNHoist.cpp. We will post the GCM implementation for review.
+
+There are a few improvements that we would like to make. First would be to
+integrate this with \LLVM{'}s new GVN-PRE implementation such that more
+congruent instructions can be scheduled. This would also expose more
+redundancies to the GVN-PRE.  In general it is a good idea to start with lower
+ranked expression first such that maximum hoisting can happen in one iteration,
+however, current implementation does not rank the expressions and iteratively
+finds a fixed point when no more candidates are available. Even this
+implementation converges quickly and no compile time regression have been
+observed because of code hoisting pass. This is not the most optimal approach
+and results in multiple data structures to be recomputed. This can be improved
+by ranking the computations \cite{rosen1988global}. Also, since GCM runs very
+early in the pass pipeline, it will be good to evaluate the codesize/performance
+impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.
+
+With the implementation of GCM in \LLVM{}, the passes which rely on the
 code-size/instruction-count to make optimization decisions needs to be
 revisited. The first candidate would be the inliner. We have seen different
-inlining decisions in Table~\ref{tab:code-size}, before and after
-code-hoisting was enabled.  Since inliner has several magic numbers tuned for
-the previous pass layout, it would need some improvement.
+inlining decisions in Table~\ref{tab:code-size}, before and after code-hoisting
+was enabled.  Since inliner has several magic numbers tuned for the previous
+pass layout, it would need some improvement.
 

### 2017-04-17 paper.tex
code size at -Os

@@
 % llvm trunk vs. llvm trunk with code-motion
-Benchmark       &\% change in performance & \% change in codesize \\\hline
-400.perlbench	& 1.00  & 1.00 \\\hline
-401.bzip2	& 1.00  & 1.00 \\\hline
-403.gcc	        & 1.00  & 1.01 \\\hline
-429.mcf	        & 1.03  & 1.00 \\\hline
-433.milc	& 1.00  & 1.00 \\\hline
-444.namd	& 1.00  & 1.01 \\\hline
-445.gobmk	& 0.99  & 1.01 \\\hline
-447.dealII	& 0.95  & 1.01 \\\hline
-450.soplex	& 0.99  & 0.99 \\\hline
-453.povray	& 0.97  & 1.00 \\\hline
-456.hmmer	& 0.99  & 1.00 \\\hline
-458.sjeng	& 1.02  & 1.00 \\\hline
-462.libquantum	& 1.00  & 1.00 \\\hline
-464.h264ref	& 1.00  & 1.00 \\\hline
-470.lbm	        & 1.00  & 1.00 \\\hline
-471.omnetpp	& 1.00  & 0.99 \\\hline
-473.astar	& 1.00  & 1.02 \\\hline
-482.sphinx3	& 0.99  & 1.00 \\\hline
-483.xalancbmk	& 1.01  & 0.99 \\\hline
-Geomean         & 0.997 & 1.00 \\\hline
+Benchmark       &\% performance uplift (\%) at -Ofast & \% codesize uplift (\%) at -Ofast & codesize uplift (\%) at -Os\\\hline
+400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
+401.bzip2	& 1.00  & 1.00 & 1.00\\\hline
+403.gcc	        & 1.00  & 1.01 & 1.00\\\hline
+429.mcf	        & 1.03  & 1.00 & 1.00\\\hline
+433.milc	& 1.00  & 1.00 & 1.00\\\hline
+444.namd	& 1.00  & 1.01 & 1.00\\\hline
+445.gobmk	& 0.99  & 1.01 & 1.00\\\hline
+447.dealII	& 0.95  & 1.01 & 0.98\\\hline
+450.soplex	& 0.99  & 0.99 & 1.00\\\hline
+453.povray	& 0.97  & 1.00 & 1.00\\\hline
+456.hmmer	& 0.99  & 1.00 & 1.00\\\hline
+458.sjeng	& 1.02  & 1.00 & 1.00\\\hline
+462.libquantum	& 1.00  & 1.00 & 1.00\\\hline
+464.h264ref	& 1.00  & 1.00 & 1.00\\\hline
+470.lbm	        & 1.00  & 1.00 & 1.00\\\hline
+471.omnetpp	& 1.00  & 0.99 & 1.00\\\hline
+473.astar	& 1.00  & 1.02 & 1.00\\\hline
+482.sphinx3	& 0.99  & 1.00 & 1.00\\\hline
+483.xalancbmk	& 1.01  & 0.99 & 1.00\\\hline
+Geomean         & 0.997 & 1.00 & 1.00\\\hline
     \end{tabular}
   \end{center}
-  \caption{Execution time and code size with and without code-motion on SPEC2006 at -Ofast}
+  \caption{Execution time and code size with and without code-motion on SPEC2006.}
   \label{tab:code-size}

### 2017-04-17 paper.tex
tables

--- 2017-04-17 paper.tex
@@
   \begin{center}
-    \begin{tabular}{|l|c|c|}
+    \begin{tabular}{|l|c|c|c|}
       \hline
@@
 % llvm trunk vs. llvm trunk with code-motion
-Benchmark       &\% change in performance & \% change in codesize \\\hline
-400.perlbench	& 1.00  & 1.00 \\\hline
-401.bzip2	& 1.00  & 1.00 \\\hline
-403.gcc	        & 1.00  & 1.01 \\\hline
-429.mcf	        & 1.03  & 1.00 \\\hline
-433.milc	& 1.00  & 1.00 \\\hline
-444.namd	& 1.00  & 1.01 \\\hline
-445.gobmk	& 0.99  & 1.01 \\\hline
-447.dealII	& 0.95  & 1.01 \\\hline
-450.soplex	& 0.99  & 0.99 \\\hline
-453.povray	& 0.97  & 1.00 \\\hline
-456.hmmer	& 0.99  & 1.00 \\\hline
-458.sjeng	& 1.02  & 1.00 \\\hline
-462.libquantum	& 1.00  & 1.00 \\\hline
-464.h264ref	& 1.00  & 1.00 \\\hline
-470.lbm	        & 1.00  & 1.00 \\\hline
-471.omnetpp	& 1.00  & 0.99 \\\hline
-473.astar	& 1.00  & 1.02 \\\hline
-482.sphinx3	& 0.99  & 1.00 \\\hline
-483.xalancbmk	& 1.01  & 0.99 \\\hline
-Geomean         & 0.997 & 1.00 \\\hline
+                &\% performance uplift & \% codesize uplift & \% codesize uplift\\
+Benchmarks      & at -Ofast& at -Ofast & at -Os\\\hline
+400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
+401.bzip2	& 1.00  & 1.00 & 1.00\\\hline
+403.gcc	        & 1.00  & 1.01 & 1.00\\\hline
+429.mcf	        & 1.03  & 1.00 & 1.00\\\hline
+433.milc	& 1.00  & 1.00 & 1.00\\\hline
+444.namd	& 1.00  & 1.01 & 1.00\\\hline
+445.gobmk	& 0.99  & 1.01 & 1.00\\\hline
+447.dealII	& 0.95  & 1.01 & 0.98\\\hline
+450.soplex	& 0.99  & 0.99 & 1.00\\\hline
+453.povray	& 0.97  & 1.00 & 1.00\\\hline
+456.hmmer	& 0.99  & 1.00 & 1.00\\\hline
+458.sjeng	& 1.02  & 1.00 & 1.00\\\hline
+462.libquantum	& 1.00  & 1.00 & 1.00\\\hline
+464.h264ref	& 1.00  & 1.00 & 1.00\\\hline
+470.lbm	        & 1.00  & 1.00 & 1.00\\\hline
+471.omnetpp	& 1.00  & 0.99 & 1.00\\\hline
+473.astar	& 1.00  & 1.02 & 1.00\\\hline
+482.sphinx3	& 0.99  & 1.00 & 1.00\\\hline
+483.xalancbmk	& 1.01  & 0.99 & 1.00\\\hline
+Geomean         & 0.997 & 1.00 & 1.00\\\hline
     \end{tabular}
   \end{center}
-  \caption{Execution time and code size with and without code-motion on SPEC2006 at -Ofast}
+  \caption{Execution time and code size with and without code-motion on SPEC2006.}
   \label{tab:code-size}
@@
       \hline
-  Benchmark       &   Baseline	        &  GCM & \%Increase     \\\hline
+  Benchmarks      &   Baseline	        &  GCM & \%Increase     \\\hline
   400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
@@
   &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
- Benchmark        & Base        & GCM	& \%  &Base     & GCM   &\%    & Base	& GCM	  &\%   & Base	& GCM	& \% \\\hline
+ Benchmarks       & Base        & GCM	& \%  &Base     & GCM   &\%    & Base	& GCM	  &\%   & Base	& GCM	& \% \\\hline
  400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline

### 2017-04-17 paper.tex
tables

@@
 % llvm trunk vs. llvm trunk with code-motion
-                &\% performance uplift & \% codesize uplift & \% codesize uplift\\
+                &\% performance uplift & \% codesize increase & \% codesize increase\\
 Benchmarks      & at -Ofast& at -Ofast & at -Os\\\hline
@@
   \end{center}
-  \caption{Execution time and code size with and without code-motion on SPEC2006.}
+  \caption{Execution time and code size with and without code-motion on SPEC2006. Some performance uplift in mcf, and sjeng
+  while code size improvement in dealII were observed.}
   \label{tab:code-size}

### 2017-04-17 paper.tex
tables

@@
 
-Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
+\citet{dhamdhere1988fast}, \citet{steven1997advanced} mention
 code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
@@
 
-Rosen \cite{rosen1988global} also briefly discuss hoisting computations with
+\citet{rosen1988global} also briefly discuss hoisting computations with
 identical value numbers from immediate successors. Their algorithm iterates on
@@
 
-Click \cite{click1995global} describe aggressive global code motion to first
+\citet{click1995global} describes aggressive global code motion to first
 schedule all the instructions as early as possible. This results in very long
@@
 
-Barany \cite{barany2013} presented a global scheduler with ILP formulation with
+\citet{barany2013} presented a global scheduler with ILP formulation with
 a goal to minimize register pressure. The results they got were not very
@@
 
-Shobaki \cite{shobaki2013} also recently presented a combinatorial global
+\citet{shobaki2013} also recently presented a combinatorial global
 scheduler with reasonable performance improvements. It is possible that both
@@
 global code motion of congruent instructions which uses the liveness analysis as
-illustrated in Das \cite{das2012} and ranking expressions explained in Rosen
-\cite{rosen1988global}.
+illustrated by \citet{das2012} and ranking expressions explained by
+\citet{rosen1988global}.
 
@@
 Code size also remained almost the same as a result of code-motion as shown in
-Table~\cite{tab:code-size}.  The code size listed here is the text size (from
+Table~\ref{tab:code-size}.  The code size listed here is the text size (from
 linux size command) of the final executables for each benchmark.  Some bencmarks
@@
       \hline
-Number of spills  &base(-Ofast) &code-motion(-Ofast) &\%change \\\hline
+Number of spills  &base(-Ofast) &GCM(-Ofast) &\%change \\\hline
 400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
@@
   \end{center}
-  \caption{Number of spills on before and after code-motion on SPEC2006 at -Ofast}
+  \caption{Number of spills before and after global-code-motion (GCM) on SPEC2006 at -Ofast}
   \label{tab:spills}
@@
 
-Spills are listed in Table~\cite{tab:spills}. Spills were reduced by an average of 2\% on the
-\SPEC{} benchmark suite. This is a result of cost-model which limits the code-motion to those
-candidates which do not increase the register pressure (except for loads Section~\cite{sec:cost-models},
-overall spills still decreases.
+Spills are listed in Table~\ref{tab:spills}. Spills were reduced by an average
+of 2\% on the \SPEC{} benchmark suite. This is a result of cost-model which
+limits the code-motion to those candidates which do not increase the register
+pressure (except for loads Section~\ref{sec:cost-models}, overall spills still
+decreases.
 
@@
   \end{center}
-  \caption{Change in compilation time with and without code-motion on SPEC2006 at -Ofast}
+  \caption{Change in compilation time with and without global-code-motion (GCM) on SPEC2006 at -Ofast}
   \label{tab:compile-time}
@@
   \end{center}
-  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
+  \caption{Static metrics before (Base) and after global-code-motion (GCM) on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
   \label{tab:stats}
@@
 
-Some useful statc metric are presented in Table~\cite{tab:stats}, which shows how code-motion
+Some useful statc metric are presented in Table~\ref{tab:stats}, which shows how code-motion
 impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant

### 2017-04-17 paper.tex
move liveness early

@@
 Shobaki not so much, of large compile times which is not feasible for industrial
-compilers like \GCC{} and \LLVM{}.
-
-With the algorithm described in the next section, we got a reduction in register
-spills, improved inliner heuristics, improved compile time and some performance
-improvements on \SPEC{} benchmarks.
+compilers like \GCC{} and \LLVM{}. With the algorithm described in the next
+section, we got a reduction in register spills, improved inliner heuristics,
+improved compile time and some performance improvements on \SPEC{} benchmarks.
 
@@
 
-The GCM pass can be broadly divided into the following steps that we will
-describe in the rest of this section:
-\begin{itemize}
-\item find candidates (congruent instructions) suitable for code-motion,
-\item compute a point in the program where it is both legal and profitable to
-  move the code,
-\item move the code to hoist point or sink point, and
-\item update data structures to continue iterative code motion.
-\end{itemize}
+Dominance graphs, Global Value Numbers and Memory SSA are already available in
+\LLVM{} but there is no facility to infer liveness of virtual registers. For
+that we implemented DJ-Graph data structure that allows us to calculate
+MergeSets which is used in the liveness analysis in our implementation of GCM.
 
@@
 instruction.
+
+The GCM pass can be broadly divided into the following steps that we will
+describe in the rest of this section:
+\begin{itemize}
+\item find candidates (congruent instructions) suitable for code-motion,
+\item compute a point in the program where it is both legal and profitable to
+  move the code,
+\item move the code to hoist point or sink point, and
+\item update data structures to continue iterative code motion.
+\end{itemize}
 

### 2017-04-17 paper.tex
section refs

@@
 \begin{itemize}
-\item find candidates (congruent instructions) suitable for code-motion,
-\item compute a point in the program where it is both legal and profitable to
-  move the code,
-\item move the code to hoist point or sink point, and
+\item find candidates (congruent instructions) suitable for code-motion (Section~\ref{subsec:finding-candidates}),
+\item compute a point in the program where it is both legal
+  (Section~\ref{subsec:legality}) and profitable
+  (Section~\ref{subsec:cost-models}) to move the code,
+\item move the code to hoist point or sink point
+  (Section~\ref{subsec:optimistic}), and
 \item update data structures to continue iterative code motion.
@@
 \subsection{Profitability check (Cost models)}
-\label{sec:cost-models}
+\label{subsec:cost-models}
 After the legality checks have passed, we check if a code-motion is profitable.
@@
 limits the code-motion to those candidates which do not increase the register
-pressure (except for loads Section~\ref{sec:cost-models}, overall spills still
-decreases.
+pressure (except for loads Section~\ref{subsec:cost-models}, overall spills
+still decreases.
 
@@
 not hoist too much we have register pressure aware cost models described in
-Section~\ref{sec:cost-models}. A part of this implementation is already merged
+Section~\ref{subsec:cost-models}. A part of this implementation is already merged
 in LLVM trunk as GVNHoist.cpp. We will post the GCM implementation for review.

### 2017-04-17 paper.tex
fix some references and formatting of cite

--- 2017-04-17 paper.tex
@@
       \hline
-Number of spills  &base(-Ofast) &GCM(-Ofast) &\%change \\\hline
+Number of spills  &base(-Ofast) &code-motion(-Ofast) &\%change \\\hline
 400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
@@
   \end{center}
-  \caption{Number of spills before and after global-code-motion (GCM) on SPEC2006 at -Ofast}
+  \caption{Number of spills on before and after code-motion on SPEC2006 at -Ofast}
   \label{tab:spills}
@@
   \end{center}
-  \caption{Change in compilation time with and without global-code-motion (GCM) on SPEC2006 at -Ofast}
+  \caption{Change in compilation time with and without code-motion on SPEC2006 at -Ofast}
   \label{tab:compile-time}
@@
   \end{center}
-  \caption{Static metrics before (Base) and after global-code-motion (GCM) on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
+  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
   \label{tab:stats}

### 2017-04-18 paper.tex
remove irrelevant old text

--- 2017-04-17 paper.tex
@@
 \subsubsection{Legality of sinking expressions}
+\label{subsec:legality-sink}
 Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
@@
 etc.). The code-motion pass was run twice in the pass pipeline (at -Ofast),
-first after EarlyCSE pass and then after the inliner. Because EarlyCSE removes
+first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
 local redundancies, GCM would have to analyze less redundant instructions, also
@@
 like 473.astar increased in code size by almost 2\%. This is because once the
-code-size of a function decreases it becomes cheaper to inline. As we can see in
-Table-\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
-447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
-more functions inlined.
-
-The results for code-hoisting are listed in
+code-size of a function decreases (due to GCM), it becomes cheaper to inline. As
+we can see in Table-\ref{tab:stats}, four more functions got inlined in
+473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all
+other benchmarks got more functions inlined.
+
+The compile time metrics of GCM are listed in
 Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
@@
 hoisted because each hoisting is performed only when at least one identical
-computation is found.
-
-Loads are hoisted the most followed by scalars, stores and calls in decreasing
-order.  This was the common trend in all our experiments. One reason why loads
-are hoisted the most is the early execution of this pass (before mem2reg) in the
-\LLVM{} pass pipeline. Passes like mem2reg, instcombine might actually remove
-those loads so this order may change should this pass be scheduled later. The number
-of instructions sunk is very less because \LLVM{} has a code sinking pass inside SimplifyCFG
-which is run before the global code-motion and sinks some instructions. Another reason is
-that instructions for which dependencies are not directly available in the successor is
-not sunk for now. Sinking those instructions would require a look ahead on the sinkability
-of the dependencies. We plan to implement this in near future.
-
-Other static metrics are listed in Table~\ref{tab:stats} and in
-Table~\ref{tab:code-motion-metric}.  Here we can see that except for
-rematerializing defs for splitting, which has an overhead of ~2\%, all other
-parameters have less than 1\% overhead. This is to explain why the performance
-does not go down with our implementation (and cost-model) of code hoisting pass.
+computation is found. Loads are hoisted the most followed by scalars, stores and
+calls in decreasing order.  This was the common trend in all our
+experiments. One reason why loads are hoisted the most is the early execution of
+this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg,
+instcombine might actually remove those loads so this order may change should
+this pass be scheduled later. The number of instructions sunk is very less
+because \LLVM{} has a code sinking pass inside SimplifyCFG which runs before the
+GCM and sinks some instructions. Another reason is that instructions for which
+dependencies are not directly available in the successor
+(Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking those
+instructions would require a look ahead on the sinkability of the
+dependencies. We plan to implement this in near future.
 

### 2017-04-18 paper.tex
reword some in a positive stance

@@
 (Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
-GCM) has a (suboptimal) code-sinking which sinks several instructions already.
+GCM) has a code-sinking which already sinks several instructions.
 
@@
 % llvm trunk vs. llvm trunk with code-motion
-                &\% performance uplift & \% codesize increase & \% codesize increase\\
-Benchmarks      & at -Ofast& at -Ofast & at -Os\\\hline
+                &\% performance uplift & \% codesize decrease & \% codesize decrease\\
+Benchmarks      & at -Ofast (high better)& at -Ofast (low better) & at -Os (low better)\\\hline
 400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
@@
       \hline
-Number of spills  &base(-Ofast) &GCM(-Ofast) &\%change \\\hline
+Number of spills  &base(-Ofast) &GCM(-Ofast) &\%change (low better) \\\hline
 400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
@@
       \hline
-  Benchmarks      &   Baseline	        & GCM           & \%Increase \\\hline
+  Benchmarks      &   Baseline	        & GCM           & \%Decrease (low better) \\\hline
   400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
@@
   \end{center}
-  \caption{Change in compilation time with and without global-code-motion (GCM) on SPEC2006 at -Ofast}
+  \caption{Change in total number of instructions executed during the
+    compilation with and without global-code-motion (GCM) on \SPEC{} at -Ofast}
   \label{tab:compile-time}

### 2017-04-18 paper.tex
fix

@@
 expressions in the same iteration but \LLVM{'}s \GVN{} infrastucture does not
-compute equivalence classes in a effective way. We found it simpler to just
+compute equivalence classes in an effective way. We found it simpler to just
 recompute the value numbers and start finding hoistable candidates again.

### 2017-04-18 paper.tex
use gcm / GCM macros

@@
 \def \SPEC {SPEC Cpu 2006}
+\def \gcm {global-code-motion}
+\def \GCM {GCM}
 
@@
 \begin{abstract}
-We present a global code motion (GCM) compiler optimization which schedules
+We present a \gcm{} (\GCM{}) compiler optimization which schedules
 congruent instructions across the program.  Not only this saves code size, it
@@
 infrastructure available in \LLVM{}. The experimental results show an average
-saving of 1\% on overall compile time on \SPEC{}. GCM enables more inlining and
+saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining and
 exposes more loop invariant code motion opportunities in majority of the
@@
 case common hoisting points for all the candidates cannot be found. We also
-formalize how register pressure will reduce as a result of code-motion, and how
-code-motion increases instruction level parallelism thereby enabling more out of
+formalize how register pressure will reduce as a result of \gcm{}, and how
+\gcm{} increases instruction level parallelism thereby enabling more out of
 order execution on modern architectures.
@@
 
-This paper describes code-motion of congruent instructions \cite{briggs1997}, a
+This paper describes \gcm{} of congruent instructions \cite{briggs1997}, a
 technique that uses the information computed for \PRE{} to detect identical
@@
 
-In fact, there are several advantages of GCM.
+In fact, there are several advantages of \GCM{}.
 \begin{itemize}
@@
 high compile time. To the best of our knowledge we have not found any reference
-which explored code-motion of congruent instructions in as much detail as we
+which explored \gcm{} of congruent instructions in as much detail as we
 have done. A part of our implementation (aggressive code hoisting) is already
-merged in \LLVM{} trunk, however, a general implementation of code-motion of
+merged in \LLVM{} trunk, however, a general implementation of \gcm{} of
 congruent (non redundant) expressions is still missing from \GCC{} and \LLVM{}
@@
 that we implemented DJ-Graph data structure that allows us to calculate
-MergeSets which is used in the liveness analysis in our implementation of GCM.
+MergeSets which is used in the liveness analysis in our implementation of \GCM{}.
 
@@
 We compute merge sets of the control flow graph which remains the same through
-the code-motion transformation. For code-motion we only want to know if a use
+the \gcm{} transformation. For \gcm{} we only want to know if a use
 operand is a kill (to compute changes in register pressure.) For that we only
@@
 
-The GCM pass can be broadly divided into the following steps that we will
+The \GCM{} pass can be broadly divided into the following steps that we will
 describe in the rest of this section:
 \begin{itemize}
-\item find candidates (congruent instructions) suitable for code-motion (Section~\ref{subsec:finding-candidates}),
+\item find candidates (congruent instructions) suitable for \gcm{} (Section~\ref{subsec:finding-candidates}),
 \item compute a point in the program where it is both legal
@@
 \label{subsec:cost-models}
-After the legality checks have passed, we check if a code-motion is profitable.
-That takes into account the impact code-motion would have on various parameters
+After the legality checks have passed, we check if a \gcm{} is profitable.
+That takes into account the impact \gcm{} would have on various parameters
 that might affect runtime performance, e.g., impact on live-range, gain in the
@@
 decrease the live-range of the defined register but increase the live-range for
-killed operands. If the live-range after code-motion is less than before it will
+killed operands. If the live-range after \gcm{} is less than before it will
 be moved. Essentially, as long as there is one killed operand, code hoisting
@@
 
-Code-motion basically consists of two parts, i.e., hoisting and sinking. This
+\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
 implementation only moves congruent instructions. An immediate guarantee of this
@@
 (Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
-GCM) has a code-sinking which already sinks several instructions.
+\GCM{}) has a code-sinking which already sinks several instructions.
 
@@
 The complexity of code hoisting is linear in number of instructions that are
-candidates for code-motion, matching the complexity of \PRE{} on \SSA{} form.
+candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
 The analyses computed for this pass are Global Value Numbering, Computation of
 DJ-graph and merge-sets, Marking Barriers, all are linear in number of
-instructions in a function. Liveness analysis expensive but only performed
-on-demand for hoistable candidates so it does not affect the overall compile
+instructions in a function. Liveness analysis is expensive and only performed
+on-demand for hoistable candidates so it does not affect the overall compilation
 time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
 are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
-Barriers for each iteration of the code-motion, we have not seen signinficant
+Barriers for each iteration of the \gcm{}, we have not seen signinficant
 increase in the compilation times. We have also provided appropriate compiler
-flags to expedite code-motion by bailing out with fewer iterations, or skip the
+flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
 liveness based profitability analysis to aggressively move the code as long as
@@
 instruction in the destination point and removes all instructions rendered
-redundant as a result of code-motion.
+redundant as a result of \gcm{}.
 
@@
 % Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
-We ran \SPEC{} with the patch. All the experiments were conducted on x86\_64
-Ubuntu-Linux machine and at -Ofast optimization level. Each sub benchmark was
+We ran \SPEC{} with the patch. All the experiments were conducted on an
+x86\_64-linux machine and at -Ofast optimization level. Each sub benchmark was
 run three times and the best result was taken. We collected execution time,
@@
 Table~\ref{tab:code-motion-metric}.  We also measured compile time
-Table~\ref{tab:compile-time} to see the impact of code-motion.
+Table~\ref{tab:compile-time} to see the impact of \gcm{}.
 
@@
   \end{center}
-  \caption{Execution time and code size with and without code-motion on SPEC2006. Some performance uplift in mcf, and sjeng
-  while code size improvement in dealII were observed.}
+  \caption{Execution time and code size with and without \gcm{}
+    (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng; code size
+    improvement in dealII at -Os.}
   \label{tab:code-size}
@@
 (429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
-etc.). The code-motion pass was run twice in the pass pipeline (at -Ofast),
+etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
 first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
-local redundancies, GCM would have to analyze less redundant instructions, also
-as inliner creates more redundancies, GCM would have more opportunities to move
+local redundancies, \GCM{} would have to analyze less redundant instructions, also
+as inliner creates more redundancies, \GCM{} would have more opportunities to move
 congruent instructions out of branches.
 
-Code size also remained almost the same as a result of code-motion as shown in
+Code size also remained almost the same as a result of \gcm{} as shown in
 Table~\ref{tab:code-size}.  The code size listed here is the text size (from
@@
 like 473.astar increased in code size by almost 2\%. This is because once the
-code-size of a function decreases (due to GCM), it becomes cheaper to inline. As
+code-size of a function decreases (due to \GCM{}), it becomes cheaper to inline. As
 we can see in Table-\ref{tab:stats}, four more functions got inlined in
@@
 
-The compile time metrics of GCM are listed in
+The compile time metrics of \GCM{} are listed in
 Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
@@
 because \LLVM{} has a code sinking pass inside SimplifyCFG which runs before the
-GCM and sinks some instructions. Another reason is that instructions for which
+\GCM{} and sinks some instructions. Another reason is that instructions for which
 dependencies are not directly available in the successor
@@
       \hline
-Number of spills  &base(-Ofast) &GCM(-Ofast) &\%change (low better) \\\hline
+Number of spills  &base(-Ofast) &\GCM{}(-Ofast) &\%change (low better) \\\hline
 400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
@@
   \end{center}
-  \caption{Number of spills before and after global-code-motion (GCM) on SPEC2006 at -Ofast}
+  \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at -Ofast}
   \label{tab:spills}
@@
 of 2\% on the \SPEC{} benchmark suite. This is a result of cost-model which
-limits the code-motion to those candidates which do not increase the register
+limits the \gcm{} to those candidates which do not increase the register
 pressure (except for loads Section~\ref{subsec:cost-models}, overall spills
@@
       \hline
-  Benchmarks      &   Baseline	        & GCM           & \%Decrease (low better) \\\hline
+  Benchmarks      &   Baseline	        & \GCM{}           & \%Decrease (low better) \\\hline
   400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
@@
   \caption{Change in total number of instructions executed during the
-    compilation with and without global-code-motion (GCM) on \SPEC{} at -Ofast}
+    compilation with and without \gcm{} (\GCM{}) on \SPEC{} at -Ofast}
   \label{tab:compile-time}
@@
 The compilation time actually reduced for most of the benchmarks as a result of code motion.
-This is because as code-motion removes (congruent) instructions, the number of instructions
+This is because as \gcm{} removes (congruent) instructions, the number of instructions
 to be processed by later compilation passes also reduces. The compilation time listed here
@@
   &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
- Benchmarks       & Base        & GCM	& \%  &Base     & GCM   &\%    & Base	& GCM	  &\%   & Base	& GCM	& \% \\\hline
+ Benchmarks       & Base        & \GCM{}	& \%  &Base     & \GCM{}   &\%    & Base	& \GCM{}	  &\%   & Base	& \GCM{}	& \% \\\hline
  400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
@@
   \end{center}
-  \caption{Static metrics before (Base) and after global-code-motion (GCM) on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
+  \caption{Static metrics before (Base) and after \gcm{} (\GCM{}) on
+    \SPEC{} at -Ofast. The \% Column show percentage increase in metric
+    w.r.t. Base. \GCM{} improves LICM and inlining}
   \label{tab:stats}
@@
 
-Some useful statc metric are presented in Table~\ref{tab:stats}, which shows how code-motion
+Some useful statc metric are presented in Table~\ref{tab:stats}, which shows how \gcm{}
 impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
@@
       \hline
- code-motion  &\multicolumn{2}{|c|}{Total Instructions} &\multicolumn{2}{|c|}{Scalars}  & \multicolumn{2}{|c|}{Loads} & \multicolumn{1}{|c|}{Stores} &Calls   &Instructions    \\\hline
+ \GCM{}  &\multicolumn{2}{|c|}{Total Instructions} &\multicolumn{2}{|c|}{Scalars}  & \multicolumn{2}{|c|}{Loads} & \multicolumn{1}{|c|}{Stores} &Calls   &Instructions    \\\hline
  metric           &  H          &R        &H      &R       &H           &R              &H=R    &H & sunk \\\hline
@@
   \end{center}
-  \caption{GCM stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are hoisted the most followed by scalars, stores and calls}
+  \caption{\GCM{} stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are hoisted the most followed by scalars, stores and calls}
   \label{tab:code-motion-metric}
@@
 Section~\ref{subsec:cost-models}. A part of this implementation is already merged
-in LLVM trunk as GVNHoist.cpp. We will post the GCM implementation for review.
+in LLVM trunk as GVNHoist.cpp. We will post the \GCM{} implementation for review.
 
@@
 and results in multiple data structures to be recomputed. This can be improved
-by ranking the computations \cite{rosen1988global}. Also, since GCM runs very
+by ranking the computations \cite{rosen1988global}. Also, since \GCM{} runs very
 early in the pass pipeline, it will be good to evaluate the codesize/performance
@@
 
-With the implementation of GCM in \LLVM{}, the passes which rely on the
+With the implementation of \GCM{} in \LLVM{}, the passes which rely on the
 code-size/instruction-count to make optimization decisions needs to be

### 2017-04-18 paper.tex
fix

@@
 experiments. One reason why loads are hoisted the most is the early execution of
-this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg,
-instcombine might actually remove those loads so this order may change should
-this pass be scheduled later. The number of instructions sunk is very less
-because \LLVM{} has a code sinking pass inside SimplifyCFG which runs before the
-\GCM{} and sinks some instructions. Another reason is that instructions for which
+this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg and
+instcombine might actually remove those loads and the number of hoisted loads may change should
+the \GCM{} pass be scheduled later. The number of instructions sunk is very less
+because \LLVM{} has a code sinking and hoisting transform inside SimplifyCFG pass which runs before
+\GCM{} and sinks and hoists some instructions with some rather severe limitations. Another reason is that instructions for which
 dependencies are not directly available in the successor

### 2017-04-18 paper.tex
clarify comment about simplifyCFG sink/hoist

@@
 this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg and
-instcombine might actually remove those loads and the number of hoisted loads may change should
-the \GCM{} pass be scheduled later. The number of instructions sunk is very less
-because \LLVM{} has a code sinking and hoisting transform inside SimplifyCFG pass which runs before
-\GCM{} and sinks and hoists some instructions with some rather severe limitations. Another reason is that instructions for which
-dependencies are not directly available in the successor
-(Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking those
-instructions would require a look ahead on the sinkability of the
+instcombine might actually remove those loads and the number of hoisted loads
+may change should the \GCM{} pass be scheduled later. We can see a few sunk
+instructions even if \LLVM{} has a code sinking and hoisting transforms in the
+SimplifyCFG pass which runs before \GCM{}: this is because sinking and hoisting
+in SimplifyCFG have some rather severe limitations to make the analysis and code
+transform very fast in order to be able to run the pass several times.  Another
+reason is that instructions for which dependencies are not directly available in
+the successor (Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking
+those instructions would require a look ahead on the ability to sink the
 dependencies. We plan to implement this in near future.

### 2017-04-18 paper.tex
use a few more gcm macros

@@
 
-\citet{click1995global} describes aggressive global code motion to first
+\citet{click1995global} describes aggressive \gcm{} to first
 schedule all the instructions as early as possible. This results in very long
@@
 
-\section{Code motion}
-
-The algorithm for code motion uses several common representations of the
+\section{Global code motion}
+
+The algorithm for \gcm{} uses several common representations of the
 program that we shortly describe below:
@@
   (Section~\ref{subsec:optimistic}), and
-\item update data structures to continue iterative code motion.
+\item update data structures to continue iterative \gcm{}.
 \end{itemize}
@@
 will either decrease or preserve the register pressure as long as there is one
-operand killed at most.  The following example explains how code motion of
+operand killed at most.  The following example explains how \gcm{} of
 identical computations can reduce the register pressure.
@@
 
-\subsection{Code generation for the optimistic code motion algorithm}
+\subsection{Code generation for the optimistic \gcm{} algorithm}
 \label{subsec:optimistic}
@@
 Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
-global code motion of congruent instructions which uses the liveness analysis as
+\gcm{} of congruent instructions which uses the liveness analysis as
 illustrated by \citet{das2012} and ranking expressions explained by
@@
 \label{sec:future-work}
-We have presented the global code motion of congruent computations in SSA
+We have presented the \gcm{} of congruent computations in SSA
 form. We saw that it improves inlining and LICM opportunities, reduces register

### 2017-04-18 paper.tex
some more rewording

@@
         if they are profitable to be hoisted at hoist-point:
-          proceed to check hoistability of subsequent instruction with same VN
+          proceed to check ability to hoist of subsequent instruction with same VN
         else
-          proceed to check if I2 and I3 (if available) is hoistable
-
-  For each hoistable VN:
+          proceed to check if I2 and I3 (if available) can be hoisted
+
+  For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
     move the first one I1 to the hoist-point
-    update the use of all other candidates to refer to I1
-    remove all others.
+    update the use of all other candidates (I2, ..., Ik) to refer to I1
+    remove all other instructions (I2, ..., Ik)
     if I1 has memory references (a load, store etc.)
@@
 \label{sec:future-work}
-We have presented the \gcm{} of congruent computations in SSA
-form. We saw that it improves inlining and LICM opportunities, reduces register
-spills, it also improves performance in some cases. To preserve performance and
-not hoist too much we have register pressure aware cost models described in
-Section~\ref{subsec:cost-models}. A part of this implementation is already merged
-in LLVM trunk as GVNHoist.cpp. We will post the \GCM{} implementation for review.
+We have presented the \gcm{} of congruent computations in SSA form. We saw that
+it improves inlining and LICM opportunities, reduces register spills, it also
+improves performance in some cases. To preserve performance and not hoist too
+much we have presented some cost models to lower register pressure. A part of
+this implementation is already merged in LLVM trunk as GVNHoist.cpp, and we
+posted the \GCM{} implementation for review.
 
@@
 finds a fixed point when no more candidates are available. Even this
-implementation converges quickly and no compile time regression have been
+implementation converges quickly and no compile time regressions have been
 observed because of code hoisting pass. This is not the most optimal approach
@@
 With the implementation of \GCM{} in \LLVM{}, the passes which rely on the
-code-size/instruction-count to make optimization decisions needs to be
+code-size/instruction-count to make optimization decisions need to be
 revisited. The first candidate would be the inliner. We have seen different

### 2017-04-18 paper.tex
tweak some algorithms to stand uncut on a page

@@
 // lnode) is not the subset of the merge-set of lnode.
-bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet Merge)
+bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet MergeSet)
   for e in (all incoming edges to lnode)
     if (e is in JE && e in V)
-      if (Merge(Source Node of e) is not a subset of Merge(lnode))
+      if (MergeSet(Source Node of e) is not a subset of MergeSet(lnode))
         return true
@@
         while (DL(tmp) >= DL(tnode))
-          Merge(tmp) = Merge(tmp) U Merge(tnode) U {tnode}
+          MergeSet(tmp) = MergeSet(tmp) U MergeSet(tnode) U {tnode}
           lnode = tmp
           tmp = dom-parent(tmp)
-        repeat = still_inconsistent(lnode, JE, Merge)
+        repeat = still_inconsistent(lnode, JE, MergeSet)
   return repeat
@@
 
-  Ms(N) = null
-  // Mergeset of N is the union of merge sets of its successors
+  MergeSet(N) = null
+  // The merge-set of N is the union of merge-sets of its successors.
   for W in successors(N)
-    Ms(N) = Ms(N) U Mr(W)
+    MergeSet(N) = MergeSet(N) U MergeSet(W)
 
@@
     while (B != null and B != def_bb(A))
-      if B in Ms(N) return true
+      if B in MergeSet(N) return true
       B = dom-parent(B)
@@
   for each basic block B in the dominator tree starting at hoist-point:
-    if Worklist is empty
-      return false // Path exists! not downward safe
+    if Worklist is empty, return false // Path exists! not downward safe
     if B is in Worklist:
@@
       remove subtree with root at B // Available at B => downward safe from B
-    if B is a leaf basic block:
-      return false  // Path exists! not downward safe
-    if B dominates hoist-point:
-      return false  // Back edge
+    if B is a leaf basic block, return false  // Path exists! not downward safe
+    if B dominates hoist-point, return false  // Back edge
   return true
@@
     Compute GVN of each expression in the function
-    int N = doCodeHoisting()
-  while (N > 0) // If any of candidates were hoisted then repeat hoisting
+  while (doCodeHoisting() > 0) // If any of candidates were hoisted then repeat hoisting
   doCodeSinking()
@@
 
-% Explain the overall algorithm, explain how anticipability is computed on demand
 \begin{verbatim}
 int doCodeHoisting()
-  For each value number VN with 2 or more instructions:
+  For each value number VN with 2 or more instructions (I1, I2, ..., Ik)
     sort the instructions according to DFSIn numbers
-    if first two I1, I2 are hoistable:
-      if they are downward-safe at hoist-point:
-        if they are profitable to be hoisted at hoist-point:
+    if first two I1, I2 are hoistable and if they are downward-safe at hoist-point
+       and if they are profitable to be hoisted at hoist-point
       proceed to check ability to hoist of subsequent instruction with same VN
@@
       proceed to check if I2 and I3 (if available) can be hoisted
-
   For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
@@
     remove all other instructions (I2, ..., Ik)
-    if I1 has memory references (a load, store etc.)
+    if I1 has memory references (a load, store, call, etc.)
       update MemorySSA of I1 and others to point to the MemorySSA reference of I1
@@
     update statistics
-
-  return number_of_hoisted_candidates
+  return number of hoisted instructions
 \end{verbatim}
@@
 void doCodeSinking()
-  For each value number VN with 2 or more instructions:
+  For each value number VN with 2 or more instructions
     if there are two instructions with same immediate successor as post-dominator
@@
            mark VN as sinkable
-
-  For each pair I1, I2 of sinkable instructions:
-    Move the first one I1 to the sink point which is just after all the PHI-nodes
-    in the post-dominator
-    update all the uses of PHI (PN) with I1
+  For each pair (I1, I2) of sinkable instructions
+    Move I1 to the sink point which is just after all the PHI-nodes in the post-dominator
+    update all the uses of PHI node (PN) with I1
     Remove I2 and PN
@@
     update statistics
-
 \end{verbatim}

### 2017-04-18 paper.tex
use some more macros

@@
 \def \GCM {GCM}
-
+\def \xlinux {x86\_64-linux}
+\def \ooo {out-of-order}
 %% Author information
@@
 infrastructure available in \LLVM{}. The experimental results show an average
-saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining and
-exposes more loop invariant code motion opportunities in majority of the
+saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining
+and exposes more loop invariant code motion opportunities in majority of the
 benchmarks. We have also seen execution time improvements in a few of \SPEC{}
 benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
-\SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
+\SPEC{} bencmarks suite when compiled for \xlinux{}. This is an optimistic
 algorithm in the sense that it considers all congruent instructions in a
@@
 formalize how register pressure will reduce as a result of \gcm{}, and how
-\gcm{} increases instruction level parallelism thereby enabling more out of
-order execution on modern architectures.
+\gcm{} increases instruction level parallelism thereby enabling more \ooo{}
+execution on modern architectures.
 \end{abstract}
@@
 computations. Code hoisting, for example, can reduce the critical path length of
-execution in out-of-order machines. As more instructions are available at the
+execution in \ooo{} machines. As more instructions are available at the
 hoisting point, the hardware has more independent instructions to reorder. The
@@
 expensive than the addition, subtraction, and multiplication operations. The
-total latency of computation across each branch is: $C_{div} + 2(C_{sub} +
-C_{mul})$.  For an out-of-order processor with two add units and two multiply
+total latency of computation across each branch is: $C_{div} + 2(C_{sub} + C_{mul})$.
+For an \ooo{} processor with two add units and two multiply
 units, the total latency of computation is $C_{div} + C_{sub} + C_{mul}$.
@@
   bubbles;
-\item it reduces branch misprediction penalty on out-of-order processors with
+\item it reduces branch misprediction penalty on \ooo{} processors with
   speculative execution of branches: by hoisting or sinking expressions out of
@@
 We ran \SPEC{} with the patch. All the experiments were conducted on an
-x86\_64-linux machine and at -Ofast optimization level. Each sub benchmark was
+\xlinux{} machine and at -Ofast optimization level. Each sub benchmark was
 run three times and the best result was taken. We collected execution time,
@@
 conditional clauses, the overall performance of that benchmark improves by
-$15\%$ on an out-of-order processor due to increased instruction level
+$15\%$ on an \ooo{} processor due to increased instruction level
 parallelism, and better scheduling of the instructions, accommodating for the

### 2017-04-18 paper.tex
grammar

@@
 
-There have been a lot of work both in industry and academia to hoist and sink
+There has been a lot of work both in industry and academia to hoist and sink
 code out of branches, and in general global scheduling
@@
 high compile time. To the best of our knowledge we have not found any reference
-which explored \gcm{} of congruent instructions in as much detail as we
-have done. A part of our implementation (aggressive code hoisting) is already
-merged in \LLVM{} trunk, however, a general implementation of \gcm{} of
-congruent (non redundant) expressions is still missing from \GCC{} and \LLVM{}
-trunk. The main contributions of this paper are:
+which explored \gcm{} of congruent instructions in as much detail us. A part of
+our implementation (aggressive code hoisting) is already merged
+in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
+redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
+contributions of this paper are:
 \begin{itemize}
-\item a new optimistic algorithm to move congruent instructiosn out of branches;
+\item a new optimistic algorithm to move congruent instructions out of branches;
 \item a cost model to reduce interference and hence, reduce spills;
@@
 
-We compute merge sets of the control flow graph which remains the same through
+We compute merge sets, of the control flow graph, which remain same throughout
 the \gcm{} transformation. For \gcm{} we only want to know if a use
@@
   (Section~\ref{subsec:optimistic}), and
-\item update data structures to continue iterative \gcm{}.
+\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
 \end{itemize}
@@
 Barriers for each iteration of the \gcm{}, we have not seen signinficant
-increase in the compilation times. We have also provided appropriate compiler
+increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
+provided appropriate compiler
 flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
 liveness based profitability analysis to aggressively move the code as long as
-they are legal (Section~\ref{sec:experimental-results}).
+they are legal.
 
@@
 % Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
-We ran \SPEC{} with the patch. All the experiments were conducted on an
-\xlinux{} machine and at -Ofast optimization level. Each sub benchmark was
-run three times and the best result was taken. We collected execution time,
-code-size which is listed in Table~\ref{tab:code-size}, other compiler stats are
-listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
+For evaluation of \gcm{}, we built \SPEC{} with the patch. All the experiments
+were conducted on an \xlinux{} machine and at -Ofast optimization level. Each
+sub benchmark was run three times and the best result was taken. We collected
+execution time and code-size which is listed in Table~\ref{tab:code-size}, other
+compiler stats are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
 Table~\ref{tab:code-motion-metric}.  We also measured compile time
@@
 
-Overall there was only a minor change in performance regressed with a few ups
+Overall there was only a minor change in the performance numbers with a few improvements 
 (429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
@@
 number of instructions removed is greater or equal to the number of instructions
-hoisted because each hoisting is performed only when at least one identical
+hoisted because each code-motion is performed only when at least one identical
 computation is found. Loads are hoisted the most followed by scalars, stores and
@@
 experiments. One reason why loads are hoisted the most is the early execution of
-this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg and
+this pass (before mem2reg pass which scalarizes some memory references) in the \LLVM{} pass pipeline. Passes like mem2reg and
 instcombine might actually remove those loads and the number of hoisted loads
@@
 The compilation time actually reduced for most of the benchmarks as a result of code motion.
-This is because as \gcm{} removes (congruent) instructions, the number of instructions
+This is because, as \gcm{} removes (congruent) instructions, the number of instructions
 to be processed by later compilation passes also reduces. The compilation time listed here
-is the total instruction count as outputted from valgrind (--tool=cachegrind).
+is the total instruction count as outputted from valgrind (--tool=cachegrind) \cite{valgrind}.
 
@@
 
-Some useful statc metric are presented in Table~\ref{tab:stats}, which shows how \gcm{}
+Some useful static metrics are presented in Table~\ref{tab:stats}, which shows how \gcm{}
 impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
@@
 significantly in several benchmarks, although number of instructions sunk did not change by much.
-Number of inlined functiosn also improved for most benchmarks. In general when more functions
+Number of inlined functions also improved for most benchmarks. In general when more functions
 were inlined, more functions were deleted as in 400.perlbench, 445.gobmk.
@@
 While benchmarking \SPEC{} we see both increase as well as decrease in
-the codesizes of the final binaries. Since the pass runs early, it affects many
-optimizations which rely on the number of instructions, lenth of the use-def
+the codesize of the final binaries. Since the pass runs early, it affects many
+optimizations which rely on the number of instructions, length of the use-def
 chain, and other metrics. For instance, the inliner is impacted by a decrease in
@@
 \label{sec:future-work}
+
 We have presented the \gcm{} of congruent computations in SSA form. We saw that
 it improves inlining and LICM opportunities, reduces register spills, it also
-improves performance in some cases. To preserve performance and not hoist too
-much we have presented some cost models to lower register pressure. A part of
-this implementation is already merged in LLVM trunk as GVNHoist.cpp, and we
-posted the \GCM{} implementation for review.
+improves performance in some cases. To preserve performance and not hoist/sink
+too much, we have implemented a register pressure aware cost model as described
+in Section~\ref{subsec:cost-models}. A part of this implementation is already
+merged in LLVM trunk as GVNHoist.cpp and we posted the \GCM{} implementation for
+review
 
@@
 redundancies to the GVN-PRE.  In general it is a good idea to start with lower
-ranked expression first such that maximum hoisting can happen in one iteration,
+ranked expressions first such that maximum hoisting can happen in one iteration,
 however, current implementation does not rank the expressions and iteratively
 finds a fixed point when no more candidates are available. Even this
-implementation converges quickly and no compile time regressions have been
-observed because of code hoisting pass. This is not the most optimal approach
+implementation converges quickly and no compile time regression have been
+observed. This is not the most optimal approach in terms of compile time complexity
 and results in multiple data structures to be recomputed. This can be improved
@@
 With the implementation of \GCM{} in \LLVM{}, the passes which rely on the
-code-size/instruction-count to make optimization decisions need to be
-revisited. The first candidate would be the inliner. We have seen different
-inlining decisions in Table~\ref{tab:code-size}, before and after code-hoisting
-was enabled.  Since inliner has several magic numbers tuned for the previous
-pass layout, it would need some improvement.
+code-size or instruction-counts to make optimization decisions need to be
+revisited. The first candidate would be the inliner. We have seen difference
+in inlining decisions (Table~\ref{tab:code-size}), when \gcm{}
+was enabled.  Since inliner has several cost-models tuned for the previous
+pass layout, that would need some improvement.
 

### 2017-04-18 paper.tex
add comment about 15% perf uplift on motivating example

@@
 exposes redundancies in some cases, it exposes more instruction level
-parallelism in the basic block to which instructions are moved, and it enables
+parallelism in the basic-block to which instructions are moved, and it enables
 other passes like loop invariant motion to remove more redundancies. The cost
@@
 phase that reduces the number of run-time computations.  Classical scalar
-optimizations like CSE \cite{dragonbook} work very well on single basic blocks but
-when it comes to detect redundancies across basic blocks these techniques fall
+optimizations like CSE \cite{dragonbook} work very well on single basic-blocks but
+when it comes to detect redundancies across basic-blocks these techniques fall
 short: more complex passes like GCSE and \PRE{} have been designed to handle these
@@
 
-This paper describes \gcm{} of congruent instructions \cite{briggs1997}, a
+This paper describes \gcm{} (\GCM{}) of congruent instructions \cite{briggs1997}, a
 technique that uses the information computed for \PRE{} to detect identical
@@
 float fun(float d, float min, float max, float a) {
-  float tmin, tmax, inv;
-
-  inv = 1.0f / d;
+  float tmin, tmax;
+  float inv = 1.0f / d;
   if (inv >= 0) { tmin = (min - a) * inv; tmax = (max - a) * inv; }
@@
 }
-
 \end{verbatim}
@@
 of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
-$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}.
-
-%GCC: 23286 has interesting test cases.
-
-In fact, there are several advantages of \GCM{}.
+$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
+seen the performance of an \ooo{} processor going up by $15\%$ on a benchmark
+containing the above motivating example running in a tight loop.  In fact, there
+are several advantages of \GCM{}:
 \begin{itemize}
@@
 \item it may improve loop vectorization by reducing a loop with control flow to
-  a loop with a single BB, should all the instructions in a conditional get
+  a loop with a single basic-block, should all the instructions in a conditional get
   hoisted or sinked;
@@
 code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
-basic blocks: the instructions of two sibling basic blocks are read at the same
+basic-blocks: the instructions of two sibling basic-blocks are read at the same
 time, and all the instructions of the blocks are hoisted to the common parent
@@
 code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
-computed which are hoisted in a basic block where the expression is
+computed which are hoisted in a basic-block where the expression is
 anticipable (all the operands are available). This algorithm would hoist as far
@@
 expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
-basic block B, i.e., values computed on all paths from B to exit and
+basic-block B, i.e., values computed on all paths from B to exit and
 AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
@@
 live ranges which is mitigated by again scheduling all the instructions as late
-as possible. They report sppeedup of as high as $23\%$.
+as possible. They report a speedup of as high as $23\%$.
 
@@
 \cite{das2012}.  It is very efficient for computing liveness and does not
-require any bitvectors to be maintained for each basic block. The underlying
+require any bitvectors to be maintained for each basic-block. The underlying
 simplicity of liveness computation is due to \SSA{} form, where the the values
@@
 (where we insert a PHI). The DJ-graph contains both these edges which allows for
-computation of merge-sets for each basic block, i.e., a set of all basic blocks
-where the values can flow from a particular basic block. We have implemented the
+computation of merge-sets for each basic-block, i.e., a set of all basic-blocks
+where the values can flow from a particular basic-block. We have implemented the
 merge-set computation based on a DJ-graph as illustrated in \cite{das2005}. A
@@
 need to know whether the use is also required later in the execution path. A
-variable is live out of a basic block B if it is used in the merge set of B. We
+variable is live out of a basic-block B if it is used in the merge set of B. We
 compute the live-out relation on-demand when profitability of hoistable/sinkable
@@
 \begin{verbatim}
-// Return true if variable A is liveout from basic block N
+// Return true if variable A is liveout from basic-block N
 bool isLiveOutUsingMergeSet(Node N, Variable A)
   if (A is defined in N)
-    if (A is used outside any basic block other than N)
+    if (A is used outside any basic-block other than N)
       return true
@@
 terminate. These two cases were missing from the algorithm and we came across
-them during implementation. If a variable is not live-out of a basic block, it
-still may be used later in the basic block: to establish whether a use of a
-variable is a kill we iterate on all the subsequent instructions in a basic
-block checking for uses, the \emph{isKill} function uses
+them during implementation. If a variable is not live-out of a basic-block, it
+still may be used later in the basic-block: to establish whether a use of a
+variable is a kill we iterate on all the subsequent instructions in a
+basic-block checking for uses, the \emph{isKill} function uses
 \emph{isLiveOutUsingMergeSet} to find out if an operand is killed in an
@@
 and anticipable sets as computed by \GCC{'}s code-hoisting but that would be a lot
-of data structures to maintain at each basic block level. For \GCC{} it makes sense
+of data structures to maintain at each basic-block level. For \GCC{} it makes sense
 because their code-hoisting is integrated with \GVN{}-\PRE{} which already has those
@@
 Barriers are based on the concept of pinned instructions \cite{click1995global}
-but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is
-actually an extended basic block because there might be calls in the middle of a
-basic block which might not return. Essentially, any instruction that cannot
+but extended to adapt to \LLVM{} IR. Since a basic-block in \LLVM{} IR is
+actually an extended basic-block because there might be calls in the middle of a
+basic-block which might not return. Essentially, any instruction that cannot
 guarantee progress is marked as a barrier. In the absence of context, as in our
@@
 computing barriers:
-for each basic block B in a function:
+for each basic-block B in a function:
   barrier_found = false
@@
 computing downward-safety of hoistable instructions at hoist-point
-  Worklist = list of all basic blocks of hoistable instructions
-  for each basic block B in the dominator tree starting at hoist-point:
+  Worklist = list of all basic-blocks of hoistable instructions
+  for each basic-block B in the dominator tree starting at hoist-point:
     if Worklist is empty, return false // Path exists! not downward safe
@@
       remove subtree with root at B // Available at B => downward safe from B
-    if B is a leaf basic block, return false  // Path exists! not downward safe
+    if B is a leaf basic-block, return false  // Path exists! not downward safe
     if B dominates hoist-point, return false  // Back edge
@@
 There is a special case for hoisting load instructions when the hoist-point is
-the predecessor basic block for all the loads. Even if the register pressure
+the predecessor basic-block for all the loads. Even if the register pressure
 would increase, we prefer to hoist loads in this case. The reason being, that
@@
 because of more inlining). The algorithm prefers hoisting to sinking. If the
-dependency of a hoistable candidate is in the same basic block as the candidate,
+dependency of a hoistable candidate is in the same basic-block as the candidate,
 then the dependency must also be hoistable, otherwise hoisting will be illegal or

### 2017-04-18 paper.tex
new page to put references

@@
 short: more complex passes like GCSE and \PRE{} have been designed to handle these
-cases based on dataflow analysis \cite{morel1979global}.  At first these
+cases based on data-flow analysis \cite{morel1979global}.  At first these
 techniques were described in the classical data-flow analysis framework, and
@@
 
-\citet{dhamdhere1988fast}, \citet{steven1997advanced} mention
-code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
+\citet{dhamdhere1988fast} and \citet{steven1997advanced} mention
+code hoisting in a data-flow framework. A list of Very Busy Expressions (VBE)
 computed which are hoisted in a basic-block where the expression is
@@
 model will be required.  Also the description of VBE is based on the classic
-dataflow model and an adaptation to a sparse \SSA{} representation is required.
+data-flow model and an adaptation to a sparse \SSA{} representation is required.
 
@@
 same target architecture. Also, both suffer from the same problem, although
-Shobaki not so much, of large compile times which is not feasible for industrial
+Shobaki not so much, of large compile times which would not suit in industrial
 compilers like \GCC{} and \LLVM{}. With the algorithm described in the next
 section, we got a reduction in register spills, improved inliner heuristics,
-improved compile time and some performance improvements on \SPEC{} benchmarks.
+improved compile time, and show performance improvements on \SPEC{} benchmarks.
 
@@
 \begin{verbatim}
-
 // Return true if the merge-set of source node of a visited J-edge (incoming edge of
@@
 // Construct merge-set of each node in Control-Flow-Graph G of a function.
-void  constructMergeSet(CFG G) {
+void constructMergeSet(CFG G)
   B = Breadth First Order of G
@@
   DL = List of Path length (from root) of each node in G.
-
   do // Call until a fixed point is reached.
@@
   while (Repeat)
-}
 \end{verbatim}
@@
 
-\section{Conclusion and Future Work}
+\section{Conclusions and Future Work}
 \label{sec:future-work}
@@
 
+\newpage
+
 %% Bibliography

### 2017-04-18 paper.tex
some more rewording

@@
 \begin{verbatim}
+
 // Return true if the merge-set of source node of a visited J-edge (incoming edge of
@@
 // Construct merge-set of each node in Control-Flow-Graph G of a function.
-void constructMergeSet(CFG G)
+void  constructMergeSet(CFG G) {
   B = Breadth First Order of G
@@
   DL = List of Path length (from root) of each node in G.
+
   do // Call until a fixed point is reached.
@@
   while (Repeat)
+}
 \end{verbatim}
@@
 
-\section{Conclusions and Future Work}
+\section{Conclusion and Future Work}
 \label{sec:future-work}
@@
 
-\newpage
-
 %% Bibliography

### 2017-04-18 paper.tex
compile time metric down

--- 2017-04-18 paper.tex
@@
 
-The compile time metrics of \GCM{} are listed in
-Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
-loads, stores and calls hoisted as well as removed. For each category, the
-number of instructions removed is greater or equal to the number of instructions
-hoisted because each code-motion is performed only when at least one identical
-computation is found. Loads are hoisted the most followed by scalars, stores and
-calls in decreasing order.  This was the common trend in all our
-experiments. One reason why loads are hoisted the most is the early execution of
-this pass (before mem2reg pass which scalarizes some memory references) in the \LLVM{} pass pipeline. Passes like mem2reg and
-instcombine might actually remove those loads and the number of hoisted loads
-may change should the \GCM{} pass be scheduled later. We can see a few sunk
-instructions even if \LLVM{} has a code sinking and hoisting transforms in the
-SimplifyCFG pass which runs before \GCM{}: this is because sinking and hoisting
-in SimplifyCFG have some rather severe limitations to make the analysis and code
-transform very fast in order to be able to run the pass several times.  Another
-reason is that instructions for which dependencies are not directly available in
-the successor (Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking
-those instructions would require a look ahead on the ability to sink the
-dependencies. We plan to implement this in near future.
-
 \begin{table}[h!]
@@
 
+The compile time metrics of \GCM{} are listed in
+Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
+loads, stores and calls hoisted as well as removed. For each category, the
+number of instructions removed is greater or equal to the number of instructions
+hoisted because each code-motion is performed only when at least one identical
+computation is found. Loads are hoisted the most followed by scalars, stores and
+calls in decreasing order.  This was the common trend in all our
+experiments. One reason why loads are hoisted the most is the early execution of
+this pass (before mem2reg pass which scalarizes some memory references) in the
+\LLVM{} pass pipeline. Passes like mem2reg and instcombine might actually remove
+those loads and the number of hoisted loads may change should the \GCM{} pass be
+scheduled later. We can see a few sunk instructions even if \LLVM{} has a code
+sinking and hoisting transforms in the SimplifyCFG pass which runs before
+\GCM{}: this is because sinking and hoisting in SimplifyCFG have some rather
+severe limitations to make the analysis and code transform very fast in order to
+be able to run the pass several times.  Another reason is that instructions for
+which dependencies are not directly available in the successor
+(Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking those
+instructions would require a look ahead on the ability to sink the
+dependencies. We plan to implement this in near future.
+
 \begin{table}[h!]

### 2017-04-18 paper.tex
fix page layout
removed redundancy

@@
 
+\newpage
 After no more candidates are hoistable, sinking is performed. Sinking is only
@@
 like 473.astar increased in code size by almost 2\%. This is because once the
-code-size of a function decreases (due to \GCM{}), it becomes cheaper to inline. As
-we can see in Table-\ref{tab:stats}, four more functions got inlined in
-473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all
-other benchmarks got more functions inlined.
+code-size of a function decreases (due to \GCM{}), it becomes cheaper to
+inline. As we can see in Table-\ref{tab:stats}, four more functions got inlined
+in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
+all other benchmarks got more functions inlined. Also, since the pass runs
+early, it affects many optimizations which rely on the number of instructions,
+length of the use-def chain, and other metrics.
 
@@
   \end{center}
-  \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at -Ofast}
+  \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at
+    -Ofast. Spills reduced by 2\% overall.}
   \label{tab:spills}
@@
 
-While benchmarking \SPEC{} we see both increase as well as decrease in
-the codesize of the final binaries. Since the pass runs early, it affects many
-optimizations which rely on the number of instructions, length of the use-def
-chain, and other metrics. For instance, the inliner is impacted by a decrease in
-the number of instructions in the caller and callee, as its heuristics estimate
-the size of functions to be inlined.
-
 The code shown in Section~\ref{sec:intro} is a reduced example that appears in a

### 2017-04-18 paper.tex
ispell-buffer

@@
 benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
-\SPEC{} bencmarks suite when compiled for \xlinux{}. This is an optimistic
+\SPEC{} benchmarks suite when compiled for \xlinux{}. This is an optimistic
 algorithm in the sense that it considers all congruent instructions in a
@@
 %% comma separated list
-\keywords{Optimizng Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional
+\keywords{Optimizing Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional
 
@@
   a loop with a single basic-block, should all the instructions in a conditional get
-  hoisted or sinked;
+  hoisted or sunk;
 \item it enables more loop invariant code motion (LICM): as LICM passes, in
@@
 instructions to bisimulate and second, the equivalence of instructions is
-computed by comparing the operands which is neither general nor scalabale.
+computed by comparing the operands which is neither general nor scalable.
 
@@
 scheduler with reasonable performance improvements. It is possible that both
-Barany and Shobaki's implmentation will have similar results when compiled for
+Barany and Shobaki's implementation will have similar results when compiled for
 same target architecture. Also, both suffer from the same problem, although
@@
 variable is live out of a basic-block B if it is used in the merge set of B. We
-compute the live-out relation on-demand when profitability of hoistable/sinkable
+compute the live-out relation on-demand when profitability of hoist/sink
 candidates is to be evaluated. A simplified version of
@@
 The availability of operand to the load (an address) is checked at the hoisting
-point. If that is not available we try to rematerialize the addresss if
+point. If that is not available we try to rematerialize the address if
 possible.  Along the path, from current position of the load instruction
@@
 (c1, c2) depend on i1 and i2 respectively which are also in their original basic
-blocks, c1 and c2 are not anticipable in B3. So without knowing the sinkability
+blocks, c1 and c2 are not anticipable in B3. So without knowing the ability to sink
 of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
@@
 
-A general global scheduling algorithm also requires checks for undefinedness
+A general global scheduling algorithm also requires checks for undefined values
 when introducing a new computation along a path. Since only very busy
 expressions are moved in the current implementation, there is no need to check
-for undefinedness resulting due to movement of instructions. This simplifies the
+for undefined resulting due to movement of instructions. This simplifies the
 implementation.
@@
 of rematerializing the instruction should it be beneficial to do so. But the
-current live-range splitting pass of \LLVM{} is not making the optimial decision
+current live-range splitting pass of \LLVM{} is not making the optimal decision
 and we have found performance regressions while hoisting aggressively.
@@
 
-\subsubsection{Partitioning the list of hoistable candidates to maximize hoisting}
+\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
 \label{subsec:partition}
@@
 In order to hoist a subset of identical instructions, we partition the list of
-all candidates in a way to maximize the total number of hoistings.  By sorting
+all candidates in a way to maximize the total number of hoist instructions.  By sorting
 the list of all the candidates in the increasing order of their depth first
@@
 After one iteration of algorithm runs through the entire function, it creates
-more oppportunities for \emph{higher ranked} computations
+more opportunities for \emph{higher ranked} computations
 \cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
@@
 intersecting side-effects, or profitability constraints, e.g., increasing
-register pressure; in those cases, this algorithmn would partition
+register pressure; in those cases, this algorithm would partition
 (Section~\ref{subsec:partition}) the list of identical instructions into subsets
@@
 After no more candidates are hoistable, sinking is performed. Sinking is only
-performed once becase currently there are very few sinkable candidates
+performed once because currently there are very few sinkable candidates
 (Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
@@
 iterate on lower ranking expressions first and then proceed to higher ranking
-expressions in the same iteration but \LLVM{'}s \GVN{} infrastucture does not
+expressions in the same iteration but \LLVM{'}s \GVN{} infrastructure does not
 compute equivalence classes in an effective way. We found it simpler to just
@@
 are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
-Barriers for each iteration of the \gcm{}, we have not seen signinficant
+Barriers for each iteration of the \gcm{}, we have not seen significant
 increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
@@
 % llvm trunk vs. llvm trunk with code-motion
-                &\% performance uplift & \% codesize decrease & \% codesize decrease\\
+                &\% performance uplift & \% code size decrease & \% code size decrease\\
 Benchmarks      & at -Ofast (high better)& at -Ofast (low better) & at -Os (low better)\\\hline
@@
 Overall there was only a minor change in the performance numbers with a few improvements 
-(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
+(429.mcf, 458.sjeng etc.) and a few regressions (447.dealII, 453.povray
 etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
@@
 Table~\ref{tab:code-size}.  The code size listed here is the text size (from
-linux size command) of the final executables for each benchmark.  Some bencmarks
+linux size command) of the final executable for each benchmark.  Some benchmarks
 like 473.astar increased in code size by almost 2\%. This is because once the
@@
 by ranking the computations \cite{rosen1988global}. Also, since \GCM{} runs very
-early in the pass pipeline, it will be good to evaluate the codesize/performance
+early in the pass pipeline, it will be good to evaluate the code size/performance
 impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.

### 2017-04-18 paper.tex
typos

@@
 limits the \gcm{} to those candidates which do not increase the register
-pressure (except for loads Section~\ref{subsec:cost-models}, overall spills
+pressure (except for loads Section~\ref{subsec:cost-models},) overall spills
 still decreases.
@@
 to be processed by later compilation passes also reduces. The compilation time listed here
-is the total instruction count as outputted from valgrind (--tool=cachegrind) \cite{valgrind}.
+is the total instruction count as outputted from valgrind -{}-tool=cachegrind \cite{valgrind}.
 

### 2017-04-18 paper.tex
move loads cost to reg-spills

@@
   general, cannot effectively reason about instructions within conditional
-  branches in the context of loops, code-hoisting is needed to move instructions
+  branches in the context of loops, code-motion is needed to move instructions
   out of conditional expressions and expose them to LICM.
@@
 do not intersect with any side-effects between the instructions to be
-hoisted/sunk and their hoisting/sinking point. It is also necessary to check if
+moved and their hoisting/sinking point. It is also necessary to check if
 there are indirect branch targets, e.g., landing pad, case statements, goto
-labels etc., along the path because it becomes difficult to prove safety checks
+labels, etc., along the path because it becomes difficult to prove safety checks
 in those cases. In our current implementation we discard candidates on those
@@
 \label{subsec:cost-models}
-After the legality checks have passed, we check if a \gcm{} is profitable.
-That takes into account the impact \gcm{} would have on various parameters
-that might affect runtime performance, e.g., impact on live-range, gain in the
-code size. Since this is mostly a code-size optimization pass, the goal is to
-not regress in performance across popular benchmarks and at the same time to reduce
-code size as much as possible. Following are the cost models which are
-implemented:
+After the legality checks have passed, we check if a \gcm{} is profitable.  That
+takes into account the impact \gcm{} would have on various parameters that might
+affect runtime performance, e.g., impact on live-range, gain in the code
+size. The current implementation makes effort to not regress in performance
+across popular benchmarks and at the same time to reduce code size as much as
+possible. Following are the cost models which are implemented:
 
@@
 improved to catch these regressions.
+
+There is a special case for hoisting load instructions when the hoist-point is
+the predecessor basic-block for all the congruent loads. Even if the register
+pressure would increase, we prefer to hoist loads in this case. The reason
+being, that would make the loaded value available by the time it is used. Also,
+because stores and calls are hoisted the least, see
+Table~\ref{tab:code-motion-metric}, the performance does not change much whether
+they are hoisted or not.
 
@@
 In our current implementation we keep as many candidates in one set as possible
-(greedy approach.) We split the list at a point where the legality checks fail
-to hoist subset of candidates which are legal to hoist and then start finding
-new hoisting point for the remaining ones.
-
-There is a special case for hoisting load instructions when the hoist-point is
-the predecessor basic-block for all the loads. Even if the register pressure
-would increase, we prefer to hoist loads in this case. The reason being, that
-would make the loaded value available by the time it is used. Also, because
-stores and calls are hoisted the least, see Table~\ref{tab:code-motion-metric},
-the performance does not change much whether they are hoisted or not.
+(greedy approach). To start with, first two candidates are tried for
+hoistability (see Section~\ref{subsec:optimistic}), if that passes then the next
+instruction is tried (if available) and so on until a non-hoistable instruction
+is found. The partition starts from that point.  We split the list at a point
+where the legality checks fail to hoist subset of candidates which are legal to
+hoist and then start finding new hoisting point for the remaining ones.
 
@@
 Overall there was only a minor change in the performance numbers with a few improvements 
-(429.mcf, 458.sjeng etc.) and a few regressions (447.dealII, 453.povray
+(429.mcf, 458.sjeng, etc.) and a few regressions (447.dealII, 453.povray,
 etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
@@
 code-size of a function decreases (due to \GCM{}), it becomes cheaper to
-inline. As we can see in Table-\ref{tab:stats}, four more functions got inlined
+inline. As we can see in Table~\ref{tab:stats}, four more functions got inlined
 in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
@@
   \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at
-    -Ofast. Spills reduced by 2\% overall.}
+    -Ofast. Spills reduced by 2\%.}
   \label{tab:spills}
@@
   \caption{Change in total number of instructions executed during the
-    compilation with and without \gcm{} (\GCM{}) on \SPEC{} at -Ofast}
+    compilation with and without \gcm{} (\GCM{}) on \SPEC{} at -Ofast.}
   \label{tab:compile-time}
@@
 
-The compilation time actually reduced for most of the benchmarks as a result of code motion.
-This is because, as \gcm{} removes (congruent) instructions, the number of instructions
-to be processed by later compilation passes also reduces. The compilation time listed here
-is the total instruction count as outputted from valgrind -{}-tool=cachegrind \cite{valgrind}.
+The compilation time actually reduced for most of the benchmarks as a result of
+code motion.  This is because, as \gcm{} removes (congruent) instructions, the
+number of instructions to be processed by later compilation passes also
+reduces. The compilation time listed here is the total instruction count
+executed during the compilation of each benchmark as output from valgrind
+-{}-tool=cachegrind \cite{valgrind}.
 
@@
     \SPEC{} at -Ofast. The \% Column show percentage increase in metric
-    w.r.t. Base. \GCM{} improves LICM and inlining}
+    w.r.t. Base. \GCM{} improves LICM and inlining.}
   \label{tab:stats}
@@
   \end{center}
-  \caption{\GCM{} stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are hoisted the most followed by scalars, stores and calls}
+  \caption{\GCM{} stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are
+    hoisted the most followed by scalars, stores and calls.}
   \label{tab:code-motion-metric}
@@
 revisited. The first candidate would be the inliner. We have seen difference
-in inlining decisions (Table~\ref{tab:code-size}), when \gcm{}
+in inlining decisions in Table~\ref{tab:code-size}, when \gcm{}
 was enabled.  Since inliner has several cost-models tuned for the previous

### 2017-04-18 paper.tex
add missing commas

@@
 there are indirect branch targets, e.g., landing pad, case statements, goto
-labels etc., along the path because it becomes difficult to prove safety checks
+labels, etc., along the path because it becomes difficult to prove safety checks
 in those cases. In our current implementation we discard candidates on those

### 2017-04-18 paper.tex
code-size in bytes

--- 2017-04-18 paper.tex
@@
                 &\% performance uplift & \% code size decrease & \% code size decrease\\
-Benchmarks      & at -Ofast (high better)& at -Ofast (low better) & at -Os (low better)\\\hline
-400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
-401.bzip2	& 1.00  & 1.00 & 1.00\\\hline
-403.gcc	        & 1.00  & 1.01 & 1.00\\\hline
-429.mcf	        & 1.03  & 1.00 & 1.00\\\hline
-433.milc	& 1.00  & 1.00 & 1.00\\\hline
-444.namd	& 1.00  & 1.01 & 1.00\\\hline
-445.gobmk	& 0.99  & 1.01 & 1.00\\\hline
-447.dealII	& 0.95  & 1.01 & 0.98\\\hline
-450.soplex	& 0.99  & 0.99 & 1.00\\\hline
-453.povray	& 0.97  & 1.00 & 1.00\\\hline
-456.hmmer	& 0.99  & 1.00 & 1.00\\\hline
-458.sjeng	& 1.02  & 1.00 & 1.00\\\hline
-462.libquantum	& 1.00  & 1.00 & 1.00\\\hline
-464.h264ref	& 1.00  & 1.00 & 1.00\\\hline
-470.lbm	        & 1.00  & 1.00 & 1.00\\\hline
-471.omnetpp	& 1.00  & 0.99 & 1.00\\\hline
-473.astar	& 1.00  & 1.02 & 1.00\\\hline
-482.sphinx3	& 0.99  & 1.00 & 1.00\\\hline
-483.xalancbmk	& 1.01  & 0.99 & 1.00\\\hline
-Geomean         & 0.997 & 1.00 & 1.00\\\hline
+Benchmarks      & at -Ofast (high better)& at -Ofast (high better) & at -Os (high better)\\\hline
+400.perlbench	& 1.00  & -9120  & 544.00  \\\hline
+401.bzip2	& 1.00  & -1600  & 128.00  \\\hline
+403.gcc	        & 1.00  & -58176 & 3384.00 \\\hline
+429.mcf	        & 1.03  & 160    & 16.00   \\\hline
+433.milc	& 1.00  & -656   & 224.00  \\\hline
+444.namd	& 1.00  & -3864  & -1168.00\\\hline
+445.gobmk	& 0.99  & -20480 & -264.00 \\\hline
+447.dealII	& 0.95  & -27836 & 50934.00\\\hline
+450.soplex	& 0.99  & 2220   & 348.00  \\\hline
+453.povray	& 0.97  & 164    & -472.00 \\\hline
+456.hmmer	& 0.99  & -3464  & -352.00 \\\hline
+458.sjeng	& 1.02  & 440    & 1456.00 \\\hline
+462.libquantum	& 1.00  & -96    & 0.00    \\\hline
+464.h264ref	& 1.00  & -3424  & 864.00  \\\hline
+470.lbm	        & 1.00  & 0      & 0.00    \\\hline
+471.omnetpp	& 1.00  & 2688   & 200.00  \\\hline
+473.astar	& 1.00  & -928   & 152.00  \\\hline
+482.sphinx3	& 0.99  & 3008   & -176.00 \\\hline
+483.xalancbmk	& 1.01  & 46492  & 4012.00 \\\hline
+Geomean         & 0.997 &  & \\\hline
     \end{tabular}

### 2017-04-18 paper.tex
split some long lines

@@
       if both are only used in the same PHI-node (PN) of the successor
-        if both have dependencies that are available or can be made available at the sink point
+        if both have dependencies that are available at the sink point
+           // available or can be made available by rematerialization
            mark VN as sinkable
   For each pair (I1, I2) of sinkable instructions
-    Move I1 to the sink point which is just after all the PHI-nodes in the post-dominator
+    Move I1 to the sink point
+    // the sink point is just after all the PHI-nodes in the post-dominator
     update all the uses of PHI node (PN) with I1
@@
 were conducted on an \xlinux{} machine and at -Ofast optimization level. Each
-sub benchmark was run three times and the best result was taken. We collected
+benchmark was run three times and the best result was taken. We collected
 execution time and code-size which is listed in Table~\ref{tab:code-size}, other
-compiler stats are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
+compiler statistics are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
 Table~\ref{tab:code-motion-metric}.  We also measured compile time

### 2017-04-18 paper.tex
make last table fit without overflow

@@
     Compute GVN of each expression in the function
-  while (doCodeHoisting() > 0) // If any of candidates were hoisted then repeat hoisting
+  while (doCodeHoisting() > 0) // Repeat hoisting if any candidates were hoisted.
   doCodeSinking()
@@
   &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
- Benchmarks       & Base        & \GCM{}	& \%  &Base     & \GCM{}   &\%    & Base	& \GCM{}	  &\%   & Base	& \GCM{}	& \% \\\hline
- 400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
- 401.bzip2	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
- 403.gcc	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
- 429.mcf	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
- 433.milc	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
- 444.namd	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
- 445.gobmk	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
- 447.dealII	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
- 450.soplex	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
- 453.povray	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
- 456.hmmer	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
- 458.sjeng	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
- 462.libquantum	  & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
- 464.h264ref	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
- 470.lbm	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
- 471.omnetpp	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
- 473.astar	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
- 482.sphinx3	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
- 483.xalancbmk	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
+ {\Small Bmarks}          & Base        & \GCM{}	& \%  &Base     & \GCM{}   &\%    & Base	& \GCM{}	  &\%   & Base	& \GCM{}	& \% \\\hline
+ {\Small 400.perlb}	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
+ {\Small 401.bzip2}	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
+ {\Small 403.gcc}	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
+ {\Small 429.mcf}	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
+ {\Small 433.milc}	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
+ {\Small 444.namd}	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
+ {\Small 445.gobmk}	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
+ {\Small 447.dealII}	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
+ {\Small 450.soplex}	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
+ {\Small 453.povray}	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
+ {\Small 456.hmmer}	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
+ {\Small 458.sjeng}	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
+ {\Small 462.libqua}     & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
+ {\Small 464.h264}	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
+ {\Small 470.lbm}	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
+ {\Small 471.omnet}	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
+ {\Small 473.astar}	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
+ {\Small 482.sphinx}	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
+ {\Small 483.xalanc}	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
     \end{tabular}

### 2017-04-18 paper.tex
code size comments

@@
 Benchmarks      & at -Ofast (high better)& at -Ofast (high better) & at -Os (high better)\\\hline
-400.perlbench	& 1.00  & -9120  & 544.00  \\\hline
-401.bzip2	& 1.00  & -1600  & 128.00  \\\hline
-403.gcc	        & 1.00  & -58176 & 3384.00 \\\hline
-429.mcf	        & 1.03  & 160    & 16.00   \\\hline
-433.milc	& 1.00  & -656   & 224.00  \\\hline
-444.namd	& 1.00  & -3864  & -1168.00\\\hline
-445.gobmk	& 0.99  & -20480 & -264.00 \\\hline
-447.dealII	& 0.95  & -27836 & 50934.00\\\hline
-450.soplex	& 0.99  & 2220   & 348.00  \\\hline
-453.povray	& 0.97  & 164    & -472.00 \\\hline
-456.hmmer	& 0.99  & -3464  & -352.00 \\\hline
-458.sjeng	& 1.02  & 440    & 1456.00 \\\hline
-462.libquantum	& 1.00  & -96    & 0.00    \\\hline
-464.h264ref	& 1.00  & -3424  & 864.00  \\\hline
-470.lbm	        & 1.00  & 0      & 0.00    \\\hline
-471.omnetpp	& 1.00  & 2688   & 200.00  \\\hline
-473.astar	& 1.00  & -928   & 152.00  \\\hline
-482.sphinx3	& 0.99  & 3008   & -176.00 \\\hline
-483.xalancbmk	& 1.01  & 46492  & 4012.00 \\\hline
+400.perlbench	& 1.00  & -9120  & 544  \\\hline
+401.bzip2	& 1.00  & -1600  & 128  \\\hline
+403.gcc	        & 1.00  & -58176 & 3384 \\\hline
+429.mcf	        & 1.03  & 160    & 16   \\\hline
+433.milc	& 1.00  & -656   & 224  \\\hline
+444.namd	& 1.00  & -3864  & -1168\\\hline
+445.gobmk	& 0.99  & -20480 & -264 \\\hline
+447.dealII	& 0.95  & -27836 & 50934\\\hline
+450.soplex	& 0.99  & 2220   & 348  \\\hline
+453.povray	& 0.97  & 164    & -472 \\\hline
+456.hmmer	& 0.99  & -3464  & -352 \\\hline
+458.sjeng	& 1.02  & 440    & 1456 \\\hline
+462.libquantum	& 1.00  & -96    & 0    \\\hline
+464.h264ref	& 1.00  & -3424  & 864  \\\hline
+470.lbm	        & 1.00  & 0      & 0    \\\hline
+471.omnetpp	& 1.00  & 2688   & 200  \\\hline
+473.astar	& 1.00  & -928   & 152  \\\hline
+482.sphinx3	& 0.99  & 3008   & -176 \\\hline
+483.xalancbmk	& 1.01  & 46492  & 4012 \\\hline
 Geomean         & 0.997 &  & \\\hline
@@
   \end{center}
-  \caption{Execution time and code size with and without \gcm{}
-    (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng; code size
-    improvement in dealII at -Os.}
+  \caption{Execution time (ratio) and code size change (bytes) with and without
+    \gcm{} (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng;
+    code size improvement in dealII at -Os.}
   \label{tab:code-size}
@@
 
-Code size also remained almost the same as a result of \gcm{} as shown in
-Table~\ref{tab:code-size}.  The code size listed here is the text size (from
-linux size command) of the final executable for each benchmark.  Some benchmarks
-like 473.astar increased in code size by almost 2\%. This is because once the
-code-size of a function decreases (due to \GCM{}), it becomes cheaper to
-inline. As we can see in Table~\ref{tab:stats}, four more functions got inlined
-in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
-all other benchmarks got more functions inlined. Also, since the pass runs
-early, it affects many optimizations which rely on the number of instructions,
-length of the use-def chain, and other metrics.
+We see some nice gains in code size at -Os with \gcm{} (403.gcc, 447.dealII, and
+483.xalancbmk) as shown in Table~\ref{tab:code-size}.  The code size listed here
+is the text size delta (from linux size command) of the final executable for
+each benchmark.  Some benchmarks like 473.astar increased in code size by almost
+2\%. This is because once the code-size of a function decreases (due to \GCM{}),
+it becomes cheaper to inline. As we can see in Table~\ref{tab:stats}, four more
+functions got inlined in 473.astar. Except for 447.dealII, 450.soplex,
+482.sphinx3 and 483.xalancbmk all other benchmarks got more functions
+inlined. Also, since the pass runs early, it affects many optimizations which
+rely on the number of instructions, length of the use-def chain, and other
+metrics.
+
+The code-size at -Ofast is very widespread with huge gains in 483.xalancbmk to
+serious regressions in 403.gcc. This is mostly because the compiler's focus is
+to improve performance at the cost of code-size at -Ofast, so passes like
+inlining, loop-unrolling, code-versioning are very aggressive at -Ofast.
 

### 2017-04-18 paper.tex
mention unit for code size decrease

@@
 % llvm trunk vs. llvm trunk with code-motion
-                &\% performance uplift & \% code size decrease & \% code size decrease\\
+                &\% performance uplift & size decrease in Bytes & size decrease in Bytes\\
 Benchmarks      & at -Ofast (high better)& at -Ofast (high better) & at -Os (high better)\\\hline

### 2017-04-18 paper.tex
grammar

@@
 partitioning the potential candidates in a way to enable partial hoisting in
-case common hoisting points for all the candidates cannot be found. We also
+case a common hoisting point for all the candidates cannot be found. We also
 formalize how register pressure will reduce as a result of \gcm{}, and how
@@
 is the text size delta (from linux size command) of the final executable for
-each benchmark.  Some benchmarks like 473.astar increased in code size by almost
-2\%. This is because once the code-size of a function decreases (due to \GCM{}),
-it becomes cheaper to inline. As we can see in Table~\ref{tab:stats}, four more
-functions got inlined in 473.astar. Except for 447.dealII, 450.soplex,
-482.sphinx3 and 483.xalancbmk all other benchmarks got more functions
-inlined. Also, since the pass runs early, it affects many optimizations which
-rely on the number of instructions, length of the use-def chain, and other
-metrics.
+each benchmark.  On the other hand, some benchmarks like 473.astar increased in
+code size by almost 2\%. This is because once the code-size of a function
+decreases (due to \GCM{}), it becomes cheaper to inline. As we can see in
+Table~\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
+447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
+more functions inlined. Also, since the pass runs early, it affects many
+optimizations which rely on the number of instructions, length of the use-def
+chain, and other metrics.
 

### 2017-04-18 paper.tex
caption

@@
   \end{center}
-  \caption{Execution time (ratio) and code size change (bytes) with and without
+  \caption{Execution time (ratio) and code size change (Bytes) with and without
     \gcm{} (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng;
@@
   \caption{Change in total number of instructions executed during the
-    compilation with and without \gcm{} (\GCM{}) on \SPEC{} at -Ofast.}
+    compilation with and without \gcm{} (\GCM{}) on \SPEC{} at -Ofast. The
+    instructions were counted using valgrind.}
   \label{tab:compile-time}

### 2017-04-18 paper.tex
unify style of algorithm

@@
   for e in (all incoming edges to lnode)
-    if (e is in JE && e in V)
-      if (MergeSet(Source Node of e) is not a subset of MergeSet(lnode))
+    if e is in JE and e in V
+      if MergeSet(Source Node of e) is not a subset of MergeSet(lnode)
         return true
@@
     for e in (all incoming edges to n)
-      if (e is in JE && e not in V)
+      if e is in JE and e not in V
         V(e) = true

### 2017-04-18 paper.tex
move punctuation within parentheses

@@
 scheduling to improve performance. Most of the recent work on global scheduling
-are done using integer linear programming (ILP), which results in prohibitively
+are done using integer linear programming (ILP,) which results in prohibitively
 high compile time. To the best of our knowledge we have not found any reference
@@
 computed which are hoisted in a basic-block where the expression is
-anticipable (all the operands are available). This algorithm would hoist as far
+anticipable (all the operands are available.) This algorithm would hoist as far
 as possible without regarding the impact on register pressure and as such a cost
@@
 only flow (from def to use) either through dominator edges or the join edges
-(where we insert a PHI). The DJ-graph contains both these edges which allows for
+(where we insert a PHI.) The DJ-graph contains both these edges which allows for
 computation of merge-sets for each basic-block, i.e., a set of all basic-blocks
@@
 \begin{itemize}
-\item find candidates (congruent instructions) suitable for \gcm{} (Section~\ref{subsec:finding-candidates}),
+\item find candidates (congruent instructions) suitable for \gcm{}
+  (Section~\ref{subsec:finding-candidates},)
 \item compute a point in the program where it is both legal
@@
 \item move the code to hoist point or sink point
-  (Section~\ref{subsec:optimistic}), and
-\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
+  (Section~\ref{subsec:optimistic},) and
+\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}.)
 \end{itemize}
@@
 instruction) or, precomputed (computing \GVN{} of all the instructions
-beforehand). Which process to choose is determined by the scope of code-hoisting
+beforehand.) Which process to choose is determined by the scope of code-hoisting
 we want to perform. In a pessimistic approach, we want to hoist a limited set of
@@
 aliasing memory references. As long as all the operands are available (or can be
-made available by rematerialization), the scalar computations can be hoisted.
+made available by rematerialization,) the scalar computations can be hoisted.
 
@@
 Hoisting upwards will decrease the live-range of its use, if it is a last use (a
-kill), but increase the live-range of its definition. Conversely, sinking will
+kill,) but increase the live-range of its definition. Conversely, sinking will
 decrease the live-range of the defined register but increase the live-range for
@@
 the list of all the candidates in the increasing order of their depth first
-search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
+search discovery time stamp \cite{clrs} (DFSIn numbers,) we make sure that
 candidates closer in the list have their common dominator nearby. In
@@
 In our current implementation we keep as many candidates in one set as possible
-(greedy approach). To start with, first two candidates are tried for
-hoistability (see Section~\ref{subsec:optimistic}), if that passes then the next
+(greedy approach.) To start with, first two candidates are tried for
+hoistability (see Section~\ref{subsec:optimistic},) if that passes then the next
 instruction is tried (if available) and so on until a non-hoistable instruction
@@
 The amount of hoisting depends on whether we collect \GVN{} of instructions
-before finding candidates (optimistic) or, on-demand (pessimistic). It also
+before finding candidates (optimistic) or, on-demand (pessimistic.) It also
 depends on the generality of the \GVN{} algorithm as mentioned earlier in
@@
 approach is gain in code size (the final executable may be of larger size
-because of more inlining). The algorithm prefers hoisting to sinking. If the
+because of more inlining.) The algorithm prefers hoisting to sinking. If the
 dependency of a hoistable candidate is in the same basic-block as the candidate,
@@
 Code hoisting opens new opportunities for other hoistable candidates which were
-of higher rank (depended on candidates which got hoisted). Ideally we could
+of higher rank (depended on candidates which got hoisted.) Ideally we could
 iterate on lower ranking expressions first and then proceed to higher ranking
@@
 Barriers for each iteration of the \gcm{}, we have not seen significant
-increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
+increase in the compilation times (see Section~\ref{tab:compile-time}.) We have also
 provided appropriate compiler
@@
 (429.mcf, 458.sjeng, etc.) and a few regressions (447.dealII, 453.povray,
-etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
+etc.) The \gcm{} pass was run twice in the pass pipeline (at -Ofast,)
 first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
@@
 code size by almost 2\%. This is because once the code-size of a function
-decreases (due to \GCM{}), it becomes cheaper to inline. As we can see in
+decreases (due to \GCM{},) it becomes cheaper to inline. As we can see in
 Table~\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
@@
   \end{center}
-  \caption{\GCM{} stats on SPEC2006 at -Ofast, H(Hoisted), R(Removed); Loads are
+  \caption{\GCM{} stats on SPEC2006 at -Ofast, H(Hoisted,) R(Removed); Loads are
     hoisted the most followed by scalars, stores and calls.}

### 2017-04-18 paper.tex
last fixes

@@
 
-The process of computing \GVN{} can be on-demand (as we come across an
-instruction) or, precomputed (computing \GVN{} of all the instructions
-beforehand.) Which process to choose is determined by the scope of code-hoisting
+The process of computing \GVN{} can be on-demand, as we come across an
+instruction, or precomputed, computing \GVN{} of all the instructions
+beforehand. Which process to choose is determined by the scope of code-hoisting
 we want to perform. In a pessimistic approach, we want to hoist a limited set of
@@
 Scalars are the easiest to hoist because we do not have to analyze them for
-aliasing memory references. As long as all the operands are available (or can be
+aliasing memory references. As long as all the operands are available, or can be
 made available by rematerialization,) the scalar computations can be hoisted.

### 2017-04-18 paper.tex
algos

@@
 \begin{verbatim}
-computing barriers:
+// Compute barriers for both hoistable and sinkable
+// candidates.
+void compute_barriers()
   for each basic-block B in a function:
@@
 \begin{verbatim}
-computing downward-safety of hoistable instructions at hoist-point
+// Compute downward safety of hoistable instructions at hoist-point
+void compute_downward_safety()
   Worklist = list of all basic-blocks of hoistable instructions
@@
 int doCodeHoisting()
+  // Collect hoistable candidates
   For each value number VN with 2 or more instructions (I1, I2, ..., Ik)
@@
       proceed to check if I2 and I3 (if available) can be hoisted
+
+  // Do the code hoisting and update statistics
   For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
@@
     update statistics
+
   return number of hoisted instructions
@@
 void doCodeSinking()
+  // Find sinkable instructions
   For each value number VN with 2 or more instructions
@@
            mark VN as sinkable
+
+  // Sink the instructions and update statistics
   For each pair (I1, I2) of sinkable instructions
@@
     if I1 is a memory reference:
-      update MemorySSA of I1 and others to point to the MemorySSA reference of I1
+      update MemorySSA of I1
+      update MemorySSA of removed instructions to point to the MemorySSA reference of I1
       remove MemorySSA reference of all others which were deleted
     update statistics
+
+  return number of sinked instructions
 \end{verbatim}

### 2017-04-18 paper.tex
formatting

@@
 \begin{verbatim}
-int doCodeHoisting()
-  // Collect hoistable candidates
+int doCodeHoisting() // Find hoistable candidates and do hoisting
   For each value number VN with 2 or more instructions (I1, I2, ..., Ik)
@@
           and if they are profitable to be hoisted at hoist-point
-      proceed to check ability to hoist of subsequent instruction with same VN
+      proceed to check the ability to hoist subsequent instruction with same VN
     else
       proceed to check if I2 and I3 (if available) can be hoisted
-
-  // Do the code hoisting and update statistics
   For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
@@
     update statistics
-
   return number of hoisted instructions
@@
 
-\newpage
+%\newpage
 After no more candidates are hoistable, sinking is performed. Sinking is only

### 2017-04-18 paper.tex
page

@@
 \begin{abstract}
-We present a \gcm{} (\GCM{}) compiler optimization which schedules
-congruent instructions across the program.  Not only this saves code size, it
-exposes redundancies in some cases, it exposes more instruction level
-parallelism in the basic-block to which instructions are moved, and it enables
-other passes like loop invariant motion to remove more redundancies. The cost
-model to drive the code motion is based on live range analysis on \SSA{}
-representation such that the (virtual) register pressure does not increase, which
-translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.
+We present a \gcm{} (\GCM{}) compiler optimization which schedules congruent
+instructions across the program.  Not only \GCM{} saves code size, it exposes
+redundancies in some cases, it exposes more instruction level parallelism in the
+basic-block to which instructions are moved, and it enables other passes like
+loop invariant motion to remove more redundancies. The cost model to drive the
+code motion is based on liveness analysis on \SSA{} representation such that the
+(virtual) register pressure does not increase resulting in 2\% fewer spills on
+the \SPEC{} benchmark suite.
 
@@
 infrastructure available in \LLVM{}. The experimental results show an average
-saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining
-and exposes more loop invariant code motion opportunities in majority of the
-benchmarks. We have also seen execution time improvements in a few of \SPEC{}
-benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
-\SPEC{} benchmarks suite when compiled for \xlinux{}. This is an optimistic
-algorithm in the sense that it considers all congruent instructions in a
-function as potential candidates. We make an extra effort to hoist candidates by
-partitioning the potential candidates in a way to enable partial hoisting in
-case a common hoisting point for all the candidates cannot be found. We also
-formalize how register pressure will reduce as a result of \gcm{}, and how
-\gcm{} increases instruction level parallelism thereby enabling more \ooo{}
-execution on modern architectures.
+saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
+inlining and exposes more loop invariant code motion opportunities in majority
+of the benchmarks. We have also seen execution time improvements in a few of
+\SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\%
+on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
+optimistic algorithm in the sense that it considers all congruent instructions
+in a function as potential candidates. We make an extra effort to hoist
+candidates by partitioning the potential candidates in a way to enable partial
+hoisting in case a common hoisting point for all the candidates cannot be
+found. We also formalize why register pressure reduces as a result of \gcm{},
+and how \gcm{} increases instruction level parallelism thereby enabling more
+\ooo{} execution on modern architectures.
 \end{abstract}
@@
 phase that reduces the number of run-time computations.  Classical scalar
-optimizations like CSE \cite{dragonbook} work very well on single basic-blocks but
-when it comes to detect redundancies across basic-blocks these techniques fall
-short: more complex passes like GCSE and \PRE{} have been designed to handle these
-cases based on data-flow analysis \cite{morel1979global}.  At first these
-techniques were described in the classical data-flow analysis framework, and
-later the use of the \SSA{} representation lowered the cost in terms of compilation
-time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
-techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
-industrial compiler.
-
-This paper describes \gcm{} (\GCM{}) of congruent instructions \cite{briggs1997}, a
-technique that uses the information computed for \PRE{} to detect identical
-computations but has a transformation phase whose goal differs from \PRE{}: it
-removes identical computations from different branches of execution.  These
-identical computations in different branches of execution are not redundant
-computations at run-time and the number of run-time computations is not
-reduced. It is not a redundancy elimination pass, and thus it has different cost
-function and heuristics than \PRE{} or CSE. It is more similar to global code
-scheduling \cite{dragonbook,click1995global} in the sense that it will only move
-computations. Code hoisting, for example, can reduce the critical path length of
-execution in \ooo{} machines. As more instructions are available at the
-hoisting point, the hardware has more independent instructions to reorder. The
-following example illustrates how hoisting can improve performance by exposing
-more ILP.
+optimizations like Common Subexpression Elimination (CSE) \cite{dragonbook} work
+very well on single basic-blocks (local level) but when it comes to detect
+redundancies across basic-blocks (global level) these techniques fall short:
+more complex passes like Global Common Subexpression Elimination (GCSE) and
+Partial Redundancy Elimination (\PRE{}) have been designed to handle these cases
+based on data-flow analysis \cite{morel1979global}.  At first these techniques
+were described in the classical data-flow analysis framework, but later the use
+of Static Single Assignment (\SSA{}) representation lowered their cost in terms
+of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
+and brought these techniques in the main stream: nowadays \SSA{} based \PRE{} is
+available in industrial compilers like \GCC{} and \LLVM{}.
+
+This paper describes \gcm{} (\GCM{}) of congruent instructions
+\cite{briggs1997}, a technique that uses the information computed for \PRE{} to
+detect identical computations but has a transformation phase whose goal differs
+from \PRE{}: it removes identical computations from different branches of
+execution.  These identical computations in different branches of execution are
+not redundant computations at run-time and the number of run-time computations
+is not reduced. It is not a redundancy elimination pass, and thus it has
+different cost function and heuristics than \PRE{} or CSE. It is similar to
+global code scheduling \cite{dragonbook,click1995global} in the sense that it
+will only move computations. Code hoisting, for example, can reduce the critical
+path length of execution in \ooo{} machines. As more instructions are available
+at the hoisting point, the hardware has more independent instructions to
+reorder. The following example illustrates how hoisting can improve performance
+by exposing more ILP.
 
@@
   return false
-
+\end{verbatim}
+
+Each basic block which has a join-edge (J-Edge) to its successor will have a
+merge-set that is a subset of the merge-set of its successor \cite{das2005}. The
+function \emph{still\_inconsistent} returns true if this invariant is not
+satisfied which allows the function \emph{constructMergeSet} to repeat.
+
+\begin{verbatim}
 // Compute a merge-set top-down in breadth first order.
@@
   return repeat
-
+\end{verbatim}
+\newpage
+\begin{verbatim}
 // Construct merge-set of each node in Control-Flow-Graph G of a function.
@@
   return false
-
+\end{verbatim}
+
+If a variable is not live-out of a basic-block, it still may be used later in
+the basic-block. To establish whether a use of a variable is a kill we iterate
+on all the subsequent instructions in a basic-block checking for uses, the
+\emph{isKill} function uses \emph{isLiveOutUsingMergeSet} to find out if an
+operand is killed in an instruction.
+
+\begin{verbatim}
 // Return true if op is the last use of the operand in I.
@@
     return false
-
-  for each I1 after I in basic_block(I)
+  for each I1 after I in basic block of I
     if (I1 uses op)
-      return false
+      return false // Not a kill
   return true
@@
 terminate. These two cases were missing from the algorithm and we came across
-them during implementation. If a variable is not live-out of a basic-block, it
-still may be used later in the basic-block: to establish whether a use of a
-variable is a kill we iterate on all the subsequent instructions in a
-basic-block checking for uses, the \emph{isKill} function uses
-\emph{isLiveOutUsingMergeSet} to find out if an operand is killed in an
-instruction.
+them during implementation.
 
@@
 In order to be able to move an instruction that has side effects, we must
-guarantee that the side-effects of that instructions before and after its motion
-will produce the same side-effects.  For hoisting in particular, we need to
-prove that on all execution paths from the hoisting point to the end of the
-program the side effects appear in the same order.
-
+guarantee that the side-effects of that instruction before and after its
+code-motion will produce the same side-effects.  For hoisting in particular, we
+need to prove that on all execution paths from the hoisting point to the end of
+the program the side effects appear in the same order. In general, we cannot
+introduce a new computation along any path of execution without checking for
+undefined behaviors. The algorithm \emph{compute\_downward\_safety} computes if
+hoisting a set of instruction to a common hoisting point is downward-safe
+\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
+downward-safety ensures that no undefined behaviors are introduced as a result
+of code-motion.
+\newline
 \begin{verbatim}
-// Compute downward safety of hoistable instructions at hoist-point
-void compute_downward_safety()
+// Return true if a hoistable instruction is downward safe at hoist-point
+bool compute_downward_safety()
   Worklist = list of all basic-blocks of hoistable instructions
   for each basic-block B in the dominator tree starting at hoist-point:
-    if Worklist is empty, return false // Path exists! not downward safe
+    if Worklist is empty
+       return false // Path exists! not downward safe
     if B is in Worklist:
@@
       remove subtree with root at B // Available at B => downward safe from B
-    if B is a leaf basic-block, return false  // Path exists! not downward safe
-    if B dominates hoist-point, return false  // Back edge
+    if B is a leaf basic-block
+      return false  // Path exists! not downward safe
+    if B dominates hoist-point
+      return false  // Back edge
   return true
@@
   Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
-  Compute DJ-graph of function and compute mergeset based on that
+  Compute DJ-graph of function
+  compute mergeset based on that (using constructMergeSet function)
+  compute barriers (using compute_barriers function)
   do
@@
 We collect the \GVN{} of all the instructions in the function and iterate on the
-list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
-sinking. So first we find the common dominator dominating all such identical
-computations and perform legality checks, as described in
+list of instructions having identical \GVN{}s. The algorithm \emph{doGCM}
+prefers hoisting (\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).
+
+For hoisting, we find a common dominator dominating all a set of congruent
+instructions and perform legality checks, as described in
 Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
@@
 (Section~\ref{subsec:partition}) the list of identical instructions into subsets
-which can be partially hoisted to their respective common dominators. However,
-it should be noted that we only hoist very busy expressions to avoid checking
-for undefined behaviors resulting because of introducing extra computation in an
-execution path.
-
+which can be partially hoisted to their respective common dominators.
+\newpage
 \begin{verbatim}

### 2017-04-18 paper.tex
shrink table size

@@
   \begin{center}
+    \scalebox{0.7}{
     \begin{tabular}{|l|c|c|c|}
@@
 % llvm trunk vs. llvm trunk with code-motion
-                &\% performance uplift & size decrease in Bytes & size decrease in Bytes\\
+                &\% performance uplift & size reduction in Bytes & size reduction in Bytes\\
 Benchmarks      & at -Ofast (high better)& at -Ofast (high better) & at -Os (high better)\\\hline
@@
 Geomean         & 0.997 &  & \\\hline
-    \end{tabular}
+    \end{tabular}}
   \end{center}
@@
   \begin{center}
+    \scalebox{0.7}{
     \begin{tabular}{|l|r|r|r|}
@@
 Grand Total	  & 43349	& 42497	& 0.98 \\\hline
-    \end{tabular}
+    \end{tabular}}
   \end{center}
@@
   483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
-    \end{tabular}
+    \end{tabular}}
   \end{center}
@@
   \begin{center}
+    \scalebox{0.8}{
     \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
@@
  {\Small 483.xalanc}	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
-    \end{tabular}
+    \end{tabular}}
   \end{center}
@@
   \begin{center}
+    \scalebox{0.8}{
     \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|}
@@
  Grand total      &34957	&38631	  &13091  &13418   &21694	&25041	        &90	&82     &10             \\\hline
-    \end{tabular}
+    \end{tabular}}
   \end{center}

### 2017-04-18 paper.tex
revert patch not needed after scaling the table

@@
   &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
- {\Small Bmarks}          & Base        & \GCM{}	& \%  &Base     & \GCM{}   &\%    & Base	& \GCM{}	  &\%   & Base	& \GCM{}	& \% \\\hline
- {\Small 400.perlb}	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
- {\Small 401.bzip2}	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
- {\Small 403.gcc}	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
- {\Small 429.mcf}	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
- {\Small 433.milc}	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
- {\Small 444.namd}	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
- {\Small 445.gobmk}	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
- {\Small 447.dealII}	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
- {\Small 450.soplex}	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
- {\Small 453.povray}	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
- {\Small 456.hmmer}	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
- {\Small 458.sjeng}	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
- {\Small 462.libqua}     & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
- {\Small 464.h264}	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
- {\Small 470.lbm}	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
- {\Small 471.omnet}	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
- {\Small 473.astar}	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
- {\Small 482.sphinx}	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
- {\Small 483.xalanc}	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
+ Benchmarks          & Base        & \GCM{}	& \%  &Base     & \GCM{}   &\%    & Base	& \GCM{}	  &\%   & Base	& \GCM{}	& \% \\\hline
+ 400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
+ 401.bzip2	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
+ 403.gcc	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
+ 429.mcf	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
+ 433.milc	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
+ 444.namd	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
+ 445.gobmk	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
+ 447.dealII	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
+ 450.soplex	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
+ 453.povray	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
+ 456.hmmer	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
+ 458.sjeng	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
+ 462.libquantum   & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
+ 464.h264ref	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
+ 470.lbm	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
+ 471.omnetpp	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
+ 473.astar	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
+ 482.sphinx3	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
+ 483.xalancbmk	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
     \end{tabular}}

### 2017-06-10 paper.tex
corr

@@
 expressions would be sunk first. And it would be illegal to sink higher ranked
-identical expressions if they are not anticipable in the common
+identical expressions if they are not fully anticipable in the common
 post-dominator. For example:
@@
 (c1, c2) depend on i1 and i2 respectively which are also in their original basic
-blocks, c1 and c2 are not anticipable in B3. So without knowing the ability to sink
+blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
 of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
@@
 expressions are moved in the current implementation, there is no need to check
-for undefined resulting due to movement of instructions. This simplifies the
+for undefinedness resulting due to movement of instructions. This simplifies the
 implementation.

### 2017-09-11 paper.tex
typos

@@
 sibling branch to a common dominator if they are very busy
-\cite{steven1997advanced}. The cost-model to mitigate register pressure is
-missing, also there is no mention of sinking congruent instructions.
+\cite{steven1997advanced}. However, the cost-model to mitigate register pressure
+is missing, also, there is no mention of sinking congruent instructions.
 
@@
 part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
-expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
-basic-block B, i.e., values computed on all paths from B to exit and
+expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions
+(VBEs) at basic-block B, i.e., values computed on all paths from B to exit and
 AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
-top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
-be computed on every path from B to exit, and can be computed in B.  It uses
-AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
-function is: for each hoist candidate, if all successors of B are dominated by
-B, then we know insertion into B will eliminate all the remaining computations.
-It then checks to see if at least one successor of B has the value available.
-This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
-ensure that B has multiple successors, since hoisting in a straight line is
-pointless.  The algorithm continues top down the dominator tree, working in
-tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
-implementation is that it works in sync with the \GVN{}-\PRE{} such that when
-new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
-hoist them.
+VBEs to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will be
+computed on every path from the basic block B to exit, and can be computed in B.
+It uses AVAIL\_OUT[B] to subtract out those values already being computed.  The
+cost function is: for each hoist candidate, if all successors of B are dominated
+by B, then we know insertion into B will eliminate all the remaining
+computations.  It then checks to see if at least one successor of B has the
+value available.  This avoids hoisting it way up the chain to ANTIC\_IN[B].  It
+also checks to ensure that B has multiple successors, since hoisting in a
+straight line is pointless.  The algorithm continues top down the dominator
+tree, working in tandem with \PRE{} until no more hoisting is possible.  One
+advantage of \GCC{} implementation is that it works in sync with the
+\GVN{}-\PRE{} such that when new hoisting opportunities are created by
+\GVN{}-\PRE{}, code-hoisting will hoist them.
 
@@
 
-\citet{shobaki2013} also recently presented a combinatorial global
-scheduler with reasonable performance improvements. It is possible that both
-Barany and Shobaki's implementation will have similar results when compiled for
-same target architecture. Also, both suffer from the same problem, although
-Shobaki not so much, of large compile times which would not suit in industrial
-compilers like \GCC{} and \LLVM{}. With the algorithm described in the next
-section, we got a reduction in register spills, improved inliner heuristics,
-improved compile time, and show performance improvements on \SPEC{} benchmarks.
+\citet{shobaki2013} also presented a combinatorial global scheduler with
+reasonable performance improvements. It is possible that both Barany and
+Shobaki's implementation will have similar results when compiled for same target
+architecture. Also, both suffer from the same problem, although Shobaki not so
+much, of large compile times which would not suit in industrial compilers like
+\GCC{} and \LLVM{}. With the algorithm described in the next section, we got a
+reduction in register spills, improved inliner heuristics, improved compile
+time, and show performance improvements on \SPEC{} benchmarks.
 
@@
 We compute merge sets, of the control flow graph, which remain same throughout
-the \gcm{} transformation. For \gcm{} we only want to know if a use
-operand is a kill (to compute changes in register pressure.) For that we only
-need to know whether the use is also required later in the execution path. A
-variable is live out of a basic-block B if it is used in the merge set of B. We
-compute the live-out relation on-demand when profitability of hoist/sink
-candidates is to be evaluated. A simplified version of
-\emph{isLiveOutUsingMergeSet} is presented here:
+the \GCM{} transformation. For \GCM{} we only want to know if a use operand is a
+kill (to compute changes in register pressure). For that we only need to know
+whether the use is also required later in the execution path. A variable is live
+out of a basic-block B if it is used in the merge set of B. We compute the
+live-out relation on-demand when profitability of hoist/sink candidates is to be
+evaluated. A simplified version of \emph{isLiveOutUsingMergeSet} is presented
+here:
 
@@
 \item find candidates (congruent instructions) suitable for \gcm{}
-  (Section~\ref{subsec:finding-candidates},)
+  (Section~\ref{subsec:finding-candidates}),
 \item compute a point in the program where it is both legal
@@
 \item move the code to hoist point or sink point
-  (Section~\ref{subsec:optimistic},) and
-\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}.)
+  (Section~\ref{subsec:optimistic}) and
+\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
 \end{itemize}

### 2017-09-11 paper.tex
typos

@@
 The amount of hoisting depends on whether we collect \GVN{} of instructions
-before finding candidates (optimistic) or, on-demand (pessimistic.) It also
+before finding candidates (optimistic) or, on-demand (pessimistic). It also
 depends on the generality of the \GVN{} algorithm as mentioned earlier in
@@
 Code hoisting opens new opportunities for other hoistable candidates which were
-of higher rank (depended on candidates which got hoisted.) Ideally we could
+of higher rank (depended on candidates which got hoisted). Ideally we could
 iterate on lower ranking expressions first and then proceed to higher ranking
@@
 Barriers for each iteration of the \gcm{}, we have not seen significant
-increase in the compilation times (see Section~\ref{tab:compile-time}.) We have also
+increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
 provided appropriate compiler
@@
 (429.mcf, 458.sjeng, etc.) and a few regressions (447.dealII, 453.povray,
-etc.) The \gcm{} pass was run twice in the pass pipeline (at -Ofast,)
+etc.) The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
 first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes

### 2017-09-16 cgo/paper.tex
adding for cgo

--- 2017-09-11 paper.tex
@@
-%% For double-blind review submission
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission
-%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission
-%\documentclass[acmlarge]{acmart}\settopmatter{}
-
+%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
+\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%% For double-blind review submission, w/ CCS and ACM Reference
+%\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
+%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
+%\documentclass[acmsmall,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%% For single-blind review submission, w/ CCS and ACM Reference
+%\documentclass[acmsmall,10pt,review]{acmart}\settopmatter{printfolios=true}
+%% For final camera-ready submission, w/ required CCS and ACM Reference
+%\documentclass[acmsmall,10pt]{acmart}\settopmatter{}
+
+%% Journal information
+%% Supplied to authors by publisher for camera-ready submission;
+%% use defaults for review submission.
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
+\acmArticle{1}
+\acmYear{2017}
+\acmMonth{1}
+\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission;
+%% use 'none' for review submission.
+\setcopyright{none}
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2017}           %% If different from \acmYear
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format should change 'acmlarge' to
-%% 'sigplan,10pt'.
+%% SIGPLAN proceedings format must update the '\documentclass' and
+%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\makeatletter\if@ACM@journal\makeatother
-
-\makeatletter\preto{\@verbatim}{\topsep=1pt \partopsep=0pt}\makeatother
-
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2017}
-\acmMonth{1}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{none}             %% For review submission
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2017}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
 \begin{document}
@@
 \title[SSA GCM]{Global code motion of congruent instructions}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-%\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-%\subtitle{Subtitle}                     %% \subtitle is optional
-%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
 
@@
 \def \ooo {out-of-order}
+
 %% Author information
@@
 %% Author with single affiliation.
-
 \author{Aditya Kumar}
@@
 \email{s.pop@samsung.com}
-
-%% Paper note
-%% The \thanks command may be used to create a "paper note" ---
-%% similar to a title note or an author note, but not explicitly
-%% associated with a particular element.  It will appear immediately
-%% above the permission/copyright statement.
-\thanks{with paper note}                %% \thanks is optional
-                                        %% can be repeated if necesary
-                                        %% contents suppressed with 'anonymous'
 
@@
 
-\ccsdesc[500]{Software and its engineering~Compilers}
-\ccsdesc[500]{Software and its engineering~Source code generation}
-\ccsdesc[500]{General and reference~Performance}
-
-%\ccsdesc[300]{Social and professional topics~History of programming languages}
+\ccsdesc[500]{Software and its engineering~General programming languages}
+\ccsdesc[300]{Social and professional topics~History of programming languages}
 %% End of generated code
@@
 %% comma separated list
-\keywords{Optimizing Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional
+\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
 
@@
 \section{Introduction}
+
 \label{sec:intro}
@@
   float inv = 1.0f / d;
-  if (inv >= 0) { tmin = (min - a) * inv; tmax = (max - a) * inv; }
-  else          { tmin = (max - a) * inv; tmax = (min - a) * inv; }
+  if (inv >= 0) {
+    tmin = (min - a) * inv;
+    tmax = (max - a) * inv;
+  } else {
+    tmin = (max - a) * inv;
+    tmax = (min - a) * inv;
+  }
   return tmax + tmin;
@@
 
-In this program the computations of tmax and tmin are identical to the
-computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
-depend on inv which depends on a division operation which is generally more
+In this program the computations of $tmax$ and $tmin$ are identical to the
+computations of $tmin$ and $tmax$ of sibling branch respectively. Both $tmax$ and $tmin$
+depend on $inv$ which depends on a division operation which is generally more
 expensive than the addition, subtraction, and multiplication operations. The
@@
 
-Now if the computation of tmax and tmin are hoisted outside the
+Now if the computation of $tmax$ and $tmin$ are hoisted outside the
 conditionals, the C code version would look like this:
@@
 
-\newpage
-
 %% Bibliography
 \bibliography{Bibliography}
+%\bibliography{bibfile}
 

### 2017-09-16 cgo/paper.tex
plot instead of table for gcm metric

@@
 \usepackage{algpseudocode}
+\usepackage[final]{pdfpages}
 
@@
 \begin{verbatim}
-// Return true if the merge-set of source node of a visited J-edge (incoming edge of
-// lnode) is not the subset of the merge-set of lnode.
-bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet MergeSet)
-  for e in (all incoming edges to lnode)
-    if e is in JE and e in V
-      if MergeSet(Source Node of e) is not a subset of MergeSet(lnode)
+// Return true if the merge-set of source node of a
+// visited J-edge (incoming edge of lnode) is not
+// the subset of the merge-set of lnode.
+bool
+still_inconsistent(Node lnode, JEdges JE,
+                   Visited V, MergeSet MergeSet)
+  for e in (all incoming edges to lnode):
+    if e is in JE and e in V:
+      if MergeSet(Source Node of e)
+         is not a subset of MergeSet(lnode):
         return true
@@
 \begin{verbatim}
-// Compute a merge-set top-down in breadth first order.
-bool constructMergeSet_1(BFSList B, JEdges JE, DomLevel DL)
+// Compute a merge-set breadth first order.
+bool constructMergeSet_1(BFSList B, JEdges JE,
+                         DomLevel DL)
   repeat = false
   V = empty list of visited edges
-  for n in B
-    for e in (all incoming edges to n)
-      if e is in JE and e not in V
+  for n in B:
+    for e in (all incoming edges to n):
+      if e is in JE and e not in V:
         V(e) = true
@@
         let lnode = NULL
-        while (DL(tmp) >= DL(tnode))
-          MergeSet(tmp) = MergeSet(tmp) U MergeSet(tnode) U {tnode}
+        while (DL(tmp) >= DL(tnode)):
+          MergeSet(tmp) = MergeSet(tmp)
+                          U MergeSet(tnode)
+                          U {tnode}
           lnode = tmp
@@
 \begin{verbatim}
-// Construct merge-set of each node in Control-Flow-Graph G of a function.
+// Construct merge-set of each node in
+// Control-Flow-Graph G of a function.
 void constructMergeSet(CFG G)
@@
   DL = List of Path length (from root) of each node in G.
-  do // Call until a fixed point is reached.
+  do: // Call until a fixed point is reached.
     Repeat = constructMergeSet_1(B, JE, DL)
@@
 % llvm trunk vs. llvm trunk with code-motion
-                &\% performance uplift & size reduction in Bytes & size reduction in Bytes\\
+                &\% performance uplift & size reduction (Bytes) & size reduction (Bytes)\\
 Benchmarks      & at -Ofast (high better)& at -Ofast (high better) & at -Os (high better)\\\hline
@@
       \hline
-Number of spills  &base(-Ofast) &\GCM{}(-Ofast) &\%change (low better) \\\hline
+Number of spills  &base(-Ofast) &\GCM{}(-Ofast) &\%change\\\hline
 400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
@@
   \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at
-    -Ofast. Spills reduced by 2\%.}
+    -Ofast. Lower is better in \%age change. Spills reduced by 2\%.}
   \label{tab:spills}
@@
       \hline
-  Benchmarks      &   Baseline	        & \GCM{}           & \%Decrease (low better) \\\hline
-  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
-  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
-  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
-  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
-  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
-  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
-  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
-  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
-  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
-  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
-  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
-  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
-  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
-  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
-  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
-  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
-  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
-  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
-  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
+      Benchmarks      &   Baseline	        & \GCM{}           & \%Decrease\\
+       & (in Millions)& (in Millions)& (lower is better)\\\hline
+  400.perlbench	  &   186	& 184	& 0.99 \\\hline
+  401.bzip2	  &   76	& 75	& 0.99 \\\hline
+  403.gcc         &   1,079	& 1,073	& 0.99 \\\hline
+  429.mcf         &   78	& 77	& 0.99 \\\hline
+  433.milc	  &   272	& 269	& 0.99 \\\hline
+  444.namd	  &   78	& 77	& 0.99 \\\hline
+  445.gobmk	  &   261	& 258	& 0.99 \\\hline
+  447.dealII	  &   643	& 638	& 0.99 \\\hline
+  450.soplex	  &   245	& 242	& 0.99 \\\hline
+  453.povray	  &   499	& 494	& 0.99 \\\hline
+  456.hmmer	  &   214	& 211	& 0.99 \\\hline
+  458.sjeng	  &   89	& 88	& 0.99 \\\hline
+  462.libquantum  &   84	& 84	& 0.99 \\\hline
+  464.h264ref	  &   151	& 149	& 0.99 \\\hline
+  470.lbm         &   71	& 71	& 1.00 \\\hline
+  471.omnetpp	  &   382	& 379	& 0.99 \\\hline
+  473.astar	  &   78	& 77	& 0.99 \\\hline
+  482.sphinx3	  &   158	& 156	& 0.99 \\\hline
+  483.xalancbmk	  &   22,485	& 22,455& 1.00 \\\hline
     \end{tabular}}
@@
     \scalebox{0.8}{
-    \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
+    \begin{tabular}{|l|r|r|r|r|}
       \hline
-  &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
- Benchmarks          & Base        & \GCM{}	& \%  &Base     & \GCM{}   &\%    & Base	& \GCM{}	  &\%   & Base	& \GCM{}	& \% \\\hline
- 400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
- 401.bzip2	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
- 403.gcc	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
- 429.mcf	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
- 433.milc	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
- 444.namd	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
- 445.gobmk	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
- 447.dealII	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
- 450.soplex	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
- 453.povray	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
- 456.hmmer	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
- 458.sjeng	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
- 462.libquantum   & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
- 464.h264ref	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
- 470.lbm	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
- 471.omnetpp	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
- 473.astar	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
- 482.sphinx3	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
- 483.xalancbmk	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
+      &\multicolumn{2}{|c|}{Loop Invariant Motion}  & \multicolumn{2}{|c|}{Functions} \\\hline
+ Benchmarks       & Hoisted \% & Sunk \% &  Inlined\%   & Deleted\% \\\hline
+ 400.perlbench	  & 2.1 & -2.8 &  0.2 &  0.9  \\\hline
+ 401.bzip2	  & 14.4& 0.0  &  4.7 &  3.1  \\\hline
+ 403.gcc	  & 13.7& 27.3 &  0.9 &  0.1  \\\hline
+ 429.mcf	  & -7.3& 0.0  &  0.0 &  0.0  \\\hline
+ 433.milc	  & 10.2& 0.0  &  8.0 &  0.0  \\\hline
+ 444.namd	  & 0.1 & 0.0  &  0.8 &  -3.5 \\\hline
+ 445.gobmk	  & 0.6 & 5.6  &  5.4 &  1.7  \\\hline
+ 447.dealII	  & 62.2& -86.7& -0.2 &  -0.8 \\\hline
+ 450.soplex	  & 3.9 & -5.6 & -0.2 &  -0.4 \\\hline
+ 453.povray	  & -4.8& 39.1 &  0.2 &  -0.1 \\\hline
+ 456.hmmer	  & 1.0 & 166.7&  0.0 &  -1.9 \\\hline
+ 458.sjeng	  & 11.5& 0.0  &  6.3 &  0.0  \\\hline
+ 462.libquantum   & 6.6 & 0.0  &  0.0 &  0.0  \\\hline
+ 464.h264ref	  & 0.5 & 0.0  &  13.0&  0.0  \\\hline
+ 470.lbm	  & 0.0 & 0.0  &  0.0 &  0.0  \\\hline
+ 471.omnetpp	  & 17.0& -6.3 &  1.1 &  -0.4 \\\hline
+ 473.astar	  & 9.1 & 0.0  &  1.1 &  0.0  \\\hline
+ 482.sphinx3	  & 5.9 & -18.2&  -2.0&  0.0  \\\hline
+ 483.xalancbmk	  & 14.6& -3.6 &  -1.0&  -1.2 \\\hline
     \end{tabular}}
   \end{center}
-  \caption{Static metrics before (Base) and after \gcm{} (\GCM{}) on
-    \SPEC{} at -Ofast. The \% Column show percentage increase in metric
-    w.r.t. Base. \GCM{} improves LICM and inlining.}
+  \caption{Change in the static compile time metrics w.r.t. \gcm{} (\GCM{}) on
+    \SPEC{} at -Ofast. The \% columns show percentage increase in metric
+    w.r.t. baseline. \GCM{} improves LICM and inlining in general.}
   \label{tab:stats}
@@
 
-\begin{table}[h!]
-  \begin{center}
-    \scalebox{0.8}{
-    \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|}
-      \hline
- \GCM{}  &\multicolumn{2}{|c|}{Total Instructions} &\multicolumn{2}{|c|}{Scalars}  & \multicolumn{2}{|c|}{Loads} & \multicolumn{1}{|c|}{Stores} &Calls   &Instructions    \\\hline
- metric           &  H          &R        &H      &R       &H           &R              &H=R    &H & sunk \\\hline
- 400.perlbench	  &  2387	& 2456	  &  855  & 857	   & 1493	& 1560		& 38	& 1	& 4	        \\\hline
- 401.bzip2	  &  176	& 300	  &  27	  & 27	   & 149	& 273		& 0	& 0 	& 0		\\\hline
- 403.gcc	  &  5670	& 5761	  &  2558 & 2567   & 3109	& 3191		& 3	& 0	& 1	        \\\hline
- 429.mcf	  &  42	        & 43	  &  5	  & 5	   & 37 	& 38		& 0	& 0 	& 0		\\\hline
- 433.milc	  &  128	& 130	  &  10	  & 10	   & 117	& 119		& 0	& 1	& 0		\\\hline
- 444.namd	  &  262	& 264	  &  113  & 113	   & 148	& 150		& 0	& 1	& 0		\\\hline
- 445.gobmk	  &  1021	& 1052	  &  312  & 314	   & 705	& 734		& 3	& 1	& 1	        \\\hline
- 447.dealII	  &  10455	& 12325	  &  4009 & 4063   & 6394	& 8210		& 20	& 32	& 0	        \\\hline
- 450.soplex	  &  2389	& 2731	  &  927  & 1004   & 1437	& 1702		& 14	& 11	& 0	        \\\hline
- 453.povray	  &  1967	& 2269	  &  560  & 590	   & 1389	& 1661		& 4	& 14	& 0	        \\\hline
- 456.hmmer	  &  408	& 427	  &  82	  & 82	   & 319	& 338		& 1	& 6	& 0	        \\\hline
- 458.sjeng	  &  372	& 399	  &  128  & 138	   & 234	& 251		& 1	& 9	& 1	        \\\hline
- 462.libquantum	  &  49	        & 49	  &  17	  & 17	   & 32	      	& 32		& 0	& 0	& 0		\\\hline
- 464.h264ref	  &  2953	& 3438	  &  712  & 779	   & 2238	& 2656		& 2	& 1	& 1	        \\\hline
- 470.lbm	  &  23	        & 23	  &  4	  & 4	   & 19		& 19		& 0	& 0	& 0		\\\hline
- 471.omnetpp	  &  391	& 408	  &  92	  & 92	   & 297	& 314		& 0	& 2	& 0		\\\hline
- 473.astar	  &  202	& 222	  &  62	  & 64	   & 140	& 158		& 0	& 0	& 0		\\\hline
- 482.sphinx3	  &  168	& 172	  &  32	  & 32	   & 136	& 140		& 0	& 0	& 0		\\\hline
- 483.xalancbmk	  &  5894	& 6162	  &  2586 & 2660   & 3301	& 3495		& 4	& 3	& 2	        \\\hline
- Grand total      &34957	&38631	  &13091  &13418   &21694	&25041	        &90	&82     &10             \\\hline
-    \end{tabular}}
-  \end{center}
-  \caption{\GCM{} stats on SPEC2006 at -Ofast, H(Hoisted,) R(Removed); Loads are
-    hoisted the most followed by scalars, stores and calls.}
+\begin{figure*}
+\includegraphics[width=\textwidth,height=10cm]{gcm-stats.pdf}
+\caption{\GCM{} stats on SPEC2006 at -Ofast. Loads are hoisted the most followed
+  by scalars, stores and calls.}
 \label{tab:code-motion-metric}
-\end{table}
+\end{figure*}
 

### 2017-09-16 cgo/paper.tex
code formatting

@@
   JE = JEdges of G
-  DL = List of Path length (from root) of each node in G.
+  DL = Path lengths (from root) of each node in G
   do: // Call until a fixed point is reached.
@@
 \begin{verbatim}
-// Return true if variable A is liveout from basic-block N
+// Return true if variable A is liveout from N
 bool isLiveOutUsingMergeSet(Node N, Variable A)
@@
   MergeSet(N) = null
-  // The merge-set of N is the union of merge-sets of its successors.
+  // The merge-set of N is the union of
+  // merge-sets of its successors.
   for W in successors(N)
@@
 
-  // Iterate over all the uses of A and see if any intersect with the merge-set of N.
+  // Iterate over all the uses of A and see
+  // if any intersect with the merge-set of N.
   for T in users(A)
@@
 // candidates.
-void compute_barriers()
+void computeBarriers()
   for each basic-block B in a function:
@@
 
-    // Find the last barrier below which instructions can be sunk.
-    // If there was no barrier in B, any instruction satisfying
-    // other legality checks can be sunk.
+    // Find the last barrier below which instructions
+    // can be sunk. If there was no barrier in B, any
+    // instruction satisfying other legality checks
+    // can be sunk.
     if barrier_found:
@@
 \begin{verbatim}
-// Return true if a hoistable instruction is downward safe at hoist-point
-bool compute_downward_safety()
-  Worklist = list of all basic-blocks of hoistable instructions
-  for each basic-block B in the dominator tree starting at hoist-point:
+// Return true if a hoistable instruction is downward
+// safe at hoist-point
+bool compute_downward_safety(Node HP)
+  Worklist = basic-blocks of hoistable instructions
+  DT = The dominator tree
+  HP = the hoist-point
+  for each basic-block B in the DT starting at HP:
     if Worklist is empty
@@
       remove B from worklist
-      remove subtree with root at B // Available at B => downward safe from B
+      // Available at B => downward safe from B
+      remove subtree with root at B
     if B is a leaf basic-block
-      return false  // Path exists! not downward safe
+      // Path exists! not downward safe
+      return false
     if B dominates hoist-point
@@
 In order to hoist a subset of identical instructions, we partition the list of
-all candidates in a way to maximize the total number of hoist instructions.  By sorting
-the list of all the candidates in the increasing order of their depth first
-search discovery time stamp \cite{clrs} (DFSIn numbers,) we make sure that
+all candidates in a way to maximize the total number of hoist instructions.  By
+sorting the list of all the candidates in the increasing order of their depth
+first search discovery time stamp \cite{clrs} (DFSIn numbers,) we make sure that
 candidates closer in the list have their common dominator nearby. In
@@
 void doGCM()
-  Analyses available: Dominator Tree, DFS Numbering, Memory SSA,
+  Analyses available:
+    Dominator Tree, DFS Numbering, Memory SSA
   Compute DJ-graph of function
-  compute mergeset based on that (using constructMergeSet function)
-  compute barriers (using compute_barriers function)
-  do
+  constructMergeSet(CGF)
+  computeBarriers()
+  do:
     Compute GVN of each expression in the function
-  while (doCodeHoisting() > 0) // Repeat hoisting if any candidates were hoisted.
+   // Repeat hoisting if any candidates were hoisted.
+  while (doCodeHoisting() > 0)
   doCodeSinking()
@@
 \begin{verbatim}
-int doCodeHoisting() // Find hoistable candidates and do hoisting
-  For each value number VN with 2 or more instructions (I1, I2, ..., Ik)
+// Find hoistable candidates and do hoisting
+int doCodeHoisting()
+  For each value number VN with 2 or more instructions
     sort the instructions according to DFSIn numbers
-    if first two I1, I2 are hoistable and if they are downward-safe at hoist-point
-          and if they are profitable to be hoisted at hoist-point
-      proceed to check the ability to hoist subsequent instruction with same VN
-    else
-      proceed to check if I2 and I3 (if available) can be hoisted
-  For each set of instructions (I1, ..., Ik) with same VN that can be hoisted:
-    move the first one I1 to the hoist-point
-    update the use of all other candidates (I2, ..., Ik) to refer to I1
+    if first two I1, I2 are hoistable
+        && downward-safe at hoist-point
+        && profitable to be hoisted at hoist-point:
+      continue the check for subsequent
+        instructions with same VN
+    else:
+      proceed to check if I2 and I3 (if available)
+      can be hoisted
+
+  For each set of hoistable instructions:
+    move the first one (I1) to the hoist-point
+    update their users others to refer to I1
     remove all other instructions (I2, ..., Ik)
-    if I1 has memory references (a load, store, call, etc.)
-      update MemorySSA of I1 and others to point to the MemorySSA reference of I1
-      remove MemorySSA reference of all others which were deleted
+    if I1 has memory references:
+      update MemorySSA of I1 and others
     update statistics
@@
 \begin{verbatim}
-void doCodeSinking()
+int doCodeSinking()
   // Find sinkable instructions
-  For each value number VN with 2 or more instructions
-    if there are two instructions with same immediate successor as post-dominator
-      if both are only used in the same PHI-node (PN) of the successor
-        if both have dependencies that are available at the sink point
-           // available or can be made available by rematerialization
-           mark VN as sinkable
+  For each value number VN with 2 or more instructions:
+    if there are two instructions with same immediate
+        successor(S) as post-dominator:
+      if both are only used in the same PHI-node of S
+          && dependencies of both are available at S:
+         mark instructions as sinkable
 
   // Sink the instructions and update statistics
-  For each pair (I1, I2) of sinkable instructions
+  For each pair (I1, I2) of sinkable instructions:
     Move I1 to the sink point
-    // the sink point is just after all the PHI-nodes in the post-dominator
+    // the sink point is just after all
+    // the PHI-nodes in the post-dominator
     update all the uses of PHI node (PN) with I1
@@
       update MemorySSA of I1
-      update MemorySSA of removed instructions to point to the MemorySSA reference of I1
-      remove MemorySSA reference of all others which were deleted
+      update MemorySSA of removed instructions
+        to point to the MemorySSA reference of I1
+      remove MemorySSA reference of all
+        others which were deleted
     update statistics

### 2017-09-16 cgo/paper.tex
final cgo

@@
 %% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2017}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+%\acmJournal{PACMPL}
+%\acmVolume{1}
+%\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
+%\acmArticle{1}
+%acmYear{2017}
+%\acmMonth{1}
+%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
 \startPage{1}
+
+%\setlength{\abovecaptionskip}{5pt plus 1pt minus 1pt} % Chosen fairly arbitrarily
+%\setlength{\belowcaptionskip}{5pt plus 1pt minus 1pt} % Chosen fairly arbitrarily
 
@@
 \usepackage{algpseudocode}
+\usepackage{enumitem}
 \usepackage[final]{pdfpages}
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Optimizing Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional
 
@@
 of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
-and brought these techniques in the main stream: nowadays \SSA{} based \PRE{} is
-available in industrial compilers like \GCC{} and \LLVM{}.
+and brought these techniques in the main stream: \SSA{} based \PRE{} is
+available in many compilers like \GCC{} and \LLVM{}.
 
@@
 at the hoisting point, the hardware has more independent instructions to
-reorder. The following example illustrates how hoisting can improve performance
-by exposing more ILP.
+reorder. The following reduced example illustrates how hoisting can improve
+performance by exposing more ILP.
 
@@
 Now if the computation of $tmax$ and $tmin$ are hoisted outside the
-conditionals, the C code version would look like this:
+conditionals, the C code version would look like:
 \begin{verbatim}
@@
 $C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
-seen the performance of an \ooo{} processor going up by $15\%$ on a benchmark
-containing the above motivating example running in a tight loop.  In fact, there
-are several advantages of \GCM{}:
-\begin{itemize}
+seen the performance of the original benchmark (on an \ooo{} processor) improve
+by $15\%$.  In fact, there are several advantages of \GCM{}:
+\begin{itemize}[leftmargin=*,topsep=0pt]
 \item it helps reduce the code size of the program by replacing multiple
@@
 contributions of this paper are:
-\begin{itemize}
+\begin{itemize}[leftmargin=*,topsep=0pt]
 \item a new optimistic algorithm to move congruent instructions out of branches;
@@
 
-The algorithm for \gcm{} uses several common representations of the
+The algorithm for \GCM{} uses several common representations of the
 program that we shortly describe below:
-\begin{itemize}
+\begin{itemize}[leftmargin=*,topsep=0pt]
 \item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
@@
 Dominance graphs, Global Value Numbers and Memory SSA are already available in
-\LLVM{} but there is no facility to infer liveness of virtual registers. For
-that we implemented DJ-Graph data structure that allows us to calculate
-MergeSets which is used in the liveness analysis in our implementation of \GCM{}.
+\LLVM{} but there is no facility to infer liveness of virtual registers. So we
+implemented DJ-Graph data structure that allowed us to calculate MergeSets which
+was used in the liveness analysis during \GCM{}.
 
@@
 bool
-still_inconsistent(Node lnode, JEdges JE,
+stillInconsistent(Node lnode, JEdges JE,
                    Visited V, MergeSet MergeSet)
@@
 merge-set that is a subset of the merge-set of its successor \cite{das2005}. The
-function \emph{still\_inconsistent} returns true if this invariant is not
+function \emph{stillInconsistent} returns true if this invariant is not
 satisfied which allows the function \emph{constructMergeSet} to repeat.
@@
 \end{verbatim}
-\newpage
+
 \begin{verbatim}
@@
 reach to the beginning of a function, in that case the inner while loop needs to
-terminate. These two cases were missing from the algorithm and we came across
-them during implementation.
+terminate. These two cases were missing from the original algorithm.
 
@@
 describe in the rest of this section:
-\begin{itemize}
+\begin{itemize}[leftmargin=*,topsep=0pt]
 \item find candidates (congruent instructions) suitable for \gcm{}
@@
 \item move the code to hoist point or sink point
-  (Section~\ref{subsec:optimistic}) and
+  (Section~\ref{subsec:optimistic}),
 \item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
@@
 discarded. To iterate on the use-def chains for memory references the MemorySSA
-infrastructure of \LLVM{} is used.
+of \LLVM{} is used.
 
@@
 loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
-instruction are available at the hoisting point, and that there are no aliasing
-loads or store along the path from the current position to the hoisting point.
+instruction are available at the hoisting point, and there are no aliasing
+memory references along the hoisting path.
 
@@
 introduce a new computation along any path of execution without checking for
-undefined behaviors. The algorithm \emph{compute\_downward\_safety} computes if
+undefined behaviors. The algorithm \emph{computeDownwardSafety} computes if
 hoisting a set of instruction to a common hoisting point is downward-safe
@@
 // safe at hoist-point
-bool compute_downward_safety(Node HP)
+bool computeDownwardSafety(Node HP)
   Worklist = basic-blocks of hoistable instructions
@@
 across popular benchmarks and at the same time to reduce code size as much as
-possible. Following are the cost models which are implemented:
+possible. Following cost models are implemented:
 
@@
     if c is true then goto B3 else goto B4
-
 B1: a0 = b<kill> + c<kill>
-
 B2: a1 = b<kill> + c<kill>
@@
 
-After hoisting a0 and a1 are removed and a copy of a0 as a01 is placed in B0
-just before the branch. Since `b' and `c' are killed in `a0' and `a1', hoisting
-the expressions will reduce the register pressure because two registers will be
-freed.
+After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ as $a1$ is placed
+in $B0$ just before the branch. Since $b$ and $c$ are killed in $a0$ and $a1$,
+hoisting the expressions will reduce the register pressure because two registers
+will be freed.
 
@@
 live-range-splitting \cite{cooper1998live} pass should make the right decision
-of rematerializing the instruction should it be beneficial to do so. But the
-current live-range splitting pass of \LLVM{} is not making the optimal decision
-and we have found performance regressions while hoisting aggressively.
-
-Moreover, \LLVM{} has a `getelementptr' instruction which computes the address
-where a load or a store would happen. It is a scalar computation and gets
-hoisted frequently even if the corresponding loads/stores would not be legal to
-hoist.  In order to reduce the register pressure while hoisting loads, we have
-restricted hoisting of address computations away from their corresponding loads
-and stores when the loads and stores cannot be moved. This restriction is only
-to mitigate the limitations of \LLVM{'}s register allocator and may be lifted in
-the future, when the register allocation rematerialization pass has been
-improved to catch these regressions.
+but the current live-range splitting pass of \LLVM{} is not making the optimal
+decision and we have found performance regressions while hoisting aggressively.
 
@@
 pressure would increase, we prefer to hoist loads in this case. The reason
-being, that would make the loaded value available by the time it is used. Also,
-because stores and calls are hoisted the least, see
-Table~\ref{tab:code-motion-metric}, the performance does not change much whether
-they are hoisted or not.
+being, that would make the loaded value available by the time it is used.
 
@@
 could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
-still be hoisted -- if legal and profitable -- to B2.
+still be hoisted - if legal and profitable - to B2.
 
@@
 updating the intermediate representation (IR) to reflect the changes. At the
-same time \MemorySSA{} is also updated to get up-to-date information about
-memory references.
+same time \MemorySSA{} is updated as well.
 
@@
 which can be partially hoisted to their respective common dominators.
-\newpage
+
 \begin{verbatim}
@@
 performed once because currently there are very few sinkable candidates
-(Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
+(Figure~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
 \GCM{}) has a code-sinking which already sinks several instructions.
@@
 compiler statistics are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
-Table~\ref{tab:code-motion-metric}.  We also measured compile time
+Figure~\ref{tab:code-motion-metric}.  We also measured compile time
 Table~\ref{tab:compile-time} to see the impact of \gcm{}.
@@
   \caption{Execution time (ratio) and code size change (Bytes) with and without
-    \gcm{} (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng;
-    code size improvement in dealII at -Os.}
+    \GCM{} on \SPEC{}: performance uplift at in mcf and sjeng;
+    code size improvement in dealII.}
   \label{tab:code-size}
@@
 congruent instructions out of branches.
+
+\begin{figure*}[h!]
+  \includegraphics[width=0.8\textwidth,height=8cm]{gcm-stats.pdf}
+  \vspace*{-1.5cm}
+\caption{\GCM{} stats on SPEC2006 at -Ofast. Loads are hoisted the most followed
+  by scalars, stores and calls.}
+\label{tab:code-motion-metric}
+\end{figure*}
 
@@
 to improve performance at the cost of code-size at -Ofast, so passes like
-inlining, loop-unrolling, code-versioning are very aggressive at -Ofast.
+inlining, loop-unrolling, code-versioning are aggressive at -Ofast.
 
@@
 
-Some useful static metrics are presented in Table~\ref{tab:stats}, which shows how \gcm{}
-impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
-code out of the loop. It may hoist the code to the loop's preheader or sink the code to loop's
-post-dominator \cite{steven1997advanced}. As we can see number of instructions hoisted increased
-significantly in several benchmarks, although number of instructions sunk did not change by much.
-Number of inlined functions also improved for most benchmarks. In general when more functions
-were inlined, more functions were deleted as in 400.perlbench, 445.gobmk.
+Some useful static metrics are presented in Table~\ref{tab:stats}, which shows
+how \gcm{} impacted important compiler optimizations like inlining and
+LICM. LICM removes loop invariant code out of the loop. It may hoist the code to
+the loop's preheader or sink the code to loop's post-dominator
+\cite{steven1997advanced}. As we can see number of instructions hoisted
+increased significantly in several benchmarks, although number of instructions
+sunk did not change by much.  Number of inlined functions also improved for most
+benchmarks. In general when more functions were inlined, more functions were
+deleted as in 400.perlbench, 445.gobmk.
 
 The compile time metrics of \GCM{} are listed in
-Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
+Figure~\ref{tab:code-motion-metric}. The table lists the number of scalars,
 loads, stores and calls hoisted as well as removed. For each category, the
@@
 which dependencies are not directly available in the successor
-(Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking those
-instructions would require a look ahead on the ability to sink the
-dependencies. We plan to implement this in near future.
-
-\begin{figure*}
-\includegraphics[width=\textwidth,height=10cm]{gcm-stats.pdf}
-\caption{\GCM{} stats on SPEC2006 at -Ofast. Loads are hoisted the most followed
-  by scalars, stores and calls.}
-\label{tab:code-motion-metric}
-\end{figure*}
-
-The code shown in Section~\ref{sec:intro} is a reduced example that appears in a
-hot loop of a proprietary benchmark.  When the expressions are hoisted from the
-conditional clauses, the overall performance of that benchmark improves by
-$15\%$ on an \ooo{} processor due to increased instruction level
-parallelism, and better scheduling of the instructions, accommodating for the
-long latency of the division operation.
+(Section~\ref{subsec:legality-sink}) is not sunk for now because that would
+require a look ahead on the ability to sink the dependencies. We plan to
+implement this in near future.
 
@@
 too much, we have implemented a register pressure aware cost model as described
-in Section~\ref{subsec:cost-models}. A part of this implementation is already
-merged in LLVM trunk as GVNHoist.cpp and we posted the \GCM{} implementation for
-review
-
-There are a few improvements that we would like to make. First would be to
-integrate this with \LLVM{'}s new GVN-PRE implementation such that more
-congruent instructions can be scheduled. This would also expose more
+in Section~\ref{subsec:cost-models}. A part of our implementation is merged in
+LLVM trunk as GVNHoist.cpp and the \GCM{} implementation is under review.
+
+We would like to integrate \GCM{} with \LLVM{'}s new GVN-PRE implementation such
+that more congruent instructions can be scheduled. This would also expose more
 redundancies to the GVN-PRE. In general it is a good idea to start with lower
@@
 implementation converges quickly and no compile time regression have been
-observed. This is not the most optimal approach in terms of compile time complexity
-and results in multiple data structures to be recomputed. This can be improved
-by ranking the computations \cite{rosen1988global}. Also, since \GCM{} runs very
-early in the pass pipeline, it will be good to evaluate the code size/performance
-impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.
-
-With the implementation of \GCM{} in \LLVM{}, the passes which rely on the
-code-size or instruction-counts to make optimization decisions need to be
-revisited. The first candidate would be the inliner. We have seen difference
-in inlining decisions in Table~\ref{tab:code-size}, when \gcm{}
-was enabled.  Since inliner has several cost-models tuned for the previous
-pass layout, that would need some improvement.
+observed. This can be improved by ranking the computations
+\cite{rosen1988global}. Also, since \GCM{} runs very early in the pass pipeline,
+it will be good to evaluate the code size/performance impact when it is run in
+sync with \GVN{}-\PRE{} just like \GCC{} does. With the implementation of \GCM{}
+in \LLVM{}, the passes which rely on the code-size or instruction-counts to make
+optimization decisions need to be revisited for example, the inliner.
 

### 2017-10-29 paper.tex
added explanation for chi

@@
 %% Title information
-\title[SSA GCM]{Global code motion of congruent instructions}         %% [Short Title] is optional;
+\title[SSA GCM]{A fast global code motion algorithm for code size optimization}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 % Experimental results: Register spills, compile time, run time.
+\section{Code hoisting of congruent computations}
+Examples to show CHI insertion.
+
+The algorithm factors out the reachability of values such that multiple queries
+to find reachability of values are fast. This is based on finding the ANTIC
+points in the CFG which do not change during hoisting. The ANTIC points are
+basically the dominance-frontiers in the inverse graph. So we introduce a data
+structure (CHI nodes) to keep track of values flowing out of a basic block. We
+only do this for values with multiple occurrences in the function as they are
+the potentially hoistable candidates. This approach allows us to hoist
+instructions to a basic block with more than two successors, as well as deal
+with infinite loops in a trivial way.
+
+Limitations: This pass does not hoist fully redundant expressions because they
+are already handled by GVN-PRE. It is advisable to run gvn-hoist before and
+after gvn-pre because gvn-pre creates opportunities for more instructions to be
+hoisted.
+
+\includegraphics[width=0.8\textwidth,height=8cm]{chi-example.pdf}
+
+A CHI keeps information about values flowing out of a basic block.  It is
+similar to PHI but in the inverse graph, and used for outgoing values on each
+edge. For conciseness, it is computed only for instructions with multiple
+occurrences in the CFG because they are the only hoistable candidates.
+
+The global value number for both I1 and I2 is V, the CHI node will save the
+instruction as well as the edge where the value is flowing to.  Where CHIArg1,
+CHIArg2 are of type CHIArg.
+
+\begin{verbatim}
+struct CHIArg {
+  // Edge destination (shows the direction of flow), may not be where the I is.
+  BasicBlock *Dest;
+  // The instruction which uses the values flowing out of CHI.
+  Instruction *I;
+  // The global value number of I.
+  GVNType VN;
+};
+
+\end{verbatim}
+
+Essentially, CHIArg1 = {V, B, I1} and CHIArg2 = {V, C, I2}
+
+For sinking we use the similar approach as \ref{SSAPRE} to track values flowing
+into a basic block.
+
+TODO: Explain ranking
 

### 2017-11-04 cgo/paper.tex
correction

@@
 decrease the live-range of the defined register but increase the live-range for
-killed operands. If the live-range after \gcm{} is less than before it will
-be moved. Essentially, as long as there is one killed operand, code hoisting
-will either decrease or preserve the register pressure.  Similarly, code-sinking
-will either decrease or preserve the register pressure as long as there is one
-operand killed at most.  The following example explains how \gcm{} of
-identical computations can reduce the register pressure.
+killed operands. If the live-range after \GCM{} will decrease, it will be
+moved. Essentially, as long as there is one killed operand, code hoisting will
+either decrease or preserve the register pressure.  Similarly, code-sinking will
+either decrease or preserve the register pressure as long as there is one
+operand killed at most.  The following example explains how \GCM{} of can reduce
+the register pressure.
 
@@
     c = n
-    if c is true then goto B3 else goto B4
+    if c is true then goto B1 else goto B2
 B1: a0 = b<kill> + c<kill>
+    goto B3
 B2: a1 = b<kill> + c<kill>
-\end{verbatim}
-
-After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ as $a1$ is placed
-in $B0$ just before the branch. Since $b$ and $c$ are killed in $a0$ and $a1$,
-hoisting the expressions will reduce the register pressure because two registers
-will be freed.
+    goto B3
+\end{verbatim}
+
+After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ is placed in
+$B0$. Since $b$ and $c$ are killed in both $a0$ and $a1$, hoisting the
+expressions will reduce the register pressure because two registers will be
+freed at the insertion point but only one register will be required to assign
+the definition of hoisted instruction $a0$.
 
@@
 Even hoisting scalars across calls is tricky because it can increase the number
-of spills. During the frame lowering of calls, the argument registers also known
-as the caller saved registers are saved because they might be modified by the
-callee and after the call they are restored. So before the call, the register
+of spills. During the frame lowering of calls, the argument registers (also
+known as the caller saved registers) are saved because they might be modified by
+the callee and after the call they are restored. Due to this the register
 pressure is high because the number of available registers are reduced by the

### 2017-11-05 cgo/paper.tex
correction from reviewers

@@
 saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
-inlining and exposes more loop invariant code motion opportunities in majority
-of the benchmarks. We have also seen execution time improvements in a few of
-\SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\%
-on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
+inlining and exposes more loop invariant code motion opportunities in the
+majority of benchmarks. We have also seen execution time improvements in a few
+of \SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by
+2\% on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
 optimistic algorithm in the sense that it considers all congruent instructions
@@
 scheduling to improve performance. Most of the recent work on global scheduling
-are done using integer linear programming (ILP,) which results in prohibitively
+are done using integer linear programming (ILP) which results in prohibitively
 high compile time. To the best of our knowledge we have not found any reference
@@
 
-We compute merge sets, of the control flow graph, which remain same throughout
-the \GCM{} transformation. For \GCM{} we only want to know if a use operand is a
-kill (to compute changes in register pressure). For that we only need to know
-whether the use is also required later in the execution path. A variable is live
-out of a basic-block B if it is used in the merge set of B. We compute the
-live-out relation on-demand when profitability of hoist/sink candidates is to be
-evaluated. A simplified version of \emph{isLiveOutUsingMergeSet} is presented
-here:
+We compute merge sets, of the control flow graph, which remains the same
+throughout the \GCM{} transformation. For \GCM{} we only want to know if a use
+operand is a kill (to compute changes in register pressure). For that we only
+need to know whether the use is also required later in the execution path. A
+variable is live out of a basic-block B if it is used in the merge set of B. We
+compute the live-out relation on-demand when profitability of hoist/sink
+candidates is to be evaluated. A simplified version of
+\emph{isLiveOutUsingMergeSet} is presented here:
 
@@
 aliasing memory references. As long as all the operands are available, or can be
-made available by rematerialization,) the scalar computations can be hoisted.
+made available by rematerialization, the scalar computations can be hoisted.
 
@@
 
-Hoisting loads/stores across calls also require precise analysis of all the
+Hoisting loads/stores across calls also requires precise analysis of all the
 memory addresses accessed by the call. The current implementation being an
@@
 Barriers are based on the concept of pinned instructions \cite{click1995global}
-but extended to adapt to \LLVM{} IR. Since a basic-block in \LLVM{} IR is
-actually an extended basic-block because there might be calls in the middle of a
-basic-block which might not return. Essentially, any instruction that cannot
-guarantee progress is marked as a barrier. In the absence of context, as in our
-current implementation, some instructions which might be safe are still
-classified as barriers, e.g., volatile loads/stores, calls with missing
-attributes.
+but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
+extended basic-block because there might be non-returning calls in the middle of
+a basic-block. Essentially, any instruction that cannot guarantee progress is
+marked as a barrier. In the absence of context, as in our current
+implementation, some instructions which might be safe are still classified as
+barriers, e.g., volatile loads/stores, calls with missing attributes.
 
@@
 Hoisting upwards will decrease the live-range of its use, if it is a last use (a
-kill,) but increase the live-range of its definition. Conversely, sinking will
+kill) but increase the live-range of its definition. Conversely, sinking will
 decrease the live-range of the defined register but increase the live-range for
@@
 %The partition algorithm is described as follows:
-In the approach, described in Section~\ref{subsec:optimistic}, it is possible
+In the approach described in Section~\ref{subsec:optimistic}, it is possible
 that a common hoisting point for all the instructions does not satisfy legality
@@
 sorting the list of all the candidates in the increasing order of their depth
-first search discovery time stamp \cite{clrs} (DFSIn numbers,) we make sure that
+first search discovery time stamp \cite{clrs} (DFSIn numbers) we make sure that
 candidates closer in the list have their common dominator nearby. In
@@
 In our current implementation we keep as many candidates in one set as possible
-(greedy approach.) To start with, first two candidates are tried for
-hoistability (see Section~\ref{subsec:optimistic},) if that passes then the next
+(greedy approach.) To start with, the first two candidates are tried for
+hoistability (see Section~\ref{subsec:optimistic}) if that passes then the next
 instruction is tried (if available) and so on until a non-hoistable instruction
@@
 code size by almost 2\%. This is because once the code-size of a function
-decreases (due to \GCM{},) it becomes cheaper to inline. As we can see in
+decreases (due to \GCM{}) it becomes cheaper to inline. As we can see in
 Table~\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
@@
       \hline
-Number of spills  &base(-Ofast) &\GCM{}(-Ofast) &\%change\\\hline
+Benchmarks        &base(-Ofast) &\GCM{}(-Ofast) &GCM/base\\\hline
 400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
@@
 limits the \gcm{} to those candidates which do not increase the register
-pressure (except for loads Section~\ref{subsec:cost-models},) overall spills
+pressure (except for loads Section~\ref{subsec:cost-models}) overall spills
 still decreases.

### 2017-11-06 cgo/paper.tex
reorder the sections
chi

@@
   identical are given the same number;
+\item Rank of an expression \cite{rosen1988global}: The expressions are ordered
+  according to the amount of dependency they have on other expressions.  For
+  example in the expression $a = b+c$, $a$ will have higher rank than rank of
+  $b$, and $c$.
 \item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
@@
 was used in the liveness analysis during \GCM{}.
-
-\subsection{Liveness using DJ-graph}
-DJ-graph \cite{Sreedhar1996} is a data structure that augments the dominator
-tree with join-edges to keep track of data-flow in a program. We use DJ-graph
-and merge-sets to compute liveness of variables as illustrated in
-\cite{das2012}.  It is very efficient for computing liveness and does not
-require any bitvectors to be maintained for each basic-block. The underlying
-simplicity of liveness computation is due to \SSA{} form, where the the values
-only flow (from def to use) either through dominator edges or the join edges
-(where we insert a PHI.) The DJ-graph contains both these edges which allows for
-computation of merge-sets for each basic-block, i.e., a set of all basic-blocks
-where the values can flow from a particular basic-block. We have implemented the
-merge-set computation based on a DJ-graph as illustrated in \cite{das2005}. A
-simplified version of the merge-set that we implemented is presented here:
-
-\begin{verbatim}
-// Return true if the merge-set of source node of a
-// visited J-edge (incoming edge of lnode) is not
-// the subset of the merge-set of lnode.
-bool
-stillInconsistent(Node lnode, JEdges JE,
-                   Visited V, MergeSet MergeSet)
-  for e in (all incoming edges to lnode):
-    if e is in JE and e in V:
-      if MergeSet(Source Node of e)
-         is not a subset of MergeSet(lnode):
-        return true
-  return false
-\end{verbatim}
-
-Each basic block which has a join-edge (J-Edge) to its successor will have a
-merge-set that is a subset of the merge-set of its successor \cite{das2005}. The
-function \emph{stillInconsistent} returns true if this invariant is not
-satisfied which allows the function \emph{constructMergeSet} to repeat.
-
-\begin{verbatim}
-// Compute a merge-set breadth first order.
-bool constructMergeSet_1(BFSList B, JEdges JE,
-                         DomLevel DL)
-  repeat = false
-  V = empty list of visited edges
-  for n in B:
-    for e in (all incoming edges to n):
-      if e is in JE and e not in V:
-        V(e) = true
-        let tmp = Source Node of e
-        let tnode = Target Node of e
-        let lnode = NULL
-        while (DL(tmp) >= DL(tnode)):
-          MergeSet(tmp) = MergeSet(tmp)
-                          U MergeSet(tnode)
-                          U {tnode}
-          lnode = tmp
-          tmp = dom-parent(tmp)
-        repeat = still_inconsistent(lnode, JE, MergeSet)
-  return repeat
-\end{verbatim}
-
-\begin{verbatim}
-// Construct merge-set of each node in
-// Control-Flow-Graph G of a function.
-void constructMergeSet(CFG G)
-  B = Breadth First Order of G
-  JE = JEdges of G
-  DL = Path lengths (from root) of each node in G
-  do: // Call until a fixed point is reached.
-    Repeat = constructMergeSet_1(B, JE, DL)
-  while (Repeat)
-\end{verbatim}
-
-We compute merge sets, of the control flow graph, which remains the same
-throughout the \GCM{} transformation. For \GCM{} we only want to know if a use
-operand is a kill (to compute changes in register pressure). For that we only
-need to know whether the use is also required later in the execution path. A
-variable is live out of a basic-block B if it is used in the merge set of B. We
-compute the live-out relation on-demand when profitability of hoist/sink
-candidates is to be evaluated. A simplified version of
-\emph{isLiveOutUsingMergeSet} is presented here:
-
-\begin{verbatim}
-// Return true if variable A is liveout from N
-bool isLiveOutUsingMergeSet(Node N, Variable A)
-  if (A is defined in N)
-    if (A is used outside any basic-block other than N)
-      return true
-    return false
-
-  MergeSet(N) = null
-  // The merge-set of N is the union of
-  // merge-sets of its successors.
-  for W in successors(N)
-    MergeSet(N) = MergeSet(N) U MergeSet(W)
-
-  // Iterate over all the uses of A and see
-  // if any intersect with the merge-set of N.
-  for T in users(A)
-    B = basic_block(T)
-    while (B != null and B != def_bb(A))
-      if B in MergeSet(N) return true
-      B = dom-parent(B)
-  return false
-\end{verbatim}
-
-If a variable is not live-out of a basic-block, it still may be used later in
-the basic-block. To establish whether a use of a variable is a kill we iterate
-on all the subsequent instructions in a basic-block checking for uses, the
-\emph{isKill} function uses \emph{isLiveOutUsingMergeSet} to find out if an
-operand is killed in an instruction.
-
-\begin{verbatim}
-// Return true if op is the last use of the operand in I.
-bool isKill(Operand op, Instruction I)
-  if (isLiveOutUsingMergeSet(op))
-    return false
-  for each I1 after I in basic block of I
-    if (I1 uses op)
-      return false // Not a kill
-  return true
-\end{verbatim}
-
-The original algorithm presented in \cite{das2012} has mismatched types in terms
-of uses and nodes because each node can have many instructions and hence many
-uses. Also, while iterating on the dominator of each user of a variable we may
-reach to the beginning of a function, in that case the inner while loop needs to
-terminate. These two cases were missing from the original algorithm.
 
@@
 
+\section{Code hoisting of congruent computations}
+Examples to show CHI insertion.
+
+The algorithm factors out the reachability of values such that multiple queries
+to find reachability of values are fast. This is based on finding the ANTIC
+points in the CFG which do not change during hoisting. The ANTIC points are
+basically the dominance-frontiers in the inverse graph. So we introduce a data
+structure (CHI nodes) to keep track of values flowing out of a basic block. We
+only do this for values with multiple occurrences in the function as they are
+the potentially hoistable candidates. This approach allows us to hoist
+instructions to a basic block with more than two successors, as well as deal
+with infinite loops in a trivial way.
+
+Limitations: This pass does not hoist fully redundant expressions because they
+are already handled by GVN-PRE. It is advisable to run gvn-hoist before and
+after gvn-pre because gvn-pre creates opportunities for more instructions to be
+hoisted.
+
+\begin{figure}[h!]
+  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
+  \vspace*{-1.0cm}
+\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}}
+\label{fig:chi-intro}
+\end{figure}
+
+A CHI keeps information about values flowing out of a basic block.  It is
+similar to PHI but in the inverse graph, and used for outgoing values on each
+edge. For conciseness, it is computed only for instructions with multiple
+occurrences in the CFG because they are the only hoistable candidates.
+
+The global value number for both I1 and I2 is V, the CHI node will save the
+instruction as well as the edge where the value is flowing to.  Where CHIArg1,
+CHIArg2 are of type CHIArg.
+
+\begin{verbatim}
+struct CHIArg {
+  // Edge destination (shows the direction of flow), may not be where the I is.
+  BasicBlock *Dest;
+  // The instruction which uses the values flowing out of CHI.
+  Instruction *I;
+  // The global value number of I.
+  GVNType VN;
+};
+
+\end{verbatim}
+
+Essentially, CHIArg1 = {V, B, I1} and CHIArg2 = {V, C, I2}
+
+For sinking we use the similar approach as \ref{SSAPRE} to track values flowing
+into a basic block.
+
+\subsection{Profitability check (Cost models)}
+\label{subsec:cost-models}
+TODO: Motivate why this is important.
+After the legality checks have passed, we check if a \gcm{} is profitable.  That
+takes into account the impact \gcm{} would have on various parameters that might
+affect runtime performance, e.g., impact on live-range, gain in the code
+size. The current implementation makes effort to not regress in performance
+across popular benchmarks and at the same time to reduce code size as much as
+possible. Following cost models are implemented:
+
+\subsubsection{Reduce register pressure}
+\label{hoist:reg-pressure}
+Hoisting upwards will decrease the live-range of its use, if it is a last use (a
+kill) but increase the live-range of its definition. Conversely, sinking will
+decrease the live-range of the defined register but increase the live-range for
+killed operands. If the live-range after \GCM{} will decrease, it will be
+moved. Essentially, as long as there is one killed operand, code hoisting will
+either decrease or preserve the register pressure.  Similarly, code-sinking will
+either decrease or preserve the register pressure as long as there is one
+operand killed at most.  The following example explains how \GCM{} of can reduce
+the register pressure.
+
+\begin{verbatim}
+B0: b = m
+    c = n
+    if c is true then goto B1 else goto B2
+B1: a0 = b<kill> + c<kill>
+    goto B3
+B2: a1 = b<kill> + c<kill>
+    goto B3
+\end{verbatim}
+
+After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ is placed in
+$B0$. Since $b$ and $c$ are killed in both $a0$ and $a1$, hoisting the
+expressions will reduce the register pressure because two registers will be
+freed at the insertion point but only one register will be required to assign
+the definition of hoisted instruction $a0$.
+
+Ideally, it should be okay to hoist all the instructions and a later
+live-range-splitting \cite{cooper1998live} pass should make the right decision
+but the current live-range splitting pass of \LLVM{} is not making the optimal
+decision and we have found performance regressions while hoisting aggressively.
+
+There is a special case for hoisting load instructions when the hoist-point is
+the predecessor basic-block for all the congruent loads. Even if the register
+pressure would increase, we prefer to hoist loads in this case. The reason
+being, that would make the loaded value available by the time it is used.
+
+\subsubsection{Hoisting an expression across a call}
+\label{cost:across-calls}
+Even hoisting scalars across calls is tricky because it can increase the number
+of spills. During the frame lowering of calls, the argument registers (also
+known as the caller saved registers) are saved because they might be modified by
+the callee and after the call they are restored. Due to this the register
+pressure is high because the number of available registers are reduced by the
+number of caller saved registers. In that situation a computation that increases
+register pressure is not profitable to hoist.
+
 \subsection{Finding candidates to move}
@@
 
-\subsection{Legality check}
-\label{subsec:legality}
-Since the equality of candidates is purely based on the value numbers, we also
-need to establish if hoisting them to a common dominator or sinking them to a
-common post-dominator would be legal. Once a common dominator (post-dominator)
-is found, we check whether all the use-operands of the set of instructions are
-available at that position. It is possible to reinstantiate or remateralize the
-use-operands in some cases when the operands are not available and make it legal
-to move the instruction.
-
-Subsequently, it is checked that the side-effects of the computations (if any)
-do not intersect with any side-effects between the instructions to be
-moved and their hoisting/sinking point. It is also necessary to check if
-there are indirect branch targets, e.g., landing pad, case statements, goto
-labels, etc., along the path because it becomes difficult to prove safety checks
-in those cases. In our current implementation we discard candidates on those
-paths.
-
-\subsubsection{Legality of hoisting scalars}
-Scalars are the easiest to hoist because we do not have to analyze them for
-aliasing memory references. As long as all the operands are available, or can be
-made available by rematerialization, the scalar computations can be hoisted.
-
-\subsubsection{Legality of hoisting loads}
-The availability of operand to the load (an address) is checked at the hoisting
-point. If that is not available we try to rematerialize the address if
-possible.  Along the path, from current position of the load instruction
-backwards on the control flow to the hoisting point, we check whether there are
-writes to memory that may alias with the load, in which case the candidate is
-discarded. To iterate on the use-def chains for memory references the MemorySSA
-of \LLVM{} is used.
-
-\subsubsection{Legality of hoisting stores}
-For stores, we check the dependency requirements similar to the hoisting of
-loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
-instruction are available at the hoisting point, and there are no aliasing
-memory references along the hoisting path.
-
-\subsubsection{Legality of hoisting calls}
-Call instructions can be divided into three categories: those calls equivalent
-to purely scalar computations, calls reading from memory, and most of the time,
-without further information, calls have to be classified as writing to memory,
-that is the most restrictive form.  Each category of call instructions is
-handled as described for scalar, load, and store instructions.
-
-Hoisting loads/stores across calls also requires precise analysis of all the
-memory addresses accessed by the call. The current implementation being an
-intraprocedural pass, cannot hoist aggressively across calls. In the presence of
-pure calls, loads can be hoisted but stores can't. Also, if a call throws
-exceptions, or if it may not return, nothing can be hoisted across that call.
-
-\subsubsection{Legality of sinking expressions}
-\label{subsec:legality-sink}
-Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
-expressions are in a landing pad, sinking of those expressions
-may reduce code size. Sinking may also reduce the register pressure in some
-cases, e.g., when the use operands are not kills. For sinking, higher ranked
-expressions would be sunk first. And it would be illegal to sink higher ranked
-identical expressions if they are not fully anticipable in the common
-post-dominator. For example:
-
-\begin{verbatim}
-B0: i0 = load B
-B1: i1 = load A
-    c1 = i1 + 10
-    d1 = i0 + 20
-    goto B3
-B2: i2 = load A
-    c2 = i2 + 10
-    d2 = i0 + 20
-    goto B3
-
-B3: phi(c1, c2)
-    phi(d1, d2)
-\end{verbatim}
-
-In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
-(c1, c2) depend on i1 and i2 respectively which are also in their original basic
-blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
-of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
-(d1, d2) can safely be sunk because their operands are readily available at the
-sink point, i.e., B3. It should also be noted that, just because the expressions
-are identical and operands are available, it still requires a unique
-post-dominating PHI to use the exact same values to be legally sinkable.
-
-A general global scheduling algorithm also requires checks for undefined values
-when introducing a new computation along a path. Since only very busy
-expressions are moved in the current implementation, there is no need to check
-for undefinedness resulting due to movement of instructions. This simplifies the
-implementation.
-
 \subsubsection{Barriers}
@@
 \end{verbatim}
+
+\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
+\label{subsec:partition}
+%The partition algorithm is described as follows:
+In the approach described in Section~\ref{subsec:optimistic}, it is possible
+that a common hoisting point for all the instructions does not satisfy legality
+or profitability checks. In these cases, it is still possible to `partially'
+hoist a subset of instructions by splitting the set of candidates and finding a
+closer hoisting point for each subset.
+
+In order to hoist a subset of identical instructions, we partition the list of
+all candidates in a way to maximize the total number of hoist instructions.  By
+sorting the list of all the candidates in the increasing order of their depth
+first search discovery time stamp \cite{clrs} (DFSIn numbers) we make sure that
+candidates closer in the list have their common dominator nearby. In
+Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
+for some reason, they cannot be hoisted at B0, then we partition the set of
+hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
+be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
+B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
+could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
+still be hoisted - if legal and profitable - to B2.
+
+\begin{figure}
+\centering
+%rankdir=LR;
+\digraph[scale=0.35]{abc}{
+  B0 -> B1
+  B0 -> B2
+  B1 -> B3
+  B1 -> B4
+  B2 -> B5
+  B2 -> B6
+}
+\caption{\CFG{} to illustrate partitioning}
+\label{fig:dfsin}
+\end{figure}
+
+% TODO: Rewrite the following.
+In our current implementation we keep as many candidates in one set as possible
+(greedy approach.) To start with, the first two candidates are tried for
+hoistability (see Section~\ref{subsec:optimistic}) if that passes then the next
+instruction is tried (if available) and so on until a non-hoistable instruction
+is found. The partition starts from that point.  We split the list at a point
+where the legality checks fail to hoist subset of candidates which are legal to
+hoist and then start finding new hoisting point for the remaining ones.
+
+\subsection{Legality check}
+\label{subsec:legality}
+Since the equality of candidates is purely based on the value numbers, we also
+need to establish if hoisting them to a common dominator or sinking them to a
+common post-dominator would be legal. Once a common dominator (post-dominator)
+is found, we check whether all the use-operands of the set of instructions are
+available at that position. It is possible to reinstantiate or remateralize the
+use-operands in some cases when the operands are not available and make it legal
+to move the instruction.
+
+Subsequently, it is checked that the side-effects of the computations (if any)
+do not intersect with any side-effects between the instructions to be
+moved and their hoisting/sinking point. It is also necessary to check if
+there are indirect branch targets, e.g., landing pad, case statements, goto
+labels, etc., along the path because it becomes difficult to prove safety checks
+in those cases. In our current implementation we discard candidates on those
+paths.
+
+\subsubsection{Legality of hoisting scalars}
+Scalars are the easiest to hoist because we do not have to analyze them for
+aliasing memory references. As long as all the operands are available, or can be
+made available by rematerialization, the scalar computations can be hoisted.
+
+\subsubsection{Legality of hoisting loads}
+The availability of operand to the load (an address) is checked at the hoisting
+point. If that is not available we try to rematerialize the address if
+possible.  Along the path, from current position of the load instruction
+backwards on the control flow to the hoisting point, we check whether there are
+writes to memory that may alias with the load, in which case the candidate is
+discarded. To iterate on the use-def chains for memory references the MemorySSA
+of \LLVM{} is used.
+
+\subsubsection{Legality of hoisting stores}
+For stores, we check the dependency requirements similar to the hoisting of
+loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
+instruction are available at the hoisting point, and there are no aliasing
+memory references along the hoisting path.
+
+\subsubsection{Legality of hoisting calls}
+Call instructions can be divided into three categories: those calls equivalent
+to purely scalar computations, calls reading from memory, and most of the time,
+without further information, calls have to be classified as writing to memory,
+that is the most restrictive form.  Each category of call instructions is
+handled as described for scalar, load, and store instructions.
+
+Hoisting loads/stores across calls also requires precise analysis of all the
+memory addresses accessed by the call. The current implementation being an
+intraprocedural pass, cannot hoist aggressively across calls. In the presence of
+pure calls, loads can be hoisted but stores can't. Also, if a call throws
+exceptions, or if it may not return, nothing can be hoisted across that call.
+
+\subsubsection{Legality of sinking expressions}
+\label{subsec:legality-sink}
+Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
+expressions are in a landing pad, sinking of those expressions
+may reduce code size. Sinking may also reduce the register pressure in some
+cases, e.g., when the use operands are not kills. For sinking, higher ranked
+expressions would be sunk first. And it would be illegal to sink higher ranked
+identical expressions if they are not fully anticipable in the common
+post-dominator. For example:
+
+\begin{verbatim}
+B0: i0 = load B
+B1: i1 = load A
+    c1 = i1 + 10
+    d1 = i0 + 20
+    goto B3
+B2: i2 = load A
+    c2 = i2 + 10
+    d2 = i0 + 20
+    goto B3
+
+B3: phi(c1, c2)
+    phi(d1, d2)
+\end{verbatim}
+
+In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
+(c1, c2) depend on i1 and i2 respectively which are also in their original basic
+blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
+of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
+(d1, d2) can safely be sunk because their operands are readily available at the
+sink point, i.e., B3. It should also be noted that, just because the expressions
+are identical and operands are available, it still requires a unique
+post-dominating PHI to use the exact same values to be legally sinkable.
+
+A general global scheduling algorithm also requires checks for undefined values
+when introducing a new computation along a path. Since only very busy
+expressions are moved in the current implementation, there is no need to check
+for undefinedness resulting due to movement of instructions. This simplifies the
+implementation.
 
@@
 \end{verbatim}
-
-\subsection{Profitability check (Cost models)}
-\label{subsec:cost-models}
-After the legality checks have passed, we check if a \gcm{} is profitable.  That
-takes into account the impact \gcm{} would have on various parameters that might
-affect runtime performance, e.g., impact on live-range, gain in the code
-size. The current implementation makes effort to not regress in performance
-across popular benchmarks and at the same time to reduce code size as much as
-possible. Following cost models are implemented:
-
-\subsubsection{Reduce register pressure}
-\label{hoist:reg-pressure}
-Hoisting upwards will decrease the live-range of its use, if it is a last use (a
-kill) but increase the live-range of its definition. Conversely, sinking will
-decrease the live-range of the defined register but increase the live-range for
-killed operands. If the live-range after \GCM{} will decrease, it will be
-moved. Essentially, as long as there is one killed operand, code hoisting will
-either decrease or preserve the register pressure.  Similarly, code-sinking will
-either decrease or preserve the register pressure as long as there is one
-operand killed at most.  The following example explains how \GCM{} of can reduce
-the register pressure.
-
-\begin{verbatim}
-B0: b = m
-    c = n
-    if c is true then goto B1 else goto B2
-B1: a0 = b<kill> + c<kill>
-    goto B3
-B2: a1 = b<kill> + c<kill>
-    goto B3
-\end{verbatim}
-
-After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ is placed in
-$B0$. Since $b$ and $c$ are killed in both $a0$ and $a1$, hoisting the
-expressions will reduce the register pressure because two registers will be
-freed at the insertion point but only one register will be required to assign
-the definition of hoisted instruction $a0$.
-
-Ideally, it should be okay to hoist all the instructions and a later
-live-range-splitting \cite{cooper1998live} pass should make the right decision
-but the current live-range splitting pass of \LLVM{} is not making the optimal
-decision and we have found performance regressions while hoisting aggressively.
-
-There is a special case for hoisting load instructions when the hoist-point is
-the predecessor basic-block for all the congruent loads. Even if the register
-pressure would increase, we prefer to hoist loads in this case. The reason
-being, that would make the loaded value available by the time it is used.
-
-\subsubsection{Hoisting an expression across a call}
-\label{cost:across-calls}
-Even hoisting scalars across calls is tricky because it can increase the number
-of spills. During the frame lowering of calls, the argument registers (also
-known as the caller saved registers) are saved because they might be modified by
-the callee and after the call they are restored. Due to this the register
-pressure is high because the number of available registers are reduced by the
-number of caller saved registers. In that situation a computation that increases
-register pressure is not profitable to hoist.
-
-\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
-\label{subsec:partition}
-%The partition algorithm is described as follows:
-In the approach described in Section~\ref{subsec:optimistic}, it is possible
-that a common hoisting point for all the instructions does not satisfy legality
-or profitability checks. In these cases, it is still possible to `partially'
-hoist a subset of instructions by splitting the set of candidates and finding a
-closer hoisting point for each subset.
-
-In order to hoist a subset of identical instructions, we partition the list of
-all candidates in a way to maximize the total number of hoist instructions.  By
-sorting the list of all the candidates in the increasing order of their depth
-first search discovery time stamp \cite{clrs} (DFSIn numbers) we make sure that
-candidates closer in the list have their common dominator nearby. In
-Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
-for some reason, they cannot be hoisted at B0, then we partition the set of
-hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
-be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
-B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
-could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
-still be hoisted - if legal and profitable - to B2.
-
-\begin{figure}
-\centering
-%rankdir=LR;
-\digraph[scale=0.35]{abc}{
-  B0 -> B1
-  B0 -> B2
-  B1 -> B3
-  B1 -> B4
-  B2 -> B5
-  B2 -> B6
-}
-\caption{\CFG{} to illustrate partitioning}
-\label{fig:dfsin}
-\end{figure}
-
-% TODO: Rewrite the following.
-In our current implementation we keep as many candidates in one set as possible
-(greedy approach.) To start with, the first two candidates are tried for
-hoistability (see Section~\ref{subsec:optimistic}) if that passes then the next
-instruction is tried (if available) and so on until a non-hoistable instruction
-is found. The partition starts from that point.  We split the list at a point
-where the legality checks fail to hoist subset of candidates which are legal to
-hoist and then start finding new hoisting point for the remaining ones.
 

### 2017-11-06 cgo/paper.tex
ssapre

@@
   \vspace*{-1.0cm}
-\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}}
+\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{V, B,
+  I1\} and CHIArg2 = \{V, C, I2\}, V is the value number for both I1, and I2}
 \label{fig:chi-intro}
@@
 struct CHIArg {
-  // Edge destination (shows the direction of flow), may not be where the I is.
+  // Edge destination (shows the direction of flow),
+  // may not be where the I is.
   BasicBlock *Dest;
-  // The instruction which uses the values flowing out of CHI.
+  // The instruction CHI-arg refers to.
   Instruction *I;
@@
 
-Essentially, CHIArg1 = {V, B, I1} and CHIArg2 = {V, C, I2}
-
-For sinking we use the similar approach as \ref{SSAPRE} to track values flowing
+Essentially, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}
+
+For sinking we use the similar approach as \cite{ssapre} to track values flowing
 into a basic block.

### 2017-11-06 cgo/paper.tex
ssapre

@@
 \usepackage{amsmath}
-\usepackage{hyperref}
 \usepackage{graphicx}
@@
 
-\section{Code hoisting of congruent computations}
-Examples to show CHI insertion.
-
+\section{A fastt Code motion of congruent computations}
 The algorithm factors out the reachability of values such that multiple queries
-to find reachability of values are fast. This is based on finding the ANTIC
-points in the CFG which do not change during hoisting. The ANTIC points are
-basically the dominance-frontiers in the inverse graph. So we introduce a data
-structure (CHI nodes) to keep track of values flowing out of a basic block. We
-only do this for values with multiple occurrences in the function as they are
-the potentially hoistable candidates. This approach allows us to hoist
-instructions to a basic block with more than two successors, as well as deal
-with infinite loops in a trivial way.
+to find reachability of values are fast. This is based on finding the dominance
+frontiers and the post-dominance frontiers of the CFG. These points, being
+structural properties of a graph, do not change during code-motion.
+
+A post-dominance frontier is the basic block where anticipability (ANTIC) of an
+instruction (I) is checked because it the point on which the basic block
+containing I is control dependent.
+
+So we introduce a data structure (CHI nodes) to keep track of values flowing out
+of a basic block. We only do this for values with multiple occurrences in the
+function as they are the potentially hoistable candidates. This approach allows
+us to hoist instructions to a basic block with more than two successors, as well
+as deal with infinite loops in a trivial way.
 
@@
 
-\begin{figure}[h!]
+A CHI keeps information about values flowing out of a basic block.  It is
+similar to PHI but in the inverse graph, and used for outgoing values on each
+edge. For conciseness, it is computed only for instructions with multiple
+occurrences in the CFG because they are the only hoistable candidates.
+
+\begin{figure}
   \includegraphics[width=0.4\textwidth]{chi-example.pdf}
@@
 \end{figure}
-
-A CHI keeps information about values flowing out of a basic block.  It is
-similar to PHI but in the inverse graph, and used for outgoing values on each
-edge. For conciseness, it is computed only for instructions with multiple
-occurrences in the CFG because they are the only hoistable candidates.
 
@@
 };
-
 \end{verbatim}
 
-Essentially, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}
-
-For sinking we use the similar approach as \cite{ssapre} to track values flowing
-into a basic block.
+Essentially, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}. For sinking we
+use the similar approach as \cite{ssapre} to track values flowing into a basic
+block.
+
+When SSAPRE inserts PHI (expression PHIs), it does so at places where
+(potentially) multiple expressions may merge. If a PHI has a missing entry, that
+means the expression is partially available.  The concept may work for hoisting
+expression to help factor out anticipability by working on inverted graph. If we
+insert a CHI, to a basic block with multiple successors, as instructions are
+hoisted upwards to the post-dominance frontier.  And we start walking the CFG to
+figure out if any CHI has a missing entry, that means the expression is not
+anticipable in the basic block having that outgoing PHI.
+
+\begin{itemize}
+\item Sort VNs according to their rank, and start with lowest ranked VN
+\item Take a VN and for each instruction with same VN
+\item  - Find the dominance frontier in the inverse graph (PDF)
+\item  - Insert the chi-node at PDF
+\item Remove the chi-nodes with missing entries
+\item Remove values from CHI-nodes which do not truly flow out, e.g.,
+\item  modified along the path.
+\item Collect the remaining values that are still anticipable
+\end{itemize}
 

### 2017-11-06 cgo/paper.tex
undef

@@
 are done using integer linear programming (ILP) which results in prohibitively
-high compile time. To the best of our knowledge we have not found any reference
-which explored \gcm{} of congruent instructions in as much detail us. A part of
-our implementation (aggressive code hoisting) is already merged
-in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
-redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
-contributions of this paper are:
+high compile time which is not suitable for industrial compilers. To the best of
+our knowledge we have not found any reference which explored \gcm{} of congruent
+instructions. A part of our implementation (aggressive code hoisting) is already
+merged in \LLVM{} trunk. The main contributions of this paper are:
 \begin{itemize}[leftmargin=*,topsep=0pt]
-\item a new optimistic algorithm to move congruent instructions out of branches;
+\item an optimistic fast algorithm to schedule congruent instructions;
 \item a cost model to reduce interference and hence, reduce spills;
-\item a technique to maximize hoisting in an optimistic approach by partitioning
-  the list of potential candidates sorted by their DFS visit number;
-\item experimental evaluation of our implementation in \LLVM{} which combines
-  \SSA{} based liveness analysis, and ranking of expressions to move very busy
-  expressions in order to reduce register spills and improve performance in some
-  cases.
+\item experimental evaluation of our implementation in \LLVM{}.
 \end{itemize}
@@
 computed by comparing the operands which is neither general nor scalable.
+
+\citet{click1995global} describes aggressive \gcm{} in sparse representation. It
+first schedules all the instructions as early as possible. This results in very
+long live ranges which is mitigated by again scheduling all the instructions as
+late as possible. They report a speedup of as high as $23\%$ but there is no
+cost model to limit amount of scheduling and could often introduce redundant
+computations. It is also not mentioned how safety is ensured while hoisting an
+instruction when it is not anticipable. The legality checks are only based on
+barriers (Section~\ref{subsec:barriers}) and availability of used operands. Even
+introducing a simple integer addition can also result in undefined behavior
+\cite{undef}.  Except from this paper we found no reference of global scheduling
+in sparse representation.
 
@@
 \GVN{}-\PRE{}, code-hoisting will hoist them.
-
-\citet{click1995global} describes aggressive \gcm{} to first
-schedule all the instructions as early as possible. This results in very long
-live ranges which is mitigated by again scheduling all the instructions as late
-as possible. They report a speedup of as high as $23\%$.
 
@@
 
-\section{A fastt Code motion of congruent computations}
+\section{A fast algorithm for code motion congruent computations}
 The algorithm factors out the reachability of values such that multiple queries
@@
 \subsubsection{Barriers}
+\label{subsec:barriers}
 Barriers are based on the concept of pinned instructions \cite{click1995global}

### 2017-11-06 cgo/paper.tex
corrections

@@
 optimistic algorithm in the sense that it considers all congruent instructions
-in a function as potential candidates. We make an extra effort to hoist
-candidates by partitioning the potential candidates in a way to enable partial
-hoisting in case a common hoisting point for all the candidates cannot be
-found. We also formalize why register pressure reduces as a result of \gcm{},
-and how \gcm{} increases instruction level parallelism thereby enabling more
-\ooo{} execution on modern architectures.
+in a function as potential candidates. We also formalize why register pressure
+reduces as a result of \gcm{}, and how \gcm{} increases instruction level
+parallelism thereby enabling more \ooo{} execution on modern architectures.
 \end{abstract}
@@
 to find reachability of values are fast. This is based on finding the dominance
-frontiers and the post-dominance frontiers of the CFG. These points, being
-structural properties of a graph, do not change during code-motion.
-
-A post-dominance frontier is the basic block where anticipability (ANTIC) of an
+frontiers and the post-dominance frontiers (PDF) of the CFG. These points, being
+structural properties of a graph, do not change during code-motion. A
+post-dominance frontier is the basic block where anticipability (ANTIC) of an
 instruction (I) is checked because it the point on which the basic block
@@
 So we introduce a data structure (CHI nodes) to keep track of values flowing out
-of a basic block. We only do this for values with multiple occurrences in the
-function as they are the potentially hoistable candidates. This approach allows
-us to hoist instructions to a basic block with more than two successors, as well
-as deal with infinite loops in a trivial way.
-
-Limitations: This pass does not hoist fully redundant expressions because they
-are already handled by GVN-PRE. It is advisable to run gvn-hoist before and
-after gvn-pre because gvn-pre creates opportunities for more instructions to be
-hoisted.
-
-A CHI keeps information about values flowing out of a basic block.  It is
-similar to PHI but in the inverse graph, and used for outgoing values on each
-edge. For conciseness, it is computed only for instructions with multiple
-occurrences in the CFG because they are the only hoistable candidates.
-
-\begin{figure}
-  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
-  \vspace*{-1.0cm}
-\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{V, B,
-  I1\} and CHIArg2 = \{V, C, I2\}, V is the value number for both I1, and I2}
-\label{fig:chi-intro}
-\end{figure}
-
-The global value number for both I1 and I2 is V, the CHI node will save the
-instruction as well as the edge where the value is flowing to.  Where CHIArg1,
-CHIArg2 are of type CHIArg.
+of a basic block with more than once successors. We only do this for values with
+multiple occurrences in the function as they are the potentially hoistable
+candidates. This approach allows us to hoist instructions to a basic block with
+more than two successors. The CHI node in a basic block $B$ has CHIArgs with
+three entries: a) the instruction (I) CHIArg refers to, b) the basic block $Dest$
+which is post-dominated by the basic block containing I, and c) the GVN of I.
 
@@
 
-Essentially, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}. For sinking we
-use the similar approach as \cite{ssapre} to track values flowing into a basic
-block.
+Limitations: This pass does not hoist fully redundant expressions because they
+are already handled by GVN-PRE. It is advisable to run gvn-hoist before and
+after gvn-pre because gvn-pre creates opportunities for more instructions to be
+hoisted.
+
+A CHI keeps information about values flowing out of a basic block.  It is
+similar to PHI but in the inverse graph, and used for outgoing values on each
+edge. Although it may be suitable for other analyses/optimizations, for \GCM{},
+it is computed only for instructions with multiple occurrences in the CFG. An
+illustration of CHI is shown in Figure~\ref{fig:chi-intro}. The \GVN{} for both
+I1 and I2 is V, the CHI node will save the instruction as well as the edge where
+the value is flowing to.  Where CHIArg1, CHIArg2 are of type CHIArg. The CHI for
+both I1 and I2 is inserted in their PDF. For sinking we use the similar approach
+as \cite{ssapre} to track values flowing into a basic block.
+
+\begin{figure}
+  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
+  \vspace*{-1.0cm}
+\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{B, I1,
+  V\} and CHIArg2 = \{C, I2, V\}, V is the value number for both I1, and I2}
+\label{fig:chi-intro}
+\end{figure}
 
@@
 anticipable in the basic block having that outgoing PHI.
-
-\begin{itemize}
-\item Sort VNs according to their rank, and start with lowest ranked VN
-\item Take a VN and for each instruction with same VN
-\item  - Find the dominance frontier in the inverse graph (PDF)
-\item  - Insert the chi-node at PDF
-\item Remove the chi-nodes with missing entries
-\item Remove values from CHI-nodes which do not truly flow out, e.g.,
-\item  modified along the path.
-\item Collect the remaining values that are still anticipable
-\end{itemize}
 
@@
     c = n
-    if c is true then goto B1 else goto B2
+    if p is true then goto B1 else goto B2
 B1: a0 = b<kill> + c<kill>
@@
     goto B3
+B3: ...
 \end{verbatim}
@@
 pressure would increase, we prefer to hoist loads in this case. The reason
-being, that would make the loaded value available by the time it is used.
+being, that would make the loaded value available by the time it is used, and
+livess would not increase by much because instruction is hoisted to immediate
+predecessor.
 
@@
 \cite{briggs1997}. This is performed by a linear scan of all instructions of the
-program and classifying them by their value numbers. We could compute available
-and anticipable sets as computed by \GCC{'}s code-hoisting but that would be a lot
-of data structures to maintain at each basic-block level. For \GCC{} it makes sense
-because their code-hoisting is integrated with \GVN{}-\PRE{} which already has those
-data structures available.
+program and classifying them by their value numbers.
 
@@
 we need to rerun the \GVN{} analysis in order to discover new candidates now
-available after having hoisted load instructions.  This limitation should be
-addressed in a new implementation of the \GVN{} based on \MemorySSA{}, that
-would better account for equivalent loads and their dependent instructions.
+available after having hoisted load instructions.
 
@@
 precomputed.
-
-Once the instructions have been classified into congruence classes, we compute
-for each group of congruent instructions, a point in the program that is both
-legal and profitable for the instructions to be moved to.
 
@@
 
-\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
-\label{subsec:partition}
-%The partition algorithm is described as follows:
-In the approach described in Section~\ref{subsec:optimistic}, it is possible
-that a common hoisting point for all the instructions does not satisfy legality
-or profitability checks. In these cases, it is still possible to `partially'
-hoist a subset of instructions by splitting the set of candidates and finding a
-closer hoisting point for each subset.
-
-In order to hoist a subset of identical instructions, we partition the list of
-all candidates in a way to maximize the total number of hoist instructions.  By
-sorting the list of all the candidates in the increasing order of their depth
-first search discovery time stamp \cite{clrs} (DFSIn numbers) we make sure that
-candidates closer in the list have their common dominator nearby. In
-Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
-for some reason, they cannot be hoisted at B0, then we partition the set of
-hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
-be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
-B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
-could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
-still be hoisted - if legal and profitable - to B2.
-
-\begin{figure}
-\centering
-%rankdir=LR;
-\digraph[scale=0.35]{abc}{
-  B0 -> B1
-  B0 -> B2
-  B1 -> B3
-  B1 -> B4
-  B2 -> B5
-  B2 -> B6
-}
-\caption{\CFG{} to illustrate partitioning}
-\label{fig:dfsin}
-\end{figure}
-
-% TODO: Rewrite the following.
-In our current implementation we keep as many candidates in one set as possible
-(greedy approach.) To start with, the first two candidates are tried for
-hoistability (see Section~\ref{subsec:optimistic}) if that passes then the next
-instruction is tried (if available) and so on until a non-hoistable instruction
-is found. The partition starts from that point.  We split the list at a point
-where the legality checks fail to hoist subset of candidates which are legal to
-hoist and then start finding new hoisting point for the remaining ones.
-
 \subsection{Legality check}
@@
 Since the equality of candidates is purely based on the value numbers, we also
-need to establish if hoisting them to a common dominator or sinking them to a
-common post-dominator would be legal. Once a common dominator (post-dominator)
-is found, we check whether all the use-operands of the set of instructions are
-available at that position. It is possible to reinstantiate or remateralize the
-use-operands in some cases when the operands are not available and make it legal
-to move the instruction.
+need to establish if hoisting them or sinking them would be legal. Once a common
+dominator (post-dominator) is found, we check whether all the use-operands of
+the set of instructions are available at that position. It is possible to
+reinstantiate or remateralize the use-operands in some cases when the operands
+are not available and make it legal to move the instruction.
 
 Subsequently, it is checked that the side-effects of the computations (if any)
-do not intersect with any side-effects between the instructions to be
-moved and their hoisting/sinking point. It is also necessary to check if
-there are indirect branch targets, e.g., landing pad, case statements, goto
-labels, etc., along the path because it becomes difficult to prove safety checks
-in those cases. In our current implementation we discard candidates on those
-paths.
+do not intersect (alias) with any side-effects between the instructions to be
+moved and their hoisting/sinking point. It is also necessary to check if there
+are indirect branch targets, e.g., landing pad, case statements, goto labels,
+etc., along the path because it becomes difficult to prove safety checks in
+those cases. In our current implementation we discard candidates on those paths.
 
@@
 
-\subsubsection{Legality of hoisting loads}
+\subsubsection{Legality of hoisting memory references}
 The availability of operand to the load (an address) is checked at the hoisting
-point. If that is not available we try to rematerialize the address if
-possible.  Along the path, from current position of the load instruction
-backwards on the control flow to the hoisting point, we check whether there are
-writes to memory that may alias with the load, in which case the candidate is
-discarded. To iterate on the use-def chains for memory references the MemorySSA
-of \LLVM{} is used.
-
-\subsubsection{Legality of hoisting stores}
+point. If that is not available we try to rematerialize the address if possible.
+Along the path, from current position of the load instruction backwards on the
+control flow to the hoisting point, we check whether there are writes to memory
+that may alias with the load, in which case the candidate is discarded. Memory
+SSA allows efficient way of iterating over the use-def chains of memory
+references on a factored graph.
+
 For stores, we check the dependency requirements similar to the hoisting of
-loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
-instruction are available at the hoisting point, and there are no aliasing
-memory references along the hoisting path.
+loads. We check that the operands of the store instruction are available at the
+hoisting point, and there are no aliasing memory references along the hoisting
+path.
 
@@
 Call instructions can be divided into three categories: those calls equivalent
-to purely scalar computations, calls reading from memory, and most of the time,
-without further information, calls have to be classified as writing to memory,
-that is the most restrictive form.  Each category of call instructions is
-handled as described for scalar, load, and store instructions.
+to purely scalar computations, calls reading from memory, and calls writing to
+memory which is the most restrictive form.  Each category of call instructions
+is handled as described for scalar, load, and store instructions.
 
@@
 memory addresses accessed by the call. The current implementation being an
-intraprocedural pass, cannot hoist aggressively across calls. In the presence of
-pure calls, loads can be hoisted but stores can't. Also, if a call throws
-exceptions, or if it may not return, nothing can be hoisted across that call.
+intraprocedural pass, cannot schedule aggressively across calls. In the presence
+of pure calls, loads can be hoisted but stores can't. Also, if a call throws
+exceptions, or if it may not return, nothing can be scheduled across that call.
 
@@
 introduce a new computation along any path of execution without checking for
-undefined behaviors. The algorithm \emph{computeDownwardSafety} computes if
-hoisting a set of instruction to a common hoisting point is downward-safe
-\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
-downward-safety ensures that no undefined behaviors are introduced as a result
-of code-motion.
-\newline
-\begin{verbatim}
-// Return true if a hoistable instruction is downward
-// safe at hoist-point
-bool computeDownwardSafety(Node HP)
-  Worklist = basic-blocks of hoistable instructions
-  DT = The dominator tree
-  HP = the hoist-point
-  for each basic-block B in the DT starting at HP:
-    if Worklist is empty
-       return false // Path exists! not downward safe
-    if B is in Worklist:
-      remove B from worklist
-      // Available at B => downward safe from B
-      remove subtree with root at B
-    if B is a leaf basic-block
-      // Path exists! not downward safe
-      return false
-    if B dominates hoist-point
-      return false  // Back edge
-  return true
-\end{verbatim}
+undefined behaviors \cite{undef}. Since \GCM{} is only dealing with VBEs, as
+hoisting happens for CHIs with complete entries, the downward safety is ensured
+by construction of CHIs.
 
@@
   Compute DJ-graph of function
-  constructMergeSet(CGF)
+  constructMergeSet(CFG)
   computeBarriers()
@@
 
-For hoisting, we find a common dominator dominating all a set of congruent
-instructions and perform legality checks, as described in
-Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
-instructions to one common dominator, due to legality constraints, e.g.,
-intersecting side-effects, or profitability constraints, e.g., increasing
-register pressure; in those cases, this algorithm would partition
-(Section~\ref{subsec:partition}) the list of identical instructions into subsets
-which can be partially hoisted to their respective common dominators.
-
 \begin{verbatim}
-// Find hoistable candidates and do hoisting
 int doCodeHoisting()
-  For each value number VN with 2 or more instructions
-    sort the instructions according to DFSIn numbers
-    if first two I1, I2 are hoistable
-        && downward-safe at hoist-point
-        && profitable to be hoisted at hoist-point:
-      continue the check for subsequent
-        instructions with same VN
-    else:
-      proceed to check if I2 and I3 (if available)
-      can be hoisted
-
-  For each set of hoistable instructions:
-    move the first one (I1) to the hoist-point
-    update their users others to refer to I1
-    remove all other instructions (I2, ..., Ik)
-    if I1 has memory references:
-      update MemorySSA of I1 and others
-    update statistics
-  return number of hoisted instructions
+  Sort VNs in ascending order
+
+  for each instruction with the same VN
+      Find its post-dominance-frontier (PDF)
+      Insert the CHI-node at PDF
+
+  for each CHI-nodes in the CFG:
+    Remove the ones with missing entries
+    Remove CHI-arg if it fails legality checks
+
+  for each CHI-nodes in the CFG:
+    if CHI  does not have missing entries:
+      Perform code hoisting
+
+  return number of sinked instructions
 \end{verbatim}

### 2017-11-06 cgo/paper.tex
ilp

@@
   count;
-\item it exposes more instruction level parallelism to the later compiler
+\item it exposes more instruction level parallelism (ILP) to the later compiler
   passes. By hoisting identical computations to be executed earlier, instruction
   schedulers can move heavy computations earlier in order to avoid pipeline
-  bubbles;
+  bubbles. Although it decreases the ILP of basic blocks from where the
+  instructions were removed, those basic blocks execute for lesser number of
+  cycles than the destination basic block.
 \item it reduces branch misprediction penalty on \ooo{} processors with

### 2017-11-06 cgo/paper.tex
sinking

@@
 \item an optimistic fast algorithm to schedule congruent instructions;
+\item introducing an analogue of PHI nodes on the inverse CFG which allows
+  faster tracking of anticipable instructions;
 \item a cost model to reduce interference and hence, reduce spills;
@@
 introduce a new computation along any path of execution without checking for
-undefined behaviors \cite{undef}. Since \GCM{} is only dealing with VBEs, as
+undefined behavior \cite{undef}. Since \GCM{} is only dealing with VBEs, as
 hoisting happens for CHIs with complete entries, the downward safety is ensured
@@
 
+Code hoisting opens new opportunities for other hoistable candidates which were
+of higher rank (depended on candidates which got hoisted) which are encountered
+subsequently as the algorithm visits instructions in the increasing order of
+their ranks.
+
 %\newpage
@@
 
+For sinking we reuse the PHI nodes instead of inserting another expression nodes
+like in \cite{ssapre} because PHI nodes ensure availability of definitions, and
+hence downward safety, as long as all instructions corresponding to a PHI are
+congruent. Also, the PHI nodes represent a complete set of instructions which
+could be potentially sunk for code-size optimizations because PHI nodes
+represent all values flowing out from a basic block into its dominance-frontier.
+
 \begin{verbatim}
@@
   // Find sinkable instructions
-  For each value number VN with 2 or more instructions:
+  For each value number VN > 1 instructions:
     if there are two instructions with same immediate
@@
 
-Code hoisting opens new opportunities for other hoistable candidates which were
-of higher rank (depended on candidates which got hoisted). Ideally we could
-iterate on lower ranking expressions first and then proceed to higher ranking
-expressions in the same iteration but \LLVM{'}s \GVN{} infrastructure does not
-compute equivalence classes in an effective way. We found it simpler to just
-recompute the value numbers and start finding hoistable candidates again.
-
 \subsection{Time complexity of algorithm}
 The complexity of code hoisting is linear in number of instructions that are
-candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
-The analyses computed for this pass are Global Value Numbering, Computation of
-DJ-graph and merge-sets, Marking Barriers, all are linear in number of
-instructions in a function. Liveness analysis is expensive and only performed
-on-demand for hoistable candidates so it does not affect the overall compilation
-time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
-are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
-Barriers for each iteration of the \gcm{}, we have not seen significant
-increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
-provided appropriate compiler
+candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.  The
+analyses computed for this pass like Global Value Numbering, Marking Barriers,
+are both linear in number of instructions in a function. Computation of DJ-graph
+and merge-sets is $O(E)$, E being number of edges in the CFG. Liveness analysis
+is not very expensive \cite{das2012} but still only performed on-demand for
+hoistable candidates. Other analyses like Alias Analysis, \MemorySSA{} and
+Dominator Tree are already available in the \LLVM{} pass pipeline.
+
+Although we recompute \GVN{}, and Barriers for each iteration of the \gcm{},
+there is still gain in the compilation times (see
+Section~\ref{tab:compile-time}). We have also provided appropriate compiler
 flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
@@
 they are legal.
-
-The analysis is followed by a simple code generation that adds the identified
-instruction in the destination point and removes all instructions rendered
-redundant as a result of \gcm{}.
 

### 2017-11-08 cgo/paper.tex
chi args

@@
 instructions across the program.  Not only \GCM{} saves code size, it exposes
-redundancies in some cases, it exposes more instruction level parallelism in the
+redundancies as well, it exposes more instruction level parallelism in the
 basic-block to which instructions are moved, and it enables other passes like
-loop invariant motion to remove more redundancies. The cost model to drive the
-code motion is based on liveness analysis on \SSA{} representation such that the
+loop invariant motion to remove redundancies. The cost model to drive the code
+motion is based on liveness analysis on \SSA{} representation such that the
 (virtual) register pressure does not increase resulting in 2\% fewer spills on
@@
 in a function as potential candidates. We also formalize why register pressure
-reduces as a result of \gcm{}, and how \gcm{} increases instruction level
-parallelism thereby enabling more \ooo{} execution on modern architectures.
+reduces, and how \GCM{} increases instruction level parallelism thereby enabling
+more \ooo{} execution on modern architectures.
 \end{abstract}
@@
 
-This paper describes \gcm{} (\GCM{}) of congruent instructions
-\cite{briggs1997}, a technique that uses the information computed for \PRE{} to
-detect identical computations but has a transformation phase whose goal differs
-from \PRE{}: it removes identical computations from different branches of
-execution.  These identical computations in different branches of execution are
-not redundant computations at run-time and the number of run-time computations
-is not reduced. It is not a redundancy elimination pass, and thus it has
-different cost function and heuristics than \PRE{} or CSE. It is similar to
-global code scheduling \cite{dragonbook,click1995global} in the sense that it
+This paper describes a fast algorithm for \gcm{} (\GCM{}) of congruent
+instructions \cite{briggs1997}, a technique that uses the information computed
+for \PRE{} to detect identical computations but has a transformation phase whose
+goal differs from \PRE{}: it removes identical computations from different
+branches of execution.  These identical computations in different branches of
+execution are not redundant computations at run-time and the number of run-time
+computations is not reduced. It is not a redundancy elimination pass, and thus
+it has different cost function and heuristics than \PRE{} or CSE. It is similar
+to global code scheduling \cite{dragonbook,click1995global} in the sense that it
 will only move computations. Code hoisting, for example, can reduce the critical
@@
 of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
-$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
-seen the performance of the original benchmark (on an \ooo{} processor) improve
-by $15\%$.  In fact, there are several advantages of \GCM{}:
+$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}.  Although
+it decreases the instruction level parallelism (ILP) of basic blocks from where
+the instructions were removed, those basic blocks cannot execute for more cycles
+than the destination basic block. We have seen the performance of a proprietary
+benchmark (on an \ooo{} processor) improve by $15\%$.  In fact, there are
+several advantages of \GCM{}:
 \begin{itemize}[leftmargin=*,topsep=0pt]
 \item it helps reduce the code size of the program by replacing multiple
-  instructions with one although the final code size may be higher because of
-  improved inlining heuristics; functions become cheaper to inline when their
-  code size decreases because the inliner heuristics depend on instruction
-  count;
-\item it exposes more instruction level parallelism (ILP) to the later compiler
-  passes. By hoisting identical computations to be executed earlier, instruction
-  schedulers can move heavy computations earlier in order to avoid pipeline
-  bubbles. Although it decreases the ILP of basic blocks from where the
-  instructions were removed, those basic blocks execute for lesser number of
-  cycles than the destination basic block.
+  instructions with one thereby making the function cheaper to inline because
+  inliner heuristics depend on instruction count;
+\item it exposes more ILP to the later compiler passes. By hoisting identical
+  computations to be executed earlier, instruction schedulers can move heavy
+  computations earlier in order to avoid pipeline bubbles.
 \item it reduces branch misprediction penalty on \ooo{} processors with
@@
   branches, it can effectively reduce the amount of code to be speculatively
-  executed and hence reduce the critical path;
+  executed;
 \item it reduces interference or register pressure with an appropriate
@@
   architectures which execute all code in branches based on masking or
-  predication;
+  predication \cite{simd};
 \item it may improve loop vectorization by reducing a loop with control flow to
@@
 
-There has been a lot of work both in industry and academia to hoist and sink
-code out of branches, and in general global scheduling
-\cite{click1995global}. Some relate code-hoisting to code-size optimization
-\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
-scheduling to improve performance. Most of the recent work on global scheduling
-are done using integer linear programming (ILP) which results in prohibitively
-high compile time which is not suitable for industrial compilers. To the best of
-our knowledge we have not found any reference which explored \gcm{} of congruent
-instructions. A part of our implementation (aggressive code hoisting) is already
-merged in \LLVM{} trunk. The main contributions of this paper are:
+There has been a lot of work both in industry and academia related to code
+hoisting/sinking, and in general global scheduling \cite{click1995global}. Some
+relate code-hoisting to code-size optimization \cite{rosen1988global} and many
+\cite{barany2013, shobaki2013} use global scheduling to improve
+performance. Most of the recent work on global scheduling are done using integer
+linear programming (ILP) which results in prohibitively high compile time which
+is not suitable for industrial compilers. To the best of our knowledge we have
+not found any reference which explored \gcm{} with a cost-model to reduce
+register-pressure. A part of our implementation (aggressive code hoisting) is
+already merged in \LLVM{} trunk. The main contributions of this paper are:
 \begin{itemize}[leftmargin=*,topsep=0pt]
 \item an optimistic fast algorithm to schedule congruent instructions;
-\item introducing an analogue of PHI nodes on the inverse CFG which allows
+\item introducing an analogue of $\phi$ nodes on the inverse CFG which allows
   faster tracking of anticipable instructions;
@@
 A CHI keeps information about values flowing out of a basic block.  It is
-similar to PHI but in the inverse graph, and used for outgoing values on each
-edge. Although it may be suitable for other analyses/optimizations, for \GCM{},
-it is computed only for instructions with multiple occurrences in the CFG. An
-illustration of CHI is shown in Figure~\ref{fig:chi-intro}. The \GVN{} for both
-I1 and I2 is V, the CHI node will save the instruction as well as the edge where
-the value is flowing to.  Where CHIArg1, CHIArg2 are of type CHIArg. The CHI for
-both I1 and I2 is inserted in their PDF. For sinking we use the similar approach
-as \cite{ssapre} to track values flowing into a basic block.
+similar to $\phi$ \cite{ssa} but in the inverse graph, and used for out-flowing
+values on each edge. Although it may be suitable for other
+analyses/optimizations, for \GCM{}, it is computed only for instructions with
+multiple occurrences in the CFG. An illustration of CHI is shown in
+Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the CHI node
+will save the instruction as well as the edge where the value is flowing to.
+Where CHIArg1, CHIArg2 are of type CHIArg. The CHI for both I1 and I2 is
+inserted in their PDF (A). For sinking we use the similar approach as
+\cite{ssapre} to track values flowing into a basic block.
 
@@
 
-When SSAPRE inserts PHI (expression PHIs), it does so at places where
-(potentially) multiple expressions may merge. If a PHI has a missing entry, that
+When SSAPRE inserts $\phi$ (expression PHIs), it does so at places where
+(potentially) multiple expressions may merge. If a $\phi$ has a missing entry, that
 means the expression is partially available.  The concept may work for hoisting
@@
 figure out if any CHI has a missing entry, that means the expression is not
-anticipable in the basic block having that outgoing PHI.
+anticipable in the basic block having CHI.
 
@@
 
-For sinking we reuse the PHI nodes instead of inserting another expression nodes
-like in \cite{ssapre} because PHI nodes ensure availability of definitions, and
-hence downward safety, as long as all instructions corresponding to a PHI are
-congruent. Also, the PHI nodes represent a complete set of instructions which
-could be potentially sunk for code-size optimizations because PHI nodes
+For sinking we reuse the $\phi$ nodes instead of inserting another expression nodes
+like in \cite{ssapre} because $\phi$ nodes ensure availability of definitions, and
+hence downward safety, as long as all instructions corresponding to a $\phi$ are
+congruent. Also, the $\phi$ nodes represent a complete set of instructions which
+could be potentially sunk for code-size optimizations because $\phi$ nodes
 represent all values flowing out from a basic block into its dominance-frontier.
@@
         successor(S) as post-dominator:
-      if both are only used in the same PHI-node of S
+      if both are only used in the same $\phi$-node of S
           && dependencies of both are available at S:

### 2017-11-08 cgo/paper.tex
chi details

@@
 will save the instruction as well as the edge where the value is flowing to.
-Where CHIArg1, CHIArg2 are of type CHIArg. The CHI for both I1 and I2 is
+Where $Arg1$, $Arg2$ are of type CHIArg. The CHI for both I1 and I2 is
 inserted in their PDF (A). For sinking we use the similar approach as
@@
   \vspace*{-1.0cm}
-\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{B, I1,
-  V\} and CHIArg2 = \{C, I2, V\}, V is the value number for both I1, and I2}
+\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1,
+  V\} and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2}
 \label{fig:chi-intro}
@@
 
-When SSAPRE inserts $\phi$ (expression PHIs), it does so at places where
-(potentially) multiple expressions may merge. If a $\phi$ has a missing entry, that
-means the expression is partially available.  The concept may work for hoisting
-expression to help factor out anticipability by working on inverted graph. If we
-insert a CHI, to a basic block with multiple successors, as instructions are
-hoisted upwards to the post-dominance frontier.  And we start walking the CFG to
-figure out if any CHI has a missing entry, that means the expression is not
-anticipable in the basic block having CHI.
+When SSAPRE inserts $\Phi$ (expression PHIs), it does so at places where
+(potentially) multiple expressions may merge. If a $\Phi$ has a missing entry,
+that means the expression is partially available.  We apply a similar concept
+for hoisting expression by working on inverted graph. We insert a CHI, to a
+basic block with multiple successors. Then for each instruction, we register its
+flow of value to its corresponding CHI which can be found in its PDF. For
+conciseness we only register values with multiple occurrences. Finally, we start
+walking the CFG and remove the CHI if it has a missing entry, that means the
+expression is not anticipable in the basic block having CHI. The completeness of
+CHI is a necessary condition for hoistability but not sufficient. This is
+because congruence of value numbers only implies that they compute same value if
+the inputs are same. So for instructions like memory references, they may
+compute different values should the memory get modified along the path. For that
+we need to check for intersecting side-effects along the path.
+
+\begin{figure}
+  \includegraphics[width=0.4\textwidth]{chi-example2.pdf}
+  \vspace*{-1.0cm}
+\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1, V\}
+  and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
+  missing entry in CHI so values are not anticipable}
+\label{fig:chi-intro2}
+\end{figure}
 

### 2017-11-08 cgo/paper.tex
move legality above

@@
 walking the CFG and remove the CHI if it has a missing entry, that means the
-expression is not anticipable in the basic block having CHI. The completeness of
-CHI is a necessary condition for hoistability but not sufficient. This is
-because congruence of value numbers only implies that they compute same value if
-the inputs are same. So for instructions like memory references, they may
-compute different values should the memory get modified along the path. For that
-we need to check for intersecting side-effects along the path.
+expression is not anticipable in the basic block having CHI. For the remaining
+CHIs anticipabilty of `values' is ensured by construction.
 
@@
 \end{figure}
+
+\subsubsection{Legality of movable instructions}
+\label{subsec:legality}
+Since the equality of candidates is purely based on the value numbers, we also
+need to establish if hoisting them or sinking them would be legal. The
+completeness of CHI is a necessary condition for hoistability but not
+sufficient. This is because congruence of value numbers only implies that they
+compute same value if the inputs are same. So for instructions like memory
+references, they may compute different values should the memory get modified
+along the path. For that we need to check for intersecting side-effects along
+the path.
+
+For hoisting, we need to prove that on all execution paths from the hoisting
+point to the end of the program the side effects appear in the same order. In
+general, we cannot introduce a new computation along any path of execution
+without checking for undefined behavior \cite{undef}. For that we need to see a)
+if I aliases with any instruction before it in its basic block and b) Walk the
+CFG from its basic block 'upwards' to its PDF and look for any interfering
+side-effects. If there aren't any then it is safe to hoist.
+
+Subsequently, it is checked that the side-effects of the computations (if any)
+do not intersect (alias) with any side-effects between the instructions to be
+moved and their hoisting/sinking point. It is also necessary to check if there
+are indirect branch targets, e.g., landing pad, case statements, goto labels,
+etc., along the path because it becomes difficult to prove safety checks in
+those cases. In our current implementation we discard candidates on those paths.
+
+\subsubsection{Legality of hoisting scalars}
+Scalars are the easiest to hoist because we do not have to analyze them for
+aliasing memory references. As long as all the operands are available, or can be
+made available by rematerialization, the scalar computations can be hoisted.
+
+\subsubsection{Legality of hoisting memory references}
+The availability of operand to the load (an address) is checked at the hoisting
+point. If that is not available we try to rematerialize the address if possible.
+Along the path, from current position of the load instruction backwards on the
+control flow to the hoisting point, we check whether there are writes to memory
+that may alias with the load, in which case the candidate is discarded. Memory
+SSA allows efficient way of iterating over the use-def chains of memory
+references on a factored graph.
+
+For stores, we check the dependency requirements similar to the hoisting of
+loads. We check that the operands of the store instruction are available at the
+hoisting point, and there are no aliasing memory references along the hoisting
+path.
+
+\subsubsection{Legality of hoisting calls}
+Call instructions can be divided into three categories: those calls equivalent
+to purely scalar computations, calls reading from memory, and calls writing to
+memory which is the most restrictive form.  Each category of call instructions
+is handled as described for scalar, load, and store instructions.
+
+Hoisting loads/stores across calls also requires precise analysis of all the
+memory addresses accessed by the call. The current implementation being an
+intraprocedural pass, cannot schedule aggressively across calls. In the presence
+of pure calls, loads can be hoisted but stores can't. Also, if a call throws
+exceptions, or if it may not return, nothing can be scheduled across that call.
+
+\subsubsection{Legality of sinking expressions}
+\label{subsec:legality-sink}
+Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
+expressions are in a landing pad, sinking of those expressions
+may reduce code size. Sinking may also reduce the register pressure in some
+cases, e.g., when the use operands are not kills. For sinking, higher ranked
+expressions would be sunk first. And it would be illegal to sink higher ranked
+identical expressions if they are not fully anticipable in the common
+post-dominator. For example:
+
+\begin{verbatim}
+B0: i0 = load B
+B1: i1 = load A
+    c1 = i1 + 10
+    d1 = i0 + 20
+    goto B3
+B2: i2 = load A
+    c2 = i2 + 10
+    d2 = i0 + 20
+    goto B3
+
+B3: phi(c1, c2)
+    phi(d1, d2)
+\end{verbatim}
+
+In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
+(c1, c2) depend on i1 and i2 respectively which are also in their original basic
+blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
+of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
+(d1, d2) can safely be sunk because their operands are readily available at the
+sink point, i.e., B3. It should also be noted that, just because the expressions
+are identical and operands are available, it still requires a unique
+post-dominating PHI to use the exact same values to be legally sinkable.
+
+A general global scheduling algorithm also requires checks for undefined values
+when introducing a new computation along a path. Since only very busy
+expressions are moved in the current implementation, there is no need to check
+for undefinedness resulting due to movement of instructions. This simplifies the
+implementation.
 
@@
 \end{verbatim}
-
-\subsection{Legality check}
-\label{subsec:legality}
-Since the equality of candidates is purely based on the value numbers, we also
-need to establish if hoisting them or sinking them would be legal. Once a common
-dominator (post-dominator) is found, we check whether all the use-operands of
-the set of instructions are available at that position. It is possible to
-reinstantiate or remateralize the use-operands in some cases when the operands
-are not available and make it legal to move the instruction.
-
-Subsequently, it is checked that the side-effects of the computations (if any)
-do not intersect (alias) with any side-effects between the instructions to be
-moved and their hoisting/sinking point. It is also necessary to check if there
-are indirect branch targets, e.g., landing pad, case statements, goto labels,
-etc., along the path because it becomes difficult to prove safety checks in
-those cases. In our current implementation we discard candidates on those paths.
-
-\subsubsection{Legality of hoisting scalars}
-Scalars are the easiest to hoist because we do not have to analyze them for
-aliasing memory references. As long as all the operands are available, or can be
-made available by rematerialization, the scalar computations can be hoisted.
-
-\subsubsection{Legality of hoisting memory references}
-The availability of operand to the load (an address) is checked at the hoisting
-point. If that is not available we try to rematerialize the address if possible.
-Along the path, from current position of the load instruction backwards on the
-control flow to the hoisting point, we check whether there are writes to memory
-that may alias with the load, in which case the candidate is discarded. Memory
-SSA allows efficient way of iterating over the use-def chains of memory
-references on a factored graph.
-
-For stores, we check the dependency requirements similar to the hoisting of
-loads. We check that the operands of the store instruction are available at the
-hoisting point, and there are no aliasing memory references along the hoisting
-path.
-
-\subsubsection{Legality of hoisting calls}
-Call instructions can be divided into three categories: those calls equivalent
-to purely scalar computations, calls reading from memory, and calls writing to
-memory which is the most restrictive form.  Each category of call instructions
-is handled as described for scalar, load, and store instructions.
-
-Hoisting loads/stores across calls also requires precise analysis of all the
-memory addresses accessed by the call. The current implementation being an
-intraprocedural pass, cannot schedule aggressively across calls. In the presence
-of pure calls, loads can be hoisted but stores can't. Also, if a call throws
-exceptions, or if it may not return, nothing can be scheduled across that call.
-
-\subsubsection{Legality of sinking expressions}
-\label{subsec:legality-sink}
-Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
-expressions are in a landing pad, sinking of those expressions
-may reduce code size. Sinking may also reduce the register pressure in some
-cases, e.g., when the use operands are not kills. For sinking, higher ranked
-expressions would be sunk first. And it would be illegal to sink higher ranked
-identical expressions if they are not fully anticipable in the common
-post-dominator. For example:
-
-\begin{verbatim}
-B0: i0 = load B
-B1: i1 = load A
-    c1 = i1 + 10
-    d1 = i0 + 20
-    goto B3
-B2: i2 = load A
-    c2 = i2 + 10
-    d2 = i0 + 20
-    goto B3
-
-B3: phi(c1, c2)
-    phi(d1, d2)
-\end{verbatim}
-
-In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
-(c1, c2) depend on i1 and i2 respectively which are also in their original basic
-blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
-of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
-(d1, d2) can safely be sunk because their operands are readily available at the
-sink point, i.e., B3. It should also be noted that, just because the expressions
-are identical and operands are available, it still requires a unique
-post-dominating PHI to use the exact same values to be legally sinkable.
-
-A general global scheduling algorithm also requires checks for undefined values
-when introducing a new computation along a path. Since only very busy
-expressions are moved in the current implementation, there is no need to check
-for undefinedness resulting due to movement of instructions. This simplifies the
-implementation.
-
-\subsubsection{Downward-safety of movable instructions}
-
-In order to be able to move an instruction that has side effects, we must
-guarantee that the side-effects of that instruction before and after its
-code-motion will produce the same side-effects.  For hoisting in particular, we
-need to prove that on all execution paths from the hoisting point to the end of
-the program the side effects appear in the same order. In general, we cannot
-introduce a new computation along any path of execution without checking for
-undefined behavior \cite{undef}. Since \GCM{} is only dealing with VBEs, as
-hoisting happens for CHIs with complete entries, the downward safety is ensured
-by construction of CHIs.
 

### 2017-11-08 cgo/paper.tex
move barriers above

@@
 
+Before we consider factors affecting the legality of \GCM{}, we introduce the
+concept of barriers.
+
+\subsection{Barriers}
+\label{subsec:barriers}
+Barriers are based on the concept of pinned instructions \cite{click1995global}
+but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
+extended basic-block because there might be non-returning calls in the middle of
+a basic-block. Essentially, any instruction that cannot guarantee progress is
+marked as a barrier. In the absence of context, as in our current
+implementation, some instructions which might be safe are still classified as
+barriers, e.g., volatile loads/stores, calls with missing attributes.
+
+\begin{verbatim}
+// Compute barriers for both hoistable and sinkable
+// candidates.
+void computeBarriers()
+  for each basic-block B in a function:
+    barrier_found = false
+    for each instruction I in B:
+      if I does not guarantee progress:
+        mark I as a barrier instruction
+        barrier_found = true
+        break;
+
+    // Find the last barrier below which instructions
+    // can be sunk. If there was no barrier in B, any
+    // instruction satisfying other legality checks
+    // can be sunk.
+    if barrier_found:
+      for each instruction I in B in reverse order:
+        if I does not guarantee progress:
+          mark I as a sink barrier
+          break
+\end{verbatim}
+
+Computing barriers allows for efficient removal of non-movable candidates, this
+makes the legality check converge faster.
+
 \subsubsection{Legality of movable instructions}
 \label{subsec:legality}
-Since the equality of candidates is purely based on the value numbers, we also
-need to establish if hoisting them or sinking them would be legal. The
+Since the equality of movable candidates is purely based on the value numbers,
+we also need to establish if hoisting them or sinking them would be legal. The
 completeness of CHI is a necessary condition for hoistability but not
@@
 general, we cannot introduce a new computation along any path of execution
-without checking for undefined behavior \cite{undef}. For that we need to see a)
-if I aliases with any instruction before it in its basic block and b) Walk the
-CFG from its basic block 'upwards' to its PDF and look for any interfering
-side-effects. If there aren't any then it is safe to hoist.
-
-Subsequently, it is checked that the side-effects of the computations (if any)
-do not intersect (alias) with any side-effects between the instructions to be
-moved and their hoisting/sinking point. It is also necessary to check if there
-are indirect branch targets, e.g., landing pad, case statements, goto labels,
-etc., along the path because it becomes difficult to prove safety checks in
-those cases. In our current implementation we discard candidates on those paths.
+without checking for undefined behavior \cite{undef}.  For that, it is checked
+that the side-effects of the computations (if any) do not intersect (alias) with
+any side-effects between the instructions to be moved and their hoisting/sinking
+point. It is also necessary to check if there are indirect branch targets, e.g.,
+landing pad, case statements, goto labels, etc., along the path because it
+becomes difficult to prove safety checks in those cases. In our current
+implementation we discard candidates on those paths.
 
@@
 post-dominating PHI to use the exact same values to be legally sinkable.
-
-A general global scheduling algorithm also requires checks for undefined values
-when introducing a new computation along a path. Since only very busy
-expressions are moved in the current implementation, there is no need to check
-for undefinedness resulting due to movement of instructions. This simplifies the
-implementation.
 
@@
 precomputed.
-
-\subsubsection{Barriers}
-\label{subsec:barriers}
-Barriers are based on the concept of pinned instructions \cite{click1995global}
-but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
-extended basic-block because there might be non-returning calls in the middle of
-a basic-block. Essentially, any instruction that cannot guarantee progress is
-marked as a barrier. In the absence of context, as in our current
-implementation, some instructions which might be safe are still classified as
-barriers, e.g., volatile loads/stores, calls with missing attributes.
-
-\begin{verbatim}
-// Compute barriers for both hoistable and sinkable
-// candidates.
-void computeBarriers()
-  for each basic-block B in a function:
-    barrier_found = false
-    for each instruction I in B:
-      if I does not guarantee progress:
-        mark I as a barrier instruction
-        barrier_found = true
-        break;
-
-    // Find the last barrier below which instructions
-    // can be sunk. If there was no barrier in B, any
-    // instruction satisfying other legality checks
-    // can be sunk.
-    if barrier_found:
-      for each instruction I in B in reverse order:
-        if I does not guarantee progress:
-          mark I as a sink barrier
-          break
-\end{verbatim}
 

### 2017-11-08 cgo/paper.tex
candidates to schedule

@@
 
-Dominance graphs, Global Value Numbers and Memory SSA are already available in
+Dominator tree, Global Value Numbers and Memory SSA are already available in
 \LLVM{} but there is no facility to infer liveness of virtual registers. So we
@@
 post-dominance frontier is the basic block where anticipability (ANTIC) of an
-instruction (I) is checked because it the point on which the basic block
-containing I is control dependent.
+instruction (I) is checked because it is the basic block on which the basic
+block containing I is control dependent.
 
@@
 
-A CHI keeps information about values flowing out of a basic block.  It is
-similar to $\phi$ \cite{ssa} but in the inverse graph, and used for out-flowing
-values on each edge. Although it may be suitable for other
+A CHI keeps information about `congruent' values flowing out of a basic block.
+It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used for
+trackig out-flowing values on each edge. Although it may be suitable for other
 analyses/optimizations, for \GCM{}, it is computed only for instructions with
@@
 will save the instruction as well as the edge where the value is flowing to.
-Where $Arg1$, $Arg2$ are of type CHIArg. The CHI for both I1 and I2 is
-inserted in their PDF (A). For sinking we use the similar approach as
-\cite{ssapre} to track values flowing into a basic block.
+Where $Arg1$, $Arg2$ are of type CHIArg. The CHI for both I1 and I2 is inserted
+in their PDF (A). For sinking we use the similar approach as \cite{ssapre} to
+track values flowing into a basic block.
 
@@
 conciseness we only register values with multiple occurrences. Finally, we start
-walking the CFG and remove the CHI if it has a missing entry, that means the
-expression is not anticipable in the basic block having CHI. For the remaining
-CHIs anticipabilty of `values' is ensured by construction.
+walking the CFG and remove the CHI if it has a missing entry (see
+Figure~\cite{fig:chi-intro2}), that means the expression is not anticipable in
+the basic block having CHI. For the remaining CHIs anticipabilty of `values' is
+ensured by construction.
 
@@
 
-\subsection{Finding candidates to move}
+\subsection{Finding candidates to schedule}
 \label{subsec:finding-candidates}

### 2017-11-08 cgo/paper.tex
codehoisting

@@
 \begin{verbatim}
-int doCodeHoisting()
+int doCodeHoisting(VNs)
+  // VNs = map (VN -> list of I with value VN)
   Sort VNs in ascending order
 
+  for each VN in VNs
     for each instruction with the same VN

### 2017-11-11 cgo/paper.tex
chi symbol

@@
 
-So we introduce a data structure (CHI nodes) to keep track of values flowing out
+So we introduce a data structure ($\chi$ nodes) to keep track of values flowing out
 of a basic block with more than once successors. We only do this for values with
@@
 candidates. This approach allows us to hoist instructions to a basic block with
-more than two successors. The CHI node in a basic block $B$ has CHIArgs with
+more than two successors. The $\chi$ node in a basic block $B$ has CHIArgs with
 three entries: a) the instruction (I) CHIArg refers to, b) the basic block $Dest$
@@
 
-A CHI keeps information about `congruent' values flowing out of a basic block.
+A $\chi$ keeps information about `congruent' values flowing out of a basic block.
 It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used for
@@
 analyses/optimizations, for \GCM{}, it is computed only for instructions with
-multiple occurrences in the CFG. An illustration of CHI is shown in
-Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the CHI node
+multiple occurrences in the CFG. An illustration of $\chi$ is shown in
+Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the $\chi$ node
 will save the instruction as well as the edge where the value is flowing to.
-Where $Arg1$, $Arg2$ are of type CHIArg. The CHI for both I1 and I2 is inserted
+Where $Arg1$, $Arg2$ are of type CHIArg. The $\chi$ for both I1 and I2 is inserted
 in their PDF (A). For sinking we use the similar approach as \cite{ssapre} to
@@
 that means the expression is partially available.  We apply a similar concept
-for hoisting expression by working on inverted graph. We insert a CHI, to a
+for hoisting expression by working on inverted graph. We insert a $\chi$, to a
 basic block with multiple successors. Then for each instruction, we register its
-flow of value to its corresponding CHI which can be found in its PDF. For
+flow of value to its corresponding $\chi$ which can be found in its PDF. For
 conciseness we only register values with multiple occurrences. Finally, we start
-walking the CFG and remove the CHI if it has a missing entry (see
+walking the CFG and remove the $\chi$ if it has a missing entry (see
 Figure~\cite{fig:chi-intro2}), that means the expression is not anticipable in
-the basic block having CHI. For the remaining CHIs anticipabilty of `values' is
-ensured by construction.
+the basic block having $\chi$. For the remaining $\chi$s anticipabilty of `values'
+is ensured by construction.
 
@@
   and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
-  missing entry in CHI so values are not anticipable}
+  missing entry in $\chi$ so values are not anticipable}
 \label{fig:chi-intro2}
@@
 implementation, some instructions which might be safe are still classified as
-barriers, e.g., volatile loads/stores, calls with missing attributes.
+barriers, e.g., calls with missing attributes.
 
@@
 we also need to establish if hoisting them or sinking them would be legal. The
-completeness of CHI is a necessary condition for hoistability but not
+completeness of $\chi$ is a necessary condition for hoistability but not
 sufficient. This is because congruence of value numbers only implies that they

### 2017-11-11 cgo/paper.tex
formatting

@@
 in a function as potential candidates. We also formalize why register pressure
-reduces, and how \GCM{} increases instruction level parallelism thereby enabling
-more \ooo{} execution on modern architectures.
+reduces as a result of \gcm{}, and how \gcm{} increases instruction level
+parallelism thereby enabling more \ooo{} execution on modern architectures.
 \end{abstract}
@@
 So we introduce a data structure ($\chi$ nodes) to keep track of values flowing out
-of a basic block with more than once successors. We only do this for values with
+of a basic block with more than one successor. We only do this for values with
 multiple occurrences in the function as they are the potentially hoistable
@@
 walking the CFG and remove the $\chi$ if it has a missing entry (see
-Figure~\cite{fig:chi-intro2}), that means the expression is not anticipable in
+Figure~\ref{fig:chi-intro2}), that means the expression is not anticipable in
 the basic block having $\chi$. For the remaining $\chi$s anticipabilty of `values'

### 2017-11-11 cgo/paper.tex
sinking live range

@@
 %% Title information
-\title[SSA GCM]{Global code motion of congruent instructions}         %% [Short Title] is optional;
+\title[SSA GCM]{A fast algorithm for global code motion of congruent instructions}         %% [Short Title] is optional;
 
@@
 \begin{abstract}
-We present a \gcm{} (\GCM{}) compiler optimization which schedules congruent
+We present a fast \gcm{} (\GCM{}) compiler optimization which schedules congruent
 instructions across the program.  Not only \GCM{} saves code size, it exposes
@@
 \section{A fast algorithm for code motion congruent computations}
-The algorithm factors out the reachability of values such that multiple queries
+Our algorithm factors out the reachability of values such that multiple queries
 to find reachability of values are fast. This is based on finding the dominance
-frontiers and the post-dominance frontiers (PDF) of the CFG. These points, being
-structural properties of a graph, do not change during code-motion. A
+frontiers (DF) and the post-dominance frontiers (PDF) of the CFG. These points,
+being structural properties of a graph, do not change during code-motion. A
 post-dominance frontier is the basic block where anticipability (ANTIC) of an
@@
 
-So we introduce a data structure ($\chi$ nodes) to keep track of values flowing out
-of a basic block with more than one successor. We only do this for values with
-multiple occurrences in the function as they are the potentially hoistable
-candidates. This approach allows us to hoist instructions to a basic block with
-more than two successors. The $\chi$ node in a basic block $B$ has CHIArgs with
-three entries: a) the instruction (I) CHIArg refers to, b) the basic block $Dest$
-which is post-dominated by the basic block containing I, and c) the GVN of I.
+So we introduce a data structure ($\chi$ nodes) to keep track of values flowing
+out of a basic block with more than one successor. This approach allows us to
+hoist instructions to a basic block with more than two successors. The $\chi$
+node in a basic block $B$ has CHIArgs with three entries: a) the instruction (I)
+CHIArg refers to, b) the basic block $Dest$ which is post-dominated by the basic
+block containing I, and c) the GVN of I.
 
@@
 
-Limitations: This pass does not hoist fully redundant expressions because they
-are already handled by GVN-PRE. It is advisable to run gvn-hoist before and
-after gvn-pre because gvn-pre creates opportunities for more instructions to be
-hoisted.
-
-A $\chi$ keeps information about `congruent' values flowing out of a basic block.
-It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used for
-trackig out-flowing values on each edge. Although it may be suitable for other
-analyses/optimizations, for \GCM{}, it is computed only for instructions with
-multiple occurrences in the CFG. An illustration of $\chi$ is shown in
-Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the $\chi$ node
-will save the instruction as well as the edge where the value is flowing to.
-Where $Arg1$, $Arg2$ are of type CHIArg. The $\chi$ for both I1 and I2 is inserted
-in their PDF (A). For sinking we use the similar approach as \cite{ssapre} to
-track values flowing into a basic block.
+A $\chi$ keeps information about `congruent' values flowing out of a basic
+block.  It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used
+for trackig out-flowing values on each edge. An illustration of $\chi$ is shown
+in Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the $\chi$
+node will save the instruction as well as the edge where the value is flowing
+to.  Where $Arg1$, $Arg2$ are of type CHIArg. The $\chi$ for both I1 and I2 is
+inserted in their PDF (A).
 
 \begin{figure}
-  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
+  \includegraphics[scale=0.55]{chi-example.pdf}
   \vspace*{-1.0cm}
@@
 
-When SSAPRE inserts $\Phi$ (expression PHIs), it does so at places where
-(potentially) multiple expressions may merge. If a $\Phi$ has a missing entry,
-that means the expression is partially available.  We apply a similar concept
-for hoisting expression by working on inverted graph. We insert a $\chi$, to a
-basic block with multiple successors. Then for each instruction, we register its
-flow of value to its corresponding $\chi$ which can be found in its PDF. For
-conciseness we only register values with multiple occurrences. Finally, we start
-walking the CFG and remove the $\chi$ if it has a missing entry (see
-Figure~\ref{fig:chi-intro2}), that means the expression is not anticipable in
-the basic block having $\chi$. For the remaining $\chi$s anticipabilty of `values'
-is ensured by construction.
+A similar idea has been used in SSAPRE \cite{ssapre} for partial redundancy
+elimination in spare form. When SSAPRE inserts $\Phi$ (expression PHIs), it does
+so at places where (potentially) multiple expressions may merge. If a $\Phi$ has
+a missing entry, that means the expression is partially available. Subsequent
+algorithms work on $\Phi$s to find out redundancies. We extend the idea for
+\GCM{} by propagating the anticipability in both directions.
 
 \begin{figure}
-  \includegraphics[width=0.4\textwidth]{chi-example2.pdf}
+  \includegraphics[scale=0.55]{chi-example2.pdf}
   \vspace*{-1.0cm}
@@
   and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
-  missing entry in $\chi$ so values are not anticipable}
+  missing entry in $\chi$ so V is not anticipable}
 \label{fig:chi-intro2}
@@
 
-Before we consider factors affecting the legality of \GCM{}, we introduce the
-concept of barriers.
+For code hoisting we insert a $\chi$, to a basic block with multiple
+successors. Then for each instruction, we register its flow of value to its
+corresponding $\chi$ which can be found in its PDF. Finally, we start walking
+the CFG and remove the $\chi$ if it has a missing entry (see
+Figure~\ref{fig:chi-intro2}), that means the expression is not anticipable in
+the basic block having incomplete $\chi$. For the remaining $\chi$s
+anticipabilty of `values' is ensured by construction. For code-sinking we insert
+$\Phi$s in basic blocks with multiple predecessors. Then for each instruction we
+register its flow of value to its correponding $\Phi$ which can be found in its
+DF.  Finally, we start walking the CFG and remove the $\Phi$ if it has a missing
+entry (see Figure~\ref{fig:phi-intro}), that means the expression is not
+available in the basic block having incomplete $\Phi$. For the remaining $\Phi$s
+availability of `values' is ensured by construction.
+
+\begin{figure}[h]
+  \includegraphics[scale=0.55]{phi-example.pdf}
+  \vspace*{-1.0cm}
+  \caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1, V\}
+  and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
+  missing entry in $\Phi$ so V is not available}
+\label{fig:phi-intro}
+\end{figure}
+
+In order to limit the number of spurious $\chi$s and $\Phi$s,
+\cite{ssapre-improv}, we insert them only for values with multiple occurrences
+in the function as they are the interesting candidates for this \GCM{}. We
+believe this factorization of anticipability/availability of expressions may be
+suitable for faster implementation of other global (across basic blocks)
+analyses/optimizations as well. Before we consider factors affecting the
+legality of \GCM{}, we introduce the concept of barriers.
 
@@
 
-\subsubsection{Legality of movable instructions}
+\subsection{Legality of movable instructions}
 \label{subsec:legality}
@@
 we also need to establish if hoisting them or sinking them would be legal. The
-completeness of $\chi$ is a necessary condition for hoistability but not
+completeness of $\chi$ and $\Phi$ are necessary for \GCM{} but not
 sufficient. This is because congruence of value numbers only implies that they
@@
 
-For hoisting, we need to prove that on all execution paths from the hoisting
-point to the end of the program the side effects appear in the same order. In
-general, we cannot introduce a new computation along any path of execution
-without checking for undefined behavior \cite{undef}.  For that, it is checked
-that the side-effects of the computations (if any) do not intersect (alias) with
-any side-effects between the instructions to be moved and their hoisting/sinking
-point. It is also necessary to check if there are indirect branch targets, e.g.,
-landing pad, case statements, goto labels, etc., along the path because it
-becomes difficult to prove safety checks in those cases. In our current
-implementation we discard candidates on those paths.
-
-\subsubsection{Legality of hoisting scalars}
-Scalars are the easiest to hoist because we do not have to analyze them for
-aliasing memory references. As long as all the operands are available, or can be
-made available by rematerialization, the scalar computations can be hoisted.
-
-\subsubsection{Legality of hoisting memory references}
-The availability of operand to the load (an address) is checked at the hoisting
-point. If that is not available we try to rematerialize the address if possible.
-Along the path, from current position of the load instruction backwards on the
-control flow to the hoisting point, we check whether there are writes to memory
-that may alias with the load, in which case the candidate is discarded. Memory
-SSA allows efficient way of iterating over the use-def chains of memory
-references on a factored graph.
-
-For stores, we check the dependency requirements similar to the hoisting of
-loads. We check that the operands of the store instruction are available at the
-hoisting point, and there are no aliasing memory references along the hoisting
-path.
-
-\subsubsection{Legality of hoisting calls}
-Call instructions can be divided into three categories: those calls equivalent
-to purely scalar computations, calls reading from memory, and calls writing to
-memory which is the most restrictive form.  Each category of call instructions
-is handled as described for scalar, load, and store instructions.
-
-Hoisting loads/stores across calls also requires precise analysis of all the
-memory addresses accessed by the call. The current implementation being an
-intraprocedural pass, cannot schedule aggressively across calls. In the presence
-of pure calls, loads can be hoisted but stores can't. Also, if a call throws
-exceptions, or if it may not return, nothing can be scheduled across that call.
-
-\subsubsection{Legality of sinking expressions}
-\label{subsec:legality-sink}
-Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
-expressions are in a landing pad, sinking of those expressions
-may reduce code size. Sinking may also reduce the register pressure in some
-cases, e.g., when the use operands are not kills. For sinking, higher ranked
-expressions would be sunk first. And it would be illegal to sink higher ranked
-identical expressions if they are not fully anticipable in the common
-post-dominator. For example:
-
-\begin{verbatim}
-B0: i0 = load B
-B1: i1 = load A
-    c1 = i1 + 10
-    d1 = i0 + 20
-    goto B3
-B2: i2 = load A
-    c2 = i2 + 10
-    d2 = i0 + 20
-    goto B3
-
-B3: phi(c1, c2)
-    phi(d1, d2)
-\end{verbatim}
-
-In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
-(c1, c2) depend on i1 and i2 respectively which are also in their original basic
-blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
-of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
-(d1, d2) can safely be sunk because their operands are readily available at the
-sink point, i.e., B3. It should also be noted that, just because the expressions
-are identical and operands are available, it still requires a unique
-post-dominating PHI to use the exact same values to be legally sinkable.
+For that, we need to prove that on all execution paths from the initial position
+of an instruction to its destination, the side effects appear in the same
+order. In general, we cannot introduce a new computation along any path of
+execution without checking for undefined behavior \cite{undef}. It is also
+necessary to check if there are indirect branch targets, e.g., landing pad, case
+statements, goto labels, etc., along the path because it becomes difficult to
+prove safety checks in those cases.
+
+Scalars are the easiest to hoist because we only need to check for availability
+of operands. For sinking the scalars we need to make sure that there are no uses
+of the instruction along the path, For instructions with memory references like
+loads, stores and calls, Memory SSA allows efficient way of iterating over the
+use-def chains of memory references on a factored graph. Hoisting loads/stores
+across calls also requires precise analysis of all the memory addresses accessed
+by the call. The current implementation being an intraprocedural pass, cannot
+schedule aggressively across calls. In the presence of pure calls, loads can be
+hoisted but stores can't. Also, if a call throws exceptions, or if it may not
+return, nothing can be scheduled across that call.
 
@@
 \label{subsec:cost-models}
-TODO: Motivate why this is important.
-After the legality checks have passed, we check if a \gcm{} is profitable.  That
-takes into account the impact \gcm{} would have on various parameters that might
+After the legality checks have passed, we check if a \GCM{} is profitable.  That
+takes into account the impact \GCM{} would have on various parameters that might
 affect runtime performance, e.g., impact on live-range, gain in the code
-size. The current implementation makes effort to not regress in performance
-across popular benchmarks and at the same time to reduce code size as much as
-possible. Following cost models are implemented:
+size. The current implementation makes effort to not regress in performance but
+at the same time to reduce code size as much as possible. Sometimes, hoisting is
+not upward-safe \cite{click1995global}, e.g., if the expressions are in a
+landing pad, sinking of those expressions may reduce code size. Following cost
+models are implemented:
 
@@
 freed at the insertion point but only one register will be required to assign
-the definition of hoisted instruction $a0$.
-
-Ideally, it should be okay to hoist all the instructions and a later
-live-range-splitting \cite{cooper1998live} pass should make the right decision
-but the current live-range splitting pass of \LLVM{} is not making the optimal
-decision and we have found performance regressions while hoisting aggressively.
-
-There is a special case for hoisting load instructions when the hoist-point is
-the predecessor basic-block for all the congruent loads. Even if the register
-pressure would increase, we prefer to hoist loads in this case. The reason
-being, that would make the loaded value available by the time it is used, and
-livess would not increase by much because instruction is hoisted to immediate
-predecessor.
+the definition of hoisted instruction $a0$. We have reduced performance
+regressions with this cost model even if LLVM has a live-range splitting
+\cite{cooper1998live} optimization.
+
+Sinking may also reduce the register pressure in some cases, e.g., when the use
+operands are not kills. For sinking, higher ranked expressions would be sunk
+first. And it would be illegal to sink higher ranked identical expressions if
+they are not fully anticipable in the common post-dominator. For example:
+
+\begin{verbatim}
+B0: i0 = load B
+B1: i1 = load A
+    c1 = i1 + 10
+    d1 = i0 + 20
+    goto B3
+B2: i2 = load A
+    c2 = i2 + 10
+    d2 = i0 + 20
+    goto B3
+
+B3: PHIc1, c2)
+    PHI(d1, d2)
+\end{verbatim}
+
+In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
+(c1, c2) depend on i1 and i2 respectively which are also in their original basic
+blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
+of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
+(d1, d2) can safely be sunk because their operands are readily available at the
+sink point, i.e., B3. It should also be noted that, just because the expressions
+are identical and operands are available, it still requires a unique
+post-dominating PHI to use the exact same values to be legally sinkable.
 
@@
 which dependencies are not directly available in the successor
-(Section~\ref{subsec:legality-sink}) is not sunk for now because that would
+(Section~\ref{subsec:legality}) is not sunk for now because that would
 require a look ahead on the ability to sink the dependencies. We plan to
@@
 LLVM trunk as GVNHoist.cpp and the \GCM{} implementation is under review.
+
+This pass does not hoist fully redundant expressions because they are already
+handled by GVN-PRE. It is advisable to run gvn-hoist before and after gvn-pre
+because gvn-pre creates opportunities for more instructions to be hoisted.
 

### 2017-11-11 cgo/paper.tex
codegen

@@
 \usepackage{enumitem}
+\usepackage{listings}
+\lstset{
+  basicstyle=\ttfamily,
+  %columns=fixed,
+  %fontadjust=true,
+  basewidth=0.4em,
+  language=C,
+  mathescape
+}
 \usepackage[final]{pdfpages}
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 float fun(float d, float min, float max, float a) {
@@
 }
-\end{verbatim}
+\end{lstlisting}
 
@@
 conditionals, the C code version would look like:
-\begin{verbatim}
+\begin{lstlisting}
 float fun(float d, float min, float max, float a) {
@@
 
-\section{A fast algorithm for code motion congruent computations}
+\section{Factoring reachability of expressions}
 Our algorithm factors out the reachability of values such that multiple queries
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 struct CHIArg {
@@
 };
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 // Compute barriers for both hoistable and sinkable
@@
           break
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 B0: b = m
@@
 B3: ...
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 B0: i0 = load B
@@
     PHI(d1, d2)
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\subsection{Finding candidates to schedule}
-\label{subsec:finding-candidates}
-The first step is to find a set of congruent instructions
-\cite{briggs1997}. This is performed by a linear scan of all instructions of the
-program and classifying them by their value numbers.
-
-The current implementation of \GVN{} in \LLVM{} has some limitations when it
-comes to loads and stores so we compute the \GVN{} of loads and stores
-separately.  Our solution is to value number the address from where the value of
-a load is to be read from memory. For stores, we value number the address as
-well as the value to be stored at that address. Another limitation of the
-current \GVN{} implementation in \LLVM{} is that the instructions dependent on
-the loads will not get numbered correctly, and so after hoisting all candidates
-we need to rerun the \GVN{} analysis in order to discover new candidates now
-available after having hoisted load instructions.
-
-The process of computing \GVN{} can be on-demand, as we come across an
-instruction, or precomputed, computing \GVN{} of all the instructions
-beforehand. Which process to choose is determined by the scope of code-hoisting
-we want to perform. In a pessimistic approach, we want to hoist a limited set of
-instructions from the sibling branches as we iterate the DFS tree bottom-up, it
-is sufficient to compute \GVN{} values on-demand. Whereas, in the optimistic
-approach, as described in Section~\ref{subsec:optimistic}, we want to move as
-many instructions as possible, and it would require \GVN{} values to be
-precomputed.
-
-\subsection{Code generation for the optimistic \gcm{} algorithm}
+\section{A fast algorithm for code motion of congruent computations}
 \label{subsec:optimistic}
 Once all the legality and profitability checks are satisfied for a set of
-congruent instructions, they are suitable candidates for hoisting or sinking. A
+congruent instructions, they are suitable candidates for code motion. A
 copy of the computation is inserted at the hoisting/sinking point along with any
 instructions which needed to be rematerialized. Thereafter, all the computations
-made redundant by the new copy are removed, and the \SSA{} form is restored by
-updating the intermediate representation (IR) to reflect the changes. At the
-same time \MemorySSA{} is updated as well.
-
-After one iteration of algorithm runs through the entire function, it creates
-more opportunities for \emph{higher ranked} computations
-\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
-pass, and so we rerun the code-hoisting algorithm until there are no more
-instructions left to be hoisted.  Obviously, this is not the most optimal
-approach and can be improved by ranking the computations \cite{rosen1988global},
-or by improving the \GVN{} analysis to correctly populate congruence classes as
-the program is modified by the code generation.
-
-Finally after the transformation is done, we verify a set of post-conditions to
-establish that program invariants are maintained: e.g., consistency of use-defs,
-and \SSA{} semantics.
-
-The amount of hoisting depends on whether we collect \GVN{} of instructions
-before finding candidates (optimistic) or, on-demand (pessimistic). It also
-depends on the generality of the \GVN{} algorithm as mentioned earlier in
-Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
-\gcm{} of congruent instructions which uses the liveness analysis as
-illustrated by \citet{das2012} and ranking expressions explained by
-\citet{rosen1988global}.
-
-\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
-implementation only moves congruent instructions. An immediate guarantee of this
-approach is gain in code size (the final executable may be of larger size
-because of more inlining.) The algorithm prefers hoisting to sinking. If the
-dependency of a hoistable candidate is in the same basic-block as the candidate,
-then the dependency must also be hoistable, otherwise hoisting will be illegal or
-would require a complicated code generation to make it legal. The current
-algorithm discards cases if the dependency is neither hoistable nor
-rematerializable.
-
-\begin{verbatim}
+made redundant by the new copy are removed.
+
+\begin{lstlisting}
 void doGCM()
@@
   while (doCodeHoisting() > 0)
+  // Code-sinking only once
   doCodeSinking()
-\end{verbatim}
-
-We collect the \GVN{} of all the instructions in the function and iterate on the
-list of instructions having identical \GVN{}s. The algorithm \emph{doGCM}
-prefers hoisting (\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).
-
-\begin{verbatim}
+\end{lstlisting}
+
+\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
+implementation only moves congruent instructions. An immediate guarantee of this
+approach is gain in code size (the final executable may be of larger size
+because of more inlining). If the dependency of a hoistable candidate is in the
+same basic-block as the candidate, then the dependency must also be hoistable,
+otherwise hoisting will be illegal or would require a complicated code
+generation to make it legal. The current algorithm discards cases if the
+dependency is neither hoistable nor rematerializable. We collect the \GVN{} of
+all the instructions in the function and iterate on the list of instructions
+having identical \GVN{}s. The algorithm \emph{doGCM} prefers hoisting
+(\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).
+
+\begin{lstlisting}
 int doCodeHoisting(VNs)
@@
   for each CHI-nodes in the CFG:
-    Remove the ones with missing entries
+    for each CHI-arg in CHI-node:
       Remove CHI-arg if it fails legality checks
@@
       Perform code hoisting
+      update data structures and statistics
+
+  return number of hoisted instructions
+\end{lstlisting}
+
+Code hoisting opens new opportunities for other higher ranked
+\cite{rosen1988global} (depended on candidates which got hoisted) which are
+encountered subsequently as the algorithm visits instructions in the increasing
+order of their ranks. Some of the congruent higher ranked computations do not
+get equal value numbers so we rerun the code-hoisting algorithm until there are
+no more instructions left to be hoisted. The hoisting converges quickly with 2-3
+iterations.
+
+\begin{lstlisting}
+int doCodeSinking()
+  // VNs = map (VN -> list of I with value VN)
+  Sort VNs in ascending order
+
+  for each VN in VNs
+    for each instruction with the same VN
+      Find its dominance-frontier (DF)
+      Insert the PHI-node at DF
+
+  for each PHI-nodes in the CFG:
+    for each PHI-arg in PHI-node:
+      Remove PHI-arg if it fails legality checks
+
+  for each PHI-nodes in the CFG:
+    if PHI does not have missing entries:
+      Perform code sinking
+      if a $\phi$ has same entries, reassign
+      update data structures and statistics
 
       return number of sinked instructions
-\end{verbatim}
-
-Code hoisting opens new opportunities for other hoistable candidates which were
-of higher rank (depended on candidates which got hoisted) which are encountered
-subsequently as the algorithm visits instructions in the increasing order of
-their ranks.
-
-%\newpage
-After no more candidates are hoistable, sinking is performed. Sinking is only
-performed once because currently there are very few sinkable candidates
-(Figure~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
-\GCM{}) has a code-sinking which already sinks several instructions.
-
-For sinking we reuse the $\phi$ nodes instead of inserting another expression nodes
-like in \cite{ssapre} because $\phi$ nodes ensure availability of definitions, and
-hence downward safety, as long as all instructions corresponding to a $\phi$ are
-congruent. Also, the $\phi$ nodes represent a complete set of instructions which
-could be potentially sunk for code-size optimizations because $\phi$ nodes
-represent all values flowing out from a basic block into its dominance-frontier.
-
-\begin{verbatim}
-int doCodeSinking()
-  // Find sinkable instructions
-  For each value number VN > 1 instructions:
-    if there are two instructions with same immediate
-        successor(S) as post-dominator:
-      if both are only used in the same $\phi$-node of S
-          && dependencies of both are available at S:
-         mark instructions as sinkable
-
-  // Sink the instructions and update statistics
-  For each pair (I1, I2) of sinkable instructions:
-    Move I1 to the sink point
-    // the sink point is just after all
-    // the PHI-nodes in the post-dominator
-    update all the uses of PHI node (PN) with I1
-    Remove I2 and PN
-    if I1 is a memory reference:
-      update MemorySSA of I1
-      update MemorySSA of removed instructions
-        to point to the MemorySSA reference of I1
-      remove MemorySSA reference of all
-        others which were deleted
-    update statistics
-
-  return number of sinked instructions
-\end{verbatim}
+\end{lstlisting}
+
+After no more candidates are hoistable, sinking is performed. It is only
+performed once because it may not open new opportunities for sinkable
+candidates. For sinking we iterate on the $\Phi$ nodes. After each code-motion
+the \SSA{} form is restored by updating the intermediate representation (IR) to
+reflect the changes. At the same time \MemorySSA{} is updated as well. Both the
+data structures can be updated very efficiently. After one full iteration of
+hoisting, the \GVN{} needs to be recomputed to explore new hoistable candidates.
+Finally after the transformation is done, we verify a set of post-conditions to
+establish that program invariants are maintained: e.g., consistency of use-defs,
+and \SSA{} semantics.
 

### 2017-11-11 cgo/paper.tex
acmart.cls

@@
 
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
 %% Keywords
@@
 \section{Introduction}
-
 \label{sec:intro}

### 2017-11-12 cgo/paper.tex
table

@@
 % llvm trunk vs. llvm trunk with code-motion
-                &\% performance uplift & size reduction (Bytes) & size reduction (Bytes)\\
-Benchmarks      & at -Ofast (high better)& at -Ofast (high better) & at -Os (high better)\\\hline
-400.perlbench	& 1.00  & -9120  & 544  \\\hline
-401.bzip2	& 1.00  & -1600  & 128  \\\hline
-403.gcc	        & 1.00  & -58176 & 3384 \\\hline
-429.mcf	        & 1.03  & 160    & 16   \\\hline
-433.milc	& 1.00  & -656   & 224  \\\hline
-444.namd	& 1.00  & -3864  & -1168\\\hline
-445.gobmk	& 0.99  & -20480 & -264 \\\hline
-447.dealII	& 0.95  & -27836 & 50934\\\hline
-450.soplex	& 0.99  & 2220   & 348  \\\hline
-453.povray	& 0.97  & 164    & -472 \\\hline
-456.hmmer	& 0.99  & -3464  & -352 \\\hline
-458.sjeng	& 1.02  & 440    & 1456 \\\hline
-462.libquantum	& 1.00  & -96    & 0    \\\hline
-464.h264ref	& 1.00  & -3424  & 864  \\\hline
-470.lbm	        & 1.00  & 0      & 0    \\\hline
-471.omnetpp	& 1.00  & 2688   & 200  \\\hline
-473.astar	& 1.00  & -928   & 152  \\\hline
-482.sphinx3	& 0.99  & 3008   & -176 \\\hline
-483.xalancbmk	& 1.01  & 46492  & 4012 \\\hline
-Geomean         & 0.997 &  & \\\hline
+                &\% performance uplift & size reduction (Bytes) \\
+Benchmarks      & at -Ofast (high better)& at -Ofast (high better)\\\hline
+400.perlbench	&  1.95  & 984   \\\hline
+401.bzip2	& -0.84  & 16    \\\hline
+403.gcc	        & -2.30  & 2128  \\\hline
+429.mcf	        & -4.36  & 16    \\\hline
+433.milc	& -1.86  & 48    \\\hline
+444.namd	& -2.40  & 544   \\\hline
+445.gobmk	& -0.52  & 56    \\\hline
+447.dealII	&  0.39  & -2560 \\\hline
+450.soplex	& -0.24  & -32   \\\hline
+453.povray	& -0.05  & 848   \\\hline
+456.hmmer	&  0.93  & 296   \\\hline
+458.sjeng	& -0.58  & 232   \\\hline
+462.libquantum	&  8.17  & -16   \\\hline
+470.lbm	        &  2.57  & 0     \\\hline
+471.omnetpp	&  5.30  & 152   \\\hline
+473.astar	& -0.17  & 48    \\\hline
+482.sphinx3	& -0.11  & 16    \\\hline
+483.xalancbmk	& -2.98  & 392   \\\hline
     \end{tabular}}
@@
   \caption{Execution time (ratio) and code size change (Bytes) with and without
-    \GCM{} on \SPEC{}: performance uplift at in mcf and sjeng;
-    code size improvement in dealII.}
+    \GCM{} on \SPEC{}}
   \label{tab:code-size}

### 2017-11-13 cgo/paper.tex
table caption

@@
   \end{center}
-  \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at
-    -Ofast. Lower is better in \%age change. Spills reduced by 2\%.}
+  \caption{Number of spills with and without\GCM{} on SPEC2006 at
+    -Ofast. Lower is better in GCM/base. Spills reduced by 2\%.}
   \label{tab:spills}
@@
   \caption{Change in total number of instructions executed during the
-    compilation with and without \gcm{} (\GCM{}) on \SPEC{} at -Ofast. The
+    compilation with and without \GCM{} on \SPEC{} at -Ofast. The
     instructions were counted using valgrind.}
@@
   \end{center}
-  \caption{Change in the static compile time metrics w.r.t. \gcm{} (\GCM{}) on
+  \caption{Change in the static compile time metrics w.r.t. \GCM{} on
     \SPEC{} at -Ofast. The \% columns show percentage increase in metric

### 2017-11-13 cgo/paper.tex
abstract

@@
 \begin{abstract}
-We present a fast \gcm{} (\GCM{}) compiler optimization which schedules congruent
-instructions across the program.  Not only \GCM{} saves code size, it exposes
-redundancies as well, it exposes more instruction level parallelism in the
-basic-block to which instructions are moved, and it enables other passes like
-loop invariant motion to remove redundancies. The cost model to drive the code
-motion is based on liveness analysis on \SSA{} representation such that the
-(virtual) register pressure does not increase resulting in 2\% fewer spills on
-the \SPEC{} benchmark suite.
-
-We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
-infrastructure available in \LLVM{}. The experimental results show an average
-saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
-inlining and exposes more loop invariant code motion opportunities in the
-majority of benchmarks. We have also seen execution time improvements in a few
-of \SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by
-2\% on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
+We present a fast \gcm{} (\GCM{}) compiler optimization which schedules
+congruent instructions across the program. We introduce data structures to help
+exploit flow of global values in sparse representation.  Not only \GCM{}
+improves performance it saves code size as well, it exposes more instruction
+level parallelism in the basic-block to which instructions are moved, and it
+enables other passes like loop invariant motion to remove redundancies. The cost
+model to drive the code motion is based on liveness analysis on \SSA{}
+representation such that the (virtual) register pressure does not increase
+resulting in 2\% fewer spills on the \SPEC{} benchmark suite. \GCM{} is an
 optimistic algorithm in the sense that it considers all congruent instructions
@@
 parallelism thereby enabling more \ooo{} execution on modern architectures.
+
+We have implemented the pass in \LLVM{}. The experimental results show
+performance gains of 8\% in 462.libquantum and 5.3\% in 471.omnetpp in \SPEC{}
+benchmark suite. It also reduces compilation time by reducing number of
+instructions to be processed by later compilation stages. \GCM{} enables more
+inlining and exposes more loop invariant code motion opportunities in the
+majority of benchmarks.
 \end{abstract}
@@
 % Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
-For evaluation of \gcm{}, we built \SPEC{} with the patch. All the experiments
-were conducted on an \xlinux{} machine and at -Ofast optimization level. Each
-benchmark was run three times and the best result was taken. We collected
-execution time and code-size which is listed in Table~\ref{tab:code-size}, other
-compiler statistics are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
+For evaluation of \gcm{}, we built \SPEC{} with and without the \GCM{}
+optimization. All the experiments were conducted on an \xlinux{} machine and at
+-Ofast optimization level with inlining and loop unrolling disabled; inliner and
+loop-unroller vary wildly with \GCM{} optimization so it was difficult to
+visualize the impact. Each benchmark was run three times and the best result was
+taken. We collected execution time and code-size which is listed in
+Table~\ref{tab:code-size}, other compiler statistics are listed in
+Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
 Figure~\ref{tab:code-motion-metric}.  We also measured compile time
-Table~\ref{tab:compile-time} to see the impact of \gcm{}.
+Table~\ref{tab:compile-time}. The benchmark 464.h264ref caused an internal
+compiler error with the trunk clang so it could not be reported.
 
@@
 
-Overall there was only a minor change in the performance numbers with a few improvements 
-(429.mcf, 458.sjeng, etc.) and a few regressions (447.dealII, 453.povray,
-etc.) The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
-first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
-local redundancies, \GCM{} would have to analyze less redundant instructions, also
-as inliner creates more redundancies, \GCM{} would have more opportunities to move
-congruent instructions out of branches.
-
-\begin{figure*}[h!]
-  \includegraphics[width=0.8\textwidth,height=8cm]{gcm-stats.pdf}
-  \vspace*{-1.5cm}
-\caption{\GCM{} stats on SPEC2006 at -Ofast. Loads are hoisted the most followed
-  by scalars, stores and calls.}
-\label{tab:code-motion-metric}
-\end{figure*}
-
-We see some nice gains in code size at -Os with \gcm{} (403.gcc, 447.dealII, and
-483.xalancbmk) as shown in Table~\ref{tab:code-size}.  The code size listed here
-is the text size delta (from linux size command) of the final executable for
-each benchmark.  On the other hand, some benchmarks like 473.astar increased in
-code size by almost 2\%. This is because once the code-size of a function
-decreases (due to \GCM{}) it becomes cheaper to inline. As we can see in
-Table~\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
-447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
-more functions inlined. Also, since the pass runs early, it affects many
+Overall, we can see some good improvements in 462.libquantum, 471omnetpp, etc.
+and some regressions in 429.mcf, 483.xalancbmk, etc. The \gcm{} pass was run once
+in the pass pipeline (at -Ofast), after local common subexpression elimination
+(EarlyCSE) pass. Because EarlyCSE removes local redundancies, \GCM{} would have
+to analyze less redundant instructions. We believe some performance regressions
+can be mitigated by running the \GCM{} few more times.
+
+We see some code size reduction in 400.perlbench, 403.gcc, and 453.povray as
+shown in Table~\ref{tab:code-size}.  The code size listed here is the text size
+delta (from size command in linux) of the final executable for each benchmark.  On
+the other hand 447.dealII increased in code size. This is because once the
+code-size of a function decreases (due to \GCM{}) it becomes cheaper to do some
+optimizations like inlining, loop unrolling, copy propagation, rematerialization
+etc. We tried to minimize the noise by disabling inliner, and loop unroller but
+still we can see some increase in the code size.  To see the impact of \GCM{} on
+inliner we ran \GCM{} with inliner enabled and collected the compile-time
+statistics produced by LLVM (See Table~\ref{tab:stats}) . Except for 447.dealII,
+450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got more
+functions inlined. Also, since the pass runs early, it affects many
 optimizations which rely on the number of instructions, length of the use-def
@@
 
-The code-size at -Ofast is very widespread with huge gains in 483.xalancbmk to
-serious regressions in 403.gcc. This is mostly because the compiler's focus is
-to improve performance at the cost of code-size at -Ofast, so passes like
-inlining, loop-unrolling, code-versioning are aggressive at -Ofast.
+\begin{table}[h!]
+  \begin{center}
+    \scalebox{0.8}{
+    \begin{tabular}{|l|r|r|r|r|}
+      \hline
+      &\multicolumn{2}{|c|}{Loop Invariant Motion}  & \multicolumn{2}{|c|}{Functions} \\\hline
+ Benchmarks       & Hoisted\% & Sunk\% &  Inlined\%   & Deleted\% \\\hline
+ 400.perlbench	  & 2.1 & -2.8 &  0.2 &  0.9  \\\hline
+ 401.bzip2	  & 14.4& 0.0  &  4.7 &  3.1  \\\hline
+ 403.gcc	  & 13.7& 27.3 &  0.9 &  0.1  \\\hline
+ 429.mcf	  & -7.3& 0.0  &  0.0 &  0.0  \\\hline
+ 433.milc	  & 10.2& 0.0  &  8.0 &  0.0  \\\hline
+ 444.namd	  & 0.1 & 0.0  &  0.8 &  -3.5 \\\hline
+ 445.gobmk	  & 0.6 & 5.6  &  5.4 &  1.7  \\\hline
+ 447.dealII	  & 62.2& -86.7& -0.2 &  -0.8 \\\hline
+ 450.soplex	  & 3.9 & -5.6 & -0.2 &  -0.4 \\\hline
+ 453.povray	  & -4.8& 39.1 &  0.2 &  -0.1 \\\hline
+ 456.hmmer	  & 1.0 & 166.7&  0.0 &  -1.9 \\\hline
+ 458.sjeng	  & 11.5& 0.0  &  6.3 &  0.0  \\\hline
+ 462.libquantum   & 6.6 & 0.0  &  0.0 &  0.0  \\\hline
+ 470.lbm	  & 0.0 & 0.0  &  0.0 &  0.0  \\\hline
+ 471.omnetpp	  & 17.0& -6.3 &  1.1 &  -0.4 \\\hline
+ 473.astar	  & 9.1 & 0.0  &  1.1 &  0.0  \\\hline
+ 482.sphinx3	  & 5.9 & -18.2&  -2.0&  0.0  \\\hline
+ 483.xalancbmk	  & 14.6& -3.6 &  -1.0&  -1.2 \\\hline
+    \end{tabular}}
+  \end{center}
+  \caption{Change in the static compile time metrics w.r.t. \GCM{} on \SPEC{} at
+    -Ofast. The \% columns show percentage increase in metric w.r.t. baseline,
+    i.e., without \GCM{}. \GCM{} improves LICM and inlining in general.}
+  \label{tab:stats}
+\end{table}
+
+Table~\ref{tab:stats}, which shows how \gcm{} impacted important compiler
+optimizations like inlining and loop invariant code motion (LICM). LICM removes
+loop invariant code out of the loop. It may hoist the code to the loop's
+preheader or sink the code to loop's post-dominator
+\cite{steven1997advanced}. As we can see number of instructions hoisted
+increased significantly in several benchmarks, although number of instructions
+sunk did not change by much.  Number of inlined functions also improved for most
+benchmarks. In general when more functions were inlined, more functions were
+deleted as in 401.bzip2, 445.gobmk.
 
@@
 462.libquantum	  & 118	        & 118	& 1.00 \\\hline
-464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
 470.lbm	          & 33	        & 33	& 1.00 \\\hline
@@
   462.libquantum  &   84	& 84	& 0.99 \\\hline
-  464.h264ref	  &   151	& 149	& 0.99 \\\hline
   470.lbm         &   71	& 71	& 1.00 \\\hline
@@
 executed during the compilation of each benchmark as output from valgrind
--{}-tool=cachegrind \cite{valgrind}.
-
-\begin{table}[h!]
-  \begin{center}
-    \scalebox{0.8}{
-    \begin{tabular}{|l|r|r|r|r|}
-      \hline
-      &\multicolumn{2}{|c|}{Loop Invariant Motion}  & \multicolumn{2}{|c|}{Functions} \\\hline
- Benchmarks       & Hoisted \% & Sunk \% &  Inlined\%   & Deleted\% \\\hline
- 400.perlbench	  & 2.1 & -2.8 &  0.2 &  0.9  \\\hline
- 401.bzip2	  & 14.4& 0.0  &  4.7 &  3.1  \\\hline
- 403.gcc	  & 13.7& 27.3 &  0.9 &  0.1  \\\hline
- 429.mcf	  & -7.3& 0.0  &  0.0 &  0.0  \\\hline
- 433.milc	  & 10.2& 0.0  &  8.0 &  0.0  \\\hline
- 444.namd	  & 0.1 & 0.0  &  0.8 &  -3.5 \\\hline
- 445.gobmk	  & 0.6 & 5.6  &  5.4 &  1.7  \\\hline
- 447.dealII	  & 62.2& -86.7& -0.2 &  -0.8 \\\hline
- 450.soplex	  & 3.9 & -5.6 & -0.2 &  -0.4 \\\hline
- 453.povray	  & -4.8& 39.1 &  0.2 &  -0.1 \\\hline
- 456.hmmer	  & 1.0 & 166.7&  0.0 &  -1.9 \\\hline
- 458.sjeng	  & 11.5& 0.0  &  6.3 &  0.0  \\\hline
- 462.libquantum   & 6.6 & 0.0  &  0.0 &  0.0  \\\hline
- 464.h264ref	  & 0.5 & 0.0  &  13.0&  0.0  \\\hline
- 470.lbm	  & 0.0 & 0.0  &  0.0 &  0.0  \\\hline
- 471.omnetpp	  & 17.0& -6.3 &  1.1 &  -0.4 \\\hline
- 473.astar	  & 9.1 & 0.0  &  1.1 &  0.0  \\\hline
- 482.sphinx3	  & 5.9 & -18.2&  -2.0&  0.0  \\\hline
- 483.xalancbmk	  & 14.6& -3.6 &  -1.0&  -1.2 \\\hline
-    \end{tabular}}
-  \end{center}
-  \caption{Change in the static compile time metrics w.r.t. \GCM{} on
-    \SPEC{} at -Ofast. The \% columns show percentage increase in metric
-    w.r.t. baseline. \GCM{} improves LICM and inlining in general.}
-  \label{tab:stats}
-\end{table}
-
-Some useful static metrics are presented in Table~\ref{tab:stats}, which shows
-how \gcm{} impacted important compiler optimizations like inlining and
-LICM. LICM removes loop invariant code out of the loop. It may hoist the code to
-the loop's preheader or sink the code to loop's post-dominator
-\cite{steven1997advanced}. As we can see number of instructions hoisted
-increased significantly in several benchmarks, although number of instructions
-sunk did not change by much.  Number of inlined functions also improved for most
-benchmarks. In general when more functions were inlined, more functions were
-deleted as in 400.perlbench, 445.gobmk.
+-{}-tool=cachegrind \cite{valgrind}. We used valgrind because it was a simple
+and reliable way to show the impact of our optimization.
+
+\begin{figure*}[h!]
+  \includegraphics[width=0.8\textwidth,height=8cm]{gcm-stats.pdf}
+  \vspace*{-1.5cm}
+\caption{\GCM{} stats on SPEC2006 at -Ofast. Loads are hoisted the most followed
+  by scalars, stores and calls.}
+\label{tab:code-motion-metric}
+\end{figure*}
 
@@
 those loads and the number of hoisted loads may change should the \GCM{} pass be
-scheduled later. We can see a few sunk instructions even if \LLVM{} has a code
-sinking and hoisting transforms in the SimplifyCFG pass which runs before
-\GCM{}: this is because sinking and hoisting in SimplifyCFG have some rather
-severe limitations to make the analysis and code transform very fast in order to
-be able to run the pass several times.  Another reason is that instructions for
-which dependencies are not directly available in the successor
-(Section~\ref{subsec:legality}) is not sunk for now because that would
-require a look ahead on the ability to sink the dependencies. We plan to
-implement this in near future.
+scheduled later. We can see only a few sunk instructions because hoisting is
+performed before sinking (Section~\ref{subsec:optimistic}) and some of the
+candidates which can either be hoisted or sunk got hoisted.
 
@@
 \label{sec:future-work}
-
-We have presented the \gcm{} of congruent computations in SSA form. We saw that
-it improves inlining and LICM opportunities, reduces register spills, it also
-improves performance in some cases. To preserve performance and not hoist/sink
-too much, we have implemented a register pressure aware cost model as described
-in Section~\ref{subsec:cost-models}. A part of our implementation is merged in
-LLVM trunk as GVNHoist.cpp and the \GCM{} implementation is under review.
-
-This pass does not hoist fully redundant expressions because they are already
-handled by GVN-PRE. It is advisable to run gvn-hoist before and after gvn-pre
-because gvn-pre creates opportunities for more instructions to be hoisted.
-
-We would like to integrate \GCM{} with \LLVM{'}s new GVN-PRE implementation such
-that more congruent instructions can be scheduled. This would also expose more
-redundancies to the GVN-PRE. In general it is a good idea to start with lower
-ranked expressions first such that maximum hoisting can happen in one iteration,
-however, current implementation does not rank the expressions and iteratively
-finds a fixed point when no more candidates are available. Even this
-implementation converges quickly and no compile time regression have been
-observed. This can be improved by ranking the computations
-\cite{rosen1988global}. Also, since \GCM{} runs very early in the pass pipeline,
-it will be good to evaluate the code size/performance impact when it is run in
-sync with \GVN{}-\PRE{} just like \GCC{} does. With the implementation of \GCM{}
-in \LLVM{}, the passes which rely on the code-size or instruction-counts to make
-optimization decisions need to be revisited for example, the inliner.
+We have presented a fast algorithm to perform \gcm{} of congruent computations
+in SSA form. We introduced data structures like $\chi$ nodes, $\Phi$ nodes which
+helps exploit flow of values in sparse representation.  Experimental results
+show that it is a useful performance and code-size optimization. It also enables
+inlining and LICM opportunities, reduces register spill. To preserve performance
+and not hoist/sink too much, we have implemented a register pressure aware cost
+model as described in Section~\ref{subsec:cost-models}. A part of our
+implementation is merged in LLVM trunk as GVNHoist.cpp and the \GCM{}
+implementation is under review. This pass does not hoist fully redundant
+expressions because they are already handled by a later partial redundancy
+elimination pass.
+
+We would like to explore other places in the optimization pipeline where \GCM{}
+could improve performance. It may be useful to integrate \GCM{} with \LLVM{'}s
+new GVN-PRE implementation such that more congruent instructions can be
+scheduled. This would also expose more redundancies to the GVN-PRE. Also, since
+\GCM{} runs very early in the pass pipeline, it will be good to evaluate the
+code size/performance impact when it is run in sync with \GVN{}-\PRE{} just like
+\GCC{} does. With the implementation of \GCM{} in \LLVM{}, the passes which rely
+on the code-size or instruction-counts to make optimization decisions need to be
+revisited for example, the inliner, the loop unroller etc.
 

### 2017-11-13 cgo/paper.tex
move sinking in legality

@@
 level parallelism in the basic-block to which instructions are moved, and it
-enables other passes like loop invariant motion to remove redundancies. The cost
-model to drive the code motion is based on liveness analysis on \SSA{}
-representation such that the (virtual) register pressure does not increase
-resulting in 2\% fewer spills on the \SPEC{} benchmark suite. \GCM{} is an
-optimistic algorithm in the sense that it considers all congruent instructions
-in a function as potential candidates. We also formalize why register pressure
-reduces as a result of \gcm{}, and how \gcm{} increases instruction level
-parallelism thereby enabling more \ooo{} execution on modern architectures.
+enables other compiler optimizations like loop invariant motion to remove
+redundancies. The cost model to drive the code motion is based on liveness
+analysis on \SSA{} representation such that the (virtual) register pressure does
+not increase resulting in 2\% fewer spills on the \SPEC{} benchmark
+suite. \GCM{} is an optimistic algorithm in the sense that it considers all
+congruent instructions in a function as potential candidates. We also formalize
+why register pressure reduces as a result of \gcm{}, and how \gcm{} increases
+instruction level parallelism thereby enabling more \ooo{} execution on modern
+architectures.
 
 We have implemented the pass in \LLVM{}. The experimental results show
-performance gains of 8\% in 462.libquantum and 5.3\% in 471.omnetpp in \SPEC{}
-benchmark suite. It also reduces compilation time by reducing number of
+performance gains of 8\% in 462.libquantum and 5.3\% in 471.omnetpp from the
+\SPEC{} benchmark suite. It also reduces compilation time by reducing number of
 instructions to be processed by later compilation stages. \GCM{} enables more
@@
 to global code scheduling \cite{dragonbook,click1995global} in the sense that it
-will only move computations. Code hoisting, for example, can reduce the critical
-path length of execution in \ooo{} machines. As more instructions are available
-at the hoisting point, the hardware has more independent instructions to
-reorder. The following reduced example illustrates how hoisting can improve
+will only move dynamic computations. Code hoisting, for example, can reduce the
+critical path length of execution in \ooo{} machines. As more instructions are
+available at the hoisting point, the hardware has more independent instructions
+to reorder. The following reduced example illustrates how hoisting can improve
 performance by exposing more ILP.
@@
 
-There has been a lot of work both in industry and academia related to code
-hoisting/sinking, and in general global scheduling \cite{click1995global}. Some
-relate code-hoisting to code-size optimization \cite{rosen1988global} and many
-\cite{barany2013, shobaki2013} use global scheduling to improve
-performance. Most of the recent work on global scheduling are done using integer
-linear programming (ILP) which results in prohibitively high compile time which
-is not suitable for industrial compilers. To the best of our knowledge we have
-not found any reference which explored \gcm{} with a cost-model to reduce
-register-pressure. A part of our implementation (aggressive code hoisting) is
-already merged in \LLVM{} trunk. The main contributions of this paper are:
-\begin{itemize}[leftmargin=*,topsep=0pt]
-\item an optimistic fast algorithm to schedule congruent instructions;
-\item introducing an analogue of $\phi$ nodes on the inverse CFG which allows
-  faster tracking of anticipable instructions;
-\item a cost model to reduce interference and hence, reduce spills;
-\item experimental evaluation of our implementation in \LLVM{}.
-\end{itemize}
-
-\section{Related Work}
-
-There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
-\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
-a more powerful code hoist transform.  The current \LLVM{} implementation of
-code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
-basic-blocks: the instructions of two sibling basic-blocks are read at the same
-time, and all the instructions of the blocks are hoisted to the common parent
-block as long as the compiler is able to prove that the instructions are
-equivalent.  This implementation does not allow for an easy extension: first in
-terms of compilation time overhead the implementation is quadratic in number of
-instructions to bisimulate and second, the equivalence of instructions is
-computed by comparing the operands which is neither general nor scalable.
-
-\citet{click1995global} describes aggressive \gcm{} in sparse representation. It
-first schedules all the instructions as early as possible. This results in very
-long live ranges which is mitigated by again scheduling all the instructions as
-late as possible. They report a speedup of as high as $23\%$ but there is no
-cost model to limit amount of scheduling and could often introduce redundant
-computations. It is also not mentioned how safety is ensured while hoisting an
-instruction when it is not anticipable. The legality checks are only based on
-barriers (Section~\ref{subsec:barriers}) and availability of used operands. Even
-introducing a simple integer addition can also result in undefined behavior
-\cite{undef}.  Except from this paper we found no reference of global scheduling
-in sparse representation.
-
-\citet{dhamdhere1988fast} and \citet{steven1997advanced} mention
-code hoisting in a data-flow framework. A list of Very Busy Expressions (VBE)
-computed which are hoisted in a basic-block where the expression is
-anticipable (all the operands are available.) This algorithm would hoist as far
-as possible without regarding the impact on register pressure and as such a cost
-model will be required.  Also the description of VBE is based on the classic
-data-flow model and an adaptation to a sparse \SSA{} representation is required.
-
-\citet{rosen1988global} also briefly discuss hoisting computations with
-identical value numbers from immediate successors. Their algorithm iterates on
-computations of same rank and move the code with identical computations from the
-sibling branch to a common dominator if they are very busy
-\cite{steven1997advanced}. However, the cost-model to mitigate register pressure
-is missing, also, there is no mention of sinking congruent instructions.
-
-\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
-part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
-expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions
-(VBEs) at basic-block B, i.e., values computed on all paths from B to exit and
-AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
-VBEs to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will be
-computed on every path from the basic block B to exit, and can be computed in B.
-It uses AVAIL\_OUT[B] to subtract out those values already being computed.  The
-cost function is: for each hoist candidate, if all successors of B are dominated
-by B, then we know insertion into B will eliminate all the remaining
-computations.  It then checks to see if at least one successor of B has the
-value available.  This avoids hoisting it way up the chain to ANTIC\_IN[B].  It
-also checks to ensure that B has multiple successors, since hoisting in a
-straight line is pointless.  The algorithm continues top down the dominator
-tree, working in tandem with \PRE{} until no more hoisting is possible.  One
-advantage of \GCC{} implementation is that it works in sync with the
-\GVN{}-\PRE{} such that when new hoisting opportunities are created by
-\GVN{}-\PRE{}, code-hoisting will hoist them.
-
-\citet{barany2013} presented a global scheduler with ILP formulation with
-a goal to minimize register pressure. The results they got were not very
-promising. It may be because they only used the scheduler for smaller functions
-($<1000$ instructions); also, they compiled the benchmarks for ARM-Cortex which
-is more resilient to register pressure because it has more registers compared to
-X86, for example.
-
-\citet{shobaki2013} also presented a combinatorial global scheduler with
-reasonable performance improvements. It is possible that both Barany and
-Shobaki's implementation will have similar results when compiled for same target
-architecture. Also, both suffer from the same problem, although Shobaki not so
-much, of large compile times which would not suit in industrial compilers like
-\GCC{} and \LLVM{}. With the algorithm described in the next section, we got a
-reduction in register spills, improved inliner heuristics, improved compile
-time, and show performance improvements on \SPEC{} benchmarks.
-
-% TODO: References
-% Sink instructions
-% Ranking of instructions from Zadeck
-% Liveness analysis from Ramakrishna
-% Experimental results: Register spills, compile time, run time.
-
-\section{Global code motion}
-
 The algorithm for \GCM{} uses several common representations of the
@@
 
-The \GCM{} pass can be broadly divided into the following steps that we will
-describe in the rest of this section:
+The \GCM{} pass can be broadly divided into the following:
 \begin{itemize}[leftmargin=*,topsep=0pt]
+\item factor the reachability of expressions (Section~\ref{sec:chi}).
 \item find candidates (congruent instructions) suitable for \gcm{}
-  (Section~\ref{subsec:finding-candidates}),
+  (Section~\ref{subsec:optimistic}),
 \item compute a point in the program where it is both legal
@@
 
-\section{Factoring reachability of expressions}
+There has been a lot of work both in industry and academia related to code
+hoisting/sinking, and in general global scheduling \cite{click1995global}. Some
+relate code-hoisting to code-size optimization \cite{rosen1988global} and many
+\cite{barany2013, shobaki2013} use global scheduling to improve
+performance. Most of the recent work on global scheduling are done using integer
+linear programming (ILP) which results in prohibitively high compile time which
+is not suitable for industrial compilers. To the best of our knowledge we have
+not found any reference which explored \gcm{} with a cost-model to reduce
+register-pressure. A part of our implementation (aggressive code hoisting) is
+already merged in \LLVM{} trunk. The main contributions of this paper are:
+\begin{itemize}[leftmargin=*,topsep=0pt]
+\item an optimistic fast algorithm to schedule congruent instructions;
+\item introducing an analogue of $\phi$ nodes which allows faster tracking of
+  anticipable instructions;
+\item a cost model to reduce register spills;
+\item experimental evaluation of our implementation in \LLVM{}.
+\end{itemize}
+
+\section{Related Work}
+There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
+\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
+a more powerful code hoist transform.  The current \LLVM{} implementation of
+code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
+basic-blocks: the instructions of two sibling basic-blocks are read at the same
+time, and all the instructions of the blocks are hoisted to the common parent
+block as long as the compiler is able to prove that the instructions are
+equivalent.  This implementation does not allow for an easy extension: first in
+terms of compilation time overhead the implementation is quadratic in number of
+instructions to bisimulate and second, the equivalence of instructions is
+computed by comparing the operands which is neither general nor scalable.
+
+\citet{click1995global} describes aggressive \gcm{} in sparse representation. It
+first schedules all the instructions as early as possible. This results in very
+long live ranges which is mitigated by again scheduling all the instructions as
+late as possible. They report a speedup of as high as $23\%$ but there is no
+cost model to limit amount of scheduling and could often introduce redundant
+computations. It is also not mentioned how safety is ensured while hoisting an
+instruction when it is not anticipable. The legality checks are only based on
+barriers (Section~\ref{subsec:barriers}) and availability of used operands. Even
+introducing a simple integer addition can also result in undefined behavior
+\cite{undef}.  Except from this paper we found no reference of global scheduling
+in sparse representation.
+
+\citet{dhamdhere1988fast} and \citet{steven1997advanced} mention
+code hoisting in a data-flow framework. A list of Very Busy Expressions (VBE)
+computed which are hoisted in a basic-block where the expression is
+anticipable (all the operands are available.) This algorithm would hoist as far
+as possible without regarding the impact on register pressure and as such a cost
+model will be required.  Also the description of VBE is based on the classic
+data-flow model and an adaptation to a sparse \SSA{} representation is required.
+
+\citet{rosen1988global} also briefly discuss hoisting computations with
+identical value numbers from immediate successors. Their algorithm iterates on
+computations of same rank and move the code with identical computations from the
+sibling branch to a common dominator if they are very busy
+\cite{steven1997advanced}. However, the cost-model to mitigate register pressure
+is missing, also, there is no mention of sinking congruent instructions.
+
+\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
+part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
+expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions
+(VBEs) at basic-block B, i.e., values computed on all paths from B to exit and
+AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
+VBEs to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will be
+computed on every path from the basic block B to exit, and can be computed in B.
+It uses AVAIL\_OUT[B] to subtract out those values already being computed.  The
+cost function is: for each hoist candidate, if all successors of B are dominated
+by B, then we know insertion into B will eliminate all the remaining
+computations.  It then checks to see if at least one successor of B has the
+value available.  This avoids hoisting it way up the chain to ANTIC\_IN[B].  It
+also checks to ensure that B has multiple successors, since hoisting in a
+straight line is pointless.  The algorithm continues top down the dominator
+tree, working in tandem with \PRE{} until no more hoisting is possible.  One
+advantage of \GCC{} implementation is that it works in sync with the
+\GVN{}-\PRE{} such that when new hoisting opportunities are created by
+\GVN{}-\PRE{}, code-hoisting will hoist them.
+
+\citet{barany2013} presented a global scheduler with ILP formulation with
+a goal to minimize register pressure. The results they got were not very
+promising. It may be because they only used the scheduler for smaller functions
+($<1000$ instructions); also, they compiled the benchmarks for ARM-Cortex which
+is more resilient to register pressure because it has more registers compared to
+X86, for example.
+
+\citet{shobaki2013} also presented a combinatorial global scheduler with
+reasonable performance improvements. It is possible that both Barany and
+Shobaki's implementation will have similar results when compiled for same target
+architecture. Also, both suffer from the same problem, although Shobaki not so
+much, of large compile times which would not suit in industrial compilers like
+\GCC{} and \LLVM{}. With the algorithm described in the next section, we got a
+reduction in register spills, improved inliner heuristics, improved compile
+time, and show performance improvements on \SPEC{} benchmarks.
+
+% TODO: References
+% Sink instructions
+% Ranking of instructions from Zadeck
+% Liveness analysis from Ramakrishna
+% Experimental results: Register spills, compile time, run time.
+\section{Global Code Motion of congruent computations}
+\label{sec:gcm}
+The \GCM{} uses the following data structures ($\chi$ and $\Phi$ nodes), that
+are introduced below. They allow us to factor the reachability of values in an
+efficient way. Together with liveness analysis using the DJ-Graph \cite{das2012}
+and Memory SSA, the entire implementation of \gcm{} only uses sparse
+representations for efficient query and update.
+
+\subsection{Factoring reachability of expressions}
+\label{sec:chi}
 Our algorithm factors out the reachability of values such that multiple queries
@@
 frontiers (DF) and the post-dominance frontiers (PDF) of the CFG. These points,
-being structural properties of a graph, do not change during code-motion. A
+being structural properties of a graph, do not change during the code-motion. A
 post-dominance frontier is the basic block where anticipability (ANTIC) of an
@@
 out of a basic block with more than one successor. This approach allows us to
-hoist instructions to a basic block with more than two successors. The $\chi$
-node in a basic block $B$ has CHIArgs with three entries: a) the instruction (I)
-CHIArg refers to, b) the basic block $Dest$ which is post-dominated by the basic
-block containing I, and c) the GVN of I.
+hoist instructions to a basic block with multiple successors. The $\chi$ node in
+a basic block $B$ has CHIArgs with three entries: a) the instruction (I) CHIArg
+refers to, b) the basic block $Dest$ which is post-dominated by the basic block
+containing I, and c) the GVN of I.
 
@@
 
-For that, we need to prove that on all execution paths from the initial position
-of an instruction to its destination, the side effects appear in the same
-order. In general, we cannot introduce a new computation along any path of
-execution without checking for undefined behavior \cite{undef}. It is also
-necessary to check if there are indirect branch targets, e.g., landing pad, case
-statements, goto labels, etc., along the path because it becomes difficult to
-prove safety checks in those cases.
+We need to prove that on all execution paths from the initial position of an
+instruction to its destination, the side effects appear in the same order. In
+general, we cannot introduce a new computation along any path of execution
+without checking for undefined behavior \cite{undef}. It is also necessary to
+check if there are indirect branch targets, e.g., landing pad, case statements,
+goto labels, etc., along the path because it becomes difficult to prove safety
+checks in those cases.
 
@@
 of operands. For sinking the scalars we need to make sure that there are no uses
-of the instruction along the path, For instructions with memory references like
+of the instruction along the path. For instructions with memory references like
 loads, stores and calls, Memory SSA allows efficient way of iterating over the
@@
 
+For sinking, higher ranked expressions would be sunk first. And it would
+be illegal to sink higher ranked identical expressions if they are not fully
+anticipable in the common post-dominator. For example:
+
+\begin{lstlisting}
+B0: i0 = load B
+B1: i1 = load A
+    c1 = i1 + 10
+    d1 = i0 + 20
+    goto B3
+B2: i2 = load A
+    c2 = i2 + 10
+    d2 = i0 + 20
+    goto B3
+
+B3: PHIc1, c2)
+    PHI(d1, d2)
+\end{lstlisting}
+
+In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
+(c1, c2) depend on i1 and i2 respectively which are also in their original basic
+blocks, c1 and c2 are not fully anticipable in B3. So without knowing the
+ability to sink `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the
+other hand (d1, d2) can safely be sunk because their operands are readily
+available at the sink point, i.e., B3. It should also be noted that, just
+because the expressions are identical and operands are available, it still
+requires a unique post-dominating $\Phi$ to use the exact same values to be
+legally sinkable.
+
 \subsection{Profitability check (Cost models)}
@@
 size. The current implementation makes effort to not regress in performance but
-at the same time to reduce code size as much as possible. Sometimes, hoisting is
-not upward-safe \cite{click1995global}, e.g., if the expressions are in a
-landing pad, sinking of those expressions may reduce code size. Following cost
-models are implemented:
+reduce register spills. We implement a cost model to reduce register pressure.
 
@@
 regressions with this cost model even if LLVM has a live-range splitting
-\cite{cooper1998live} optimization.
-
-Sinking may also reduce the register pressure in some cases, e.g., when the use
-operands are not kills. For sinking, higher ranked expressions would be sunk
-first. And it would be illegal to sink higher ranked identical expressions if
-they are not fully anticipable in the common post-dominator. For example:
-
-\begin{lstlisting}
-B0: i0 = load B
-B1: i1 = load A
-    c1 = i1 + 10
-    d1 = i0 + 20
-    goto B3
-B2: i2 = load A
-    c2 = i2 + 10
-    d2 = i0 + 20
-    goto B3
-
-B3: PHIc1, c2)
-    PHI(d1, d2)
-\end{lstlisting}
-
-In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
-(c1, c2) depend on i1 and i2 respectively which are also in their original basic
-blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
-of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
-(d1, d2) can safely be sunk because their operands are readily available at the
-sink point, i.e., B3. It should also be noted that, just because the expressions
-are identical and operands are available, it still requires a unique
-post-dominating PHI to use the exact same values to be legally sinkable.
+\cite{cooper1998live} optimization. Sinking may also reduce the register
+pressure when the use operands are not kills.
 

### 2017-11-13 cgo/paper.tex
reformat

@@
 
-  for each CHI-nodes in the CFG:
-    for each CHI-arg in CHI-node:
-      Remove CHI-arg if it fails legality checks
-
-  for each CHI-nodes in the CFG:
-    if CHI  does not have missing entries:
+  for each $\chi$-nodes in the CFG:
+    for each $\chi$-arg in $\chi$-node:
+      Remove $\chi$-arg if it fails legality checks
+
+  for each $\chi$-nodes in the CFG:
+    if $\chi$ does not have missing entries:
       Perform code hoisting
@@
       Find its dominance-frontier (DF)
-      Insert the PHI-node at DF
-
-  for each PHI-nodes in the CFG:
-    for each PHI-arg in PHI-node:
-      Remove PHI-arg if it fails legality checks
-
-  for each PHI-nodes in the CFG:
-    if PHI does not have missing entries:
+      Insert the $\Phi$-node at DF
+
+  for each $\Phi$-nodes in the CFG:
+    for each $\Phi$-arg in PHI-node:
+      Remove $\Phi$-arg if it fails legality checks
+
+  for each $\Phi$-nodes in the CFG:
+    if $\Phi$ does not have missing entries:
       Perform code sinking
@@
 Table~\ref{tab:code-size}, other compiler statistics are listed in
-Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
+Table~\ref{tab:stats}, Table~\ref{tab:spills}, and
 Figure~\ref{tab:code-motion-metric}.  We also measured compile time
@@
 etc. We tried to minimize the noise by disabling inliner, and loop unroller but
-still we can see some increase in the code size.  To see the impact of \GCM{} on
-inliner we ran \GCM{} with inliner enabled and collected the compile-time
-statistics produced by LLVM (See Table~\ref{tab:stats}) . Except for 447.dealII,
-450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got more
-functions inlined. Also, since the pass runs early, it affects many
-optimizations which rely on the number of instructions, length of the use-def
-chain, and other metrics.
+still we can see some increase in the code size.
+
+To see the impact of \GCM{} on inliner we ran \GCM{} with inliner enabled and
+collected the compile-time statistics produced by LLVM (See
+Table~\ref{tab:stats}) . Except for 447.dealII, 450.soplex, 482.sphinx3 and
+483.xalancbmk all other benchmarks got more functions inlined. Also, since the
+pass runs early, it affects many optimizations which rely on the number of
+instructions, length of the use-def chain, and other metrics.
 
@@
 implementation is under review. This pass does not hoist fully redundant
-expressions because they are already handled by a later partial redundancy
-elimination pass.
+expressions because they are already removed by a later partial redundancy
+elimination pass in LLVM.
 

### 2017-11-13 cgo/paper.tex
call and fast algo

@@
 of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
-and brought these techniques in the main stream: \SSA{} based \PRE{} is
-available in many compilers like \GCC{} and \LLVM{}.
+and brought these techniques in the main stream: \SSA{} based \PRE{} has been
+implemented in many optimizing compilers like \GCC{} and \LLVM{}.
 
@@
 a basic-block. Essentially, any instruction that cannot guarantee progress is
-marked as a barrier. In the absence of context, as in our current
-implementation, some instructions which might be safe are still classified as
-barriers, e.g., calls with missing attributes.
+marked as a barrier. A barrier will prevent any instruction occurring lexically
+below it from getting hoisted. On the other hand, a sink barrier will prevent
+any instruction above it from getting sunk. If there are no barriers in a basic
+block, any instruction satisfying other legality checks can be sunk.
 
 \begin{lstlisting}
-// Compute barriers for both hoistable and sinkable
-// candidates.
+// Compute both hoist and sink barriers
 void computeBarriers()
@@
     // Find the last barrier below which instructions
-    // can be sunk. If there was no barrier in B, any
-    // instruction satisfying other legality checks
     // can be sunk.
@@
 
-Computing barriers allows for efficient removal of non-movable candidates, this
-makes the legality check converge faster.
+In the absence of context, as in our current implementation, some instructions
+which might be safe are still classified as barriers, e.g., calls with missing
+attributes.  Computing barriers allows for efficient removal of non-movable
+candidates, this makes the legality check converge faster.
 
@@
 check if there are indirect branch targets, e.g., landing pad, case statements,
-goto labels, etc., along the path because it becomes difficult to prove safety
-checks in those cases.
+or other forms of goto labels along the path because it becomes difficult to
+prove safety checks in those cases.
 
@@
 \begin{lstlisting}
-B0: b = m
+B0:
+    b = m
     c = n
     if p is true then goto B1 else goto B2
-B1: a0 = b<kill> + c<kill>
+B1:
+    a0 = b<kill> + c<kill>
     goto B3
-B2: a1 = b<kill> + c<kill>
+B2:
+    a1 = b<kill> + c<kill>
     goto B3
@@
 \label{cost:across-calls}
-Even hoisting scalars across calls is tricky because it can increase the number
-of spills. During the frame lowering of calls, the argument registers (also
-known as the caller saved registers) are saved because they might be modified by
-the callee and after the call they are restored. Due to this the register
-pressure is high because the number of available registers are reduced by the
-number of caller saved registers. In that situation a computation that increases
-register pressure is not profitable to hoist.
+Call expressions are difficult to analyze in an intraprocedural optimization
+pass like \GCM{}. Even hoisting scalars across calls is tricky because it can
+increase the number of spills. During the register allocation of calls
+expressions, the argument registers (also known as the caller saved registers)
+are saved because they might be modified by the callee and after the call they
+are restored. Due to this the register pressure is high because the number of
+available registers are reduced by the number of caller saved registers. In that
+situation a computation that reduces register pressure may be beneficial to
+hoist. Similarly, a computation that introduces a register (without killing any
+other operand) before a call may be beneficial to sink.
 
@@
 Once all the legality and profitability checks are satisfied for a set of
-congruent instructions, they are suitable candidates for code motion. A
-copy of the computation is inserted at the hoisting/sinking point along with any
+congruent instructions, they are suitable candidates for code motion. A copy of
+the computation is inserted at the hoisting/sinking point along with any
 instructions which needed to be rematerialized. Thereafter, all the computations
-made redundant by the new copy are removed.
+made redundant by the new copy are removed and affected data structuers are
+updated.
 
@@
 \GCM{} basically consists of two parts, i.e., hoisting and sinking. This
-implementation only moves congruent instructions. An immediate guarantee of this
-approach is gain in code size (the final executable may be of larger size
-because of more inlining). If the dependency of a hoistable candidate is in the
-same basic-block as the candidate, then the dependency must also be hoistable,
-otherwise hoisting will be illegal or would require a complicated code
-generation to make it legal. The current algorithm discards cases if the
-dependency is neither hoistable nor rematerializable. We collect the \GVN{} of
-all the instructions in the function and iterate on the list of instructions
-having identical \GVN{}s. The algorithm \emph{doGCM} prefers hoisting
-(\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).
+implementation only moves congruent instructions which makes it easier to ensure
+that we do not introduce a redundant computation along any path of execution. If
+the dependency of a hoistable candidate is in the same basic-block as the
+candidate, then the dependency must also be hoistable, otherwise hoisting will
+be illegal or would require a complicated code generation to make it legal. The
+current algorithm discards cases if the dependency is neither hoistable nor
+rematerializable. We collect the \GVN{} of all the instructions in the function
+and iterate on the list of instructions having identical \GVN{}s. The algorithm
+\emph{doGCM} prefers hoisting (\emph{doCodeHoisting}) to sinking
+(\emph{doCodeSinking}).
 

### 2017-12-21 cgo/cc-2018-reviews-rebuttal.md
Review Rebuttal

Dear Aditya Kumar,

Thank you again for your submission to CC 2018. You are now
given an opportunity to respond to the comments made to your submission.
The author response period is 20 -- 21 December, 2017 (midnight, AOE).

During this time, you will have access to the reviews for your paper.

Please note that every paper received at least three reviews and
these reviews are completed by its assigned PC members, without any
coordination between them. Thus, there may be inconsistencies.
Reviews and scores are likely going to be updated, to take into account
the discussion among the reviewers and your rebuttal. So please do not
assume that a paper with all positive reviews will necessarily get accepted
or a paper with all negative reviews will necessarily get rejected.

You have 800 words for your rebuttal. The author response is not
compulsory. However, if you feel that your paper may get accepted,
I strongly encourage you to use this opportunity to answer the reviewers'
questions.  You may also use the space to focus on any factual errors
in the reviews. Please try to be as concise and to the point as possible
and please do not provide new research results or reformulate the
presentation.

Please also try to be polite and constructive. Please be very careful not
to reveal your identity in your response. If your response breaks the
double-blind reviewing policy, then your paper may be rejected outright.

The program committee will read your response carefully and take this
information into account during the PC discussions.

The reviews on your paper are attached to this email.

Best wishes,

       Title: A fast algorithm for global code motion of congruent
              instructions
  Paper site: https://cc18.hotcrp.com/paper/66

Use the link below to sign in to the site.

https://cc18.hotcrp.com/?email=hiraditya%40gmail.com

--- Jingling
Program Chair for CC 2018

Review #66A
===========================================================================

Overall merit
-------------
3. Weak accept

Paper summary
-------------
This work describes the design and implementation of a global code-motion optimization pass in LLVM. It provides a detailed overview of the algorithm, analyses it relies on, data-structures used, and cost-model considerations. Evaluation of several different metrics are presented on the Spec2006 benchmark suite, and clear reduction in number of spills is shown, with some other effects on inlining and LICM, while incurring no increase in compilation time.

Strengths:
        - Provides a practical report on an actual implementation in an industry-strength compiler. The work is also (in the process of being) contributed to open-source.

        - The paper is well written. In particular I thought the motivation (benefits of GCM) is set up very clearly and convincingly, prior-art is discussed in detail, the data-structures are presented clearly, and the overall algorithm is easy to follow with the pseudo-code provided.

        - Experimental results on the entire SPEC2006 benchmark suite are presented, and several metrics are examined (code size, compile time, runtime) including an account of individual effects on number of spills, number of instructions hoisted/sunk, and effects on LICM and inlining.

Weaknesses:
        - Overall the experimental results are not entirely conclusive. There is a clear 2% improvement in number of spills (Table 3), but otherwise the bottom line picture is not very clear (e.g. as reflected in Tables 1,2).

        - Performance is shown only with inliner and loop-unroller disabled. It's ok that results are shown in a setup that leaves out some effects that make it difficult to isolate the "pure" impact of GCM, but results on a more realistic setup are still also relevant to get an idea about.
(I understand though the difficulty to draw conclusions and make an informed analysis with various compilation passes being affected in various ways, and I appreciate the effort to compensate for that with the different metrics that Section 5 explores.)

        - I would have liked to see more analysis of the experimental results, and more discussion of the tradeoffs of various decisions in the implementation at hand (some examples below).

Comments for author
-------------------
- I would have liked to see more analysis of the experimental results:
        ○ Regarding Table 1: Can anything be said about the benchmarks with the larger impact? Were the effects scattered all over these benchmarks or occurred in a few hotspots? Can you say something on what was the primary factor for the performance effect (e.g. from the benefits mentioned in the introduction: reduced register pressure / increased ILP / reduced branch mis-peculation penalty? Were there any effects on vectorization?). Also, how significant are the absolute code-size effects?
        ○ Regarding Table2: can you say anything about the situations where GCM resulted in less inlining/less instructions deleted? About the LICM effects: were the LICM improvements indeed due to conditional code simplified by GCM?
        ○ Regarding Table 3: Can anything be said about the cases that exhibited significant register-spill increase? Can anything be done to further reduce it? How would the picture look like if the more aggressive GCM was used (that is, with the compiler flag that makes it not  consider register pressure)?
        ○ Regarding Figure 4: Can you tell how many of these hoists would have happened anyhow, without GCM in the picture?
        ○ Regarding Table 4: How does the compile-time increase look like in terms of actual compilation time?

- Other comments on the experimental section:
        ○ What was the effect on the memory footprint of the compiler? (how much extra storage was required)?
        ○ Is it possible to show the impact on other platforms? (what would be the benefit on a platform that is less sensitive to register pressure?)
        ○ Is it possible to compare some of the metrics to GCC's GCM implementation? Can anything be learned from that on the tradeoffs between the heuristics the two implementations use?
        ○ No overall (average) impact is indicated over all the benchmarks (except for Table 3).

- Were any other heuristics considered along the way? For example, how about giving hoisting preference to higher-latency instructions?

- You mention that scheduling GCM so early in the compilation pass had significant implications. Did you consider scheduling the pass at a different (later) place in the compiler?

Typos:

Page4: trackig -> tracking
Page5: PHIc1, c2) -> PHI(c1, c2)
Page6: explains how GCM of can reduce -> remove 'of'?
Page6: allocation of calls expressions -> call expressions
Page6: for each x-nodes in the CFG -> node

* * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *

Review #66B
===========================================================================

Overall merit
-------------
3. Weak accept

Paper summary
-------------
The paper "A fast algorithm for global code motion of congruent instructions"
extends global code motion with a register-aware cost model.
The primary trick seems to be to consider the life times of values (virtual registers)
as a surrogate for (hardware) register pressure.
The heuristic is to hoist or sink instructions
if more life times are shortened than extended by a motion.
The approach was implemented into LLVM and evaluated with SPEC 2006.

The work is well motivated from practical and theoretic angles
due to bug reports and related work.
The idea is nicely presented and the steps of the algorithm are described in sufficient detail.
Related work and the evaluation can be improved.

Comments for author
-------------------
The novelty of the core idea is not clear to me.
For example, "A lifetime optimal algorithm for speculative PRE"
by Jingling Xue and Qiong Cai seem to have a similar idea,
although for the related optimization PRE.
Also, Knoop and Steffen have multiple publications on code motion,
but none is discussed or even cited.

The evaluation could be improved with more rigorous setup and statistics.
The hardware should be described in more detail than "x86_64-linux machine".
At least the CPU should be described like "Intel i7-3770".
For execution time statistics at least deviations should be considered.
If deviations are practically zero that fact should be mentioned and all is well.
If they are closer to the performance changes
that must be honestly reported as well.
One step further,
testing the significance of the changes might be worthwhile (e.g. Student's T-test).
An improvement like libquantum's 8% is probably significant,
and povray's 0.05% is probably just noise,
but what about hmmer's 0.93%?
Unfortunately, this requires more extensive benchmarking than "run three times".

Overall, the paper is a nice contribution and publication worthy.
While related work and evaluation can be improved,
I believe the algorithm to be new
and the evalation demonstrating real improvements.

Insignificant nitpicking:

Line 575 mentions "reduced performance regressions with this cost model",
but I cannot find which regressions this refers to?

Line 733 seems to have a superfluous "ref".

The code examples are missing ":" at a few loops

Figure 4 should not use a line as the elements on the x-axis are independent.
A table might be sufficient?

* * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *

Review #66C
===========================================================================

Overall merit
-------------
1. Reject

Paper summary
-------------
This manuscript intent is to describe the author's implementation of
global code motion (GCM) in LLVM with what they claim to be a new
technique to take register pressure into consideration.

Comments for author
-------------------

While this manuscript is ok as an experience report in LLVM, the style
in which it is written makes it a very poor contribution to the
literature of code optimization. The contribution is described as a
mashing up of important existing techniques but with no concern to
explaining them at any level or even to provide definitions for the
most important concepts in the paper. Here are some specific examples:

- the concepts of "congruent values" or "congruent expressions" or
  "congruent instructions" are never defined.
- The concept of a Barrier is never precisely defined. What does it
  mean for an instruction to "not guarantee progress"? (section 3.2)
 - What does it mean for a value to have "multiple occurrences in a
  function"? (end of section 3.1). Is it multiple uses? multiple
  definitions? Does an use inside a loop counts as multiple uses?
 - What is a "landing pad" (section 3.3)?
- Precision: "no use of the instruction" (section 3.3). Instructions
  are not used. The value computed by an instruction is used.
 - "Anticipable" (last parag. of Section 3.3) is not defined.
 - The key insight in the whole manuscript appears split between the
  last lines of page 5 and first lines of page 6 --- it is the idea of
  examining opearnd that are killed by the instructions that will be
  hoisted. This insight should be more prominent in the manuscript.
  - there is an inconsistency between the text before and after the
  code on the left column of page 6: "as long as there is one operant
  killed at most" and "when the use opearnds are not kills"?
  - how is the ranking of a code hoisting opportunity defined? (end of
  page 6)
  - I do not now what is a "dependency of a hoistable candidate" is it
  an instruction that generates a value that is used by the hoistable
  candidate? Or is it an instruction that uses a value that is defined
  by the hoistable candidate? This type of imprecision in the text
  makes it very difficult for other to understand or try to reproduce
  the technique described int the manuscript.

Experimental Evaluation:

- In the experimental evaluation, what is the rational to report the
  best result of three runs? What was the methodology used to execute
  the experiments? Was the order of execution of the experiments
  randomized or were they executed in some fix order?

- The machine used for the experiments is described as "an
  x86_64-linux" machine? The experimental machine must be described in
  more detail. Does this processor has CPU frequency scaling
  (https://wiki.archlinux.org/index.php/CPU_frequency_scaling and
  https://www.kernel.org/doc/html/v4.13/admin-guide/pm/cpufreq.html)?
  If it does and this feature was not turned off, all the reported
  results may have been severely influenced by this feature.

- It does not seem to make sense to report the result only with loop
  unrolling and inlining disabled in Table 1. If these code transformations
  interfere with the GCM optimization, there is even more reason to
  report the performance results with and without these code
  transformations.

- The loop unrolling algorithm was disabled in both versions of LLVM
   (with new GMC optimization and without) because the results "vary
   wildly with GCM optimization". If the implementation is producing
   unpredictable results when a common optimization is applied to the
   code, then these results should be investigated carefully to
   determine what is causing these variations. Moreover, it would be
   valuable to compare with the performance of the unmodifed LLVM
   compiler.

- The manuscript claims that there is an increase in code size of the
   447.dealII benchmark and attribute this increase to "it becoming
   cheaper to do some optimizations like inlining, loop unrolling,
   copy...." but earlier it was stated that loop unrolling is disabled.
   It is not clear if the code was compiled with loop
   unrolling enabled to evaluate the code size variations.

- It is very strange that the impact on compilation in Table 4 is
  reported as changes in number of instructions executed during
  compilation. Anyone using the LLVM compiler will be most interested
  in the compilation time measured in seconds.

- Figure 4 does not make much sense. Why is this a line graph? A line
  graph should be use when there is an order between the points in the
  horizontal axis. In this case there is no order. Also, the vertical
  axis of the plot have no label or unit. Also, how is the
  "GrandTotal" computed for each of the metrics shown in that plot?

Editorial issues:

Large algorithms without labels in sections 3.1, 3.2, 3.3, 3.4.1,
4. Ideally algorithms should be typeset as figures with captions so
that the text can refer to them (LaTeX provides several packages for
typesetting algorithms). Also, most of the algorithms do not provide
a clear description of what input each algorithm expects and what
output it generates.

Line 314: following with the names within parenthesis is an unusual
sentence construction in English.

Line 531: "post-dominaing \phi to use"...

Line 554: "how GCM of can reduce"

Line 587: "Due to this" This what?

Line 604: "structuers"

Line 702: "is linear in (the) number"

Line 723: Cpu -> CPU (also many other places)

Line 732: "compile time (as shown in) Table 4."

Rebuttal questions
------------------

Thank you for the rigorous review of the paper. 
General response:
- The CPU used was Intel(R) Xeon(R) CPU E5-2660 v2 with frequency scaling disabled and the benchmark was run with the standard reportable configuratino where 'test runs' were followed by 'ref runs'.
- Some benchmarks regressed w.r.t. code size even if loop unrolling and inlining was disabled because other optimizations became cheaper like versioning of code, vectorization, copy-propagation, if-conversion etc. Some scheduling and register allocation decisions in the compiler backend could also change because of GCM.
- Because many optimizations in LLVM are still not mature for example: constant propagation may increase register pressure, but instead of computing the affect on register pressure, the optimization algorithm may use a heuristic to propagate values only so far (say next basic block only). Similarly for if-conversion, loop unrolling, inliner etc. have approaches based on separate cost-models. That would limit some useful gains which our algorithm could have shown.
- The compilation time was reported in terms of valgrind instructions because it was deterministic. We could report the number with other benchmarks but because we were sticking to SPEC we did not want to explore the details of its build configuration.
- Figure 4 is mostly to show the relative comparison of different kinds of instructions scheduled.

Following are the responses to specific questions:

@Reviewer1:
- Theoretically GCM would improve vectorization but we haven't explored a case in those benchmarks. For example, consider a loop:

for (int i = 0; i < N; i++) {
  if (p) a = b+c;
  else e = b+c;
  //...
}

The conditional within a loop may prevent vectorization. PRE may not remove this because there is no partial redundancy.  Loop Invariant Motion cannot remove this because none of the computations of 'b+c' dominate the overall execution of the loop.
GCM would hoist 'b+c' and then the if-else can be converted to a csel (conditional select) thus enabling vectorization. In some cases the conditional can be completely removed because of GCM.

- The GCM does not have access to profile information so the scheduling is oblivious to hotspots. We have not investigated which code-motion caused a specific benchmark to improve.

- GCC's GCM is implemented in the backend (RTL) level so any such comparison will be imprecise. The algorithms used there is quite different because of level of abstractions both are dealing with. Our cost-model is geared towards reducing code-size and reducing register pressure while GCC's cost-model is classical, improving ILP with minimal impact to register pressure.

-Table2:
Because inliner (in LLVM) relies on instruction counts, some functions can get chance to be inlined thereby preventing subsequent inlining. e.g. consider a callgraph with inlinecost for each A(100)->B(20)->[C(125), D(90)] and if the inliner inlines functions with a maximum cost of 120. Without GCM, D will get inlined in B, and then B will still get inlined in A. With GCM, if C reduced in code size and its cost became, say 110, only C will get inlined in B preventing the inlining of B.

@Reviewer2:
The algorithm relies on SSA so early references to code-motion in dataflow model cannot exploit the kind of abstraction we have explained in the paper. There are plenty of papers related to GCM but none in SSA and realizing that code-hoisting or code-sinking can be performed with an algorithm similar to PHI-insertion. Also, the cost model to compute the difference in liveness (before and after GCM) is novel. The paper by Jingling Xue et al. is based on bit-vectors and is computationally an order of magnitude worse in terms of time and no better than us in space.

@Reviewer3:
- We should have defined those terms, I apologize.
- Multiple occurrence in a function referred to multiple expressions with same value number.
- Congruency is with respect to global values as mentioned in [Briggs 1997].
- When an instruction does not guarantee progress, that means that instruction may cause an exception and subsequent instructions are not guaranteed to execute. With limited information available statically, call instructions which may not return, some memory operations etc. are assumed to not guarantee progress.
- The place where an invocation continues after exception (https://llvm.org/docs/ExceptionHandling.html) is called landing pad. In simple terms, it is a basic block which is the target of an indirect branch.
- There is no inconsistency, the first argument discusses when there is at least one use which is a kill, then sinking will not increase the register pressure. The second argument says, if none of the use operands are kills then register pressure may reduce (for example: a<def> = b+c, and b, c are not killed here then sinking this will reduce the live-range of a).
- Ranking is defined in page 2 second column.
- Dependency of hoistable candidate: instructions defining the operands used by a hoistable instruction.

### 2018-02-21 cgo/paper.tex
other files
remove refs
quotes

@@
 not found any reference which explored \gcm{} with a cost-model to reduce
-register-pressure. A part of our implementation (aggressive code hoisting) is
-already merged in \LLVM{} trunk. The main contributions of this paper are:
+register-pressure. The main contributions of this paper are:
 \begin{itemize}[leftmargin=*,topsep=0pt]
@@
 reduces. The compilation time listed here is the total instruction count
-executed during the compilation of each benchmark as output from valgrind
--{}-tool=cachegrind \cite{valgrind}. We used valgrind because it was a simple
+executed during the compilation of each benchmark as output from `valgrind
+-{}-tool=cachegrind' \cite{valgrind}. We used valgrind because it was a simple
 and reliable way to show the impact of our optimization.
@@
 and not hoist/sink too much, we have implemented a register pressure aware cost
-model as described in Section~\ref{subsec:cost-models}. A part of our
-implementation is merged in LLVM trunk as GVNHoist.cpp and the \GCM{}
-implementation is under review. This pass does not hoist fully redundant
-expressions because they are already removed by a later partial redundancy
-elimination pass in LLVM.
+model as described in Section~\ref{subsec:cost-models}. This pass does not hoist
+fully redundant expressions because they are already removed by a later partial
+redundancy elimination pass in LLVM.
 

### 2018-02-23 cgo/paper.tex
cc comments

@@
 
+\renewcommand\lstlistingname{Algorithm}
+
 \begin{document}
@@
 \def \GVN {GVN}
-\def \SPEC {SPEC Cpu 2006}
+\def \SPEC {SPEC CPU 2006}
 \def \gcm {global-code-motion}
@@
 but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
-extended basic-block because there might be non-returning calls in the middle of
-a basic-block. Essentially, any instruction that cannot guarantee progress is
-marked as a barrier. A barrier will prevent any instruction occurring lexically
-below it from getting hoisted. On the other hand, a sink barrier will prevent
-any instruction above it from getting sunk. If there are no barriers in a basic
-block, any instruction satisfying other legality checks can be sunk.
-
-\begin{lstlisting}
+extended basic-block \cite{ebb} because there might be non-returning calls in
+the middle of a basic-block. There might be other instructions like stores,
+where compiler may not guarantee progress of control flow, all such instructions
+are marked as barriers. A barrier will prevent any instruction occurring
+lexically below it from getting hoisted. On the other hand, a sink barrier will
+prevent any instruction above it from getting sunk. If there are no barriers in
+a basic block, any instruction satisfying other legality checks can be sunk.
+
+\begin{lstlisting}[caption={Compute hoist and sink barriers},captionpos=b]
 // Compute both hoist and sink barriers
@@
 B0: i0 = load B
+    if p is true then goto B1 else goto B2
 B1: i1 = load A
@@
 
-In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
-(c1, c2) depend on i1 and i2 respectively which are also in their original basic
-blocks, c1 and c2 are not fully anticipable in B3. So without knowing the
+In this example (c1, c2) or (d1, d2) are potential sinkable candidates.  (c1,
+c2) depend on i1 and i2 respectively which are also in their original basic
+blocks; c1 and c2 are not fully anticipable in B3. So without knowing the
 ability to sink `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the
 other hand (d1, d2) can safely be sunk because their operands are readily
-available at the sink point, i.e., B3. It should also be noted that, just
-because the expressions are identical and operands are available, it still
-requires a unique post-dominating $\Phi$ to use the exact same values to be
-legally sinkable.
+available at the sink point, i.e., B3. It should also be noted that, even if the
+expressions are identical and operands are available, it still requires a both
+instructions to be post-dominated by a unique $\Phi$ to be legally sinkable.
 
@@
 affect runtime performance, e.g., impact on live-range, gain in the code
-size. The current implementation makes effort to not regress in performance but
-reduce register spills. We implement a cost model to reduce register pressure.
+size. The current implementation makes effort to not regress in performance and
+still reduce register spills. We implement a cost model to reduce register
+pressure.
 
@@
 killed operands. If the live-range after \GCM{} will decrease, it will be
-moved. Essentially, as long as there is one killed operand, code hoisting will
-either decrease or preserve the register pressure.  Similarly, code-sinking will
-either decrease or preserve the register pressure as long as there is one
-operand killed at most.  The following example explains how \GCM{} of can reduce
-the register pressure.
+moved. Essentially, if there is one killed operand, code hoisting will either
+decrease or preserve the register pressure as long as the hoisting of killed
+operands does not cross a use.  Similarly, code-sinking (of an instructions
+which defines a register) will either decrease or preserve the register pressure
+as long as there is one operand killed at most and the instruction is not sunk
+below the killing use of a non-killed operand. The following example explains
+how \GCM{} can reduce the register pressure.
 
@@
 pass like \GCM{}. Even hoisting scalars across calls is tricky because it can
-increase the number of spills. During the register allocation of calls
+increase the number of spills. During the register allocation of call
 expressions, the argument registers (also known as the caller saved registers)
 are saved because they might be modified by the callee and after the call they
-are restored. Due to this the register pressure is high because the number of
-available registers are reduced by the number of caller saved registers. In that
-situation a computation that reduces register pressure may be beneficial to
-hoist. Similarly, a computation that introduces a register (without killing any
-other operand) before a call may be beneficial to sink.
+are restored. Due to this frame setup the register pressure is high because the
+number of available registers are reduced by the number of caller saved
+registers. In that situation a computation that reduces register pressure may be
+beneficial to hoist. Similarly, a computation that introduces a register
+(without killing any other operand) before a call may be beneficial to sink.
 
@@
 instructions which needed to be rematerialized. Thereafter, all the computations
-made redundant by the new copy are removed and affected data structuers are
+made redundant by the new copy are removed and affected data structures are
 updated.
 
-\begin{lstlisting}
+\begin{lstlisting}[caption={Global Code Motion},captionpos=b]
 void doGCM()
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption={Hoist instructions},captionpos=b]
 int doCodeHoisting(VNs)
@@
 
-  for each VN in VNs
-    for each instruction with the same VN
+  for each VN in VNs:
+    for each instruction with the same VN:
       Find its post-dominance-frontier (PDF)
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption={Sink instructions},captionpos=b]
 int doCodeSinking()
@@
 
-  for each VN in VNs
-    for each instruction with the same VN
+  for each VN in VNs:
+    for each instruction with the same VN:
       Find its dominance-frontier (DF)
@@
 \subsection{Time complexity of algorithm}
-The complexity of code hoisting is linear in number of instructions that are
+The complexity of code hoisting is linear in the number of instructions that are
 candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.  The
@@
 Table~\ref{tab:stats}, Table~\ref{tab:spills}, and
-Figure~\ref{tab:code-motion-metric}.  We also measured compile time
+Figure~\ref{tab:code-motion-metric}.  We also measured compile time as shown in
 Table~\ref{tab:compile-time}. The benchmark 464.h264ref caused an internal

### 2018-02-23 cgo/paper.tex
sinkable contents

@@
 where compiler may not guarantee progress of control flow, all such instructions
-are marked as barriers. A barrier will prevent any instruction occurring
-lexically below it from getting hoisted. On the other hand, a sink barrier will
-prevent any instruction above it from getting sunk. If there are no barriers in
-a basic block, any instruction satisfying other legality checks can be sunk.
-
-\begin{lstlisting}[caption={Compute hoist and sink barriers},captionpos=b]
+are marked as barriers as illustrated in Algorithm~\ref{algo:barriers}.
+
+\begin{lstlisting}[caption={Compute hoist and sink barriers},captionpos=b,label={algo:barriers}]
 // Compute both hoist and sink barriers
@@
 
-In the absence of context, as in our current implementation, some instructions
-which might be safe are still classified as barriers, e.g., calls with missing
-attributes.  Computing barriers allows for efficient removal of non-movable
-candidates, this makes the legality check converge faster.
+A barrier will prevent any instruction occurring lexically below it from getting
+hoisted. On the other hand, a sink barrier will prevent any instruction above it
+from getting sunk. If there are no barriers in a basic block, any instruction
+satisfying other legality checks can be sunk.  In the absence of interprocedural
+context, as in our current implementation, some instructions which might be safe
+are still classified as barriers, e.g., calls with missing attributes.
+Computing barriers allows for efficient removal of non-movable candidates, this
+makes the legality check converge faster.
 
@@
 
-\begin{lstlisting}[caption={Global Code Motion},captionpos=b]
+\begin{lstlisting}[caption={Global Code Motion},captionpos=b,label={algo:gcm}]
 void doGCM()
-  Analyses available:
-    Dominator Tree, DFS Numbering, Memory SSA
-  Compute DJ-graph of function
+  Analyses available: Dominator Tree, DFS Numbers, Memory SSA
+  Compute DJ-graph of CFG
   constructMergeSet(CFG)
@@
   while (doCodeHoisting() > 0)
-  // Code-sinking only once
-  doCodeSinking()
+  doCodeSinking() // Code-sinking only once
 \end{lstlisting}
 
-\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
+\GCM{} (Algorithm~\ref{algo:gcm}) basically consists of two parts, i.e., hoisting
+(Algorithm~\ref{algo:hoist}) and sinking (Algorithm~\ref{algo:sink}). This
 implementation only moves congruent instructions which makes it easier to ensure
@@
 
-\begin{lstlisting}[caption={Hoist instructions},captionpos=b]
+\begin{lstlisting}[caption={Hoist instructions},captionpos=b,label={algo:hoist}]
 int doCodeHoisting(VNs)
@@
 
-\begin{lstlisting}[caption={Sink instructions},captionpos=b]
+After no more candidates are hoistable, sinking is performed. To sink the
+instructions $\Phi$ (Section~\ref{fig:phi-intro}) nodes are considered as they
+keep track of expressions which are available at the point of their definition.
+Sinking is only performed once because there were far fewer sinkable
+instructions than hoistable instructions during our experiments.
+
+Moreover, as the direction of movement of sinkable instruction is in the same
+direction as program flow, the sinkable instructions which are dependent on
+other sinkable instructions can be sunk in the same iteration as algorithm
+iterates bottom up.
+
+\begin{lstlisting}[caption={Sink instructions},captionpos=b,label={algo:sink}]
 int doCodeSinking()
@@
 
-After no more candidates are hoistable, sinking is performed. It is only
-performed once because it may not open new opportunities for sinkable
-candidates. For sinking we iterate on the $\Phi$ nodes. After each code-motion
-the \SSA{} form is restored by updating the intermediate representation (IR) to
-reflect the changes. At the same time \MemorySSA{} is updated as well. Both the
-data structures can be updated very efficiently. After one full iteration of
-hoisting, the \GVN{} needs to be recomputed to explore new hoistable candidates.
-Finally after the transformation is done, we verify a set of post-conditions to
-establish that program invariants are maintained: e.g., consistency of use-defs,
-and \SSA{} semantics.
+After each code-motion the \SSA{} form is restored by updating the
+intermediate representation (IR) to reflect the changes. At the same time
+\MemorySSA{} is updated as well. Both the data structures can be updated very
+efficiently. After one full iteration of hoisting, the \GVN{} needs to be
+recomputed to explore new hoistable candidates.  Finally after the
+transformation is done, we verify a set of post-conditions to establish that
+program invariants are maintained: e.g., consistency of use-defs, and \SSA{}
+semantics.
 

### 2019-03-22 paper.tex
correction
Update results for OpenCV

@@
 \usepackage{algpseudocode}
+\usepackage[export]{adjustbox}
 
@@
     c = n
-    if c is true then goto B3 else goto B4
+    if c is true then goto B1 else goto B2
 
@@
 
+We compiled a popular open source computer vision library, OpenCV.
+We used OpenCV performance tests and scripts provided by them to measure performance gain with \gcm{}.
+We compared performance for every test in different test suites.
+We observe some gains with minor regressions.
+Feature2D test suite resulted in performance gain of $1-2\%$ without any performance regression.
+Ratio of performance with \gcm{} for different test suites is shown in Figure~\ref{fig:opencv-perf-result}.
+
+\begin{figure}[H]
+
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=0.9\linewidth, height=4cm]{calib-perf-result.png} 
+\caption{Calib3D}
+\label{fig:subim1}
+\end{subfigure}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=0.9\linewidth, height=4cm]{core-perf-result.png}
+\caption{Core}
+\label{fig:subim2}
+\end{subfigure}
+
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=0.9\linewidth, height=4cm]{dnn-perf-result.png} 
+\caption{dnn}
+\label{fig:subim3}
+\end{subfigure}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=0.9\linewidth, height=4cm]{feature-perf-result.png}
+\caption{Feature2D}
+\label{fig:subim4}
+\end{subfigure}
+
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=0.9\linewidth, height=4cm]{imgproc-perf-result.png} 
+\caption{ImgProc}
+\label{fig:subim5}
+\end{subfigure}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=0.9\linewidth, height=4cm]{objdetect-perf-result.png}
+\caption{ObjDetect}
+\label{fig:subim6}
+\end{subfigure}
+
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=0.9\linewidth, height=4cm, center]{video-perf-result.png} 
+\caption{Video}
+\label{fig:subim7}
+\end{subfigure}
+
+\caption{Ratio of performance for OpenCV perf test with \gcm{}}
+\label{fig:opencv-perf-result}
+\end{figure}
+
+We report decrease in size of libopencv\_core library by $0.17\%$.
+There is no regression observed in size of other libraries.
+We cross-compiled OpenCV for Android using Android ntk 16b.
+We report decrease in size in Figure~\ref{fig:opencv-android-size-result}.
+Notably with \gcm{} we observed $0.4\%$ decrease in size for OpenCV machine learning library libopencv\_ml.so* and it's core library libopencv\_core.so*.
+
+\begin{figure}[H]
+\includegraphics[width=0.9\linewidth, height=8cm, center]{opencv-android-size-result.png} 
+\caption{Percentage decrease in size of OpenCV libraries on Android with \gcm{}}
+\label{fig:opencv-android-size-result}
+\end{figure}
+
 The code shown in Section~\ref{sec:intro} is a reduced example that appears in a
